### 2017-08-29 PartProgNuclPhys/ppnp.tex
Updated PPNP files

\documentclass[preprint,12pt,authoryear]{elsarticle}
\usepackage{graphicx,amsmath,amssymb,bm}
\usepackage{amsfonts}
\usepackage{color}
\usepackage{float}
\usepackage{dcolumn}
\usepackage{subfigure}
\usepackage{multirow}

\newcommand{\bra}[1]{\left\langle #1 \right|}
\newcommand{\ket}[1]{\left| #1 \right\rangle}

\begin{document}

\begin{frontmatter}
\title{Beyond Hartree-Fock in Nuclear Physics}
\author{M.~Hjorth-Jensen}
\address{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan
  State University, East Lansing, MI 48824-1321, USA}
\address{Department of Physics, University of Oslo, N-0316 Oslo, Norway}
\author{G.~R.~Jansen}
\address{National Center for Computational Sciences and Physics Division,, Oak Ridge National Laboratory,
Oak Ridge, TN 37831, USA}

\begin{abstract}
To get a proper handle on correlations in nuclear systems and relate
these to the underlying forces, forms an essential part in basic
nuclear physics research and has wide ranging implications for our
understanding of how subatomic matter organizes itself and what
phenomena emerge. To identify the mechanisms which are
responsible for the occurence or not of so-called magic numbers in
various regions of the nuclear chart, requires systematic studies of
correlations and nuclear forces.
In this work we analyze the evolution of effective
   single-particle energies for the chain of oxygen, calcium, nickel
   and tin isotopes and study the role of two- and three-body forces
   in terms of their central, spin-orbit and tensor force components.
   The aim is to understand the role of these components towards the
   respective driplines of the above isotopic chains. Nuclear forces
   based on chiral effective field theory and systematically fit to
   nuclear data are employed in these studies.

\end{abstract}

\begin{keyword}
\PACS 21.10.-k \sep 21.10.D \sep 21.30.Fe \sep 21.60.De

\end{keyword}

\end{frontmatter}

\section{Introduction}\label{sec:intro}

The understanding of nuclear structure and reactions and thereby
properties of nuclei in terms of the underlying strong force and the
pertinent laws of motion, involves a number of approximations that
pose a great challenge to nuclear theory.  Similarly, our interpretations of
experimentally measurable quantities rely as well on several assumptions and
approximations.  As an example, concepts such as an independent
particle motion and various mean-field approaches based thereupon,
play an important role in the analysis and interpretation of
experimental results. Eventual deviations from such a mean-field
picture are interpreted as a possible measure of correlations, with
the potential of providing a better understanding of the role of the
underlying strong force in a nuclear many-body environment.  A popular
theoretical approach for studying correlations beyond a given mean field, is
the nuclear shell model, see for example
Refs.~\cite{lawson1980,talmi1993,caurier2005}. It rests on the assumption that the
state functions used in nuclear structure studies can be approximated
by various Slater determinants based on a particular orthonormal set of
single-particle states. The shell model has been rather successful in
describing many of the observed features of nuclei and experimental
probes like nucleon transfer and knockout reactions have been
essential in supporting a mean-field picture of
nuclei. Single-particle properties have been extracted by measurements
of nucleon-adding and nucleon-removing transfer reactions,
establishing thereby a link between observables and theoretical
interpretations, see for example
Refs.~\cite{schiffer2012,schiffer2013} and discussions therein.

However, in nuclear physics, the analysis and theoretical
interpretations of correlations beyond a mean field picture are
complicated by the fact that the strong force is represented by
various effective models and Hamiltonians in the energy regime of
nuclear structure studies. These effective Hamiltonians are nowadays
based on  chiral effective field theory
(EFT),  see for example
Refs.~\cite{weinberg1990,weinberg1991,ordonez1992,ordonez1994,ordonez1996,vankolck1999,machleidt2011,epelbaum2009,ekstrom2013}.  
The EFT nuclear interactions that are currently
employed in nuclear structure calculations employ normally pions and nucleons
as effective degrees of freedom. Compared
with former interaction models inspired by a one-boson-exchange
approximation (see Refs.~\cite{machleidt1989,machleidt2001} for
relevant reviews), interactions based on chiral EFT exhibit currents
that are consistent with the underlying Lagrangian that defines the
interactions. Furthermore, chiral EFT interactions allow for systematic
improvements since they are based on power counting in terms of the
probed momentum scale over a given cutoff scale $\Lambda$.  The power counting
introduces also a systematic recipe for constructing nucleon-nucleon
($NN$) forces, three-nucleon ($3N$) forces, and forces of higher rank.

To get a proper handle on correlations in nuclear systems and relate
these to the abovementioned underlying forces, forms an essential part
in basic nuclear physics research and has wide ranging implications
for our understanding of how subatomic matter organizes itself and
what phenomena emerge. For example, to identify the mechanisms which
are responsible for the occurence or not of so-called magic numbers in
various regions of the nuclear chart, will most likely require
systematic studies of correlations beyond standard shell-model
calculations, which often are limited to one or at most two major
oscillator shells.  For exotic isotopes close to the so-called drip
lines, continuum degrees of freedom and more complicated many-body
forces may become important, as demonstrated recently in
Refs.~\cite{hagen2012a,hagen2012b} for the chains of oxygen and
calcium isotopes.

The $NN$ interaction has traditionally been studied
in terms of its central, spin-orbit and tensor force.  These terms
accomodate to a large degree our phenomenological knowledge of the
strong interaction, which, when applied in a nuclear many-body
context, is subjected to various degrees of renormalization.  In
particular, the tensor force, a non-central component of the nuclear
force, has been shown to play an important role for the development of
magic numbers of nuclei with large $N/Z$ ratios, see
Refs.~\cite{otsuka2001,otsuka2005,otsuka2010b,tsunoda2011,smirnova2010}.
Furthermore, the spin-orbit force which arises from the nuclear
forces, is of the order of the average binding energy per
nucleon and plays an important role in defining nuclear
single-particle fields. Its importance is essential in order to account
for magic numbers and large shell gaps in nuclei and lead, in the
early days of nuclear physics, to the introduction of an empirical
one-body spin-orbit force \cite{mayer1949,haxel1949}. To relate such
an empirical spin-orbit force to the underlying nuclear forces is an
unresolved and outstanding problem in nuclear physics, see for example
Refs.~\cite{ando1981,burgunder2014}.  In
Refs.~\cite{ando1981,pieper1993}, the large splittings in energy
between states that can be interpreted as representing single-particle
spin-orbit partners in for $A=15$ and $A=39$, could be related both to
the two-body spin-orbit force and a contribution to the spin-orbit
force from three-body forces. In Ref.~\cite{pieper1993}, the authors
showed that the two-nucleon spin-orbit force gave approximately half
of the observed splitting, while the other half came from pion
exchange interactions between three or more nucleons.

In order to shed light on these issues, we present in this work an
analysis of effective single-particle energies and their evolution as
function of neutron numbers for selected oxygen, calcium, nickel and
tin isotopes. The interactions which define the single-particle
energies contain two-body and three-body forces defined within the
framework of chiral effective field theory.  We optimize the chiral
interaction up to next-to-next-to leading order (NNLO) using the
recently developed model-based, derivative-free algorithm POUNDerS
\cite{taoman}.  The cutoff dependence of the chiral forces and the
various optimizations allow us in turn to study the convergence
properties of chiral forces in a nuclear many-body
environment. Furthrmore, these interactions are in turn decomposed in
terms of a central, spin-orbit and tensor force
component~\cite{elliott1968,kirson1973,brown1988,osnes1992,richter1991,smirnova2010}. This
decomposition allows thereby for a systematic analysis of the
spin-orbit force and the tensor force as functions of varying neutron
and proton numbers.  The single-particle energies for the above
isotopes are computed using Hartree-Fock theory, many-body
perturbation theory and coupled-cluster theory
\cite{shavittbartlett2009}. The effective single-particle energies are defined
as one particle (protons and neutrons) on top of nuclei with
$j$-filled shells. Although this definition of an effective single-particle
energy is questionable, see for example the discussion in
Ref.~\cite{duguet2012}, it has been widely used in a 
nuclear structure, see for example
Refs.~\cite{otsuka2005,otsuka2010b,smirnova2010,nowacki2012,smirnova2012,sorlin2008}.

After these introductory remarks, we present in the next section our
theoretical framework, with the definition of effective
single-particle energies and their relation to the decomposition of
the nuclear forces in central, spin-orbit and tensor components. The
nuclear interactions we include in our analysis are only briefly
reviewed here. Where details are needed, we refer the reader to our
recent parametrization of chiral forces in
Ref.~\cite{carlsson2014}. Section \ref{sec:results} presents our main
findings while Sec.~\ref{sec:conclusions} outlines our conclusions
and perspectives.

\section{Theoretical framework}\label{sec:formalism}

In this section we present first our basic definitions of Hamiltonians
with and without three-body forces and link these expressions with the
derivation the so-called monopole interaction and its connection with
single-particle energies at a mean-field level. The interactions are
in turn analyzed in terms of a multi-component expansion of the
nuclear forces, with an emphasis of the central, spin-orbit and tensor
components. These terms accomodate our basic phenomenological
knowledge of the strong interaction. We end this section with a
summary on how to parametrize chiral interactions. These interactions,
which are constrained to reproduce several nuclear observables via an
optimization procedure, form the input to our analysis.

\subsection{Definitions}\label{subsec:definitions}

Our Hamiltonian contains one-body, two-body and three-body contributions and in 
the equations below, we  label states below the Fermi level $F$ as $i,j,\ldots$ while states 
above the Fermi level are defined by $a,b,\ldots$. General single-particle states
are given by the letters  $p,q \dots$. The
quantities $pq\dots$ represent the quantum numbers of various
single-particle states, namely
$p=(n_p,l_p,j_p,m_{j_p},t_{z_p})$.   
The commutation relations for creation and annihilations operators with respect to
a given reference state are then given by
\[
   \left\{a_p^\dagger, a_q \right\}= \delta_{pq}, p, q \leq F \hspace{0.5cm} \left\{a_p, a_q^\dagger \right\} =
        \delta_{pq}, p, q > F.
\]
The action of the creation and annihilation operators with respect to a reference state $\Phi_0$ are then given by
 $a_i|\Phi_0\rangle = |\Phi_i\rangle$ where a state labeled by $|\Phi_i\rangle$ means that a particle in a 
single-particle state $i$ has been removed. Similarly, we have $a_a^\dagger|\Phi_0\rangle = |\Phi^a\rangle$, $a_i^\dagger|\Phi_0\rangle = 0$ 
and $a_a|\Phi_0\rangle = 0$.  With the above definitions, we write our Hamiltonian as
\[
\hat{H}=\hat{H}_0+\hat{V}+\hat{W},
\]
where 
the single-particle part is given by
 \[
    \hat{H}_0 = \sum_{pq} \langle p|\hat{h}_0|q\rangle a_p^\dagger a_q.
 \]
This part of the Hamiltonian is commonly defined in terms of some external potential like the three-dimensional 
harmonic oscillator or a particular mean-field basis. 
Similarly, the two-body part of the Hamiltonian
is given by  
\[
  \hat{V} = \frac{1}{4}\sum_{pqrs} \langle pq|\hat{v}|rs\rangle_{\mathrm{AS}} a_p^\dagger a_q^\dagger a_s a_r
\]
where we have employed antisymmetric matrix elements defined as
\[
\langle pq|\hat{v}|rs\rangle_{\mathrm{AS}}=\langle pq|\hat{v}|rs\rangle-\langle pq|\hat{v}|sr\rangle.
\]
We will assume throughout this work that the  two-body operator $\hat{v}$ is given by a nucleon-nucleon interaction.
The models for the two-nucleon interaction will be defined in Sec.~\ref{subsec:forcemodels}. 
Finally, the three-body part of our Hamiltonian operator  is defined by   
\[
   \hat{W} =\frac{1}{36} \sum_{pqrstu} \langle pqr|\hat{w}|stu\rangle_{\mathrm{AS}}a_p^\dagger a_q^\dagger a_r^\dagger a_u a_t a_s,
\]
where we have defined the antisymmetric matrix elements
\[
          \langle pqr|\hat{w}|stu\rangle_{\mathrm{AS}}= \langle pqr|\hat{w}|stu\rangle + \langle pqr|\hat{w}|tus\rangle + \langle pqr|\hat{w}|ust\rangle- \langle pqr|\hat{w}|sut\rangle - \langle pqr|\hat{w}|tsu\rangle - \langle pqr|\hat{w}|uts\rangle.
\]
We will in the discussions to come drop the $\mathrm{AS}$ subscript,
assuming thereby that all matrix elements are antisymmetrized.
Introducing a reference state $|\Phi_0\rangle$ as our new vacuum state
leads to a redefinition of the Hamiltonian in terms of a constant
reference energy $E_0$ defined as
\[
E_0 = \sum_{i\le F}\langle i | \hat{h}_0|i\rangle+\frac{1}{2}\sum_{ij\le F} \langle ij|\hat{v}|ij\rangle +\frac{1}{6}\sum_{ijk\le F} \langle ijk|\hat{w}|ijk\rangle,
\]
and a normal-ordered Hamiltonian 
\[
  \hat{H}_N=\sum_{pq}\langle p|\tilde{f}|q\rangle a^\dagger_p a_q+\frac{1}{4} \sum_{pqrs} \langle pq|\tilde{v}|rs\rangle a^\dagger_p a^\dagger_q a_s  a_r +\frac{1}{36} \sum_{\substack{pqr \\stu}}
                 \langle pqr|\hat{w}|stu\rangle a^\dagger_p a^\dagger_q a^\dagger_r a_u a_t a_s 
\]
where
\[
\langle p| \tilde{f}|q\rangle = \langle p|\hat{h}_0|q\rangle +\sum_{i\le F} \langle pi|\hat{v}|qi\rangle+\frac{1}{2}\sum_{ij\le\alpha_F} \langle pij|\hat{w}|qij\rangle,
\] 
represents a correction to the single-particle operator $\hat{h}_0$ due to contributions from the nucleons below the Fermi level.
The two-body matrix elements are now modified in order to account for medium-modified contributions from the three-body interaction, resulting in
\begin{equation}
\langle pq|\tilde{v}|rs\rangle=\langle pq|\hat{v}|rs\rangle+\sum_{i\le F} \langle pqi|\hat{w}|rsi\rangle.
\label{eq:twobodyveff}
\end{equation}
In Eq.~(\ref{eq:twobodyveff}), the effective two-body interaction
$\tilde{v}$ can contain both a standard two-nucleon interaction and a
density dependent contribution stemming from a three-body interaction
$\hat{w}$.

An important ingredient in studies of effective interactions and their
applications to nuclear structure, is the so-called monopole
interaction, normally defined in terms of a nucleon-nucleon
interaction $\hat{v}$ \cite{otsuka2001,otsuka2005,otsuka2010b,smirnova2010,nowacki2012,smirnova2012,sorlin2008,poves1981a,poves1981b,zuker2003}
\begin{equation}
\bar{V}_{\alpha_p\alpha_q} = \frac{\sum_{J}(2J+1) \langle (\alpha_p\alpha_q)J | \hat{v} | (\alpha_p\alpha_q)J \rangle }{\sum_{J}(2J+1)},
\label{eq:monopole1}
\end{equation}
where the total angular momentum of a two-body state $J$ runs over all
possible values.  
In the above equation we have defined a nucleon-nucleon interaction in a so-called angular-momentum 
coupled representation with the symbol $\alpha_{p,q}$ representing all possible quantum numbers except the 
magnetic substates $m_{j_{p,q}}$
The monopole Hamiltonian can be interpreted as an angle-averaged matrix element.  
We have assumed that the single-particle angular momenta $j_p$ and $j_q$ couple to a total
two-particle angular momentum $J$.  The summation over $J$ with the value $2J+1$ can be replaced
by $\sum_J(2J+1)=(2j_p+1)(2j_q+1)$ if $\alpha_p\ne \alpha_q$. If $\alpha_p=\alpha_q$ we can generalize this equation to, assuming that 
our states can represent either protons or neutrons,
\begin{equation}
\sum_{J}(2J+1)=(2j_p+1)(2j_q+1-\delta_{\alpha_p\alpha_q}).
\label{eq:jvalue}
\end{equation}
The spherical single-particle states,
provide an important ingredient for the formation of shells and
interplay between spherical configurations and deformation in
nuclei. Large shell gaps obtained from a monopole Hamiltonian are a
prerequisite to obtain certain magic numbers. 
Equation~(\ref{eq:monopole1}) can also be expressed in terms of the
medium-modified two-body interaction defined in
Eq.~(\ref{eq:twobodyveff}), that is we can have 
\begin{equation}
\tilde{V}_{\alpha_p\alpha_q} = \frac{\sum_{J}(2J+1) \langle (\alpha_p\alpha_q)J | \tilde{v} | (\alpha_p\alpha_q)J \rangle }{\sum_{J}(2J+1)}.
\label{eq:monopole2}
\end{equation}
As stated in the introduction, one of the
aims of this work is to study the role of three-body interactions in
nuclear structure, with an emphasis on the evolution of
single-particle energies.

The single-particle energy $\epsilon_p$ resulting from for example a self-consistent Hartree-Fock field, 
or from  first order in many-body perturbation theory, is given by (in an uncoupled basis)
\[
\epsilon_p=\langle p| \tilde{f}|p\rangle = \langle p|\hat{h}_0|p\rangle +\sum_{i\le F} \langle pi|\hat{v}|pi\rangle+\frac{1}{2}\sum_{ij\le\alpha_F} \langle pij|\hat{w}|pij\rangle,
\]
where we have included the three-body interaction as well.
We can rewrite this equation in an angular coupled basis ($jj$-coupled basis) as 
\begin{equation}
\epsilon_{\alpha_p}= \langle \alpha_p|\hat{h}_0|\alpha_p\rangle+\frac{1}{2j_p+1}\sum_{\alpha_i\le \alpha_F}\sum_{J} (2J+1)\langle (\alpha_p\alpha_i)J | \hat{v} | (\alpha_p\alpha_i)J \rangle,
\label{eq:spequation1}
\end{equation}
or
\begin{equation}
\epsilon_{\alpha_p}= \langle \alpha_p|\hat{h}_0|\alpha_p\rangle+\frac{1}{2j_p+1}\sum_{\alpha_i\le \alpha_F}\sum_{J} (2J+1)\langle (\alpha_p\alpha_i)J | \tilde{v} | (\alpha_p\alpha_i)J \rangle,
\label{eq:spequation2}
\end{equation}
where the first equation contains a two-body force only while
Eq.~(\ref{eq:spequation2}) includes the medium-modified contribution
from the three-body interaction as well. In
Eqs.~(\ref{eq:spequation1}) and (\ref{eq:spequation2}), we have
used a compact notation for the single-particle states, with the
symbol $\alpha_p$ etc representing all possible quantum numbers except
the magnetic substates $m_{j_p}$, that is $\alpha_p=(n_p,l_p,j_p,t_{z_p})$. The symbol $\alpha_F$ stands now for
all single-particle states up to the Fermi level, excluding again the
magnetic substates.  In the above two-body interaction matrix elements
$\langle (\alpha_p\alpha_i)J | \hat{v}(\tilde{v}) |(\alpha_p\alpha_i)J \rangle$ we have dropped additional quantum
numbers like the isospin projection. Our interactions are diagonal in
the projection of the total isospin but breaks both isospin symmetry
and charge symmetry.

Depending on the choice of single-particle Hamiltonian, the quantity
$\langle \alpha_p|\hat{h}_0|\alpha_p\rangle$ could represent the expectation value of the single-particle
kinetic energy or the eigenstate of a single-particle Hamiltonian
$\hat{h}_0$. The latter could for example be the solution of
Schr\"odinger's equation for a particle moving in a Woods-Saxon like
single-particle potential, or the widely employed harmonic oscillator
in three dimensions.

Using the definition of the single-particle energy in
Eq.~(\ref{eq:spequation1}), the definition of the monopole matrix
element in Eqs.~(\ref{eq:monopole1}) or (\ref{eq:monopole2}) and Eq.~(\ref{eq:jvalue}), we can rewrite
Eq.~(\ref{eq:spequation1}) as
\begin{equation}
\epsilon_{\alpha_{p}}=\langle \alpha_p|\hat{h}_0|\alpha_p\rangle+\sum_{\alpha_i\le \alpha_F}N_{\alpha_i}\bar{V}_{\alpha_p\alpha_i},
\label{eq:spmono1}
\end{equation}
with $N_{\alpha_i}=2\alpha_i+1$, and Eq.~(\ref{eq:spequation2}) as 
\begin{equation}
\epsilon_{\alpha_{p}}=\langle \alpha_p|\hat{h}_0|\alpha_p\rangle+\sum_{\alpha_i\le \alpha_F}N_{\alpha_i}\tilde{V}_{\alpha_p\alpha_i}.
\label{eq:spmono2}
\end{equation}

\subsection{Spin-tensor decomposition}\label{subsec:spintensor}

The effective interaction discussed in the previous subsection is a
scalar two-body operator. A general scalar two-body operator
$\hat{v}$ can be written as
\begin{equation}
\hat{v} = \sum_{k} \hat{v}_{k} = \sum_{k} \mbox{\boldmath $C^{(k)}\cdot
Q^{(k)}$},
\end{equation}
where the operators \mbox{\boldmath $C^{(k)}$} and
\mbox{\boldmath $Q^{(k)}$} are irreducible spherical tensor
operators of rank $k$, acting in spin and coordinate space,
respectively. The value of $k$ is limited to $k\le 2$ since the total eigenspin of the two-nucleon
system is either $0$ or $1$. The term with
$k=0$ refers to the central component of the two-body
operator.
The values of $k=1$ and $k=2$ are called
the vector and the tensor components, respectively. The vector term is also called the two-body spin-orbit term, although it
also contains the anti-symmetric spin-orbit term, see for example Ref.~\cite{conze1973} for further details.
Using standard angular momentum algebra it is rather straightforward to relate the matrix elements  $\hat{v}_{k}$ to
those of say $\hat{v}$ or $\tilde{v}$.

One possible decomposition of the effective interaction is to
express the $k$-th component of the interaction $\langle
(\alpha_p\alpha_q)J|\hat{v}_{k}|(\alpha_r\alpha_s)J\rangle $ in a $jj$-coupled basis, where
$\hat{v}_k$ is related to the matrix
elements $\langle (\alpha_p\alpha_q)J|\hat{v}|(\alpha_r\alpha_s)J\rangle $ (or $\langle (\alpha_p\alpha_q)J|\tilde{v}|(\alpha_r\alpha_s)J\rangle $ 
through the relation
\begin{eqnarray}
\langle (\alpha_p\alpha_q)J|\hat{v}_{k}|(\alpha_r\alpha_s)J\rangle&=&(-1)^{J}(2k+1)
\sum_{LL'SS'}\langle
\alpha_p\alpha_q|LSJ\rangle \langle \alpha_r\alpha_s|L'S'J\rangle
\left\{ \begin{array}{ccc}
		       L&S&J\\
		       S'&L'&k
		       \end{array}
		 \right\} \nonumber \\
& &\times \sum_{J'}(-1)^{J'}(2J'+1)\left\{ \begin{array}{ccc}
		       L&S&J'\\
		       S'&L'&k
		       \end{array}
		 \right\}
\sum_{\alpha_{p}'\alpha_{q}'\alpha_{r}'\alpha_{s}'}\langle \alpha_p'\alpha_q'|LSJ'\rangle
\nonumber\\
& &\times \langle \alpha_r'\alpha_s'|L'S'J'\rangle \langle
(\alpha_p'\alpha_q')J'|\hat{v}|(\alpha_r'\alpha_s')J'\rangle .  \label{eq:jj}
\end{eqnarray}
The two-particle matrix elements are normalized and antisymmetrized.
A similar expression applies to the medium-modified two-body interaction $\tilde{v}$ of 
Eq.~(\ref{eq:twobodyveff}) as well.
The symbol $\langle \alpha_p\alpha_q|LSJ \rangle $ is a shorthand
for the $LS-jj$ transformation coefficient,
\[
\langle \alpha_p\alpha_q|\lambda SJ \rangle = \sqrt{(2j_{p}+1)(2j_{q}+1)(2\lambda+1)(2S+1)}
\left\{
\begin{array}{ccc}
       l_{p}&\frac{1}{2}&j_{p}\\
       l_{q}&\frac{1}{2}&j_{q}\\
       \lambda    &S          &J
\end{array}
\right\},
\]
see for example Ref.~\cite{lawson1980}. 
The transformation from an $LS$ basis to a $jj$-coupled scheme is then given by the relation
\[
|(\alpha_p\alpha_q)J\rangle  = \sum_{LS}\langle \alpha_p\alpha_q|LSJ \rangle |(\tilde{\alpha}_p\tilde{\alpha}_q)LSJ\rangle,
\] 
where the symbol like $\tilde{\alpha}_{p}$ refers to the quantum numbers in an $LS$ basis, that is
$\tilde{\alpha}_{p}=(n_{p},l_{p},s_{p},t_{z_{p}})$.

To derive Eq.~(\ref{eq:jj}), we have used the fact that the two-body
matrix elements of $\hat{v}_k$ can also be interpreted in the
representation of the $LS$-coupling scheme as in
Refs.~\cite{elliott1968,kirson1973,brown1988,osnes1992,richter1991,smirnova2010}. This
representation allows for a more direct comparison with the
nucleon-nucleon interaction, a quantity normally defined in terms of
the partial waves of the center-of-mass and relative motion system,
see for example Ref.~\cite{machleidt2011}.  Similar to the
decomposition in the $jj$-scheme, the $LS$-coupled matrix element of a
given component $k$ $\bra{(\tilde{\alpha}_p\tilde{\alpha}_q
  )LSJ'T}\hat{v}_{k}\ket{(\tilde{\alpha}_r\tilde{\alpha}_s )LSJ'T} $
are related to the corresponding matrix elements of the total
interaction in the $jj$-scheme by
\[
\begin{array}{ll}
&\\
\bra{(\tilde{\alpha}_p\tilde{\alpha}_q )LSJ'T}\hat{v}_{k}\ket{(\tilde{\alpha}_r\tilde{\alpha}_s )L'S'J'T}&=
{\displaystyle\frac{1}{\sqrt{(1+\delta_{\tilde{\alpha}_p\tilde{\alpha}_q})(1+\delta_{\tilde{\alpha}_r\tilde{\alpha}_s})}}}
(-1)^{J'}\hat{k}\left\{\begin{array}{ccc}L&S&J'\\S'&L'&k
\end{array}\right\}
\\&\\
&\times {\displaystyle\sum_{J}}(-1)^{J}\hat{J}\left\{\begin{array}{ccc}L&S&J\\S'&L'&k
\end{array}\right\}
{\displaystyle \sum_{\alpha_p \alpha_q \alpha_r \alpha_s}}
\langle \alpha_p\alpha_q|LSJ\rangle
\langle \alpha_r\alpha_s|L'S'J\rangle
\\&\\
&\times\sqrt{(1+\delta_{\alpha_p\alpha_q})(1+\delta_{\alpha_r\alpha_s})}\bra{(\alpha_p\alpha_q)JT}\hat{v}\ket{(\alpha_r\alpha_s)JT}
.\end{array}
\]

In this work we will use Eq.~(\ref{eq:jj}) in our analysis of the various matrix elements in Eqs.~(\ref{eq:monopole1}) 
and (\ref{eq:monopole2}) and the single-particle energies of Eqs.~(\ref{eq:spmono1})
and (\ref{eq:spmono2}).

\section{Results}\label{sec:results}

\begin{figure}%[hbtp]
     \begin{center}
        \subfigure[Results for the $0f_{7/2}$ single-particle state]{
            \label{fig:f72}
            \includegraphics[width=0.38\textwidth]{Figs/protonf72.pdf}
        }
        \subfigure[Results for the $0f_{5/2}$ single-particle state ]{
           \label{fig:f52}
           \includegraphics[width=0.38\textwidth]{Figs/protonf52.pdf}
        }\\ %  ------- End of the first row ----------------------%
        \subfigure[Results for the $1p_{3/2}$ single-particle state]{
            \label{fig:p32}
            \includegraphics[width=0.38\textwidth]{Figs/protonp32.pdf}
        }
        \subfigure[Results for the $1p_{1/2}$ single-particle state]{
            \label{fig:p12}
            \includegraphics[width=0.38\textwidth]{Figs/protonp12.pdf}
        }
    \end{center}
\caption{Contributions from the central, spin-orbit and tensor components to  
the proton single-particle energies of $1p0f$ single-particle states. 
The closed-shell cores are $^{40}$Ca, $^{48}$Ca, $^{52}$Ca, 
$^{54}$Ca and $^{60}$Ca. The results have been obtained with a harmonic oscillator basis and an 
oscillator energy $\hbar\omega =10.5$ MeV using the N$^3$LO interaction model of Ref.~\cite{machleidt2011}.}
\label{fig:hoprotonCa}
\end{figure}

\section{Conclusions and perspectives}\label{sec:conclusions}

\section*{Acknowledgement}
M.~H.~J. acknowledges U.S. NSF Grants No. PHY-1404159 and No. PHY-1713901 (Michigan State University).

\bibliography{paper}
%\bibliographystyle{apsrev}

\end{document}

### 2017-08-29 PartProgNuclPhys/Figs/plot.py

@@ -1,0 +1,32 @@
+#!/usr/bin/python
+from numpy import *
+from pylab import *
+from matplotlib import rc, rcParams
+import matplotlib.units as units
+import matplotlib.ticker as ticker
+import sys
+import os
+
+
+rc('text',usetex=True)
+rc('font',**{'family':'serif','serif':['New fits of nuclear forces']})
+
+data = loadtxt("protonf72.dat")
+
+l = len(data[:,0])
+
+axis([39,61,-41, -24])
+xlabel(r'$A$',fontsize=20)
+ylabel(r'$\epsilon_{f_{7/2}}$',fontsize=20)
+plot(data[:,0], data[:,1],'b-*', data[:,0], data[:,1]+data[:,2],'r:.', data[:,0], data[:,4], 'm:v',markersize=7)
+
+#title(r'Proton single-particle energies for $A=40-60$', fontsize=20, horizontalalignment='center')
+legend(('Central','Central+Spin-orbit','All terms'),
+           'upper right', shadow=False, fancybox=False,prop={"size":18})
+legend(loc='upper right')
+
+xticks( [40,42,44,46,48,50,52,54,56,58,60])
+# Save the figure in a separate file
+savefig('protonf72.pdf', format='pdf')
+# Draw the plot to screen
+show()

### 2018-02-15 doc/ElectronGas/backup.tex
Added electron gas  paper

\documentclass[aps,twocolumn,showpacs,floatfix,nofootinbib,preprintnumbers,superscriptaddress,amsmath,amssymb]{revtex4-1}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{bm}
\usepackage{color}
\usepackage{float}
\usepackage{dcolumn}
\usepackage{multirow} 
\usepackage{subfigure}

\newcommand{\Ho}{\hat{H}^{\rm od}}
\newcommand{\expec}[1]{\langle #1 \rangle}
\DeclareMathOperator\erf{erf}
\DeclareMathOperator\erfc{erfc}

\begin{document}

\title{Many-body approaches to the homogeneous electron gas in two and three dimensions}

\author{Justin G.~Lietz}
\affiliation{National Superconducting
  Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}

\author{Scott K.~Bogner}
\affiliation{National Superconducting
  Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}

\author{Morten Hjorth-Jensen} 
\affiliation{National Superconducting Cyclotron
  Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\affiliation{Department of Physics, University of Oslo, N-0316 Oslo, Norway}

\begin{abstract} 
We present results for the two- and three-dimensional homogeneous
electron gas using first principle methods like coupled cluster
theory and two recently developed methods, the in-medium Similarity Renormalization. In particular, we present
new results for the correlation energy for the two-dimensional
electron gas and analyze our results in terms of finite size effects and basis functions.
\end{abstract}

\pacs{02.70.Ss, 31.15.A-, 31.15.bw, 31.15.V-, 71.10.Ca, 71.15.-m}

\maketitle

\section{Introduction}
A proper understanding of the role of correlations beyond a mean-field
description is of great importance for quantum many-particle
theories. For all possible quantum-mechanical systems, either finite
ones or infinite systems like the homogeneous electron gas in two or
three dimensions or nuclear or neutron star matter, the concept of an
independent particle motion continues to play a fundamental role in
studies of many-particle systems.  Deviations from such an independent
particle, or mean-field, picture have normally been interpreted as a
possible measure of correlations.  The latter are expected to reveal
important features of both the structure and the dynamics of a
many-particle system beyond a mean-field approximation.

The homogeneous electron gas is one of the few examples of a system of many
interacting particles that allows for a solution of the mean-field
Hartree-Fock equations on a closed form.  To first order in the
electron-electron interaction, this applies to ground state properties
like the energy and its pertinent equation of state as well.  The
homogeneus electron gas is a system of electrons that is not
influenced by external forces except by an attraction provided by a
uniform background of ions. These ions give rise to a uniform
background charge.  The ions are stationary and the system as a whole
is neutral.
Irrespective of this simplicity, this system, in both two and
three-dimensions, has eluded a proper description of correlations in
terms of various first principle methods, except perhaps for quantum
Monte Carlo methods. In particular, the diffusion Monte Carlo
calculations of Ceperley {\em et al.}
\cite{ceperley1980,tanatar1989}, are presently still considered as the
best possible benchmarks for the two- and three-dimensional electron
gas.

The electron gas, in 
two or three dimensions is thus interesting as a test-bed for 
electron-electron correlations. The three-dimensional 
electron gas is particularly important as a cornerstone 
of the local-density approximation in density-functional 
theory. In the physical world, systems 
similar to the three-dimensional electron gas can be 
found in, for example, alkali metals and doped 
semiconductors. Two-dimensional electron fluids are 
observed on metal and liquid-helium surfaces, as well as 
at metal-oxide-semiconductor interfaces. However, the Coulomb 
interaction has an infinite range, and therefore 
long-range correlations play an essential role in the
electron gas. At low densities, the electrons become 
localized and form a lattice. This so-called Wigner 
crystallization \cite{wigner1934} is a direct consequence 
of the long-ranged repulsive interaction. At higher
densities, the electron gas is better described as a
liquid \cite{ceperley1978,ceperley1980,tanatar1989,bishop1982}. 
When using, for example, Monte Carlo methods 
\cite{foulkes2001}, the electron gas must be approximated 
by a finite system. The long-range Coulomb interaction 
in the electron gas causes additional finite-size effects 
\cite{fraser1996,chiesa2006,drummond2008} that are not
present in other infinite systems like nuclear matter or neutron star matter in 
\cite{hagen2014}. 
This poses additional challenges to many-body methods when applied 
to the electron gas.

The aim of this work is to present results for the infinite electron
gas using several many-body methods.  We focus in particular on
coupled-cluster theory (add ref), the recently develop in-medium
similarity renormalization group methods, full-configuration
interaction quantmum Monte Carlo \cite{booth2009,shepherd2012a,booth2013}
We present results for all these methods and study in particular both finite size effects and effects due to
a truncation in the single-particle basis (add more refs).

More

\section{Formalism}
\label{sec:formalism}

\subsection{Previous studies of the electron gas}

The electron gas has been studied using a large number 
of different approaches, and we will here only mention
some of the most important works that are relevant here. 
We start with the three-dimensional electron 
gas, which has got most attention in the literature.
The three-dimensional electron gas will mainly serve as a benchmark for our calculations,

It is a well-kown fact that the correlation energy
of the three-dimensional electron gas diverges 
at second order in perturbation theory. The particle-hole ring diagrams 
dominate at the limit of high density. Even though
all these diagrams diverge when calculated separately,
the energy converges when summing all direct 
particle-hole ring diagrams to infinite order.
Gell-Mann and Brueckner obtained for this so-called
random-phase approximation (RPA) the first terms of
the exact energy in the high-density limit 
\cite{gellmann1957}.

At metallic and lower densities, short-ranged correlations
become significant. Singwi \emph{et al.} \cite{singwi1968}
and Lowy and Brown \cite{lowy1975} came up with early attempts 
to combine RPA with contributions for short-ranged 
correlations. The calculations of Singwi \emph{et al.}
were based on dielectric function theory, whereas Lowy and
Brown interpolated between short- and long-range models
using a diagrammatic technique.

In 1978, Ceperley used \cite{ceperley1978} 
the variational Monte Carlo (VMC) method to 
study the two- and three-dimensional electron gas. The
obtained VMC ground-state energies were shown to be close
to other results at that time. Higher accuracy can be obtained with
the diffusion Monte Carlo (DMC) method (in nuclear physics,
this method is commonly named Green's function Monte Carlo). 
The DMC results of Ceperley and Alder from 1980 
\cite{ceperley1980} are still, more than thirty years later, 
among the most accurate energy estimates for the 
three-dimensional electron gas. In Ref.~\cite{ceperley1980},
the fermion sign problem is handled using 
a released-node approximation. Similar accuracy has been 
obtained in more recent calculations using the 
backflow-correlation technique \cite{kwon1998}, which is used 
to relax the simpler fixed-node approximation.

When modelling the electron gas using a finite box, as is
commonly done in Monte Carlo methods, the energy has an error
compared to the electron gas at the thermodynamic limit. 
At the thermodynamic limit, plane-wave single-particle states
fill the Fermi sphere with a continuous spectrum. When the
system is approximated using a finite box, the 
single-particle spectrum becomes discrete. This 
discretization gives an 
error that is common to systems with short-ranged and 
long-ranged interactions. The perhaps most obvious way to
correct for finite-size effects related to a discrete 
single-particle basis, is by using an extrapolation formula.
As is described in Ref.~\cite{drummond2008}, results from
Hartree-Fock and density-functional theory calculations 
can be used to construct extrapolation methods. Another
approach that efficiently reduces the finite-size error
is a technique that utilizes so-called twisted boundary 
conditions \cite{lin2001}. Twisted boundary conditions 
means that the single-particle wave function is 
multiplied by a complex phase factor when moving from one 
simulation cell to a neighbouring cell. When averaging
over results obtained with different twist angles, the
energy estimates become much more accurate than when
using periodic boundary conditions \cite{lin2001}.

Finite-size approximations of extended Coulombic systems, 
such as the electron gas, have additional errors that are 
caused by the long-ranged interaction \cite{fraser1996,chiesa2006,drummond2008}. 
Interactions with electrons in
neighbouring cells can be summed using, for example,
Ewald's method \cite{ewald1921,fraser1996,wood2004,drummond2008}. In 
Ewald's approach, each electron in the simulation cell
interacts with an infinite number of image charges located
at the same local position in all other cells. As is shown 
in Refs.~\cite{fraser1996,chiesa2006}, 
Ewald's interaction cannot describe the exchange-correlation
energy correctly. Chiesa \emph{et al.} \cite{chiesa2006}
used a static structure factor and a Jastrow factor
derived from RPA to estimate the correction to Ewald's
method. The correction technique of Chiesa \emph{et al.}
is directly applicable only to Monte Carlo methods.
Fraser \emph{et al.} \cite{fraser1996} suggested two 
different effective interactions that avoid the screening 
effects caused by Ewald's interaction . The most successful
alternative was to use a normal Coulomb interaction combined 
with the minimum-image convention \cite{fraser1996}.
The reader is referred to Ref.~\cite{drummond2008} for
more details about finite-size effects and different
approaches to correct for the related errors.

The full configuration-interaction quantum Monte Carlo 
(FCIQMC) met\-hod \cite{booth2009} is a new approach to the 
quantum many-body problem, in which the full 
configuration-interaction (FCI) equations \cite{harris1992} are 
solved approximately using a Monte Carlo technique. Similarly 
as in the DMC or GFMC methods, the Schr{\"o}dinger equation is 
written as a diffusion equation with an imaginary time 
variable \cite{shepherd2012a}. In the FCI method, the total 
wave-function ansatz 
is expressed as a linear combination of Slater determinants 
constructed from a given single-particle basis. In FCIQMC, 
the coefficients in the expansion of Slater determinants 
are obtained as the large-time limit of a random walk 
\cite{shepherd2012a}. As we discussed above, Monte Carlo
calculations of the electron gas have an error related to 
the finite number of particles in the simulation cell. 
Methods such as FCIQMC, CC, and partial summations derived 
from many-body perturbation theory have and additional error 
when studying systems with a discrete single-particle basis: 
The result depends on the given set of both occupied and 
unoccupied single-particle states \cite{shepherd2012b}. 
This error can be corrected for by using, for example, the 
single-point extrapolation technique \cite{shepherd2012b,
shepherd2012c} introduced by Shepherd \emph{et al.} The 
FCIQMC method has recently been applied to finite 
electron-gas systems \cite{shepherd2012a,shepherd2012b,
shepherd2012c}, giving results in close agreement
with DMC calculations utilizing backflow correlations 
\cite{shepherd2012a}.

Singal and Das \cite{singal1973} were the first to study 
the electron gas using a CC approach. The approximation
they used is similar to the BHF method \cite{haftel1970}, 
and does not properly include particle-hole ring diagrams.
Later, Freeman did CC calculations \cite{freeman1977} in 
which only ring diagrams and their exchange parts were 
retained in a CC doubles approximation. The results of both
Singal and Das \cite{singal1973} and Freeman 
\cite{freeman1977} compared well with dielectric-function
approaches. Presently, the most accurate CC caclulations of 
the three-dimensional electron gas are those of Bishop and
L{\"u}hrmann \cite{bishop1978,bishop1982}. Bishop and
L{\"u}hrmann derived a CC SUB2 approximation (also called
CC doubles (CCD) \cite{shavittbartlett2009}) for the electron gas, 
extended with some ladder contributions from higher-order
amplitudes. As the authors describe in 
Ref.~\cite{bishop1978}, the CCD approximation 
contains many more diagrammatic classes than
partial-summation techniques derived from perturbation 
theory. The CCD approximation takes account of 
particle-particle and hole-hole ladders, particle-hole
ring diagrams including exchange terms, and many other
diagrams to infinite order in perturbation theory. 
In their CC approximations, Bishop and L{\"u}hrmann 
replaced summations over hole states by averages. In
fact, the nine-dimensional CC amplitudes were simplified
to one-dimensional objects, with the absolute value
of the transfer momentum being the only variable. 
The authors showed that the state-average approximation
is accurate in the RPA approximation.
Bishop and L{\"u}hrmann neglected some diagrammatic 
classes, such as the hole-hole ladders and mixed ladders, 
which they assumed to be small. Despite all these 
approximations, the final CC energies are very accurate 
in a typical metallic density range. Relative 
differences of less than one percent
compared to DMC calculations of Ceperley and Alder
may indicate that the extended CCD approximation describes
most of the relevant correlations in the three-dimensional
electron gas \cite{bishop1982}.

The CC study of Bishop and L{\"u}hrmann was very successful, but still
the calculations are based on a large number of approximations. It
would be desireable to apply CC theory to the electron gas without the
same approximations.  Recently, Shepherd \emph{et al.}
\cite{shepherd2012b,shepherd2013a,shepherd2013b,shepherd2013c} have
taken up again the CC effort for the three-dimensional electron
gas. The CC calculations for the electron gas from the 1970s and 1980s
were all done at the thermodynamic limit
\cite{singal1973,freeman1977,bishop1978,bishop1982}. Unfortunately,
singularities related to the Coulomb two-body interaction may
complicate numerical calculations at the thermodynamic limit.
Shepherd \emph{et al.} approximate the electron gas using finite cubic
boxes
\cite{shepherd2012b,shepherd2013a,shepherd2013b,shepherd2013c}. When
using finite-size systems, the CC equations are simpler and most of
the problems with ill-bahaved functions are avoided.  Instead, one has
to deal with errors related to finite particle numbers and
single-particle bases, similarly as in the FCIQMC method
\cite{shepherd2012a,shepherd2012b,shepherd2012c}.

Let us finally turn our attention to the two-dimensional
electron gas. The two-dimensional electron gas is defined
in the same way as the three-dimensional counterpart, and 
similar approaches can therefore often be used to 
study both systems. As a first example, the classical 
derivation of the high-density RPA approximation by Gell-Mann 
and Brueckner \cite{gellmann1957} has been extended to the 
two-dimensional electron gas by Rajagopal and Kimball 
\cite{rajagopal1977}.

As far as we know, the only CC calculations that have been
done for the two-dimensional electron gas are the ring
\cite{freeman1978} and particle-particle ladder \cite{freeman1983}
approximations of Freeman. Beacuse of stronger correlations
in the purely two-dimensional system, the ring approximation
was not as reliable in two dimensions as it was in the 
three-dimensional case \cite{freeman1978}. According to the 
studies of Freeman, the CC ring-diagram approximation is 
less accurate than the CC ladder approximation at 
intermediate and low densities, whereas both methods give 
similar results at the high-density limit \cite{freeman1983}.

In the earliest CC studies of the 
electron gas \cite{singal1973,freeman1977,bishop1978,
freeman1978,bishop1982,freeman1983}, the system was treated 
at the thermodynamic limit. 
Some of the most accurate calculations of the electron gas
\cite{ceperley1980,tanatar1989,shepherd2012a} have been done using finite boxes and Monte
Carlo methods. Recently, CC theory has been applied in 
several studies \cite{shepherd2012b,shepherd2013a,
shepherd2013b,shepherd2013c,roggero2013} to finite-particle 
approximations of the homogeneous electron gas. 
 Freeman has done CC studies of the two-dimensional 
electron gas including only ring \cite{freeman1978} or 
only particle-particle ladder \cite{freeman1983} diagrams 
from the CCD approximation, but to the best of our knowledge, 
nobody has done a complete CCD calculation of the 
two-dimensional homogeneous electron gas. 
As an extension of the CC calculations by Shepherd 
\emph{et al.}~\cite{shepherd2012b,shepherd2013a,shepherd2013b,shepherd2013c}
and Roggero \emph{et al.}~\cite{roggero2013}, we study
the two-dimensional electron gas in the CCD approximation
using a finite number of particles.
We validate our methods by comparing results 
for the three-dimensional electron gas with Shepherd 
\emph{et al.} \cite{shepherd2012b,shepherd2012c}.
using momentum discretized in cartesian coordinates, see discussions below.

\subsection{Model Hamiltonian and Definitions}

The homogeneous electron gas is defined as an infinite system
consisting of electrons distributed with a constant density 
through the entire real space. The system is assumed to 
contain a constant positive background charge which cancels
the negative charges of the electrons. The Hamiltonian 
operator of the homogeneous electron gas can be written as
\cite{fetter}
\begin{align}
  \hat{H} = \hat{H}_{\text{kin}} + \hat{H}_{ee} + \hat{H}_{eb} 
  + \hat{H}_{bb},
\end{align}
where $\hat{H}_{\text{kin}}$ is the kinetic-energy operator,
$\hat{H}_{ee}$ models the electron-electron interaction,
$\hat{H}_{eb}$ represents the interaction between electrons 
and the positive background charge, and the operator 
$\hat{H}_{bb}$ gives the interaction energy of the background
charge with itself.

Similarly as in
Refs.~\cite{shepherd2012b,shepherd2013a,shepherd2013b,shepherd2013c,roggero2013},
we approximate the homogeneous electron gas by a finite hypercube
filled with electrons. The summation over
an infinite number of Coulomb interaction terms can be efficiently
calculated using, for example, Ewald's summation technique
\cite{ewald1921,fraser1996}.  In Ewald's approach, the Coulomb
interaction is split into a short-ranged part, which is calculated
in real space, and a long-ranged part, which is summed in Fourier
space.  The kinetic energy operator is
\begin{align}
  \hat{H}_{\text{kin}} = -\frac{\hbar^{2}}{2m}\sum_{i=1}^{N}
  \nabla_{i}^{2},
\end{align}
where the sum is taken over all particles in the finite
box. The Ewald electron-electron interaction operator 
can be written as \cite{drummond2008}
\begin{align}
  \hat{H}_{ee} = \sum_{i<j}^{N} 
  v_{E}\left( \mathbf{r}_{i}-\mathbf{r}_{j}\right)
  + \frac{1}{2}Nv_{0},
\end{align}
where $v_{E}(\mathbf{r})$ is the effective two-body 
interaction and $v_{0}$ is the self-interaction, defined 
as $v_{0} = \lim_{\mathbf{r} \rightarrow 0} 
\left\{ v_{E}(\mathbf{r}) - 1/r\right\} $. The negative 
electron charges are neutralized by a positive, homogeneous 
background charge. Fraser \emph{et al.} explain 
\cite{fraser1996} how the
electron-background and background-background terms, 
$\hat{H}_{eb}$ and $\hat{H}_{bb}$, vanish
when using Ewald's interaction for the three-dimensional
electron gas. Using the same arguments, one can show that
these terms are also zero in the corresponding 
two-dimensional system.

In the three-dimensional electron gas, the Ewald 
interaction is \cite{drummond2008}
\begin{align}
  v_{E}(\mathbf{r}) &= \sum_{\mathbf{k} \neq \mathbf{0}}
  \frac{4\pi }{L^{3}k^{2}}e^{i\mathbf{k}\cdot \mathbf{r}}
  e^{-\eta^{2}k^{2}/4} \nonumber \\
  &+ \sum_{\mathbf{R}}\frac{1}{\left| \mathbf{r}
    -\mathbf{R}\right| } \erfc \left( \frac{\left| 
    \mathbf{r}-\mathbf{R}\right|}{\eta }\right)
  - \frac{\pi \eta^{2}}{L^{3}},
\end{align}
where $L$ is the box side length, $\erfc(x)$ is the 
complementary error function, and $\eta $ is a free
parameter that can take any value in the interval 
$(0, \infty )$. The translational vector 
\begin{align}
  \mathbf{R} = L\left(n_{x}\mathbf{u}_{x} + n_{y}
  \mathbf{u}_{y} + n_{z}\mathbf{u}_{z}\right) ,
\end{align}
where $\mathbf{u}_{i}$ is the unit vector for dimension $i$,
is defined for all integers $n_{x}$, $n_{y}$, and 
$n_{z}$. These vectors are used to obtain all image
cells in the entire real space. 
The parameter $\eta $ decides how 
the Coulomb interaction is divided into a short-ranged
and long-ranged part, and does not alter the total
function. However, the number of operations needed
to calculate the Ewald interaction with a desired 
accuracy depends on $\eta $, and $\eta $ is therefore 
often chosen to optimize the convergency as a function
of the simulation-cell size \cite{drummond2008}. In
our calculations, we choose $\eta $ to be an infinitesimally
small positive number, similarly as was done in 
Refs.~\cite{shepherd2012b,shepherd2013a,shepherd2013b,shepherd2013c,roggero2013}.
This gives an interaction that is evaluated only in
Fourier space.

When studying the two-dimensional electron gas, we
use an Ewald interaction that is quasi two-dimensional.
The interaction is derived in three dimensions, with 
Fourier discretization in only two dimensions 
\cite{wood2004}. The Ewald effective
interaction has the form \cite{wood2004}
\begin{align}
  v_{E}(\mathbf{r}) &= \sum_{\mathbf{k} \neq \mathbf{0}} 
  \frac{\pi }{L^{2}k}\left\{ e^{-kz} \erfc \left(
  \frac{\eta k}{2} - \frac{z}{\eta }\right)+ \right. \nonumber \\
  & \left. e^{kz}\erfc \left( \frac{\eta k}{2} + \frac{z}{\eta }
  \right) \right\} e^{i\mathbf{k}\cdot \mathbf{r}_{xy}} 
  \nonumber \\
  & + \sum_{\mathbf{R}}\frac{1}{\left| \mathbf{r}-\mathbf{R}
    \right| } \erfc \left( \frac{\left| \mathbf{r}-\mathbf{R}
    \right|}{\eta }\right) \nonumber \\
  & - \frac{2\pi}{L^{2}}\left\{ z\erf \left( \frac{z}{\eta }
  \right) + \frac{\eta }{\sqrt{\pi }}e^{-z^{2}/\eta^{2}}\right\},
\end{align}
where the Fourier vectors $\mathbf{k}$ and the position vector
$\mathbf{r}_{xy}$ are defined in the $(x,y)$ plane. When
applying the interaction $v_{E}(\mathbf{r})$ to two-dimensional
systems, we set $z$ to zero. Similarly as in the 
three-dimensional case, and as suggested in 
Ref.~\cite{drummond2008} for two dimensions, also here we 
choose $\eta $ to approach zero from above. The resulting 
Fourier-transformed interaction is
\begin{align}
  v_{E}^{\eta = 0, z = 0}(\mathbf{r}) = \sum_{\mathbf{k} \neq \mathbf{0}} 
  \frac{2\pi }{L^{2}k}e^{i\mathbf{k}\cdot \mathbf{r}_{xy}}. 
\end{align}
The self-interaction $v_{0}$ is a constant that can be 
included in the reference energy.

In the three-dimensional electron gas, the antisymmetrized
matrix elements are
\begin{align} \label{eq:vmat_3dheg}
  & \langle \mathbf{k}_{p}m_{s_{p}}\mathbf{k}_{q}m_{s_{q}}
  |\tilde{v}|\mathbf{k}_{r}m_{s_{r}}\mathbf{k}_{s}m_{s_{s}}\rangle_{AS} 
  \nonumber \\
  & = \frac{4\pi }{L^{3}}\delta_{\mathbf{k}_{p}+\mathbf{k}_{q},
    \mathbf{k}_{r}+\mathbf{k}_{s}}\left\{ 
  \delta_{m_{s_{p}}m_{s_{r}}}\delta_{m_{s_{q}}m_{s_{s}}}
  \left( 1 - \delta_{\mathbf{k}_{p}\mathbf{k}_{r}}\right) 
  \frac{1}{|\mathbf{k}_{r}-\mathbf{k}_{p}|^{2}}
  \right. \nonumber \\
  & \left. - \delta_{m_{s_{p}}m_{s_{s}}}\delta_{m_{s_{q}}m_{s_{r}}}
  \left( 1 - \delta_{\mathbf{k}_{p}\mathbf{k}_{s}} \right)
  \frac{1}{|\mathbf{k}_{s}-\mathbf{k}_{p}|^{2}} 
  \right\} ,
\end{align}
where the Kronecker delta functions 
$\delta_{\mathbf{k}_{p}\mathbf{k}_{r}}$ and
$\delta_{\mathbf{k}_{p}\mathbf{k}_{s}}$ ensure that the 
contribution with zero momentum transfer vanishes.
Similarly, the matrix elements for the two-dimensional
electron gas are
\begin{align} \label{eq:vmat_2dheg}
  & \langle \mathbf{k}_{p}m_{s_{p}}\mathbf{k}_{q}m_{s_{q}}
  |v|\mathbf{k}_{r}m_{s_{r}}\mathbf{k}_{s}m_{s_{s}}\rangle_{AS} 
  \nonumber \\
  & = \frac{2\pi }{L^{2}}
  \delta_{\mathbf{k}_{p}+\mathbf{k}_{q},\mathbf{k}_{r}+\mathbf{k}_{s}}
  \left\{ \delta_{m_{s_{p}}m_{s_{r}}}\delta_{m_{s_{q}}m_{s_{s}}} 
  \left( 1 - \delta_{\mathbf{k}_{p}\mathbf{k}_{r}}\right)
  \frac{1}{
    |\mathbf{k}_{r}-\mathbf{k}_{p}|} \right.
  \nonumber \\
  & - \left. \delta_{m_{s_{p}}m_{s_{s}}}\delta_{m_{s_{q}}m_{s_{r}}}
  \left( 1 - \delta_{\mathbf{k}_{p}\mathbf{k}_{s}}\right)
  \frac{1}{ 
    |\mathbf{k}_{s}-\mathbf{k}_{p}|}
  \right\} ,
\end{align}
where the single-particle momentum vectors $\mathbf{k}_{p,q,r,s}$
are now defined in two dimensions. In our calculations, the 
self-interaction constant is included in the reference 
energy, as in Eq.~(\ref{eq:ene_ref}). We therefore get the 
Fock-operator matrix elements 
\begin{align}
  \langle \mathbf{k}_{p}|f|\mathbf{k}_{q} \rangle
  = \frac{\hbar^{2}k_{p}^{2}}{2m}\delta_{\mathbf{k}_{p},
  \mathbf{k}_{q}} + \sum_{\mathbf{k}_{i}}\langle 
  \mathbf{k}_{p}\mathbf{k}_{i}|v|\mathbf{k}_{q}
  \mathbf{k}_{i}\rangle_{AS}.
  \label{eq:fock_heg}
\end{align}
In Ref.~\cite{shepherd2013b}, the matrix elements were 
defined with the self-interaction constant included in the
two-body interaction. This gives Fock-operator matrix 
elements with a gap constant. The definition used in 
Ref.~\cite{shepherd2013b} may give numerically more
stable calculations, as the gap constant 
prevents the energy denominator from becoming too small
in the vicinity of the Fermi surface. However, when using
Fock matrix elements as defined in Eq.~(\ref{eq:fock_heg}), 
the energy denominator in the CC equations never vanishes 
unless the numerator is zero. 
When using periodic boundary conditions, the 
discrete-momentum single-particle basis functions 
\[
\phi_{\mathbf{k}}(\mathbf{r}) =
e^{i\mathbf{k}\cdot \mathbf{r}}/L^{d/2}
\]
are associated with 
the single-particle energy   
\begin{align}
  \varepsilon_{n_{x}, n_{y}} = \frac{\hbar^{2}}{2m} 
  \left( \frac{2\pi }{L}\right)^{2}
  \left( n_{x}^{2} + n_{y}^{2}\right)
\end{align}
for two-dimensional sytems and 
\begin{align}
  \varepsilon_{n_{x}, n_{y}, n_{z}} = \frac{\hbar^{2}}{2m}
  \left( \frac{2\pi }{L}\right)^{2}
  \left( n_{x}^{2} + n_{y}^{2} + n_{z}^{2}\right)
\end{align} 
for three-dimensional systems. Similarly as in, for example, 
Refs.~\cite{shepherd2012a,roggero2013,hagen2014}, we choose 
the single-particle basis such that both the occupied and 
unoccupied single-particle spaces have a closed-shell 
structure. This means that all single-particle states 
corresponding to energies below a chosen cutoff are
included in the basis. We study only the unpolarized spin
phase, in which all orbitals are occupied with one spin-up 
and one spin-down electron. Table \ref{tab:orbitals_2dheg}
shows the lowest-lying spin-orbitals and the cumulated
numbers of single-particle states for a two-dimensional
electron box with periodic boundary conditions.
\begin{table}
  \begin{center}
  \begin{tabular}{c|c|c|c|c}
    $n_{x}^{2}+n_{y}^{2}$ & $n_{x}$ & $n_{y}$ & 
    $N_{\uparrow \downarrow }$ & $N_{\uparrow \uparrow }$ \\
    \hline
    0 & 0 & 0 & 2 & 1 \\
    \hline
    1 & -1 &  0 & & \\
    &  1 &  0 & & \\
    &  0 & -1 & & \\
    &  0 &  1 & 10 & 5 \\
    \hline
    2  & -1 & -1 & & \\
    & -1 &  1 & & \\
    &  1 & -1 & & \\
    &  1 &  1 & 18 & 9 \\
    \hline
    4  & -2 & 0 & & \\
    &  2 & 0 & & \\
    &  0 & -2 & & \\
    &  0 &  2 & 26 & 13 \\
    \hline
    5  & -2 & -1 & & \\
    &  2 & -1 & & \\
    & -2 &  1 & & \\
    &  2 &  1 & & \\
    & -1 & -2 & & \\
    & -1 &  2 & & \\
    &  1 & -2 & & \\
    &  1 &  2 & 42 & 21 \\
    \hline
  \end{tabular}
  \end{center}
  \caption{Illustration of how single-particle energies
    fill energy shells in a two-dimensional electron box.
  Here $n_{x}$ and $n_{y}$ are the momentum quantum numbers,
  $n_{x}^{2} + n_{y}^{2}$ determines the single-particle 
  energy level, $N_{\uparrow \downarrow }$ represents the 
  cumulated number of spin-orbitals in an unpolarized spin
  phase, and $N_{\uparrow \uparrow }$ stands for the
  cumulated number of spin-orbitals in a spin-polarized
  system.} 
  \label{tab:orbitals_2dheg}
\end{table}

Finally, a useful benchmark for our calculations is the expression for
the reference energy $E_0$ per particle.
Defining the $T=0$ density $\rho_0$, we can in turn determine in three
dimensions the radius $r_0$ of a sphere representing the volume an
electron occupies (the classical electron radius) as
\[
r_0= \left(\frac{3}{4\pi \rho}\right)^{1/3}.
\]
In two dimensions the corresponding quantity is
\[
r_0= \left(\frac{1}{\pi \rho}\right)^{1/2}.
\]
One can then express the reference energy per electron in terms of the
dimensionless quantity $r_s=r_0/a_0$, where we have introduced the
Bohr radius $a_0=\hbar^2/e^2m$. The energy per electron computed with
the reference Slater determinant can then be written as \cite{fetter}
(using hereafter only atomic units, meaning that $\hbar = m = e = 1$)
\[
E_0/N=\frac{1}{2}\left[\frac{2.21}{r_s^2}-\frac{0.916}{r_s}\right],
\]
for the three-dimensional electron gas.  For the two-dimensional gas
the corresponding expression is, see for example
Refs.~\cite{rajagopal1977,tanatar1989},
\[
E_0/N=\frac{1}{r_s^2}-\frac{8\sqrt{2}}{3\pi r_s}.
\]

For an infinite homogeneous system, there are some particular
simplications due to the conservation of the total momentum of the
particles.  By symmetry considerations, the total momentum of the
system has to be zero zero. Both the kinetic energy operator and the
total Hamiltonian $\hat{H}$ are assumed to be diagonal in the total
momentum $\mathbf{K}$. Hence, both the reference state $\Phi_{0}$ and
the correlated ground state $\Psi$ must be eigenfunctions of the
operator $\mathbf{\hat{K}}$ with the corresponding eigenvalue
$\mathbf{K} = \mathbf{0}$ \cite{day1967}.  This leads to important
simplications to our different many-body methods. In coupled cluster
theory for example, see the discussion in the next subsection, all
terms that involve single particle-hole excitations vanish.

\subsection{Coupled-cluster theory for the homogeneous electron gas}

In this subsection we present the coupled-cluster equations needed to
obtain the correlation energy at the singles and doubles level. Due to
conservation of momentum, the coupled cluster equations simplify
considerably.  The so-called singles equa

The foundation for most many-body methods is to express the correct
wave function by an expansion in a set of basis functions.  One
example is the Hartree-Fock~(HF) method which employs a unitary
transformation of the single-particle wave functions,
\begin{equation}
|\lambda \rangle = \sum_{\psi} C_{\lambda \psi} |\psi \rangle ,
\end{equation}
and approximates the ground state with a reference Slater determinant
built up by these transformed wave functions.  Another example is
configuration interaction~(CI) where the reference determinant is set
to a linear expansion of determinants, including the initial reference
determinant, 1p-1h excitations, 2p-2h excitations and so on, i.e.
\begin{equation}
|\Psi_{0}^{CI} \rangle = C_0 |\Phi_0 \rangle + \sum_{ia} C_i^a
|\Phi_i^a\rangle + \sum_{ijab} C_{ij}^{ab} |\Phi_{ij}^{ab} \rangle +
\cdots \hspace{2mm}.
\end{equation}
In all these methods one needs to solve a set of coupled equations to
find the coefficients.

The coupled-cluster method also expands the exact solution in a set of
Slater determinants, but employs a non-linear expansion through the
exponential ansatz,
\begin{equation}
\label{eq:CC:expon}
|\Psi_0^{CC} \rangle = e^{\hat{T}} |\Phi_0 \rangle ,
\end{equation}
where $\hat{T}$ is the cluster operator including \textit{all}
possible excitations on the reference determinant.  Sorting
excitations by the number of excited electrons, we may generally
express this general cluster operator as a sum of a 1p-1h operator, a
2p-2h operator, and so on,
\begin{equation}
\hat{T} = \hat{T_1} + \hat{T_2} + \hat{T_3} + \cdots \hspace{2mm}.
\end{equation}
In the form of second-quantized operators the 1p-1h cluster operator
is defined as
\begin{equation}
\hat{T}_1 = \sum_{ia} t_i^a \hat{a}^{\dagger} \hat{i},
\end{equation}
the 2p-2h cluster operator as
\begin{equation}
\hat{T}_2 = \frac{1}{4} \sum_{ijab} t_{ij}^{ab} \hat{a}^{\dagger}
\hat{b}^{\dagger} \hat{j} \hat{i},
\end{equation}
continuing up to
\begin{equation}
\hat{T}_n = \left( \frac{1}{n!}\right)^2 \sum_{ij\cdots ab\cdots}
t_{ij\cdots}^{ab\cdots} \hat{a}^{\dagger} \hat{b}^{\dagger} \cdots
\hat{j} \hat{i} .
\end{equation}
As long as we have a complete single-particle basis and include all
possible excitations up to $n$p-$n$h in a system with $n$ particles,
we should find the exact solution for both CI, $|\Psi_0^{CI}\rangle$,
and CC, $|\Psi_0^{CC}\rangle$.

Adding all terms together we get the complete $\hat{T}_1$ amplitude
equations;
\begin{equation}
\label{eq:CC:t1eq_raw}
\begin{split}
0 =& f_{ai} + \sum_d f_{ad} t_i^d - \sum_l f_{li} t_l^a + \sum_{ld}
\langle la||di \rangle t_l^d \\ &+ \sum_{ld} f_{ld} t_{il}^{ad} +
\frac{1}{2} \sum_{lde} \langle al||de \rangle t_{il}^{de} -
\frac{1}{2} \sum_{lmd} \langle lm||di \rangle t_{lm}^{da} - \sum_{ld}
f_{ld} t_i^d t_l^a \\ & + \sum_{lde} \langle al||de \rangle t_i^d
t_l^e - \sum_{lmd} \langle lm||di \rangle t_l^d t_m^a + \frac{1}{2}
\sum_{lmde} \langle lm||de \rangle t_i^d t_{lm}^{ea} \\ &+ \frac{1}{2}
\sum_{lmde} \langle lm||de \rangle t_m^a t_{il}^{de} + \sum_{lmde}
\langle lm||de \rangle t_m^e t_{il}^{ad} + \sum_{lmde} \langle lm||de
\rangle t_i^d t_l^e t_m^a .
\end{split}
\end{equation}
\begin{equation}
\label{eq:CC:t2eq_raw}
\begin{split}
0 =& \langle ab || ij \rangle +\frac{1}{2} \langle ab||de \rangle
t_{ij}^{de} \\ & - \hat{P}_{ij} \left[ f_{li} + f_{ld} t_{i}^d +
  \langle ml||di \rangle t_m^d + \langle ml||de \rangle t_m^d t_i^e
  +\frac{1}{2}\langle ml||de \rangle t_{mi}^{de} \right] t_{lj}^{ab}
\\ & + \frac{1}{2} \left[ \langle lm||ij \rangle + \hat{P}_{ij}
  \langle lm||dj \rangle t_i^d +\frac{1}{2} \langle lm||de \rangle
  t_{ij}^{de} +\hat{P}_{ij} \frac{1}{2} \langle lm||de \rangle t_i^d
  t_j^e \right] t_{lm}^{ab} \\ & + \hat{P}_{ab} \left[ f_{bd} - f_{ld}
  t_l^b + \langle bl||de \rangle t_l^e - \langle lm||de \rangle t_m^e
  t_l^b + \frac{1}{2} \langle lm||de \rangle t_{lm}^{eb} \right]
t_{ij}^{ad} \\ & +\hat{P}_{ij} \hat{P}_{ab} \left[ \langle lb||dj
  \rangle - \langle lm||dj \rangle t_{m}^b + \langle bl||ed \rangle
  t_{j}^e - \langle lm||de \rangle t_j^e t_m^b + \frac{1}{2} \langle
  lm||de \rangle t_{mj}^{eb} \right] t_{il}^{ad} \\ & -\hat{P}_{ab}
\left[ \langle al||ij \rangle + \frac{1}{2} \langle al||de \rangle
  t_{ij}^{de} + \hat{P}_{ij} \langle al||dj \rangle t_i^d +
  \hat{P}_{ij} \frac{1}{2} \langle al||de \rangle t_i^d t_j^e \right.
  \\& \left. + \frac{1}{2} \langle lm||ij \rangle t_m^a + \frac{1}{4}
  \langle lm||de \rangle t_m^a t_{ij}^{de} + \hat{P}_{ij} \frac{1}{2}
  \langle lm||dj \rangle t_i^d t_m^a + \hat{P}_{ij} \frac{1}{4}
  \langle lm||de \rangle t_i^d t_j^e t_m^a \right] t_l^b \\ & +
\hat{P}_{ij} \left[ \langle ab || dj \rangle + \frac{1}{2} \langle
  ab||de \rangle t_j^e \right] t_i^d .
\end{split}
\end{equation}

\subsection{The Similarity Renormalization Group flow equations}
\label{subsec:SRG}
The Similarity Renormalization Group~(SRG) method was introduced
independently by Glazek and Wilson
\cite{PhysRevD.48.5863,*PhysRevD.49.4214} and Wegner
\cite{PhysRepWegner0,*PhysRepWegner} as a new way to implement the
principle of energy scale separation.  The method uses a continuous
sequence of unitary transformations to decouple the high- and
low-energy matrix elements of a given interaction, thus driving the
Hamiltonian towards a band- or block-diagonal form.

Let us consider the initial Hamiltonian
\[
 \hat{H} = \hat{H}^{\rm d} + \Ho,
\]
where $\hat{H}^{\rm d}$ and $\Ho$ denote its ``diagonal'' and
``off-diagonal'' parts, respectively.
 Introducing a flow parameter $s$, there exits a unitary
 transformation $U_s$, such that
\begin{equation}
 \hat{H}_s = U_s^\dagger \hat{H} U_s \equiv \hat{H}^{\rm d}_s + \Ho_s,
\end{equation}
with the relations $U_{s=0} = \mathbf{1}$, and $\hat{H}_{s= 0} =\hat{H}$.  
The transformation $U_s$ is parametrized as
\[
U_s = T_s \exp \left(\int_0^s \! ds'\hat{\eta}_{s'} \right),
\]
where the anti-hermitian operator $\hat{\eta}_s$ serves as generator
of the transformation. With $T_s$ we denote $s$-ordering, which is
defined equivalently to usual time-ordering.  Taking the derivative of
$\hat{H}_s$ with respect to $s$ gives
\begin{equation}
 \frac{d \hat{H}_s}{ds} = \frac{d U_s}{ds}\hat{H} U_s^\dagger + U_s
 \hat{H} \frac{d U_s^\dagger}{ds}.
\label{eq:flow_long}
\end{equation}
Utilizing that for our particular form of $U_s$, we have that 
\begin{equation}
\hat{\eta}_s = \frac{d U_s}{ds} U_s^\dagger = - U_s \frac{d
  U_s^\dagger}{ds} = -\hat{\eta}_s,
\label{eq:eta}
\end{equation} we obtain that 
\begin{equation} \frac{d \hat{H}_s}{ds} = \hat{\eta}_s \hat{H}_s
- \hat{H}_s \hat{\eta}_s = \left[\hat{\eta}_s, \hat{H}_s \right].
\label{eq:flowEquations}
\end{equation} This is the key expression of the SRG method, describing the flow
of the Hamiltonian.  The specific unitary transformation is determined
by the choice of $\hat{\eta}_s$.  Through different choices of
$\hat{\eta}_s$, the SRG evolution can be adapted to the features of a
particular problem.\\

One possibility to solve the flow equations is to choose a basis with
respect to the physical vacuum state, set up the Hamiltonian matrix in
this basis and solve Eq.~(\ref{eq:flowEquations}) as a set of coupled
first-order differential equations. However, since the size of the
problem grows enormously with the number of particles and the size of
the model space, the applicability of this free-space SRG method is
restricted to comparatively small systems.

Instead of performing SRG in free space, the evolution can be done at
finite density, i.e.~directly in the $N$-body system
\cite{kehrein2006flow}. This approach has recently been applied very
successfully in nuclear physics \cite{IMSRG,PhysRevLett.106.222502}
and is called in-medium SRG (IM-SRG). The method allows the evolution
of $3,...,N$-body operators using only two-body machinery, with the
simplifications arising from the use of normal-ordering with respect
to a reference state.

Integrating the flow equations~(\ref{eq:flowEquations}), we face one
of the major challenges of the SRG method, namely the generation of
higher and higher order interaction terms during the flow. With each
evaluation of the commutator, the Hamiltonian gains terms of higher
order, and these induced contributions will in subsequent integration
steps contribute to terms of lower order. In principle, this continues
to infinity.\\ To make the method computationally possible, we have to
close the IM-SRG flow equations, suggesting that we are forced to
truncate the equations to a certain order. We choose to truncate both
$\hat{H}_s$ and $\hat{\eta}_s$ at the two-body level, an approach
which is referred to as IM-SRG(2).  This normal-ordered two-body
approximation seems to be sufficient in many cases and has yielded
excellent results for several nuclei
\cite{PhysRevLett.106.222502,PhysRevLett.109.052501,IMSRG}.\\ The
commutator in the flow equations (\ref{eq:flowEquations}) guarantees
that the IM-SRG wave function $U_s^\dagger|\Phi\rangle$ can be
expanded in terms of linked diagrams only
\cite{shavittbartlett2009,ISI:A1981MN73700014}, which suggests that IM-SRG
is size-extensive. Regarding the quality of the SRG results, it means
that the error introduced by truncating the many-body expansions
scales linearly with the number of particles~$N$. With this
truncation, the generator $\hat{\eta}$ can be written as
%\[
%\hat{\eta} = \sum_{pq} \eta_{pq}^{(1)} \lbrace a_p^\lbrace\dagger\rbrace a_q\rbrace + \frac{1}{4}\sum%\limits_{pqrs}\eta_{pqrs}^{(2)} \lbrace a_p\dagger a_q\dagger a_s a_r \rrb,
%\]
where $\eta_{pq}^{(1)}$ and $ \eta_{pqrs}^{(2)}$ are the one- and
two-body elements, respectively. Making use of the permutation
operator $ \hat{P}_{pq}f(p,q) = f(q,p)$, the IM-SRG(2) flow equations
are given by
Note that for brevity, we skipped all $s$-dependence in the equations.

\subsubsection*{Choice of generator}
To determine the specific unitary transformation, one needs to specify
the generator~$\hat{\eta}$. Through different choices, the SRG flow
can be adapted to the features of a particular problem.\\

Apart from this canonical generator, there exist several other ones in
literature. One of them is White's choice
\cite{White:cond-mat0201346}, which makes numerical approaches much
more efficient.  The problem with Wegner's generator are the widely
varying decaying speeds of the elements, removing first terms with
large energy differences and then subsequently those ones with smaller
energy separations.  That way the flow equations become a stiff set of
coupled differential equations, which often gets numerically
unstable.\\ White takes an alternative approach, which is especially
suited for problems where one is interested in the ground state of a
system. Instead of driving all off-diagonal elements of the
Hamiltonian to zero, he focuses solely on those ones that are
connected to the reference state $|\Phi_0\rangle$, aiming to decouple
the reference state from the remaining Hamiltonian. With a suitable
transformation, the elements get similar decaying speeds, which solves
the problem of stiffness of the flow equations.  The generator is
explicitly constructed the following way
\cite{PhysRevLett.106.222502,White:cond-mat0201346}
\begin{align}
\hat{\eta} &= \sum_{ai} \frac{f_{ai}}{f_a-f_i-v_{aiai}}\lbrace
a_a\dagger a_i\rbrace -\text{hc} \notag \\ & + \sum_{abij}
\frac{v_{abij}}{f_a+f_b-f_i-f_j+A_{abij}}\lbrace a_a\dagger a_b\dagger
a_j a_i\rbrace - \text{hc},
\label{eq:WhiteFull}
\end{align}
with $f_p \equiv f_{pp}$, 'hc' denoting the Hermitian conjugate, and
\[
A_{abij} = v_{abab} + v_{ijij} - v_{aiai} - v_{ajaj} - v_{bibi} -
v_{bjbj}.
\label{eq:White7}
\]
Compared to Wegner's canonical generator, where the final flow
equations involve third order powers of the $f$- and $v$-elements,
these elements contribute only linearly with White's generator, which
results in much better numerical properties.

\subsection{The FCIQMC algorithm}

The recently developed Full Configuration Interaction Quantum Monte
Carlo (FCIQMC) \cite{booth2009,cleland2010,shepherd2012a} promises to
calculate the Full Configuration Interaction (FCI) ground state energy
within a given Hilbert space, and often at a much lower computational
cost than conventional FCI solvers.
%
The FCI energy is given by the equation
\begin{equation}
	E_0^{FCI} = \min{\langle \Psi|\hat H|\Psi\rangle},
	\label{}
\end{equation}
with $|\Psi\rangle\in\mathcal H _{FCI}$. The FCI space $\mathcal H_{FCI}$ is spanned by all
determinants that can be constructed from single-particle states with
an energy lower than the cutoff $\mathbf{k}_c^2/2$.
%
This problem is usually solved by using brute force diagonalizion
techniques like the Lanczos algorithm [mat. comp.], but due to the
large memory footprint of such methods, these are limited to systems
with ``small'' Hilbert spaces. The largest simulations of this kind is
performed with a basis with up to $\sim 10^{11}$(?) determinants [? ?]
which in practice means that only systems with a relatively low number
of particles or small single particle basis can be simulated. The
FCIQMC algorithm does however make it possible threat much larger
Hilbert spaces. One example is the simulations performed by Shepherd
{\it et. al.} \cite{shepherd2012c} where the algorithm is applied on
the three dimensional homogenous electron gas with a bases consisting
of up to $\sim10^{108}$ determinants.

In the following we will briefly introduce the FCIQMC algorithm. 
This is a so called Projector Monte Carlo algorithm where the FCI ground state is found by calculating the projected state
\begin{align}
	\lim_{t\to\infty}e^{-t(\hat H-S)}|\Psi_0\rangle,
	\label{eq:projectedstate}
\end{align}
where the shift $S\in\mathcal R$. The projected state will be
proportional to the FCI ground state $\Psi_0^{FCI}$ if the shift is
equal to $E_0^{FCI}$ and $\expec{\Psi_0|\Psi_0^{FCI}}\neq 0$.  This is
the same approach that is used in other Projector Monte Carlo methods
like Diffusion Monte Carlo or Greens Function Monte Carlo, but the
integration is not performed in the coordinate space but in the
discrete space of Slater determinants. Because of the inherent
anti-symmetric properties of this basis, the infamous sign problem is
not as severe for FCIQMC as for other Projector Monte Carlo methods,
and in principle the exact FCI energies can be calculated. See for
example Refs. \cite{kolodrubetz2013,spencer2012} for a thorough
analysis of the sign problem in FCIQMC.

The large $t$ projected state is found by iterating over the equation
\begin{align}
	|\Psi_0\rangle \gets e^{-\tau(\hat H-S)} |\Psi_0\rangle,
	\label{eqMarkovChain}
\end{align}
where $\tau$ is a small time step. %In this case, $\ket{\Psi_0}$ will become proportional to the FCI ground state in the limit of a large number of iterations.
%
The initial state $|\Psi_0\rangle$ is now expanded on the FCI basis
\begin{align}
	|\Psi_0\rangle = \sum_i C_i |\Phi_i\rangle,
	\label{}
\end{align}
and by approximating the projection operator as its first order Taylor expansion
we can write Eq.~\eqref{eqMarkovChain} in terms of the coefficients $\left\{ C_i \right\}$
\begin{align}
	%\sum_i C_{i}^{(n+1)}\ket{\Phi_i} & \approx (1 - (\hat H - S)\tau) \sum_i C_{i}^{(n)}\ket{\Phi_i}\\
	%&\Downarrow\notag\\
	C_{i} %&= C_{i}^{(n)} - \sum_j(H_{ij}-\partial_{ij}S)\tau C_{j}^{(n)}\notag\\
	%&\gets  C_{i} - (H_{ii}-S)\tau C_{i} - \sum_{j\neq i}H_{ij}\tau C_{j},
	&\gets  C_{i} - \sum_jp_d(i|j) C_{j},
	\label{eq:fciequation}
\end{align}
where we have used $p_d(i|j) = \tau(\hat H_{ij}-S\delta_{ij})$ and
$H_{ij}=\expec{\Phi_i|\hat H|\Phi_j}$.  In the limit of a large number
of iterations, the coefficients $\left\{ C_i \right\}$ will be
proportional to those of the ground state.
%
Note that this approach is exact when the time step $\tau$ is smaller
than $2/(E_M-S)$, where $E_M$ is the largest single particle energy
$E_M=\max(\expec{\Phi_i|\hat H|\Phi_i})$.

A DMC inspired approach is used to find the steady state solution of
Eq. \eqref{eq:fciequation}.  The coefficients $\{C_i\}$ are
represented by a population of $N_W$ walkers $\alpha$ which are
distributed on the determinants $\Phi_{i_\alpha}$ and with a sign
$s_\alpha=\pm$. The amplitude $\{C_i\}$ is then defined to be
\begin{align}
	C_i\propto n_i=\sum_\alpha s_\alpha\delta_{ii_\alpha},
\end{align} 
which is a signed sum where two walkers of opposite signs on the same
determinant will make zero contribution.

It is now possible to define a set of dynamical rules for the walkers
of such a nature that they converge to a distribution
$(n_1,n_2,\dots)\propto (C_1,C_2,\dots)$.  These dynamical rules are
derived from the last term of the right hand side of
Eq. \eqref{eq:fciequation} which describes the rate of which the
coefficients $C_i$ changes during one iteration.

Numerically we do the following. Given a walker on determinant $j$, we
add $N_a=\text{floor}(|p_d(i|j)|)$ walkers on determinant $i$, and one
extra with the probability $N_a-|p_d(i|j)|$. The new walkers have the
same sign as the walker on determinant $j$ is $p_d(i|j)$ is negative
and the opposite sign otherwise.

The sum over all elements $p_d{(i|j)}$ is numerically very expensive
for large FCI bases, and to be able to solve
Eq. \eqref{eq:fciequation} efficiently, Monte Carlo sampling is used
to approximate the sum over the off diagonal terms
\begin{align}
	\sum_j\sum_{i\neq j} p_d{(i|j)} C_{j} \approx \sum_j \sum _{s \in \mathcal S_j} p_d{(s|j)} / p_{gen}(s|j) C_j,\notag\\
	\label{eq:samplingrule}
\end{align}
where $p_{gen}(s|j)$ is the suggestion probability of $s$ from $j$,
and $\mathcal S_j$ consists of $n_j$ elements that are sampled from
$p_{gen}(s|j)$. This reduce the number of operations to $\mathcal
O(N_w)$ each iteration.
%
This equation is only correct when $p_{gen}(s|j)$ is different from
zero for all connected determinants $|\Phi_j\rangle$ and $|\Phi_s\rangle$
\begin{align}
	p_{gen}(s|j)\neq0 \quad\forall\quad H_{js} \neq 0.
\end{align}

In principle, there is no other constraint on $p_{gen}(s|j)$, and as
long as the number of iterations is large enough the simulations will
eventually converge to the correct results, but in practice certain
choices of suggestion probability distributions are expected to lead
to a more efficient sampling.  In our implementation we have used a
$p_{gen}(s|j)$ which is uniform for all $s$ that corresponds to an
interaction $\expec{\Phi_s|\hat H|\Phi_j}$ % \neq 0$ 
where the spin
and momentum are conserved and zero otherwise.

It has been shown that the number of walkers must be above a certain
threshold $N_C$ for the population to converge to the ground state
distribution \cite{spencer2012,bestref?booth?}. This number is known
to be highly system dependent and to increase roughly linearly with
the size of the FCI space \cite{bestref?}.

Cleland {\it et. al.}  (2010) \cite{cleland2010} demonstrated that by
introducing a set of new dynamical rules for the walkers, the
performance of the algorithm can be improved drastically while still
obtaining results with a precision of $10^{-3}$ Ry or better. The
number of walkers that is necessary in such simulations are typically
several orders of magnitude smaller than $N_C$.
%
The improved algorithm is called initiator-FCIQMC or i-FCIQMC, and
does only differ from the original algorithm in how the spawning
probabilites $\{p_d(i|j);i\neq j\}$ are calculated.

First, a so called initiator space is defined. This is a set of
determinants which are assumed to have a relatively large average
population (large ampiludes $|C_i|$). All walkers that live on a
determinant in the initiator space are called initiators, and
initiators follows the same rules as walkers in the basic FCIQMC
algorithm.  The first rule of i-FCIQMC concerns the non-initiators and
states that:
\begin{itemize}
	\item ({\it Spawning rule I}): Non-initiators are only allowed
          to spawn on determinants that are previously populated.
\end{itemize}
With this rule the Hilbert space is in effect reduced to the states in
the initiator space and the single and double excitations of
these. Next, two additional rules are added to improve the energy and
to give the walkers access to the entire FCI space:
\begin{itemize}
			\item ({\it Dynamic enlargement of the
                          initiator space}): We introduce the
                          initiator threshold $N_I$ which is a
                          parameter that is set before a simulations
                          is started.  All determinants $\Phi_i$ with
                          a population $|n_i|\ge N_{I}$ are
                          automatically included in the initiator
                          space for as long the population is above
                          this limit. Here $n_i$ is the signed
                          population of $\Phi_i$. % and $N_i$ is a
                          positive integer that we will refer to as
                          the initiator threshold. The initiator
                          threshold is set before the simulation is
                          started. %Note that with $N_I=1$ i-FCIQMC is
                          identical to FCIQMC.
			\item ({\it Spawning rule II}): If the
                          absolute value of the signed sum
                          $\sum_\alpha s_\alpha$ of non-initiator
                          walkers $\alpha$ that attempts to spawn on
                          an unpopulated determinant is larger than
                          two, then these spawning events are
                          accepted.
\end{itemize}
%
Although the initiator adaption of the algorithm has been shown to
increase the efficiency, it is only correct according to
Eq.~\eqref{eq:fciequation} in the limit of a large number of of
walkers or an initiator threshold $N_I=1$. In this case the initiator
space will grow to include the entire FCI space, and i-FCIQMC will
tend to FCIQMC.
%
This means that the i-FCIQMC energies are systematically improvable
and that the so called initiator error will decay to zero when the
number of walkers is increased.

The initiator spaces are defined using a so called Complete Active
Space (CAS) criterion, which is often used within quantum chemistry
together with methods like many-body pertubation theory or the
configuration interaction method to pick out physically relevant
states of a Hilbert space.
%
Because of the dynamic enlargement of the initiator space, the final
result will be the same regardles of which CAS we choose. But, as
Cleland {\it et. al.} \cite{cleland2010} has shown, an appropriate CAS
might sometimes improve the convergence.

The population of walkers will on average increase if the shift $S$ is larger than $E^{FCI}_0$ and decrease if $S$ is smaller than $E_0^{FCI}$. 
%
We can therefore control the population by varying $S$. We use the formula of Umrigar {\em et al} 
\cite{Umrigar_1993} which has been used for population control in the context of DMC
\begin{align}
	S\to S^{(i)} = S^{(i-1)}- \frac{\xi}\tau \log\left[\frac{N_W^{(i)}}{N_W^{(i-1)}}\right],
	\label{eqvaryshift}
\end{align}
where $\xi$ is a real number of which the optimal value must be found by experiment, and $N_W^{(i)}$ and $S^{(i)}$ are the population and shift at the time $i\tau$. 
%
The shift $S$ that stabilizes the number of walkers must be the ground state energy, and this means 
%
that the above equation provides us with a measure of $E_0^{FCI}$.

We use two different statistical estimators to calculate the energy. The first is the one that we just mentioned, which can be taken as the long time average of the shift 
\begin{align}
	E_0\approx\expec S=(1/N) \sum_{i=1}^N S^{(i)},
	\label{eqvaryshift}
\end{align}
and which we will refer to as the generational estimator of the energy.
%
The second is the projected estimator that is calculated by sampling the projected energy
\begin{align}
	&E_0 = \frac{ \expec{ \Phi_0|\hat H|\Psi_0^{FCI} } }{\expec{ \Phi_0|\Psi_0^{FCI} }} 
	= \frac{ \expec{ \Phi_0|\hat H\sum_iC_i|\Phi_i} }{\expec{ \Phi_0|\sum_iC_i|\Phi_i }} \label{eqProje1},
\end{align}
which is rewritten on the form
\begin{align}
	E_0 \approx &\expec{E_P} = H_{00} + \sum_{i\neq0}H_{0i} \expec{n_i}/\expec{n_0}.
	\label{}
\end{align}

An FCIQMC or i-FCIQMC simulation is usually started with one walker on
the Hartree-Fock (HF) determinant and a shift $S$ that is equal to the
HF energy. Since the HF energy is larger than the FCI energy, the
number of walkers $N_W$ will increase rapidly. To stop the population
growth and stabilize the population around the desired number of
walkers, the shift $S$ is adjusted according to
Eq. \eqref{eqvaryshift}. After a thermalization phase we start
sampling the generational and the projected estimator.  These
estimators are statistical measures, and there will always be a
certain statistical error associated with them. The statistical errors
of both estimators are calculated using Flyvbjerg-Pedersen analysis
[citation], and the energy with the smallest statistical error is
chosen as the final result.

A series of simulations with an increasing number of walkers must be
performed to converge the i-FCIQMC energy. (A study of the convergence
of the initiator error is provided by Sheperd {\it et. al.}
\cite{shepherd2012c}).

\section{Results}
\label{sec:results}
\subsection{Benchmark results for the three-dimensional electron gas}

\subsection{The two-dimensional electron gas}

\begin{figure}[hbtp]
     \begin{center}

            \includegraphics[width=0.4\textwidth]{figures/nOrbitsEneN10Rs10.pdf}
    \end{center}
    \caption{Correlation energy per particle as function of the number of
single-particle orbits for a ten-particle electron gas at $r_s=1.0$, using CCD, IMSRG(2) 
and FCIQMC. The IMSRG(2)
correlation energies are in excellent agreement with the almost
exact FCIQMC results. For comparison, we have marked the correlation
energy in the thermodynamic limit, as obtained by extrapolating GFMC
energies. The width of the horizontal line shows the statistical error of
the GFMC calculations.}
            \label{fig:N10}
\end{figure}

\begin{figure}[hbtp]
     \begin{center}

            \includegraphics[width=0.4\textwidth]{figures/nOrbitsEneN26Rs10.pdf}
    \end{center}
    \caption{Correlation energy per particle as function of the number of
single-particle orbits for a 26 particle electron gas at $r_s=1.0$, using CCD, IMSRG(2) 
and FCIQMC. Here we see a larger discrepancy between the IMSRG(2) results and the FCIQMC values. 
For comparison, we have marked the correlation
energy in the thermodynamic limit, as obtained by extrapolating GFMC
energies.}
            \label{fig:N26}
\end{figure}

\begin{figure}[hbtp]
     \begin{center}

            \includegraphics[width=0.4\textwidth]{figures/nOrbitsEneNRs01.pdf}
    \end{center}
    \caption{CCD correlation energy per particle as function of the number of
single-particle orbits for varying  electron numebrs $N$ $r_s=0.1$.}
            \label{fig:ccd01}
\end{figure}

\begin{figure}[hbtp]
     \begin{center}

            \includegraphics[width=0.4\textwidth]{figures/nOrbitsEneNRs05.pdf}
    \end{center}
    \caption{CCD correlation energy per particle as function of the number of
single-particle orbits for varying  electron numebrs $N$ $r_s=0.5$.}
            \label{fig:ccd05}
\end{figure}

\begin{figure}[hbtp]
     \begin{center}

            \includegraphics[width=0.4\textwidth]{figures/nOrbitsEneNRs10.pdf}
    \end{center}
    \caption{CCD correlation energy per particle as function of the number of
single-particle orbits for varying  electron numebrs $N$ $r_s=1.0$.}
            \label{fig:ccd05}
\end{figure}

\begin{figure}[hbtp]
     \begin{center}

            \includegraphics[width=0.4\textwidth]{figures/nOrbitsEneNRs20.pdf}
    \end{center}
    \caption{CCD correlation energy per particle as function of the number of
single-particle orbits for varying  electron numebrs $N$ $r_s=2.0$.}
            \label{fig:ccd05}
\end{figure}

\begin{align}
\frac{d E_0}{ds} &= \sum_{ia}\left( \eta_{ia}^{(1)}f_{ai} -
\eta_{ai}^{(1)}f_{ia}\right) +
\frac{1}{2}\sum_{ijab}\eta_{ijab}^{(2)}v_{abij},
\label{eq:flow1}\\
\frac{d f_{pq}}{ds}&= \sum_r \left( \eta_{pr}^{(1)}f_{rq} +
\eta_{qr}^{(1)}f_{rp}\right) + \sum_{ia}\left( 1-\hat{P}_{ia}\right)
\left( \eta_{ia}^{(1)}v_{apiq} - f_{ia}\eta_{apiq}^{(2)} \right)
\notag \\ & +\frac{1}{2} \sum_{aij} \left( 1+\hat{P}_{pq}\right)
\eta_{apij}^{(2)}v_{ijaq} + \frac{1}{2}\sum_{abi}\left(
1+\hat{P}_{pq}\right) \eta_{ipab}^{(2)}v_{abiq},
\label{eq:flow2}
\end{align}

\begin{align}
\frac{d v_{pqrs}}{ds} &= \sum_t \left( 1-\hat{P}_{pq} \right) \left(
\eta_{pt}^{(1)}v_{tqrs}-f_{pt}\eta_{tqrs}^{(2)}\right) -\sum_t \left(
1-\hat{P}_{rs} \right) \left( \eta_{tr}^{(1)} v_{pqts} - f_{tr}
\eta_{pqts}^{(2)}\right) \notag \\ & +\frac{1}{2}\sum_{ab}
\lb\eta_{pqab}^{(2)} v_{abrs} - v_{pqab}\eta_{abrs}^{(2)}\right) -
\frac{1}{2}\sum_{ij} \lb\eta_{pqij}^{(2)} v_{ijrs} -
v_{pqij}\eta_{ijrs}^{(2)}\right) \notag \\ & -\sum_{ia} \left( 1-
\hat{P}_{ia} \right) \left( 1-\hat{P}_{pq}\right) \left(
1-\hat{P}_{rs} \right) \eta_{aqis}^{(2)}v_{ipar}.
\label{eq:flow3}
\end{align}

\section{Conclusions}
\label{sec:conclusions}
%
\begin{acknowledgments}
  This work was supported by the

\end{acknowledgments}

\bibliography{paper} \bibliographystyle{apsrev4-1}

\end{document}

  \begin{itemize}
  \item The first study of the 2D electron gas with SRG, FCIQMC, and CC  
  \item How well can the methods describe correlations?
  \item At what densities do SRG and CC break down?
  \item Theoretical explanation of breakdown?
  \item Ring and ladder approximations vs. CCD 
  \item What are differences between finite and periodic systems
    (QD vs. 2DEG)
  \item Compare 2D with finite thickness 2D+1D (or 3D?)
  \item Virial theorem (J{\o}rgen's thesis)?
  \item More ideas?
  \end{itemize}

### 2018-02-15 doc/ElectronGas/paper.tex
Added electron gas  paper

--- 2018-02-15 doc/ElectronGas/backup.tex
@@
   Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
+
+\author{Sean B.~S.~Miller}
+\affiliation{Depatment of Physics, Chalmers Technical University, G\"oteborg, Sweden}
 
@@
 
-Let us consider the initial Hamiltonian
-\[
- \hat{H} = \hat{H}^{\rm d} + \Ho,
-\]
-where $\hat{H}^{\rm d}$ and $\Ho$ denote its ``diagonal'' and
-``off-diagonal'' parts, respectively.
- Introducing a flow parameter $s$, there exits a unitary
- transformation $U_s$, such that
-\begin{equation}
- \hat{H}_s = U_s^\dagger \hat{H} U_s \equiv \hat{H}^{\rm d}_s + \Ho_s,
-\end{equation}
-with the relations $U_{s=0} = \mathbf{1}$, and $\hat{H}_{s= 0} =\hat{H}$.  
-The transformation $U_s$ is parametrized as
-\[
-U_s = T_s \exp \left(\int_0^s \! ds'\hat{\eta}_{s'} \right),
-\]
-where the anti-hermitian operator $\hat{\eta}_s$ serves as generator
-of the transformation. With $T_s$ we denote $s$-ordering, which is
-defined equivalently to usual time-ordering.  Taking the derivative of
-$\hat{H}_s$ with respect to $s$ gives
-\begin{equation}
- \frac{d \hat{H}_s}{ds} = \frac{d U_s}{ds}\hat{H} U_s^\dagger + U_s
- \hat{H} \frac{d U_s^\dagger}{ds}.
-\label{eq:flow_long}
-\end{equation}
-Utilizing that for our particular form of $U_s$, we have that 
-\begin{equation}
-\hat{\eta}_s = \frac{d U_s}{ds} U_s^\dagger = - U_s \frac{d
-  U_s^\dagger}{ds} = -\hat{\eta}_s,
-\label{eq:eta}
-\end{equation} we obtain that 
-\begin{equation} \frac{d \hat{H}_s}{ds} = \hat{\eta}_s \hat{H}_s
-- \hat{H}_s \hat{\eta}_s = \left[\hat{\eta}_s, \hat{H}_s \right].
-\label{eq:flowEquations}
-\end{equation} This is the key expression of the SRG method, describing the flow
-of the Hamiltonian.  The specific unitary transformation is determined
-by the choice of $\hat{\eta}_s$.  Through different choices of
-$\hat{\eta}_s$, the SRG evolution can be adapted to the features of a
-particular problem.\\
-
-One possibility to solve the flow equations is to choose a basis with
-respect to the physical vacuum state, set up the Hamiltonian matrix in
-this basis and solve Eq.~(\ref{eq:flowEquations}) as a set of coupled
-first-order differential equations. However, since the size of the
-problem grows enormously with the number of particles and the size of
-the model space, the applicability of this free-space SRG method is
-restricted to comparatively small systems.
-
-Instead of performing SRG in free space, the evolution can be done at
-finite density, i.e.~directly in the $N$-body system
-\cite{kehrein2006flow}. This approach has recently been applied very
-successfully in nuclear physics \cite{IMSRG,PhysRevLett.106.222502}
-and is called in-medium SRG (IM-SRG). The method allows the evolution
-of $3,...,N$-body operators using only two-body machinery, with the
-simplifications arising from the use of normal-ordering with respect
-to a reference state.
-
-Integrating the flow equations~(\ref{eq:flowEquations}), we face one
-of the major challenges of the SRG method, namely the generation of
-higher and higher order interaction terms during the flow. With each
-evaluation of the commutator, the Hamiltonian gains terms of higher
-order, and these induced contributions will in subsequent integration
-steps contribute to terms of lower order. In principle, this continues
-to infinity.\\ To make the method computationally possible, we have to
-close the IM-SRG flow equations, suggesting that we are forced to
-truncate the equations to a certain order. We choose to truncate both
-$\hat{H}_s$ and $\hat{\eta}_s$ at the two-body level, an approach
-which is referred to as IM-SRG(2).  This normal-ordered two-body
-approximation seems to be sufficient in many cases and has yielded
-excellent results for several nuclei
-\cite{PhysRevLett.106.222502,PhysRevLett.109.052501,IMSRG}.\\ The
-commutator in the flow equations (\ref{eq:flowEquations}) guarantees
-that the IM-SRG wave function $U_s^\dagger|\Phi\rangle$ can be
-expanded in terms of linked diagrams only
-\cite{shavittbartlett2009,ISI:A1981MN73700014}, which suggests that IM-SRG
-is size-extensive. Regarding the quality of the SRG results, it means
-that the error introduced by truncating the many-body expansions
-scales linearly with the number of particles~$N$. With this
-truncation, the generator $\hat{\eta}$ can be written as
-%\[
-%\hat{\eta} = \sum_{pq} \eta_{pq}^{(1)} \lbrace a_p^\lbrace\dagger\rbrace a_q\rbrace + \frac{1}{4}\sum%\limits_{pqrs}\eta_{pqrs}^{(2)} \lbrace a_p\dagger a_q\dagger a_s a_r \rrb,
-%\]
-where $\eta_{pq}^{(1)}$ and $ \eta_{pqrs}^{(2)}$ are the one- and
-two-body elements, respectively. Making use of the permutation
-operator $ \hat{P}_{pq}f(p,q) = f(q,p)$, the IM-SRG(2) flow equations
-are given by
-Note that for brevity, we skipped all $s$-dependence in the equations.
-
-\subsubsection*{Choice of generator}
-To determine the specific unitary transformation, one needs to specify
-the generator~$\hat{\eta}$. Through different choices, the SRG flow
-can be adapted to the features of a particular problem.\\
-
-Apart from this canonical generator, there exist several other ones in
-literature. One of them is White's choice
-\cite{White:cond-mat0201346}, which makes numerical approaches much
-more efficient.  The problem with Wegner's generator are the widely
-varying decaying speeds of the elements, removing first terms with
-large energy differences and then subsequently those ones with smaller
-energy separations.  That way the flow equations become a stiff set of
-coupled differential equations, which often gets numerically
-unstable.\\ White takes an alternative approach, which is especially
-suited for problems where one is interested in the ground state of a
-system. Instead of driving all off-diagonal elements of the
-Hamiltonian to zero, he focuses solely on those ones that are
-connected to the reference state $|\Phi_0\rangle$, aiming to decouple
-the reference state from the remaining Hamiltonian. With a suitable
-transformation, the elements get similar decaying speeds, which solves
-the problem of stiffness of the flow equations.  The generator is
-explicitly constructed the following way
-\cite{PhysRevLett.106.222502,White:cond-mat0201346}
-\begin{align}
-\hat{\eta} &= \sum_{ai} \frac{f_{ai}}{f_a-f_i-v_{aiai}}\lbrace
-a_a\dagger a_i\rbrace -\text{hc} \notag \\ & + \sum_{abij}
-\frac{v_{abij}}{f_a+f_b-f_i-f_j+A_{abij}}\lbrace a_a\dagger a_b\dagger
-a_j a_i\rbrace - \text{hc},
-\label{eq:WhiteFull}
-\end{align}
-with $f_p \equiv f_{pp}$, 'hc' denoting the Hermitian conjugate, and
-\[
-A_{abij} = v_{abab} + v_{ijij} - v_{aiai} - v_{ajaj} - v_{bibi} -
-v_{bjbj}.
-\label{eq:White7}
-\]
-Compared to Wegner's canonical generator, where the final flow
-equations involve third order powers of the $f$- and $v$-elements,
-these elements contribute only linearly with White's generator, which
-results in much better numerical properties.
-
-\subsection{The FCIQMC algorithm}
-
-The recently developed Full Configuration Interaction Quantum Monte
-Carlo (FCIQMC) \cite{booth2009,cleland2010,shepherd2012a} promises to
-calculate the Full Configuration Interaction (FCI) ground state energy
-within a given Hilbert space, and often at a much lower computational
-cost than conventional FCI solvers.
-%
-The FCI energy is given by the equation
-\begin{equation}
-	E_0^{FCI} = \min{\langle \Psi|\hat H|\Psi\rangle},
-	\label{}
-\end{equation}
-with $|\Psi\rangle\in\mathcal H _{FCI}$. The FCI space $\mathcal H_{FCI}$ is spanned by all
-determinants that can be constructed from single-particle states with
-an energy lower than the cutoff $\mathbf{k}_c^2/2$.
-%
-This problem is usually solved by using brute force diagonalizion
-techniques like the Lanczos algorithm [mat. comp.], but due to the
-large memory footprint of such methods, these are limited to systems
-with ``small'' Hilbert spaces. The largest simulations of this kind is
-performed with a basis with up to $\sim 10^{11}$(?) determinants [? ?]
-which in practice means that only systems with a relatively low number
-of particles or small single particle basis can be simulated. The
-FCIQMC algorithm does however make it possible threat much larger
-Hilbert spaces. One example is the simulations performed by Shepherd
-{\it et. al.} \cite{shepherd2012c} where the algorithm is applied on
-the three dimensional homogenous electron gas with a bases consisting
-of up to $\sim10^{108}$ determinants.
-
-In the following we will briefly introduce the FCIQMC algorithm. 
-This is a so called Projector Monte Carlo algorithm where the FCI ground state is found by calculating the projected state
-\begin{align}
-	\lim_{t\to\infty}e^{-t(\hat H-S)}|\Psi_0\rangle,
-	\label{eq:projectedstate}
-\end{align}
-where the shift $S\in\mathcal R$. The projected state will be
-proportional to the FCI ground state $\Psi_0^{FCI}$ if the shift is
-equal to $E_0^{FCI}$ and $\expec{\Psi_0|\Psi_0^{FCI}}\neq 0$.  This is
-the same approach that is used in other Projector Monte Carlo methods
-like Diffusion Monte Carlo or Greens Function Monte Carlo, but the
-integration is not performed in the coordinate space but in the
-discrete space of Slater determinants. Because of the inherent
-anti-symmetric properties of this basis, the infamous sign problem is
-not as severe for FCIQMC as for other Projector Monte Carlo methods,
-and in principle the exact FCI energies can be calculated. See for
-example Refs. \cite{kolodrubetz2013,spencer2012} for a thorough
-analysis of the sign problem in FCIQMC.
-
-The large $t$ projected state is found by iterating over the equation
-\begin{align}
-	|\Psi_0\rangle \gets e^{-\tau(\hat H-S)} |\Psi_0\rangle,
-	\label{eqMarkovChain}
-\end{align}
-where $\tau$ is a small time step. %In this case, $\ket{\Psi_0}$ will become proportional to the FCI ground state in the limit of a large number of iterations.
-%
-The initial state $|\Psi_0\rangle$ is now expanded on the FCI basis
-\begin{align}
-	|\Psi_0\rangle = \sum_i C_i |\Phi_i\rangle,
-	\label{}
-\end{align}
-and by approximating the projection operator as its first order Taylor expansion
-we can write Eq.~\eqref{eqMarkovChain} in terms of the coefficients $\left\{ C_i \right\}$
-\begin{align}
-	%\sum_i C_{i}^{(n+1)}\ket{\Phi_i} & \approx (1 - (\hat H - S)\tau) \sum_i C_{i}^{(n)}\ket{\Phi_i}\\
-	%&\Downarrow\notag\\
-	C_{i} %&= C_{i}^{(n)} - \sum_j(H_{ij}-\partial_{ij}S)\tau C_{j}^{(n)}\notag\\
-	%&\gets  C_{i} - (H_{ii}-S)\tau C_{i} - \sum_{j\neq i}H_{ij}\tau C_{j},
-	&\gets  C_{i} - \sum_jp_d(i|j) C_{j},
-	\label{eq:fciequation}
-\end{align}
-where we have used $p_d(i|j) = \tau(\hat H_{ij}-S\delta_{ij})$ and
-$H_{ij}=\expec{\Phi_i|\hat H|\Phi_j}$.  In the limit of a large number
-of iterations, the coefficients $\left\{ C_i \right\}$ will be
-proportional to those of the ground state.
-%
-Note that this approach is exact when the time step $\tau$ is smaller
-than $2/(E_M-S)$, where $E_M$ is the largest single particle energy
-$E_M=\max(\expec{\Phi_i|\hat H|\Phi_i})$.
-
-A DMC inspired approach is used to find the steady state solution of
-Eq. \eqref{eq:fciequation}.  The coefficients $\{C_i\}$ are
-represented by a population of $N_W$ walkers $\alpha$ which are
-distributed on the determinants $\Phi_{i_\alpha}$ and with a sign
-$s_\alpha=\pm$. The amplitude $\{C_i\}$ is then defined to be
-\begin{align}
-	C_i\propto n_i=\sum_\alpha s_\alpha\delta_{ii_\alpha},
-\end{align} 
-which is a signed sum where two walkers of opposite signs on the same
-determinant will make zero contribution.
-
-It is now possible to define a set of dynamical rules for the walkers
-of such a nature that they converge to a distribution
-$(n_1,n_2,\dots)\propto (C_1,C_2,\dots)$.  These dynamical rules are
-derived from the last term of the right hand side of
-Eq. \eqref{eq:fciequation} which describes the rate of which the
-coefficients $C_i$ changes during one iteration.
-
-Numerically we do the following. Given a walker on determinant $j$, we
-add $N_a=\text{floor}(|p_d(i|j)|)$ walkers on determinant $i$, and one
-extra with the probability $N_a-|p_d(i|j)|$. The new walkers have the
-same sign as the walker on determinant $j$ is $p_d(i|j)$ is negative
-and the opposite sign otherwise.
-
-The sum over all elements $p_d{(i|j)}$ is numerically very expensive
-for large FCI bases, and to be able to solve
-Eq. \eqref{eq:fciequation} efficiently, Monte Carlo sampling is used
-to approximate the sum over the off diagonal terms
-\begin{align}
-	\sum_j\sum_{i\neq j} p_d{(i|j)} C_{j} \approx \sum_j \sum _{s \in \mathcal S_j} p_d{(s|j)} / p_{gen}(s|j) C_j,\notag\\
-	\label{eq:samplingrule}
-\end{align}
-where $p_{gen}(s|j)$ is the suggestion probability of $s$ from $j$,
-and $\mathcal S_j$ consists of $n_j$ elements that are sampled from
-$p_{gen}(s|j)$. This reduce the number of operations to $\mathcal
-O(N_w)$ each iteration.
-%
-This equation is only correct when $p_{gen}(s|j)$ is different from
-zero for all connected determinants $|\Phi_j\rangle$ and $|\Phi_s\rangle$
-\begin{align}
-	p_{gen}(s|j)\neq0 \quad\forall\quad H_{js} \neq 0.
-\end{align}
-
-In principle, there is no other constraint on $p_{gen}(s|j)$, and as
-long as the number of iterations is large enough the simulations will
-eventually converge to the correct results, but in practice certain
-choices of suggestion probability distributions are expected to lead
-to a more efficient sampling.  In our implementation we have used a
-$p_{gen}(s|j)$ which is uniform for all $s$ that corresponds to an
-interaction $\expec{\Phi_s|\hat H|\Phi_j}$ % \neq 0$ 
-where the spin
-and momentum are conserved and zero otherwise.
-
-It has been shown that the number of walkers must be above a certain
-threshold $N_C$ for the population to converge to the ground state
-distribution \cite{spencer2012,bestref?booth?}. This number is known
-to be highly system dependent and to increase roughly linearly with
-the size of the FCI space \cite{bestref?}.
-
-Cleland {\it et. al.}  (2010) \cite{cleland2010} demonstrated that by
-introducing a set of new dynamical rules for the walkers, the
-performance of the algorithm can be improved drastically while still
-obtaining results with a precision of $10^{-3}$ Ry or better. The
-number of walkers that is necessary in such simulations are typically
-several orders of magnitude smaller than $N_C$.
-%
-The improved algorithm is called initiator-FCIQMC or i-FCIQMC, and
-does only differ from the original algorithm in how the spawning
-probabilites $\{p_d(i|j);i\neq j\}$ are calculated.
-
-First, a so called initiator space is defined. This is a set of
-determinants which are assumed to have a relatively large average
-population (large ampiludes $|C_i|$). All walkers that live on a
-determinant in the initiator space are called initiators, and
-initiators follows the same rules as walkers in the basic FCIQMC
-algorithm.  The first rule of i-FCIQMC concerns the non-initiators and
-states that:
-\begin{itemize}
-	\item ({\it Spawning rule I}): Non-initiators are only allowed
-          to spawn on determinants that are previously populated.
-\end{itemize}
-With this rule the Hilbert space is in effect reduced to the states in
-the initiator space and the single and double excitations of
-these. Next, two additional rules are added to improve the energy and
-to give the walkers access to the entire FCI space:
-\begin{itemize}
-			\item ({\it Dynamic enlargement of the
-                          initiator space}): We introduce the
-                          initiator threshold $N_I$ which is a
-                          parameter that is set before a simulations
-                          is started.  All determinants $\Phi_i$ with
-                          a population $|n_i|\ge N_{I}$ are
-                          automatically included in the initiator
-                          space for as long the population is above
-                          this limit. Here $n_i$ is the signed
-                          population of $\Phi_i$. % and $N_i$ is a
-                          positive integer that we will refer to as
-                          the initiator threshold. The initiator
-                          threshold is set before the simulation is
-                          started. %Note that with $N_I=1$ i-FCIQMC is
-                          identical to FCIQMC.
-			\item ({\it Spawning rule II}): If the
-                          absolute value of the signed sum
-                          $\sum_\alpha s_\alpha$ of non-initiator
-                          walkers $\alpha$ that attempts to spawn on
-                          an unpopulated determinant is larger than
-                          two, then these spawning events are
-                          accepted.
-\end{itemize}
-%
-Although the initiator adaption of the algorithm has been shown to
-increase the efficiency, it is only correct according to
-Eq.~\eqref{eq:fciequation} in the limit of a large number of of
-walkers or an initiator threshold $N_I=1$. In this case the initiator
-space will grow to include the entire FCI space, and i-FCIQMC will
-tend to FCIQMC.
-%
-This means that the i-FCIQMC energies are systematically improvable
-and that the so called initiator error will decay to zero when the
-number of walkers is increased.
-
-The initiator spaces are defined using a so called Complete Active
-Space (CAS) criterion, which is often used within quantum chemistry
-together with methods like many-body pertubation theory or the
-configuration interaction method to pick out physically relevant
-states of a Hilbert space.
-%
-Because of the dynamic enlargement of the initiator space, the final
-result will be the same regardles of which CAS we choose. But, as
-Cleland {\it et. al.} \cite{cleland2010} has shown, an appropriate CAS
-might sometimes improve the convergence.
-
-The population of walkers will on average increase if the shift $S$ is larger than $E^{FCI}_0$ and decrease if $S$ is smaller than $E_0^{FCI}$. 
-%
-We can therefore control the population by varying $S$. We use the formula of Umrigar {\em et al} 
-\cite{Umrigar_1993} which has been used for population control in the context of DMC
-\begin{align}
-	S\to S^{(i)} = S^{(i-1)}- \frac{\xi}\tau \log\left[\frac{N_W^{(i)}}{N_W^{(i-1)}}\right],
-	\label{eqvaryshift}
-\end{align}
-where $\xi$ is a real number of which the optimal value must be found by experiment, and $N_W^{(i)}$ and $S^{(i)}$ are the population and shift at the time $i\tau$. 
-%
-The shift $S$ that stabilizes the number of walkers must be the ground state energy, and this means 
-%
-that the above equation provides us with a measure of $E_0^{FCI}$.
-
-We use two different statistical estimators to calculate the energy. The first is the one that we just mentioned, which can be taken as the long time average of the shift 
-\begin{align}
-	E_0\approx\expec S=(1/N) \sum_{i=1}^N S^{(i)},
-	\label{eqvaryshift}
-\end{align}
-and which we will refer to as the generational estimator of the energy.
-%
-The second is the projected estimator that is calculated by sampling the projected energy
-\begin{align}
-	&E_0 = \frac{ \expec{ \Phi_0|\hat H|\Psi_0^{FCI} } }{\expec{ \Phi_0|\Psi_0^{FCI} }} 
-	= \frac{ \expec{ \Phi_0|\hat H\sum_iC_i|\Phi_i} }{\expec{ \Phi_0|\sum_iC_i|\Phi_i }} \label{eqProje1},
-\end{align}
-which is rewritten on the form
-\begin{align}
-	E_0 \approx &\expec{E_P} = H_{00} + \sum_{i\neq0}H_{0i} \expec{n_i}/\expec{n_0}.
-	\label{}
-\end{align}
-
-An FCIQMC or i-FCIQMC simulation is usually started with one walker on
-the Hartree-Fock (HF) determinant and a shift $S$ that is equal to the
-HF energy. Since the HF energy is larger than the FCI energy, the
-number of walkers $N_W$ will increase rapidly. To stop the population
-growth and stabilize the population around the desired number of
-walkers, the shift $S$ is adjusted according to
-Eq. \eqref{eqvaryshift}. After a thermalization phase we start
-sampling the generational and the projected estimator.  These
-estimators are statistical measures, and there will always be a
-certain statistical error associated with them. The statistical errors
-of both estimators are calculated using Flyvbjerg-Pedersen analysis
-[citation], and the energy with the smallest statistical error is
-chosen as the final result.
-
-A series of simulations with an increasing number of walkers must be
-performed to converge the i-FCIQMC energy. (A study of the convergence
-of the initiator error is provided by Sheperd {\it et. al.}
-\cite{shepherd2012c}).
-
 \section{Results}
@@
 
-\begin{align}
-\frac{d E_0}{ds} &= \sum_{ia}\left( \eta_{ia}^{(1)}f_{ai} -
-\eta_{ai}^{(1)}f_{ia}\right) +
-\frac{1}{2}\sum_{ijab}\eta_{ijab}^{(2)}v_{abij},
-\label{eq:flow1}\\
-\frac{d f_{pq}}{ds}&= \sum_r \left( \eta_{pr}^{(1)}f_{rq} +
-\eta_{qr}^{(1)}f_{rp}\right) + \sum_{ia}\left( 1-\hat{P}_{ia}\right)
-\left( \eta_{ia}^{(1)}v_{apiq} - f_{ia}\eta_{apiq}^{(2)} \right)
-\notag \\ & +\frac{1}{2} \sum_{aij} \left( 1+\hat{P}_{pq}\right)
-\eta_{apij}^{(2)}v_{ijaq} + \frac{1}{2}\sum_{abi}\left(
-1+\hat{P}_{pq}\right) \eta_{ipab}^{(2)}v_{abiq},
-\label{eq:flow2}
-\end{align}
-
-\begin{align}
-\frac{d v_{pqrs}}{ds} &= \sum_t \left( 1-\hat{P}_{pq} \right) \left(
-\eta_{pt}^{(1)}v_{tqrs}-f_{pt}\eta_{tqrs}^{(2)}\right) -\sum_t \left(
-1-\hat{P}_{rs} \right) \left( \eta_{tr}^{(1)} v_{pqts} - f_{tr}
-\eta_{pqts}^{(2)}\right) \notag \\ & +\frac{1}{2}\sum_{ab}
-\lb\eta_{pqab}^{(2)} v_{abrs} - v_{pqab}\eta_{abrs}^{(2)}\right) -
-\frac{1}{2}\sum_{ij} \lb\eta_{pqij}^{(2)} v_{ijrs} -
-v_{pqij}\eta_{ijrs}^{(2)}\right) \notag \\ & -\sum_{ia} \left( 1-
-\hat{P}_{ia} \right) \left( 1-\hat{P}_{pq}\right) \left(
-1-\hat{P}_{rs} \right) \eta_{aqis}^{(2)}v_{ipar}.
-\label{eq:flow3}
-\end{align}
-
 \section{Conclusions}

### 2018-02-28 doc/ElectronGas/triplets.tex
added some backup material

	\chapter{Coupled Cluster Theory}

	Coupled-cluster (CC) theory was developed by Fritz Coester and Hermann K\"ummel in the late 1950s and early 1960s (\cite{Coester58},\cite{CoesterKummel60},\cite{Kummel62}) with their work on the exponential ansatz\footnote{This term will be explained soon, as well as the coupled-cluster equations.}, and furthermore in the early 1970s by Hermann K\"ummel and K.H. L\"uhrmann (\cite{Kummel71},\cite{KummeLuhrmann72},\cite{KummeLuhrmann72_2}) with explicit derivation of the coupled-cluster equations. The enthusiasm for the CC method came after Ji\v{r}i \v{C}\'{i}\v{z}ek's Ph.D. thesis in 1965, published in 1966 \cite{Cizek66}, where he used the CC-doubles method to perform several calculations on the benzene molecule.\\

	Coupled-cluster theory has been named the "golden standard" in theoretical chemistry, due its high accuracy with relatively low computational cost. It may be reviewed at great length, but there already exists excellent literature covering the entire field, such as the work by Shavitt and Bartlett \cite{ShavittBartlett09}. Therefore, this chapter will mainly cover that which is deemed necessary at a basic level to understand and apply the method. It will mainly follow the trail of thought presented in \cite{ShavittBartlett09}.

	\section{The exponential Ansatz}
	We start by introducing a very simple way to organise our equations. An arbitrary state can expressed as a linear combination of Slater determinants

	\begin{equation}
		|\Psi\rangle = |\Phi\rangle + \sum_{a=N+1}^\infty\sum_{i=1}^Nc_i^a|\Phi_i^a\rangle + \frac{1}{4}\sum_{a,b=N+1}^\infty\sum_{i,j=1}^Nc_{ij}^{ab}|\Phi_{ij}^{ab}\rangle +\ldots \:,
		\label{EQ 3.1 general WF}
	\end{equation}

	\noindent where $c_{ijk\ldots}^{abc\ldots}$ are the linear coefficients of each base state. This means all states can be expressed from the Hartree-Fock solution. The linear combination can be rewritten by the use of orbital cluster operators:

	\begin{align}
		\hat{T}_1 &\equiv \sum_{a=N+1}^\infty\sum_{i=1}^Nt_i^a a_a^\dagger a_i \:,\\
		\hat{T}_2 &\equiv \frac{1}{2!}\sum_{a,b=N+1}^\infty\frac{1}{2!}\sum_{i,j=1}^Nt_{ij}^{ab} a_a^\dagger a_b^\dagger a_ ia_j \:,\\
		\hat{T}_3 &\equiv \frac{1}{3!}\sum_{a,b,c=N+1}^\infty\frac{1}{3!}\sum_{i,j,k=1}^Nt_{ijk}^{abc} a_a^\dagger a_b^\dagger a_c^\dagger a_ia_ja_k \:.\\
		&\quad\quad\quad\quad\quad\vdots
	\end{align}

	In general we define $\hat{T}_n$ as

	\begin{equation}
		\hat{T}_n \equiv \left(\frac{1}{n!}\right)^2\sum_{a,b,\ldots=N+1}^\infty\sum_{i,j,\ldots=1}^N t_{ij\ldots}^{ab\ldots} a_a^\dagger a_b^\dagger \ldots a_ia_j\ldots \:.
	\end{equation}

	We can only annihilate as many fermions, say $N$, from the reference state $|\Phi\rangle$ as are present in the system. Since $\hat{T}_{N+1}$ annihilates $N+1$ states from $|\Phi\rangle$, leaving zero as the result of annihilating the vacuum, our series ends with $\hat{T}_N$. With these operators, equation (\ref{EQ 3.1 general WF}) can be rewritten as

	\begin{align}
		\begin{split}
		|\Psi\rangle = \Big(\mathds{1} &+ \hat{T}_1 + \frac{1}{2!}\hat{T}_1^2 + \frac{1}{3!}\hat{T}_1^3 + \ldots + \frac{1}{N!}\hat{T}_1^N\\
		&+ \hat{T}_2 + \frac{1}{2}\hat{T}_2^2 + \frac{1}{3!}\hat{T}_2^3 + \ldots + \frac{1}{(N/2)!}\hat{T}_2^{N/2}\\
		&+ \ldots + \hat{T}_N \\
		&+ \hat{T}_2\hat{T}_1 + \frac{1}{2!}\hat{T}_2\hat{T}_1^2 + \ldots\Big)|\Phi\rangle \:.
		\end{split}
	\end{align}

	Notice that we now need to find the coefficients $t_i^a$, $t_{ij}^{ab}$, $\ldots$, rather than $c_i^a$, $c_{ij}^{ab}$, $\ldots$, and that this is the main problem of the CC method. It is customary to call the $t$-coefficients for \emph{amplitudes}. If we now define a new operator $\hat{T} \equiv \hat{T}_1 + \hat{T}_2 + \hat{T}_3 + \ldots + \hat{T}_N$, then the series above may be rewritten:

	\begin{align}
		\begin{split}
		|\Psi\rangle &= \left(\sum_{n=0}^\infty \frac{1}{n!}\hat{T}^n\right)|\Phi\rangle\\
		&= e^{\hat{T}}|\Phi\rangle
		\end{split}
	\end{align}

	\noindent where $\hat{T}$ is called the \emph{cluster operator}. An $N$ fermion system is fully represented by $\hat{T}|\Phi\rangle$, where $\hat{T} = \mathds{1} + \hat{T}_1 +\ldots+\hat{T}_N$.\\

	There is a very fortuitous advantage to the exponential Ansatz, which is that size-consistency follows naturally from it. As discussed in the previous chapter, size-consistency requires that the total energy of two non-interacting systems $A$ and $B$, is

	\begin{equation}
		E(A+B) = E(A) + E(B)
	\end{equation}

	If the two systems are non-interacting, then the ground state ought to be separable:

	\begin{equation}
		\Phi_0(AB) = \Phi_0(A)\Phi_0(B)
	\end{equation}

	Furthermore, if the cluster operator is additive,

	\begin{equation}
		\hat{T}(AB) = \hat{T}(A) + \hat{T}(B)\:,
	\end{equation}

	\noindent then we can derive for the full wavefunction:

	\begin{align}
		\begin{split}
			\Psi(AB) &= e^{\hat{T}(AB)}\Phi_0(AB) \\
			&= e^{\hat{T}(A)}e^{\hat{T}(B)}\Phi_0(A)\Phi_0(B) \\
			&= e^{\hat{T}(A)}\Phi_0(A)e^{\hat{T}(B)}\Phi_0(B) \\
			&= \Psi(A)\Psi(B) 
		\end{split}
	\end{align}

	As we see, even the wavefunctions for the systems are independent. Last, we explicitly check size-consistency;

	\begin{align}
		\begin{split}
			\hat{H}(AB)\Psi(AB) &= \left[ \hat{H}(A) + \hat{H}(B) \right]\Psi(A)\Psi(B) \\
			&= \left[ \hat{H}(A)\Psi(A)\right]\Psi(B) + \Psi(A)\left[\hat{H}(B)\Psi(B)\right] \\
			&= \left[E(A) + E(B)\right]\Psi(AB)
		\end{split}
	\end{align}

	\section{The coupled cluster equations}
	With the exponential ansatz at hand, we can, in theory, tackle any system, since if we introduce more interactions to our model, we simply add more cluster operators. However, the reason we do not, just as for the Hartree-Fock equations, MBPT, and CI, is the huge computational effort required for each new operator introduced. In fact, even the triples operators are so time consuming that they are often omitted in computational studies, not to mention $n\geq 4$ operators. The reason why the computational time increases so rapidly can be seen from the \emph{coupled-cluster equations}.\\

	The coupled-cluster equations are used to find the correlation amplitudes $t_{ij\ldots}^{\alpha\beta\ldots}$, which are needed to calculate the coupled-cluster energy $E_{CC}$. A possible starting point is then:

	\begin{equation}
		\overbar{H}|\Phi_0\rangle = E_{CC}|\Phi_0\rangle
		\label{EQ 3.2 general energy equation}
	\end{equation}

	\noindent where $\bar{H}$ is the similarity transform of the Hamiltonian operator, i.e.

	\begin{equation}
		\overbar{H} = e^{-\hat{T}}\hat{H}_Ne^{\hat{T}}
	\end{equation}

	\noindent where

	\begin{equation}
		\hat{H}_N = \hat{H} - \langle\Phi_0|\hat{H}|\Phi_0\rangle
	\end{equation}

	\noindent is the system Hamiltonian but with the reference energy subtracted. If we now act from the left in equation (\ref{EQ 3.2 general energy equation}) with either $\langle\Phi_0|$ or $\langle\Phi^*| = \langle\Phi_{ij\ldots}^{\alpha\beta\ldots}|$, we get two results:

	\begin{align}
	\langle\Phi_0|\overbar{H}|\Phi_0\rangle &= E_{CC}
	\label{EQ 3.2 general CC equation for energy}\\
	\langle\Phi^*|\overbar{H}|\Phi_0\rangle &= 0
	\label{EQ 3.3 general CC equation for amplitudes}
	\end{align}

	These are the CC equations. The CC amplitudes are unknown, so we need to solve the CC equations to find them, and then use the amplitudes to calculate $E_{CC}$ using equation (\ref{EQ 3.2 general CC equation for energy}). The coupled cluster method therefore seems to consist of writing down all the CC equations (equation (\ref{EQ 3.3 general CC equation for amplitudes})), solving them by finding the correct amplitudes, and finally using the amplitudes to calculate $E_{CC}$. Logically simple but practically difficult.\\

	The big problem is finding the CC amplitudes. Since they are unknown, where do we start? It turns out we may solve them iteratively, by using the amplitudes at one iteration to find the amplitudes at the next. Therefore we need an equation for the iterative process.\\

	Before ending this section, we simplify the problem by deciding to rewrite $\overbar{H}$ using the Baker-Campbell-Hausdorff expansion to get \cite{ShavittBartlett09};

	\begin{equation}
		\overbar{H} = \hat{H}_N + \left[ \hat{H}_N,\hat{T} \right] + \frac{1}{2!}\left[\left[ \hat{H}_N,\hat{T} \right],\hat{T}\right] + \frac{1}{3!}\left[\left[ \left[\hat{H}_N,\hat{T} \right],\hat{T}\right],\hat{T}\right] + \ldots\:,
		\label{EQ 3.2 BCH expansion}
	\end{equation}

	The generalized Wick's theorem states that contractions within a normal-ordered string give zero contribution in a product of normal-ordered strings, i.e:

	\begin{equation}
		\{S_1\}\{S_2\}\{S_3\}\ldots = \{S_1S_2S_3\ldots\} + \sum_{\text{(1)}}\{\contraction[2pt]{}{S}{{}_1}{S}S_1S_2S_3\ldots\} + \sum_{\text{(2)}}\{\contraction[4pt]{S_1S_2}{S}{{}_3}{\ldots}\contraction[2pt]{}{S}{{}_1}{S}S_1S_2S_3\ldots\} + \ldots
	\end{equation}

	\noindent where $S_i$ are strings of operators. Since both $\hat{H}_N$ and $\hat{T}$ are written in normal-ordered form, we express all products in equation (\ref{EQ 3.2 BCH expansion}) using Wick's theorem. From the section on second quantisation, we know the commutation relations of the creation and annihilation operators within $\hat{H}_N$ and $\hat{T}$, and that the only non-zero contractions are:

	\begin{align}
		\contraction[2pt]{}{\hat{a}}{}{\hat{b}}\hat{a}\hat{b}^\dagger &= \delta_{ab}\\
		\contraction[2pt]{}{\hat{i}}{{}^\dagger}{\hat{j}}\hat{i}^\dagger\hat{j} &= \delta_{ij}
	\end{align}

	From these two identities, it is easy to realize that all contractions between cluster operators are zero, and the same goes for all contractions between $\hat{H}_N$ and $\hat{T}$ where $\hat{T}$ is to the left of $\hat{H}_N$. The only remaining, non-zero terms are:

	\begin{align}
		\begin{split}
		\overbar{H} &= \hat{H}_N + \frac{1}{2} \contraction[2pt]{}{\hat{H}}{{}_N}{\hat{T}}\hat{H}_N\hat{T} + \frac{1}{3!}\contraction[2pt]{}{\hat{H}}{{}_N}{\hat{T}}
		\contraction[2pt]{}{\hat{H}}{{}_N\hat{T}}{\hat{T}}\hat{H}_N\hat{T}\hat{T} + \ldots \\
		&= \left(\hat{H}_Ne^{\hat{T}}\right)_C
		\end{split}
	\end{align}

	Here we have introduced the notation $(\ldots)_C$, where the subscript means the leftmost operator is contracted at least once with every operator to the right.\\

	Now, the CC equations look as follows:

	\begin{align}
	\langle\Phi_0|\left(\hat{H}_Ne^{\hat{T}}\right)_C|\Phi_0\rangle &= E_{CC}\\
	\langle\Phi^*|\left(\hat{H}_Ne^{\hat{T}}\right)_C|\Phi_0\rangle &= 0
	\end{align}

	We have greatly reduced the number of terms compared to the earlier CC equations.

	\section{Diagrammatic approach}
	The easiest way to derive the CC amplitude equations is by the use of \emph{Feynman-Goldstone diagrams}. This method is often called the diagrammatic approach, and is favourable because it is visually intuitive, reducing the possibility of algebraic mistakes, and because we automatically sum over equivalent terms in a single diagram, meaning we do not fill a whole library every time we write down the equations. Those acquainted with basic Feynman diagrams, and the reasoning behind them, will recognise these qualities of the diagrammatic approach.

	\subsection{Diagram rules}
	Like Feynman diagrams, CC diagrams (or Goldstone diagrams) follow a series of rules regarding lines, arrows, and vertices. There are 10 rules \cite{ShavittBartlett09} regarding the interpretation of diagrams.

	\fbox{\parbox{14cm}{Rules for interpretation of diagrams:
			\begin{enumerate}
				\item \emph{Labelling}: Open lines are labelled as holes if they point downwards and as particles if they point upwards. These labels are the indices of the "bra" in equation (\ref{EQ 3.3 general CC equation for amplitudes}).

				\item \emph{One-body prefactor}: A one-particle interaction gives a factor $f_{out,in}$.

				\item \emph{Two-body prefactor}: A two-particle interaction gives a factor\\ $\langle\: \text{left-out} \: \text{right-out} \:||\: \text{left-in} \: \text{right-in} \:\rangle$.

				\item \emph{Amplitude prefactors}: A connection gives a factor $t_{ij\ldots}^{ab\ldots}$.

				\item \emph{Summation of indices}: All internal lines, or labels, are in sums.

				\item \emph{Equivalent internal lines}: Equivalent internal lines give a factor $\frac{1}{2}$. Two internal lines are considered equivalent when they point in the same direction and connect the same two vertices.

				\item \emph{Equivalent $T$-vertices}: Equivalent $\hat{T}_m$ vertices give a factor $\frac{1}{2}$. Such vertices are considered equivalent if they have the same number of line pairs and connect in the same way to the interaction vertex.

				\item \emph{Phase factor}: Each diagram has a factor $(-1)^{h-l}$, where $h$ is the number of hole lines and $l$ is the number of loops. External lines are considered "looped", so for a bra $\langle \Phi_{ij}^{ab}|$ we would immediately add +2 to $l$, since $(i,a)$ and $(j,b)$ are considered to form "quasiloops".

				\item \emph{Permutation of external lines}: For a pair of external lines that are connected to the same vertex, we consider them equivalent. We sum over all permutations of inequivalent external lines, with a prefactor $(-1)^{\sigma(P)}$ for each term.

				\item \emph{Factor cancellation}: We cancel a factor $\frac{1}{2}$ caused by equivalent $\hat{T}_m$-vertices for each pair of external lines connected to equivalent vertices.
			\end{enumerate}}}

	\subsection{Sign labels}
	%page 299 Shavitt and Bartlett
	A tool that lets us easily construct all possible Goldstone diagrams for some CC equation is known as \emph{sign labels}. A sign label is an assignment of signs (pluses or minuses) to a vertex diagram. We can then match different vertex diagrams such that the combined diagram has a certain excitation level. For example, if we want to draw all the diagrams included in the equation $\langle\Phi_{ij}^{ab}|\left(\hat{H}_Ne^{\hat
		T}\right)|\Phi_0\rangle = 0$, then we need to include all diagrams where the excitation level is 2, since $\langle\Phi_{ij}^{ab}|$ has excitation level 2. We will explain more further on.\\
	The convention is as follows.\\

	\emph{Sign labels:} To each open hole line we assign a minus, and to each open particle line we assign a plus. Internal lines are not assigned any sign, which we will symbolize by a zero, i.e.\\

	\begin{figure}[h]
		\centering
		$
		\underset{\Scale[1]{(-)}\hspace{0.005cm}}{\bdiags
		\dTtd{1}{t}
		\dTt{1}{tt}
		\dline{t1}{tt1}
		\ediag}
		$
		\hspace{2cm}
		$
		\underset{\Scale[1]{(0)}\hspace{0.005cm}}{\bdiags
			\dnoarrow
			\dTv{1}{t}
			\dTdv{1}{td}
			\dline{t}{td}
			\ediag}
		$
		\hspace{2cm}
		$
		\underset{\Scale[1]{(+)}\hspace{0.005cm}}{\bdiags
			\dTtd{1}{t}
			\dTt{1}{tt}
			\dline{tt1}{t1}
			\ediag}
			$
	\end{figure}

	A good example of the sign convention is the terms in the normal-product form of the Hamiltonian operator ($\hat{H}_N$). We know we can write $\hat{H}_N = \hat{F}_N + \hat{W}$, where we have:

	\begin{equation}
		\hat{F}_N = \sum_{pq} f_{pq}\{\hat{p}^\dagger \hat{q}\} = \sum_{ab} f_{ab}\{\hat{a}^\dagger \hat{b}\} + \sum_{ij} f_{ij}\{\hat{i}^\dagger \hat{j}\} + \sum_{ai} f_{ai}\{\hat{a}^\dagger \hat{i}\} + \sum_{ia} f_{ia}\{\hat{i}^\dagger \hat{a}\} \:,
	\end{equation}

	\noindent as the one-body term, and

	\begin{align}
		\begin{split}
		\hat{W} = \frac{1}{2}\sum_{pqrs}\langle pq||rs\rangle \{\hat{p}^\dagger\hat{q}^\dagger\hat{r}\hat{s}\} = &\frac{1}{4}\sum_{abcd}\langle ab||cd\rangle \{\hat{a}^\dagger\hat{b}^\dagger\hat{d}\hat{c}\} + \frac{1}{4}\sum_{ijkl}\langle ij||kl\rangle \{\hat{i}^\dagger\hat{j}^\dagger\hat{k}\hat{l}\} + \sum_{ijab}\langle ai||bj\rangle \{\hat{a}^\dagger\hat{i}^\dagger\hat{j}\hat{b}\} \\
		+&\frac{1}{2}\sum_{abci}\langle ab||ci\rangle \{\hat{a}^\dagger\hat{b}^\dagger\hat{i}\hat{c}\} + \frac{1}{2}\sum_{ijka}\langle ia||jk\rangle \{\hat{i}^\dagger\hat{a}^\dagger\hat{k}\hat{j}\} + \frac{1}{2}\sum_{abci}\langle ai||bc\rangle \{\hat{a}^\dagger\hat{i}^\dagger\hat{c}\hat{b}\} \\
		+&\frac{1}{2}\sum_{ijka}\langle ij||ka\rangle \{\hat{i}^\dagger\hat{j}^\dagger\hat{a}\hat{k}\} + \frac{1}{4}\sum_{abij}\langle ab||ij\rangle \{\hat{a}^\dagger\hat{b}^\dagger\hat{j}\hat{i}\} + \frac{1}{4}\sum_{ijab}\langle ij||ab\rangle \{\hat{i}^\dagger\hat{j}^\dagger\hat{b}\hat{a}\} \:,
		\end{split}
	\end{align}

	\noindent as the two-body term of the normal-ordered Hamiltonian. Perhaps not pretty, but certainly systematic. We are now curious as to how this looks in diagrammatic form. Recalling that annihilation operators remove states that are already present in $|\Phi\rangle$ (our reference Slater determinant), the annihilation operators symbolise states that are "inbound" to some interaction vertex, i.e. lines coming in from below. In the same manner, the creation operator creates states, "outbound" states, meaning lines that move from the vertex and up. If we study equations 9.105 and 9.106 in \cite{ShavittBartlett09}, shown below, it is easy to understand how the diagrams represent the algebraic terms.

	\input{One_body_diagrams}
	\input{Two_body_diagrams}

	We see that it is fairly easy to interpret the diagrams in this simple form. Consider now the signs of each diagram:

	\input{Hamiltonian_signs}
	\newpage
	Since the CC equations consist of terms where the Hamiltonian is contracted with orbital cluster operators, we need to know how to construct the CC operators as diagrams. It is important to note that since there are always summations of the indices in the CC operators, we need not write arrows on the lines, only how many of each sign there are\footnote{If we really wanted to, we would be free to write arrows. It is just unnecessary.}. Below we show the first 3 cluster operator vertices.

	\begin{figure}[h]
		\centering
		$
		\underset{+\quad-}{\bdiag
		\dhscale{1.5}
		\dTs{1}{t}
		\dnoarrow
		\dline{tv1}{t}
		\dline{t}{tv2}
		\ediag}
		$
		\hspace{2cm}
		$
		\underset{+\quad+\quad-\quad-}{\bdiag
		\dhscale{1.5}
		\dTs{2}{tt}
		\dnoarrow
		\dline{tt1v1}{tt1}
		\dline{tt1}{tt1v2}
		\dline{tt2v1}{tt2}
		\dline{tt2}{tt2v2}
		\ediag}
		$
		\hspace{2cm}
		$
		\underset{+\quad+\quad+\quad-\quad-\quad-}{\bdiag
		\dhscale{1.5}
		\dTs{3}{ttt}
		\dnoarrow
		\dline{ttt1v1}{ttt1}
		\dline{ttt1}{ttt1v2}
		\dline{ttt2v1}{ttt2}
		\dline{ttt2}{ttt2v2}
		\dline{ttt3v1}{ttt3}
		\dline{ttt3}{ttt3v2}
		\ediag}
		$
	\end{figure}

	We now have all we need to start constructing diagrams. As explained at the start of this section, we need to match vertex diagrams to form combined diagrams with a certain excitation level. All combined diagrams are then added to give all the terms of the same CC equation. Since the energy equation has an excitation level of zero, then the amplitude equations follow in ascending order of excitation.\\

	An example of how we may now proceed might be useful. Say we wish to construct the diagram that gives the term $\hat{H}_{N,8}\hat{T}_1\hat{T}_2$, where $\hat{H}_{N,8}$ is the 8th diagram of the above Hamiltonian diagrams. Below we show all distinct diagrams as well as their algebraic interpretations.

	\begin{figure}[h]
		\centering
		$\bdiag
			\dmoveH{1}
			\dT{1}{t}
			\dWs{w1}{w2}
			\dmoveT{1}
			\dT{2}{tt}
			\dcurcur{w1}{t}
			\dcurcur{t}{w1}
			\dline{tt1}{w2}
			\dline[$\:a$]{w2}{w2v1}
			\dline[$\:i$]{tt1v2}{tt1}
			\dline[\raisebox{1cm}{$j$}]{tt2v1}{tt2}
			\dline[$\:b$]{tt2}{tt2v2}
			\ediag
		$
		\hspace{2cm}
		$
		\bdiag
		\dmoveH{1.65}
		\dT{1}{t}
		\dWs{w1}{w2}
		\dmoveT{0.8}
		\dT{2}{tt}
		\dline[\raisebox{1cm}{$i$}]{tv1}{t}
		\dline{t}{w1}
		\dline[$\:a$]{w1}{tv2}
		\dcurcur{tt1}{w2}
		\dcurcur{w2}{tt1}
		\dline[\raisebox{1cm}{$j$}]{tt2v1}{tt2}
		\dline[$\:b$]{tt2}{tt2v2}
		\ediag
		$
		\hspace{2cm}
		$
		\bdiag
		\dmoveH{0.6}
		\dT{1}{t}
		\dscaleop{1.55}
		\dWs{w1}{w2}
		\dscaleop{1}
		\dmoveT{1}
		\dT{2}{tt}
		\dline[\raisebox{1cm}{$i$}]{tv1}{t}
		\dline{t}{w1}
		\dline[$\:a$]{w1}{tv2}
		\dline[\raisebox{1cm}{$\:j$}]{tt1v1}{tt1}
		\dline{tt1}{w2}
		\dline{w2}{tt2}
		\dline[$\:b$]{tt2}{tt2v2}
		\ediag
		$
	\end{figure}

	Finally, there is also the diagram:

	\begin{figure}[h]
		\centering
		$
		\bdiag
		\dmoveH{6}
		\dT{1}{t}
		\dscaleop{0.9}
		\dWs{w1}{w2}
		\dscaleop{1}
		\dmoveT{1}
		\dT{2}{tt}
		\dline[\raisebox{1cm}{$i$}]{tv1}{t}
		\dline[$\:a$]{t}{tv2}
		\dcurcur{tt1}{w1}
		\dcurcur{w1}{tt1}
		\dline[$\:j$]{tt2v2}{tt2}
		\dline{tt2}{w2}
		\dline[$\:b$]{w2}{tt2v1}
		\ediag
		$
	\end{figure}

	There is, however, one last theorem we need in order to fully utilize the diagrammatic method. Up until now, we have not made any distinction between unlinked diagrams, where we multiply two separate diagrams to give us the desired excitation level, and linked diagrams, where all vertices are connected (or linked) in some way. Above, we had three linked diagrams and one unlinked. While this does not seem so bad, we would find that the number of diagrams increases dramatically if we always included all possible diagrams with excitation level 2. Fortunately, there is exists a theorem, named the \emph{linked-diagram theorem}, which states that only terms associated with linked diagrams contribute to the wavefunction energy. This means that any composite diagram, made up of two or more closed diagrams which are not all connected in some way, \emph{will not} contribute to the CC energy. The proof is not too long, but we shall simply refer the reader to chapters 5 and 6 of \cite{ShavittBartlett09}.

	\section{Coupled-cluster models}
	Now that we have covered the basic formalism of the CC method, we can move onto their analytical expressions. We list the tables in the CCD and CCDT models\footnote{These are the models we shall be using. Other methods are listed in detail in \cite{ShavittBartlett09}.} below such that the reader may easily which of the CC diagrammatic rules are used to give which factors. To perform a CC analysis, we shall have to implement the following expressions in computer code.

	\subsection{A first example: coupled-cluster doubles (CCD)}
	With the rules for constructing and interpreting the CC diagrams, we can begin looking at various coupled-cluster models, starting with the so-called Coupled-Cluster doubles equations. If we set $\hat{T} = \hat{T}_2$, we choose to consider CCD. That is, we consider only double excitations in our CC operator. This means our amplitude equations are given by the sum of all possible linked diagrams with excitation level 2. In table \ref{CC | table | "CCD amp eq derivation"} below, we show all CCD amplitude diagrams together with their algebraic interpretation in such a manner that it is easy to see which rule gives which factor. Note that rule 10 does not apply to any diagram here.

	\begin{table}[h]
	\centering
	\caption{All terms for the CCD amplitude equations. Note that rule 10 does nothing here.}
	\begin{tabular}{cccccccccc}
		Name & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & \\ \hline
		& $f_{k,i}$ &  & $t_{jk}^{ab}$ & $\sum_{k}$ &  &  & $(-1)^{2-2}$ & $P(ij)$ &\\
		& $f_{a,c}$ &  & $t_{ij}^{cb}$ & $\sum_{c}$ &  &  & $(-1)^{3-2}$ & $P(ab)$ &\\
		$L_a$ & & $\langle kl||ij \rangle$ & $t_{kl}^{ab}$ & $\sum_{kl}$ & $\frac{1}{2}$ &  & $(-1)^{2-2}$ &  &\\
		$L_b$ & & $\langle ab||cd \rangle$ & $t_{ij}^{cd}$ & $\sum_{cd}$ & $\frac{1}{2}$ &  & $(-1)^{2-2}$ &  &\\
		$L_c$ & & $\langle kb||ic \rangle$ & $t_{kj}^{ac}$ & $\sum_{ck}$ &  &  & $(-1)^{3-2}$ & $P(ab)P(ij)$ &\\
		$Q_a$ & & $\langle kl||cd \rangle$ & $t_{kl}^{ab}$ $t_{ji}^{dc}$ & $\sum_{klcd}$ & $\frac{1}{4}$ &  & $(-1)^{4-2}$ &  &\\
		$Q_b$ & & $\langle kl||cd \rangle$ & $t_{ki}^{ac}$ $t_{lj}^{bd}$ & $\sum_{klcd}$ &  & $\frac{1}{2}$ & $(-1)^{4-4}$ & $P(ab)P(ij)$ &\\
		$Q_c$ & & $\langle kl||cd \rangle$ & $t_{ki}^{cd}$ $t_{lj}^{bd}$ & $\sum_{klcd}$ & $\frac{1}{2}$ &  & $(-1)^{4-3}$ & $P(ij)$ &\\
		$Q_d$ & & $\langle kl||cd \rangle$ & $t_{lk}^{ac}$ $t_{ji}^{bd}$ & $\sum_{klcd}$ & $\frac{1}{2}$ &  & $(-1)^{4-3}$ & $P(ab)$ &\\ \hline
	\end{tabular}
	\label{CC | table | "CCD amp eq derivation"}
	\end{table}

	By adding all CCD amplitude terms, we now arrive at\footnote{We must also add the lone Hamiltonian term, which is the first term seen here.}

	\begin{align}
		\begin{split}
		\langle ab||ij\rangle &+ \hat{P}(ab)\sum_{c}f_{bc}t_{ij}^{ac} - \hat{P}(ij)\sum_{k}f_{kj}t_{kj}^{ab}\\
		&+ \frac{1}{2}\sum_{cd}\langle ab||cd\rangle t_{ij}^{cd} + \frac{1}{2}\sum_{kl}\langle kl||ij\rangle t_{kl}^{ab} + \hat{P}(ij|ab)\sum_{kc}\langle kb||cj\rangle t_{ik}^{ac}\\
		&+ \frac{1}{4}\sum_{klcd}\langle kl||cd\rangle t_{ij}^{cd}t_{kl}^{ab} + \hat{P}(ij)\sum_{klcd}\langle kl||cd\rangle t_{ik}^{ac}t_{jl}^{bd}\\
		&- \frac{1}{2}\hat{P}(ij)\sum_{klcd}\langle kl||cd\rangle t_{ik}^{dc}t_{lj}^{ab} - \frac{1}{2}\hat{P}(ab)\sum_{klcd}\langle kl||cd\rangle t_{lk}^{ac}t_{ij}^{db}\\
		&= 0\:.
		\end{split}
	\end{align}

	As mentioned earlier, with a canonical Hartree-Fock basis, all off-diagonal elements of the Fock matrix ($f_{\alpha\beta}$) are zero, and only the self-energy terms remain, which we now denote $\epsilon_i \equiv f_{ii} = h_i + \sum_{j}\langle ij||ij\rangle$. We may then rewrite the above equation as

	\begin{equation}
		(\epsilon_i+\epsilon_j-\epsilon_a-\epsilon_b)t_{ij}^{ab} = \langle ab||ij\rangle + L(t) + Q(t) \:,
	\end{equation}

	\noindent or

	\begin{equation}
		t_{ij}^{ab} = \frac{\langle ab||ij\rangle + L(t) + Q(t)}{\epsilon_i+\epsilon_j-\epsilon_a-\epsilon_b} \:,
		\label{CC | eq | "CCD amplitudes equation"}
	\end{equation}

	\noindent where $L(t)$ is the sum over all the ladder diagrams ($L_a$ to $L_c$) and $Q(t)$ is the sum over all quadratic $\hat{T}_2$ terms ($Q_a$ to $Q_d$). Each term is already shown in table \ref{CC | table | "CCD amp eq derivation"}.

	\subsection{Coupled-cluster singles and doubles (CCSD)}
	Deriving the CCSD equations is not very difficult, but it requires much caution as we are very prone to mistakes. For a full derivation we simply refer the reader to excellent derivations as presented in \cite{ShavittBartlett09} and \cite{Hansen15}. As we now have two different kinds of amplitudes, $t_i^a$ and $t_{ij}^{ab}$, we get two distinct sets of amplitude equations. We can further note that the amplitudes contribute to one another, which we already expected from the exponential ansatz and the contracted form of the CC equations ($\langle\Phi^*|\{\hat{H}_Ne^{\hat{T}}\}_C|\Phi_0\rangle = 0$).

	\subsection{Coupled-cluster doubles and triples (CCDT)}
	We define the cluster operator as $\hat{T} = \hat{T}_2 + \hat{T}_3$. In addition to the CCD diagrams (table \ref{CC | table | "CCD amp eq derivation"}), we get two additional diagrams that contribute to the $T_2$ amplitudes, shown in table \ref{CC | table | "CCDT T2 amp eq derivation"}. In table \ref{CC | table | "CCDT T3 amp eq derivation"} we show all $T_3$ amplitude diagrams.

	\begin{table}[h]
		\centering
		\caption{$T_3$ contributions to $T_2$ amplitude equations. Note that rules 6, 7, and 10 do nothing here.}
		\begin{tabular}{cccccccccc}
			Name & 2 & 3 & 4 & 5 & 8 & 9 & \\ \hline
			$D_{10b}$ & & $\langle ak||cd \rangle$ & $t_{ijk}^{cdb}$ & $\sum_{cdk}$ & $(-1)^{2-2}$ & $P(ab)$ &\\
			$D_{10c}$ & & $\langle kl||cj \rangle$ & $t_{ikl}^{cab}$ & $\sum_{ckl}$ & $(-1)^{3-2}$ & $P(ij)$ &\\ \hline
		\end{tabular}
		\label{CC | table | "CCDT T2 amp eq derivation"}
	\end{table}

	\begin{table}[h]
		\centering
		\caption{All terms for the $T_3$ amplitude equations in the CCDT model. Note that rules 7 and 10 do nothing here.}
		\begin{tabular}{ccccccccc}
			Name & 2 & 3 & 4 & 5 & 6 & 8 & 9 & \\ \hline
			& $f_{c,d}$ & & $t_{ijk}^{abd}$ & $\sum_d$ & & $(-1)^{3-3}$ & $P(c/ab)$ & \\
			& $f_{l,k}$ & & $t_{ljk}^{abc}$ & $\sum_l$ & & $(-1)^{4-3}$ & $P(k/ij)$ & \\
			$T_{1a}$ & & $\langle bc||dk \rangle$ & $t_{ij}^{ad}$ & $\sum_{d}$ & & $(-1)^{3-3}$ & $P(a/bc)P(k/ij)$ &\\
			$T_{1b}$ & & $\langle lc||jk \rangle$ & $t_{il}^{ab}$ & $\sum_{l}$ & & $(-1)^{4-3}$ & $P(c/ba)P(i/jk)$ &\\
			$T_{2c}$ & & $\langle ab||de \rangle$ & $t_{ijk}^{dec}$ & $\sum_{de}$ & $\frac{1}{2}$ & $(-1)^{3-3}$ & $P(c/ab)$ &\\
			$T_{2d}$ & & $\langle lm||ij \rangle$ & $t_{lmk}^{abc}$ & $\sum_{lm}$ & $\frac{1}{2}$ & $(-1)^{5-3}$ & $P(k/ij)$ &\\
			$T_{2e}$ & & $\langle al||id \rangle$ & $t_{ljk}^{dbc}$ & $\sum_{ld}$ & & $(-1)^{4-4}$ & $P(a/bc)P(i/jk)$ &\\
			$T_{3b}$ & & $\langle lb||de \rangle$ & $t_{id}^{al}t_{jk}^{ec}$ & $\sum_{dle}$ & & $(-1)^{4-4}$ & $P(abc)P(i/jk)$ &\\
			$T_{3c}$ & & $\langle lm||dj \rangle$ & $t_{id}^{al}t_{mk}^{bc}$ & $\sum_{dlm}$ & & $(-1)^{5-4}$ & $P(a/bc)P(ijk)$ &\\
			$T_{3d}$ & & $\langle lc||de \rangle$ & $t_{il}^{ab}t_{jk}^{de}$ & $\sum_{dle}$ & $\frac{1}{2}$  & $(-1)^{4-3}$ & $P(c/ba)P(i/jk)$ &\\
			$T_{3e}$ & & $\langle lm||dk \rangle$ & $t_{ij}^{ad}t_{lm}^{bc}$ & $\sum_{dlm}$ & $\frac{1}{2}$ & $(-1)^{5-3}$ & $P(a/bc)P(k/ij)$ &\\
			$T_{5a}$ & & $\langle lm||de \rangle$ & $t_{il}^{ad}t_{mjk}^{ebc}$ & $\sum_{lmde}$ & & $(-1)^{5-5}$ & $P(a/bc)P(i/jk)$ &\\
			$T_{5b}$ & & $\langle lm||de \rangle$ & $t_{li}^{de}t_{mjk}^{abc}$ & $\sum_{lmde}$ & $\frac{1}{2}$ & $(-1)^{5-4}$ & $P(i/jk)$ &\\
			$T_{5c}$ & & $\langle lm||de \rangle$ & $t_{lm}^{da}t_{ijk}^{ebc}$ & $\sum_{lmde}$ & $\frac{1}{2}$ & $(-1)^{5-4}$ & $P(a/bc)$ &\\
			$T_{5d}$ & & $\langle lm||de \rangle$ & $t_{ij}^{ad}t_{lmk}^{bec}$ & $\sum_{lmde}$ & $\frac{1}{2}$ & $(-1)^{5-4}$ & $P(a/bc)P(k/ij)$ &\\
			$T_{5e}$ & & $\langle lm||de \rangle$ & $t_{il}^{ab}t_{jmk}^{dec}$ & $\sum_{lmde}$ & $\frac{1}{2}$ & $(-1)^{5-4}$ & $P(c/ab)P(i/jk)$ &\\
			$T_{5f}$ & & $\langle lm||de \rangle$ & $t_{ij}^{de}t_{lmk}^{abc}$ & $\sum_{lmde}$ & $\frac{1}{4}$ & $(-1)^{5-3}$ & $P(k/ij)$ &\\
			$T_{5g}$ & & $\langle lm||de \rangle$ & $t_{lm}^{ab}t_{ijk}^{dec}$ & $\sum_{lmde}$ & $\frac{1}{4}$ & $(-1)^{5-3}$ & $P(c/ab)$ &\\ \hline
		\end{tabular}
		\label{CC | table | "CCDT T3 amp eq derivation"}
	\end{table}

	We have only listed the CCDT amplitude terms, which might seem strange. After all, by writing the full CCSDT equations, we can easily get the CCDT equations by setting all singles amplitudes ($t_i^a$) to zero. However, as we have mentioned, working with Bloch wavefunctions means we already have a Hartree-Fock basis. With a Hartree-Fock basis, all off-diagonal Fock matrix elements ($f_{p,q}$) are zero\footnote{Remember that a Hartree-Fock basis is found by diagonalising the Fock matrix.}.

	\section{Convergence of iterative series}
	We have not mentioned exactly how the CC amplitude equations are solved, but we did mention that it was important to write them in the form of equation (\ref{CC | eq | "CCD amplitudes equation"}). We say we have a self-consistent equation. Every $(t_{ij\ldots}^{ab\ldots})_{n+1}$ is defined from $(t_{ij\ldots}^{ab\ldots})_{n}$, and begins with some reasonable initial guess. In the case of CCD, we start with MBPT2, as mentioned. We keep going through this algorithm until a constant expression for $E_{CC}$ is reached, which rarely takes more than $50$ iterations. However, sometimes we get divergent results. This is typical for the nuclear potentials we will discuss later on. To deal with divergences, a suppression is put on the equations. That is, we use a \emph{relaxation parameter} $\alpha$ such that

	\begin{equation}
		(t_{ij\ldots}^{ab\ldots})_{n+1} = \alpha(t_{ij\ldots}^{ab\ldots})_{n+1} + (1-\alpha)(t_{ij\ldots}^{ab\ldots})_{n}
	\end{equation}

	This relaxation parameter reduces the "oscillation" the CC iterations as they converge towards self-consistent solutions. Furthermore, the relaxation reduces the number of iterations greatly, for a fine-tuned $\alpha$.

	Most of the doubles and triples diagrams require one alignment. The $T_{3b-e}$-diagrams require an additional realignment in-between matrix products, as there is no possible arrangement of indices for all three matrices that allows for a matrix contraction. This realignment can be problematic when it comes to parallelisation, but we shall discuss this point further later on. In table \ref{Implementation | table | "alignments"}, we show all alignments as they have been implemented. It is important to remember that, even though a matrix has been aligned, the elements still need to appear in the correct positions for the matrix product to be correct. This is where the use of \emph{dictionaries}, or \emph{maps}, is important.\\

	\begin{table}[h]
		\centering
		\captionsetup{width=.8\textwidth}
		\caption{The alignments, realignments (for $T_{3b-e}$), and product forms for the CCDT implementation. Symbols with $\tilde{a}$ are aligned, while those without are in their original form. Note that some diagrams require no ($L_a,\:L_b,\:Q_a$) or little ($T_{5f},\:T_{5g}$) alignment.}
		\begin{tabular}{ccccc}
			Diagram & Unaligned & Aligned & Realignment & Product \\ \hline
			$L_a$ & $v_{ab}^{cd}t_{ij}^{cd}$ & $v_{ab}^{cd}t_{ij}^{cd}$ & & $A_{ij}^{ab}$\\
			$L_b$ & $v_{kl}^{ij}t_{kl}^{ab}$ & $v_{kl}^{ij}t_{kl}^{ab}$ & & $A_{ij}^{ab}$\\
			$L_c$ & $v_{ak}^{cj}t_{ik}^{cb}$ & $\tilde{v}_{aj}^{ck}\tilde{t}_{ck}^{ib}$ & & $\tilde{A}_{ib}^{aj}$\\
			$Q_a$ & $t_{ij}^{cd}v_{kl}^{cd}t_{kl}^{ab}$ & $t_{ij}^{cd}v_{kl}^{cd}t_{kl}^{ab}$ & & $A_{ij}^{ab}$\\
			$Q_b$ & $t_{ik}^{ca}v_{kl}^{cd}t_{jl}^{db}$ & $\tilde{t}_{kc}^{ai}\tilde{v}_{kc}^{dl}\tilde{t}_{jb}^{dl}$ & & $\tilde{A}_{jb}^{ai}$\\
			$Q_c$ & $t_{kl}^{ca}v_{kl}^{cd}t_{ij}^{db}$ & $\tilde{t}_{klc}^{a}\tilde{v}_{klc}^{d}\tilde{t}_{ijb}^{d}$ & & $\tilde{A}_{ijb}^a$ \\
			$Q_d$ & $t_{ik}^{cd}v_{kl}^{cd}t_{jl}^{ab}$ & $\tilde{t}_{i}^{cdk}\tilde{v}_{l}^{cdk}\tilde{t}_{l}^{abj}$ & & $\tilde{A}_{i}^{abj}$ \\
			$D_{10b}$ & $v_{ak}^{cd}t_{ijk}^{cdb}$ & $\tilde{v}^{cdk}_{a}\tilde{t}_{ijb}^{cdk}$ & & $\tilde{A}_{ijb}^a$ \\
			$D_{10c}$ & $v_{kl}^{jc}t_{ikl}^{cab}$ & $\tilde{v}_{klc}^{j}\tilde{t}_{klc}^{abi}$ & & $\tilde{A}_{j}^{abi}$ \\
			$T_{1a}$ & $v_{bc}^{dk}t_{ij}^{ad}$ & $\tilde{v}_{bck}^{d}\tilde{t}_{ija}^{d}$ & & $\tilde{A}_{bck}^{ija}$ \\
			$T_{1b}$ & $v_{lc}^{jk}t_{il}^{ab}$ & $\tilde{v}_{l}^{jkc}\tilde{t}_{l}^{abi}$ & & $\tilde{A}_{jkc}^{abi}$ \\
			$T_{2c}$ & $v_{ab}^{de}t_{ijk}^{dec}$ & $\tilde{v}_{ab}^{de}\tilde{t}_{ijkc}^{de}$ & & $\tilde{A}_{ijkc}^{ab}$ \\
			$T_{2d}$ & $v_{ij}^{lm}t_{lmk}^{abc}$ & $\tilde{v}_{ij}^{lm}\tilde{t}_{lm}^{abck}$ & & $\tilde{A}_{ij}^{abck}$ \\
			$T_{2e}$ & $v_{id}^{al}t_{ljk}^{dbc}$ & $\tilde{v}_{ia}^{ld}\tilde{t}_{ld}^{bcjk}$ & & $\tilde{A}_{ia}^{bcjk}$ \\
			$T_{3b}$ & $t_{il}^{ad}v_{de}^{bl}t_{jk}^{ec}$ & $\left(\tilde{t}_{ld}^{ai}\tilde{v}_{eb}^{ld}\right)t_{jk}^{ec}$ & $(\tilde{t}\tilde{v})_{bai}^{e}\tilde{t}_{jkc}^{e}$ & $\tilde{A}_{jkc}^{bai}$ \\
			$T_{3c}$ & $t_{il}^{ad}v_{lm}^{jd}t_{mk}^{bc}$ & $\left(\tilde{t}_{ia}^{dl}\tilde{v}_{mj}^{dl}\right)t_{jk}^{ec}$ & $(\tilde{t}\tilde{v})_{ija}^{m}\tilde{t}_{m}^{bck}$ & $\tilde{A}_{ija}^{bck}$ \\
			$T_{3d}$ & $t_{il}^{ab}v_{de}^{cl}t_{jk}^{de}$ & $t_{il}^{ab}\left(\tilde{v}_{de}^{cl}\tilde{t}_{jk}^{de}\right)$ & $\tilde{t}_{l}^{abi}(\tilde{v}\tilde{t})_{jkc}^{l}$ & $\tilde{A}_{jkc}^{abi}$ \\
			$T_{3e}$ & $t_{ij}^{ad}v_{lm}^{kd}t_{lm}^{bc}$ & $\left(\tilde{t}_{ija}^{d}\tilde{v}_{lmk}^{d}\right)t_{lm}^{bc}$ & $(\tilde{t}\tilde{v})_{ijka}^{lm}\tilde{t}_{lm}^{bc}$ & $\tilde{A}_{ijka}^{bc}$ \\
			$T_{5a}$ & $t_{il}^{ad}v_{lm}^{de}t_{mjk}^{ebc}$ & $\tilde{t}_{ia}^{dl}\tilde{v}_{me}^{dl}\tilde{t}_{me}^{bcjk}$ & & $\tilde{A}_{ia}^{bcjk}$ \\
			$T_{5b}$ & $t_{li}^{de}v_{lm}^{de}t_{mjk}^{abc}$ & $\tilde{t}_{i}^{del}\tilde{v}_{m}^{del}\tilde{t}_{m}^{abcjk}$ & & $\tilde{A}_{i}^{abcjk}$ \\
			$T_{5c}$ & $t_{lm}^{da}v_{lm}^{de}t_{ijk}^{ebc}$ & $\tilde{t}_{lmd}^{a}\tilde{v}_{lmd}^{e}\tilde{t}_{ijkbc}^{e}$ & & $\tilde{A}_{ijkbc}^{a}$ \\
			$T_{5d}$ & $t_{ij}^{ad}v_{lm}^{de}t_{lmk}^{bec}$ & $\tilde{t}_{ija}^{d}\tilde{v}_{lme}^{d}\tilde{t}_{lme}^{bck}$ & & $\tilde{A}_{ija}^{bck}$ \\
			$T_{5e}$ & $t_{il}^{ab}v_{lm}^{de}t_{jmk}^{dec}$ & $\tilde{t}_{l}^{abi}\tilde{v}_{l}^{dem}\tilde{t}_{jkc}^{dem}$ & & $\tilde{A}_{jkc}^{abi}$ \\
			$T_{5f}$ & $t_{ij}^{de}v_{lm}^{de}t_{lmk}^{abc}$ & $t_{ij}^{de}v_{lm}^{de}\tilde{t}_{lm}^{abck}$ & & $\tilde{A}_{ij}^{abck}$ \\
			$T_{5g}$ & $t_{lm}^{ab}v_{lm}^{de}t_{ijk}^{dec}$ & $t_{lm}^{ab}v_{lm}^{de}\tilde{t}_{ijkc}^{de}$ & & $\tilde{A}_{ab}^{ijkc}$
		\end{tabular}
		\label{Implementation | table | "alignments"}
	\end{table}

	A dictionary is a object used in computing that, given some \emph{key}, returns a value. That is, we can make a dictionary $D$ that looks as

	\begin{center}
	\begin{tabular}{cc}
		key & value \\
		car & 12 \\
		candlestick & -17 \\
		frumblesnuff & 34
	\end{tabular}
	\end{center}

	Now, if we call $D(\text{candlestick})$, we get back the value $-17$. Of course, the keys can be whatever we desire, like string, integers, floats, etc, and the same goes for the values. In our case, the keys will be unique identifiers for a set of states. For example, the $T_2$ amplitudes can be gotten for a set of states $i$, $j$, $a$, and $b$ by some unique identifier,

	\begin{equation}
		Id(i,j,a,b) = a + bN_p + iN_p^2 + jN_p^2N_h\:,
	\end{equation}

	which will serve as integer keys. Obviously the amplitudes themselves will be floats, so the values of the map will be floats. This extends easily to all the other matrices as well. In C++, we use the standard library maps, which has the exact same functionality as dictionaries.\\

	Consider a dictionary $D$, that takes 6 arguments, $\{i,\:j,\:k,\:a,\:b,\:c\}$, and returns an index, which, when used together with an array that holds the $T_3$ amplitudes, gives us element $t_{ijk}^{abc}$. If we let $A$ be the array of amplitudes, then

	\begin{align}
	D(i,j,k,a,b,c) &= I \\
	A(I) = t_{ijk}^{abc}
	\label{Implementation | eq | "T3 element retrieval"}
	\end{align}

	Our matrices are constructed with the use of index matrices, such as those we explained previously. For example, the unaligned matrix $t_{ijk}^{abc}$ would have rows where each row is some selection $\{ijk\}$, where\footnote{Later on, when we start discussing index matrices, we shall be using phrases such as \texttt{ppmm\_hhpp} to signify which how we have done shift of unique identifiers. The notation means plus-plus-minus-minus for hole-hole-particle-particle, i.e. $+k_u(h1)+k_u(h2)-k_u(p1)-k_u(p2)$.}

	\begin{equation}
	k_i+k_j+k_k = k_u
	\end{equation}

	The columns would be the same, such that each column is some selection $\{abc\}$ where

	\begin{equation}
	k_a+k_b+k_c = k_u
	\end{equation}

	In order to insert the correct amplitude for a row and column, we have to find it in the amplitude array $A$. From the column and row, we can figure out which indices we have (due to the index matrices predefined), and insert them into the dictionary $D$, just like we described above. If we do this for the whole matrix, we will have constructed channel $k_u$ for the $T_3$ matrix.\\

	It should be apparent that this approach does not change with alignment. If we instead want to make a matrix in the shape of $t_{ij}^{abck}$, then while we will be working with another set of channels and index matrices, the approach is the same. However, since the CC diagrams contain no apparent repeated manner in which the index summations are done, it is difficult to make a general alignment function that places the amplitudes correctly in matrices for all diagrams. This is why, for our implementation, each diagram has its own matrix construction function. A golden rule when it comes to code writing is to never write repeated code, as the possibility of error increases with each line, but we found no way of avoiding it.\\

	This approach is the one used throughout our implementation. Although we will discuss the downsides of the use of dictionaries later on, we note that making a matrix from an array \emph{must} use some sort of search function, whether it is done through the std library, some other library, or our own search function, perhaps even in parallel. Such a search function will use time, and the most speedup to be gained in any implementation like this is by optimising the search function.

	\subsection{Permutations}
	\label{Implementation | sec | "permutations"}
	Most of the CC diagrams have permutations of indices. In the CCD diagrams, the possible permutations are $\hat{P}_{ij}$, $\hat{P}_{ab}$, and $\hat{P}_{ij}\hat{P}_{ab}$. This is not very many possible permutations, and they are quite simple to perform. That is, if we stored the terms from a diagram sum in a dictionary, we could perform a permutation by simply calling different elements of the sum. Assume we are calculating a diagram in some form

	\begin{equation}
		t_{ij}^{ab} \leftarrow \hat{P}D_{ij}^{ab}\:,
	\end{equation}

	where $D_{ij}^{ab}$ is the outcome of the summation of internal indices, and we have stored if as a map, say $M_D$, using the same identity function to generate keys as the one we use for the $T_2$ amplitude map. If now $\hat{P} = \hat{P}_{ab} = 1-P_{ab}$, then the permutation is simply

	\begin{equation}
		t_{ij}^{ab} \leftarrow \left[D_{ij}^{ab} - D_{ij}^{ba}\right] \:,
	\end{equation}

	where we find the two elements $D_{ij}^{ab}$, $D_{ij}^{ba}$ by

	\begin{align}
		\begin{split}
		D_{ij}^{ab} &= M_D\left(\text{id}(i,j,a,b)\right) \\
		D_{ij}^{ba} &= M_D\left(\text{id}(i,j,b,a)\right)
		\end{split}
	\end{align}

	We have not bothered with a more optimised scheme with which to do the CCD permutations, since, as we have mentioned at the start of this chapter, the $T_2$ amplitudes are few in comparison to the $T_3$ amplitudes. This means that the $T_2$ diagram maps are much smaller and take much less time to search through, than the $T_3$ diagram maps. In CCDT, however, there are up to 4 permutation operators on a single diagrams. Take, for example, the diagram $T_{1a}$:

	\begin{align}
		\begin{split}
		T_{1a} :&= \hat{P}(a/bc)\hat{P}(k/ij)\sum_d \langle bc|| dk\rangle t_{ij}^{ad} \\
		& = (1-\hat{P}_{ab} - \hat{P}_{ac})(1-\hat{P}_{ki}-\hat{P}_{kj})D_{ijk}^{abc} \\
		&= \big[1-\hat{P}_{ab}-\hat{P}_{ac}-\hat{P}_{ki}-\hat{P}_{kj} \\
		&\quad\quad\: + \hat{P}_{ab}\hat{P}_{ki} + \hat{P}_{ab}\hat{P}_{kj} + \hat{P}_{ac}\hat{P}_{ki} + \hat{P}_{ac}\hat{P}_{kj} \big]D_{ijk}^{abc}
		\end{split}
	\end{align}

	where $D_{ijk}^{abc}$ is the sum for some diagram. If we were still to use dictionaries to perform permutations, we would have to do as many search calls as there are terms in the equation above. As stated, we would like to minimise search calls as much as possible, so an alternate approach is better. In the process of writing, we attempted two different approaches, mostly out of necessity.\\

	The first approach consisted of storing the permutations beforehand. This is very memory intensive for aligned matrices\footnote{We would have to store all the permutations, for each diagram.}. Fortunately, if we use re-alignment of diagrams, i.e. write $\tilde{D}_{ija}^{cbk}$ as $D_{ijk}^{abc}$, we can simply store the permutations of columns and rows, since there will never be a permutation between hole- and particle-indices. Therefore, to do permutations on the amplitude contribution $\tilde{D}$, we use the dictionary to set all the elements of $\tilde{D}$ into a temporary, empty array\footnote{By empty, we mean all components are zero.} $A'$ in same layout as for the amplitude array $A$, meaning the element number $i$ in $A'$ has the same state indices as element $i$ in $A$. We then construct the direct channels $D$ using this temporary array. We have then effectively done a remapping

	\begin{equation}
		\tilde{D}_{ija}^{cbk} \:\rightarrow\: D_{ijk}^{abc}
	\end{equation}

	By now swapping the rows and columns of $D$, using some object telling us, for example, which rows go where if we interchange $i\leftrightarrow k$. To be specific, we can consider how to do this permutation, meaning we want to find $\hat{P}(ik)t_{ijk}^{abc} = t_{ijk}^{abc} - t_{kji}^{abc}$. Now, due to our whole channel regime, we cannot just change the indices in some 4- or 6-dimensional tensor. Instead, we need to interchange columns and rows in matrices, such if that two rows, $(i,j,k)$ and $(l,m,n)$, that become permuted as

	\begin{align}
	(i,j,k) \quad&\overset{\hat{P}}{\longrightarrow}\quad (i',j',k') \\
	(l,m,n) \quad&\overset{\hat{P}}{\longrightarrow}\quad (l',m',n')
	\end{align}

	and fulfil

	\begin{align}
	(i,j,k) &= (l',m',n') \\
	(l,m,n) &= (i',j',k') \:,
	\end{align}

	then we switch the two rows $(i,j,k)$ and $(l,m,n)$. In matrix form, such a permutation can, for example, be seen as

	\begin{equation*}
	\begin{matrix}[cccc|cccccc]
	&   &   & a & 10 & 9  & 15 & 11 & 17 & 14\\
	&   &   & b & 11 & 14 & 14 & 18 & 23 & 15\\
	&   &   & c & 16 & 17 & 13 & 20 & 20 & 23\\
	i & j & k &   &    &    &    &    &    &   \\ \hline
	0 & 1 & 2 & & a_{11} & a_{12} & a_{13} & a_{14} & a_{15} & a_{16}\\
	0 & 2 & 1 & & a_{21} & a_{22} & a_{23} & a_{24} & a_{25} & a_{26}\\
	1 & 0 & 2 & & a_{31} & a_{32} & a_{33} & a_{34} & a_{35} & a_{36}\\
	1 & 2 & 0 & & a_{41} & a_{42} & a_{43} & a_{44} & a_{45} & a_{46}\\
	2 & 0 & 1 & & a_{51} & a_{52} & a_{53} & a_{54} & a_{55} & a_{56}\\
	2 & 1 & 0 & & a_{61} & a_{62} & a_{63} & a_{64} & a_{65} & a_{66}\\
	\end{matrix}
	\quad\overset{\hat{P}(ik)}{\longrightarrow}\quad
	\begin{matrix}[cccc|cccccc]
	&   &   & a & 10 & 9  & 15 & 11 & 17 & 14\\
	&   &   & b & 11 & 14 & 14 & 18 & 23 & 15\\
	&   &   & c & 16 & 17 & 13 & 20 & 20 & 23\\
	k & j & i &   &    &    &    &    &    &   \\ \hline
	2 & 1 & 0 & & a_{61} & a_{62} & a_{63} & a_{64} & a_{65} & a_{66}\\
	1 & 2 & 0 & & a_{41} & a_{42} & a_{43} & a_{44} & a_{45} & a_{46}\\
	2 & 0 & 1 & & a_{51} & a_{52} & a_{53} & a_{54} & a_{55} & a_{56}\\
	0 & 2 & 1 & & a_{21} & a_{22} & a_{23} & a_{24} & a_{25} & a_{26}\\
	1 & 0 & 2 & & a_{31} & a_{32} & a_{33} & a_{34} & a_{35} & a_{36}\\
	0 & 1 & 2 & & a_{11} & a_{12} & a_{13} & a_{14} & a_{15} & a_{16}
	\end{matrix}
	\end{equation*}

	By adding these matrices, we will have calculated $\hat{P}(ik)t_{ijk}^{abc}$.\\

	The second approach consisted of assuming that every element in the diagram sums were some permutation for another element. To explicate, assume we have some diagram element $D_{ijk}^{abc}$. We know that, without permutation, this element contributes directly to the same amplitude, i.e.

	\begin{equation}
		t_{ijk}^{abc} \leftarrow D_{ijk}^{abc}
	\end{equation}

	However, we know\footnote{To realise this, go back to the way in which we do the $T_2$ permutations, and consider to what the map actually returns when we swap two indices.} that this element is also a $i\leftrightarrow j$ permutation of the element $D_{i'j'k'}^{a'b'c'}$, where

	\begin{align}
	\begin{split}
	i &= j' \\
	j &= i' \\
	k &= k'
	\end{split}
	\end{align}

	In other words, we may use

	\begin{align}
		\begin{split}
		t_{i'j'k'}^{a'b'c'} &= \hat{P}(ik)D_{i'j'k'}^{a'b'c'} \\
		&= D_{i'j'k'}^{a'b'c'} - D_{ijk}^{abc}
		\end{split}
	\end{align}

	At first, this might seem utterly pointless, since we would have to search through all the $D$ elements to find $D_{ijk}^{abc}$, when working with $D_{i'j'k'}^{a'b'c'}$. However, knowing that this specific diagram has a permutation of $i$ and $j$, we can, when going through all the elements of $D$, simply let each element $D_{ijk}^{abc}$ contribute to both\footnote{It is important to be mindful of signs when doing this. Each swap of indices gives a minus sign.} $t_{ijk}^{abc}$ and $t_{jik}^{abc}$. Mathematically, this approach is show by

	\begin{align}
		\begin{split}
			t_{ijk}^{abc} &\leftarrow \hat{P}(ik) D_{ijk}^{abc} \\
			\underset{\hat{P}(ik)\big|}{\longrightarrow}\quad \hat{P}(ik)t_{ijk}^{abc} &\leftarrow \hat{P}^2(ik)D_{ijk}^{abc} = D_{ijk}^{abc}
		\end{split}
	\end{align}

	The latter approach is nice in the sense that we need not do any sort of remapping, but it is at the cost of having to do another map search per permutation operator, per diagram element. However, it is problematic when it comes to parallellisation. Since we, ultimately, would like to work on channels in parallel, the second approach means we might try to write to the same element from several processes. We will discuss parallellisation in more depth below.

	\section{Parallelisation}
	Parallelisation is a necessity for CCDT, even for the most efficient implementations, simply due to the immense increase in computational time and memory requirements when we move up in system size. There are two options for C++ when it comes to parallelisation, one being use of MP (multi-processing) and the other being MPI (multi-processing interface). We have attempted both options as a means of speedup, but found complications in both cases, which severely limited the use of parallel computation.\\

### 2018-04-13 doc/TimeDependence/Tex/Article.tex
Added time dependence paper

\documentclass[a4paper,10pt, twocolumn]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{braket}
\usepackage{verbatim}
\usepackage{mdframed} %nice frames
\usepackage[hidelinks]{hyperref}
\usepackage{cite}
\usepackage[toc,page]{appendix}
\newtheorem{theorem}{Theorem}
\renewcommand{\thesection}{\Roman{section}}
\renewcommand\thesubsection{\Alph{subsection}}
%opening
\title{Article}
\author{Kristiansen, Håkon Emil}

\begin{document}

\maketitle

\begin{abstract}
We present a working implementation of a simplified version of the orbital-adaptive time-dependent coupled-cluster method which we refer to as 
just the time-dependent coupled-cluster (TDCC) method.  OATDCC/TDCC approximates solutions to the quantum $N$-body problem 
and for $N=2$ TDCC is equivalent to the time-dependent configuration-interaction method (TDCI). 
A TDCI solver is implemented and shown to be correct by comparing with a previous study 
and is then used as a foundation for verifying the implementation of the TDCC method. For one and two dimensional quantum dot 
systems with $N=2$ we demonstrate that the methods are equal to numerical precision. Furthermore, for a small number of basis functions we use the TDCC solver 
to compute the time evolution of six-electron isotropic quantum dot.  
\end{abstract}

%\tableofcontents

\section{Introduction}
Quantum mechanics is a theory that describes/models the properties of microscopic systems. It postulates that given the so-called wavefunction, $\Psi(\mathbf{r},t)$, we can 
in principle compute all that there is to know about the system. 
Given suitable initial conditions $\Psi(\mathbf{r},t)$ can be determined for all future time by solving the time-dependent Schrödinger equation (TDSE),
\begin{equation}
 i \hbar \frac{\partial }{\partial t} \Psi(\mathbf{r},t) = \hat{H} \Psi(\mathbf{r},t).
\end{equation}
The initial condition is typically taken as a linear combination of what is called stationary states, which in turn can be found by solving what is known as the 
time-independent Schrödinger (TISE) equation
\begin{equation}
 \hat{H} \Psi(\mathbf{r}) = E \Psi(\mathbf{r}).
\end{equation}
Exact/analytical solutions to the TISE is possible only for the simplest systems and general solutions to the TDSE is even rarer. Thus, 
one must resort to approximative/numerical methods. Of particular interest is the so-called many-body problem where one considers systems of a large number of interacting particles, 
where large can be anywhere from two to infinity.

Several approaches for solving the many-body TISE numerically/approximatively has been devised such as Hartree-Fock (HF), Density Functional Theory (DFT), 
Configuration Interaction (CI) and Coupled Cluster (CC). While efficient, Hartree-Fock and DFT 
are insufficient if one wants a high degree of accuracy. Configuration Interaction and Coupled Cluster methods are hierarchical in the sense that one can 
systematically construct increasly accurate approximations. If the CI method is not truncated we have what is known as Full Configuration Interaction (FCI). FCI 
can be seen as exact (within some finite space), however it suffers from exponential scaling. Truncated CI methods which would achieve polynomial scaling 
are problematic since they are not size-consistent and extensive. Truncated CC methods one the other hand are size-consistent and extensive and achieves polynomial 
scaling. Due to this fact Coupled Cluster is considered the gold standard of many-body techniques if high accuracy is desired.

Similarly, there exists different approximations to the solution of the TDSE with 
the Multiconfiguration Time-Dependent Hartree-Fock method (MCTDHF) being considered the most accurate. MCTDHF, which is combination of CI and HF generalized to 
the time domain, suffers from exponential scaling. In a recent article \cite{Kvaal12}, Simen Kvaal proposes a method based on Coupled-Cluster theory, 
the so-called Orbital Adaptive Time Dependent Coupled Cluster method (OATDCC), which is a hierarchical approximation to the MCTDHF method. The OATDCC inherits 
size-consisteny and extensivity from the CC method and achieves polynomial scaling. 
\section{Theory/Formalism}
\subsection{The model Hamiltonian}
We shall model the quantum dot systems as a collection of $N$ nonrelativistic electrons of mass $m$ in $d$-dimensional space trapped by a confining potential 
$v_{\text{conf}}(\mathbf{\hat{r}})$. Here $\hat{\mathbf{r}}$ is the position operator in $d$-dimensional space 
\begin{equation}
 \hat{\mathbf{r}} \equiv \sum_{i=1}^d \hat{\mathbf{x}}_i \hat{\mathbf{e}}_i
\end{equation}
with $\hat{\mathbf{e}}_i$ a unit vector.

The electrons interact through the Coulomb interaction $e^2/(4\pi \epsilon R)$, where $e$ is the electron charge, $\epsilon$ the permitivity of the medium and $R$ is the 
distance between two interacting electrons.

For convenience, we will use atomic untis, chosen such that $\hbar \equiv m \equiv e \equiv 4\pi \epsilon \equiv 1$. Thus, all energies are presented in hartrees per $\hbar$.

In atomic units, the many-body problem is described by the Hamiltonian on second quantized form
\begin{align}
 \hat{H} = \hat{H}_1 + \hat{H}_2 = &\sum_{pq} \braket{p|\hat{h}_0|q}\hat{a}_p^\dagger\hat{a}_q \nonumber \\
 + \frac{1}{4}&\sum_{pqrs}\braket{pq|\hat{u}|rs}_{\text{AS}}\hat{a}_p^\dagger\hat{a}_q^\dagger\hat{a}_s\hat{a}_r \label{FieldFreeHamiltonian}
\end{align}
where $\hat{H}_1$ and $\hat{H}_2$ are its one- and two-body parts.

The operator $\hat{h}_0$ is the single-particle Hamiltonian
\begin{equation}
 \hat{h}_0 = -\frac{1}{2}\nabla^2 + v_{\text{conf}}(\mathbf{\hat{r}})
\end{equation}
while $\hat{u}$ represents the Coulomb interaction. Here $\ket{p}$ is a single-particle basis function which in its position representation is given by
\begin{equation}
 \braket{\mathbf{r}|p} \equiv \phi_p(\mathbf{r}).
\end{equation}
In practice, we have to choose some single-particle basis to compute within. One typical choice is to take the single-particle basis to be the eigenfunctions of $\hat{h}_0$ such that 
\begin{equation}
 \hat{h}_0 \ket{p} = \epsilon_p \ket{p},
\end{equation}
where $\epsilon_p$ is the corresponding eigenvalue.

The quantities $\braket{p|\hat{h}_0|q}$ and $\braket{pq|\hat{u}|rs}_{\text{AS}}$ are referred to as \textit{matrix elements} and must 
be computed relative to the chosen single-particle basis. In particular, $\braket{pq|\hat{u}|rs}_{\text{AS}}$ are \textit{antisymmetrized} matrix elements, defined  by 
\begin{equation}
 \braket{pq|\hat{u}|rs}_{\text{AS}} \equiv u^{pq}_{rs} - u^{pq}_{sr}
\end{equation}
where $u^{pq}_{rs}$ is the interaction integral, 
\begin{equation}
 u^{pq}_{rs} \equiv \int \phi^*_p(\mathbf{r}_1) \phi^*_q(\mathbf{r}_2) \hat{u}(\mathbf{r}_1,\mathbf{r}_2) \phi_r(\mathbf{r}_1)\phi_s(\mathbf{r}_2) d\mathbf{r}_1 d\mathbf{r}_2.
\end{equation}
In the rest of this paper we will drop the subscript AS for the antisymmetrized matrix elements.

In order to generate dynamics the electrons are coupled to a time dependent monochromatic laser field, with frequency $\omega$ and amplitude $\mathcal{E}_0$, polarized along the $x$-axis \cite{Zanghellini04}
\begin{equation}
 v_{\text{laser}}(\mathbf{\hat{r},t}) = \hat{\mathbf{x}}_1 \mathcal{E}_0 \sin(\omega t),
\end{equation}
resulting in a time dependent single-particle Hamiltonian
\begin{equation}
 \hat{h}_t = \hat{h}_0 + v_{\text{laser}}(\mathbf{\hat{r},t}).
\end{equation}
The choice of polarization axis is arbitrary.

The total time dependent Hamiltonian of the system is then given by
\begin{align}
 \hat{H}(t) = &\sum_{pq} \braket{p|\hat{h}_t|q}\hat{a}_p^\dagger\hat{a}_q \nonumber \\
 + \frac{1}{4}&\sum_{pqrs}\braket{pq|\hat{u}|rs}_{\text{AS}}\hat{a}_p^\dagger\hat{a}_q^\dagger\hat{a}_s\hat{a}_r.
\end{align}
It is important to note that this implies that we have to update the matrix elements $\braket{p|\hat{h}_t|q}$ during the simulation. However, in this work we consider only methods where 
the single-particle basis is held fixed in time, hence the two-body matrix elements can be computed $\braket{pq|\hat{u}|rs}$ once and for all.

We will handle the time dependent problem in two steps. First we prepare a system in its groundstate by solving the time independent Schrödinger equation (TISE) for the field-free Hamiltonian \eqref{FieldFreeHamiltonian}. 
Then we apply the time dependent laser field and compute the time evolution from the time dependent Schrödinger equation (TDSE). 
Before we discuss the metods for solving the TISE and TDSE respectively we will adress the choice of single-particle basis.

\subsection{Choice of single-particle basis}
The choice of single-particle basis will in general depend on the problem. In quantum chemistry it is customary to expand the eigenfunctions, $\{ \psi_k \}$, 
of $\hat{h}_0$ in (non-orthogonal) Gaussian basis functions
$g_p(\mathbf{r})$
\begin{equation}
 \hat{h}_0 \ket{\psi_k} = \hat{h}_0 \sum_{p=1}^L g_p(\mathbf{r}) = \epsilon_k \sum_{p=1}^L g_p(\mathbf{r})
\end{equation}
resulting in a generalized eigenvalue problem which must be solved for the eigenfunctions. Ideally we would let $L \rightarrow \infty$ but must be taken as finite in practice. 
The choice of Gaussian basis functions is popular due to the fact that matrix elemtents can be evaluated efficiently \cite{Helgaker00book}.

Another popular choice, which is the one chosen in this work, is to use the eigenfunctions of the Harmonic oscillator (HO) 
\begin{equation}
 \left(-\frac{1}{2}\nabla^2 + \frac{1}{2} \omega_{\text{HO}}\hat{\mathbf{r}}^2 \right) \phi_k(\mathbf{r}) = \epsilon_k \phi_k(\mathbf{r})  
\end{equation}
as a single-particle basis. The solutions to the HO-problem is well known and the eigenfunctions have closed form expressions.

Of particular importance for our purposes is the fact that the interaction integrals
\begin{equation}
 u^{pq}_{rs} \equiv \int \phi^*_p(\mathbf{r}_1) \phi^*_q(\mathbf{r}_2) \hat{u}(\mathbf{r}_1,\mathbf{r}_2) \phi_r(\mathbf{r}_1)\phi_s(\mathbf{r}_2) d\mathbf{r}_1 d\mathbf{r}_2
\end{equation}
can be solved analytically in two- and three-dimensions, using HO-basis functions in polar and spherical coordinates respectively \cite{Anisimovas98, Vorrath2003}. If the confining potential 
is chosen to be the Harmonic oscillator potential we are essentially done and can move on to the many-body treatment of the problem.

Additionally we may approximate the eigenfunctions of more general confining potentials by expanding the solutions in the HO-functions.

Consider the general problem
\begin{equation}
 \hat{h}_0 \ket{\psi_k} = \epsilon_k \ket{\psi_k} \label{EigenfuncEq}
\end{equation}
where 
\begin{equation}
 \hat{h}_0 = -\frac{1}{2}\nabla^2 + v_{\text{conf}}(\mathbf{r}).
\end{equation}
Expand the eigenfunctions $\ket{\psi_k}$ in the HO-functions $\ket{\phi_p}$
\begin{equation}
 \ket{\psi_k} = \sum_{p=1}^L c_{p,k} \ket{\phi_p} 
\end{equation}
and insert into Eq.\eqref{EigenfuncEq}. Left-projection with $\bra{\phi_q}$ results in the following expression
\begin{equation}
 \sum_{p=1}^L c_{p,k} \braket{\phi_q|\hat{h}_0|\phi_p} = \epsilon_k \sum_{p=1}^L c_{p,k} \delta_{q,p}
\end{equation}
where we have used the orthogonality of the HO-functions. This can be written on matrix form as 
\begin{equation}
 HC = C\mathcal{E}
\end{equation}
where $H_{qp} \equiv \braket{\phi_q|\hat{h}_0|\phi_p}$, $\mathcal{E}$ is a diagonal matrix containing the eigenvalues on its diagonal, and $C$ is matrix conatining 
the expansions coefficients for basis function $k$ in its columns. The eigenproblem must be solved for the expansions coefficients, resulting in a approximation of the interaction integrals 
in terms of the known intercation integrals $\braket{\phi_\alpha \phi_\beta|\hat{u}|\phi_\gamma \phi_\delta}$ given by 
\begin{equation}
 \braket{\psi_p \psi_q| \hat{u}| \psi_r \psi_s} = \sum_{\alpha \beta \gamma \delta} c^*_{p,\alpha} c^*_{q,\beta} c^*_{r,\gamma} c^*_{s,\delta} \braket{\phi_\alpha \phi_\beta|\hat{u}|\phi_\gamma \phi_\delta}.
\end{equation}

The quality of such an approximation will naturally depend on how close the confining potential is to the Harmonic oscillator potential. Nevertheless it constitutes a convenient starting point for 
investigations of more general potentials.

\subsection{Configuration interaction theory}

\subsubsection{Time independent Configuration interaction}
The conceptually simplest method for solving the many-problem is probably that of Configuration interaction (CI).

Let us first consider the time independent Schrödinger equation 
\begin{equation}
 \hat{H} \ket{\Psi_k} = E_k \ket{\Psi_k}
\end{equation}
where $\hat{H}$ is the field-free Hamiltonian given by Eq. \eqref{FieldFreeHamiltonian}, $\ket{\Psi_k}$ is the many-body eigenfunction and $E_k$ its corresponding eigenvalue. It is customary 
to use a (finite in practice) basis of orthogonal Slater determinants (SD), $\{ \Phi_p \}_{p=1}^{N_{\text{sd}}}$ where $N_{\text{sd}}$ is the number of determinants, built from the single-particle basis functions described in the previous section.

The principal idea of CI is to expand the wavefunction in the SD-basis
\begin{equation}
 \hat{H} \sum_{p=1}^{N_{\text{sd}}} c_{p,k} \ket{\Phi_p} = E_k \sum_{p=1}^{N_{\text{sd}}} c_{p,k} \ket{\Phi_p}.
\end{equation}
If we left-project with $\bra{\Phi_q}$ the above equation can be written on matrix form as 
\begin{equation}
 HC = CE
\end{equation}
where $H_{qp} \equiv \braket{\Phi_q|\hat{H}|\Phi_p}$, $E$ is a diagonal matrix with the eigenvalues along its diagonal and $C$ is a matrix containing the expansion coefficients of the 
$k$-th eigenfunction in column $k$. Thus if we can compute the matrix elements $H_{qp}$ the TISE can be solved by numerical diagonalization.

The evaluation of $H_{qp}$ is simplified by what is known as the Slater-Condon rules. If $\ket{\Phi}$ is any SD, 
the only non-zero contributions to $H$ are given by
\begin{align}
 \braket{\Phi| \left(\hat{H}_1 + \hat{H}_2\right)|\Phi} &= \sum_i \braket{i|\hat{h}_0|i} + \frac{1}{2}\sum_{ij}\braket{ij|\hat{u}|ij} \label{SlaterCondonEqual}\\
 \braket{\Phi| \left(\hat{H}_1 + \hat{H}_2\right)|\Phi_k^c} &= \braket{k|\hat{h}_0|c} + \sum_{i} \braket{ki|\hat{u}|ci} \label{SlaterCondonDiffByOne} \\
 \braket{\Phi| \left(\hat{H}_1 + \hat{H}_2\right)|\Phi_{kl}^{cd}} &= \braket{kl|\hat{u}|cd} \label{SlaterCondonDiffByTwo}.
\end{align}
In other words, we only have to consider the overlap between determinants that differ by at most two occupied single-particle functions. 
Here we use the convention that $i,j,k,l = 1,...,N$, with $N$ being the number of electrons and $a,b,c,d = N+1,...$ and $p,q,r,s = 1,...,L$. Thus, 
we see that if the matrix elements $\braket{p|\hat{h}_0|q}$ and $\braket{pq|\hat{u}|rs}$ are known we can set up the Hamilton matrix and diagonalize it, obtaining the expansion coefficients $C$.

Ideally we would like $N_{\text{sd}} \rightarrow \infty$ which is impossible in practice. If we choose a finite number, $L$, of single-particle basis functions and use all the linearly independent 
Slater determinants we can build from this basis we get what is known as the Full Configuration Interaction (FCI). However, the problem with FCI is that the number of linearly independent 
determinants are given by 
\begin{equation}
 N_{\text{sd}} = \binom{L}{N}
\end{equation}
with $N$ being the number of electrons. This means that the numbers of SD's becomes huge very fast, making FCI practical/feasible only for the smallest systems. On the other hand, FCI is exact within 
a given computational space, thus it is an invaluable tool in aiding us determining the validity of more sophisticated methods. Furthermore, CI is variational in the sense that it guarantees an upper bound 
on the groundstate energy.

One way to adress the problems with FCI is to consider a \textit{reference} determinant $\ket{\Phi}$, typically taken to be the Hartree-Fock state, and consider excitations upon this state. That is, 
we make the ansatz
\begin{equation}
 \ket{\Psi} = A_0 \ket{\Phi} + \sum_{ia} A_{ia} \ket{\Phi_{i}^a} + \frac{1}{4} \sum_{ijab} A_{ijab} \ket{\Phi_{ij}^{ab}} + \cdots
\end{equation}
where $\ket{\Phi_i^a}$ is a singly-excited determinant, $\ket{\Phi_{ij}^{ab}}$ doubly-excited etc. and truncate at some excitation level. 
This results in a hierarchy of approximations known as CIS (only single-excitations), CID (only double-excitations), CISD (singles and doubles and so on. In this way one obtains polynomial 
scaling of the number of determinants. The problem with this approach is that the truncated CI-wavefunction is not size-consistent and extensive \cite{Helgaker00book}. It turns out 
that this problem is resolved by the Coupled Cluster method, which makes it the de facto standard of many-body methods today.

\subsubsection{Time dependent Configuration interaction}
Another advantage with CI is that it is straightforward to generalize to time dependent problems. Recall that the time evolution of an  
arbitrary state, $\ket{\Psi(t)}$ with initial condition $\ket{\Psi(t_0)}$, is given by the time dependent Schrödinger equation, 
\begin{equation}
 i \hbar \frac{\partial}{\partial t}\ket{\Psi(t)} = \hat{H}(t) \ket{\Psi(t)}.
\end{equation}
We will take the initial condition to be some (normalized) linear combination of the CI eigenfunctions. In particular, we are interested in the 
time evolution of the groundstate.

The time-dependent Configuration Interaction (TDCI) method is obtained by placing a time-dependence on 
the expansion coefficients in the CI-wavefunction while keeping the single-particle basis fixed. Analogous to what we did in the previous section, we 
expand $\ket{\Psi(t)}$ in a given orthonormal Slater determinant basis $\{ \Phi_p\}_{p=1}^{N_{\text{sd}}}$
\begin{equation}
 \ket{\Psi(t)} = \sum_{p=1}^{N_{\text{sd}}} A_p(t) \ket{\Phi_p}.
\end{equation}
If we now insert this expansion into the time-dependent Schrödinger equation we obtain,
\begin{equation}
 i \hbar  \sum_{p=1}^{N_{\text{sd}}} \frac{\partial A_p(t)}{\partial t} \ket{\Phi_p} =  \sum_{p=1}^{N_{\text{sd}}} A_p(t) \hat{H}(t) \ket{\Phi_p}. 
\end{equation}
Left-projecting the above equation with $\bra{\Phi_q}$ and using the orthonormality of the Slater determinants, 
we are left with a differential equation for each expansion coefficient $A_J(t)$
which can be written as the matrix-vector equation
\begin{equation}
 i\hbar \frac{\partial A(t)}{\partial t} = H(t)A(t). \label{TDFCIEquation}
\end{equation}
Here $H(t)$ is the time dependent Hamilton matrix defined in terms of its elements,
\begin{equation}
 H_{qp}(t) = \braket{\Phi_q|\hat{H}(t)|\Phi_p}
\end{equation}
and $A(t) = [A_0(t), A_1(t), \cdots]$ is the vector containing the expansion coefficients. We refer to this as the 
time dependent Configuration Interaction (TDCI) method.

Completely analogous to the time independent case we have a hierarchy of TDCI methods defined by the Slater determinant basis. Thus we have 
a series of possible approximations ranging from TD-CIS to TD-FCI.
If we did not have to truncate the Slater determinant basis, TDFCI would be an exact solution to the TDSE.
However, we must in practice truncate the basis and TDFCI suffers from the fact that the size of FCI-space grows exponentially. 
In order to obtain a good approximation to the exact solution we would need a huge fixed basis\cite{Kvaal12}.

On the other hand it is quite straightforward to implement, especially when a program for
computing the CI-groundstate already has been written. 
Again, the fact that the CI and CC space are equivalent for $N=2$ makes it worthwhile to write a TDCI program, 
which we can use to compare with the time dependent Coupled Cluster method which we discuss in the next chapter. 
\begin{comment}
\subsection{Coupled cluster theory}
In this section we review Coupled Cluster (CC) theory. 
We will look at the "classical" way of obtaining the CC equations. The presentation borrows heavily from the 
excellent review by Crawford and Schaefer\cite{Crawford2000}.

Furthermore we look at a different approach which in addition to the usual CC equations also 
gives a method to solve the time dependent Schrödinger equation, the so-called Orbital Adaptive Coupled Cluster (OATDCC) method\cite{Kvaal12}.

\subsubsection{The exponential ansatz}
As usual we start by considering the time-independent Schrödinger equation
\begin{equation*}
\hat{H} \ket{\Psi} = E\ket{\Psi}.
\end{equation*}
In Coupled Cluster theory an exponential ansatz 
\begin{equation}
\ket{\Psi}_{CC} \equiv e^{\hat{T}} \ket{\Phi} \label{ExponantialAnsatzCC},
\end{equation}
is used to approximate the exact solution. Here $\ket{\Phi}$ is a reference Slater determinant 
and $\hat{T} = \sum_{i}^n \hat{T}_i$ is the sum over n-orbital cluster operators.

The the one- and two-orbital cluster operators are defined as
\begin{equation}
\hat{T}_1 \equiv \sum_{ia} t_i^a c_a^\dagger c_i
\end{equation}
and 
\begin{equation}
\hat{T}_2 \equiv \frac{1}{4} \sum_{ijab} t_{ij}^{ab} c_b^\dagger c_j c_a^\dagger c_i.
\end{equation}
Generally an n-orbital cluster operator is defined as 
\begin{equation}
\hat{T}_n \equiv \left(\frac{1}{n!}\right)^2 \sum_{ijk...abc...}t_{ijk...}^{abc...} ...c_c^\dagger c_k c_b^\dagger c_j c_a^\dagger   c_i.
\end{equation}
%We see that the cluster operators are linear combinations of excitation operators as defined by \eqref{DefinitionOfExciationOperator}. 
Recalling the exponential series we can write,
\begin{equation}
 \ket{\Psi}_{CC} = e^{\hat{T}} \ket{\Phi} = \left( 1+\sum_{k=1}^\infty \frac{1}{k!}\hat{T}^k \right) \ket{\Phi}. \label{CCansatzSeriesExpansion}
\end{equation}

\subsection{The Coupled Cluster Equations}
We now proceed by inserting expression \eqref{ExponantialAnsatzCC} into the time-independent Schrödinger equation
\begin{equation}
\hat{H}e^{\hat{T}} \ket{\Phi} = Ee^{\hat{T}} \ket{\Phi} \label{ExponantialAnsatzCCInsertedInSE}.
\end{equation}
Left-projecting with $\bra{\Phi}$ we get an expression for the energy,
\begin{align*}
 \braket{\Phi|\hat{H}e^{\hat{T}}|\Phi} &= E \braket{\Phi|e^{\hat{T}}|\Phi} = E \braket{\Phi|\Psi_{CC}} \\
				       &= E
\end{align*}
where intermediate normalization, $\braket{\Phi|\Psi_{CC}} = 1$, is assumed. Furthermore one can obtain an expression for the amplitude $t^{ab...}_{ij...}$ 
upon left-projection with $\bra{\Phi^{ab...}_{ij...}}$,
\begin{equation}
 \braket{\Phi^{ab...}_{ij...}|\hat{H}e^{\hat{T}}|\Phi} = E\braket{\Phi^{ab...}_{ij...}|e^{\hat{T}}|\Phi}.
\end{equation}
Notice that this equations are non-linear (due to the presence of $e^{\hat{T}}$) and depend on the energy equation. This is impractical and is resolved by performing 
a similarity transformation of the Hamiltonian.

Multiply Eq. \eqref{ExponantialAnsatzCCInsertedInSE} with $e^{-\hat{T}}$ and left-project with $\bra{\Phi}$ in order to obtain
\begin{equation}
 \braket{\Phi|e^{-\hat{T}} \hat{H} e^{\hat{T}} | \Phi} = E \braket{\Phi|e^{-\hat{T}}e^{\hat{T}}|\Phi} = E. \label{SimTransEnergyEquation}
\end{equation}
Left-projection with $\bra{\Phi^{ab...}_{ij...}}$ now results in,
\begin{equation}
 \braket{\Phi^{ab...}_{ij...}|e^{-\hat{T}}\hat{H}e^{\hat{T}}|\Phi} = E \braket{\Phi^{ab...}_{ij...}|\Phi} = 0. \label{SimTransAmplitudeEquation}
\end{equation}
Equations \eqref{SimTransEnergyEquation} and \eqref{SimTransAmplitudeEquation} are commonly referred to as the Coupled Cluster equations. Notice now that the equation 
for the amplitudes does not depend on the energy. The amplitude equation is non-linear and has to be solved iteratively. We say that we have a self-consistent solution to CC-equations
if the energy converges to some pre-defined precision.

In order to implement the Coupled Cluster method we need algebraic expressions for the CC-equations. 
\begin{theorem}[Baker-Campbell-Hausdorff expansion]
For any matrices $A$ and $S$ the Baker-Campbell-Hausdorff (BCH) expansion is given by,
\begin{equation}
e^{-S}Ae^{S} = A + [A,S] + \frac{1}{2}[[A,S],S] + \frac{1}{3!}[[[A,S],S],S] + ... \label{BCH},
\end{equation}
where 
$$[A,S] \equiv AS - SA$$ 
is the commutator between $A$ and $S$. 
\end{theorem}
Under the assumption that $\hat{H}$ contains at most a two-body operator the BCH-expansion truncates after the first 
five terms on the right-hand side of eq. \eqref{BCH}\cite{Crawfordtruncates}. 
This fact greatly simplifies the evaluation of the left-hand sides of the equations \eqref{SimTransEnergyEquation} and \eqref{SimTransAmplitudeEquation}.

These expressions can be evaluated by hand using the second quantization formalism as demonstrated in\cite{Crawford2000} or by diagrammatic
techniques\cite{ShavittBartlett}. Alternatively one can use the second quantization toolbox in the Python library Sympy\footnote{\href{http://docs.sympy.org/latest/modules/physics/secondquant.html}{http://docs.sympy.org/latest/modules/physics/secondquant.html}} 
to compute these expressions.

\subsection{A variational CC theory?}
One drawback with the CC-method is that it does not satisfy the variational principle \eqref{variationalTheorem}. Consider the operator
$e^{-\hat{T}}\hat{H}e^{T}$ which appears in the similarity transformed energy equation \eqref{SimTransEnergyEquation}. This operator is not Hermitian. In order to see this compute $\hat{T}_1^\dagger$,
\begin{equation}
 \hat{T}_1^\dagger = \left(\sum_{ia} t_i^a c_a^\dagger c_i \right) = \sum_{ia} (t_i^a)^* c_i^\dagger c_a \neq \hat{T}_1.
\end{equation}
Then it follows that
\begin{equation}
 \left(e^{-\hat{T}}\hat{H}e^{\hat{T}}\right)^\dagger = (e^{\hat{T}})^\dagger \hat{H} (e^{-\hat{T}})^\dagger = e^{\hat{T}^\dagger} \hat{H} e^{-\hat{T}^\dagger} \neq e^{-\hat{T}}\hat{H}e^{\hat{T}}.
\end{equation}
Hermiticity of the operator was a necessary condition for the variational theorem, thus Eq. \eqref{SimTransEnergyEquation} is not variational. 
However, if $\hat{T}$ is not truncated the spectrum of $e^{-\hat{T}}\hat{H}e^{T}$
is identical to the original Hermitian operator $\hat{H}$, which justifies its use in quantum mechanical models\cite{CrawfordNonVar}.

There has been attempts to construct a variational solution by deriving the amplitude equations by minimizing the functional,
\begin{equation}
 \frac{\braket{\Psi|\hat{H}|\Psi}}{\braket{\Psi|\Psi}} = \frac{\braket{\Phi|(e^{\hat{T}})^\dagger \hat{H}e^{T}|\Phi}}{\braket{\Phi|(e^{\hat{T}})^\dagger e^{\hat{T}}|\Phi}}.
\end{equation}
Notice then that $(e^{\hat{T}})^\dagger \hat{H}e^{T}$ does not conform to the Baker-Campbell-Hausdorff formula and the series has no natural truncation, complicating 
matters greatly. 
\subsection{The Orbital Adaptive Time Dependent Coupled Cluster Method}
In the article\cite{Kvaal12} Simen Kvaal demonstrates that it is possible to 
derive equations of motion for the CC-ansatz by computing critical points of a bivariational expectation value functional. 
In his approach both the orbitals and CC-amplitudes are allowed to vary and the method can be used to solve the time-dependent Schrödinger equation.  
The method is called the Orbital Adaptive Time-Dependent Coupled Cluster (OATDCC) and is a hierarchical approximation to MCTDHF. Furthermore, if the orbitals are 
held fixed in time the method gives a approximation to the Time-Dependent CI method which we refer to as the Time-Dependent Coupled Cluster (TDCC) method. 
While TDFCI and MCTDHF suffers from exponential scaling, OATDCC and TDCC achieves polynomial scaling.

The main goal of this article is to implement the TDCC method for a possibly time-dependent Hamiltonian. 
As such we will in the following summarize the key ingredients needed to achieve this. 
For mathematical details the interested reader is referred to the article by Kvaal.

\subsubsection{The Bivariational Principle}
As we have established, the Coupled Cluster method is not variational in the usual sense since the similarity transformed Hamiltonian,
$$ \bar{H} = e^{-\hat{T}}\hat{H}e^{\hat{T}} $$
is not Hermitian. Now following Kvaal, let $A$ be an operator (possibly non-Hermitian) over Hilbert space $\mathcal{H}$ and consider the expectation value functional
\begin{equation}
 \mathcal{E}_A: \mathcal{H}^{\prime}\times \mathcal{H} \rightarrow \mathbb{C}, \ \ \mathcal{E}_A [\bra{\Psi^{\prime}}, \ket{\Psi}] = \frac{\braket{\Psi^{\prime} |A| \Psi} }{\braket{\Psi^{\prime}|\Psi}}. 
\end{equation}
This is referred to as the the bivariational expectation value functional. In contrast to the usual 
variational principle $\bra{\Psi^{\prime}}$ and $\ket{\Psi}$ are treated as independent parameters. The conditions for 
$\delta \mathcal{E}_A = 0$, for all independent variations of $\bra{\Psi^{\prime}}$ and $\ket{\Psi}$ is 
\begin{equation}
 (A-a)\ket{\Psi} = 0 \ \ \text{and} \ \ \bra{\Psi^{\prime}}(A-a) = 0,
\end{equation}
with 
$$a = \mathcal{E}_A[\bra{\Psi^{\prime}},\ket{\Psi}] $$
being the value of $\mathcal{E}_A$ at the critical point.

Equations of motion can be derived from so-called time-dependent variational principles. One such variational is set up 
using the Lagrange formulation\cite{BECK20001}, 
\begin{equation}
 \delta \int_0^T L[\ket{\Psi}]dt = 0,
\end{equation}
where the Lagrangian functional is given by 
\begin{equation}
L[\ket{\Psi}] = \braket{\Psi(t)|\hat{H}-i\hbar \frac{\partial}{\partial t}|\Psi(t)}
\end{equation}
with the boundary conditions $\delta L(t_1) = \delta L(t_2) = 0$. The functional
\begin{equation}
\mathcal{S}[\ket{\Psi}] \equiv \int_0^T \braket{\Psi(t)|\hat{H}-i\hbar \frac{\partial}{\partial t}|\Psi(t)} dt \label{ActionFunctional1}
\end{equation}
is referred to as the action functional and computing 
equations of motion from $\delta \mathcal{S} = 0$ is known as the prinicipal of least action.

By considering a bivariational generalization of the action functional \eqref{ActionFunctional1},
\begin{align}
 \mathcal{S}[\bra{\Psi^{\prime}}, \ket{\Psi}] \equiv &\int_0^T \frac{\braket{\Psi^{\prime}(t)|i\hbar \frac{\partial}{\partial t}-\hat{H}|\Psi(t)}}{\braket{\Psi^{\prime}(t)|\Psi(t)}} dt \\
 &\int_0^T i\hbar \frac{\braket{\Psi^{\prime}(t)|\frac{\partial}{\partial t}\Psi(t)}}{\braket{\Psi^{\prime}(t)|\Psi(t)}} - \mathcal{E}_H[\bra{\Psi^{\prime}(t)}, \ket{\Psi(t)}] dt. \label{ActionFunctional}
\end{align}
Kvaal derives equations of motions generalizing the Coupled Cluster method to the time domain. 
Here, one should note the appearance of the bivariational generalization of the energy expectation value,
\begin{equation}
 \mathcal{E}_H[\bra{\Psi^{\prime}}, \ket{\Psi}] = \frac{\braket{\Psi^{\prime} |\hat{H}| \Psi} }{\braket{\Psi^{\prime}|\Psi}} \label{BivarEnergy},
\end{equation}
which is required to be complex analytic as shown in Ref.\cite{Kvaal13}.

\subsubsection{The Coupled Cluster Ansatz}
The fundamental idea in\cite{Kvaal12} is to introduce different exponential parametrizations $\ket{\Psi}$, $\bra{\Psi^\prime}$ and 
compute $\delta \mathcal{S} = 0$, with $\mathcal{S}$ given by \eqref{ActionFunctional}, which ultimately results in time dependent Coupled Cluster equations.

In particular one make the ansatzes 
\begin{align}
 \ket{\Psi} &= e^{\hat{T}}\ket{\phi} \\ 
 \bra{\Psi^{\prime}} &= \bra{\tilde{\phi}}e^{\hat{T}^\prime},
\end{align}
where $\hat{T}$ is the familiar excitation operator
\begin{equation}
 \hat{T} = \sum_{ia} \tau_i^a c_a^\dagger \tilde{c}_i + \frac{1}{2!^2} \sum_{ijab} \tau_{ij}^{ab} c_a^\dagger \tilde{c}_i c_b^\dagger \tilde{c}_j + \cdots
\end{equation}
and $\hat{T}^{\prime}$ is a de-excitation operator on the form
\begin{equation}
 \hat{T}^{\prime} = \sum_{ia} (\tau^{\prime})_a^i c_i^\dagger \tilde{c}_a + \frac{1}{2!^2} \sum_{ijab} (\tau^{\prime})_{ab}^{ij} c_i^\dagger \tilde{c}_a c_j^\dagger \tilde{c}_b + \cdots.
\end{equation}
Here $\ket{\phi} = \ket{\varphi_{p_1},\cdots,\varphi_{p_N}}$ and $\bra{\tilde{\phi}} = \bra{\tilde{\varphi}_{p_1},\cdots,\tilde{\varphi}_{p_N}}$ are Slater determinants 
built from biorthogonal single-particle functions, $\braket{\tilde{\varphi}_p|\varphi_q} = \delta_{pq}$,  belonging to separate 
Hilbert spaces $\mathcal{V}$ and $\tilde{\mathcal{V}}$. The biorthogonality condition implies the anticommutator 
relation 
\begin{equation}
 \{ \tilde{c}_p, c_q^\dagger \} \equiv \tilde{c}_p c_q^\dagger + c_q^\dagger \tilde{c}_p = \delta_{pq},
\end{equation}
which preserves Wicks theorem. Here $\tilde{c}_p, \tilde{c}_p^\dagger, c_p$ and $c_p^\dagger$ are creation and annihilation opeartotrs associated with 
the biorthogonal single-particle functions. It should be noted that when $\tilde{c}_p$ or $\tilde{c}_p^\dagger$ acts on the determinant $\ket{\phi}$ it 
is only responsible for removing or adding $\varphi_p$ (not $\tilde{\varphi}_p$) from the determinant\cite{Kvaal12}. The notation 
$\Phi = (\varphi_1,\cdots,\varphi_L)$ and $\tilde{\Phi} = (\tilde{\varphi}_1,\cdots,\tilde{\varphi}_L)$ to denote the set of orbitals used to build 
the Slater determinants $\ket{\phi}$ and $\bra{\tilde{\phi}}$.

In standard CC theory and virtually every other many-body method, the single-particle functions 
are taken to be orthonogal such that $\mathcal{V} \equiv \tilde{\mathcal{V}}^\prime$. This relaxation of
orthonormality of the single-particle functions is necessary to ensure that the bivariational functional is
complex analytic if the single-particle functions are to be treated as variational parameters\cite{Kvaal12}.

Next, a change of variables from $(T^{\prime},T)$ to $(\Lambda, T)$, with
\begin{equation}
\Lambda = \sum_{ia} \lambda_a^i c_i^\dagger \tilde{c}_a + \frac{1}{2!^2} \sum_{ijab} \lambda_{ab}^{ij} c_i^\dagger \tilde{c}_a c_j^\dagger \tilde{c}_b + \cdots 
\end{equation} 
a de-excitation operator is performed such that
\begin{equation}
 \bra{\tilde{\Psi}} = \frac{\bra{\Psi^{\prime}}}{\braket{\Psi^{\prime}|\Psi}} = \bra{\tilde{\phi}}(1 + \Lambda)e^{-T}.
\end{equation}
Using this parametrization $\braket{\tilde{\Psi}|\Psi} = 1$. The amplitudes $\{\tau_i^a,\tau_{ij}^{ab},\cdots \}$ are the usual Coupled Cluster amplitudes
which we refer to as the excitation amplitudes or $\tau$-amplitudes. 
Also we have introduced a second set of amplitudes $\{ \lambda_a^i, \lambda_{ab}^{ij},\cdots \}$ 
which we will refer to as the de-excitation amplitudes or $\lambda$-amplitudes.

Now the bivariational expectation value functional \eqref{BivarEnergy} reads 
\begin{equation}
 \mathcal{E}_H [\lambda, \tau, \tilde{\Phi}, \Phi] = \braket{\tilde{\phi}|(I+\Lambda)e^{-T}\hat{H}e^{T}|\phi}.
\end{equation}
Inserting the wavefunction ansatzes into the action functional \eqref{ActionFunctional} one arrives at the following expression,
\begin{equation}
 \mathcal{S}[\lambda,\tau,\tilde{\Phi},\Phi] = \int_0^T i\hbar \braket{\tilde{\phi}|(I+\Lambda)e^{-T}\frac{\partial}{\partial t} e^{T}|\phi} - \mathcal{E}_H[\lambda(t),\tau(t),\tilde{\Phi}(t), \Phi(t)] dt.
\end{equation}
Kvaal then shows that the action can be written,
\begin{align}
 \mathcal{S}[\lambda,\tau,\tilde{\Phi},\Phi] &= \int_0^T i \hbar \sum_\mu \lambda_\mu \dot{\tau}^\mu - \mathcal{E}_{H-i\hbar D_0}[\lambda, \tau, \tilde{\Phi}, \Phi] dt \\ 
 &=\int_0^T i \hbar \lambda_\mu \dot{\tau}^\mu + \rho_p^q(h_q^p - i \hbar \eta_q^p) + \frac{1}{4} \rho_{pr}^{qs}u^{pr}_{qs} dt,
\end{align}
where the operator $D_0$ is defined as,
\begin{equation}
 D_0 \equiv \sum_{pq} \braket{\tilde{\varphi}_p|\dot{\varphi}_q}c_p^\dagger \tilde{c}_q.
\end{equation}
Here $\mu$ denotes a general excitation $\{ia, ijab, \cdots\}$. Furthermore,  
\begin{align*}
 \rho_p^q &= \rho_p^q(\lambda,\tau) \equiv \braket{\tilde{\Psi}|c_p^\dagger \tilde{c}_q|\Psi}, \\
 \rho_{pr}^{qs} &= \rho_{pr}^{qs}(\lambda,\tau) \equiv  \braket{\tilde{\Psi}|c_p^\dagger c_r^\dagger \tilde{c}_s \tilde{c}_q|\Psi}, \\
 h^p_q &= h^p_q(\tilde{\Phi}, \Phi) \equiv \braket{\tilde{\varphi}_p|h|\varphi_q}, \\
 \eta_q^p &= \eta_q^p(\tilde{\Phi}, \dot{\Phi}) \equiv \braket{\tilde{\varphi}_p|\dot{\varphi}_q}, \\
 u^{pr}_{qs} &= u^{pr}_{qs}(\tilde{\Phi},\Phi) \equiv \braket{\tilde{\varphi}_p \tilde{\varphi}_r|u|\varphi_q \varphi_s}_{AS}. \\
\end{align*}
The quantities $\rho_p^q$ and $\rho_{pr}^{qs}$ are the CC reduced one- and two-body density matrices. Note that these do not depend explicitly on the orbitals since 
they are evaluated by Wick's theorem which only depends on the anti-commutator relations \eqref{AntiCommutatorRelations}. Thus, they only depend on the amplitudes.

\subsubsection{OATDCC equations}
Finally, equations of motion are derived by demanding $\delta \mathcal{S}[\bra{\tilde{\Psi}},\ket{\Psi}] = 0$ for all independent variations 
of $\bra{\tilde{\Psi}}$ and $\ket{\Psi}$. In general both the single-particle functions and the amplitudes are allowed to vary resulting in equations of motion 
for both the amplitudes and the single-particle functions. A detailed derivation can be found in the supplementary material to\cite{Kvaal12}.

For the amplitudes the equations of motion read,
\begin{align}
 i \hbar \dot{\tau}^\mu &= \frac{\partial}{\partial \lambda_\mu } \mathcal{E}_{H-i\hbar D_0} [\lambda, \tau, \tilde{\Phi}, \Phi ] = \braket{\tilde{\phi}_\mu | e^{-T} \left(H-i \hbar D_0\right) e^T | \phi} \label{TAmplitudesEquation}, \\
 -i \hbar \dot{\lambda}_\mu &= \frac{\partial}{\partial \tau^\mu } \mathcal{E}_{H-i\hbar D_0} [\lambda, \tau, \tilde{\Phi}, \Phi ] = \braket{ \tilde{\phi} | (1+\Lambda)e^{-T} [H-i \hbar D_0,X_\mu] e^T | \phi} \label{LAmplitudesEquation}.
\end{align}
Here $X_\mu$ are the excitation operators, 
\begin{align*}
 X_i^a &= c_a^\dagger \tilde{c}_i, \\
 X_{ij}^{ab} &= c_a^\dagger \tilde{c}_i c_b^\dagger \tilde{c}_j, \\ 
 &\vdots \\
\end{align*}
Note that in the absence of $D_0$ the right-hand side of \eqref{TAmplitudesEquation} is equivalent with the usual amplitude equations \eqref{SimTransAmplitudeEquation}.

The equations for the unknowns $\eta_q^p(\tilde{\Phi}, \dot{\Phi})$ are given by
\begin{align}
 i\hbar \sum_{bj} A_{aj}^{ib} \eta_b^j &=   \sum_p \rho_p^i h_a^j - \sum_q \rho_a^q h_b^i + \frac{1}{2} \left[ \sum_{prs} \rho_{pr}^{is}u_{as}^{pr} - \sum_{rqs} \rho_{ar}^{qs}u_{qs}^{ir} \right] \\
 -i\hbar \sum_{bj} A_{bi}^{ja} \eta_j^b &=  \sum_p \rho_p^a h_i^b -\sum_q \rho_i^q h_j^a + \frac{1}{2} \left[ \sum_{prs} \rho_{pr}^{as}u_{is}^{pr} - \sum_{rqs} \rho_{ir}^{qs}u_{qs}^{ar} \right] + i \hbar \dot{\rho}_i^a,
\end{align}
where 
\begin{equation*}
 A_{aj}^{ib} \equiv \langle \tilde{\Psi} | [c_j^\dagger \tilde{c}_b, c_a^\dagger \tilde{c}_i] | \Psi \rangle = \delta_a^b \rho_j^i - \delta_j^i \rho_a^b. 
\end{equation*}
For the single-particle functions the equations of motion are given by 
\begin{align}
  i\hbar \sum_q \rho_p^q Q \ket{\dot{\varphi}_q} &= \sum_q \rho_p^q Q h \ket{\varphi_q} + \sum_{qrs} \rho_{pr}^{qs} Q W_s^r \ket{\varphi_q} \label{QspaceketEq}\\
 -i\hbar \sum_p \rho_p^q\bra{\dot{\tilde{\varphi}}_p}Q &= \sum_p \rho_p^q\bra{\tilde{\varphi}_p}hQ + \sum_{prs} \rho_{pr}^{qs} \bra{\tilde{\varphi}_p}W_s^rQ,  \label{QspacebraEq}
\end{align}
where $Q = I-P$ with $P$ being the projection operator
\begin{equation}
 \Phi \tilde{\Phi} \equiv \sum_p \ket{\varphi_p}\bra{\tilde{\varphi}_p}.
\end{equation}
The operators $W^r_s$ are defined by,
\begin{equation}
 W^r_s \ket{\psi} \equiv \braket{\cdot \tilde{\varphi}_r|u|\psi \varphi_s} \equiv \int dx_1 \int dx_2 \tilde{\varphi}_r(x_2) u(x_1,x_2) \psi(x_1) \varphi_s(x_2) 
\end{equation}
Equations \eqref{TAmplitudesEquation}-\eqref{QspacebraEq} is collectively referred to as the OATDCC equations.

It is not at all obvious how one should compute the 
initial state for the OATDCC equations when the single-particle functions are allowed to vary 
and Kvaal proposes several different solutions in his article. 
\subsubsection{The TDCC equations}
A special case of OATDCC arises if the orbitals are held fixed in time. Then the operator $D_0$ drops and the OATDCC equations is reduced to only the amplitude 
equations
\begin{align}
 i \hbar \dot{\tau}^\mu &= \frac{\partial}{\partial \lambda_\mu } \mathcal{E}_{H} [\lambda, \tau, \tilde{\Phi}, \Phi ] = \langle \tilde{\phi}_\mu | e^{-T} H e^T | \phi \rangle \label{TAmplitudesEquationTDCC} \\
 -i \hbar \dot{\lambda}_\mu &= \frac{\partial}{\partial \tau^\mu } \mathcal{E}_{H} [\lambda, \tau, \tilde{\Phi}, \Phi ] = \langle \tilde{\phi} | (1+\Lambda)e^{-T} [H,X_\mu] e^T | \phi \rangle \label{LAmplitudesEquationTDCC},
\end{align}
which we refer to as the TDCC-equations. Notice now that equations \eqref{TAmplitudesEquationTDCC} and \eqref{SimTransAmplitudeEquation} are equivalent. Furthermore, 
we take $\tilde{\Phi} = \Phi^*$.

The initial amplitudes $(\lambda^{(0)},\tau^{(0)})$ are obtained by solving the non-linear equations
\begin{align}
 \langle \tilde{\phi}_\mu | e^{-T} H e^T | \phi \rangle &= 0\label{TAmplitudesEquationTDCCGS} \\
 \langle \tilde{\phi} | (1+\Lambda)e^{-T} [H,X_\mu] e^T | \phi \rangle &= 0 \label{LAmplitudesEquationTDCCGS}.
\end{align}

For the time evolution one then propgates the amplitudes by solving equations \eqref{TAmplitudesEquationTDCC} and \eqref{LAmplitudesEquationTDCC}. 
The energy is taken as the real part of
\begin{equation}
 \mathcal{E}_H [\lambda(t), \tau(t)] = \braket{\tilde{\phi}|(I+\Lambda)e^{-T}\hat{H}e^{T}|\phi} \label{CCexpecval},
\end{equation}
where the imaginary part should be "small".

Approximations to the TDCC equations are obtained by truncating $T$ and $\Lambda$. For example 
$T = T_1 + T_2$ and $\Lambda = \Lambda_1 + \Lambda_2$ corresponds to the TDCCSD approximation while $T = T_2$ and $\Lambda = \Lambda_2$ 
gives the TDCCD approximation.

It is shown in \cite{Kvaal12} that the singles amplitudes are redundant when the single-particle functions are varied. We will therefore restrict ourselves to the 
TDCCD approximation, using the Hartree-Fock state as reference determinant. This allows for flexibility if one wants to extend the program written in this work 
to the full OATDCC method.

\subsubsection{The TDCCD approximation} \label{TDCCD approximation section}
In order to write a computer program that solves the TDCC equations we need algebraic expressions for the right-hand sides of equations \eqref{TAmplitudesEquationTDCC}
and \eqref{LAmplitudesEquationTDCC}. Now, in the TDCCD approximation we take $T = T_2$ and $\Lambda = \Lambda_2$. 
We want expressions for the amplitudes and the one- and two-body density matrices. Thus we must evaluate the vacuum expectation values,
\begin{align}
 \frac{\partial}{\partial \lambda_{ab}^{ij} } \mathcal{E}_{H} &= \braket{\phi_{ij}^{ab}|e^{-T_2}He^{T_2}|\phi} \label{ExpecTamplitude}\\
 \frac{\partial}{\partial \tau^{ab}_{ij} } \mathcal{E}_{H} &= \braket{\phi|(1+\Lambda_2)e^{-T_2}[H, X_{ij}^{ab}]e^{T_2}|\phi}, \label{ExpexLamplitude}\\
 \rho_p^q &= \braket{\tilde{\Psi}|c_p^\dagger \tilde{c}_q|\Psi}, \\
 \rho_{pr}^{qs} &=  \braket{\tilde{\Psi}|c_p^\dagger c_r^\dagger \tilde{c}_s \tilde{c}_q|\Psi}.
\end{align}
These are generated by the \textit{CoupledClusterSympy.py} script, which uses the second quantization package in SymPy, 
and is found in the src folder at the github page\footnote{\href{https://github.com/haakoek/PythonVersionMaster/tree/master/src}{https://github.com/haakoek/PythonVersionMaster/tree/master/src}}. 
The expressions for the amplitude equations and one- and two-body density matrices are listed in the appendices.

Finally, the energy is taken as the real part of the expectation value \eqref{CCexpecval} which in the CCD approximation is given by,
\begin{align*}
 \mathcal{E}_H [\lambda(t), \tau(t)] &= \braket{\tilde{\phi}|(I+\Lambda_2)e^{-T_2}\hat{H}e^{T_2}|\phi} \\
				     &= \braket{\tilde{\phi}|e^{-T_2}\hat{H}e^{T_2}|\phi} + \braket{\tilde{\phi}|\Lambda_2e^{-T_2}\hat{H}e^{T_2}|\phi} \\
				     &= \braket{\tilde{\phi}|e^{-T_2}\hat{H}e^{T_2}|\phi} + \frac{1}{4} \sum_{ijab} \lambda_{ij}^{ab}\braket{\tilde{\phi}_{ij}^{ab}|e^{-T_2}\hat{H}e^{T_2}|\phi}
\end{align*}
where 
\begin{equation}
 \braket{\tilde{\phi}|e^{-T_2}\hat{H}e^{T_2}|\phi} = h_i^i + \frac{1}{2}u_{ij}^{ij} + \frac{1}{4}\tau_{ij}^{ab}u_{ab}^{ij}. \label{ECCDGroundstate}
\end{equation}
Notice that in the groundstate,
\begin{equation*}
 \braket{\tilde{\phi}_{ij}^{ab}|e^{-T_2}\hat{H}e^{T_2}|\phi} = 0
\end{equation*}
which means that the energy for the groundstate is simply given by \eqref{ECCDGroundstate}.

Note also that the one- and two-body density matrices depend only on the amplitudes $(\tau,\lambda)$. This is an advantage when we want to verify the implementation of 
the $\lambda$-amplitudes since we can compare the density matrices with the density matrices computed with the CI method.

In order to verify that the $\tau$-amplitudes has been computed correctly for the groundstate 
we compute the groundstate energy and compare with previous published results.

\section{Computational details}

\section{Results}

\section{Conclusions}

\begin{appendices}
 \section{Properties of Cluster operators.}
\begin{itemize}
\item $\hat{T}_{N+1} \ket{\Phi} = 0$, where $\hat{T} = \hat{T}_1 + \hat{T}_2 + \cdots$, $\ket{\Phi}$ a reference determinant and $N$ the number of particles 
in $\ket{\Phi}$. This implies that for a given $N$ the cluster operator $\hat{T}$ has a finite number of terms,
\begin{equation}
 \hat{T} = \hat{T}_1 + \hat{T}_2 + \cdots + \hat{T}_N.
\end{equation}
 \item The rank of 
an excitation/cluster operator is defined as the number of creation or annihilation operators, i.e rank$\left(\hat{T}_2\right) = 2$.
\item The product of two exciation/cluster operators is an exciation/cluster operator.
\item All excitation/cluster operators commute. We can prove this by considering the action of the commutator on a reference determinant as follows,
\begin{align*}
 [c_b^\dagger c_j, c_a^\dagger c_i] \ket{\Phi} &= \left(c_b^\dagger c_j c_a^\dagger c_i - c_a^\dagger c_i c_b^\dagger c_j \right) \ket{\Phi} \\
 &= \ket{\Phi_{ij}^{ab}} \ket{\Phi_{ji}^{ba}} \\
 &= \ket{\Phi_{ij}^{ab}} - \ket{\Phi_{ij}^{ab}} \\
 &= 0 \cdot \ket{\Phi_{ij}^{ab}}.
\end{align*}
Thus $c_b^\dagger c_j c_a^\dagger c_i = c_a^\dagger c_i c_b^\dagger c_j$, meaning that $c_b^\dagger c_j$ and $c_a^\dagger c_i$ commute. The result for a 
general excitation operator follows by induction.
\item rank$\left(\hat{T}_n \hat{T}_m \right) = n+m$.
\item Excitation operators are nilpotent, 
\begin{equation}
 \hat{\tau}_X^2 = \left(\cdots c_c^\dagger c_k c_b^\dagger c_j c_a^\dagger c_i \right)^2 = 0.
\end{equation}
This follows immediately from the fact that, 
\begin{align*}
 \left(c_a^\dagger c_i\right)^2 \ket{\Phi} = c_a^\dagger c_i \ket{\Phi_i^a} = 0 
\end{align*}
since $c_i \ket{\Phi_i^a} = 0$.
\item $\hat{\tau}_X^\dagger = \left(\cdots c_c^\dagger c_k c_b^\dagger c_j c_a^\dagger c_i \right)^\dagger = \left(c_i^\dagger c_a c_j^\dagger c_b c_k^\dagger c_c \cdots\right) $ 
is a de-excitation operator. Thus,
\begin{equation}
 \bra{\Phi}\hat{\tau}_X = \hat{\tau}^\dagger_X \ket{\Phi} \equiv 0.
\end{equation}
\item Since all cluster operators commute, $e^{\hat{T}+\hat{T}^\prime} = e^{\hat{T}} e^{\hat{T}^\prime}$.
\end{itemize}

\section{Existence of cluster expansion.}
Suppose $\braket{\Phi|\Psi} = 1$. Then there exists cluster operators $\hat{T}$ and $\hat{A}$ such that 
\begin{equation}
 \ket{\Psi} = \left( 1 + \hat{A} \right) \ket{\Phi} = e^{\hat{T}} \ket{\Phi}.
\end{equation}
\subparagraph*{Proof:} Let $X$ denote a general excitation index and let $\tau_X$ be a general excitation operator. Assume that 
$\braket{\Phi|\Psi} = 1$ and let $$\{ \ket{\Phi}, \ket{\Phi_X} | \text{ for all } X \}$$ be a Slater determinant basis where $\braket{\Phi|\Phi_X} = 0$. We can expand $\ket{\Psi}$ in the basis, giving
\begin{equation}
 \ket{\Psi} = A_0 \ket{\Phi} + \sum_X A_X \ket{\Phi_X}, \ \ A_0, A_X \in \mathbb{C}.
\end{equation}
Note that $\braket{\Phi|\Psi} = 1$ implies that $A_0 = 1$. Using $\ket{\Phi_X} = \hat{\tau}_X \ket{\Phi}$ we get, 
\begin{equation}
 \ket{\Psi} = \ket{\Phi} + \sum_X A_X \hat{\tau}_X \ket{\Phi} = \left(1 + \hat{A}\right) \ket{\Phi},
\end{equation}
where we have defined $\hat{A} \equiv \sum_X A_X \hat{\tau}_X$. We now want to find $\hat{T}$ such that $e^{\hat{T}} = 1 + \hat{A}$. Rewrite $\hat{A}$ 
as,
\begin{align*}
 \hat{A} = \sum_X A_X \hat{\tau}_X &= \sum_{ia} A_i^a \hat{\tau}_i^a + \frac{1}{2!^2}\sum_{ijab} A_{ij}^{ab} \hat{\tau}_{ij}^{ab} + \cdots \\
 &= \hat{A}_1 + \hat{A}_2 + \cdots
\end{align*}
By expanding $e^{\hat{T}}$ as a series
\begin{equation}
 e^{\hat{T}} = 1 + \hat{T} + \frac{1}{2} \hat{T}^2 + \cdots \label{ExponantialExpansion},
\end{equation}
and note that $e^{\hat{T}} = 1 + \hat{A}$ if they have the same singles, doubles, etc. parts. Let $N$ be given and insert $\hat{T} = \hat{T}_1 + \cdots + \hat{T}_N$
into eq. \eqref{ExponantialExpansion}, expand each power and group terms of equal rank together,
\begin{align*}
 e^{\hat{T}} &= 1 + \hat{T} + \frac{1}{2} \hat{T}^2 + \frac{1}{6} \hat{T}^3 + \cdots \\
 &= 1 + \left(\hat{T}_1 + \cdots + \hat{T}_N \right) + \frac{1}{2} \left( \hat{T}_1 + \cdots + \hat{T}_N \right)^2 \\
 &+ \frac{1}{6} \left( \hat{T}_1 + \cdots + \hat{T}_N \right)^3 + \frac{1}{24} \left( \hat{T}_1 + \cdots + \hat{T}_N \right)^4 + \cdots \\
 &= 1 + \underbrace{\hat{T}_1}_{\text{rank} 1} + \underbrace{\left( \hat{T}_2 + \frac{1}{2}\hat{T}_1^2 \right)}_{\text{rank} 2} + \underbrace{\left( \hat{T}_3 + \frac{1}{6} \hat{T}_1^3 + \hat{T}_1 \hat{T}_2 \right)}_{\text{rank} 3} + \cdots \\
\end{align*}
Note that in the last equation, the rank $k$ term contains only one term from $\hat{T}_k$ while all other terms are products of cluster operators of lower rank.
For the above expression to be equal to $1 + \hat{A}_1 + \hat{A}_2 + \cdots$ each rank $k$ term must equal $\hat{A}_k$.
\begin{align*}
 \hat{A}_1 = &\hat{T}_1 \\
 \hat{A}_2 = &\hat{T}_2 + \frac{1}{2}\hat{T}_1^2 \\
 \hat{A}_3 = &\hat{T}_3 + \frac{1}{6} \hat{T}_1^3 + \hat{T}_1 \hat{T}_2 \\
  &\vdots
\end{align*}
Then we can find $\hat{T}_k$ recursively, 
\begin{align}
 \hat{T}_1 = &\hat{A}_1 \\
 \hat{T}_2 = &\hat{A}_2 - \frac{1}{2}\hat{T}_1^2 = \hat{A}_2 - \frac{1}{2}\hat{A}_1^2  \\
 \hat{T}_3 = &\hat{A}_3 - \frac{1}{6} \hat{T}_1^3 - \hat{T}_1 \hat{T}_2 \\
           = &\hat{A}_3 - \frac{1}{6} \hat{A}_1^3 - \hat{A}_1 \left( \hat{A}_2 - \frac{1}{2}\hat{A}_1^2 \right) \\
 & \vdots
\end{align}
Hence, given $\ket{\Psi}$, we can find $\hat{A}$ which in turn gives $\hat{T}$.

We notice that the expansion $\ket{\Psi} = (1+\hat{A}) \ket{\Phi}$ actually is the CI expansion. Thus, the 
CC expansion $e^{\hat{T}}\ket{\Phi}$ is a nonlinear reparametrization of the linear CI parametrization. In practice, for a system containing $N$ particles, 
one must truncate $\hat{T} = \hat{T}_1 + \cdots + \hat{T}_N$ and $\hat{A} = \hat{A}_1 + \cdots + \hat{A}_N$. Now, truncate the expansion at the 
$k$-th operator in the sum with $1 \leq k \leq N$ such that $\hat{T} = \hat{T}_1 + \cdots + \hat{T}_k$ and $\hat{A} = \hat{A}_1 + \cdots + \hat{A}_k$. 
Observe that 
\begin{align*}
\ket{\Psi} &= (1+\hat{A}_1 + \cdots + \hat{A}_k) \ket{\Phi} \\
\end{align*}
is a linear combination where the highest excitation contribution has rank $k$. The exponential ansatz on the other hand 
\begin{align*}
 e^{\hat{T}_1 + \cdots + \hat{T}_k} \ket{\Phi} = \left( 1 + \left( \hat{T}_1 + \cdots + \hat{T}_k \right) + \frac{1}{2} \left( \hat{T}_1 + \cdots + \hat{T}_k \right)^2 + \cdots \right) \ket{\Phi}
\end{align*}
contains contributions from determinants of higher excitation than rank $k$ which is immediately clear from the inclusion of the term $\hat{T}_k^2$ 
in the second parenthesis in the above expression. It turns out that this gives the truncated CC ansatz an advantage over 
the truncated CI ansatz. The latter is not size-consistent and extensive while the former is.

Loosely speaking size-consistency and extensivity 
means that if $A$ and $B$ are two noninteracting subsystems, we should get the same energy for the supersystem $AB$ irrespective of whether we have
carried out the calculations for each subsystem separately or for both subsystems simultaneously. In general 
truncated exponential wavefunctions have this property while truncated linear wavefunctions does not\cite{Helgaker00book}.

Depending on the truncation we get a sequence of more and more 
accurate CC approximations:
\begin{itemize}
 \item $\ket{\Psi_{CCS}} = e^{\hat{T}_1} \ket{\Phi}$ coupled-cluster singles (CCS)
 \item $\ket{\Psi_{CCD}} = e^{\hat{T}_2} \ket{\Phi}$ coupled-cluster doubles (CCD)
 \item $\ket{\Psi_{CCSD}} = e^{\hat{T}_1 + \hat{T}_2} \ket{\Phi}$ coupled-cluster singles doubles (CCSD)
 \item $\ket{\Psi_{CCSDT}} = e^{\hat{T}_1 + \hat{T}_2 + \hat{T}_3} \ket{\Phi}$ coupled-cluster singles doubles triples (CCSDT)
\end{itemize}

\section{Equality of CC and CI for $N=2$.}

An important special case is the case $N=2$. In that case the CC expansion equals the CI expansion. This is invaluable information when we want to implement 
the CC method since we can then compare it with the corresponding CI method to verify the implementation. To see this let $N=2$ and let 
$\ket{\Phi}$ be a reference determinant as usual. Observe that $\hat{A}_3 \ket{\Phi} = 0 $ and $\hat{T}_3 = 0$. Thus, 
$\hat{A} = \hat{A}_1 + \hat{A}_2$ and $\hat{T} = \hat{T}_1 + \hat{T}_2$ will give all non-zero contributions. Additionaly all products of cluster operators 
where the product has rank larger than $2$, acting on the reference will give a zero result.
Inserting 
 $\hat{T}_1 = \hat{A}_1$ and $\hat{T}_2 = \hat{A}_2 - \frac{1}{2}\hat{A}_1^2$ into the CC expansionincluding only terms of at most rank $2$ we obtain,  
 \begin{align*}
  e^{\hat{T}_1 + \hat{T}_2} \ket{\Phi} &= \left( 1 + \underbrace{\hat{T}_1}_{\text{rank} 1} + \underbrace{\left( \hat{T}_2 + \frac{1}{2}\hat{T}_1^2 \right)}_{\text{rank} 2} \right) \ket{\Phi} \\
  &= \left(1+ \hat{A}_1 + \hat{A}_2 - \frac{1}{2}\hat{A}_1^2 + \frac{1}{2} \hat{A}_1^2 \right) \ket{\Phi} \\
  &= \left(1+ \hat{A}_1 + \hat{A}_2 \right) \ket{\Phi}
 \end{align*}
which indeed is the CI expansion. For $N=2$, $\hat{A} = \hat{A}_1 + \hat{A}_2$ corresponds to a full configuration interaction treatment, while 
$\hat{T} = \hat{T}_1 + \hat{T}_2$ CCSD, but in this case CCSD is actually equal to FCI. Morover, if we consider only $\hat{T} = \hat{T}_2$ (CCD), we get 
\begin{align*}
 \ket{\Psi}_{\text{CCD}} = e^{\hat{T}_2} \ket{\Phi} &= \left(1+ \hat{T}_2 \right) \ket{\Phi} \\
 &= \left( 1 + \hat{A}_2 - \frac{1}{2}\hat{A}_1^2 + \frac{1}{2} \hat{A}_1^2 \right) \ket{\Phi} \\
 &= \left(1 + \hat{A}_2 \right) \ket{\Phi} = \ket{\Psi}_{\text{CID}}.
\end{align*}
The importance of these two results is that one can verify a CCD implementation by comparing with a CID implementation and a CCSD implementation can be 
verified by comparing with a FCI/CISD implementation.

\section{Amplitude equations}
Recall that the Hamiltonian in second quantization is given by,
\begin{equation}
 \hat{H} = \hat{H}^{(1)} + \hat{H}^{(2)} = \sum_{pq} h^p_q c_p^\dagger c_q + \frac{1}{4} \sum_{pqrs} u^{pr}_{qs} c_p^\dagger c_r^\dagger c_s c_q
\end{equation}
where $ u^{pr}_{qs} = \braket{\varphi_p \varphi_r|\hat{H}^{(2)}|\varphi_q \varphi_s}_{AS}$.
Evaluation of the right hand side of equation \eqref{ExpecTamplitude} gives,
\begin{equation}
 \frac{\partial}{\partial \lambda_{ij}^{ab}}E_{H^{(1)}} = h^{k}_{i} \tau^{ab}_{jk} P(ij) + h^{a}_{c} \tau^{bc}_{ji} P(ab) \label{AlgebraicTampH1},
\end{equation}
\begin{align}
 \frac{\partial}{\partial \lambda_{ab}^{ij}}E_{H^{(2)}} = &-\tau^{ab}_{ik} u^{kl}_{jl} P(ij) + \tau^{ab}_{il} \chi_j^l P(ij) \nonumber \\
 &+\tau^{ab}_{kl} \chi^{kl}_{ij} +\tau^{ac}_{ij}\chi_c^b P(ab) \nonumber \\
 &+\tau^{ac}_{ik} \chi_{cj}^{kb}P(ab)P(ij) -\frac{1}{2}\tau^{dc}_{ji} u^{ab}_{dc} \nonumber \\ 
 &+u^{ab}_{ij} \label{AlgebraicTampH2},
\end{align}
where $P(ij)$ is an anti-symmtrizer in the sense that $f(ij)P(ij) = f(ij)-f(ji)$ and similarly for $P(ab)$. We have defined the so-called intermediates,
\begin{align}
 \chi^{kl}_{ij} &= \frac{1}{4} \tau^{cd}_{ij} u^{kl}_{cd} + \frac{1}{2} u^{kl}_{ij} \label{Xklij} \\
 \chi_c^b &= \frac{1}{2} \tau^{bd}_{kl}  u^{kl}_{dc}  +  u^{bk}_{ck} \label{Xcb} \\
 \chi_j^l &= \frac{1}{2}\tau^{dc}_{jk} u^{kl}_{dc} \label{Xjl} \\
 \chi_{cj}^{kb} &= u^{kb}_{cj} + \frac{1}{2} u^{kl}_{cd}\tau^{bd}_{jl} \label{Xcjkb} .
\end{align}
We use the Einstein summation convention over repeated indices of opposite vertical placement. The appearance of a $P(ij)$ or a $P(ab)$ should be ignored
for the invocation of the summation convention. This means that for example the expression $h^{k}_{i} \tau^{ab}_{jk} P(ij)$ is interpreted as,
\begin{equation*}
 h^{k}_{i} \tau^{ab}_{jk} P(ij) = \sum_k \left(h^{k}_{i} \tau^{ab}_{jk} - h^{k}_{j} \tau^{ab}_{ik} \right).
\end{equation*}
In other words, we sum over all indices that does not appear on the left-hand side of all following expressions.

Evaluation of the right hand side of equation \eqref{ExpexLamplitude} results in,
\begin{align}
 \frac{\partial}{\partial \tau_{ij}^{ab}}E_{H^{(1)}} = h^{c}_{a} \lambda^{ji}_{bc} P(ab) + h^{i}_{k} \lambda^{jk}_{ab} P(ij) \label{AlgebraicLampH1},
\end{align}
\begin{align}
 \frac{\partial}{\partial \tau_{ij}^{ab}}E_{H^{(2)}} = &\lambda^{kl}_{ac} \xi^{cji}_{klb} P(ab) +\lambda^{ik}_{dc} \xi^{dcj}_{kab}  P(ij) \nonumber \\
 +&\lambda^{kl}_{ab}\xi_{kl}^{ij} +\lambda^{ik}_{ac}\xi_{kb}^{cj} P(ab) P(ij) \nonumber \\
 +&\lambda^{ik}_{ab}\xi_k^j P(ij) +\lambda^{ji}_{dc}\xi_{ab}^{dc} \nonumber \\
 +&\lambda^{ji}_{ac} \xi_b^c P(ab) -u^{ji}_{ab} \label{AlgebraicLampH2},
\end{align}
where we have defined the intermediates,
\begin{align}
 \xi^{cji}_{klb}      &= - \frac{1}{2}\tau^{dc}_{kl} u^{ji}_{bd} \label{Ecjiklb} \\
 \xi^{dcj}_{kab}            &= - \frac{1}{2}\tau^{dc}_{kl} u^{jl}_{ab} \label{Edcjkab} \\
 \xi_{kl}^{ij} &= -\frac{1}{2} u^{ji}_{kl}  - \frac{1}{4} \tau^{dc}_{kl}u^{ji}_{dc} \label{Eklij} \\
 \xi_{kb}^{cj} &= - \tau^{dc}_{kl} u^{jl}_{bd} +  u^{jc}_{bk} \\
 \xi_k^j &= -\frac{1}{2} \tau^{dc}_{kl} u^{jl}_{dc}  -  u^{jl}_{kl} \\
 \xi_{ab}^{dc} &= -\frac{1}{4}\tau^{dc}_{kl}u^{kl}_{ab} - \frac{1}{2}u^{dc}_{ab} \\
 \xi_b^c &= -\frac{1}{2} \tau^{dc}_{kl} u^{kl}_{bd} - u^{ck}_{bk} \label{Ebc}.
\end{align}
If we precompute and store all intermediates equation \eqref{AlgebraicTampH2} scale as 
$$\mathcal{O}\left(N^2(L-N)^4\right)$$ 
while 
equation \eqref{AlgebraicLampH2} scale as
$$\mathcal{O}\left(N^3(L-N)^4\right).$$
\section{Density matrices}
For the onebody density matrix we get
\begin{align}
 \rho_i^j &= \delta_{i j} - \frac{1}{2}\lambda^{kj}_{ab} \tau^{ab}_{ki}, \label{RHOij} \\
 \rho_a^b &= \frac{1}{2}\lambda^{ij}_{ac} \tau^{bc}_{ij}, \\
 \rho_i^a &= \rho_a^i = 0.
\end{align}
The non-zero elements of the two-body density matrix are given by,
\begin{align}
 \rho_{ij}^{kl} &= - \delta_{i l} \delta_{j k} P(ij) - \frac{1}{2}P(ij) P(kl)\delta_{j l}\lambda^{km}_{dc} \tau^{dc}_{im}   + \frac{1}{2}\lambda^{lk}_{dc} \tau^{dc}_{ji}, \\
 \rho_{ab}^{ij} &= \lambda^{ij}_{ab} \\
 \rho_{ia}^{jb} &= \frac{1}{2}\delta_{i j} \lambda^{kl}_{ac} \tau^{bc}_{kl}  - \lambda^{jk}_{ac} \tau^{bc}_{ik} = -\rho_{ia}^{bj} = -\rho_{ai}^{jb} = \rho_{ai}^{bj} \\
 \rho_{ab}^{cd} &= \frac{1}{2}\lambda^{ij}_{ab} \tau^{cd}_{ij} \\
 \rho_{ij}^{ab} &= - \frac{1}{4} \lambda^{kl}_{dc} \tau^{ab}_{ik} \tau^{dc}_{jl} P(ab) P(ij) - \frac{1}{4} \lambda^{kl}_{dc} \tau^{ab}_{kl} \tau^{dc}_{ji} - \frac{1}{4} \lambda^{kl}_{dc} \tau^{ac}_{kl} \tau^{bd}_{ji} P(ij) \nonumber \\
 &+ \lambda^{kl}_{dc} \tau^{ac}_{il} \tau^{bd}_{jk} P(ij) - \frac{1}{4} \lambda^{kl}_{dc} \tau^{ac}_{ji} \tau^{bd}_{kl} P(ij) - \tau^{ab}_{ji} \label{RHOijab}.
\end{align}

\end{appendices}
\end{comment}
\clearpage
\bibliographystyle{ieeetr}
\bibliography{references}

\end{document}

### 2018-04-14 doc/TimeDependence/Tex/Article.tex
Update on article

@@
-\documentclass[a4paper,10pt, twocolumn]{article}
+\documentclass[aip,jcp,reprint,floatfix]{revtex4-1}
+
+\usepackage{amsmath,amssymb}
+\usepackage{epsfig}
+\usepackage{graphicx}
+\usepackage{dcolumn}
+\usepackage{bbold}
+\usepackage{natbib}
+\usepackage[normalem]{ulem} % to strike out plain text
+\usepackage[colorlinks=true,citecolor={blue},urlcolor={blue}, linkcolor=blue]{hyperref}%
 \usepackage[utf8]{inputenc}
-\usepackage{amsmath}
 \usepackage{amsthm}
-\usepackage{amssymb}
 \usepackage{braket}
 \usepackage{verbatim}
-\usepackage{mdframed} %nice frames
-\usepackage[hidelinks]{hyperref}
-\usepackage{cite}
-\usepackage[toc,page]{appendix}
+\usepackage{mdframed}
 \newtheorem{theorem}{Theorem}
-\renewcommand{\thesection}{\Roman{section}}
-\renewcommand\thesubsection{\Alph{subsection}}
-%opening
-\title{Article}
-\author{Kristiansen, Håkon Emil}
 
@@
 
+\title{Time-dependent optimized coupled-cluster method for multielectron dynamics}
+
+\author{Håkon Emil Kristiansen}
+\email[Electronic mail:]{haakoek@student.matnat.uio.no}
+\affiliation{Department of Physics, University of Oslo, N-0316 Oslo, Norway}
+
+\begin{abstract}
+We present a working implementation of a simplified version of the
+orbital-adaptive time-dependent coupled-cluster method which we refer
+to as just the time-dependent coupled-cluster (TDCC) method.
+OATDCC/TDCC approximates solutions to the quantum $N$-body problem and
+for $N=2$ TDCC is equivalent to the time-dependent
+configuration-interaction method (TDCI).  A TDCI solver is implemented
+and shown to be correct by comparing with a previous study and is then
+used as a foundation for verifying the implementation of the TDCC
+method. For one and two dimensional quantum dot systems with $N=2$ we
+demonstrate that the methods are equal to numerical
+precision. Furthermore, for a small number of basis functions we use
+the TDCC solver to compute the time evolution of six-electron
+isotropic quantum dot.
+\end{abstract}
+
+\pacs{32.80.Rm, 31.15.A-, 42.65.Ky}
+
 \maketitle
 
-\begin{abstract}
-We present a working implementation of a simplified version of the orbital-adaptive time-dependent coupled-cluster method which we refer to as 
-just the time-dependent coupled-cluster (TDCC) method.  OATDCC/TDCC approximates solutions to the quantum $N$-body problem 
-and for $N=2$ TDCC is equivalent to the time-dependent configuration-interaction method (TDCI). 
-A TDCI solver is implemented and shown to be correct by comparing with a previous study 
-and is then used as a foundation for verifying the implementation of the TDCC method. For one and two dimensional quantum dot 
-systems with $N=2$ we demonstrate that the methods are equal to numerical precision. Furthermore, for a small number of basis functions we use the TDCC solver 
-to compute the time evolution of six-electron isotropic quantum dot.  
-\end{abstract}
+\section{Introduction}
 
-%\tableofcontents
-
-\section{Introduction}
-Quantum mechanics is a theory that describes/models the properties of microscopic systems. It postulates that given the so-called wavefunction, $\Psi(\mathbf{r},t)$, we can 
-in principle compute all that there is to know about the system. 
-Given suitable initial conditions $\Psi(\mathbf{r},t)$ can be determined for all future time by solving the time-dependent Schrödinger equation (TDSE),
+Quantum mechanics is a theory that describes/models the properties of
+microscopic systems. It postulates that given the so-called
+wavefunction, $\Psi(\mathbf{r},t)$, we can in principle compute all
+that there is to know about the system.  Given suitable initial
+conditions $\Psi(\mathbf{r},t)$ can be determined for all future time
+by solving the time-dependent Schrödinger equation (TDSE),
 \begin{equation}
@@
 \end{equation}
-Exact/analytical solutions to the TISE is possible only for the simplest systems and general solutions to the TDSE is even rarer. Thus, 
-one must resort to approximative/numerical methods. Of particular interest is the so-called many-body problem where one considers systems of a large number of interacting particles, 
-where large can be anywhere from two to infinity.
+Exact/analytical solutions to the TISE is possible only for the
+simplest systems and general solutions to the TDSE is even
+rarer. Thus, one must resort to approximative/numerical methods. Of
+particular interest is the so-called many-body problem where one
+considers systems of a large number of interacting particles, where
+large can be anywhere from two to infinity.
 
-Several approaches for solving the many-body TISE numerically/approximatively has been devised such as Hartree-Fock (HF), Density Functional Theory (DFT), 
-Configuration Interaction (CI) and Coupled Cluster (CC). While efficient, Hartree-Fock and DFT 
-are insufficient if one wants a high degree of accuracy. Configuration Interaction and Coupled Cluster methods are hierarchical in the sense that one can 
-systematically construct increasly accurate approximations. If the CI method is not truncated we have what is known as Full Configuration Interaction (FCI). FCI 
-can be seen as exact (within some finite space), however it suffers from exponential scaling. Truncated CI methods which would achieve polynomial scaling 
-are problematic since they are not size-consistent and extensive. Truncated CC methods one the other hand are size-consistent and extensive and achieves polynomial 
-scaling. Due to this fact Coupled Cluster is considered the gold standard of many-body techniques if high accuracy is desired.
+Several approaches for solving the many-body TISE
+numerically/approximatively has been devised such as Hartree-Fock
+(HF), Density Functional Theory (DFT), Configuration Interaction (CI)
+and Coupled Cluster (CC). While efficient, Hartree-Fock and DFT are
+insufficient if one wants a high degree of accuracy. Configuration
+Interaction and Coupled Cluster methods are hierarchical in the sense
+that one can systematically construct increasly accurate
+approximations. If the CI method is not truncated we have what is
+known as Full Configuration Interaction (FCI). FCI can be seen as
+exact (within some finite space), however it suffers from exponential
+scaling. Truncated CI methods which would achieve polynomial scaling
+are problematic since they are not size-consistent and
+extensive. Truncated CC methods one the other hand are size-consistent
+and extensive and achieves polynomial scaling. Due to this fact
+Coupled Cluster is considered the gold standard of many-body
+techniques if high accuracy is desired.
 
-Similarly, there exists different approximations to the solution of the TDSE with 
-the Multiconfiguration Time-Dependent Hartree-Fock method (MCTDHF) being considered the most accurate. MCTDHF, which is combination of CI and HF generalized to 
-the time domain, suffers from exponential scaling. In a recent article \cite{Kvaal12}, Simen Kvaal proposes a method based on Coupled-Cluster theory, 
-the so-called Orbital Adaptive Time Dependent Coupled Cluster method (OATDCC), which is a hierarchical approximation to the MCTDHF method. The OATDCC inherits 
-size-consisteny and extensivity from the CC method and achieves polynomial scaling. 
+Similarly, there exists different approximations to the solution of
+the TDSE with the Multiconfiguration Time-Dependent Hartree-Fock
+method (MCTDHF) being considered the most accurate. MCTDHF, which is
+combination of CI and HF generalized to the time domain, suffers from
+exponential scaling. In a recent article \cite{Kvaal12}, Simen Kvaal
+proposes a method based on Coupled-Cluster theory, the so-called
+Orbital Adaptive Time Dependent Coupled Cluster method (OATDCC), which
+is a hierarchical approximation to the MCTDHF method. The OATDCC
+inherits size-consisteny and extensivity from the CC method and
+achieves polynomial scaling.
+
 \section{Theory/Formalism}
 \subsection{The model Hamiltonian}
-We shall model the quantum dot systems as a collection of $N$ nonrelativistic electrons of mass $m$ in $d$-dimensional space trapped by a confining potential 
-$v_{\text{conf}}(\mathbf{\hat{r}})$. Here $\hat{\mathbf{r}}$ is the position operator in $d$-dimensional space 
+We shall model the quantum dot systems as a collection of $N$
+nonrelativistic electrons of mass $m$ in $d$-dimensional space trapped
+by a confining potential $v_{\text{conf}}(\mathbf{\hat{r}})$. Here
+$\hat{\mathbf{r}}$ is the position operator in $d$-dimensional space
 \begin{equation}
@@
 
-The electrons interact through the Coulomb interaction $e^2/(4\pi \epsilon R)$, where $e$ is the electron charge, $\epsilon$ the permitivity of the medium and $R$ is the 
-distance between two interacting electrons.
+The electrons interact through the Coulomb interaction $e^2/(4\pi
+\epsilon R)$, where $e$ is the electron charge, $\epsilon$ the
+permitivity of the medium and $R$ is the distance between two
+interacting electrons.
 
-For convenience, we will use atomic untis, chosen such that $\hbar \equiv m \equiv e \equiv 4\pi \epsilon \equiv 1$. Thus, all energies are presented in hartrees per $\hbar$.
+For convenience, we will use atomic untis, chosen such that $\hbar
+\equiv m \equiv e \equiv 4\pi \epsilon \equiv 1$. Thus, all energies
+are presented in hartrees per $\hbar$.
 
-In atomic units, the many-body problem is described by the Hamiltonian on second quantized form
+In atomic units, the many-body problem is described by the Hamiltonian
+on second quantized form
 \begin{align}
@@
 \end{equation}
-while $\hat{u}$ represents the Coulomb interaction. Here $\ket{p}$ is a single-particle basis function which in its position representation is given by
+while $\hat{u}$ represents the Coulomb interaction. Here $\ket{p}$ is
+a single-particle basis function which in its position representation
+is given by
 \begin{equation}
@@
 \end{equation}
-In practice, we have to choose some single-particle basis to compute within. One typical choice is to take the single-particle basis to be the eigenfunctions of $\hat{h}_0$ such that 
+In practice, we have to choose some single-particle basis to compute
+within. One typical choice is to take the single-particle basis to be
+the eigenfunctions of $\hat{h}_0$ such that
 \begin{equation}
@@
 
-The quantities $\braket{p|\hat{h}_0|q}$ and $\braket{pq|\hat{u}|rs}_{\text{AS}}$ are referred to as \textit{matrix elements} and must 
-be computed relative to the chosen single-particle basis. In particular, $\braket{pq|\hat{u}|rs}_{\text{AS}}$ are \textit{antisymmetrized} matrix elements, defined  by 
+The quantities $\braket{p|\hat{h}_0|q}$ and
+$\braket{pq|\hat{u}|rs}_{\text{AS}}$ are referred to as \textit{matrix
+  elements} and must be computed relative to the chosen
+single-particle basis. In particular,
+$\braket{pq|\hat{u}|rs}_{\text{AS}}$ are \textit{antisymmetrized}
+matrix elements, defined by
 \begin{equation}
@@
 \end{equation}
-In the rest of this paper we will drop the subscript AS for the antisymmetrized matrix elements.
+In the rest of this paper we will drop the subscript AS for the
+antisymmetrized matrix elements.
 
-In order to generate dynamics the electrons are coupled to a time dependent monochromatic laser field, with frequency $\omega$ and amplitude $\mathcal{E}_0$, polarized along the $x$-axis \cite{Zanghellini04}
+In order to generate dynamics the electrons are coupled to a time
+dependent monochromatic laser field, with frequency $\omega$ and
+amplitude $\mathcal{E}_0$, polarized along the $x$-axis
+\cite{Zanghellini04}
 \begin{equation}
@@
 \end{align}
-It is important to note that this implies that we have to update the matrix elements $\braket{p|\hat{h}_t|q}$ during the simulation. However, in this work we consider only methods where 
-the single-particle basis is held fixed in time, hence the two-body matrix elements can be computed $\braket{pq|\hat{u}|rs}$ once and for all.
+It is important to note that this implies that we have to update the
+matrix elements $\braket{p|\hat{h}_t|q}$ during the
+simulation. However, in this work we consider only methods where the
+single-particle basis is held fixed in time, hence the two-body matrix
+elements can be computed $\braket{pq|\hat{u}|rs}$ once and for all.
 
-We will handle the time dependent problem in two steps.
-First we prepare a system in its groundstate by solving the time independent Schrödinger equation (TISE) for the field-free Hamiltonian \eqref{FieldFreeHamiltonian}.
-Then we apply the time dependent laser field and compute the time evolution from the time dependent Schrödinger equation (TDSE). 
-Before we discuss the metods for solving the TISE and TDSE respectively we will adress the choice of single-particle basis.
+We will handle the time dependent problem in two steps. First we
+prepare a system in its groundstate by solving the time independent
+Schrödinger equation (TISE) for the field-free Hamiltonian
+\eqref{FieldFreeHamiltonian}.  Then we apply the time dependent laser
+field and compute the time evolution from the time dependent
+Schrödinger equation (TDSE).  Before we discuss the metods for solving
+the TISE and TDSE respectively we will adress the choice of
+single-particle basis.
 
 \subsection{Choice of single-particle basis}
-The choice of single-particle basis will in general depend on the problem. In quantum chemistry it is customary to expand the eigenfunctions, $\{ \psi_k \}$, 
-of $\hat{h}_0$ in (non-orthogonal) Gaussian basis functions
-$g_p(\mathbf{r})$
+The choice of single-particle basis will in general depend on the
+problem. In quantum chemistry it is customary to expand the
+eigenfunctions, $\{ \psi_k \}$, of $\hat{h}_0$ in (non-orthogonal)
+Gaussian basis functions $g_p(\mathbf{r})$
 \begin{equation}
@@
 \end{equation}
-resulting in a generalized eigenvalue problem which must be solved for the eigenfunctions. Ideally we would let $L \rightarrow \infty$ but must be taken as finite in practice. 
-The choice of Gaussian basis functions is popular due to the fact that matrix elemtents can be evaluated efficiently \cite{Helgaker00book}.
+resulting in a generalized eigenvalue problem which must be solved for
+the eigenfunctions. Ideally we would let $L \rightarrow \infty$ but
+must be taken as finite in practice.  The choice of Gaussian basis
+functions is popular due to the fact that matrix elemtents can be
+evaluated efficiently \cite{Helgaker00book}.
 
-Another popular choice, which is the one chosen in this work, is to use the eigenfunctions of the Harmonic oscillator (HO) 
+Another popular choice, which is the one chosen in this work, is to
+use the eigenfunctions of the Harmonic oscillator (HO)
 \begin{equation}
@@
 \end{equation}
-as a single-particle basis. The solutions to the HO-problem is well known and the eigenfunctions have closed form expressions.
+as a single-particle basis. The solutions to the HO-problem is well
+known and the eigenfunctions have closed form expressions.
 
-Of particular importance for our purposes is the fact that the interaction integrals
+Of particular importance for our purposes is the fact that the
+interaction integrals
 \begin{equation}
@@
 \end{equation}
-can be solved analytically in two- and three-dimensions, using HO-basis functions in polar and spherical coordinates respectively \cite{Anisimovas98, Vorrath2003}. If the confining potential 
-is chosen to be the Harmonic oscillator potential we are essentially done and can move on to the many-body treatment of the problem.
+can be solved analytically in two- and three-dimensions, using
+HO-basis functions in polar and spherical coordinates respectively
+\cite{Anisimovas98, Vorrath2003}. If the confining potential is chosen
+to be the Harmonic oscillator potential we are essentially done and
+can move on to the many-body treatment of the problem.
 
-Additionally we may approximate the eigenfunctions of more general confining potentials by expanding the solutions in the HO-functions.
+Additionally we may approximate the eigenfunctions of more general
+confining potentials by expanding the solutions in the HO-functions.
 
@@
 \end{equation}
-and insert into Eq.\eqref{EigenfuncEq}. Left-projection with $\bra{\phi_q}$ results in the following expression
+and insert into Eq.\eqref{EigenfuncEq}. Left-projection with
+$\bra{\phi_q}$ results in the following expression
 \begin{equation}
@@
 \end{equation}
-where $H_{qp} \equiv \braket{\phi_q|\hat{h}_0|\phi_p}$, $\mathcal{E}$ is a diagonal matrix containing the eigenvalues on its diagonal, and $C$ is matrix conatining 
-the expansions coefficients for basis function $k$ in its columns. The eigenproblem must be solved for the expansions coefficients, resulting in a approximation of the interaction integrals 
-in terms of the known intercation integrals $\braket{\phi_\alpha \phi_\beta|\hat{u}|\phi_\gamma \phi_\delta}$ given by 
+where $H_{qp} \equiv \braket{\phi_q|\hat{h}_0|\phi_p}$, $\mathcal{E}$
+is a diagonal matrix containing the eigenvalues on its diagonal, and
+$C$ is matrix conatining the expansions coefficients for basis
+function $k$ in its columns. The eigenproblem must be solved for the
+expansions coefficients, resulting in a approximation of the
+interaction integrals in terms of the known intercation integrals
+$\braket{\phi_\alpha \phi_\beta|\hat{u}|\phi_\gamma \phi_\delta}$
+given by
 \begin{equation}
@@
 
-The quality of such an approximation will naturally depend on how close the confining potential is to the Harmonic oscillator potential. Nevertheless it constitutes a convenient starting point for 
+The quality of such an approximation will naturally depend on how
+close the confining potential is to the Harmonic oscillator
+potential. Nevertheless it constitutes a convenient starting point for
 investigations of more general potentials.
@@
 \subsubsection{Time independent Configuration interaction}
-The conceptually simplest method for solving the many-problem is probably that of Configuration interaction (CI).
+The conceptually simplest method for solving the many-problem is
+probably that of Configuration interaction (CI).
 
@@
 \end{equation}
-where $\hat{H}$ is the field-free Hamiltonian given by Eq. \eqref{FieldFreeHamiltonian}, $\ket{\Psi_k}$ is the many-body eigenfunction and $E_k$ its corresponding eigenvalue. It is customary 
-to use a (finite in practice) basis of orthogonal Slater determinants (SD), $\{ \Phi_p \}_{p=1}^{N_{\text{sd}}}$ where $N_{\text{sd}}$ is the number of determinants, built from the single-particle basis functions described in the previous section.
+where $\hat{H}$ is the field-free Hamiltonian given by
+Eq. \eqref{FieldFreeHamiltonian}, $\ket{\Psi_k}$ is the many-body
+eigenfunction and $E_k$ its corresponding eigenvalue. It is customary
+to use a (finite in practice) basis of orthogonal Slater determinants
+(SD), $\{ \Phi_p \}_{p=1}^{N_{\text{sd}}}$ where $N_{\text{sd}}$ is
+the number of determinants, built from the single-particle basis
+functions described in the previous section.
 
@@
 \end{equation}
-where $H_{qp} \equiv \braket{\Phi_q|\hat{H}|\Phi_p}$, $E$ is a diagonal matrix with the eigenvalues along its diagonal and $C$ is a matrix containing the expansion coefficients of the 
-$k$-th eigenfunction in column $k$. Thus if we can compute the matrix elements $H_{qp}$ the TISE can be solved by numerical diagonalization.
+where $H_{qp} \equiv \braket{\Phi_q|\hat{H}|\Phi_p}$, $E$ is a
+diagonal matrix with the eigenvalues along its diagonal and $C$ is a
+matrix containing the expansion coefficients of the $k$-th
+eigenfunction in column $k$. Thus if we can compute the matrix
+elements $H_{qp}$ the TISE can be solved by numerical diagonalization.
 
-The evaluation of $H_{qp}$ is simplified by what is known as the Slater-Condon rules. If $\ket{\Phi}$ is any SD, 
-the only non-zero contributions to $H$ are given by
+The evaluation of $H_{qp}$ is simplified by what is known as the
+Slater-Condon rules. If $\ket{\Phi}$ is any SD, the only non-zero
+contributions to $H$ are given by
 \begin{align}
@@
 \end{align}
-In other words, we only have to consider the overlap between determinants that differ by at most two occupied single-particle functions. 
-Here we use the convention that $i,j,k,l = 1,...,N$, with $N$ being the number of electrons and $a,b,c,d = N+1,...$ and $p,q,r,s = 1,...,L$. Thus, 
-we see that if the matrix elements $\braket{p|\hat{h}_0|q}$ and $\braket{pq|\hat{u}|rs}$ are known we can set up the Hamilton matrix and diagonalize it, obtaining the expansion coefficients $C$.
+In other words, we only have to consider the overlap between
+determinants that differ by at most two occupied single-particle
+functions.  Here we use the convention that $i,j,k,l = 1,...,N$, with
+$N$ being the number of electrons and $a,b,c,d = N+1,...$ and $p,q,r,s
+= 1,...,L$. Thus, we see that if the matrix elements
+$\braket{p|\hat{h}_0|q}$ and $\braket{pq|\hat{u}|rs}$ are known we can
+set up the Hamilton matrix and diagonalize it, obtaining the expansion
+coefficients $C$.
 
-Ideally we would like $N_{\text{sd}} \rightarrow \infty$ which is impossible in practice. If we choose a finite number, $L$, of single-particle basis functions and use all the linearly independent 
-Slater determinants we can build from this basis we get what is known as the Full Configuration Interaction (FCI). However, the problem with FCI is that the number of linearly independent 
-determinants are given by 
+Ideally we would like $N_{\text{sd}} \rightarrow \infty$ which is
+impossible in practice. If we choose a finite number, $L$, of
+single-particle basis functions and use all the linearly independent
+Slater determinants we can build from this basis we get what is known
+as the Full Configuration Interaction (FCI). However, the problem with
+FCI is that the number of linearly independent determinants are given
+by
 \begin{equation}
@@
 \end{equation}
-with $N$ being the number of electrons. This means that the numbers of SD's becomes huge very fast, making FCI practical/feasible only for the smallest systems. On the other hand, FCI is exact within 
-a given computational space, thus it is an invaluable tool in aiding us determining the validity of more sophisticated methods.
-Furthermore, CI is variational in the sense that it guarantees an upper bound
-on the groundstate energy.
+with $N$ being the number of electrons. This means that the numbers of
+SD's becomes huge very fast, making FCI practical/feasible only for
+the smallest systems. On the other hand, FCI is exact within a given
+computational space, thus it is an invaluable tool in aiding us
+determining the validity of more sophisticated methods. Furthermore,
+CI is variational in the sense that it guarantees an upper bound on
+the groundstate energy.
 
-One way to adress the problems with FCI is to consider a \textit{reference} determinant $\ket{\Phi}$, typically taken to be the Hartree-Fock state, and consider excitations upon this state. That is, 
+One way to adress the problems with FCI is to consider a
+\textit{reference} determinant $\ket{\Phi}$, typically taken to be the
+Hartree-Fock state, and consider excitations upon this state. That is,
 we make the ansatz
@@
 \end{equation}
-where $\ket{\Phi_i^a}$ is a singly-excited determinant, $\ket{\Phi_{ij}^{ab}}$ doubly-excited etc. and truncate at some excitation level. 
-This results in a hierarchy of approximations known as CIS (only single-excitations), CID (only double-excitations), CISD (singles and doubles and so on. In this way one obtains polynomial 
-scaling of the number of determinants. The problem with this approach is that the truncated CI-wavefunction is not size-consistent and extensive \cite{Helgaker00book}. It turns out 
-that this problem is resolved by the Coupled Cluster method, which makes it the de facto standard of many-body methods today.
+where $\ket{\Phi_i^a}$ is a singly-excited determinant,
+$\ket{\Phi_{ij}^{ab}}$ doubly-excited etc. and truncate at some
+excitation level.  This results in a hierarchy of approximations known
+as CIS (only single-excitations), CID (only double-excitations), CISD
+(singles and doubles and so on. In this way one obtains polynomial
+scaling of the number of determinants. The problem with this approach
+is that the truncated CI-wavefunction is not size-consistent and
+extensive \cite{Helgaker00book}. It turns out that this problem is
+resolved by the Coupled Cluster method, which makes it the de facto
+standard of many-body methods today.
 
 \subsubsection{Time dependent Configuration interaction}
-Another advantage with CI is that it is straightforward to generalize to time dependent problems. Recall that the time evolution of an  
-arbitrary state, $\ket{\Psi(t)}$ with initial condition $\ket{\Psi(t_0)}$, is given by the time dependent Schrödinger equation, 
+Another advantage with CI is that it is straightforward to generalize
+to time dependent problems. Recall that the time evolution of an
+arbitrary state, $\ket{\Psi(t)}$ with initial condition
+$\ket{\Psi(t_0)}$, is given by the time dependent Schrödinger
+equation,
 \begin{equation}
@@
 \end{equation}
-We will take the initial condition to be some (normalized) linear combination of the CI eigenfunctions. In particular, we are interested in the 
-time evolution of the groundstate.
+We will take the initial condition to be some (normalized) linear
+combination of the CI eigenfunctions. In particular, we are interested
+in the time evolution of the groundstate.
 
-The time-dependent Configuration Interaction (TDCI) method is obtained by placing a time-dependence on 
-the expansion coefficients in the CI-wavefunction while keeping the single-particle basis fixed. Analogous to what we did in the previous section, we 
-expand $\ket{\Psi(t)}$ in a given orthonormal Slater determinant basis $\{ \Phi_p\}_{p=1}^{N_{\text{sd}}}$
+The time-dependent Configuration Interaction (TDCI) method is obtained
+by placing a time-dependence on the expansion coefficients in the
+CI-wavefunction while keeping the single-particle basis
+fixed. Analogous to what we did in the previous section, we expand
+$\ket{\Psi(t)}$ in a given orthonormal Slater determinant basis $\{
+\Phi_p\}_{p=1}^{N_{\text{sd}}}$
 \begin{equation}
@@
 \end{equation}
-Left-projecting the above equation with $\bra{\Phi_q}$ and using the orthonormality of the Slater determinants, 
-we are left with a differential equation for each expansion coefficient $A_J(t)$
-which can be written as the matrix-vector equation
+Left-projecting the above equation with $\bra{\Phi_q}$ and using the
+orthonormality of the Slater determinants, we are left with a
+differential equation for each expansion coefficient $A_J(t)$ which
+can be written as the matrix-vector equation
 \begin{equation}
@@
 \end{equation}
-and $A(t) = [A_0(t), A_1(t), \cdots]$ is the vector containing the expansion coefficients. We refer to this as the 
-time dependent Configuration Interaction (TDCI) method.
+and $A(t) = [A_0(t), A_1(t), \cdots]$ is the vector containing the
+expansion coefficients. We refer to this as the time dependent
+Configuration Interaction (TDCI) method.
 
-Completely analogous to the time independent case we have a hierarchy of TDCI methods defined by the Slater determinant basis. Thus we have 
-a series of possible approximations ranging from TD-CIS to TD-FCI.
-If we did not have to truncate the Slater determinant basis, TDFCI would be an exact solution to the TDSE.
-However, we must in practice truncate the basis and TDFCI suffers from the fact that the size of FCI-space grows exponentially. 
-In order to obtain a good approximation to the exact solution we would need a huge fixed basis\cite{Kvaal12}.
+Completely analogous to the time independent case we have a hierarchy
+of TDCI methods defined by the Slater determinant basis. Thus we have
+a series of possible approximations ranging from TD-CIS to TD-FCI.  If
+we did not have to truncate the Slater determinant basis, TDFCI would
+be an exact solution to the TDSE.  However, we must in practice
+truncate the basis and TDFCI suffers from the fact that the size of
+FCI-space grows exponentially.  In order to obtain a good
+approximation to the exact solution we would need a huge fixed
+basis\cite{Kvaal12}.
 
-On the other hand it is quite straightforward to implement, especially when a program for
-computing the CI-groundstate already has been written. 
-Again, the fact that the CI and CC space are equivalent for $N=2$ makes it worthwhile to write a TDCI program, 
-which we can use to compare with the time dependent Coupled Cluster method which we discuss in the next chapter. 
+On the other hand it is quite straightforward to implement, especially
+when a program for computing the CI-groundstate already has been
+written.  Again, the fact that the CI and CC space are equivalent for
+$N=2$ makes it worthwhile to write a TDCI program, which we can use to
+compare with the time dependent Coupled Cluster method which we
+discuss in the next chapter.
 \begin{comment}
 \subsection{Coupled cluster theory}
-In this section we review Coupled Cluster (CC) theory. 
-We will look at the "classical" way of obtaining the CC equations. The presentation borrows heavily from the 
-excellent review by Crawford and Schaefer\cite{Crawford2000}.
+In this section we review Coupled Cluster (CC) theory.  We will look
+at the "classical" way of obtaining the CC equations. The presentation
+borrows heavily from the excellent review by Crawford and
+Schaefer\cite{Crawford2000}.
 
-Furthermore we look at a different approach which in addition to the usual CC equations also 
-gives a method to solve the time dependent Schrödinger equation, the so-called Orbital Adaptive Coupled Cluster (OATDCC) method\cite{Kvaal12}.
+Furthermore we look at a different approach which in addition to the
+usual CC equations also gives a method to solve the time dependent
+Schrödinger equation, the so-called Orbital Adaptive Coupled Cluster
+(OATDCC) method\cite{Kvaal12}.
 
@@
 \end{equation}
-is used to approximate the exact solution. Here $\ket{\Phi}$ is a reference Slater determinant 
-and $\hat{T} = \sum_{i}^n \hat{T}_i$ is the sum over n-orbital cluster operators.
+is used to approximate the exact solution. Here $\ket{\Phi}$ is a
+reference Slater determinant and $\hat{T} = \sum_{i}^n \hat{T}_i$ is
+the sum over n-orbital cluster operators.
 
@@
 \subsection{The Coupled Cluster Equations}
-We now proceed by inserting expression \eqref{ExponantialAnsatzCC} into the time-independent Schrödinger equation
+We now proceed by inserting expression \eqref{ExponantialAnsatzCC}
+into the time-independent Schrödinger equation
 \begin{equation}
@@
 \end{align*}
-where intermediate normalization, $\braket{\Phi|\Psi_{CC}} = 1$, is assumed. Furthermore one can obtain an expression for the amplitude $t^{ab...}_{ij...}$ 
-upon left-projection with $\bra{\Phi^{ab...}_{ij...}}$,
+where intermediate normalization, $\braket{\Phi|\Psi_{CC}} = 1$, is
+assumed. Furthermore one can obtain an expression for the amplitude
+$t^{ab...}_{ij...}$ upon left-projection with
+$\bra{\Phi^{ab...}_{ij...}}$,
 \begin{equation}
@@
 \end{equation}
-Notice that this equations are non-linear (due to the presence of $e^{\hat{T}}$) and depend on the energy equation. This is impractical and is resolved by performing 
-a similarity transformation of the Hamiltonian.
+Notice that this equations are non-linear (due to the presence of
+$e^{\hat{T}}$) and depend on the energy equation. This is impractical
+and is resolved by performing a similarity transformation of the
+Hamiltonian.
 
-Multiply Eq. \eqref{ExponantialAnsatzCCInsertedInSE} with $e^{-\hat{T}}$ and left-project with $\bra{\Phi}$ in order to obtain
+Multiply Eq. \eqref{ExponantialAnsatzCCInsertedInSE} with
+$e^{-\hat{T}}$ and left-project with $\bra{\Phi}$ in order to obtain
 \begin{equation}
@@
 \end{equation}
-Equations \eqref{SimTransEnergyEquation} and \eqref{SimTransAmplitudeEquation} are commonly referred to as the Coupled Cluster equations. Notice now that the equation 
-for the amplitudes does not depend on the energy. The amplitude equation is non-linear and has to be solved iteratively. We say that we have a self-consistent solution to CC-equations
-if the energy converges to some pre-defined precision.
+Equations \eqref{SimTransEnergyEquation} and
+\eqref{SimTransAmplitudeEquation} are commonly referred to as the
+Coupled Cluster equations. Notice now that the equation for the
+amplitudes does not depend on the energy. The amplitude equation is
+non-linear and has to be solved iteratively. We say that we have a
+self-consistent solution to CC-equations if the energy converges to
+some pre-defined precision.
 
-In order to implement the Coupled Cluster method we need algebraic expressions for the CC-equations. 
+In order to implement the Coupled Cluster method we need algebraic
+expressions for the CC-equations.
 \begin{theorem}[Baker-Campbell-Hausdorff expansion]
@@
 \end{theorem}
-Under the assumption that $\hat{H}$ contains at most a two-body operator the BCH-expansion truncates after the first 
-five terms on the right-hand side of eq. \eqref{BCH}\cite{Crawfordtruncates}. 
-This fact greatly simplifies the evaluation of the left-hand sides of the equations \eqref{SimTransEnergyEquation} and \eqref{SimTransAmplitudeEquation}.
+Under the assumption that $\hat{H}$ contains at most a two-body
+operator the BCH-expansion truncates after the first five terms on the
+right-hand side of eq. \eqref{BCH}\cite{Crawfordtruncates}.  This fact
+greatly simplifies the evaluation of the left-hand sides of the
+equations \eqref{SimTransEnergyEquation} and
+\eqref{SimTransAmplitudeEquation}.
 
-These expressions can be evaluated by hand using the second quantization formalism as demonstrated in\cite{Crawford2000} or by diagrammatic
-techniques\cite{ShavittBartlett}.
-Alternatively one can use the second quantization toolbox in the Python library Sympy\footnote{\href{http://docs.sympy.org/latest/modules/physics/secondquant.html}{http://docs.sympy.org/latest/modules/physics/secondquant.html}}
+These expressions can be evaluated by hand using the second
+quantization formalism as demonstrated in\cite{Crawford2000} or by
+diagrammatic techniques\cite{ShavittBartlett}. Alternatively one can
+use the second quantization toolbox in the Python library
+Sympy\footnote{\href{http://docs.sympy.org/latest/modules/physics/secondquant.html}{http://docs.sympy.org/latest/modules/physics/secondquant.html}}
 to compute these expressions.
@@
 \subsection{A variational CC theory?}
-One drawback with the CC-method is that it does not satisfy the variational principle \eqref{variationalTheorem}. Consider the operator
-$e^{-\hat{T}}\hat{H}e^{T}$ which appears in the similarity transformed energy equation \eqref{SimTransEnergyEquation}. This operator is not Hermitian. In order to see this compute $\hat{T}_1^\dagger$,
+One drawback with the CC-method is that it does not satisfy the
+variational principle \eqref{variationalTheorem}. Consider the
+operator $e^{-\hat{T}}\hat{H}e^{T}$ which appears in the similarity
+transformed energy equation \eqref{SimTransEnergyEquation}. This
+operator is not Hermitian. In order to see this compute
+$\hat{T}_1^\dagger$,
 \begin{equation}
@@
 \end{equation}
-Hermiticity of the operator was a necessary condition for the variational theorem, thus Eq. \eqref{SimTransEnergyEquation} is not variational. 
-However, if $\hat{T}$ is not truncated the spectrum of $e^{-\hat{T}}\hat{H}e^{T}$
-is identical to the original Hermitian operator $\hat{H}$, which justifies its use in quantum mechanical models\cite{CrawfordNonVar}.
+Hermiticity of the operator was a necessary condition for the
+variational theorem, thus Eq. \eqref{SimTransEnergyEquation} is not
+variational.  However, if $\hat{T}$ is not truncated the spectrum of
+$e^{-\hat{T}}\hat{H}e^{T}$ is identical to the original Hermitian
+operator $\hat{H}$, which justifies its use in quantum mechanical
+models\cite{CrawfordNonVar}.
 
-There has been attempts to construct a variational solution by deriving the amplitude equations by minimizing the functional,
+There has been attempts to construct a variational solution by
+deriving the amplitude equations by minimizing the functional,
 \begin{equation}
@@
 \subsection{The Orbital Adaptive Time Dependent Coupled Cluster Method}
-In the article\cite{Kvaal12} Simen Kvaal demonstrates that it is possible to 
-derive equations of motion for the CC-ansatz by computing critical points of a bivariational expectation value functional. 
-In his approach both the orbitals and CC-amplitudes are allowed to vary and the method can be used to solve the time-dependent Schrödinger equation.  
-The method is called the Orbital Adaptive Time-Dependent Coupled Cluster (OATDCC) and is a hierarchical approximation to MCTDHF. Furthermore, if the orbitals are 
-held fixed in time the method gives a approximation to the Time-Dependent CI method which we refer to as the Time-Dependent Coupled Cluster (TDCC) method. 
-While TDFCI and MCTDHF suffers from exponential scaling, OATDCC and TDCC achieves polynomial scaling.
+In the article\cite{Kvaal12} Simen Kvaal demonstrates that it is
+possible to derive equations of motion for the CC-ansatz by computing
+critical points of a bivariational expectation value functional.  In
+his approach both the orbitals and CC-amplitudes are allowed to vary
+and the method can be used to solve the time-dependent Schrödinger
+equation.  The method is called the Orbital Adaptive Time-Dependent
+Coupled Cluster (OATDCC) and is a hierarchical approximation to
+MCTDHF. Furthermore, if the orbitals are held fixed in time the method
+gives a approximation to the Time-Dependent CI method which we refer
+to as the Time-Dependent Coupled Cluster (TDCC) method.  While TDFCI
+and MCTDHF suffers from exponential scaling, OATDCC and TDCC achieves
+polynomial scaling.
 
-The main goal of this article is to implement the TDCC method for a possibly time-dependent Hamiltonian. 
-As such we will in the following summarize the key ingredients needed to achieve this. 
-For mathematical details the interested reader is referred to the article by Kvaal.
+The main goal of this article is to implement the TDCC method for a
+possibly time-dependent Hamiltonian.  As such we will in the following
+summarize the key ingredients needed to achieve this.  For
+mathematical details the interested reader is referred to the article
+by Kvaal.
 
 \subsubsection{The Bivariational Principle}
-As we have established, the Coupled Cluster method is not variational in the usual sense since the similarity transformed Hamiltonian,
-$$ \bar{H} = e^{-\hat{T}}\hat{H}e^{\hat{T}} $$
-is not Hermitian. Now following Kvaal, let $A$ be an operator (possibly non-Hermitian) over Hilbert space $\mathcal{H}$ and consider the expectation value functional
+As we have established, the Coupled Cluster method is not variational
+in the usual sense since the similarity transformed Hamiltonian,
+$$ \bar{H} = e^{-\hat{T}}\hat{H}e^{\hat{T}} $$ is not Hermitian. Now
+following Kvaal, let $A$ be an operator (possibly non-Hermitian) over
+Hilbert space $\mathcal{H}$ and consider the expectation value
+functional
 \begin{equation}
@@
 \end{equation}
-This is referred to as the the bivariational expectation value functional. In contrast to the usual 
-variational principle $\bra{\Psi^{\prime}}$ and $\ket{\Psi}$ are treated as independent parameters. The conditions for 
-$\delta \mathcal{E}_A = 0$, for all independent variations of $\bra{\Psi^{\prime}}$ and $\ket{\Psi}$ is 
+This is referred to as the the bivariational expectation value
+functional. In contrast to the usual variational principle
+$\bra{\Psi^{\prime}}$ and $\ket{\Psi}$ are treated as independent
+parameters. The conditions for $\delta \mathcal{E}_A = 0$, for all
+independent variations of $\bra{\Psi^{\prime}}$ and $\ket{\Psi}$ is
 \begin{equation}
@@
 \end{equation}
-with 
-$$a = \mathcal{E}_A[\bra{\Psi^{\prime}},\ket{\Psi}] $$
+with $$a = \mathcal{E}_A[\bra{\Psi^{\prime}},\ket{\Psi}] $$
 being the value of $\mathcal{E}_A$ at the critical point.
 
-Equations of motion can be derived from so-called time-dependent variational principles. One such variational is set up 
-using the Lagrange formulation\cite{BECK20001}, 
+Equations of motion can be derived from so-called time-dependent
+variational principles. One such variational is set up using the
+Lagrange formulation\cite{BECK20001},
 \begin{equation}
@@
 \end{equation}
-is referred to as the action functional and computing 
-equations of motion from $\delta \mathcal{S} = 0$ is known as the prinicipal of least action.
+is referred to as the action functional and computing equations of
+motion from $\delta \mathcal{S} = 0$ is known as the prinicipal of
+least action.
 
-By considering a bivariational generalization of the action functional \eqref{ActionFunctional1},
+By considering a bivariational generalization of the action functional
+\eqref{ActionFunctional1},
 \begin{align}
@@
 \end{align}
-Kvaal derives equations of motions generalizing the Coupled Cluster method to the time domain. 
-Here, one should note the appearance of the bivariational generalization of the energy expectation value,
+Kvaal derives equations of motions generalizing the Coupled Cluster
+method to the time domain.  Here, one should note the appearance of
+the bivariational generalization of the energy expectation value,
 \begin{equation}
@@
 \subsubsection{The Coupled Cluster Ansatz}
-The fundamental idea in\cite{Kvaal12} is to introduce different exponential parametrizations $\ket{\Psi}$, $\bra{\Psi^\prime}$ and 
-compute $\delta \mathcal{S} = 0$, with $\mathcal{S}$ given by \eqref{ActionFunctional}, which ultimately results in time dependent Coupled Cluster equations.
+The fundamental idea in\cite{Kvaal12} is to introduce different
+exponential parametrizations $\ket{\Psi}$, $\bra{\Psi^\prime}$ and
+compute $\delta \mathcal{S} = 0$, with $\mathcal{S}$ given by
+\eqref{ActionFunctional}, which ultimately results in time dependent
+Coupled Cluster equations.
 
@@
 \begin{equation}
- \hat{T} = \sum_{ia} \tau_i^a c_a^\dagger \tilde{c}_i + \frac{1}{2!^2} \sum_{ijab} \tau_{ij}^{ab} c_a^\dagger \tilde{c}_i c_b^\dagger \tilde{c}_j + \cdots
+ \hat{T} = \sum_{ia} \tau_i^a c_a^\dagger \tilde{c}_i + \frac{1}{2!^2}
+ \sum_{ijab} \tau_{ij}^{ab} c_a^\dagger \tilde{c}_i c_b^\dagger
+ \tilde{c}_j + \cdots
 \end{equation}
@@
 
-An important special case is the case $N=2$. In that case the CC expansion equals the CI expansion. This is invaluable information when we want to implement 
-the CC method since we can then compare it with the corresponding CI method to verify the implementation. To see this let $N=2$ and let 
-$\ket{\Phi}$ be a reference determinant as usual. Observe that $\hat{A}_3 \ket{\Phi} = 0 $ and $\hat{T}_3 = 0$. Thus, 
-$\hat{A} = \hat{A}_1 + \hat{A}_2$ and $\hat{T} = \hat{T}_1 + \hat{T}_2$ will give all non-zero contributions. Additionaly all products of cluster operators 
-where the product has rank larger than $2$, acting on the reference will give a zero result.
-Inserting 
- $\hat{T}_1 = \hat{A}_1$ and $\hat{T}_2 = \hat{A}_2 - \frac{1}{2}\hat{A}_1^2$ into the CC expansionincluding only terms of at most rank $2$ we obtain,  
+An important special case is the case $N=2$. In that case the CC
+expansion equals the CI expansion. This is invaluable information when
+we want to implement the CC method since we can then compare it with
+the corresponding CI method to verify the implementation. To see this
+let $N=2$ and let $\ket{\Phi}$ be a reference determinant as
+usual. Observe that $\hat{A}_3 \ket{\Phi} = 0 $ and $\hat{T}_3 =
+0$. Thus, $\hat{A} = \hat{A}_1 + \hat{A}_2$ and $\hat{T} = \hat{T}_1 +
+\hat{T}_2$ will give all non-zero contributions. Additionaly all
+products of cluster operators where the product has rank larger than
+$2$, acting on the reference will give a zero result.  Inserting
+$\hat{T}_1 = \hat{A}_1$ and $\hat{T}_2 = \hat{A}_2 -
+\frac{1}{2}\hat{A}_1^2$ into the CC expansionincluding only terms of
+at most rank $2$ we obtain,
  \begin{align*}
-  e^{\hat{T}_1 + \hat{T}_2} \ket{\Phi} &= \left( 1 + \underbrace{\hat{T}_1}_{\text{rank} 1} + \underbrace{\left( \hat{T}_2 + \frac{1}{2}\hat{T}_1^2 \right)}_{\text{rank} 2} \right) \ket{\Phi} \\
-  &= \left(1+ \hat{A}_1 + \hat{A}_2 - \frac{1}{2}\hat{A}_1^2 + \frac{1}{2} \hat{A}_1^2 \right) \ket{\Phi} \\
-  &= \left(1+ \hat{A}_1 + \hat{A}_2 \right) \ket{\Phi}
+  e^{\hat{T}_1 + \hat{T}_2} \ket{\Phi} &= \left( 1 +
+  \underbrace{\hat{T}_1}_{\text{rank} 1} + \underbrace{\left(
+    \hat{T}_2 + \frac{1}{2}\hat{T}_1^2 \right)}_{\text{rank} 2}
+  \right) \ket{\Phi} \\ &= \left(1+ \hat{A}_1 + \hat{A}_2 -
+  \frac{1}{2}\hat{A}_1^2 + \frac{1}{2} \hat{A}_1^2 \right) \ket{\Phi}
+  \\ &= \left(1+ \hat{A}_1 + \hat{A}_2 \right) \ket{\Phi}
  \end{align*}
-which indeed is the CI expansion. For $N=2$, $\hat{A} = \hat{A}_1 + \hat{A}_2$ corresponds to a full configuration interaction treatment, while 
-$\hat{T} = \hat{T}_1 + \hat{T}_2$ CCSD, but in this case CCSD is actually equal to FCI. Morover, if we consider only $\hat{T} = \hat{T}_2$ (CCD), we get 
+which indeed is the CI expansion. For $N=2$, $\hat{A} = \hat{A}_1 +
+\hat{A}_2$ corresponds to a full configuration interaction treatment,
+while $\hat{T} = \hat{T}_1 + \hat{T}_2$ CCSD, but in this case CCSD is
+actually equal to FCI. Morover, if we consider only $\hat{T} =
+\hat{T}_2$ (CCD), we get
 \begin{align*}
- \ket{\Psi}_{\text{CCD}} = e^{\hat{T}_2} \ket{\Phi} &= \left(1+ \hat{T}_2 \right) \ket{\Phi} \\
- &= \left( 1 + \hat{A}_2 - \frac{1}{2}\hat{A}_1^2 + \frac{1}{2} \hat{A}_1^2 \right) \ket{\Phi} \\
- &= \left(1 + \hat{A}_2 \right) \ket{\Phi} = \ket{\Psi}_{\text{CID}}.
+ \ket{\Psi}_{\text{CCD}} = e^{\hat{T}_2} \ket{\Phi} &= \left(1+
+ \hat{T}_2 \right) \ket{\Phi} \\ &= \left( 1 + \hat{A}_2 -
+ \frac{1}{2}\hat{A}_1^2 + \frac{1}{2} \hat{A}_1^2 \right) \ket{\Phi}
+ \\ &= \left(1 + \hat{A}_2 \right) \ket{\Phi} =
+ \ket{\Psi}_{\text{CID}}.
 \end{align*}
-The importance of these two results is that one can verify a CCD implementation by comparing with a CID implementation and a CCSD implementation can be 
-verified by comparing with a FCI/CISD implementation.
+The importance of these two results is that one can verify a CCD
+implementation by comparing with a CID implementation and a CCSD
+implementation can be verified by comparing with a FCI/CISD
+implementation.
 
@@
 \end{equation}
-where $ u^{pr}_{qs} = \braket{\varphi_p \varphi_r|\hat{H}^{(2)}|\varphi_q \varphi_s}_{AS}$.
-Evaluation of the right hand side of equation \eqref{ExpecTamplitude} gives,
+where $ u^{pr}_{qs} = \braket{\varphi_p
+  \varphi_r|\hat{H}^{(2)}|\varphi_q \varphi_s}_{AS}$.  Evaluation of
+the right hand side of equation \eqref{ExpecTamplitude} gives,
 \begin{equation}
- \frac{\partial}{\partial \lambda_{ij}^{ab}}E_{H^{(1)}} = h^{k}_{i} \tau^{ab}_{jk} P(ij) + h^{a}_{c} \tau^{bc}_{ji} P(ab) \label{AlgebraicTampH1},
+ \frac{\partial}{\partial \lambda_{ij}^{ab}}E_{H^{(1)}} = h^{k}_{i}
+ \tau^{ab}_{jk} P(ij) + h^{a}_{c} \tau^{bc}_{ji}
+ P(ab) \label{AlgebraicTampH1},
 \end{equation}
 \begin{align}
- \frac{\partial}{\partial \lambda_{ab}^{ij}}E_{H^{(2)}} = &-\tau^{ab}_{ik} u^{kl}_{jl} P(ij) + \tau^{ab}_{il} \chi_j^l P(ij) \nonumber \\
- &+\tau^{ab}_{kl} \chi^{kl}_{ij} +\tau^{ac}_{ij}\chi_c^b P(ab) \nonumber \\
- &+\tau^{ac}_{ik} \chi_{cj}^{kb}P(ab)P(ij) -\frac{1}{2}\tau^{dc}_{ji} u^{ab}_{dc} \nonumber \\ 
- &+u^{ab}_{ij} \label{AlgebraicTampH2},
+ \frac{\partial}{\partial \lambda_{ab}^{ij}}E_{H^{(2)}} =
+ &-\tau^{ab}_{ik} u^{kl}_{jl} P(ij) + \tau^{ab}_{il} \chi_j^l P(ij)
+ \nonumber \\ &+\tau^{ab}_{kl} \chi^{kl}_{ij} +\tau^{ac}_{ij}\chi_c^b
+ P(ab) \nonumber \\ &+\tau^{ac}_{ik} \chi_{cj}^{kb}P(ab)P(ij)
+ -\frac{1}{2}\tau^{dc}_{ji} u^{ab}_{dc} \nonumber
+ \\ &+u^{ab}_{ij} \label{AlgebraicTampH2},
 \end{align}
-where $P(ij)$ is an anti-symmtrizer in the sense that $f(ij)P(ij) = f(ij)-f(ji)$ and similarly for $P(ab)$. We have defined the so-called intermediates,
+where $P(ij)$ is an anti-symmtrizer in the sense that $f(ij)P(ij) =
+f(ij)-f(ji)$ and similarly for $P(ab)$. We have defined the so-called
+intermediates,
 \begin{align}
@@
 \end{align}
-We use the Einstein summation convention over repeated indices of opposite vertical placement. The appearance of a $P(ij)$ or a $P(ab)$ should be ignored
-for the invocation of the summation convention. This means that for example the expression $h^{k}_{i} \tau^{ab}_{jk} P(ij)$ is interpreted as,
+We use the Einstein summation convention over repeated indices of
+opposite vertical placement. The appearance of a $P(ij)$ or a $P(ab)$
+should be ignored for the invocation of the summation convention. This
+means that for example the expression $h^{k}_{i} \tau^{ab}_{jk} P(ij)$
+is interpreted as,
 \begin{equation*}
@@
 \end{equation*}
-In other words, we sum over all indices that does not appear on the left-hand side of all following expressions.
+In other words, we sum over all indices that does not appear on the
+left-hand side of all following expressions.
 
-Evaluation of the right hand side of equation \eqref{ExpexLamplitude} results in,
+Evaluation of the right hand side of equation \eqref{ExpexLamplitude}
+results in,
 \begin{align}
@@
 \end{comment}
-\clearpage
 \bibliographystyle{ieeetr}

### 2019-01-15 doc/MachineLearning/Article.tex
added ml paper

--- 2018-04-14 doc/TimeDependence/Tex/Article.tex
@@
 
-\title{Time-dependent optimized coupled-cluster method for multielectron dynamics}
-
-\author{Håkon Emil Kristiansen}
-\email[Electronic mail:]{haakoek@student.matnat.uio.no}
+\title{Machine Learning and the Many-body Problem}
+
+\author{Jane Kim}
+\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
 \affiliation{Department of Physics, University of Oslo, N-0316 Oslo, Norway}
 
+\author{Vilde Flugsrud}
+\affiliation{Department of Physics, University of Oslo, N-0316 Oslo, Norway}
+
+\author{Sebastian G.~Winther-Larsen}
+\affiliation{Department of Physics, University of Oslo, N-0316 Oslo, Norway}
+
+\author{Morten Hjorth-Jensen}
+\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
+\affiliation{Department of Physics and Center for Computing in Science Education, University of Oslo, N-0316 Oslo, Norway}
+
 \begin{abstract}
-We present a working implementation of a simplified version of the
-orbital-adaptive time-dependent coupled-cluster method which we refer
-to as just the time-dependent coupled-cluster (TDCC) method.
-OATDCC/TDCC approximates solutions to the quantum $N$-body problem and
-for $N=2$ TDCC is equivalent to the time-dependent
-configuration-interaction method (TDCI).  A TDCI solver is implemented
-and shown to be correct by comparing with a previous study and is then
-used as a foundation for verifying the implementation of the TDCC
-method. For one and two dimensional quantum dot systems with $N=2$ we
-demonstrate that the methods are equal to numerical
-precision. Furthermore, for a small number of basis functions we use
-the TDCC solver to compute the time evolution of six-electron
-isotropic quantum dot.
+
 \end{abstract}

### 2019-01-15 doc/Correlations/Figs/plot.py

@@ -11,7 +11,7 @@ import os
 rc('text',usetex=True)
 rc('font',**{'family':'serif','serif':['New fits of nuclear forces']})
 
-data = loadtxt("protonf72.dat")
+data = loadtxt("f72.dat")
 
 l = len(data[:,0])
 

### 2019-01-15 doc/MachineLearning/Article.tex
added ml paper

@@
 
-\author{Sebastian G.~Winther-Larsen}
+\author{Alocias Madarson}
 \affiliation{Department of Physics, University of Oslo, N-0316 Oslo, Norway}
@@
 
-Quantum mechanics is a theory that describes/models the properties of
-microscopic systems. It postulates that given the so-called
-wavefunction, $\Psi(\mathbf{r},t)$, we can in principle compute all
-that there is to know about the system.  Given suitable initial
-conditions $\Psi(\mathbf{r},t)$ can be determined for all future time
-by solving the time-dependent Schrödinger equation (TDSE),
-\begin{equation}
- i \hbar \frac{\partial }{\partial t} \Psi(\mathbf{r},t) = \hat{H} \Psi(\mathbf{r},t).
-\end{equation}
-The initial condition is typically taken as a linear combination of what is called stationary states, which in turn can be found by solving what is known as the 
-time-independent Schrödinger (TISE) equation
-\begin{equation}
- \hat{H} \Psi(\mathbf{r}) = E \Psi(\mathbf{r}).
-\end{equation}
-Exact/analytical solutions to the TISE is possible only for the
-simplest systems and general solutions to the TDSE is even
-rarer. Thus, one must resort to approximative/numerical methods. Of
-particular interest is the so-called many-body problem where one
-considers systems of a large number of interacting particles, where
-large can be anywhere from two to infinity.
-
-Several approaches for solving the many-body TISE
-numerically/approximatively has been devised such as Hartree-Fock
-(HF), Density Functional Theory (DFT), Configuration Interaction (CI)
-and Coupled Cluster (CC). While efficient, Hartree-Fock and DFT are
-insufficient if one wants a high degree of accuracy. Configuration
-Interaction and Coupled Cluster methods are hierarchical in the sense
-that one can systematically construct increasly accurate
-approximations. If the CI method is not truncated we have what is
-known as Full Configuration Interaction (FCI). FCI can be seen as
-exact (within some finite space), however it suffers from exponential
-scaling. Truncated CI methods which would achieve polynomial scaling
-are problematic since they are not size-consistent and
-extensive. Truncated CC methods one the other hand are size-consistent
-and extensive and achieves polynomial scaling. Due to this fact
-Coupled Cluster is considered the gold standard of many-body
-techniques if high accuracy is desired.
-
-Similarly, there exists different approximations to the solution of
-the TDSE with the Multiconfiguration Time-Dependent Hartree-Fock
-method (MCTDHF) being considered the most accurate. MCTDHF, which is
-combination of CI and HF generalized to the time domain, suffers from
-exponential scaling. In a recent article \cite{Kvaal12}, Simen Kvaal
-proposes a method based on Coupled-Cluster theory, the so-called
-Orbital Adaptive Time Dependent Coupled Cluster method (OATDCC), which
-is a hierarchical approximation to the MCTDHF method. The OATDCC
-inherits size-consisteny and extensivity from the CC method and
-achieves polynomial scaling.
-
-\section{Theory/Formalism}
-\subsection{The model Hamiltonian}
-We shall model the quantum dot systems as a collection of $N$
-nonrelativistic electrons of mass $m$ in $d$-dimensional space trapped
-by a confining potential $v_{\text{conf}}(\mathbf{\hat{r}})$. Here
-$\hat{\mathbf{r}}$ is the position operator in $d$-dimensional space
-\begin{equation}
- \hat{\mathbf{r}} \equiv \sum_{i=1}^d \hat{\mathbf{x}}_i \hat{\mathbf{e}}_i
-\end{equation}
-with $\hat{\mathbf{e}}_i$ a unit vector.
-
-The electrons interact through the Coulomb interaction $e^2/(4\pi
-\epsilon R)$, where $e$ is the electron charge, $\epsilon$ the
-permitivity of the medium and $R$ is the distance between two
-interacting electrons.
-
-For convenience, we will use atomic untis, chosen such that $\hbar
-\equiv m \equiv e \equiv 4\pi \epsilon \equiv 1$. Thus, all energies
-are presented in hartrees per $\hbar$.
-
-In atomic units, the many-body problem is described by the Hamiltonian
-on second quantized form
-\begin{align}
- \hat{H} = \hat{H}_1 + \hat{H}_2 = &\sum_{pq} \braket{p|\hat{h}_0|q}\hat{a}_p^\dagger\hat{a}_q \nonumber \\
- + \frac{1}{4}&\sum_{pqrs}\braket{pq|\hat{u}|rs}_{\text{AS}}\hat{a}_p^\dagger\hat{a}_q^\dagger\hat{a}_s\hat{a}_r \label{FieldFreeHamiltonian}
-\end{align}
-where $\hat{H}_1$ and $\hat{H}_2$ are its one- and two-body parts.
-
-The operator $\hat{h}_0$ is the single-particle Hamiltonian
-\begin{equation}
- \hat{h}_0 = -\frac{1}{2}\nabla^2 + v_{\text{conf}}(\mathbf{\hat{r}})
-\end{equation}
-while $\hat{u}$ represents the Coulomb interaction. Here $\ket{p}$ is
-a single-particle basis function which in its position representation
-is given by
-\begin{equation}
- \braket{\mathbf{r}|p} \equiv \phi_p(\mathbf{r}).
-\end{equation}
-In practice, we have to choose some single-particle basis to compute
-within. One typical choice is to take the single-particle basis to be
-the eigenfunctions of $\hat{h}_0$ such that
-\begin{equation}
- \hat{h}_0 \ket{p} = \epsilon_p \ket{p},
-\end{equation}
-where $\epsilon_p$ is the corresponding eigenvalue.
-
-The quantities $\braket{p|\hat{h}_0|q}$ and
-$\braket{pq|\hat{u}|rs}_{\text{AS}}$ are referred to as \textit{matrix
-  elements} and must be computed relative to the chosen
-single-particle basis. In particular,
-$\braket{pq|\hat{u}|rs}_{\text{AS}}$ are \textit{antisymmetrized}
-matrix elements, defined by
-\begin{equation}
- \braket{pq|\hat{u}|rs}_{\text{AS}} \equiv u^{pq}_{rs} - u^{pq}_{sr}
-\end{equation}
-where $u^{pq}_{rs}$ is the interaction integral, 
-\begin{equation}
- u^{pq}_{rs} \equiv \int \phi^*_p(\mathbf{r}_1) \phi^*_q(\mathbf{r}_2) \hat{u}(\mathbf{r}_1,\mathbf{r}_2) \phi_r(\mathbf{r}_1)\phi_s(\mathbf{r}_2) d\mathbf{r}_1 d\mathbf{r}_2.
-\end{equation}
-In the rest of this paper we will drop the subscript AS for the
-antisymmetrized matrix elements.
-
-In order to generate dynamics the electrons are coupled to a time
-dependent monochromatic laser field, with frequency $\omega$ and
-amplitude $\mathcal{E}_0$, polarized along the $x$-axis
-\cite{Zanghellini04}
-\begin{equation}
- v_{\text{laser}}(\mathbf{\hat{r},t}) = \hat{\mathbf{x}}_1 \mathcal{E}_0 \sin(\omega t),
-\end{equation}
-resulting in a time dependent single-particle Hamiltonian
-\begin{equation}
- \hat{h}_t = \hat{h}_0 + v_{\text{laser}}(\mathbf{\hat{r},t}).
-\end{equation}
-The choice of polarization axis is arbitrary.
-
-The total time dependent Hamiltonian of the system is then given by
-\begin{align}
- \hat{H}(t) = &\sum_{pq} \braket{p|\hat{h}_t|q}\hat{a}_p^\dagger\hat{a}_q \nonumber \\
- + \frac{1}{4}&\sum_{pqrs}\braket{pq|\hat{u}|rs}_{\text{AS}}\hat{a}_p^\dagger\hat{a}_q^\dagger\hat{a}_s\hat{a}_r.
-\end{align}
-It is important to note that this implies that we have to update the
-matrix elements $\braket{p|\hat{h}_t|q}$ during the
-simulation. However, in this work we consider only methods where the
-single-particle basis is held fixed in time, hence the two-body matrix
-elements can be computed $\braket{pq|\hat{u}|rs}$ once and for all.
-
-We will handle the time dependent problem in two steps. First we
-prepare a system in its groundstate by solving the time independent
-Schrödinger equation (TISE) for the field-free Hamiltonian
-\eqref{FieldFreeHamiltonian}.  Then we apply the time dependent laser
-field and compute the time evolution from the time dependent
-Schrödinger equation (TDSE).  Before we discuss the metods for solving
-the TISE and TDSE respectively we will adress the choice of
-single-particle basis.
-
-\subsection{Choice of single-particle basis}
-The choice of single-particle basis will in general depend on the
-problem. In quantum chemistry it is customary to expand the
-eigenfunctions, $\{ \psi_k \}$, of $\hat{h}_0$ in (non-orthogonal)
-Gaussian basis functions $g_p(\mathbf{r})$
-\begin{equation}
- \hat{h}_0 \ket{\psi_k} = \hat{h}_0 \sum_{p=1}^L g_p(\mathbf{r}) = \epsilon_k \sum_{p=1}^L g_p(\mathbf{r})
-\end{equation}
-resulting in a generalized eigenvalue problem which must be solved for
-the eigenfunctions. Ideally we would let $L \rightarrow \infty$ but
-must be taken as finite in practice.  The choice of Gaussian basis
-functions is popular due to the fact that matrix elemtents can be
-evaluated efficiently \cite{Helgaker00book}.
-
-Another popular choice, which is the one chosen in this work, is to
-use the eigenfunctions of the Harmonic oscillator (HO)
-\begin{equation}
- \left(-\frac{1}{2}\nabla^2 + \frac{1}{2} \omega_{\text{HO}}\hat{\mathbf{r}}^2 \right) \phi_k(\mathbf{r}) = \epsilon_k \phi_k(\mathbf{r})  
-\end{equation}
-as a single-particle basis. The solutions to the HO-problem is well
-known and the eigenfunctions have closed form expressions.
-
-Of particular importance for our purposes is the fact that the
-interaction integrals
-\begin{equation}
- u^{pq}_{rs} \equiv \int \phi^*_p(\mathbf{r}_1) \phi^*_q(\mathbf{r}_2) \hat{u}(\mathbf{r}_1,\mathbf{r}_2) \phi_r(\mathbf{r}_1)\phi_s(\mathbf{r}_2) d\mathbf{r}_1 d\mathbf{r}_2
-\end{equation}
-can be solved analytically in two- and three-dimensions, using
-HO-basis functions in polar and spherical coordinates respectively
-\cite{Anisimovas98, Vorrath2003}. If the confining potential is chosen
-to be the Harmonic oscillator potential we are essentially done and
-can move on to the many-body treatment of the problem.
-
-Additionally we may approximate the eigenfunctions of more general
-confining potentials by expanding the solutions in the HO-functions.
-
-Consider the general problem
-\begin{equation}
- \hat{h}_0 \ket{\psi_k} = \epsilon_k \ket{\psi_k} \label{EigenfuncEq}
-\end{equation}
-where 
-\begin{equation}
- \hat{h}_0 = -\frac{1}{2}\nabla^2 + v_{\text{conf}}(\mathbf{r}).
-\end{equation}
-Expand the eigenfunctions $\ket{\psi_k}$ in the HO-functions $\ket{\phi_p}$
-\begin{equation}
- \ket{\psi_k} = \sum_{p=1}^L c_{p,k} \ket{\phi_p} 
-\end{equation}
-and insert into Eq.\eqref{EigenfuncEq}. Left-projection with
-$\bra{\phi_q}$ results in the following expression
-\begin{equation}
- \sum_{p=1}^L c_{p,k} \braket{\phi_q|\hat{h}_0|\phi_p} = \epsilon_k \sum_{p=1}^L c_{p,k} \delta_{q,p}
-\end{equation}
-where we have used the orthogonality of the HO-functions. This can be written on matrix form as 
-\begin{equation}
- HC = C\mathcal{E}
-\end{equation}
-where $H_{qp} \equiv \braket{\phi_q|\hat{h}_0|\phi_p}$, $\mathcal{E}$
-is a diagonal matrix containing the eigenvalues on its diagonal, and
-$C$ is matrix conatining the expansions coefficients for basis
-function $k$ in its columns. The eigenproblem must be solved for the
-expansions coefficients, resulting in a approximation of the
-interaction integrals in terms of the known intercation integrals
-$\braket{\phi_\alpha \phi_\beta|\hat{u}|\phi_\gamma \phi_\delta}$
-given by
-\begin{equation}
- \braket{\psi_p \psi_q| \hat{u}| \psi_r \psi_s} = \sum_{\alpha \beta \gamma \delta} c^*_{p,\alpha} c^*_{q,\beta} c^*_{r,\gamma} c^*_{s,\delta} \braket{\phi_\alpha \phi_\beta|\hat{u}|\phi_\gamma \phi_\delta}.
-\end{equation}
-
-The quality of such an approximation will naturally depend on how
-close the confining potential is to the Harmonic oscillator
-potential. Nevertheless it constitutes a convenient starting point for
-investigations of more general potentials.
-
-\subsection{Configuration interaction theory}
-
-\subsubsection{Time independent Configuration interaction}
-The conceptually simplest method for solving the many-problem is
-probably that of Configuration interaction (CI).
-
-Let us first consider the time independent Schrödinger equation 
-\begin{equation}
- \hat{H} \ket{\Psi_k} = E_k \ket{\Psi_k}
-\end{equation}
-where $\hat{H}$ is the field-free Hamiltonian given by
-Eq. \eqref{FieldFreeHamiltonian}, $\ket{\Psi_k}$ is the many-body
-eigenfunction and $E_k$ its corresponding eigenvalue. It is customary
-to use a (finite in practice) basis of orthogonal Slater determinants
-(SD), $\{ \Phi_p \}_{p=1}^{N_{\text{sd}}}$ where $N_{\text{sd}}$ is
-the number of determinants, built from the single-particle basis
-functions described in the previous section.
-
-The principal idea of CI is to expand the wavefunction in the SD-basis
-\begin{equation}
- \hat{H} \sum_{p=1}^{N_{\text{sd}}} c_{p,k} \ket{\Phi_p} = E_k \sum_{p=1}^{N_{\text{sd}}} c_{p,k} \ket{\Phi_p}.
-\end{equation}
-If we left-project with $\bra{\Phi_q}$ the above equation can be written on matrix form as 
-\begin{equation}
- HC = CE
-\end{equation}
-where $H_{qp} \equiv \braket{\Phi_q|\hat{H}|\Phi_p}$, $E$ is a
-diagonal matrix with the eigenvalues along its diagonal and $C$ is a
-matrix containing the expansion coefficients of the $k$-th
-eigenfunction in column $k$. Thus if we can compute the matrix
-elements $H_{qp}$ the TISE can be solved by numerical diagonalization.
-
-The evaluation of $H_{qp}$ is simplified by what is known as the
-Slater-Condon rules. If $\ket{\Phi}$ is any SD, the only non-zero
-contributions to $H$ are given by
-\begin{align}
- \braket{\Phi| \left(\hat{H}_1 + \hat{H}_2\right)|\Phi} &= \sum_i \braket{i|\hat{h}_0|i} + \frac{1}{2}\sum_{ij}\braket{ij|\hat{u}|ij} \label{SlaterCondonEqual}\\
- \braket{\Phi| \left(\hat{H}_1 + \hat{H}_2\right)|\Phi_k^c} &= \braket{k|\hat{h}_0|c} + \sum_{i} \braket{ki|\hat{u}|ci} \label{SlaterCondonDiffByOne} \\
- \braket{\Phi| \left(\hat{H}_1 + \hat{H}_2\right)|\Phi_{kl}^{cd}} &= \braket{kl|\hat{u}|cd} \label{SlaterCondonDiffByTwo}.
-\end{align}
-In other words, we only have to consider the overlap between
-determinants that differ by at most two occupied single-particle
-functions.  Here we use the convention that $i,j,k,l = 1,...,N$, with
-$N$ being the number of electrons and $a,b,c,d = N+1,...$ and $p,q,r,s
-= 1,...,L$. Thus, we see that if the matrix elements
-$\braket{p|\hat{h}_0|q}$ and $\braket{pq|\hat{u}|rs}$ are known we can
-set up the Hamilton matrix and diagonalize it, obtaining the expansion
-coefficients $C$.
-
-Ideally we would like $N_{\text{sd}} \rightarrow \infty$ which is
-impossible in practice. If we choose a finite number, $L$, of
-single-particle basis functions and use all the linearly independent
-Slater determinants we can build from this basis we get what is known
-as the Full Configuration Interaction (FCI). However, the problem with
-FCI is that the number of linearly independent determinants are given
-by
-\begin{equation}
- N_{\text{sd}} = \binom{L}{N}
-\end{equation}
-with $N$ being the number of electrons. This means that the numbers of
-SD's becomes huge very fast, making FCI practical/feasible only for
-the smallest systems. On the other hand, FCI is exact within a given
-computational space, thus it is an invaluable tool in aiding us
-determining the validity of more sophisticated methods. Furthermore,
-CI is variational in the sense that it guarantees an upper bound on
-the groundstate energy.
-
-One way to adress the problems with FCI is to consider a
-\textit{reference} determinant $\ket{\Phi}$, typically taken to be the
-Hartree-Fock state, and consider excitations upon this state. That is,
-we make the ansatz
-\begin{equation}
- \ket{\Psi} = A_0 \ket{\Phi} + \sum_{ia} A_{ia} \ket{\Phi_{i}^a} + \frac{1}{4} \sum_{ijab} A_{ijab} \ket{\Phi_{ij}^{ab}} + \cdots
-\end{equation}
-where $\ket{\Phi_i^a}$ is a singly-excited determinant,
-$\ket{\Phi_{ij}^{ab}}$ doubly-excited etc. and truncate at some
-excitation level.  This results in a hierarchy of approximations known
-as CIS (only single-excitations), CID (only double-excitations), CISD
-(singles and doubles and so on. In this way one obtains polynomial
-scaling of the number of determinants. The problem with this approach
-is that the truncated CI-wavefunction is not size-consistent and
-extensive \cite{Helgaker00book}. It turns out that this problem is
-resolved by the Coupled Cluster method, which makes it the de facto
-standard of many-body methods today.
-
-\subsubsection{Time dependent Configuration interaction}
-Another advantage with CI is that it is straightforward to generalize
-to time dependent problems. Recall that the time evolution of an
-arbitrary state, $\ket{\Psi(t)}$ with initial condition
-$\ket{\Psi(t_0)}$, is given by the time dependent Schrödinger
-equation,
-\begin{equation}
- i \hbar \frac{\partial}{\partial t}\ket{\Psi(t)} = \hat{H}(t) \ket{\Psi(t)}.
-\end{equation}
-We will take the initial condition to be some (normalized) linear
-combination of the CI eigenfunctions. In particular, we are interested
-in the time evolution of the groundstate.
-
-The time-dependent Configuration Interaction (TDCI) method is obtained
-by placing a time-dependence on the expansion coefficients in the
-CI-wavefunction while keeping the single-particle basis
-fixed. Analogous to what we did in the previous section, we expand
-$\ket{\Psi(t)}$ in a given orthonormal Slater determinant basis $\{
-\Phi_p\}_{p=1}^{N_{\text{sd}}}$
-\begin{equation}
- \ket{\Psi(t)} = \sum_{p=1}^{N_{\text{sd}}} A_p(t) \ket{\Phi_p}.
-\end{equation}
-If we now insert this expansion into the time-dependent Schrödinger equation we obtain,
-\begin{equation}
- i \hbar  \sum_{p=1}^{N_{\text{sd}}} \frac{\partial A_p(t)}{\partial t} \ket{\Phi_p} =  \sum_{p=1}^{N_{\text{sd}}} A_p(t) \hat{H}(t) \ket{\Phi_p}. 
-\end{equation}
-Left-projecting the above equation with $\bra{\Phi_q}$ and using the
-orthonormality of the Slater determinants, we are left with a
-differential equation for each expansion coefficient $A_J(t)$ which
-can be written as the matrix-vector equation
-\begin{equation}
- i\hbar \frac{\partial A(t)}{\partial t} = H(t)A(t). \label{TDFCIEquation}
-\end{equation}
-Here $H(t)$ is the time dependent Hamilton matrix defined in terms of its elements,
-\begin{equation}
- H_{qp}(t) = \braket{\Phi_q|\hat{H}(t)|\Phi_p}
-\end{equation}
-and $A(t) = [A_0(t), A_1(t), \cdots]$ is the vector containing the
-expansion coefficients. We refer to this as the time dependent
-Configuration Interaction (TDCI) method.
-
-Completely analogous to the time independent case we have a hierarchy
-of TDCI methods defined by the Slater determinant basis. Thus we have
-a series of possible approximations ranging from TD-CIS to TD-FCI.  If
-we did not have to truncate the Slater determinant basis, TDFCI would
-be an exact solution to the TDSE.  However, we must in practice
-truncate the basis and TDFCI suffers from the fact that the size of
-FCI-space grows exponentially.  In order to obtain a good
-approximation to the exact solution we would need a huge fixed
-basis\cite{Kvaal12}.
-
-On the other hand it is quite straightforward to implement, especially
-when a program for computing the CI-groundstate already has been
-written.  Again, the fact that the CI and CC space are equivalent for
-$N=2$ makes it worthwhile to write a TDCI program, which we can use to
-compare with the time dependent Coupled Cluster method which we
-discuss in the next chapter.
-\begin{comment}
-\subsection{Coupled cluster theory}
-In this section we review Coupled Cluster (CC) theory.  We will look
-at the "classical" way of obtaining the CC equations. The presentation
-borrows heavily from the excellent review by Crawford and
-Schaefer\cite{Crawford2000}.
-
-Furthermore we look at a different approach which in addition to the
-usual CC equations also gives a method to solve the time dependent
-Schrödinger equation, the so-called Orbital Adaptive Coupled Cluster
-(OATDCC) method\cite{Kvaal12}.
-
-\subsubsection{The exponential ansatz}
-As usual we start by considering the time-independent Schrödinger equation
-\begin{equation*}
-\hat{H} \ket{\Psi} = E\ket{\Psi}.
-\end{equation*}
-In Coupled Cluster theory an exponential ansatz 
-\begin{equation}
-\ket{\Psi}_{CC} \equiv e^{\hat{T}} \ket{\Phi} \label{ExponantialAnsatzCC},
-\end{equation}
-is used to approximate the exact solution. Here $\ket{\Phi}$ is a
-reference Slater determinant and $\hat{T} = \sum_{i}^n \hat{T}_i$ is
-the sum over n-orbital cluster operators.
-
-The the one- and two-orbital cluster operators are defined as
-\begin{equation}
-\hat{T}_1 \equiv \sum_{ia} t_i^a c_a^\dagger c_i
-\end{equation}
-and 
-\begin{equation}
-\hat{T}_2 \equiv \frac{1}{4} \sum_{ijab} t_{ij}^{ab} c_b^\dagger c_j c_a^\dagger c_i.
-\end{equation}
-Generally an n-orbital cluster operator is defined as 
-\begin{equation}
-\hat{T}_n \equiv \left(\frac{1}{n!}\right)^2 \sum_{ijk...abc...}t_{ijk...}^{abc...} ...c_c^\dagger c_k c_b^\dagger c_j c_a^\dagger   c_i.
-\end{equation}
-%We see that the cluster operators are linear combinations of excitation operators as defined by \eqref{DefinitionOfExciationOperator}. 
-Recalling the exponential series we can write,
-\begin{equation}
- \ket{\Psi}_{CC} = e^{\hat{T}} \ket{\Phi} = \left( 1+\sum_{k=1}^\infty \frac{1}{k!}\hat{T}^k \right) \ket{\Phi}. \label{CCansatzSeriesExpansion}
-\end{equation}
-
-\subsection{The Coupled Cluster Equations}
-We now proceed by inserting expression \eqref{ExponantialAnsatzCC}
-into the time-independent Schrödinger equation
-\begin{equation}
-\hat{H}e^{\hat{T}} \ket{\Phi} = Ee^{\hat{T}} \ket{\Phi} \label{ExponantialAnsatzCCInsertedInSE}.
-\end{equation}
-Left-projecting with $\bra{\Phi}$ we get an expression for the energy,
-\begin{align*}
- \braket{\Phi|\hat{H}e^{\hat{T}}|\Phi} &= E \braket{\Phi|e^{\hat{T}}|\Phi} = E \braket{\Phi|\Psi_{CC}} \\
-				       &= E
-\end{align*}
-where intermediate normalization, $\braket{\Phi|\Psi_{CC}} = 1$, is
-assumed. Furthermore one can obtain an expression for the amplitude
-$t^{ab...}_{ij...}$ upon left-projection with
-$\bra{\Phi^{ab...}_{ij...}}$,
-\begin{equation}
- \braket{\Phi^{ab...}_{ij...}|\hat{H}e^{\hat{T}}|\Phi} = E\braket{\Phi^{ab...}_{ij...}|e^{\hat{T}}|\Phi}.
-\end{equation}
-Notice that this equations are non-linear (due to the presence of
-$e^{\hat{T}}$) and depend on the energy equation. This is impractical
-and is resolved by performing a similarity transformation of the
-Hamiltonian.
-
-Multiply Eq. \eqref{ExponantialAnsatzCCInsertedInSE} with
-$e^{-\hat{T}}$ and left-project with $\bra{\Phi}$ in order to obtain
-\begin{equation}
- \braket{\Phi|e^{-\hat{T}} \hat{H} e^{\hat{T}} | \Phi} = E \braket{\Phi|e^{-\hat{T}}e^{\hat{T}}|\Phi} = E. \label{SimTransEnergyEquation}
-\end{equation}
-Left-projection with $\bra{\Phi^{ab...}_{ij...}}$ now results in,
-\begin{equation}
- \braket{\Phi^{ab...}_{ij...}|e^{-\hat{T}}\hat{H}e^{\hat{T}}|\Phi} = E \braket{\Phi^{ab...}_{ij...}|\Phi} = 0. \label{SimTransAmplitudeEquation}
-\end{equation}
-Equations \eqref{SimTransEnergyEquation} and
-\eqref{SimTransAmplitudeEquation} are commonly referred to as the
-Coupled Cluster equations. Notice now that the equation for the
-amplitudes does not depend on the energy. The amplitude equation is
-non-linear and has to be solved iteratively. We say that we have a
-self-consistent solution to CC-equations if the energy converges to
-some pre-defined precision.
-
-In order to implement the Coupled Cluster method we need algebraic
-expressions for the CC-equations.
-\begin{theorem}[Baker-Campbell-Hausdorff expansion]
-For any matrices $A$ and $S$ the Baker-Campbell-Hausdorff (BCH) expansion is given by,
-\begin{equation}
-e^{-S}Ae^{S} = A + [A,S] + \frac{1}{2}[[A,S],S] + \frac{1}{3!}[[[A,S],S],S] + ... \label{BCH},
-\end{equation}
-where 
-$$[A,S] \equiv AS - SA$$ 
-is the commutator between $A$ and $S$. 
-\end{theorem}
-Under the assumption that $\hat{H}$ contains at most a two-body
-operator the BCH-expansion truncates after the first five terms on the
-right-hand side of eq. \eqref{BCH}\cite{Crawfordtruncates}.  This fact
-greatly simplifies the evaluation of the left-hand sides of the
-equations \eqref{SimTransEnergyEquation} and
-\eqref{SimTransAmplitudeEquation}.
-
-These expressions can be evaluated by hand using the second
-quantization formalism as demonstrated in\cite{Crawford2000} or by
-diagrammatic techniques\cite{ShavittBartlett}. Alternatively one can
-use the second quantization toolbox in the Python library
-Sympy\footnote{\href{http://docs.sympy.org/latest/modules/physics/secondquant.html}{http://docs.sympy.org/latest/modules/physics/secondquant.html}}
-to compute these expressions.
-
-\subsection{A variational CC theory?}
-One drawback with the CC-method is that it does not satisfy the
-variational principle \eqref{variationalTheorem}. Consider the
-operator $e^{-\hat{T}}\hat{H}e^{T}$ which appears in the similarity
-transformed energy equation \eqref{SimTransEnergyEquation}. This
-operator is not Hermitian. In order to see this compute
-$\hat{T}_1^\dagger$,
-\begin{equation}
- \hat{T}_1^\dagger = \left(\sum_{ia} t_i^a c_a^\dagger c_i \right) = \sum_{ia} (t_i^a)^* c_i^\dagger c_a \neq \hat{T}_1.
-\end{equation}
-Then it follows that
-\begin{equation}
- \left(e^{-\hat{T}}\hat{H}e^{\hat{T}}\right)^\dagger = (e^{\hat{T}})^\dagger \hat{H} (e^{-\hat{T}})^\dagger = e^{\hat{T}^\dagger} \hat{H} e^{-\hat{T}^\dagger} \neq e^{-\hat{T}}\hat{H}e^{\hat{T}}.
-\end{equation}
-Hermiticity of the operator was a necessary condition for the
-variational theorem, thus Eq. \eqref{SimTransEnergyEquation} is not
-variational.  However, if $\hat{T}$ is not truncated the spectrum of
-$e^{-\hat{T}}\hat{H}e^{T}$ is identical to the original Hermitian
-operator $\hat{H}$, which justifies its use in quantum mechanical
-models\cite{CrawfordNonVar}.
-
-There has been attempts to construct a variational solution by
-deriving the amplitude equations by minimizing the functional,
-\begin{equation}
- \frac{\braket{\Psi|\hat{H}|\Psi}}{\braket{\Psi|\Psi}} = \frac{\braket{\Phi|(e^{\hat{T}})^\dagger \hat{H}e^{T}|\Phi}}{\braket{\Phi|(e^{\hat{T}})^\dagger e^{\hat{T}}|\Phi}}.
-\end{equation}
-Notice then that $(e^{\hat{T}})^\dagger \hat{H}e^{T}$ does not conform to the Baker-Campbell-Hausdorff formula and the series has no natural truncation, complicating 
-matters greatly. 
-\subsection{The Orbital Adaptive Time Dependent Coupled Cluster Method}
-In the article\cite{Kvaal12} Simen Kvaal demonstrates that it is
-possible to derive equations of motion for the CC-ansatz by computing
-critical points of a bivariational expectation value functional.  In
-his approach both the orbitals and CC-amplitudes are allowed to vary
-and the method can be used to solve the time-dependent Schrödinger
-equation.  The method is called the Orbital Adaptive Time-Dependent
-Coupled Cluster (OATDCC) and is a hierarchical approximation to
-MCTDHF. Furthermore, if the orbitals are held fixed in time the method
-gives a approximation to the Time-Dependent CI method which we refer
-to as the Time-Dependent Coupled Cluster (TDCC) method.  While TDFCI
-and MCTDHF suffers from exponential scaling, OATDCC and TDCC achieves
-polynomial scaling.
-
-The main goal of this article is to implement the TDCC method for a
-possibly time-dependent Hamiltonian.  As such we will in the following
-summarize the key ingredients needed to achieve this.  For
-mathematical details the interested reader is referred to the article
-by Kvaal.
-
-\subsubsection{The Bivariational Principle}
-As we have established, the Coupled Cluster method is not variational
-in the usual sense since the similarity transformed Hamiltonian,
-$$ \bar{H} = e^{-\hat{T}}\hat{H}e^{\hat{T}} $$ is not Hermitian. Now
-following Kvaal, let $A$ be an operator (possibly non-Hermitian) over
-Hilbert space $\mathcal{H}$ and consider the expectation value
-functional
-\begin{equation}
- \mathcal{E}_A: \mathcal{H}^{\prime}\times \mathcal{H} \rightarrow \mathbb{C}, \ \ \mathcal{E}_A [\bra{\Psi^{\prime}}, \ket{\Psi}] = \frac{\braket{\Psi^{\prime} |A| \Psi} }{\braket{\Psi^{\prime}|\Psi}}.
-\end{equation}
-This is referred to as the the bivariational expectation value
-functional. In contrast to the usual variational principle
-$\bra{\Psi^{\prime}}$ and $\ket{\Psi}$ are treated as independent
-parameters. The conditions for $\delta \mathcal{E}_A = 0$, for all
-independent variations of $\bra{\Psi^{\prime}}$ and $\ket{\Psi}$ is
-\begin{equation}
- (A-a)\ket{\Psi} = 0 \ \ \text{and} \ \ \bra{\Psi^{\prime}}(A-a) = 0,
-\end{equation}
-with $$a = \mathcal{E}_A[\bra{\Psi^{\prime}},\ket{\Psi}] $$
-being the value of $\mathcal{E}_A$ at the critical point.
-
-Equations of motion can be derived from so-called time-dependent
-variational principles. One such variational is set up using the
-Lagrange formulation\cite{BECK20001},
-\begin{equation}
- \delta \int_0^T L[\ket{\Psi}]dt = 0,
-\end{equation}
-where the Lagrangian functional is given by 
-\begin{equation}
-L[\ket{\Psi}] = \braket{\Psi(t)|\hat{H}-i\hbar \frac{\partial}{\partial t}|\Psi(t)}
-\end{equation}
-with the boundary conditions $\delta L(t_1) = \delta L(t_2) = 0$. The functional
-\begin{equation}
-\mathcal{S}[\ket{\Psi}] \equiv \int_0^T \braket{\Psi(t)|\hat{H}-i\hbar \frac{\partial}{\partial t}|\Psi(t)} dt \label{ActionFunctional1}
-\end{equation}
-is referred to as the action functional and computing equations of
-motion from $\delta \mathcal{S} = 0$ is known as the prinicipal of
-least action.
-
-By considering a bivariational generalization of the action functional
-\eqref{ActionFunctional1},
-\begin{align}
- \mathcal{S}[\bra{\Psi^{\prime}}, \ket{\Psi}] \equiv &\int_0^T \frac{\braket{\Psi^{\prime}(t)|i\hbar \frac{\partial}{\partial t}-\hat{H}|\Psi(t)}}{\braket{\Psi^{\prime}(t)|\Psi(t)}} dt \\
- &\int_0^T i\hbar \frac{\braket{\Psi^{\prime}(t)|\frac{\partial}{\partial t}\Psi(t)}}{\braket{\Psi^{\prime}(t)|\Psi(t)}} - \mathcal{E}_H[\bra{\Psi^{\prime}(t)}, \ket{\Psi(t)}] dt.
-\label{ActionFunctional}
-\end{align}
-Kvaal derives equations of motions generalizing the Coupled Cluster
-method to the time domain.  Here, one should note the appearance of
-the bivariational generalization of the energy expectation value,
-\begin{equation}
- \mathcal{E}_H[\bra{\Psi^{\prime}}, \ket{\Psi}] = \frac{\braket{\Psi^{\prime} |\hat{H}| \Psi} }{\braket{\Psi^{\prime}|\Psi}} \label{BivarEnergy},
-\end{equation}
-which is required to be complex analytic as shown in Ref.\cite{Kvaal13}.
-
-\subsubsection{The Coupled Cluster Ansatz}
-The fundamental idea in\cite{Kvaal12} is to introduce different
-exponential parametrizations $\ket{\Psi}$, $\bra{\Psi^\prime}$ and
-compute $\delta \mathcal{S} = 0$, with $\mathcal{S}$ given by
-\eqref{ActionFunctional}, which ultimately results in time dependent
-Coupled Cluster equations.
-
-In particular one make the ansatzes 
-\begin{align}
- \ket{\Psi} &= e^{\hat{T}}\ket{\phi} \\ 
- \bra{\Psi^{\prime}} &= \bra{\tilde{\phi}}e^{\hat{T}^\prime},
-\end{align}
-where $\hat{T}$ is the familiar excitation operator
-\begin{equation}
- \hat{T} = \sum_{ia} \tau_i^a c_a^\dagger \tilde{c}_i + \frac{1}{2!^2}
- \sum_{ijab} \tau_{ij}^{ab} c_a^\dagger \tilde{c}_i c_b^\dagger
- \tilde{c}_j + \cdots
-\end{equation}
-and $\hat{T}^{\prime}$ is a de-excitation operator on the form
-\begin{equation}
- \hat{T}^{\prime} = \sum_{ia} (\tau^{\prime})_a^i c_i^\dagger \tilde{c}_a + \frac{1}{2!^2} \sum_{ijab} (\tau^{\prime})_{ab}^{ij} c_i^\dagger \tilde{c}_a c_j^\dagger \tilde{c}_b + \cdots.
-\end{equation}
-Here $\ket{\phi} = \ket{\varphi_{p_1},\cdots,\varphi_{p_N}}$ and $\bra{\tilde{\phi}} = \bra{\tilde{\varphi}_{p_1},\cdots,\tilde{\varphi}_{p_N}}$ are Slater determinants 
-built from biorthogonal single-particle functions, $\braket{\tilde{\varphi}_p|\varphi_q} = \delta_{pq}$,  belonging to separate 
-Hilbert spaces $\mathcal{V}$ and $\tilde{\mathcal{V}}$. The biorthogonality condition implies the anticommutator 
-relation 
-\begin{equation}
- \{ \tilde{c}_p, c_q^\dagger \} \equiv \tilde{c}_p c_q^\dagger + c_q^\dagger \tilde{c}_p = \delta_{pq},
-\end{equation}
-which preserves Wicks theorem. Here $\tilde{c}_p, \tilde{c}_p^\dagger, c_p$ and $c_p^\dagger$ are creation and annihilation opeartotrs associated with 
-the biorthogonal single-particle functions. It should be noted that when $\tilde{c}_p$ or $\tilde{c}_p^\dagger$ acts on the determinant $\ket{\phi}$ it 
-is only responsible for removing or adding $\varphi_p$ (not $\tilde{\varphi}_p$) from the determinant\cite{Kvaal12}. The notation 
-$\Phi = (\varphi_1,\cdots,\varphi_L)$ and $\tilde{\Phi} = (\tilde{\varphi}_1,\cdots,\tilde{\varphi}_L)$ to denote the set of orbitals used to build 
-the Slater determinants $\ket{\phi}$ and $\bra{\tilde{\phi}}$.
-
-In standard CC theory and virtually every other many-body method, the single-particle functions 
-are taken to be orthonogal such that $\mathcal{V} \equiv \tilde{\mathcal{V}}^\prime$. This relaxation of
-orthonormality of the single-particle functions is necessary to ensure that the bivariational functional is
-complex analytic if the single-particle functions are to be treated as variational parameters\cite{Kvaal12}.
-
-Next, a change of variables from $(T^{\prime},T)$ to $(\Lambda, T)$, with
-\begin{equation}
-\Lambda = \sum_{ia} \lambda_a^i c_i^\dagger \tilde{c}_a + \frac{1}{2!^2} \sum_{ijab} \lambda_{ab}^{ij} c_i^\dagger \tilde{c}_a c_j^\dagger \tilde{c}_b + \cdots 
-\end{equation} 
-a de-excitation operator is performed such that
-\begin{equation}
- \bra{\tilde{\Psi}} = \frac{\bra{\Psi^{\prime}}}{\braket{\Psi^{\prime}|\Psi}} = \bra{\tilde{\phi}}(1 + \Lambda)e^{-T}.
-\end{equation}
-Using this parametrization $\braket{\tilde{\Psi}|\Psi} = 1$. The amplitudes $\{\tau_i^a,\tau_{ij}^{ab},\cdots \}$ are the usual Coupled Cluster amplitudes
-which we refer to as the excitation amplitudes or $\tau$-amplitudes. 
-Also we have introduced a second set of amplitudes $\{ \lambda_a^i, \lambda_{ab}^{ij},\cdots \}$ 
-which we will refer to as the de-excitation amplitudes or $\lambda$-amplitudes.
-
-Now the bivariational expectation value functional \eqref{BivarEnergy} reads 
-\begin{equation}
- \mathcal{E}_H [\lambda, \tau, \tilde{\Phi}, \Phi] = \braket{\tilde{\phi}|(I+\Lambda)e^{-T}\hat{H}e^{T}|\phi}.
-\end{equation}
-Inserting the wavefunction ansatzes into the action functional \eqref{ActionFunctional} one arrives at the following expression,
-\begin{equation}
- \mathcal{S}[\lambda,\tau,\tilde{\Phi},\Phi] = \int_0^T i\hbar \braket{\tilde{\phi}|(I+\Lambda)e^{-T}\frac{\partial}{\partial t} e^{T}|\phi} - \mathcal{E}_H[\lambda(t),\tau(t),\tilde{\Phi}(t), \Phi(t)] dt.
-\end{equation}
-Kvaal then shows that the action can be written,
-\begin{align}
- \mathcal{S}[\lambda,\tau,\tilde{\Phi},\Phi] &= \int_0^T i \hbar \sum_\mu \lambda_\mu \dot{\tau}^\mu - \mathcal{E}_{H-i\hbar D_0}[\lambda, \tau, \tilde{\Phi}, \Phi] dt \\ 
- &=\int_0^T i \hbar \lambda_\mu \dot{\tau}^\mu + \rho_p^q(h_q^p - i \hbar \eta_q^p) + \frac{1}{4} \rho_{pr}^{qs}u^{pr}_{qs} dt,
-\end{align}
-where the operator $D_0$ is defined as,
-\begin{equation}
- D_0 \equiv \sum_{pq} \braket{\tilde{\varphi}_p|\dot{\varphi}_q}c_p^\dagger \tilde{c}_q.
-\end{equation}
-Here $\mu$ denotes a general excitation $\{ia, ijab, \cdots\}$. Furthermore,  
-\begin{align*}
- \rho_p^q &= \rho_p^q(\lambda,\tau) \equiv \braket{\tilde{\Psi}|c_p^\dagger \tilde{c}_q|\Psi}, \\
- \rho_{pr}^{qs} &= \rho_{pr}^{qs}(\lambda,\tau) \equiv  \braket{\tilde{\Psi}|c_p^\dagger c_r^\dagger \tilde{c}_s \tilde{c}_q|\Psi}, \\
- h^p_q &= h^p_q(\tilde{\Phi}, \Phi) \equiv \braket{\tilde{\varphi}_p|h|\varphi_q}, \\
- \eta_q^p &= \eta_q^p(\tilde{\Phi}, \dot{\Phi}) \equiv \braket{\tilde{\varphi}_p|\dot{\varphi}_q}, \\
- u^{pr}_{qs} &= u^{pr}_{qs}(\tilde{\Phi},\Phi) \equiv \braket{\tilde{\varphi}_p \tilde{\varphi}_r|u|\varphi_q \varphi_s}_{AS}. \\
-\end{align*}
-The quantities $\rho_p^q$ and $\rho_{pr}^{qs}$ are the CC reduced one- and two-body density matrices. Note that these do not depend explicitly on the orbitals since 
-they are evaluated by Wick's theorem which only depends on the anti-commutator relations \eqref{AntiCommutatorRelations}. Thus, they only depend on the amplitudes.
-
-\subsubsection{OATDCC equations}
-Finally, equations of motion are derived by demanding $\delta \mathcal{S}[\bra{\tilde{\Psi}},\ket{\Psi}] = 0$ for all independent variations 
-of $\bra{\tilde{\Psi}}$ and $\ket{\Psi}$. In general both the single-particle functions and the amplitudes are allowed to vary resulting in equations of motion 
-for both the amplitudes and the single-particle functions. A detailed derivation can be found in the supplementary material to\cite{Kvaal12}.
-
-For the amplitudes the equations of motion read,
-\begin{align}
- i \hbar \dot{\tau}^\mu &= \frac{\partial}{\partial \lambda_\mu } \mathcal{E}_{H-i\hbar D_0} [\lambda, \tau, \tilde{\Phi}, \Phi ] = \braket{\tilde{\phi}_\mu | e^{-T} \left(H-i \hbar D_0\right) e^T | \phi} \label{TAmplitudesEquation}, \\
- -i \hbar \dot{\lambda}_\mu &= \frac{\partial}{\partial \tau^\mu } \mathcal{E}_{H-i\hbar D_0} [\lambda, \tau, \tilde{\Phi}, \Phi ] = \braket{ \tilde{\phi} | (1+\Lambda)e^{-T} [H-i \hbar D_0,X_\mu] e^T | \phi} \label{LAmplitudesEquation}.
-\end{align}
-Here $X_\mu$ are the excitation operators, 
-\begin{align*}
- X_i^a &= c_a^\dagger \tilde{c}_i, \\
- X_{ij}^{ab} &= c_a^\dagger \tilde{c}_i c_b^\dagger \tilde{c}_j, \\ 
- &\vdots \\
-\end{align*}
-Note that in the absence of $D_0$ the right-hand side of \eqref{TAmplitudesEquation} is equivalent with the usual amplitude equations \eqref{SimTransAmplitudeEquation}.
-
-The equations for the unknowns $\eta_q^p(\tilde{\Phi}, \dot{\Phi})$ are given by
-\begin{align}
- i\hbar \sum_{bj} A_{aj}^{ib} \eta_b^j &=   \sum_p \rho_p^i h_a^j - \sum_q \rho_a^q h_b^i + \frac{1}{2} \left[ \sum_{prs} \rho_{pr}^{is}u_{as}^{pr} - \sum_{rqs} \rho_{ar}^{qs}u_{qs}^{ir} \right] \\
- -i\hbar \sum_{bj} A_{bi}^{ja} \eta_j^b &=  \sum_p \rho_p^a h_i^b -\sum_q \rho_i^q h_j^a + \frac{1}{2} \left[ \sum_{prs} \rho_{pr}^{as}u_{is}^{pr} - \sum_{rqs} \rho_{ir}^{qs}u_{qs}^{ar} \right] + i \hbar \dot{\rho}_i^a,
-\end{align}
-where 
-\begin{equation*}
- A_{aj}^{ib} \equiv \langle \tilde{\Psi} | [c_j^\dagger \tilde{c}_b, c_a^\dagger \tilde{c}_i] | \Psi \rangle = \delta_a^b \rho_j^i - \delta_j^i \rho_a^b. 
-\end{equation*}
-For the single-particle functions the equations of motion are given by 
-\begin{align}
-  i\hbar \sum_q \rho_p^q Q \ket{\dot{\varphi}_q} &= \sum_q \rho_p^q Q h \ket{\varphi_q} + \sum_{qrs} \rho_{pr}^{qs} Q W_s^r \ket{\varphi_q} \label{QspaceketEq}\\
- -i\hbar \sum_p \rho_p^q\bra{\dot{\tilde{\varphi}}_p}Q &= \sum_p \rho_p^q\bra{\tilde{\varphi}_p}hQ + \sum_{prs} \rho_{pr}^{qs} \bra{\tilde{\varphi}_p}W_s^rQ,  \label{QspacebraEq}
-\end{align}
-where $Q = I-P$ with $P$ being the projection operator
-\begin{equation}
- \Phi \tilde{\Phi} \equiv \sum_p \ket{\varphi_p}\bra{\tilde{\varphi}_p}.
-\end{equation}
-The operators $W^r_s$ are defined by,
-\begin{equation}
- W^r_s \ket{\psi} \equiv \braket{\cdot \tilde{\varphi}_r|u|\psi \varphi_s} \equiv \int dx_1 \int dx_2 \tilde{\varphi}_r(x_2) u(x_1,x_2) \psi(x_1) \varphi_s(x_2) 
-\end{equation}
-Equations \eqref{TAmplitudesEquation}-\eqref{QspacebraEq} is collectively referred to as the OATDCC equations.
-
-It is not at all obvious how one should compute the 
-initial state for the OATDCC equations when the single-particle functions are allowed to vary 
-and Kvaal proposes several different solutions in his article. 
-\subsubsection{The TDCC equations}
-A special case of OATDCC arises if the orbitals are held fixed in time. Then the operator $D_0$ drops and the OATDCC equations is reduced to only the amplitude 
-equations
-\begin{align}
- i \hbar \dot{\tau}^\mu &= \frac{\partial}{\partial \lambda_\mu } \mathcal{E}_{H} [\lambda, \tau, \tilde{\Phi}, \Phi ] = \langle \tilde{\phi}_\mu | e^{-T} H e^T | \phi \rangle \label{TAmplitudesEquationTDCC} \\
- -i \hbar \dot{\lambda}_\mu &= \frac{\partial}{\partial \tau^\mu } \mathcal{E}_{H} [\lambda, \tau, \tilde{\Phi}, \Phi ] = \langle \tilde{\phi} | (1+\Lambda)e^{-T} [H,X_\mu] e^T | \phi \rangle \label{LAmplitudesEquationTDCC},
-\end{align}
-which we refer to as the TDCC-equations. Notice now that equations \eqref{TAmplitudesEquationTDCC} and \eqref{SimTransAmplitudeEquation} are equivalent. Furthermore, 
-we take $\tilde{\Phi} = \Phi^*$.
-
-The initial amplitudes $(\lambda^{(0)},\tau^{(0)})$ are obtained by solving the non-linear equations
-\begin{align}
- \langle \tilde{\phi}_\mu | e^{-T} H e^T | \phi \rangle &= 0\label{TAmplitudesEquationTDCCGS} \\
- \langle \tilde{\phi} | (1+\Lambda)e^{-T} [H,X_\mu] e^T | \phi \rangle &= 0 \label{LAmplitudesEquationTDCCGS}.
-\end{align}
-
-For the time evolution one then propgates the amplitudes by solving equations \eqref{TAmplitudesEquationTDCC} and \eqref{LAmplitudesEquationTDCC}. 
-The energy is taken as the real part of
-\begin{equation}
- \mathcal{E}_H [\lambda(t), \tau(t)] = \braket{\tilde{\phi}|(I+\Lambda)e^{-T}\hat{H}e^{T}|\phi} \label{CCexpecval},
-\end{equation}
-where the imaginary part should be "small".
-
-Approximations to the TDCC equations are obtained by truncating $T$ and $\Lambda$. For example 
-$T = T_1 + T_2$ and $\Lambda = \Lambda_1 + \Lambda_2$ corresponds to the TDCCSD approximation while $T = T_2$ and $\Lambda = \Lambda_2$ 
-gives the TDCCD approximation.
-
-It is shown in \cite{Kvaal12} that the singles amplitudes are redundant when the single-particle functions are varied. We will therefore restrict ourselves to the 
-TDCCD approximation, using the Hartree-Fock state as reference determinant. This allows for flexibility if one wants to extend the program written in this work 
-to the full OATDCC method.
-
-\subsubsection{The TDCCD approximation} \label{TDCCD approximation section}
-In order to write a computer program that solves the TDCC equations we need algebraic expressions for the right-hand sides of equations \eqref{TAmplitudesEquationTDCC}
-and \eqref{LAmplitudesEquationTDCC}. Now, in the TDCCD approximation we take $T = T_2$ and $\Lambda = \Lambda_2$. 
-We want expressions for the amplitudes and the one- and two-body density matrices. Thus we must evaluate the vacuum expectation values,
-\begin{align}
- \frac{\partial}{\partial \lambda_{ab}^{ij} } \mathcal{E}_{H} &= \braket{\phi_{ij}^{ab}|e^{-T_2}He^{T_2}|\phi} \label{ExpecTamplitude}\\
- \frac{\partial}{\partial \tau^{ab}_{ij} } \mathcal{E}_{H} &= \braket{\phi|(1+\Lambda_2)e^{-T_2}[H, X_{ij}^{ab}]e^{T_2}|\phi}, \label{ExpexLamplitude}\\
- \rho_p^q &= \braket{\tilde{\Psi}|c_p^\dagger \tilde{c}_q|\Psi}, \\
- \rho_{pr}^{qs} &=  \braket{\tilde{\Psi}|c_p^\dagger c_r^\dagger \tilde{c}_s \tilde{c}_q|\Psi}.
-\end{align}
-These are generated by the \textit{CoupledClusterSympy.py} script, which uses the second quantization package in SymPy, 
-and is found in the src folder at the github page\footnote{\href{https://github.com/haakoek/PythonVersionMaster/tree/master/src}{https://github.com/haakoek/PythonVersionMaster/tree/master/src}}. 
-The expressions for the amplitude equations and one- and two-body density matrices are listed in the appendices.
-
-Finally, the energy is taken as the real part of the expectation value \eqref{CCexpecval} which in the CCD approximation is given by,
-\begin{align*}
- \mathcal{E}_H [\lambda(t), \tau(t)] &= \braket{\tilde{\phi}|(I+\Lambda_2)e^{-T_2}\hat{H}e^{T_2}|\phi} \\
-				     &= \braket{\tilde{\phi}|e^{-T_2}\hat{H}e^{T_2}|\phi} + \braket{\tilde{\phi}|\Lambda_2e^{-T_2}\hat{H}e^{T_2}|\phi} \\
-				     &= \braket{\tilde{\phi}|e^{-T_2}\hat{H}e^{T_2}|\phi} + \frac{1}{4} \sum_{ijab} \lambda_{ij}^{ab}\braket{\tilde{\phi}_{ij}^{ab}|e^{-T_2}\hat{H}e^{T_2}|\phi}
-\end{align*}
-where 
-\begin{equation}
- \braket{\tilde{\phi}|e^{-T_2}\hat{H}e^{T_2}|\phi} = h_i^i + \frac{1}{2}u_{ij}^{ij} + \frac{1}{4}\tau_{ij}^{ab}u_{ab}^{ij}. \label{ECCDGroundstate}
-\end{equation}
-Notice that in the groundstate,
-\begin{equation*}
- \braket{\tilde{\phi}_{ij}^{ab}|e^{-T_2}\hat{H}e^{T_2}|\phi} = 0
-\end{equation*}
-which means that the energy for the groundstate is simply given by \eqref{ECCDGroundstate}.
-
-Note also that the one- and two-body density matrices depend only on the amplitudes $(\tau,\lambda)$. This is an advantage when we want to verify the implementation of 
-the $\lambda$-amplitudes since we can compare the density matrices with the density matrices computed with the CI method.
-
-In order to verify that the $\tau$-amplitudes has been computed correctly for the groundstate 
-we compute the groundstate energy and compare with previous published results.
-
-\section{Computational details}
-
-\section{Results}
-
-\section{Conclusions}
-
-\begin{appendices}
- \section{Properties of Cluster operators.}
-\begin{itemize}
-\item $\hat{T}_{N+1} \ket{\Phi} = 0$, where $\hat{T} = \hat{T}_1 + \hat{T}_2 + \cdots$, $\ket{\Phi}$ a reference determinant and $N$ the number of particles 
-in $\ket{\Phi}$. This implies that for a given $N$ the cluster operator $\hat{T}$ has a finite number of terms,
-\begin{equation}
- \hat{T} = \hat{T}_1 + \hat{T}_2 + \cdots + \hat{T}_N.
-\end{equation}
- \item The rank of 
-an excitation/cluster operator is defined as the number of creation or annihilation operators, i.e rank$\left(\hat{T}_2\right) = 2$.
-\item The product of two exciation/cluster operators is an exciation/cluster operator.
-\item All excitation/cluster operators commute. We can prove this by considering the action of the commutator on a reference determinant as follows,
-\begin{align*}
- [c_b^\dagger c_j, c_a^\dagger c_i] \ket{\Phi} &= \left(c_b^\dagger c_j c_a^\dagger c_i - c_a^\dagger c_i c_b^\dagger c_j \right) \ket{\Phi} \\
- &= \ket{\Phi_{ij}^{ab}} \ket{\Phi_{ji}^{ba}} \\
- &= \ket{\Phi_{ij}^{ab}} - \ket{\Phi_{ij}^{ab}} \\
- &= 0 \cdot \ket{\Phi_{ij}^{ab}}.
-\end{align*}
-Thus $c_b^\dagger c_j c_a^\dagger c_i = c_a^\dagger c_i c_b^\dagger c_j$, meaning that $c_b^\dagger c_j$ and $c_a^\dagger c_i$ commute. The result for a 
-general excitation operator follows by induction.
-\item rank$\left(\hat{T}_n \hat{T}_m \right) = n+m$.
-\item Excitation operators are nilpotent, 
-\begin{equation}
- \hat{\tau}_X^2 = \left(\cdots c_c^\dagger c_k c_b^\dagger c_j c_a^\dagger c_i \right)^2 = 0.
-\end{equation}
-This follows immediately from the fact that, 
-\begin{align*}
- \left(c_a^\dagger c_i\right)^2 \ket{\Phi} = c_a^\dagger c_i \ket{\Phi_i^a} = 0 
-\end{align*}
-since $c_i \ket{\Phi_i^a} = 0$.
-\item $\hat{\tau}_X^\dagger = \left(\cdots c_c^\dagger c_k c_b^\dagger c_j c_a^\dagger c_i \right)^\dagger = \left(c_i^\dagger c_a c_j^\dagger c_b c_k^\dagger c_c \cdots\right) $ 
-is a de-excitation operator. Thus,
-\begin{equation}
- \bra{\Phi}\hat{\tau}_X = \hat{\tau}^\dagger_X \ket{\Phi} \equiv 0.
-\end{equation}
-\item Since all cluster operators commute, $e^{\hat{T}+\hat{T}^\prime} = e^{\hat{T}} e^{\hat{T}^\prime}$.
-\end{itemize}
-
-\section{Existence of cluster expansion.}
-Suppose $\braket{\Phi|\Psi} = 1$. Then there exists cluster operators $\hat{T}$ and $\hat{A}$ such that 
-\begin{equation}
- \ket{\Psi} = \left( 1 + \hat{A} \right) \ket{\Phi} = e^{\hat{T}} \ket{\Phi}.
-\end{equation}
-\subparagraph*{Proof:} Let $X$ denote a general excitation index and let $\tau_X$ be a general excitation operator. Assume that 
-$\braket{\Phi|\Psi} = 1$ and let $$\{ \ket{\Phi}, \ket{\Phi_X} | \text{ for all } X \}$$ be a Slater determinant basis where $\braket{\Phi|\Phi_X} = 0$. We can expand $\ket{\Psi}$ in the basis, giving
-\begin{equation}
- \ket{\Psi} = A_0 \ket{\Phi} + \sum_X A_X \ket{\Phi_X}, \ \ A_0, A_X \in \mathbb{C}.
-\end{equation}
-Note that $\braket{\Phi|\Psi} = 1$ implies that $A_0 = 1$. Using $\ket{\Phi_X} = \hat{\tau}_X \ket{\Phi}$ we get, 
-\begin{equation}
- \ket{\Psi} = \ket{\Phi} + \sum_X A_X \hat{\tau}_X \ket{\Phi} = \left(1 + \hat{A}\right) \ket{\Phi},
-\end{equation}
-where we have defined $\hat{A} \equiv \sum_X A_X \hat{\tau}_X$. We now want to find $\hat{T}$ such that $e^{\hat{T}} = 1 + \hat{A}$. Rewrite $\hat{A}$ 
-as,
-\begin{align*}
- \hat{A} = \sum_X A_X \hat{\tau}_X &= \sum_{ia} A_i^a \hat{\tau}_i^a + \frac{1}{2!^2}\sum_{ijab} A_{ij}^{ab} \hat{\tau}_{ij}^{ab} + \cdots \\
- &= \hat{A}_1 + \hat{A}_2 + \cdots
-\end{align*}
-By expanding $e^{\hat{T}}$ as a series
-\begin{equation}
- e^{\hat{T}} = 1 + \hat{T} + \frac{1}{2} \hat{T}^2 + \cdots \label{ExponantialExpansion},
-\end{equation}
-and note that $e^{\hat{T}} = 1 + \hat{A}$ if they have the same singles, doubles, etc. parts. Let $N$ be given and insert $\hat{T} = \hat{T}_1 + \cdots + \hat{T}_N$
-into eq. \eqref{ExponantialExpansion}, expand each power and group terms of equal rank together,
-\begin{align*}
- e^{\hat{T}} &= 1 + \hat{T} + \frac{1}{2} \hat{T}^2 + \frac{1}{6} \hat{T}^3 + \cdots \\
- &= 1 + \left(\hat{T}_1 + \cdots + \hat{T}_N \right) + \frac{1}{2} \left( \hat{T}_1 + \cdots + \hat{T}_N \right)^2 \\
- &+ \frac{1}{6} \left( \hat{T}_1 + \cdots + \hat{T}_N \right)^3 + \frac{1}{24} \left( \hat{T}_1 + \cdots + \hat{T}_N \right)^4 + \cdots \\
- &= 1 + \underbrace{\hat{T}_1}_{\text{rank} 1} + \underbrace{\left( \hat{T}_2 + \frac{1}{2}\hat{T}_1^2 \right)}_{\text{rank} 2} + \underbrace{\left( \hat{T}_3 + \frac{1}{6} \hat{T}_1^3 + \hat{T}_1 \hat{T}_2 \right)}_{\text{rank} 3} + \cdots \\
-\end{align*}
-Note that in the last equation, the rank $k$ term contains only one term from $\hat{T}_k$ while all other terms are products of cluster operators of lower rank.
-For the above expression to be equal to $1 + \hat{A}_1 + \hat{A}_2 + \cdots$ each rank $k$ term must equal $\hat{A}_k$.
-\begin{align*}
- \hat{A}_1 = &\hat{T}_1 \\
- \hat{A}_2 = &\hat{T}_2 + \frac{1}{2}\hat{T}_1^2 \\
- \hat{A}_3 = &\hat{T}_3 + \frac{1}{6} \hat{T}_1^3 + \hat{T}_1 \hat{T}_2 \\
-  &\vdots
-\end{align*}
-Then we can find $\hat{T}_k$ recursively, 
-\begin{align}
- \hat{T}_1 = &\hat{A}_1 \\
- \hat{T}_2 = &\hat{A}_2 - \frac{1}{2}\hat{T}_1^2 = \hat{A}_2 - \frac{1}{2}\hat{A}_1^2  \\
- \hat{T}_3 = &\hat{A}_3 - \frac{1}{6} \hat{T}_1^3 - \hat{T}_1 \hat{T}_2 \\
-           = &\hat{A}_3 - \frac{1}{6} \hat{A}_1^3 - \hat{A}_1 \left( \hat{A}_2 - \frac{1}{2}\hat{A}_1^2 \right) \\
- & \vdots
-\end{align}
-Hence, given $\ket{\Psi}$, we can find $\hat{A}$ which in turn gives $\hat{T}$.
-
-We notice that the expansion $\ket{\Psi} = (1+\hat{A}) \ket{\Phi}$ actually is the CI expansion. Thus, the 
-CC expansion $e^{\hat{T}}\ket{\Phi}$ is a nonlinear reparametrization of the linear CI parametrization. In practice, for a system containing $N$ particles, 
-one must truncate $\hat{T} = \hat{T}_1 + \cdots + \hat{T}_N$ and $\hat{A} = \hat{A}_1 + \cdots + \hat{A}_N$. Now, truncate the expansion at the 
-$k$-th operator in the sum with $1 \leq k \leq N$ such that $\hat{T} = \hat{T}_1 + \cdots + \hat{T}_k$ and $\hat{A} = \hat{A}_1 + \cdots + \hat{A}_k$. 
-Observe that 
-\begin{align*}
-\ket{\Psi} &= (1+\hat{A}_1 + \cdots + \hat{A}_k) \ket{\Phi} \\
-\end{align*}
-is a linear combination where the highest excitation contribution has rank $k$. The exponential ansatz on the other hand 
-\begin{align*}
- e^{\hat{T}_1 + \cdots + \hat{T}_k} \ket{\Phi} = \left( 1 + \left( \hat{T}_1 + \cdots + \hat{T}_k \right) + \frac{1}{2} \left( \hat{T}_1 + \cdots + \hat{T}_k \right)^2 + \cdots \right) \ket{\Phi}
-\end{align*}
-contains contributions from determinants of higher excitation than rank $k$ which is immediately clear from the inclusion of the term $\hat{T}_k^2$ 
-in the second parenthesis in the above expression. It turns out that this gives the truncated CC ansatz an advantage over 
-the truncated CI ansatz. The latter is not size-consistent and extensive while the former is.
-
-Loosely speaking size-consistency and extensivity 
-means that if $A$ and $B$ are two noninteracting subsystems, we should get the same energy for the supersystem $AB$ irrespective of whether we have
-carried out the calculations for each subsystem separately or for both subsystems simultaneously. In general 
-truncated exponential wavefunctions have this property while truncated linear wavefunctions does not\cite{Helgaker00book}.
-
-Depending on the truncation we get a sequence of more and more
-accurate CC approximations:
-\begin{itemize}
- \item $\ket{\Psi_{CCS}} = e^{\hat{T}_1} \ket{\Phi}$ coupled-cluster singles (CCS)
- \item $\ket{\Psi_{CCD}} = e^{\hat{T}_2} \ket{\Phi}$ coupled-cluster doubles (CCD)
- \item $\ket{\Psi_{CCSD}} = e^{\hat{T}_1 + \hat{T}_2} \ket{\Phi}$ coupled-cluster singles doubles (CCSD)
- \item $\ket{\Psi_{CCSDT}} = e^{\hat{T}_1 + \hat{T}_2 + \hat{T}_3} \ket{\Phi}$ coupled-cluster singles doubles triples (CCSDT)
-\end{itemize}
-
-\section{Equality of CC and CI for $N=2$.}
-
-An important special case is the case $N=2$. In that case the CC
-expansion equals the CI expansion. This is invaluable information when
-we want to implement the CC method since we can then compare it with
-the corresponding CI method to verify the implementation. To see this
-let $N=2$ and let $\ket{\Phi}$ be a reference determinant as
-usual. Observe that $\hat{A}_3 \ket{\Phi} = 0 $ and $\hat{T}_3 =
-0$. Thus, $\hat{A} = \hat{A}_1 + \hat{A}_2$ and $\hat{T} = \hat{T}_1 +
-\hat{T}_2$ will give all non-zero contributions. Additionaly all
-products of cluster operators where the product has rank larger than
-$2$, acting on the reference will give a zero result.  Inserting
-$\hat{T}_1 = \hat{A}_1$ and $\hat{T}_2 = \hat{A}_2 -
-\frac{1}{2}\hat{A}_1^2$ into the CC expansionincluding only terms of
-at most rank $2$ we obtain,
- \begin{align*}
-  e^{\hat{T}_1 + \hat{T}_2} \ket{\Phi} &= \left( 1 +
-  \underbrace{\hat{T}_1}_{\text{rank} 1} + \underbrace{\left(
-    \hat{T}_2 + \frac{1}{2}\hat{T}_1^2 \right)}_{\text{rank} 2}
-  \right) \ket{\Phi} \\ &= \left(1+ \hat{A}_1 + \hat{A}_2 -
-  \frac{1}{2}\hat{A}_1^2 + \frac{1}{2} \hat{A}_1^2 \right) \ket{\Phi}
-  \\ &= \left(1+ \hat{A}_1 + \hat{A}_2 \right) \ket{\Phi}
- \end{align*}
-which indeed is the CI expansion. For $N=2$, $\hat{A} = \hat{A}_1 +
-\hat{A}_2$ corresponds to a full configuration interaction treatment,
-while $\hat{T} = \hat{T}_1 + \hat{T}_2$ CCSD, but in this case CCSD is
-actually equal to FCI. Morover, if we consider only $\hat{T} =
-\hat{T}_2$ (CCD), we get
-\begin{align*}
- \ket{\Psi}_{\text{CCD}} = e^{\hat{T}_2} \ket{\Phi} &= \left(1+
- \hat{T}_2 \right) \ket{\Phi} \\ &= \left( 1 + \hat{A}_2 -
- \frac{1}{2}\hat{A}_1^2 + \frac{1}{2} \hat{A}_1^2 \right) \ket{\Phi}
- \\ &= \left(1 + \hat{A}_2 \right) \ket{\Phi} =
- \ket{\Psi}_{\text{CID}}.
-\end{align*}
-The importance of these two results is that one can verify a CCD
-implementation by comparing with a CID implementation and a CCSD
-implementation can be verified by comparing with a FCI/CISD
-implementation.
-
-\section{Amplitude equations}
-Recall that the Hamiltonian in second quantization is given by,
-\begin{equation}
- \hat{H} = \hat{H}^{(1)} + \hat{H}^{(2)} = \sum_{pq} h^p_q c_p^\dagger c_q + \frac{1}{4} \sum_{pqrs} u^{pr}_{qs} c_p^\dagger c_r^\dagger c_s c_q
-\end{equation}
-where $ u^{pr}_{qs} = \braket{\varphi_p
-  \varphi_r|\hat{H}^{(2)}|\varphi_q \varphi_s}_{AS}$.  Evaluation of
-the right hand side of equation \eqref{ExpecTamplitude} gives,
-\begin{equation}
- \frac{\partial}{\partial \lambda_{ij}^{ab}}E_{H^{(1)}} = h^{k}_{i}
- \tau^{ab}_{jk} P(ij) + h^{a}_{c} \tau^{bc}_{ji}
- P(ab) \label{AlgebraicTampH1},
-\end{equation}
-\begin{align}
- \frac{\partial}{\partial \lambda_{ab}^{ij}}E_{H^{(2)}} =
- &-\tau^{ab}_{ik} u^{kl}_{jl} P(ij) + \tau^{ab}_{il} \chi_j^l P(ij)
- \nonumber \\ &+\tau^{ab}_{kl} \chi^{kl}_{ij} +\tau^{ac}_{ij}\chi_c^b
- P(ab) \nonumber \\ &+\tau^{ac}_{ik} \chi_{cj}^{kb}P(ab)P(ij)
- -\frac{1}{2}\tau^{dc}_{ji} u^{ab}_{dc} \nonumber
- \\ &+u^{ab}_{ij} \label{AlgebraicTampH2},
-\end{align}
-where $P(ij)$ is an anti-symmtrizer in the sense that $f(ij)P(ij) =
-f(ij)-f(ji)$ and similarly for $P(ab)$. We have defined the so-called
-intermediates,
-\begin{align}
- \chi^{kl}_{ij} &= \frac{1}{4} \tau^{cd}_{ij} u^{kl}_{cd} + \frac{1}{2} u^{kl}_{ij} \label{Xklij} \\
- \chi_c^b &= \frac{1}{2} \tau^{bd}_{kl}  u^{kl}_{dc}  +  u^{bk}_{ck} \label{Xcb} \\
- \chi_j^l &= \frac{1}{2}\tau^{dc}_{jk} u^{kl}_{dc} \label{Xjl} \\
- \chi_{cj}^{kb} &= u^{kb}_{cj} + \frac{1}{2} u^{kl}_{cd}\tau^{bd}_{jl} \label{Xcjkb} .
-\end{align}
-We use the Einstein summation convention over repeated indices of
-opposite vertical placement. The appearance of a $P(ij)$ or a $P(ab)$
-should be ignored for the invocation of the summation convention. This
-means that for example the expression $h^{k}_{i} \tau^{ab}_{jk} P(ij)$
-is interpreted as,
-\begin{equation*}
- h^{k}_{i} \tau^{ab}_{jk} P(ij) = \sum_k \left(h^{k}_{i} \tau^{ab}_{jk} - h^{k}_{j} \tau^{ab}_{ik} \right).
-\end{equation*}
-In other words, we sum over all indices that does not appear on the
-left-hand side of all following expressions.
-
-Evaluation of the right hand side of equation \eqref{ExpexLamplitude}
-results in,
-\begin{align}
- \frac{\partial}{\partial \tau_{ij}^{ab}}E_{H^{(1)}} = h^{c}_{a} \lambda^{ji}_{bc} P(ab) + h^{i}_{k} \lambda^{jk}_{ab} P(ij) \label{AlgebraicLampH1},
-\end{align}
-\begin{align}
- \frac{\partial}{\partial \tau_{ij}^{ab}}E_{H^{(2)}} = &\lambda^{kl}_{ac} \xi^{cji}_{klb} P(ab) +\lambda^{ik}_{dc} \xi^{dcj}_{kab}  P(ij) \nonumber \\
- +&\lambda^{kl}_{ab}\xi_{kl}^{ij} +\lambda^{ik}_{ac}\xi_{kb}^{cj} P(ab) P(ij) \nonumber \\
- +&\lambda^{ik}_{ab}\xi_k^j P(ij) +\lambda^{ji}_{dc}\xi_{ab}^{dc} \nonumber \\
- +&\lambda^{ji}_{ac} \xi_b^c P(ab) -u^{ji}_{ab} \label{AlgebraicLampH2},
-\end{align}
-where we have defined the intermediates,
-\begin{align}
- \xi^{cji}_{klb}      &= - \frac{1}{2}\tau^{dc}_{kl} u^{ji}_{bd} \label{Ecjiklb} \\
- \xi^{dcj}_{kab}            &= - \frac{1}{2}\tau^{dc}_{kl} u^{jl}_{ab} \label{Edcjkab} \\
- \xi_{kl}^{ij} &= -\frac{1}{2} u^{ji}_{kl}  - \frac{1}{4} \tau^{dc}_{kl}u^{ji}_{dc} \label{Eklij} \\
- \xi_{kb}^{cj} &= - \tau^{dc}_{kl} u^{jl}_{bd} +  u^{jc}_{bk} \\
- \xi_k^j &= -\frac{1}{2} \tau^{dc}_{kl} u^{jl}_{dc}  -  u^{jl}_{kl} \\
- \xi_{ab}^{dc} &= -\frac{1}{4}\tau^{dc}_{kl}u^{kl}_{ab} - \frac{1}{2}u^{dc}_{ab} \\
- \xi_b^c &= -\frac{1}{2} \tau^{dc}_{kl} u^{kl}_{bd} - u^{ck}_{bk} \label{Ebc}.
-\end{align}
-If we precompute and store all intermediates equation \eqref{AlgebraicTampH2} scale as 
-$$\mathcal{O}\left(N^2(L-N)^4\right)$$ 
-while 
-equation \eqref{AlgebraicLampH2} scale as
-$$\mathcal{O}\left(N^3(L-N)^4\right).$$
-\section{Density matrices}
-For the onebody density matrix we get
-\begin{align}
- \rho_i^j &= \delta_{i j} - \frac{1}{2}\lambda^{kj}_{ab} \tau^{ab}_{ki}, \label{RHOij} \\
- \rho_a^b &= \frac{1}{2}\lambda^{ij}_{ac} \tau^{bc}_{ij}, \\
- \rho_i^a &= \rho_a^i = 0.
-\end{align}
-The non-zero elements of the two-body density matrix are given by,
-\begin{align}
- \rho_{ij}^{kl} &= - \delta_{i l} \delta_{j k} P(ij) - \frac{1}{2}P(ij) P(kl)\delta_{j l}\lambda^{km}_{dc} \tau^{dc}_{im}   + \frac{1}{2}\lambda^{lk}_{dc} \tau^{dc}_{ji}, \\
- \rho_{ab}^{ij} &= \lambda^{ij}_{ab} \\
- \rho_{ia}^{jb} &= \frac{1}{2}\delta_{i j} \lambda^{kl}_{ac} \tau^{bc}_{kl}  - \lambda^{jk}_{ac} \tau^{bc}_{ik} = -\rho_{ia}^{bj} = -\rho_{ai}^{jb} = \rho_{ai}^{bj} \\
- \rho_{ab}^{cd} &= \frac{1}{2}\lambda^{ij}_{ab} \tau^{cd}_{ij} \\
- \rho_{ij}^{ab} &= - \frac{1}{4} \lambda^{kl}_{dc} \tau^{ab}_{ik} \tau^{dc}_{jl} P(ab) P(ij) - \frac{1}{4} \lambda^{kl}_{dc} \tau^{ab}_{kl} \tau^{dc}_{ji} - \frac{1}{4} \lambda^{kl}_{dc} \tau^{ac}_{kl} \tau^{bd}_{ji} P(ij) \nonumber \\
- &+ \lambda^{kl}_{dc} \tau^{ac}_{il} \tau^{bd}_{jk} P(ij) - \frac{1}{4} \lambda^{kl}_{dc} \tau^{ac}_{ji} \tau^{bd}_{kl} P(ij) - \tau^{ab}_{ji} \label{RHOijab}.
-\end{align}
-
-\end{appendices}
-\end{comment}
 \bibliographystyle{ieeetr}

### 2019-01-16 doc/MachineLearning/article1.tex
update on article

\documentclass[aip,jcp,reprint,floatfix]{revtex4-1}

\usepackage{amsmath,amssymb}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{dcolumn}
\usepackage{bbold}
\usepackage{natbib}
\usepackage[normalem]{ulem} % to strike out plain text
\usepackage[colorlinks=true,citecolor={blue},urlcolor={blue}, linkcolor=blue]{hyperref}%
\usepackage[utf8]{inputenc}
\usepackage{amsthm}
\usepackage{braket}
\usepackage{verbatim}
\usepackage{mdframed}
\newtheorem{theorem}{Theorem}

\begin{document}

\title{Machine Learning and the Many-body Problem}

\author{Jane Kim}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}

\author{Vilde Moe Flugsrud}
\affiliation{Department of Physics, University of Oslo, N-0316 Oslo, Norway}

\author{Alfred Alocias Mariadason}
\affiliation{Department of Physics, University of Oslo, N-0316 Oslo, Norway}

\author{Morten Hjorth-Jensen}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\affiliation{Department of Physics and Center for Computing in Science Education, University of Oslo, N-0316 Oslo, Norway}

\begin{abstract}

\end{abstract}

\pacs{32.80.Rm, 31.15.A-, 42.65.Ky}

\maketitle

\section{Introduction}

\end{document}

%%
%% Automatically generated file from DocOnce source
%% (https://github.com/hplgit/doconce/)
%%
%%

%-------------------- begin preamble ----------------------

\documentclass[%
oneside,                 % oneside: electronic viewing, twoside: printing
final,                   % draft: marks overfull hboxes, figures with paths
10pt]{article}

\listfiles               %  print all files needed to compile this document

\usepackage{relsize,makeidx,color,setspace,amsmath,amsfonts,amssymb}
\usepackage[table]{xcolor}
\usepackage{bm,ltablex,microtype}

\usepackage[pdftex]{graphicx}

\usepackage{fancyvrb} % packages needed for verbatim environments
\usepackage{minted}
\usemintedstyle{default}

\usepackage[T1]{fontenc}
%\usepackage[latin1]{inputenc}
\usepackage{ucs}
\usepackage[utf8x]{inputenc}

\usepackage{lmodern}         % Latin Modern fonts derived from Computer Modern

% Hyperlinks in PDF:
\definecolor{linkcolor}{rgb}{0,0,0.4}
\usepackage{hyperref}
\hypersetup{
    breaklinks=true,
    colorlinks=true,
    linkcolor=linkcolor,
    urlcolor=linkcolor,
    citecolor=black,
    filecolor=black,
    %filecolor=blue,
    pdfmenubar=true,
    pdftoolbar=true,
    bookmarksdepth=3   % Uncomment (and tweak) for PDF bookmarks with more levels than the TOC
    }
%\hyperbaseurl{}   % hyperlinks are relative to this root

\setcounter{tocdepth}{2}  % levels in table of contents

% Tricks for having figures close to where they are defined:
% 1. define less restrictive rules for where to put figures
\setcounter{topnumber}{2}
\setcounter{bottomnumber}{2}
\setcounter{totalnumber}{4}
\renewcommand{\topfraction}{0.95}
\renewcommand{\bottomfraction}{0.95}
\renewcommand{\textfraction}{0}
\renewcommand{\floatpagefraction}{0.75}
% floatpagefraction must always be less than topfraction!
% 2. ensure all figures are flushed before next section
\usepackage[section]{placeins}
% 3. enable begin{figure}[H] (often leads to ugly pagebreaks)
%\usepackage{float}\restylefloat{figure}

% --- fancyhdr package for fancy headers ---
\usepackage{fancyhdr}
\fancyhf{} % sets both header and footer to nothing
\renewcommand{\headrulewidth}{0pt}
\fancyfoot[LE,RO]{\thepage}
% Ensure copyright on titlepage (article style) and chapter pages (book style)
\fancypagestyle{plain}{
  \fancyhf{}
  \fancyfoot[C]{{\footnotesize \copyright\ 1999-2019, Morten Hjorth-Jensen. Released under CC Attribution-NonCommercial 4.0 license}}
%  \renewcommand{\footrulewidth}{0mm}
  \renewcommand{\headrulewidth}{0mm}
}
% Ensure copyright on titlepages with \thispagestyle{empty}
\fancypagestyle{empty}{
  \fancyhf{}
  \fancyfoot[C]{{\footnotesize \copyright\ 1999-2019, Morten Hjorth-Jensen. Released under CC Attribution-NonCommercial 4.0 license}}
  \renewcommand{\footrulewidth}{0mm}
  \renewcommand{\headrulewidth}{0mm}
}

\pagestyle{fancy}

\usepackage[framemethod=TikZ]{mdframed}

% --- begin definitions of admonition environments ---

% --- end of definitions of admonition environments ---

% prevent orhpans and widows
\clubpenalty = 10000
\widowpenalty = 10000

% --- end of standard preamble for documents ---

% insert custom LaTeX commands...

\raggedbottom
\makeindex
\usepackage[totoc]{idxlayout}   % for index in the toc
\usepackage[nottoc]{tocbibind}  % for references/bibliography in the toc

%-------------------- end preamble ----------------------

\begin{document}

% matching end for #ifdef PREAMBLE

\newcommand{\exercisesection}[1]{\subsection*{#1}}

% ------------------- main content ----------------------

% ----------------- title -------------------------

\thispagestyle{empty}

\begin{center}
{\LARGE\bf
\begin{spacing}{1.25}
Solving quantum mechanical problems with Machine Learning
\end{spacing}
}
\end{center}

% ----------------- author(s) -------------------------

\begin{center}
{\bf Morten Hjorth-Jensen${}^{1, 2}$} \\ [0mm]
\end{center}

\begin{center}
% List of all institutions:
\centerline{{\small ${}^1$Department of Physics and Astronomy and National Superconducting Cyclotron Laboratory, Michigan State University, USA}}
\centerline{{\small ${}^2$Department of Physics and Center for Computing in Science Education, University of Oslo, Norway}}
\end{center}

% ----------------- end author(s) -------------------------

% --- begin date ---
\begin{center}
Jan 16, 2019 
\end{center}
% --- end date ---

\vspace{1cm}

% !split
\subsection*{What is this talk about?}

% --- begin paragraph admon ---
\paragraph{}
The main aim is to give you a short and pedestrian introduction to  how we can use Machine Learning methods
to solve quantum mechanical many-body problems. And why this could be of interest.

The hope is that after this talk you have gotten the basic ideas to get you started. Peeping into \href{{https://github.com/mhjensenseminars/MachineLearningTalk}}{\nolinkurl{https://github.com/mhjensenseminars/MachineLearningTalk}}, you'll find a Jupyter notebook, slides, codes etc that will allow you to reproduce the simulations discussed here, and perhaps run your own very first calculations.

Try it out and please don't hesitate to swing by if something is unclear.
% --- end paragraph admon ---

% !split
\subsection*{Why?}

How can we avoid the dimensionality curse? Many possibilities
\begin{enumerate}
\item smarter basis functions

\item resummation of specific correlations

\item stochastic sampling of high-lying states (stochastic FCI, CC and SRG/IMSRG)

\item many more
\end{enumerate}

\noindent
Machine Learning and Quantum Computing hold also great promise in tackling the 
ever increasing dimensionalities. Here we will focus on Machine Learning.

% !split
\subsection*{Overview}

% --- begin paragraph admon ---
\paragraph{}
\begin{itemize}
\item Short intro to Machine Learning

\item Variational Monte Carlo (Markov Chain Monte Carlo, $\mathrm{MC}^2$) and two-electron quantum dots, solving quantum mechanical problems in a stochastic way. It  will serve as  our motivation for switching to Machine Learning.

\item From Variational Monte Carlo to Boltzmann Machines and Machine Learning
\end{itemize}

\noindent
% --- end paragraph admon ---

% !split
\subsection*{What are the Machine Learning calculations here based on?}

This work is inspired by the idea of representing the wave function with
a restricted Boltzmann machine (RBM), presented recently by \href{{http://science.sciencemag.org/content/355/6325/602}}{G. Carleo and M. Troyer, Science \textbf{355}, Issue 6325, pp. 602-606 (2017)}. They
named such a wave function/network a \emph{neural network quantum state} (NQS). In their article they apply it to the quantum mechanical
spin lattice systems of the Ising model and Heisenberg model, with
encouraging results.

Thanks to Jane Kim (MSU), Vilde Flugsrud (UiO), Alfred Alocias Mariadason (UiO) for many discussions and interpretations.

% !split
\subsection*{A new world}

% --- begin paragraph admon ---
\paragraph{}
Machine learning (ML) is an extremely rich field, in spite of its young age. The
increases we have seen during the last three decades in computational
capabilities have been followed by developments of methods and
techniques for analyzing and handling large date sets, relying heavily
on statistics, computer science and mathematics.  The field is rather
new and developing rapidly.

Popular software packages written in Python for ML are

\begin{itemize}
\item \href{{http://scikit-learn.org/stable/}}{Scikit-learn},

\item \href{{https://www.tensorflow.org/}}{Tensorflow},

\item \href{{http://pytorch.org/}}{PyTorch} and

\item \href{{https://keras.io/}}{Keras}.
\end{itemize}

\noindent
These are all freely available at their respective GitHub sites. They 
encompass communities of developers in the thousands or more. And the number
of code developers and contributors keeps increasing.
% --- end paragraph admon ---

% !split
\subsection*{Lots of room for creativity}

% --- begin paragraph admon ---
\paragraph{}
Not all the
algorithms and methods can be given a rigorous mathematical
justification, opening up thereby for experimenting
and trial and error and thereby exciting new developments.
% --- end paragraph admon ---

% --- begin paragraph admon ---
\paragraph{}
A solid command of linear algebra, multivariate theory, 
probability theory, statistical data analysis,
understanding errors and Monte Carlo methods is important in order to understand many of the 
various algorithms and methods.
% --- end paragraph admon ---

\textbf{Job market, a personal statement}: \href{{https://www.analyticsindiamag.com/top-countries-hiring-most-number-of-artificial-intelligence-machine-learning-experts/}}{A familiarity with ML is almost becoming a prerequisite for many of the most exciting employment opportunities}. And add quantum computing and there you are!

% !split
\subsection*{Knowledge of Statistical analysis and optimization of data}
Some key elements that enter much of the discussion on ML:
\begin{enumerate}
\item Basic concepts, expectation values, variance, covariance, correlation functions and errors;

\item Simpler models, binomial distribution, the Poisson distribution, simple and multivariate normal distributions;

\item Central elements of Bayesian statistics and modeling;

\item Central elements from linear algebra

\item Gradient methods for data optimization

\item Monte Carlo methods, Markov chains, Metropolis-Hastings algorithm;

\item Estimation of errors using cross-validation, blocking, bootstrapping and jackknife methods;

\item Practical optimization using Singular-value decomposition and least squares for parameterizing data.

\item Principal Component Analysis.
\end{enumerate}

\noindent
% !split 
\subsection*{Some members of the ML family}

% --- begin paragraph admon ---
\paragraph{}
\begin{enumerate}
\item Linear regression and its variants, Logistic regression

\item Decision tree algorithms, from simpler to more complex ones like random forests

\item Bayesian statistics

\item Support vector machines and finally various variants of

\item Artifical neural networks and deep learning

\item Convolutional NN, autoenconders

\item and many more
\end{enumerate}

\noindent
% --- end paragraph admon ---

% !split
\subsection*{What are the basic ingredients?}

% --- begin paragraph admon ---
\paragraph{}
Almost every problem in ML and data science starts with the same ingredients:
\begin{itemize}
\item The dataset $\mathbf{x}$ (could be some observable quantity of the system we are studying)

\item A model which is a function of a set of parameters $\mathbf{\alpha}$ that relates to the dataset, say a likelihood  function $p(\mathbf{x}\vert \mathbf{\alpha})$ or just a simple model $f(\mathbf{\alpha})$

\item A so-called \textbf{cost} function $\mathcal{C} (\mathbf{x}, f(\mathbf{\alpha}))$ which allows us to decide how well our model represents the dataset. 
\end{itemize}

\noindent
We seek to minimize the function $\mathcal{C} (\mathbf{x}, f(\mathbf{\alpha}))$ by finding the parameter values which minimize $\mathcal{C}$. This leads to  various minimization algorithms.
% --- end paragraph admon ---

% !split
\subsection*{What is Machine Learning?}

% --- begin paragraph admon ---
\paragraph{}
Machine learning is the science of giving computers the ability to
learn without being explicitly programmed.  The idea is that there
exist generic algorithms which can be used to find patterns in a broad
class of data sets without having to write code specifically for each
problem. The algorithm will build its own logic based on the data.

Machine learning is a subfield of computer science, and is closely
related to computational statistics.  It evolved from the study of
pattern recognition in artificial intelligence (AI) research, and has
made contributions to AI tasks like computer vision, natural language
processing and speech recognition. It has also, especially in later
years, found applications in a wide variety of other areas, including
bioinformatics, economy, physics, finance and marketing.

You will notice however that many of the basic ideas discussed do come from Physics!
% --- end paragraph admon ---

% !split
\subsection*{Types of Machine Learning}

% --- begin paragraph admon ---
\paragraph{}
The approaches to machine learning are many, but are often split into two main categories. 
In \emph{supervised learning} we know the answer to a problem,
and let the computer deduce the logic behind it. On the other hand, \emph{unsupervised learning}
is a method for finding patterns and relationship in data sets without any prior knowledge of the system.
Some authours also operate with a third category, namely \emph{reinforcement learning}. This is a paradigm 
of learning inspired by behavioural psychology, where learning is achieved by trial-and-error, 
solely from rewards and punishment.

Another way to categorize machine learning tasks is to consider the desired output of a system.
Some of the most common tasks are:

\begin{itemize}
  \item Classification: Outputs are divided into two or more classes. The goal is to   produce a model that assigns inputs into one of these classes. An example is to identify  digits based on pictures of hand-written ones. Classification is typically supervised learning.

  \item Regression: Finding a functional relationship between an input data set and a reference data set.   The goal is to construct a function that maps input data to continuous output values.

  \item Clustering: Data are divided into groups with certain common traits, without knowing the different groups beforehand.  It is thus a form of unsupervised learning.
\end{itemize}

\noindent
% --- end paragraph admon ---

% !split
\subsection*{References}

% --- begin paragraph admon ---
\paragraph{}
\begin{itemize}
\item An excellent reference, \href{{https://arxiv.org/abs/1803.08823}}{Mehta \emph{et al.}, arXiv:1803.08823 and Physics Reports in press (2018)}

\item A cute paper by \href{{https://journals.aps.org/prc/abstract/10.1103/PhysRevC.97.014306}}{Utama and Piekarewicz, Validating neural-network refinements of nuclear mass models, Phys. Rev. C 97, 014306}

\item \href{{https://journals.aps.org/prl/abstract/10.1103/PhysRevLett.120.156001}}{Every issue of Physical Review Letters has now one or more articles on ML}

\item \href{{https://github.com/CompPhysics/MachineLearning}}{Books and lectures notes} and see also the course \href{{https://www.uio.no/studier/emner/matnat/fys/FYS-STK4155/h18/index.html}}{FYS-STK3155/4155}

\item See also Metha and Schwab, arXiv.1410.3831, how to link Variational renormalization group theory with deep learning (recommended read)
\end{itemize}

\noindent
% --- end paragraph admon ---

% !split
\subsection*{Another interesting article}

Here we will use so-called \textbf{reduced Boltzmann Machines} to simulate quantum many-body problems. For Monte Carlo aficionados, there is a very close similarity with what are called \textbf{shadow wave functions}, see the work of \href{{https://journals.aps.org/pre/abstract/10.1103/PhysRevE.90.053304}}{Pederiva and Kalos and collaborators, Phys Rev. E 90, 053304 (2014)}.

% !split
\subsection*{Courses on Data science and Machine Learning  at UiO}

The \href{{https://www.mn.uio.no/english/research/about/centre-focus/innovation/data-science/studies/}}{link here}  gives an excellent overview of courses on Machine learning at UiO.

\begin{enumerate}
\item \href{{http://www.uio.no/studier/emner/matnat/math/STK2100/index-eng.html}}{STK2100 Machine learning and statistical methods for prediction and classification}.

\item \href{{https://www.uio.no/studier/emner/matnat/ifi/IN3050/index-eng.html}}{IN3050 Introduction to Artificial Intelligence and Machine Learning}. Introductory course in machine learning and AI with an algorithmic approach.

\item \href{{http://www.uio.no/studier/emner/matnat/math/STK-INF3000/index-eng.html}}{STK-INF3000/4000 Selected Topics in Data Science}. The course provides insight into selected contemporary relevant topics within Data Science.

\item \href{{https://www.uio.no/studier/emner/matnat/ifi/IN4080/index.html}}{IN4080 Natural Language Processing}. Probabilistic and machine learning techniques applied to natural language processing.

\item \href{{https://www.uio.no/studier/emner/matnat/math/STK-IN4300/index-eng.html}}{STK-IN4300 – Statistical learning methods in Data Science}. An advanced introduction to statistical and machine learning. For students with a good mathematics and statistics background.

\item \href{{http://www.uio.no/studier/emner/matnat/ifi/INF4490/}}{INF4490 Biologically Inspired Computing}. An introduction to self-adapting methods also called artificial intelligence or machine learning.

\item \href{{https://www.uio.no/studier/emner/matnat/ifi/IN-STK5000/index-eng.html}}{IN-STK5000  Adaptive Methods for Data-Based Decision Making}. Methods for adaptive collection and processing of data based on machine learning techniques.

\item \href{{https://www.uio.no/studier/emner/matnat/ifi/IN5400/}}{IN5400/INF5860 – Machine Learning for Image Analysis}. An introduction to deep learning with particular emphasis on applications within Image analysis, but useful for other application areas too.

\item \href{{https://www.uio.no/studier/emner/matnat/its/TEK5040/}}{TEK5040 – Deep learning}. The course addresses advanced algorithms and architectures for deep learning with neural networks. The course provides an introduction to how deep-learning techniques can be used in the construction of key parts of advanced autonomous systems that exist in physical environments and cyber environments.
\end{enumerate}

\noindent
% !split
\subsection*{Additional courses of interest}

\begin{enumerate}
\item \href{{https://www.uio.no/studier/emner/matnat/math/STK4051/index-eng.html}}{STK4051 Computational Statistics}

\item \href{{https://www.uio.no/studier/emner/matnat/math/STK4021/index-eng.html}}{STK4021 Applied Bayesian Analysis and Numerical Methods}
\end{enumerate}

\noindent
% !split
\subsection*{Decision trees and Regression}
\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=7mm]{python}
import numpy as np
import matplotlib.pyplot as plt
from sklearn.preprocessing import PolynomialFeatures
from sklearn.linear_model import LinearRegression

steps=250

distance=0
x=0
distance_list=[]
steps_list=[]
while x<steps:
    distance+=np.random.randint(-1,2)
    distance_list.append(distance)
    x+=1
    steps_list.append(x)
plt.plot(steps_list,distance_list, color='green', label="Random Walk Data")

steps_list=np.asarray(steps_list)
distance_list=np.asarray(distance_list)

X=steps_list[:,np.newaxis]

#Polynomial fits

#Degree 2
poly_features=PolynomialFeatures(degree=2, include_bias=False)
X_poly=poly_features.fit_transform(X)

lin_reg=LinearRegression()
poly_fit=lin_reg.fit(X_poly,distance_list)
b=lin_reg.coef_
c=lin_reg.intercept_
print ("2nd degree coefficients:")
print ("zero power: ",c)
print ("first power: ", b[0])
print ("second power: ",b[1])

z = np.arange(0, steps, .01)
z_mod=b[1]*z**2+b[0]*z+c

fit_mod=b[1]*X**2+b[0]*X+c
plt.plot(z, z_mod, color='r', label="2nd Degree Fit")
plt.title("Polynomial Regression")

plt.xlabel("Steps")
plt.ylabel("Distance")

#Degree 10
poly_features10=PolynomialFeatures(degree=10, include_bias=False)
X_poly10=poly_features10.fit_transform(X)

poly_fit10=lin_reg.fit(X_poly10,distance_list)

y_plot=poly_fit10.predict(X_poly10)
plt.plot(X, y_plot, color='black', label="10th Degree Fit")

plt.legend()
plt.show()

#Decision Tree Regression
from sklearn.tree import DecisionTreeRegressor
regr_1=DecisionTreeRegressor(max_depth=2)
regr_2=DecisionTreeRegressor(max_depth=5)
regr_3=DecisionTreeRegressor(max_depth=7)
regr_1.fit(X, distance_list)
regr_2.fit(X, distance_list)
regr_3.fit(X, distance_list)

X_test = np.arange(0.0, steps, 0.01)[:, np.newaxis]
y_1 = regr_1.predict(X_test)
y_2 = regr_2.predict(X_test)
y_3=regr_3.predict(X_test)

# Plot the results
plt.figure()
plt.scatter(X, distance_list, s=2.5, c="black", label="data")
plt.plot(X_test, y_1, color="red",
         label="max_depth=2", linewidth=2)
plt.plot(X_test, y_2, color="green", label="max_depth=5", linewidth=2)
plt.plot(X_test, y_3, color="m", label="max_depth=7", linewidth=2)

plt.xlabel("Data")
plt.ylabel("Darget")
plt.title("Decision Tree Regression")
plt.legend()
plt.show()

\end{minted}

% !split
\subsection*{Artificial neurons}

% --- begin paragraph admon ---
\paragraph{}

The field of artificial neural networks has a long history of
development, and is closely connected with the advancement of computer
science and computers in general. A model of artificial neurons was
first developed by McCulloch and Pitts in 1943 to study signal
processing in the brain and has later been refined by others. The
general idea is to mimic neural networks in the human brain, which is
composed of billions of neurons that communicate with each other by
sending electrical signals.  Each neuron accumulates its incoming
signals, which must exceed an activation threshold to yield an
output. If the threshold is not overcome, the neuron remains inactive,
i.e.~has zero output.

This behaviour has inspired a simple mathematical model for an artificial neuron.

\[
 y = f\left(\sum_{i=1}^n w_ix_i\right) = f(u)
\]
Here, the output $y$ of the neuron is the value of its activation function, which have as input
a weighted sum of signals $x_i, \dots ,x_n$ received by $n$ other neurons.
% --- end paragraph admon ---

% !split
\subsection*{A simple perceptron model}

\vspace{6mm}

% inline figure
\centerline{\includegraphics[width=0.9\linewidth]{figures/perceptron.png}}

\vspace{6mm}

% !split
\subsection*{Neural network types}

% --- begin paragraph admon ---
\paragraph{}
An artificial neural network (NN), is a computational model that consists of layers of connected neurons, or \emph{nodes}. 
It is supposed to mimic a biological nervous system by letting each neuron interact with other neurons
by sending signals in the form of mathematical functions between layers. 
A wide variety of different NNs have
been developed, but most of them consist of an input layer, an output layer and eventual layers in-between, called
\emph{hidden layers}. All layers can contain an arbitrary number of nodes, and each connection between two nodes
is associated with a weight variable.

\vspace{6mm}

% inline figure
\centerline{\includegraphics[width=0.6\linewidth]{figures/dnn.png}}

\vspace{6mm}
% --- end paragraph admon ---

% !split
\subsection*{The system: two electrons in a harmonic oscillator trap in two dimensions}

The Hamiltonian of the quantum dot is given by
\[ \hat{H} = \hat{H}_0 + \hat{V}, 
\]
where $\hat{H}_0$ is the many-body HO Hamiltonian, and $\hat{V}$ is the
inter-electron Coulomb interactions. In dimensionless units,
\[ \hat{V}= \sum_{i < j}^N \frac{1}{r_{ij}},
\]
with $r_{ij}=\sqrt{\mathbf{r}_i^2 - \mathbf{r}_j^2}$.

This leads to the  separable Hamiltonian, with the relative motion part given by ($r_{ij}=r$)
\[ 
\hat{H}_r=-\nabla^2_r + \frac{1}{4}\omega^2r^2+ \frac{1}{r},
\]
plus a standard Harmonic Oscillator problem  for the center-of-mass motion.
This system has analytical solutions in two and three dimensions (\href{{https://journals.aps.org/pra/abstract/10.1103/PhysRevA.48.3561}}{M. Taut 1993 and 1994}).

% !split
\subsection*{Quantum Monte Carlo Motivation}

% --- begin paragraph admon ---
\paragraph{}
Given a hamiltonian $H$ and a trial wave function $\Psi_T$, the variational principle states that the expectation value of $\langle H \rangle$, defined through 
\[
   \langle E \rangle =
   \frac{\int d\bm{R}\Psi^{\ast}_T(\bm{R})H(\bm{R})\Psi_T(\bm{R})}
        {\int d\bm{R}\Psi^{\ast}_T(\bm{R})\Psi_T(\bm{R})},
\]
is an upper bound to the ground state energy $E_0$ of the hamiltonian $H$, that is 
\[
    E_0 \le \langle H \rangle .
\]
In general, the integrals involved in the calculation of various  expectation values  are multi-dimensional ones. Traditional integration methods such as the Gauss-Legendre will not be adequate for say the  computation of the energy of a many-body system.
% --- end paragraph admon ---

% !split
\subsection*{Quantum Monte Carlo Motivation}

% --- begin paragraph admon ---
\paragraph{Basic steps.}
Choose a trial wave function
$\psi_T(\bm{R})$.
\[
   P(\bm{R},\bm{\alpha})= \frac{\left|\psi_T(\bm{R},\bm{\alpha})\right|^2}{\int \left|\psi_T(\bm{R},\bm{\alpha})\right|^2d\bm{R}}.
\]
This is our model, or likelihood/probability distribution function  (PDF). It depends on some variational parameters $\bm{\alpha}$.
The approximation to the expectation value of the Hamiltonian is now 
\[
   \langle E[\bm{\alpha}] \rangle = 
   \frac{\int d\bm{R}\Psi^{\ast}_T(\bm{R},\bm{\alpha})H(\bm{R})\Psi_T(\bm{R},\bm{\alpha})}
        {\int d\bm{R}\Psi^{\ast}_T(\bm{R},\bm{\alpha})\Psi_T(\bm{R},\bm{\alpha})}.
\]
% --- end paragraph admon ---

% !split
\subsection*{Quantum Monte Carlo Motivation}

% --- begin paragraph admon ---
\paragraph{Define a new quantity.}
\[
   E_L(\bm{R},\bm{\alpha})=\frac{1}{\psi_T(\bm{R},\bm{\alpha})}H\psi_T(\bm{R},\bm{\alpha}),
\]
called the local energy, which, together with our trial PDF yields
\[
  E[\bm{\alpha}]=\int P(\bm{R})E_L(\bm{R},\bm{\alpha}) d\bm{R}\approx \frac{1}{N}\sum_{i=1}^NE_L(\bm{R_i},\bm{\alpha})
\]
with $N$ being the number of Monte Carlo samples.
% --- end paragraph admon ---

% !split
\subsection*{Quantum Monte Carlo}

% --- begin paragraph admon ---
\paragraph{}
The Algorithm for performing a variational Monte Carlo calculations runs thus as this

\begin{itemize}
   \item Initialisation: Fix the number of Monte Carlo steps. Choose an initial $\bm{R}$ and variational parameters $\alpha$ and calculate $\left|\psi_T(\bm{R},\bm{\alpha})\right|^2$.

   \item Initialise the energy and the variance and start the Monte Carlo calculation by looping over trials.
\begin{itemize}

      \item Calculate  a trial position  $\bm{R}_p=\bm{R}+r*step$ where $r$ is a random variable $r \in [0,1]$.

      \item Metropolis algorithm to accept or reject this move  $w = P(\bm{R}_p,\bm{\alpha})/P(\bm{R},\bm{\alpha})$.

      \item If the step is accepted, then we set $\bm{R}=\bm{R}_p$.

      \item Update averages

\end{itemize}

\noindent
   \item Finish and compute final averages.
\end{itemize}

\noindent
Observe that the jumping in space is governed by the variable \emph{step}. This is often called brute-force sampling.
Need importance sampling to get more relevant sampling.
% --- end paragraph admon ---

% !split
\subsection*{The trial wave function}

% --- begin paragraph admon ---
\paragraph{}
We want to perform  a Variational Monte Carlo calculation of the ground state of two electrons in a quantum dot well with different oscillator energies, assuming total spin $S=0$.
Our trial wave function has the following form
\begin{equation}
   \psi_{T}(\bm{r}_1,\bm{r}_2) = 
   C\exp{\left(-\alpha_1\omega(r_1^2+r_2^2)/2\right)}
   \exp{\left(\frac{r_{12}}{(1+\alpha_2 r_{12})}\right)}, 
\label{eq:trial}
\end{equation}
where the $\alpha$s represent our variational parameters, two in this case.

Why does the trial function look like this? How did we get there? \textbf{This will be our main motivation} for switching to
Machine Learning.
% --- end paragraph admon ---

% !split
\subsection*{The correlation part of the wave function}

To find an ansatz for the correlated part of the wave function, it is useful to rewrite the two-particle
local energy in terms of the relative and center-of-mass motion.
Let us denote the distance between the two electrons as
$r_{12}$. We omit the center-of-mass motion since we are only interested in the case when
$r_{12} \rightarrow 0$. The contribution from the center-of-mass (CoM) variable $\bm{R}_{\mathrm{CoM}}$
gives only a finite contribution.
We focus only on the terms that are relevant for $r_{12}$ and for three dimensions. The relevant local energy becomes then
\[
\lim_{r_{12} \rightarrow 0}E_L(R)=
    \frac{1}{{\cal R}_T(r_{12})}\left(2\frac{d^2}{dr_{ij}^2}+\frac{4}{r_{ij}}\frac{d}{dr_{ij}}+
\frac{2}{r_{ij}}-\frac{l(l+1)}{r_{ij}^2}+2E
\right){\cal R}_T(r_{12}) = 0.
\]
Set $l=0$ and we have the so-called \textbf{cusp} condition
\[
\frac{d {\cal R}_T(r_{12})}{dr_{12}} = -\frac{1}{2(l+1)}
{\cal R}_T(r_{12})\qquad r_{12}\to 0
\]

% !split
\subsection*{Resulting ansatz}
The above  results in
\[
{\cal R}_T  \propto \exp{(r_{ij}/2)}, 
\]
for anti-parallel spins and 
\[
{\cal R}_T  \propto \exp{(r_{ij}/4)}, 
\]
for anti-parallel spins. 
This is the so-called cusp condition for the relative motion, resulting in a minimal requirement
for the correlation part of the wave fuction.
For general systems containing more than say two electrons, we have this
condition for each electron pair $ij$.

% !split
\subsection*{The VMC code}

\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=7mm]{python}
# Importing various packages
from math import exp, sqrt
from random import random, seed
import numpy as np
import matplotlib.pyplot as plt
from mpl_toolkits.mplot3d import Axes3D
from matplotlib import cm
from matplotlib.ticker import LinearLocator, FormatStrFormatter
import sys

#Trial wave function for quantum dots in two dims
def WaveFunction(r,alpha,beta):
    r1 = r[0,0]**2 + r[0,1]**2
    r2 = r[1,0]**2 + r[1,1]**2
    r12 = sqrt((r[0,0]-r[1,0])**2 + (r[0,1]-r[1,1])**2)
    deno = r12/(1+beta*r12)
    return exp(-0.5*alpha*(r1+r2)+deno)

#Local energy  for quantum dots in two dims, using analytical local energy
def LocalEnergy(r,alpha,beta):

    r1 = (r[0,0]**2 + r[0,1]**2)
    r2 = (r[1,0]**2 + r[1,1]**2)
    r12 = sqrt((r[0,0]-r[1,0])**2 + (r[0,1]-r[1,1])**2)
    deno = 1.0/(1+beta*r12)
    deno2 = deno*deno
    return 0.5*(1-alpha*alpha)*(r1 + r2) +2.0*alpha + 1.0/r12+deno2*(alpha*r12-deno2+2*beta*deno-1.0/r12)

# The Monte Carlo sampling with the Metropolis algo
def MonteCarloSampling():

    NumberMCcycles= 100000
    StepSize = 1.0
    # positions
    PositionOld = np.zeros((NumberParticles,Dimension), np.double)
    PositionNew = np.zeros((NumberParticles,Dimension), np.double)
    # seed for rng generator
    seed()
    # start variational parameter
    alpha = 0.9
    for ia in range(MaxVariations):
        alpha += .025
        AlphaValues[ia] = alpha
        beta = 0.2 
        for jb in range(MaxVariations):
            beta += .01
            BetaValues[jb] = beta
            energy = energy2 = 0.0
            DeltaE = 0.0
            #Initial position
            for i in range(NumberParticles):
                for j in range(Dimension):
                    PositionOld[i,j] = StepSize * (random() - .5)
            wfold = WaveFunction(PositionOld,alpha,beta)

            #Loop over MC MCcycles
            for MCcycle in range(NumberMCcycles):
                #Trial position
                for i in range(NumberParticles):
                    for j in range(Dimension):
                        PositionNew[i,j] = PositionOld[i,j] + StepSize * (random() - .5)
                wfnew = WaveFunction(PositionNew,alpha,beta)

                #Metropolis test to see whether we accept the move
                if random() < wfnew**2 / wfold**2:
                   PositionOld = PositionNew.copy()
                   wfold = wfnew
                   DeltaE = LocalEnergy(PositionOld,alpha,beta)
                energy += DeltaE
                energy2 += DeltaE**2

            #We calculate mean, variance and error ...
            energy /= NumberMCcycles
            energy2 /= NumberMCcycles
            variance = energy2 - energy**2
            error = sqrt(variance/NumberMCcycles)
            Energies[ia,jb] = energy    
    return Energies, AlphaValues, BetaValues

#Here starts the main program with variable declarations
NumberParticles = 2
Dimension = 2
MaxVariations = 10
Energies = np.zeros((MaxVariations,MaxVariations))
AlphaValues = np.zeros(MaxVariations)
BetaValues = np.zeros(MaxVariations)
(Energies, AlphaValues, BetaValues) = MonteCarloSampling()

# Prepare for plots
fig = plt.figure()
ax = fig.gca(projection='3d')
# Plot the surface.
X, Y = np.meshgrid(AlphaValues, BetaValues)
surf = ax.plot_surface(X, Y, Energies,cmap=cm.coolwarm,linewidth=0, antialiased=False)
# Customize the z axis.
zmin = np.matrix(Energies).min()
zmax = np.matrix(Energies).max()
ax.set_zlim(zmin, zmax)
ax.set_xlabel(r'$\alpha$')
ax.set_ylabel(r'$\beta$')
ax.set_zlabel(r'$\langle E \rangle$')
ax.zaxis.set_major_locator(LinearLocator(10))
ax.zaxis.set_major_formatter(FormatStrFormatter('%.02f'))
# Add a color bar which maps values to colors.
fig.colorbar(surf, shrink=0.5, aspect=5)
plt.show()

\end{minted}

% !split
\subsection*{Technical aspect, improvements and how to define the cost function}

% --- begin paragraph admon ---
\paragraph{}

\textbf{The above procedure is not the smartest one}. Looping over all variational parameters becomes expensive.
Also, we don't use importance sampling and optimizations of the standard deviation (blocking, bootstrap, jackknife). 
Such codes are included in the above Github address.

We can also be smarter and use minimization methods to find the \textbf{optimal} variational parameters with fewer Monte Carlo cycles and then 
fire up our heavy artillery.

One way to achieve this is to minimize the energy as function of the variational parameters.
% --- end paragraph admon ---

% !split
\subsection*{Energy derivatives}

% --- begin paragraph admon ---
\paragraph{}
To find the derivatives of the local energy expectation value as function of the variational parameters, we can use the chain rule and the hermiticity of the Hamiltonian.

Let us define 
\[
\bar{E}_{\alpha_i}=\frac{d\langle  E_L\rangle}{d\alpha_i}.
\]
as the derivative of the energy with respect to the variational parameter $\alpha_i$
We define also the derivative of the trial function (skipping the subindex $T$) as 
\[
\bar{\Psi}_{i}=\frac{d\Psi}{d\alpha_i}.
\]
% --- end paragraph admon ---

% !split
\subsection*{Derivatives of the local energy}

% --- begin paragraph admon ---
\paragraph{}
The elements of the gradient of the local energy are then (using the chain rule and the hermiticity of the Hamiltonian)
\[
\bar{E}_{i}= 2\left( \langle \frac{\bar{\Psi}_{i}}{\Psi}E_L\rangle -\langle \frac{\bar{\Psi}_{i}}{\Psi}\rangle\langle E_L \rangle\right).
\]
From a computational point of view it means that you need to compute the expectation values of 
\[
\langle \frac{\bar{\Psi}_{i}}{\Psi}E_L\rangle,
\]
and
\[
\langle \frac{\bar{\Psi}_{i}}{\Psi}\rangle\langle E_L\rangle
\]
These integrals are evaluted using MC intergration (with all its possible error sources). 
We can then use methods like stochastic gradient or other minimization methods to find the optimal variational parameters (I don't discuss this topic here, but these methods are very important in ML).
% --- end paragraph admon ---

% !split
\subsection*{How do we define our cost function?}

% --- begin paragraph admon ---
\paragraph{}
We have a model, our likelihood function.

How should we define the cost function?
% --- end paragraph admon ---

% !split
\subsection*{Meet the variance and its derivatives}

% --- begin paragraph admon ---
\paragraph{Why the variance?}
Suppose the trial function (our model) is the exact wave function. The action of the hamiltionan
on the wave function
\[
   H\Psi = \mathrm{constant}\times \Psi,
\]
The integral which defines various 
expectation values involving moments of the hamiltonian becomes then
\[
   \langle E^n \rangle =   \langle H^n \rangle =
   \frac{\int d\bm{R}\Psi^{\ast}(\bm{R})H^n(\bm{R})\Psi(\bm{R})}
        {\int d\bm{R}\Psi^{\ast}(\bm{R})\Psi(\bm{R})}=
\mathrm{constant}\times\frac{\int d\bm{R}\Psi^{\ast}(\bm{R})\Psi(\bm{R})}
        {\int d\bm{R}\Psi^{\ast}(\bm{R})\Psi(\bm{R})}=\mathrm{constant}.
\]
\textbf{This gives an important information: If I want the variance, the exact wave function leads to zero variance!}
The variance is defined as 
\[
\sigma_E = \langle E^2\rangle - \langle E\rangle^2.
\]
Variation is then performed by minimizing both the energy and the variance.
% --- end paragraph admon ---

% !split
\subsection*{The variance defines the cost function}

% --- begin paragraph admon ---
\paragraph{}
We can then take the derivatives of 
\[
\sigma_E = \langle E^2\rangle - \langle E\rangle^2,
\]
with respect to the variational parameters. The derivatives of the variance can then be used to defined the
so-called Hessian matrix, which in turn allows us to use minimization methods like Newton's method or 
standard gradient methods.

This leads to however a more complicated expression, with obvious errors when evaluating integrals by Monte Carlo integration. Less used, see however \href{{https://journals.aps.org/prl/abstract/10.1103/PhysRevLett.94.150201}}{Filippi and Umrigar}. The expression becomes complicated
\[
\bar{E}_{ij} = 2\left[ \langle (\frac{\bar{\Psi}_{ij}}{\Psi}+\frac{\bar{\Psi}_{j}}{\Psi}\frac{\bar{\Psi}_{i}}{\Psi})(E_L-\langle E\rangle)\rangle -\langle \frac{\bar{\Psi}_{i}}{\Psi}\rangle\bar{E}_j-\langle \frac{\bar{\Psi}_{j}}{\Psi}\rangle\bar{E}_i\right] +\langle \frac{\bar{\Psi}_{i}}{\Psi}E_L{_j}\rangle +\langle \frac{\bar{\Psi}_{j}}{\Psi}E_L{_i}\rangle -\langle \frac{\bar{\Psi}_{i}}{\Psi}\rangle\langle E_L{_j}\rangle \langle \frac{\bar{\Psi}_{j}}{\Psi}\rangle\langle E_L{_i}\rangle.  
\]

Evaluating the cost function means having to evaluate the above second derivative of the energy.
% --- end paragraph admon ---

% !split
\subsection*{Why Boltzmann machines?}

What is known as restricted Boltzmann Machines (RMB) have received a lot of attention lately. 
One of the major reasons is that they can be stacked layer-wise to build deep neural networks that capture complicated statistics.

The original RBMs had just one visible layer and a hidden layer, but recently so-called Gaussian-binary RBMs have gained quite some popularity in imaging since they are capable of modeling continuous data that are common to natural images.

Furthermore, they have been used to solve complicated quantum mechanical many-particle problems or classical statistical physics problems like the Ising and Potts classes of models.

% !split
\subsection*{Boltzmann Machines}

Why use a generative model rather than the more well known discriminative deep neural networks (DNN)?

\begin{itemize}
\item Discriminitave methods have several limitations: They are mainly supervised learning methods, thus requiring labeled data. And there are tasks they cannot accomplish, like drawing new examples from an unknown probability distribution.

\item A generative model can learn to represent and sample from a probability distribution. The core idea is to learn a parametric model of the probability distribution from which the training data was drawn. As an example
\begin{enumerate}

 \item A model for images could learn to draw new examples of cats and dogs, given a training dataset of images of cats and dogs.

 \item Generate a sample of an ordered or disordered Ising model phase, having been given samples of such phases.

 \item Model the trial function for Monte Carlo calculations
\end{enumerate}

\noindent
\end{itemize}

\noindent
% !split
\subsection*{Some similarities and differences from DNNs}

\begin{enumerate}
\item Both use gradient-descent based learning procedures for minimizing cost functions

\item Energy based models don't use backpropagation and automatic differentiation for computing gradients, instead turning to Markov Chain Monte Carlo methods.

\item DNNs often have several hidden layers. A restricted Boltzmann machine has only one hidden layer, however several RBMs can be stacked to make up Deep Belief Networks, of which they constitute the building blocks.
\end{enumerate}

\noindent
History: The RBM was developed by amongst others Geoffrey Hinton, called by some the "Godfather of Deep Learning", working with the University of Toronto and Google.

% !split
\subsection*{Boltzmann machines (BM)}

% --- begin paragraph admon ---
\paragraph{}
A BM is what we would call an undirected probabilistic graphical model
with stochastic continuous or discrete units.
% --- end paragraph admon ---

% --- begin paragraph admon ---
\paragraph{}
It is interpreted as a stochastic recurrent neural network where the
state of each unit(neurons/nodes) depends on the units it is connected
to. The weights in the network represent thus the strength of the
interaction between various units/nodes.
% --- end paragraph admon ---

% --- begin paragraph admon ---
\paragraph{}
It turns into a Hopfield network if we choose deterministic rather
than stochastic units. In contrast to a Hopfield network, a BM is a
so-called generative model. It allows us to generate new samples from
the learned distribution.
% --- end paragraph admon ---

% !split
\subsection*{A standard BM setup}

% --- begin paragraph admon ---
\paragraph{}
A standard BM network is divided into a set of observable and visible units $\hat{x}$ and a set of unknown hidden units/nodes $\hat{h}$.
% --- end paragraph admon ---

% --- begin paragraph admon ---
\paragraph{}
Additionally there can be bias nodes for the hidden and visible layers. These biases are normally set to $1$.
% --- end paragraph admon ---

% --- begin paragraph admon ---
\paragraph{}
BMs are stackable, meaning they cwe can train a BM which serves as input to another BM. We can construct deep networks for learning complex PDFs. The layers can be trained one after another, a feature which makes them popular in deep learning
% --- end paragraph admon ---

However, they are often hard to train. This leads to the introduction of so-called restricted BMs, or RBMS.
Here we take away all lateral connections between nodes in the visible layer as well as connections between nodes in the hidden layer. The network is illustrated in the figure below.

% !split
\subsection*{The structure of the RBM network}

\vspace{6mm}

% inline figure
\centerline{\includegraphics[width=1.0\linewidth]{figures/RBM.pdf}}

\vspace{6mm}

% !split
\subsection*{The network}

\textbf{The network layers}:
\begin{enumerate}
 \item A function $\mathbf{x}$ that represents the visible layer, a vector of $M$ elements (nodes). This layer represents both what the RBM might be given as training input, and what we want it to be able to reconstruct. This might for example be the pixels of an image, the spin values of the Ising model, or coefficients representing speech.

 \item The function $\mathbf{h}$ represents the hidden, or latent, layer. A vector of $N$ elements (nodes). Also called "feature detectors".
\end{enumerate}

\noindent
% !split
\subsection*{Goals}

The goal of the hidden layer is to increase the model's expressive power. We encode complex interactions between visible variables by introducing additional, hidden variables that interact with visible degrees of freedom in a simple manner, yet still reproduce the complex correlations between visible degrees in the data once marginalized over (integrated out).

Examples of this trick being employed in physics: 
\begin{enumerate}
 \item The Hubbard-Stratonovich transformation

 \item The introduction of ghost fields in gauge theory

 \item Shadow wave functions in Quantum Monte Carlo simulations
\end{enumerate}

\noindent
\textbf{The network parameters, to be optimized/learned}:
\begin{enumerate}
 \item $\mathbf{a}$ represents the visible bias, a vector of same length as $\mathbf{x}$.

 \item $\mathbf{b}$ represents the hidden bias, a vector of same lenght as $\mathbf{h}$.

 \item $W$ represents the interaction weights, a matrix of size $M\times N$.
\end{enumerate}

\noindent
% !split
\subsection*{Joint distribution}

The restricted Boltzmann machine is described by a Boltzmann distribution
\begin{align}
	P_{rbm}(\mathbf{x},\mathbf{h}) = \frac{1}{Z} e^{-\frac{1}{T_0}E(\mathbf{x},\mathbf{h})},
\end{align}
where $Z$ is the normalization constant or partition function, defined as 
\begin{align}
	Z = \int \int e^{-\frac{1}{T_0}E(\mathbf{x},\mathbf{h})} d\mathbf{x} d\mathbf{h}.
\end{align}
It is common to ignore $T_0$ by setting it to one.

% !split
\subsection*{Network Elements, the energy function}

The function $E(\mathbf{x},\mathbf{h})$ gives the \textbf{energy} of a
configuration (pair of vectors) $(\mathbf{x}, \mathbf{h})$. The lower
the energy of a configuration, the higher the probability of it. This
function also depends on the parameters $\mathbf{a}$, $\mathbf{b}$ and
$W$. Thus, when we adjust them during the learning procedure, we are
adjusting the energy function to best fit our problem.

An expression for the energy function is
\[
E(\hat{x},\hat{h}) = -\sum_{ia}^{NA}b_i^a \alpha_i^a(x_i)-\sum_{jd}^{MD}c_j^d \beta_j^d(h_j)-\sum_{ijad}^{NAMD}b_i^a \alpha_i^a(x_i)c_j^d \beta_j^d(h_j)w_{ij}^{ad}.
\]

Here $\beta_j^d(h_j)$ and $\alpha_i^a(x_j)$ are so-called transfer functions that map a given input value to a desired feature value. The labels $a$ and $d$ denote that there can be multiple transfer functions per variable. The first sum depends only on the visible units. The second on the hidden ones. \textbf{Note} that there is no connection between nodes in a layer.

The quantities $b$ and $c$ can be interpreted as the visible and hidden biases, respectively.

The connection between the nodes in the two layers is given by the weights $w_{ij}$.

% !split
\subsection*{Defining different types of RBMs}
There are different variants of RBMs, and the differences lie in the types of visible and hidden units we choose as well as in the implementation of the energy function $E(\mathbf{x},\mathbf{h})$.

% --- begin paragraph admon ---
\paragraph{Binary-Binary RBM:}

RBMs were first developed using binary units in both the visible and hidden layer. The corresponding energy function is defined as follows:
\begin{align}
	E(\mathbf{x}, \mathbf{h}) = - \sum_i^M x_i a_i- \sum_j^N b_j h_j - \sum_{i,j}^{M,N} x_i w_{ij} h_j,
\end{align}
where the binary values taken on by the nodes are most commonly 0 and 1.
% --- end paragraph admon ---

% --- begin paragraph admon ---
\paragraph{Gaussian-Binary RBM:}

Another varient is the RBM where the visible units are Gaussian while the hidden units remain binary:
\begin{align}
	E(\mathbf{x}, \mathbf{h}) = \sum_i^M \frac{(x_i - a_i)^2}{2\sigma_i^2} - \sum_j^N b_j h_j - \sum_{i,j}^{M,N} \frac{x_i w_{ij} h_j}{\sigma_i^2}. 
\end{align}
% --- end paragraph admon ---

% !split
\subsection*{More about RBMs}
\begin{enumerate}
\item Useful when we model continuous data (i.e., we wish $\mathbf{x}$ to be continuous)

\item Requires a smaller learning rate, since there's no upper bound to the value a component might take in the reconstruction
\end{enumerate}

\noindent
Other types of units include:
\begin{enumerate}
\item Softmax and multinomial units

\item Gaussian visible and hidden units

\item Binomial units

\item Rectified linear units
\end{enumerate}

\noindent
% !split
\subsection*{Sampling: Metropolis sampling}
In order to sample from the RBM probability distribution it is common to use Markov Chain Monte Carlo (MCMC) algorithms such as Metropolis-Hastings or Gibbs sampling.

Metropolis sampling starts by suggesting a new configuration $\bm{x}^{k+1}$. In the brute force method this is done by some random change of the visible units. The new configuration is then accepted with the acceptance probability
\begin{align}
	A(\bm{x}^k \rightarrow \bm{x}^{k+1}) = \text{min} (1, \frac{P(\bm{x}^{k+1})}{P(\bm{x}^k)}),
\end{align} 
where we need the marginalized probability
\begin{align}
	P(\bm{x})  &= \sum_\mathbf{h} P_{rbm}(\mathbf{x}, \mathbf{h}) \\
				&= \frac{1}{Z}\sum_\mathbf{h} e^{-E(\mathbf{x}, \mathbf{h})}.
\end{align}

% !split
\subsection*{Sampling: Gibbs sampling}

In this method we sample from the joint probability $P_{rbm} (\mathbf{x}, \mathbf{h})$ by way of a two step sampling process. We alternately update the visible and hidden units.
New samples are generated according to the conditional probabilities $P(x_i|\mathbf{h})$ and $P(h_j|\mathbf{x})$ respectively and accepted with the probability of $1$. While the the visible nodes are dependent on the hidden nodes and vice versa, the nodes are independent of other nodes within the same layer. This is due to there being no intra layer interactions in the restricted Boltzmann machine.

The conditional probabilities are often referred to as the activitation functions in the neural networks context due to their role in determining the node outputs. For the binary-binary RBM they are
\begin{align}
	P(h_j = 1 | \bm{x}) &= \frac{1}{1 + e^{-b_j - \sum_i x_i w_{ij}}} \\
	P(x_i = 1 | \bm{h}) &= \frac{1}{1 + e^{-a_j - \sum_j h_j w_{ij}}},
\end{align}
where we recognize the logistic sigmoid function $\sigma (x) = 1/(1+exp(-x))$.

% !split
\subsection*{Gaussian RBM}
For the Gaussian-Binary RBM the conditional probabilities are
\begin{align}
	P(x_i|\mathbf{h}) &= \mathcal{N}(x_i; a_i+ \sum_j h_j w_{ij}, \sigma^2) \\
	P(h_j=1|\mathbf{x}) &=  \frac{1}{1+e^{-b_j-\frac{1}{\sigma^2} \sum_i x_i w_{ij}}},
\end{align}
while the visible units now follow a normal distribution, we see the hidden units again follow the logistic sigmoid function.

% !split
\subsection*{Cost function}

When working with a training dataset, the most common training approach is maximizing the log-likelihood of the training data. The log likelihood characterizes the log-probability of generating the observed data using our generative model. Using this method our cost function is chosen as the negative log-likelihood. The learning then consists of trying to find parameters that maximize the probability of the dataset, and is known as Maximum Likelihood Estimation (MLE).
Denoting the parameters as $\bm{\theta} = a_1,...,a_M,b_1,...,b_N,w_{11},...,w_{MN}$, the log-likelihood is given by
\begin{align}
	\mathcal{L}(\{ \theta_i \}) &= \langle \text{log} P_\theta(\bm{x}) \rangle_{data} \\
	&= - \langle E(\bm{x}; \{ \theta_i\}) \rangle_{data} - \text{log} Z(\{ \theta_i\}),
\end{align}
where we used that the normalization constant does not depend on the data, $\langle \text{log} Z(\{ \theta_i\}) \rangle = \text{log} Z(\{ \theta_i\})$
Our cost function is the negative log-likelihood, $\mathcal{C}(\{ \theta_i \}) = - \mathcal{L}(\{ \theta_i \})$

% !split
\subsection*{Optimization / Training}

The training procedure of choice often is Stochastic Gradient Descent (SGD). It consists of a series of iterations where we update the parameters according to the equation
\begin{align}
	\bm{\theta}_{k+1} = \bm{\theta}_k - \eta \nabla \mathcal{C} (\bm{\theta}_k)
\end{align}
at each $k$-th iteration. There are a range of variants of the algorithm which aim at making the learning rate $\eta$ more adaptive so the method might be more efficient while remaining stable.

We now need the gradient of the cost function in order to minimize it. We find that
\begin{align}
	\frac{\partial \mathcal{C}(\{ \theta_i\})}{\partial \theta_i}
	&= \langle \frac{\partial E(\bm{x}; \theta_i)}{\partial \theta_i} \rangle_{data}
	+ \frac{\partial \text{log} Z(\{ \theta_i\})}{\partial \theta_i} \\
	&= \langle O_i(\bm{x}) \rangle_{data} - \langle O_i(\bm{x}) \rangle_{model},
\end{align}
where in order to simplify notation we defined the "operator"
\begin{align}
	O_i(\bm{x}) = \frac{\partial E(\bm{x}; \theta_i)}{\partial \theta_i}, 
\end{align}
and used the statistical mechanics relationship between expectation values and the log-partition function:
\begin{align}
	\langle O_i(\bm{x}) \rangle_{model} = \text{Tr} P_\theta(\bm{x})O_i(\bm{x}) = - \frac{\partial \text{log} Z(\{ \theta_i\})}{\partial \theta_i}.
\end{align}

% !split 
\subsection*{More on RBMs}

The data-dependent term in the gradient is known as the positive phase of the gradient, while the model-dependent term is known as the negative phase of the gradient. The aim of the training is to lower the energy of configurations that are near observed data points (increasing their probability), and raising the energy of configurations that are far from observed data points (decreasing their probability).

The gradient of the negative log-likelihood cost function of a Binary-Binary RBM is then
\begin{align}
	\frac{\partial \mathcal{C} (w_{ij}, a_i, b_j)}{\partial w_{ij}} =& \langle x_i h_j \rangle_{data} - \langle x_i h_j \rangle_{model} \\
	\frac{\partial \mathcal{C} (w_{ij}, a_i, b_j)}{\partial a_{ij}} =& \langle x_i \rangle_{data} - \langle x_i \rangle_{model} \\
	\frac{\partial \mathcal{C} (w_{ij}, a_i, b_j)}{\partial b_{ij}} =& \langle h_i \rangle_{data} - \langle h_i \rangle_{model}. \\
\end{align}
To get the expecation values with respect to the \emph{data}, we set the visible units to each of the observed samples in the training data, then update the hidden units according to the conditional probability found before. We then average over all samples in the training data to calculate expectation values with respect to the data.

% !split
\subsection*{Which sampling to use}

To get the expectation values with respect to the \emph{model}, we use Gibbs sampling. We can either initialize the $\bm{x}$ randomly or with a training sample. While we ideally want a large number of Gibbs iterations $n\rightarrow n$, one might decide to truncate it earlier for efficiency. Doing this while having intialized $\bm{x}$ with a training data vector is referred to as contrastive divergence (CD), because one is then closer to approximating the gradient of this function than the negative log-likelihood. The contrastive divergence function is the difference between two Kullback-Leibler divergences (also called relative entropy), which measure how one probability distribution diverges from a second, expected probability distribution (in this case the estimated one from the ground truth one).

% !split
\subsection*{RBMs for the quantum many body problem}

The idea of applying RBMs to quantum many body problems was presented by G. Carleo and M. Troyer, working with ETH Zurich and Microsoft Research.

Some of their motivation included

\begin{itemize}
\item "The wave function $\Psi$ is a monolithic mathematical quantity that contains all the information on a quantum state, be it a single particle or a complex molecule. In principle, an exponential amount of information is needed to fully encode a generic many-body quantum state."

\item There are still interesting open problems, including fundamental questions ranging from the dynamical properties of high-dimensional systems to the exact ground-state properties of strongly interacting fermions.

\item The difficulty lies in finding a general strategy to reduce the exponential complexity of the full many-body wave function down to its most essential features. That is
\begin{enumerate}

\item $\rightarrow$ Dimensional reduction

\item $\rightarrow$ Feature extraction

\end{enumerate}

\noindent
\item Among the most successful techniques to attack these challenges, artifical neural networks play a prominent role.

\item Want to understand whether an artifical neural network may adapt to describe a quantum system.
\end{itemize}

\noindent
% !split
\subsection*{Choose the right RBM}

Carleo and Troyer applied the RBM to the quantum mechanical spin lattice systems of the Ising model and Heisenberg model, with encouraging results. Our goal is to test the method on systems of moving particles. For the spin lattice systems it was natural to use a binary-binary RBM, with the nodes taking values of 1 and -1. For moving particles, on the other hand, we want the visible nodes to be continuous, representing position coordinates. Thus, we start by choosing a Gaussian-binary RBM, where the visible nodes are continuous and hidden nodes take on values of 0 or 1. If eventually we would like the hidden nodes to be continuous as well the rectified linear units seem like the most relevant choice.

% !split
\subsection*{Representing the wave function}
The wavefunction should be a probability amplitude depending on $\bm{x}$. The RBM model is given by the joint distribution of $\bm{x}$ and $\bm{h}$
\begin{align}
	F_{rbm}(\mathbf{x},\mathbf{h}) = \frac{1}{Z} e^{-\frac{1}{T_0}E(\mathbf{x},\mathbf{h})}.
\end{align}
To find the marginal distribution of $\bm{x}$ we set:
\begin{align}
	F_{rbm}(\mathbf{x}) &= \sum_\mathbf{h} F_{rbm}(\mathbf{x}, \mathbf{h}) \\
				&= \frac{1}{Z}\sum_\mathbf{h} e^{-E(\mathbf{x}, \mathbf{h})}.
\end{align}
Now this is what we use to represent the wave function, calling it a neural-network quantum state (NQS)
\begin{align}
	\Psi (\mathbf{X}) &= F_{rbm}(\mathbf{x}) \\
	&= \frac{1}{Z}\sum_{\bm{h}} e^{-E(\mathbf{x}, \mathbf{h})} \\
	&= \frac{1}{Z} \sum_{\{h_j\}} e^{-\sum_i^M \frac{(x_i - a_i)^2}{2\sigma^2} + \sum_j^N b_j h_j + \sum_{i,j}^{M,N} \frac{x_i w_{ij} h_j}{\sigma^2}} \\
	&= \frac{1}{Z} e^{-\sum_i^M \frac{(x_i - a_i)^2}{2\sigma^2}} \prod_j^N (1 + e^{b_j + \sum_i^M \frac{x_i w_{ij}}{\sigma^2}}). \\
\end{align}

% !split
\subsection*{Choose the cost function}
Now we don't necessarily have training data (unless we generate it by using some other method). However, what we do have is the variational principle which allows us to obtain the ground state wave function by minimizing the expectation value of the energy of a trial wavefunction (corresponding to the untrained NQS). Similarly to the traditional variational Monte Carlo method then, it is the local energy we wish to minimize. The gradient to use for the stochastic gradient descent procedure is
\begin{align}
	G_i = \frac{\partial \langle E_L \rangle}{\partial \theta_i}
	= 2(\langle E_L \frac{1}{\Psi}\frac{\partial \Psi}{\partial \theta_i} \rangle - \langle E_L \rangle \langle \frac{1}{\Psi}\frac{\partial \Psi}{\partial \theta_i} \rangle ),
\end{align}
where the local energy is given by
\begin{align}
	E_L = \frac{1}{\Psi} \hat{\mathbf{H}} \Psi.
\end{align}

% !split
\subsection*{Running the codes}

% --- begin paragraph admon ---
\paragraph{}
You can find the codes for the simple two-electron case at the Github repository \href{{https://github.com/mhjensenseminars/MachineLearningTalk/tree/master/doc/Programs/MLcpp/src}}{\nolinkurl{https://github.com/mhjensenseminars/MachineLearningTalk/tree/master/doc/Programs/MLcpp/src}}. Python codes to come, only c++ as of now.

The trial wave function is based on the product of a Slater determinant with Gaussian orbitals, a simple Jastrow factor $\exp{(r_{ij})}$ and the reduced Boltzmann machines.

The Broyden-Fletcher-Goldfarb-Shanno algorithm was used for the minimization. We used $14$ hidden nodes in the calculations below.
% --- end paragraph admon ---

% !split
\subsection*{Energy as function of iterations, $N=2$ electrons}

% --- begin paragraph admon ---
\paragraph{}

\vspace{6mm}

% inline figure
\centerline{\includegraphics[width=0.9\linewidth]{figures/figN2.pdf}}

\vspace{6mm}
% --- end paragraph admon ---

% !split
\subsection*{Energy as function of iterations, $N=6$ electrons}

% --- begin paragraph admon ---
\paragraph{}

\vspace{6mm}

% inline figure
\centerline{\includegraphics[width=0.9\linewidth]{figures/figN6.pdf}}

\vspace{6mm}
% --- end paragraph admon ---

% !split
\subsection*{Conclusions and where do we stand}

% --- begin paragraph admon ---
\paragraph{}
\begin{itemize}
\item A simple extension of the work of \href{{http://science.sciencemag.org/content/355/6325/602}}{G. Carleo and M. Troyer, Science \textbf{355}, Issue 6325, pp. 602-606 (2017)} gives excellent results for two-electron systems as well as good agreement with standard VMC calculations for $N=6$ and $N=12$ electrons.

\item Minimization problem can be tricky.

\item Anti-symmetry dealt with multiplying the trail wave function with an optimized Slater determinant.

\item To come: Analysis of wave function from ML and compare with diffusion and Variational Monte Carlo calculations as well as the analytical results of Taut for the two-electron case.

\item Extend to more fermions. How do we deal with the antisymmetry of the multi-fermion wave function?
\begin{enumerate}

 \item Here we used standard Hartree-Fock theory to define an optimal Slater determinant. Takes care of the antisymmetry. What about constructing an anti-symmetrized network function?

 \item Use thereafter ML to determine the correlated part of the wafe function (including a standard Jastrow factor).

 \item Test this for multi-fermion systems and compare with other many-body methods.

\end{enumerate}

\noindent
\item Can we use ML to find out which correlations are relevant and thereby diminish the dimensionality problem in say CC or SRG theories?
\end{itemize}

\noindent
% --- end paragraph admon ---

% !split
\subsection*{Additional material}

% !split
\subsection*{Kullback-Leibler relative entropy}

When the goal of the training is to approximate a probability
distribution, as it is in generative modeling, another relevant
measure is the \textbf{Kullback-Leibler divergence}, also known as the
relative entropy or Shannon entropy. It is a non-symmetric measure of the
dissimilarity between two probability density functions $p$ and
$q$. If $p$ is the unkown probability which we approximate with $q$,
we can measure the difference by
\begin{align}
	\text{KL}(p||q) = \int_{-\infty}^{\infty} p (\bm{x}) \log \frac{p(\bm{x})}{q(\bm{x})}  d\bm{x}.
\end{align}

Thus, the Kullback-Leibler divergence between the distribution of the
training data $f(\bm{x})$ and the model distribution $p(\bm{x}|
\bm{\theta})$ is

\begin{align}
	\text{KL} (f(\bm{x})|| p(\bm{x}| \bm{\theta})) =& \int_{-\infty}^{\infty}
	f (\bm{x}) \log \frac{f(\bm{x})}{p(\bm{x}| \bm{\theta})} d\bm{x} \\
	=& \int_{-\infty}^{\infty} f(\bm{x}) \log f(\bm{x}) d\bm{x} - \int_{-\infty}^{\infty} f(\bm{x}) \log
	p(\bm{x}| \bm{\theta}) d\bm{x} \\
	%=& \mathbb{E}_{f(\bm{x})} (\log f(\bm{x})) - \mathbb{E}_{f(\bm{x})} (\log p(\bm{x}| \bm{\theta}))
	=& \langle \log f(\bm{x}) \rangle_{f(\bm{x})} - \langle \log p(\bm{x}| \bm{\theta}) \rangle_{f(\bm{x})} \\
	=& \langle \log f(\bm{x}) \rangle_{data} + \langle E(\bm{x}) \rangle_{data} + \log Z \\
	=& \langle \log f(\bm{x}) \rangle_{data} + \mathcal{C}_{LL} .
\end{align}

The first term is constant with respect to $\bm{\theta}$ since $f(\bm{x})$ is independent of $\bm{\theta}$. Thus the Kullback-Leibler Divergence is minimal when the second term is minimal. The second term is the log-likelihood cost function, hence minimizing the Kullback-Leibler divergence is equivalent to maximizing the log-likelihood.

% !split
\subsection*{Optimizing the cost function}

To further understand generative models it is useful to study the
gradient of the cost function which is needed in order to minimize it
using methods like stochastic gradient descent.

The partition function is the generating function of
expectation values, in particular there are mathematical relationships
between expectation values and the log-partition function. In this
case we have
\begin{align}
	\langle \frac{ \partial E(\bm{x}; \theta_i) } { \partial \theta_i} \rangle_{model}
	= \int p(\bm{x}| \bm{\theta}) \frac{ \partial E(\bm{x}; \theta_i) } { \partial \theta_i} d\bm{x} 
	= -\frac{\partial \log Z(\theta_i)}{ \partial  \theta_i} .
\end{align}

Here $\langle \cdot \rangle_{model}$ is the expectation value over the model probability distribution $p(\bm{x}| \bm{\theta})$.

% !split
\subsection*{Setting up for gradient descent calculations}

Using the previous relationship we can express the gradient of the cost function as

\begin{align}
	\frac{\partial \mathcal{C}_{LL}}{\partial \theta_i}
	=& \langle \frac{ \partial E(\bm{x}; \theta_i) } { \partial \theta_i} \rangle_{data} + \frac{\partial \log Z(\theta_i)}{ \partial  \theta_i} \\
	=& \langle \frac{ \partial E(\bm{x}; \theta_i) } { \partial \theta_i} \rangle_{data} - \langle \frac{ \partial E(\bm{x}; \theta_i) } { \partial \theta_i} \rangle_{model} \\
	%=& \langle O_i(\bm{x}) \rangle_{data} - \langle O_i(\bm{x}) \rangle_{model}
\end{align}

This expression shows that the gradient of the log-likelihood cost
function is a \textbf{difference of moments}, with one calculated from
the data and one calculated from the model. The data-dependent term is
called the \textbf{positive phase} and the model-dependent term is
called the \textbf{negative phase} of the gradient. We see now that
minimizing the cost function results in lowering the energy of
configurations $\bm{x}$ near points in the training data and
increasing the energy of configurations not observed in the training
data. That means we increase the model's probability of configurations
similar to those in the training data.

% !split
\subsection*{More interpretations}

The gradient of the cost function also demonstrates why gradients of
unsupervised, generative models must be computed differently from for
those of for example FNNs. While the data-dependent expectation value
is easily calculated based on the samples $\bm{x}_i$ in the training
data, we must sample from the model in order to generate samples from
which to caclulate the model-dependent term. We sample from the model
by using MCMC-based methods. We can not sample from the model directly
because the partition function $Z$ is generally intractable.

As in supervised machine learning problems, the goal is also here to
perform well on \textbf{unseen} data, that is to have good
generalization from the training data. The distribution $f(x)$ we
approximate is not the \textbf{true} distribution we wish to estimate,
it is limited to the training data. Hence, in unsupervised training as
well it is important to prevent overfitting to the training data. Thus
it is common to add regularizers to the cost function in the same
manner as we discussed for say linear regression.

% ------------------- end of main content ---------------

\end{document}

### 2019-01-16 doc/MachineLearning/article2.tex
update on article

--- 2019-01-16 doc/MachineLearning/article1.tex
@@
 \end{document}
-
-%%
-%% Automatically generated file from DocOnce source
-%% (https://github.com/hplgit/doconce/)
-%%
-%%
-
-%-------------------- begin preamble ----------------------
-
-\documentclass[%
-oneside,                 % oneside: electronic viewing, twoside: printing
-final,                   % draft: marks overfull hboxes, figures with paths
-10pt]{article}
-
-\listfiles               %  print all files needed to compile this document
-
-\usepackage{relsize,makeidx,color,setspace,amsmath,amsfonts,amssymb}
-\usepackage[table]{xcolor}
-\usepackage{bm,ltablex,microtype}
-
-\usepackage[pdftex]{graphicx}
-
-\usepackage{fancyvrb} % packages needed for verbatim environments
-\usepackage{minted}
-\usemintedstyle{default}
-
-\usepackage[T1]{fontenc}
-%\usepackage[latin1]{inputenc}
-\usepackage{ucs}
-\usepackage[utf8x]{inputenc}
-
-\usepackage{lmodern}         % Latin Modern fonts derived from Computer Modern
-
-% Hyperlinks in PDF:
-\definecolor{linkcolor}{rgb}{0,0,0.4}
-\usepackage{hyperref}
-\hypersetup{
-    breaklinks=true,
-    colorlinks=true,
-    linkcolor=linkcolor,
-    urlcolor=linkcolor,
-    citecolor=black,
-    filecolor=black,
-    %filecolor=blue,
-    pdfmenubar=true,
-    pdftoolbar=true,
-    bookmarksdepth=3   % Uncomment (and tweak) for PDF bookmarks with more levels than the TOC
-    }
-%\hyperbaseurl{}   % hyperlinks are relative to this root
-
-\setcounter{tocdepth}{2}  % levels in table of contents
-
-% Tricks for having figures close to where they are defined:
-% 1. define less restrictive rules for where to put figures
-\setcounter{topnumber}{2}
-\setcounter{bottomnumber}{2}
-\setcounter{totalnumber}{4}
-\renewcommand{\topfraction}{0.95}
-\renewcommand{\bottomfraction}{0.95}
-\renewcommand{\textfraction}{0}
-\renewcommand{\floatpagefraction}{0.75}
-% floatpagefraction must always be less than topfraction!
-% 2. ensure all figures are flushed before next section
-\usepackage[section]{placeins}
-% 3. enable begin{figure}[H] (often leads to ugly pagebreaks)
-%\usepackage{float}\restylefloat{figure}
-
-% --- fancyhdr package for fancy headers ---
-\usepackage{fancyhdr}
-\fancyhf{} % sets both header and footer to nothing
-\renewcommand{\headrulewidth}{0pt}
-\fancyfoot[LE,RO]{\thepage}
-% Ensure copyright on titlepage (article style) and chapter pages (book style)
-\fancypagestyle{plain}{
-  \fancyhf{}
-  \fancyfoot[C]{{\footnotesize \copyright\ 1999-2019, Morten Hjorth-Jensen. Released under CC Attribution-NonCommercial 4.0 license}}
-%  \renewcommand{\footrulewidth}{0mm}
-  \renewcommand{\headrulewidth}{0mm}
-}
-% Ensure copyright on titlepages with \thispagestyle{empty}
-\fancypagestyle{empty}{
-  \fancyhf{}
-  \fancyfoot[C]{{\footnotesize \copyright\ 1999-2019, Morten Hjorth-Jensen. Released under CC Attribution-NonCommercial 4.0 license}}
-  \renewcommand{\footrulewidth}{0mm}
-  \renewcommand{\headrulewidth}{0mm}
-}
-
-\pagestyle{fancy}
-
-\usepackage[framemethod=TikZ]{mdframed}
-
-% --- begin definitions of admonition environments ---
-
-% --- end of definitions of admonition environments ---
-
-% prevent orhpans and widows
-\clubpenalty = 10000
-\widowpenalty = 10000
-
-% --- end of standard preamble for documents ---
-
-% insert custom LaTeX commands...
-
-\raggedbottom
-\makeindex
-\usepackage[totoc]{idxlayout}   % for index in the toc
-\usepackage[nottoc]{tocbibind}  % for references/bibliography in the toc
-
-%-------------------- end preamble ----------------------
-
-\begin{document}
-
-% matching end for #ifdef PREAMBLE
-
-\newcommand{\exercisesection}[1]{\subsection*{#1}}
-
-% ------------------- main content ----------------------
-
-% ----------------- title -------------------------
-
-\thispagestyle{empty}
-
-\begin{center}
-{\LARGE\bf
-\begin{spacing}{1.25}
-Solving quantum mechanical problems with Machine Learning
-\end{spacing}
-}
-\end{center}
-
-% ----------------- author(s) -------------------------
-
-\begin{center}
-{\bf Morten Hjorth-Jensen${}^{1, 2}$} \\ [0mm]
-\end{center}
-
-\begin{center}
-% List of all institutions:
-\centerline{{\small ${}^1$Department of Physics and Astronomy and National Superconducting Cyclotron Laboratory, Michigan State University, USA}}
-\centerline{{\small ${}^2$Department of Physics and Center for Computing in Science Education, University of Oslo, Norway}}
-\end{center}
-
-% ----------------- end author(s) -------------------------
-
-% --- begin date ---
-\begin{center}
-Jan 16, 2019 
-\end{center}
-% --- end date ---
-
-\vspace{1cm}
-
-% !split
-\subsection*{What is this talk about?}
-
-% --- begin paragraph admon ---
-\paragraph{}
-The main aim is to give you a short and pedestrian introduction to  how we can use Machine Learning methods
-to solve quantum mechanical many-body problems. And why this could be of interest.
-
-The hope is that after this talk you have gotten the basic ideas to get you started.
-Peeping into \href{{https://github.com/mhjensenseminars/MachineLearningTalk}}{\nolinkurl{https://github.com/mhjensenseminars/MachineLearningTalk}}, you'll find a Jupyter notebook, slides, codes etc that will allow you to reproduce the simulations discussed here, and perhaps run your own very first calculations.
-
-Try it out and please don't hesitate to swing by if something is unclear.
-% --- end paragraph admon ---
-
-% !split
-\subsection*{Why?}
-
-How can we avoid the dimensionality curse? Many possibilities
-\begin{enumerate}
-\item smarter basis functions
-
-\item resummation of specific correlations
-
-\item stochastic sampling of high-lying states (stochastic FCI, CC and SRG/IMSRG)
-
-\item many more
-\end{enumerate}
-
-\noindent
-Machine Learning and Quantum Computing hold also great promise in tackling the 
-ever increasing dimensionalities. Here we will focus on Machine Learning.
-
-% !split
-\subsection*{Overview}
-
-% --- begin paragraph admon ---
-\paragraph{}
-\begin{itemize}
-\item Short intro to Machine Learning
-
-\item Variational Monte Carlo (Markov Chain Monte Carlo, $\mathrm{MC}^2$) and two-electron quantum dots, solving quantum mechanical problems in a stochastic way.
-It  will serve as  our motivation for switching to Machine Learning.
-
-\item From Variational Monte Carlo to Boltzmann Machines and Machine Learning
-\end{itemize}
-
-\noindent
-% --- end paragraph admon ---
-
-% !split
-\subsection*{What are the Machine Learning calculations here based on?}
-
-This work is inspired by the idea of representing the wave function with
-a restricted Boltzmann machine (RBM), presented recently by \href{{http://science.sciencemag.org/content/355/6325/602}}{G.
-Carleo and M.
-Troyer, Science \textbf{355}, Issue 6325, pp.
-602-606 (2017)}.
-They
-named such a wave function/network a \emph{neural network quantum state} (NQS). In their article they apply it to the quantum mechanical
-spin lattice systems of the Ising model and Heisenberg model, with
-encouraging results.
-
-Thanks to Jane Kim (MSU), Vilde Flugsrud (UiO), Alfred Alocias Mariadason (UiO) for many discussions and interpretations.
-
-% !split
-\subsection*{A new world}
-
-% --- begin paragraph admon ---
-\paragraph{}
-Machine learning (ML) is an extremely rich field, in spite of its young age. The
-increases we have seen during the last three decades in computational
-capabilities have been followed by developments of methods and
-techniques for analyzing and handling large date sets, relying heavily
-on statistics, computer science and mathematics.  The field is rather
-new and developing rapidly.
-
-Popular software packages written in Python for ML are
-
-\begin{itemize}
-\item \href{{http://scikit-learn.org/stable/}}{Scikit-learn},
-
-\item \href{{https://www.tensorflow.org/}}{Tensorflow},
-
-\item \href{{http://pytorch.org/}}{PyTorch} and
-
-\item \href{{https://keras.io/}}{Keras}.
-\end{itemize}
-
-\noindent
-These are all freely available at their respective GitHub sites. They 
-encompass communities of developers in the thousands or more. And the number
-of code developers and contributors keeps increasing.
-% --- end paragraph admon ---
-
-% !split
-\subsection*{Lots of room for creativity}
-
-% --- begin paragraph admon ---
-\paragraph{}
-Not all the
-algorithms and methods can be given a rigorous mathematical
-justification, opening up thereby for experimenting
-and trial and error and thereby exciting new developments.
-% --- end paragraph admon ---
-
-% --- begin paragraph admon ---
-\paragraph{}
-A solid command of linear algebra, multivariate theory, 
-probability theory, statistical data analysis,
-understanding errors and Monte Carlo methods is important in order to understand many of the 
-various algorithms and methods.
-% --- end paragraph admon ---
-
-\textbf{Job market, a personal statement}: \href{{https://www.analyticsindiamag.com/top-countries-hiring-most-number-of-artificial-intelligence-machine-learning-experts/}}{A familiarity with ML is almost becoming a prerequisite for many of the most exciting employment opportunities}.
-And add quantum computing and there you are!
-
-% !split
-\subsection*{Knowledge of Statistical analysis and optimization of data}
-Some key elements that enter much of the discussion on ML:
-\begin{enumerate}
-\item Basic concepts, expectation values, variance, covariance, correlation functions and errors;
-
-\item Simpler models, binomial distribution, the Poisson distribution, simple and multivariate normal distributions;
-
-\item Central elements of Bayesian statistics and modeling;
-
-\item Central elements from linear algebra
-
-\item Gradient methods for data optimization
-
-\item Monte Carlo methods, Markov chains, Metropolis-Hastings algorithm;
-
-\item Estimation of errors using cross-validation, blocking, bootstrapping and jackknife methods;
-
-\item Practical optimization using Singular-value decomposition and least squares for parameterizing data.
-
-\item Principal Component Analysis.
-\end{enumerate}
-
-\noindent
-% !split 
-\subsection*{Some members of the ML family}
-
-% --- begin paragraph admon ---
-\paragraph{}
-\begin{enumerate}
-\item Linear regression and its variants, Logistic regression
-
-\item Decision tree algorithms, from simpler to more complex ones like random forests
-
-\item Bayesian statistics
-
-\item Support vector machines and finally various variants of
-
-\item Artifical neural networks and deep learning
-
-\item Convolutional NN, autoenconders
-
-\item and many more
-\end{enumerate}
-
-\noindent
-% --- end paragraph admon ---
-
-% !split
-\subsection*{What are the basic ingredients?}
-
-% --- begin paragraph admon ---
-\paragraph{}
-Almost every problem in ML and data science starts with the same ingredients:
-\begin{itemize}
-\item The dataset $\mathbf{x}$ (could be some observable quantity of the system we are studying)
-
-\item A model which is a function of a set of parameters $\mathbf{\alpha}$ that relates to the dataset, say a likelihood  function $p(\mathbf{x}\vert \mathbf{\alpha})$ or just a simple model $f(\mathbf{\alpha})$
-
-\item A so-called \textbf{cost} function $\mathcal{C} (\mathbf{x}, f(\mathbf{\alpha}))$ which allows us to decide how well our model represents the dataset. 
-\end{itemize}
-
-\noindent
-We seek to minimize the function $\mathcal{C} (\mathbf{x}, f(\mathbf{\alpha}))$ by finding the parameter values which minimize $\mathcal{C}$. This leads to  various minimization algorithms.
-% --- end paragraph admon ---
-
-% !split
-\subsection*{What is Machine Learning?}
-
-% --- begin paragraph admon ---
-\paragraph{}
-Machine learning is the science of giving computers the ability to
-learn without being explicitly programmed.  The idea is that there
-exist generic algorithms which can be used to find patterns in a broad
-class of data sets without having to write code specifically for each
-problem. The algorithm will build its own logic based on the data.
-
-Machine learning is a subfield of computer science, and is closely
-related to computational statistics.  It evolved from the study of
-pattern recognition in artificial intelligence (AI) research, and has
-made contributions to AI tasks like computer vision, natural language
-processing and speech recognition. It has also, especially in later
-years, found applications in a wide variety of other areas, including
-bioinformatics, economy, physics, finance and marketing.
-
-You will notice however that many of the basic ideas discussed do come from Physics!
-% --- end paragraph admon ---
-
-% !split
-\subsection*{Types of Machine Learning}
-
-% --- begin paragraph admon ---
-\paragraph{}
-The approaches to machine learning are many, but are often split into two main categories. 
-In \emph{supervised learning} we know the answer to a problem,
-and let the computer deduce the logic behind it. On the other hand, \emph{unsupervised learning}
-is a method for finding patterns and relationship in data sets without any prior knowledge of the system.
-Some authours also operate with a third category, namely \emph{reinforcement learning}. This is a paradigm 
-of learning inspired by behavioural psychology, where learning is achieved by trial-and-error, 
-solely from rewards and punishment.
-
-Another way to categorize machine learning tasks is to consider the desired output of a system.
-Some of the most common tasks are:
-
-\begin{itemize}
-  \item Classification: Outputs are divided into two or more classes.
-The goal is to   produce a model that assigns inputs into one of these classes.
-An example is to identify  digits based on pictures of hand-written ones.
-Classification is typically supervised learning.
-
-  \item Regression: Finding a functional relationship between an input data set and a reference data set.   The goal is to construct a function that maps input data to continuous output values.
-
-  \item Clustering: Data are divided into groups with certain common traits, without knowing the different groups beforehand.  It is thus a form of unsupervised learning.
-\end{itemize}
-
-\noindent
-% --- end paragraph admon ---
-
-% !split
-\subsection*{References}
-
-% --- begin paragraph admon ---
-\paragraph{}
-\begin{itemize}
-\item An excellent reference, \href{{https://arxiv.org/abs/1803.08823}}{Mehta \emph{et al.}, arXiv:1803.08823 and Physics Reports in press (2018)}
-
-\item A cute paper by \href{{https://journals.aps.org/prc/abstract/10.1103/PhysRevC.97.014306}}{Utama and Piekarewicz, Validating neural-network refinements of nuclear mass models, Phys.
-Rev.
-C 97, 014306}
-
-\item \href{{https://journals.aps.org/prl/abstract/10.1103/PhysRevLett.120.156001}}{Every issue of Physical Review Letters has now one or more articles on ML}
-
-\item \href{{https://github.com/CompPhysics/MachineLearning}}{Books and lectures notes} and see also the course \href{{https://www.uio.no/studier/emner/matnat/fys/FYS-STK4155/h18/index.html}}{FYS-STK3155/4155}
-
-\item See also Metha and Schwab, arXiv.1410.3831, how to link Variational renormalization group theory with deep learning (recommended read)
-\end{itemize}
-
-\noindent
-% --- end paragraph admon ---
-
-% !split
-\subsection*{Another interesting article}
-
-Here we will use so-called \textbf{reduced Boltzmann Machines} to simulate quantum many-body problems.
-For Monte Carlo aficionados, there is a very close similarity with what are called \textbf{shadow wave functions}, see the work of \href{{https://journals.aps.org/pre/abstract/10.1103/PhysRevE.90.053304}}{Pederiva and Kalos and collaborators, Phys Rev.
-E 90, 053304 (2014)}.
-
-% !split
-\subsection*{Courses on Data science and Machine Learning  at UiO}
-
-The \href{{https://www.mn.uio.no/english/research/about/centre-focus/innovation/data-science/studies/}}{link here}  gives an excellent overview of courses on Machine learning at UiO.
-
-\begin{enumerate}
-\item \href{{http://www.uio.no/studier/emner/matnat/math/STK2100/index-eng.html}}{STK2100 Machine learning and statistical methods for prediction and classification}.
-
-\item \href{{https://www.uio.no/studier/emner/matnat/ifi/IN3050/index-eng.html}}{IN3050 Introduction to Artificial Intelligence and Machine Learning}.
-Introductory course in machine learning and AI with an algorithmic approach.
-
-\item \href{{http://www.uio.no/studier/emner/matnat/math/STK-INF3000/index-eng.html}}{STK-INF3000/4000 Selected Topics in Data Science}.
-The course provides insight into selected contemporary relevant topics within Data Science.
-
-\item \href{{https://www.uio.no/studier/emner/matnat/ifi/IN4080/index.html}}{IN4080 Natural Language Processing}. Probabilistic and machine learning techniques applied to natural language processing.
-
-\item \href{{https://www.uio.no/studier/emner/matnat/math/STK-IN4300/index-eng.html}}{STK-IN4300 – Statistical learning methods in Data Science}.
-An advanced introduction to statistical and machine learning.
-For students with a good mathematics and statistics background.
-
-\item \href{{http://www.uio.no/studier/emner/matnat/ifi/INF4490/}}{INF4490 Biologically Inspired Computing}.
-An introduction to self-adapting methods also called artificial intelligence or machine learning.
-
-\item \href{{https://www.uio.no/studier/emner/matnat/ifi/IN-STK5000/index-eng.html}}{IN-STK5000  Adaptive Methods for Data-Based Decision Making}.
-Methods for adaptive collection and processing of data based on machine learning techniques.
-
-\item \href{{https://www.uio.no/studier/emner/matnat/ifi/IN5400/}}{IN5400/INF5860 – Machine Learning for Image Analysis}.
-An introduction to deep learning with particular emphasis on applications within Image analysis, but useful for other application areas too.
-
-\item \href{{https://www.uio.no/studier/emner/matnat/its/TEK5040/}}{TEK5040 – Deep learning}.
-The course addresses advanced algorithms and architectures for deep learning with neural networks.
-The course provides an introduction to how deep-learning techniques can be used in the construction of key parts of advanced autonomous systems that exist in physical environments and cyber environments.
-\end{enumerate}
-
-\noindent
-% !split
-\subsection*{Additional courses of interest}
-
-\begin{enumerate}
-\item \href{{https://www.uio.no/studier/emner/matnat/math/STK4051/index-eng.html}}{STK4051 Computational Statistics}
-
-\item \href{{https://www.uio.no/studier/emner/matnat/math/STK4021/index-eng.html}}{STK4021 Applied Bayesian Analysis and Numerical Methods}
-\end{enumerate}
-
-\noindent
-% !split
-\subsection*{Decision trees and Regression}
-\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=7mm]{python}
-import numpy as np
-import matplotlib.pyplot as plt
-from sklearn.preprocessing import PolynomialFeatures
-from sklearn.linear_model import LinearRegression
-
-steps=250
-
-distance=0
-x=0
-distance_list=[]
-steps_list=[]
-while x<steps:
-    distance+=np.random.randint(-1,2)
-    distance_list.append(distance)
-    x+=1
-    steps_list.append(x)
-plt.plot(steps_list,distance_list, color='green', label="Random Walk Data")
-
-steps_list=np.asarray(steps_list)
-distance_list=np.asarray(distance_list)
-
-X=steps_list[:,np.newaxis]
-
-#Polynomial fits
-
-#Degree 2
-poly_features=PolynomialFeatures(degree=2, include_bias=False)
-X_poly=poly_features.fit_transform(X)
-
-lin_reg=LinearRegression()
-poly_fit=lin_reg.fit(X_poly,distance_list)
-b=lin_reg.coef_
-c=lin_reg.intercept_
-print ("2nd degree coefficients:")
-print ("zero power: ",c)
-print ("first power: ", b[0])
-print ("second power: ",b[1])
-
-z = np.arange(0, steps, .01)
-z_mod=b[1]*z**2+b[0]*z+c
-
-fit_mod=b[1]*X**2+b[0]*X+c
-plt.plot(z, z_mod, color='r', label="2nd Degree Fit")
-plt.title("Polynomial Regression")
-
-plt.xlabel("Steps")
-plt.ylabel("Distance")
-
-#Degree 10
-poly_features10=PolynomialFeatures(degree=10, include_bias=False)
-X_poly10=poly_features10.fit_transform(X)
-
-poly_fit10=lin_reg.fit(X_poly10,distance_list)
-
-y_plot=poly_fit10.predict(X_poly10)
-plt.plot(X, y_plot, color='black', label="10th Degree Fit")
-
-plt.legend()
-plt.show()
-
-#Decision Tree Regression
-from sklearn.tree import DecisionTreeRegressor
-regr_1=DecisionTreeRegressor(max_depth=2)
-regr_2=DecisionTreeRegressor(max_depth=5)
-regr_3=DecisionTreeRegressor(max_depth=7)
-regr_1.fit(X, distance_list)
-regr_2.fit(X, distance_list)
-regr_3.fit(X, distance_list)
-
-X_test = np.arange(0.0, steps, 0.01)[:, np.newaxis]
-y_1 = regr_1.predict(X_test)
-y_2 = regr_2.predict(X_test)
-y_3=regr_3.predict(X_test)
-
-# Plot the results
-plt.figure()
-plt.scatter(X, distance_list, s=2.5, c="black", label="data")
-plt.plot(X_test, y_1, color="red",
-         label="max_depth=2", linewidth=2)
-plt.plot(X_test, y_2, color="green", label="max_depth=5", linewidth=2)
-plt.plot(X_test, y_3, color="m", label="max_depth=7", linewidth=2)
-
-plt.xlabel("Data")
-plt.ylabel("Darget")
-plt.title("Decision Tree Regression")
-plt.legend()
-plt.show()
-
-\end{minted}
-
-% !split
-\subsection*{Artificial neurons}
-
-% --- begin paragraph admon ---
-\paragraph{}
-
-The field of artificial neural networks has a long history of
-development, and is closely connected with the advancement of computer
-science and computers in general. A model of artificial neurons was
-first developed by McCulloch and Pitts in 1943 to study signal
-processing in the brain and has later been refined by others. The
-general idea is to mimic neural networks in the human brain, which is
-composed of billions of neurons that communicate with each other by
-sending electrical signals.  Each neuron accumulates its incoming
-signals, which must exceed an activation threshold to yield an
-output. If the threshold is not overcome, the neuron remains inactive,
-i.e.~has zero output.
-
-This behaviour has inspired a simple mathematical model for an artificial neuron.
-
-\[
- y = f\left(\sum_{i=1}^n w_ix_i\right) = f(u)
-\]
-Here, the output $y$ of the neuron is the value of its activation function, which have as input
-a weighted sum of signals $x_i, \dots ,x_n$ received by $n$ other neurons.
-% --- end paragraph admon ---
-
-% !split
-\subsection*{A simple perceptron model}
-
-\vspace{6mm}
-
-% inline figure
-\centerline{\includegraphics[width=0.9\linewidth]{figures/perceptron.png}}
-
-\vspace{6mm}
-
-% !split
-\subsection*{Neural network types}
-
-% --- begin paragraph admon ---
-\paragraph{}
-An artificial neural network (NN), is a computational model that consists of layers of connected neurons, or \emph{nodes}. 
-It is supposed to mimic a biological nervous system by letting each neuron interact with other neurons
-by sending signals in the form of mathematical functions between layers. 
-A wide variety of different NNs have
-been developed, but most of them consist of an input layer, an output layer and eventual layers in-between, called
-\emph{hidden layers}. All layers can contain an arbitrary number of nodes, and each connection between two nodes
-is associated with a weight variable.
-
-\vspace{6mm}
-
-% inline figure
-\centerline{\includegraphics[width=0.6\linewidth]{figures/dnn.png}}
-
-\vspace{6mm}
-% --- end paragraph admon ---
-
-% !split
-\subsection*{The system: two electrons in a harmonic oscillator trap in two dimensions}
-
-The Hamiltonian of the quantum dot is given by
-\[ \hat{H} = \hat{H}_0 + \hat{V}, 
-\]
-where $\hat{H}_0$ is the many-body HO Hamiltonian, and $\hat{V}$ is the
-inter-electron Coulomb interactions. In dimensionless units,
-\[ \hat{V}= \sum_{i < j}^N \frac{1}{r_{ij}},
-\]
-with $r_{ij}=\sqrt{\mathbf{r}_i^2 - \mathbf{r}_j^2}$.
-
-This leads to the  separable Hamiltonian, with the relative motion part given by ($r_{ij}=r$)
-\[ 
-\hat{H}_r=-\nabla^2_r + \frac{1}{4}\omega^2r^2+ \frac{1}{r},
-\]
-plus a standard Harmonic Oscillator problem  for the center-of-mass motion.
-This system has analytical solutions in two and three dimensions (\href{{https://journals.aps.org/pra/abstract/10.1103/PhysRevA.48.3561}}{M. Taut 1993 and 1994}).
-
-% !split
-\subsection*{Quantum Monte Carlo Motivation}
-
-% --- begin paragraph admon ---
-\paragraph{}
-Given a hamiltonian $H$ and a trial wave function $\Psi_T$, the variational principle states that the expectation value of $\langle H \rangle$, defined through 
-\[
-   \langle E \rangle =
-   \frac{\int d\bm{R}\Psi^{\ast}_T(\bm{R})H(\bm{R})\Psi_T(\bm{R})}
-        {\int d\bm{R}\Psi^{\ast}_T(\bm{R})\Psi_T(\bm{R})},
-\]
-is an upper bound to the ground state energy $E_0$ of the hamiltonian $H$, that is 
-\[
-    E_0 \le \langle H \rangle .
-\]
-In general, the integrals involved in the calculation of various  expectation values  are multi-dimensional ones.
-Traditional integration methods such as the Gauss-Legendre will not be adequate for say the  computation of the energy of a many-body system.
-% --- end paragraph admon ---
-
-% !split
-\subsection*{Quantum Monte Carlo Motivation}
-
-% --- begin paragraph admon ---
-\paragraph{Basic steps.}
-Choose a trial wave function
-$\psi_T(\bm{R})$.
-\[
-   P(\bm{R},\bm{\alpha})= \frac{\left|\psi_T(\bm{R},\bm{\alpha})\right|^2}{\int \left|\psi_T(\bm{R},\bm{\alpha})\right|^2d\bm{R}}.
-\]
-This is our model, or likelihood/probability distribution function  (PDF). It depends on some variational parameters $\bm{\alpha}$.
-The approximation to the expectation value of the Hamiltonian is now 
-\[
-   \langle E[\bm{\alpha}] \rangle = 
-   \frac{\int d\bm{R}\Psi^{\ast}_T(\bm{R},\bm{\alpha})H(\bm{R})\Psi_T(\bm{R},\bm{\alpha})}
-        {\int d\bm{R}\Psi^{\ast}_T(\bm{R},\bm{\alpha})\Psi_T(\bm{R},\bm{\alpha})}.
-\]
-% --- end paragraph admon ---
-
-% !split
-\subsection*{Quantum Monte Carlo Motivation}
-
-% --- begin paragraph admon ---
-\paragraph{Define a new quantity.}
-\[
-   E_L(\bm{R},\bm{\alpha})=\frac{1}{\psi_T(\bm{R},\bm{\alpha})}H\psi_T(\bm{R},\bm{\alpha}),
-\]
-called the local energy, which, together with our trial PDF yields
-\[
-  E[\bm{\alpha}]=\int P(\bm{R})E_L(\bm{R},\bm{\alpha}) d\bm{R}\approx \frac{1}{N}\sum_{i=1}^NE_L(\bm{R_i},\bm{\alpha})
-\]
-with $N$ being the number of Monte Carlo samples.
-% --- end paragraph admon ---
-
-% !split
-\subsection*{Quantum Monte Carlo}
-
-% --- begin paragraph admon ---
-\paragraph{}
-The Algorithm for performing a variational Monte Carlo calculations runs thus as this
-
-\begin{itemize}
-   \item Initialisation: Fix the number of Monte Carlo steps. Choose an initial $\bm{R}$ and variational parameters $\alpha$ and calculate $\left|\psi_T(\bm{R},\bm{\alpha})\right|^2$.
-
-   \item Initialise the energy and the variance and start the Monte Carlo calculation by looping over trials.
-\begin{itemize}
-
-      \item Calculate  a trial position  $\bm{R}_p=\bm{R}+r*step$ where $r$ is a random variable $r \in [0,1]$.
-
-      \item Metropolis algorithm to accept or reject this move  $w = P(\bm{R}_p,\bm{\alpha})/P(\bm{R},\bm{\alpha})$.
-
-      \item If the step is accepted, then we set $\bm{R}=\bm{R}_p$.
-
-      \item Update averages
-
-\end{itemize}
-
-\noindent
-   \item Finish and compute final averages.
-\end{itemize}
-
-\noindent
-Observe that the jumping in space is governed by the variable \emph{step}. This is often called brute-force sampling.
-Need importance sampling to get more relevant sampling.
-% --- end paragraph admon ---
-
-% !split
-\subsection*{The trial wave function}
-
-% --- begin paragraph admon ---
-\paragraph{}
-We want to perform  a Variational Monte Carlo calculation of the ground state of two electrons in a quantum dot well with different oscillator energies, assuming total spin $S=0$.
-Our trial wave function has the following form
-\begin{equation}
-   \psi_{T}(\bm{r}_1,\bm{r}_2) = 
-   C\exp{\left(-\alpha_1\omega(r_1^2+r_2^2)/2\right)}
-   \exp{\left(\frac{r_{12}}{(1+\alpha_2 r_{12})}\right)}, 
-\label{eq:trial}
-\end{equation}
-where the $\alpha$s represent our variational parameters, two in this case.
-
-Why does the trial function look like this? How did we get there? \textbf{This will be our main motivation} for switching to
-Machine Learning.
-% --- end paragraph admon ---
-
-% !split
-\subsection*{The correlation part of the wave function}
-
-To find an ansatz for the correlated part of the wave function, it is useful to rewrite the two-particle
-local energy in terms of the relative and center-of-mass motion.
-Let us denote the distance between the two electrons as
-$r_{12}$. We omit the center-of-mass motion since we are only interested in the case when
-$r_{12} \rightarrow 0$. The contribution from the center-of-mass (CoM) variable $\bm{R}_{\mathrm{CoM}}$
-gives only a finite contribution.
-We focus only on the terms that are relevant for $r_{12}$ and for three dimensions. The relevant local energy becomes then
-\[
-\lim_{r_{12} \rightarrow 0}E_L(R)=
-    \frac{1}{{\cal R}_T(r_{12})}\left(2\frac{d^2}{dr_{ij}^2}+\frac{4}{r_{ij}}\frac{d}{dr_{ij}}+
-\frac{2}{r_{ij}}-\frac{l(l+1)}{r_{ij}^2}+2E
-\right){\cal R}_T(r_{12}) = 0.
-\]
-Set $l=0$ and we have the so-called \textbf{cusp} condition
-\[
-\frac{d {\cal R}_T(r_{12})}{dr_{12}} = -\frac{1}{2(l+1)}
-{\cal R}_T(r_{12})\qquad r_{12}\to 0
-\]
-
-% !split
-\subsection*{Resulting ansatz}
-The above  results in
-\[
-{\cal R}_T  \propto \exp{(r_{ij}/2)}, 
-\]
-for anti-parallel spins and 
-\[
-{\cal R}_T  \propto \exp{(r_{ij}/4)}, 
-\]
-for anti-parallel spins. 
-This is the so-called cusp condition for the relative motion, resulting in a minimal requirement
-for the correlation part of the wave fuction.
-For general systems containing more than say two electrons, we have this
-condition for each electron pair $ij$.
-
-% !split
-\subsection*{The VMC code}
-
-\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=7mm]{python}
-# Importing various packages
-from math import exp, sqrt
-from random import random, seed
-import numpy as np
-import matplotlib.pyplot as plt
-from mpl_toolkits.mplot3d import Axes3D
-from matplotlib import cm
-from matplotlib.ticker import LinearLocator, FormatStrFormatter
-import sys
-
-#Trial wave function for quantum dots in two dims
-def WaveFunction(r,alpha,beta):
-    r1 = r[0,0]**2 + r[0,1]**2
-    r2 = r[1,0]**2 + r[1,1]**2
-    r12 = sqrt((r[0,0]-r[1,0])**2 + (r[0,1]-r[1,1])**2)
-    deno = r12/(1+beta*r12)
-    return exp(-0.5*alpha*(r1+r2)+deno)
-
-#Local energy  for quantum dots in two dims, using analytical local energy
-def LocalEnergy(r,alpha,beta):
-
-    r1 = (r[0,0]**2 + r[0,1]**2)
-    r2 = (r[1,0]**2 + r[1,1]**2)
-    r12 = sqrt((r[0,0]-r[1,0])**2 + (r[0,1]-r[1,1])**2)
-    deno = 1.0/(1+beta*r12)
-    deno2 = deno*deno
-    return 0.5*(1-alpha*alpha)*(r1 + r2) +2.0*alpha + 1.0/r12+deno2*(alpha*r12-deno2+2*beta*deno-1.0/r12)
-
-# The Monte Carlo sampling with the Metropolis algo
-def MonteCarloSampling():
-
-    NumberMCcycles= 100000
-    StepSize = 1.0
-    # positions
-    PositionOld = np.zeros((NumberParticles,Dimension), np.double)
-    PositionNew = np.zeros((NumberParticles,Dimension), np.double)
-    # seed for rng generator
-    seed()
-    # start variational parameter
-    alpha = 0.9
-    for ia in range(MaxVariations):
-        alpha += .025
-        AlphaValues[ia] = alpha
-        beta = 0.2 
-        for jb in range(MaxVariations):
-            beta += .01
-            BetaValues[jb] = beta
-            energy = energy2 = 0.0
-            DeltaE = 0.0
-            #Initial position
-            for i in range(NumberParticles):
-                for j in range(Dimension):
-                    PositionOld[i,j] = StepSize * (random() - .5)
-            wfold = WaveFunction(PositionOld,alpha,beta)
-
-            #Loop over MC MCcycles
-            for MCcycle in range(NumberMCcycles):
-                #Trial position
-                for i in range(NumberParticles):
-                    for j in range(Dimension):
-                        PositionNew[i,j] = PositionOld[i,j] + StepSize * (random() - .5)
-                wfnew = WaveFunction(PositionNew,alpha,beta)
-
-                #Metropolis test to see whether we accept the move
-                if random() < wfnew**2 / wfold**2:
-                   PositionOld = PositionNew.copy()
-                   wfold = wfnew
-                   DeltaE = LocalEnergy(PositionOld,alpha,beta)
-                energy += DeltaE
-                energy2 += DeltaE**2
-
-            #We calculate mean, variance and error ...
-            energy /= NumberMCcycles
-            energy2 /= NumberMCcycles
-            variance = energy2 - energy**2
-            error = sqrt(variance/NumberMCcycles)
-            Energies[ia,jb] = energy    
-    return Energies, AlphaValues, BetaValues
-
-#Here starts the main program with variable declarations
-NumberParticles = 2
-Dimension = 2
-MaxVariations = 10
-Energies = np.zeros((MaxVariations,MaxVariations))
-AlphaValues = np.zeros(MaxVariations)
-BetaValues = np.zeros(MaxVariations)
-(Energies, AlphaValues, BetaValues) = MonteCarloSampling()
-
-# Prepare for plots
-fig = plt.figure()
-ax = fig.gca(projection='3d')
-# Plot the surface.
-X, Y = np.meshgrid(AlphaValues, BetaValues)
-surf = ax.plot_surface(X, Y, Energies,cmap=cm.coolwarm,linewidth=0, antialiased=False)
-# Customize the z axis.
-zmin = np.matrix(Energies).min()
-zmax = np.matrix(Energies).max()
-ax.set_zlim(zmin, zmax)
-ax.set_xlabel(r'$\alpha$')
-ax.set_ylabel(r'$\beta$')
-ax.set_zlabel(r'$\langle E \rangle$')
-ax.zaxis.set_major_locator(LinearLocator(10))
-ax.zaxis.set_major_formatter(FormatStrFormatter('%.02f'))
-# Add a color bar which maps values to colors.
-fig.colorbar(surf, shrink=0.5, aspect=5)
-plt.show()
-
-\end{minted}
-
-% !split
-\subsection*{Technical aspect, improvements and how to define the cost function}
-
-% --- begin paragraph admon ---
-\paragraph{}
-
-\textbf{The above procedure is not the smartest one}. Looping over all variational parameters becomes expensive.
-Also, we don't use importance sampling and optimizations of the standard deviation (blocking, bootstrap, jackknife). 
-Such codes are included in the above Github address.
-
-We can also be smarter and use minimization methods to find the \textbf{optimal} variational parameters with fewer Monte Carlo cycles and then 
-fire up our heavy artillery.
-
-One way to achieve this is to minimize the energy as function of the variational parameters.
-% --- end paragraph admon ---
-
-% !split
-\subsection*{Energy derivatives}
-
-% --- begin paragraph admon ---
-\paragraph{}
-To find the derivatives of the local energy expectation value as function of the variational parameters, we can use the chain rule and the hermiticity of the Hamiltonian.
-
-Let us define 
-\[
-\bar{E}_{\alpha_i}=\frac{d\langle  E_L\rangle}{d\alpha_i}.
-\]
-as the derivative of the energy with respect to the variational parameter $\alpha_i$
-We define also the derivative of the trial function (skipping the subindex $T$) as 
-\[
-\bar{\Psi}_{i}=\frac{d\Psi}{d\alpha_i}.
-\]
-% --- end paragraph admon ---
-
-% !split
-\subsection*{Derivatives of the local energy}
-
-% --- begin paragraph admon ---
-\paragraph{}
-The elements of the gradient of the local energy are then (using the chain rule and the hermiticity of the Hamiltonian)
-\[
-\bar{E}_{i}= 2\left( \langle \frac{\bar{\Psi}_{i}}{\Psi}E_L\rangle -\langle \frac{\bar{\Psi}_{i}}{\Psi}\rangle\langle E_L \rangle\right).
-\]
-From a computational point of view it means that you need to compute the expectation values of 
-\[
-\langle \frac{\bar{\Psi}_{i}}{\Psi}E_L\rangle,
-\]
-and
-\[
-\langle \frac{\bar{\Psi}_{i}}{\Psi}\rangle\langle E_L\rangle
-\]
-These integrals are evaluted using MC intergration (with all its possible error sources). 
-We can then use methods like stochastic gradient or other minimization methods to find the optimal variational parameters (I don't discuss this topic here, but these methods are very important in ML).
-% --- end paragraph admon ---
-
-% !split
-\subsection*{How do we define our cost function?}
-
-% --- begin paragraph admon ---
-\paragraph{}
-We have a model, our likelihood function.
-
-How should we define the cost function?
-% --- end paragraph admon ---
-
-% !split
-\subsection*{Meet the variance and its derivatives}
-
-% --- begin paragraph admon ---
-\paragraph{Why the variance?}
-Suppose the trial function (our model) is the exact wave function. The action of the hamiltionan
-on the wave function
-\[
-   H\Psi = \mathrm{constant}\times \Psi,
-\]
-The integral which defines various 
-expectation values involving moments of the hamiltonian becomes then
-\[
-   \langle E^n \rangle =   \langle H^n \rangle =
-   \frac{\int d\bm{R}\Psi^{\ast}(\bm{R})H^n(\bm{R})\Psi(\bm{R})}
-        {\int d\bm{R}\Psi^{\ast}(\bm{R})\Psi(\bm{R})}=
-\mathrm{constant}\times\frac{\int d\bm{R}\Psi^{\ast}(\bm{R})\Psi(\bm{R})}
-        {\int d\bm{R}\Psi^{\ast}(\bm{R})\Psi(\bm{R})}=\mathrm{constant}.
-\]
-\textbf{This gives an important information: If I want the variance, the exact wave function leads to zero variance!}
-The variance is defined as 
-\[
-\sigma_E = \langle E^2\rangle - \langle E\rangle^2.
-\]
-Variation is then performed by minimizing both the energy and the variance.
-% --- end paragraph admon ---
-
-% !split
-\subsection*{The variance defines the cost function}
-
-% --- begin paragraph admon ---
-\paragraph{}
-We can then take the derivatives of 
-\[
-\sigma_E = \langle E^2\rangle - \langle E\rangle^2,
-\]
-with respect to the variational parameters. The derivatives of the variance can then be used to defined the
-so-called Hessian matrix, which in turn allows us to use minimization methods like Newton's method or 
-standard gradient methods.
-
-This leads to however a more complicated expression, with obvious errors when evaluating integrals by Monte Carlo integration.
-Less used, see however \href{{https://journals.aps.org/prl/abstract/10.1103/PhysRevLett.94.150201}}{Filippi and Umrigar}.
-The expression becomes complicated
-\[
-\bar{E}_{ij} = 2\left[ \langle (\frac{\bar{\Psi}_{ij}}{\Psi}+\frac{\bar{\Psi}_{j}}{\Psi}\frac{\bar{\Psi}_{i}}{\Psi})(E_L-\langle E\rangle)\rangle -\langle \frac{\bar{\Psi}_{i}}{\Psi}\rangle\bar{E}_j-\langle \frac{\bar{\Psi}_{j}}{\Psi}\rangle\bar{E}_i\right] +\langle \frac{\bar{\Psi}_{i}}{\Psi}E_L{_j}\rangle +\langle \frac{\bar{\Psi}_{j}}{\Psi}E_L{_i}\rangle -\langle \frac{\bar{\Psi}_{i}}{\Psi}\rangle\langle E_L{_j}\rangle \langle \frac{\bar{\Psi}_{j}}{\Psi}\rangle\langle E_L{_i}\rangle.
-\]
-
-Evaluating the cost function means having to evaluate the above second derivative of the energy.
-% --- end paragraph admon ---
-
-% !split
-\subsection*{Why Boltzmann machines?}
-
-What is known as restricted Boltzmann Machines (RMB) have received a lot of attention lately. 
-One of the major reasons is that they can be stacked layer-wise to build deep neural networks that capture complicated statistics.
-
-The original RBMs had just one visible layer and a hidden layer, but recently so-called Gaussian-binary RBMs have gained quite some popularity in imaging since they are capable of modeling continuous data that are common to natural images.
-
-Furthermore, they have been used to solve complicated quantum mechanical many-particle problems or classical statistical physics problems like the Ising and Potts classes of models.
-
-% !split
-\subsection*{Boltzmann Machines}
-
-Why use a generative model rather than the more well known discriminative deep neural networks (DNN)?
-
-\begin{itemize}
-\item Discriminitave methods have several limitations: They are mainly supervised learning methods, thus requiring labeled data.
-And there are tasks they cannot accomplish, like drawing new examples from an unknown probability distribution.
-
-\item A generative model can learn to represent and sample from a probability distribution.
-The core idea is to learn a parametric model of the probability distribution from which the training data was drawn.
-As an example
-\begin{enumerate}
-
- \item A model for images could learn to draw new examples of cats and dogs, given a training dataset of images of cats and dogs.
-
- \item Generate a sample of an ordered or disordered Ising model phase, having been given samples of such phases.
-
- \item Model the trial function for Monte Carlo calculations
-\end{enumerate}
-
-\noindent
-\end{itemize}
-
-\noindent
-% !split
-\subsection*{Some similarities and differences from DNNs}
-
-\begin{enumerate}
-\item Both use gradient-descent based learning procedures for minimizing cost functions
-
-\item Energy based models don't use backpropagation and automatic differentiation for computing gradients, instead turning to Markov Chain Monte Carlo methods.
-
-\item DNNs often have several hidden layers.
-A restricted Boltzmann machine has only one hidden layer, however several RBMs can be stacked to make up Deep Belief Networks, of which they constitute the building blocks.
-\end{enumerate}
-
-\noindent
-History: The RBM was developed by amongst others Geoffrey Hinton, called by some the "Godfather of Deep Learning", working with the University of Toronto and Google.
-
-% !split
-\subsection*{Boltzmann machines (BM)}
-
-% --- begin paragraph admon ---
-\paragraph{}
-A BM is what we would call an undirected probabilistic graphical model
-with stochastic continuous or discrete units.
-% --- end paragraph admon ---
-
-% --- begin paragraph admon ---
-\paragraph{}
-It is interpreted as a stochastic recurrent neural network where the
-state of each unit(neurons/nodes) depends on the units it is connected
-to. The weights in the network represent thus the strength of the
-interaction between various units/nodes.
-% --- end paragraph admon ---
-
-% --- begin paragraph admon ---
-\paragraph{}
-It turns into a Hopfield network if we choose deterministic rather
-than stochastic units. In contrast to a Hopfield network, a BM is a
-so-called generative model. It allows us to generate new samples from
-the learned distribution.
-% --- end paragraph admon ---
-
-% !split
-\subsection*{A standard BM setup}
-
-% --- begin paragraph admon ---
-\paragraph{}
-A standard BM network is divided into a set of observable and visible units $\hat{x}$ and a set of unknown hidden units/nodes $\hat{h}$.
-% --- end paragraph admon ---
-
-% --- begin paragraph admon ---
-\paragraph{}
-Additionally there can be bias nodes for the hidden and visible layers. These biases are normally set to $1$.
-% --- end paragraph admon ---
-
-% --- begin paragraph admon ---
-\paragraph{}
-BMs are stackable, meaning they cwe can train a BM which serves as input to another BM.
-We can construct deep networks for learning complex PDFs.
-The layers can be trained one after another, a feature which makes them popular in deep learning
-% --- end paragraph admon ---
-
-However, they are often hard to train. This leads to the introduction of so-called restricted BMs, or RBMS.
-Here we take away all lateral connections between nodes in the visible layer as well as connections between nodes in the hidden layer. The network is illustrated in the figure below.
-
-% !split
-\subsection*{The structure of the RBM network}
-
-\vspace{6mm}
-
-% inline figure
-\centerline{\includegraphics[width=1.0\linewidth]{figures/RBM.pdf}}
-
-\vspace{6mm}
-
-% !split
-\subsection*{The network}
-
-\textbf{The network layers}:
-\begin{enumerate}
- \item A function $\mathbf{x}$ that represents the visible layer, a vector of $M$ elements (nodes).
-This layer represents both what the RBM might be given as training input, and what we want it to be able to reconstruct.
-This might for example be the pixels of an image, the spin values of the Ising model, or coefficients representing speech.
-
- \item The function $\mathbf{h}$ represents the hidden, or latent, layer. A vector of $N$ elements (nodes). Also called "feature detectors".
-\end{enumerate}
-
-\noindent
-% !split
-\subsection*{Goals}
-
-The goal of the hidden layer is to increase the model's expressive power.
-We encode complex interactions between visible variables by introducing additional, hidden variables that interact with visible degrees of freedom in a simple manner, yet still reproduce the complex correlations between visible degrees in the data once marginalized over (integrated out).
-
-Examples of this trick being employed in physics: 
-\begin{enumerate}
- \item The Hubbard-Stratonovich transformation
-
- \item The introduction of ghost fields in gauge theory
-
- \item Shadow wave functions in Quantum Monte Carlo simulations
-\end{enumerate}
-
-\noindent
-\textbf{The network parameters, to be optimized/learned}:
-\begin{enumerate}
- \item $\mathbf{a}$ represents the visible bias, a vector of same length as $\mathbf{x}$.
-
- \item $\mathbf{b}$ represents the hidden bias, a vector of same lenght as $\mathbf{h}$.
-
- \item $W$ represents the interaction weights, a matrix of size $M\times N$.
-\end{enumerate}
-
-\noindent
-% !split
-\subsection*{Joint distribution}
-
-The restricted Boltzmann machine is described by a Boltzmann distribution
-\begin{align}
-	P_{rbm}(\mathbf{x},\mathbf{h}) = \frac{1}{Z} e^{-\frac{1}{T_0}E(\mathbf{x},\mathbf{h})},
-\end{align}
-where $Z$ is the normalization constant or partition function, defined as 
-\begin{align}
-	Z = \int \int e^{-\frac{1}{T_0}E(\mathbf{x},\mathbf{h})} d\mathbf{x} d\mathbf{h}.
-\end{align}
-It is common to ignore $T_0$ by setting it to one.
-
-% !split
-\subsection*{Network Elements, the energy function}
-
-The function $E(\mathbf{x},\mathbf{h})$ gives the \textbf{energy} of a
-configuration (pair of vectors) $(\mathbf{x}, \mathbf{h})$. The lower
-the energy of a configuration, the higher the probability of it. This
-function also depends on the parameters $\mathbf{a}$, $\mathbf{b}$ and
-$W$. Thus, when we adjust them during the learning procedure, we are
-adjusting the energy function to best fit our problem.
-
-An expression for the energy function is
-\[
-E(\hat{x},\hat{h}) = -\sum_{ia}^{NA}b_i^a \alpha_i^a(x_i)-\sum_{jd}^{MD}c_j^d \beta_j^d(h_j)-\sum_{ijad}^{NAMD}b_i^a \alpha_i^a(x_i)c_j^d \beta_j^d(h_j)w_{ij}^{ad}.
-\]
-
-Here $\beta_j^d(h_j)$ and $\alpha_i^a(x_j)$ are so-called transfer functions that map a given input value to a desired feature value.
-The labels $a$ and $d$ denote that there can be multiple transfer functions per variable.
-The first sum depends only on the visible units.
-The second on the hidden ones.
-\textbf{Note} that there is no connection between nodes in a layer.
-
-The quantities $b$ and $c$ can be interpreted as the visible and hidden biases, respectively.
-
-The connection between the nodes in the two layers is given by the weights $w_{ij}$.
-
-% !split
-\subsection*{Defining different types of RBMs}
-There are different variants of RBMs, and the differences lie in the types of visible and hidden units we choose as well as in the implementation of the energy function $E(\mathbf{x},\mathbf{h})$.
-
-% --- begin paragraph admon ---
-\paragraph{Binary-Binary RBM:}
-
-RBMs were first developed using binary units in both the visible and hidden layer. The corresponding energy function is defined as follows:
-\begin{align}
-	E(\mathbf{x}, \mathbf{h}) = - \sum_i^M x_i a_i- \sum_j^N b_j h_j - \sum_{i,j}^{M,N} x_i w_{ij} h_j,
-\end{align}
-where the binary values taken on by the nodes are most commonly 0 and 1.
-% --- end paragraph admon ---
-
-% --- begin paragraph admon ---
-\paragraph{Gaussian-Binary RBM:}
-
-Another varient is the RBM where the visible units are Gaussian while the hidden units remain binary:
-\begin{align}
-	E(\mathbf{x}, \mathbf{h}) = \sum_i^M \frac{(x_i - a_i)^2}{2\sigma_i^2} - \sum_j^N b_j h_j - \sum_{i,j}^{M,N} \frac{x_i w_{ij} h_j}{\sigma_i^2}. 
-\end{align}
-% --- end paragraph admon ---
-
-% !split
-\subsection*{More about RBMs}
-\begin{enumerate}
-\item Useful when we model continuous data (i.e., we wish $\mathbf{x}$ to be continuous)
-
-\item Requires a smaller learning rate, since there's no upper bound to the value a component might take in the reconstruction
-\end{enumerate}
-
-\noindent
-Other types of units include:
-\begin{enumerate}
-\item Softmax and multinomial units
-
-\item Gaussian visible and hidden units
-
-\item Binomial units
-
-\item Rectified linear units
-\end{enumerate}
-
-\noindent
-% !split
-\subsection*{Sampling: Metropolis sampling}
-In order to sample from the RBM probability distribution it is common to use Markov Chain Monte Carlo (MCMC) algorithms such as Metropolis-Hastings or Gibbs sampling.
-
-Metropolis sampling starts by suggesting a new configuration $\bm{x}^{k+1}$.
-In the brute force method this is done by some random change of the visible units.
-The new configuration is then accepted with the acceptance probability
-\begin{align}
-	A(\bm{x}^k \rightarrow \bm{x}^{k+1}) = \text{min} (1, \frac{P(\bm{x}^{k+1})}{P(\bm{x}^k)}),
-\end{align} 
-where we need the marginalized probability
-\begin{align}
-	P(\bm{x})  &= \sum_\mathbf{h} P_{rbm}(\mathbf{x}, \mathbf{h}) \\
-				&= \frac{1}{Z}\sum_\mathbf{h} e^{-E(\mathbf{x}, \mathbf{h})}.
-\end{align}
-
-% !split
-\subsection*{Sampling: Gibbs sampling}
-
-In this method we sample from the joint probability $P_{rbm} (\mathbf{x}, \mathbf{h})$ by way of a two step sampling process. We alternately update the visible and hidden units.
-New samples are generated according to the conditional probabilities $P(x_i|\mathbf{h})$ and $P(h_j|\mathbf{x})$ respectively and accepted with the probability of $1$.
-While the the visible nodes are dependent on the hidden nodes and vice versa, the nodes are independent of other nodes within the same layer.
-This is due to there being no intra layer interactions in the restricted Boltzmann machine.
-
-The conditional probabilities are often referred to as the activitation functions in the neural networks context due to their role in determining the node outputs. For the binary-binary RBM they are
-\begin{align}
-	P(h_j = 1 | \bm{x}) &= \frac{1}{1 + e^{-b_j - \sum_i x_i w_{ij}}} \\
-	P(x_i = 1 | \bm{h}) &= \frac{1}{1 + e^{-a_j - \sum_j h_j w_{ij}}},
-\end{align}
-where we recognize the logistic sigmoid function $\sigma (x) = 1/(1+exp(-x))$.
-
-% !split
-\subsection*{Gaussian RBM}
-For the Gaussian-Binary RBM the conditional probabilities are
-\begin{align}
-	P(x_i|\mathbf{h}) &= \mathcal{N}(x_i; a_i+ \sum_j h_j w_{ij}, \sigma^2) \\
-	P(h_j=1|\mathbf{x}) &=  \frac{1}{1+e^{-b_j-\frac{1}{\sigma^2} \sum_i x_i w_{ij}}},
-\end{align}
-while the visible units now follow a normal distribution, we see the hidden units again follow the logistic sigmoid function.
-
-% !split
-\subsection*{Cost function}
-
-When working with a training dataset, the most common training approach is maximizing the log-likelihood of the training data.
-The log likelihood characterizes the log-probability of generating the observed data using our generative model.
-Using this method our cost function is chosen as the negative log-likelihood.
-The learning then consists of trying to find parameters that maximize the probability of the dataset, and is known as Maximum Likelihood Estimation (MLE).
-Denoting the parameters as $\bm{\theta} = a_1,...,a_M,b_1,...,b_N,w_{11},...,w_{MN}$, the log-likelihood is given by
-\begin{align}
-	\mathcal{L}(\{ \theta_i \}) &= \langle \text{log} P_\theta(\bm{x}) \rangle_{data} \\
-	&= - \langle E(\bm{x}; \{ \theta_i\}) \rangle_{data} - \text{log} Z(\{ \theta_i\}),
-\end{align}
-where we used that the normalization constant does not depend on the data, $\langle \text{log} Z(\{ \theta_i\}) \rangle = \text{log} Z(\{ \theta_i\})$
-Our cost function is the negative log-likelihood, $\mathcal{C}(\{ \theta_i \}) = - \mathcal{L}(\{ \theta_i \})$
-
-% !split
-\subsection*{Optimization / Training}
-
-The training procedure of choice often is Stochastic Gradient Descent (SGD). It consists of a series of iterations where we update the parameters according to the equation
-\begin{align}
-	\bm{\theta}_{k+1} = \bm{\theta}_k - \eta \nabla \mathcal{C} (\bm{\theta}_k)
-\end{align}
-at each $k$-th iteration. There are a range of variants of the algorithm which aim at making the learning rate $\eta$ more adaptive so the method might be more efficient while remaining stable.
-
-We now need the gradient of the cost function in order to minimize it. We find that
-\begin{align}
-	\frac{\partial \mathcal{C}(\{ \theta_i\})}{\partial \theta_i}
-	&= \langle \frac{\partial E(\bm{x}; \theta_i)}{\partial \theta_i} \rangle_{data}
-	+ \frac{\partial \text{log} Z(\{ \theta_i\})}{\partial \theta_i} \\
-	&= \langle O_i(\bm{x}) \rangle_{data} - \langle O_i(\bm{x}) \rangle_{model},
-\end{align}
-where in order to simplify notation we defined the "operator"
-\begin{align}
-	O_i(\bm{x}) = \frac{\partial E(\bm{x}; \theta_i)}{\partial \theta_i}, 
-\end{align}
-and used the statistical mechanics relationship between expectation values and the log-partition function:
-\begin{align}
-	\langle O_i(\bm{x}) \rangle_{model} = \text{Tr} P_\theta(\bm{x})O_i(\bm{x}) = - \frac{\partial \text{log} Z(\{ \theta_i\})}{\partial \theta_i}.
-\end{align}
-
-% !split 
-\subsection*{More on RBMs}
-
-The data-dependent term in the gradient is known as the positive phase of the gradient, while the model-dependent term is known as the negative phase of the gradient.
-The aim of the training is to lower the energy of configurations that are near observed data points (increasing their probability), and raising the energy of configurations that are far from observed data points (decreasing their probability).
-
-The gradient of the negative log-likelihood cost function of a Binary-Binary RBM is then
-\begin{align}
-	\frac{\partial \mathcal{C} (w_{ij}, a_i, b_j)}{\partial w_{ij}} =& \langle x_i h_j \rangle_{data} - \langle x_i h_j \rangle_{model} \\
-	\frac{\partial \mathcal{C} (w_{ij}, a_i, b_j)}{\partial a_{ij}} =& \langle x_i \rangle_{data} - \langle x_i \rangle_{model} \\
-	\frac{\partial \mathcal{C} (w_{ij}, a_i, b_j)}{\partial b_{ij}} =& \langle h_i \rangle_{data} - \langle h_i \rangle_{model}. \\
-\end{align}
-To get the expecation values with respect to the \emph{data}, we set the visible units to each of the observed samples in the training data, then update the hidden units according to the conditional probability found before.
-We then average over all samples in the training data to calculate expectation values with respect to the data.
-
-% !split
-\subsection*{Which sampling to use}
-
-To get the expectation values with respect to the \emph{model}, we use Gibbs sampling.
-We can either initialize the $\bm{x}$ randomly or with a training sample.
-While we ideally want a large number of Gibbs iterations $n\rightarrow n$, one might decide to truncate it earlier for efficiency.
-Doing this while having intialized $\bm{x}$ with a training data vector is referred to as contrastive divergence (CD), because one is then closer to approximating the gradient of this function than the negative log-likelihood.
-The contrastive divergence function is the difference between two Kullback-Leibler divergences (also called relative entropy), which measure how one probability distribution diverges from a second, expected probability distribution (in this case the estimated one from the ground truth one).
-
-% !split
-\subsection*{RBMs for the quantum many body problem}
-
-The idea of applying RBMs to quantum many body problems was presented by G. Carleo and M. Troyer, working with ETH Zurich and Microsoft Research.
-
-Some of their motivation included
-
-\begin{itemize}
-\item "The wave function $\Psi$ is a monolithic mathematical quantity that contains all the information on a quantum state, be it a single particle or a complex molecule.
-In principle, an exponential amount of information is needed to fully encode a generic many-body quantum state."
-
-\item There are still interesting open problems, including fundamental questions ranging from the dynamical properties of high-dimensional systems to the exact ground-state properties of strongly interacting fermions.
-
-\item The difficulty lies in finding a general strategy to reduce the exponential complexity of the full many-body wave function down to its most essential features. That is
-\begin{enumerate}
-
-\item $\rightarrow$ Dimensional reduction
-
-\item $\rightarrow$ Feature extraction
-
-\end{enumerate}
-
-\noindent
-\item Among the most successful techniques to attack these challenges, artifical neural networks play a prominent role.
-
-\item Want to understand whether an artifical neural network may adapt to describe a quantum system.
-\end{itemize}
-
-\noindent
-% !split
-\subsection*{Choose the right RBM}
-
-Carleo and Troyer applied the RBM to the quantum mechanical spin lattice systems of the Ising model and Heisenberg model, with encouraging results.
-Our goal is to test the method on systems of moving particles.
-For the spin lattice systems it was natural to use a binary-binary RBM, with the nodes taking values of 1 and -1.
-For moving particles, on the other hand, we want the visible nodes to be continuous, representing position coordinates.
-Thus, we start by choosing a Gaussian-binary RBM, where the visible nodes are continuous and hidden nodes take on values of 0 or 1.
-If eventually we would like the hidden nodes to be continuous as well the rectified linear units seem like the most relevant choice.
-
-% !split
-\subsection*{Representing the wave function}
-The wavefunction should be a probability amplitude depending on $\bm{x}$. The RBM model is given by the joint distribution of $\bm{x}$ and $\bm{h}$
-\begin{align}
-	F_{rbm}(\mathbf{x},\mathbf{h}) = \frac{1}{Z} e^{-\frac{1}{T_0}E(\mathbf{x},\mathbf{h})}.
-\end{align}
-To find the marginal distribution of $\bm{x}$ we set:
-\begin{align}
-	F_{rbm}(\mathbf{x}) &= \sum_\mathbf{h} F_{rbm}(\mathbf{x}, \mathbf{h}) \\
-				&= \frac{1}{Z}\sum_\mathbf{h} e^{-E(\mathbf{x}, \mathbf{h})}.
-\end{align}
-Now this is what we use to represent the wave function, calling it a neural-network quantum state (NQS)
-\begin{align}
-	\Psi (\mathbf{X}) &= F_{rbm}(\mathbf{x}) \\
-	&= \frac{1}{Z}\sum_{\bm{h}} e^{-E(\mathbf{x}, \mathbf{h})} \\
-	&= \frac{1}{Z} \sum_{\{h_j\}} e^{-\sum_i^M \frac{(x_i - a_i)^2}{2\sigma^2} + \sum_j^N b_j h_j + \sum_{i,j}^{M,N} \frac{x_i w_{ij} h_j}{\sigma^2}} \\
-	&= \frac{1}{Z} e^{-\sum_i^M \frac{(x_i - a_i)^2}{2\sigma^2}} \prod_j^N (1 + e^{b_j + \sum_i^M \frac{x_i w_{ij}}{\sigma^2}}). \\
-\end{align}
-
-% !split
-\subsection*{Choose the cost function}
-Now we don't necessarily have training data (unless we generate it by using some other method).
-However, what we do have is the variational principle which allows us to obtain the ground state wave function by minimizing the expectation value of the energy of a trial wavefunction (corresponding to the untrained NQS).
-Similarly to the traditional variational Monte Carlo method then, it is the local energy we wish to minimize.
-The gradient to use for the stochastic gradient descent procedure is
-\begin{align}
-	G_i = \frac{\partial \langle E_L \rangle}{\partial \theta_i}
-	= 2(\langle E_L \frac{1}{\Psi}\frac{\partial \Psi}{\partial \theta_i} \rangle - \langle E_L \rangle \langle \frac{1}{\Psi}\frac{\partial \Psi}{\partial \theta_i} \rangle ),
-\end{align}
-where the local energy is given by
-\begin{align}
-	E_L = \frac{1}{\Psi} \hat{\mathbf{H}} \Psi.
-\end{align}
-
-% !split
-\subsection*{Running the codes}
-
-% --- begin paragraph admon ---
-\paragraph{}
-You can find the codes for the simple two-electron case at the Github repository \href{{https://github.com/mhjensenseminars/MachineLearningTalk/tree/master/doc/Programs/MLcpp/src}}{\nolinkurl{https://github.com/mhjensenseminars/MachineLearningTalk/tree/master/doc/Programs/MLcpp/src}}.
-Python codes to come, only c++ as of now.
-
-The trial wave function is based on the product of a Slater determinant with Gaussian orbitals, a simple Jastrow factor $\exp{(r_{ij})}$ and the reduced Boltzmann machines.
-
-The Broyden-Fletcher-Goldfarb-Shanno algorithm was used for the minimization. We used $14$ hidden nodes in the calculations below.
-% --- end paragraph admon ---
-
-% !split
-\subsection*{Energy as function of iterations, $N=2$ electrons}
-
-% --- begin paragraph admon ---
-\paragraph{}
-
-\vspace{6mm}
-
-% inline figure
-\centerline{\includegraphics[width=0.9\linewidth]{figures/figN2.pdf}}
-
-\vspace{6mm}
-% --- end paragraph admon ---
-
-% !split
-\subsection*{Energy as function of iterations, $N=6$ electrons}
-
-% --- begin paragraph admon ---
-\paragraph{}
-
-\vspace{6mm}
-
-% inline figure
-\centerline{\includegraphics[width=0.9\linewidth]{figures/figN6.pdf}}
-
-\vspace{6mm}
-% --- end paragraph admon ---
-
-% !split
-\subsection*{Conclusions and where do we stand}
-
-% --- begin paragraph admon ---
-\paragraph{}
-\begin{itemize}
-\item A simple extension of the work of \href{{http://science.sciencemag.org/content/355/6325/602}}{G.
-Carleo and M.
-Troyer, Science \textbf{355}, Issue 6325, pp.
-602-606 (2017)} gives excellent results for two-electron systems as well as good agreement with standard VMC calculations for $N=6$ and $N=12$ electrons.
-
-\item Minimization problem can be tricky.
-
-\item Anti-symmetry dealt with multiplying the trail wave function with an optimized Slater determinant.
-
-\item To come: Analysis of wave function from ML and compare with diffusion and Variational Monte Carlo calculations as well as the analytical results of Taut for the two-electron case.
-
-\item Extend to more fermions. How do we deal with the antisymmetry of the multi-fermion wave function?
-\begin{enumerate}
-
- \item Here we used standard Hartree-Fock theory to define an optimal Slater determinant. Takes care of the antisymmetry. What about constructing an anti-symmetrized network function?
-
- \item Use thereafter ML to determine the correlated part of the wafe function (including a standard Jastrow factor).
-
- \item Test this for multi-fermion systems and compare with other many-body methods.
-
-\end{enumerate}
-
-\noindent
-\item Can we use ML to find out which correlations are relevant and thereby diminish the dimensionality problem in say CC or SRG theories?
-\end{itemize}
-
-\noindent
-% --- end paragraph admon ---
-
-% !split
-\subsection*{Additional material}
-
-% !split
-\subsection*{Kullback-Leibler relative entropy}
-
-When the goal of the training is to approximate a probability
-distribution, as it is in generative modeling, another relevant
-measure is the \textbf{Kullback-Leibler divergence}, also known as the
-relative entropy or Shannon entropy. It is a non-symmetric measure of the
-dissimilarity between two probability density functions $p$ and
-$q$. If $p$ is the unkown probability which we approximate with $q$,
-we can measure the difference by
-\begin{align}
-	\text{KL}(p||q) = \int_{-\infty}^{\infty} p (\bm{x}) \log \frac{p(\bm{x})}{q(\bm{x})}  d\bm{x}.
-\end{align}
-
-Thus, the Kullback-Leibler divergence between the distribution of the
-training data $f(\bm{x})$ and the model distribution $p(\bm{x}|
-\bm{\theta})$ is
-
-\begin{align}
-	\text{KL} (f(\bm{x})|| p(\bm{x}| \bm{\theta})) =& \int_{-\infty}^{\infty}
-	f (\bm{x}) \log \frac{f(\bm{x})}{p(\bm{x}| \bm{\theta})} d\bm{x} \\
-	=& \int_{-\infty}^{\infty} f(\bm{x}) \log f(\bm{x}) d\bm{x} - \int_{-\infty}^{\infty} f(\bm{x}) \log
-	p(\bm{x}| \bm{\theta}) d\bm{x} \\
-	%=& \mathbb{E}_{f(\bm{x})} (\log f(\bm{x})) - \mathbb{E}_{f(\bm{x})} (\log p(\bm{x}| \bm{\theta}))
-	=& \langle \log f(\bm{x}) \rangle_{f(\bm{x})} - \langle \log p(\bm{x}| \bm{\theta}) \rangle_{f(\bm{x})} \\
-	=& \langle \log f(\bm{x}) \rangle_{data} + \langle E(\bm{x}) \rangle_{data} + \log Z \\
-	=& \langle \log f(\bm{x}) \rangle_{data} + \mathcal{C}_{LL} .
-\end{align}
-
-The first term is constant with respect to $\bm{\theta}$ since $f(\bm{x})$ is independent of $\bm{\theta}$.
-Thus the Kullback-Leibler Divergence is minimal when the second term is minimal.
-The second term is the log-likelihood cost function, hence minimizing the Kullback-Leibler divergence is equivalent to maximizing the log-likelihood.
-
-% !split
-\subsection*{Optimizing the cost function}
-
-To further understand generative models it is useful to study the
-gradient of the cost function which is needed in order to minimize it
-using methods like stochastic gradient descent.
-
-The partition function is the generating function of
-expectation values, in particular there are mathematical relationships
-between expectation values and the log-partition function. In this
-case we have
-\begin{align}
-	\langle \frac{ \partial E(\bm{x}; \theta_i) } { \partial \theta_i} \rangle_{model}
-	= \int p(\bm{x}| \bm{\theta}) \frac{ \partial E(\bm{x}; \theta_i) } { \partial \theta_i} d\bm{x} 
-	= -\frac{\partial \log Z(\theta_i)}{ \partial  \theta_i} .
-\end{align}
-
-Here $\langle \cdot \rangle_{model}$ is the expectation value over the model probability distribution $p(\bm{x}| \bm{\theta})$.
-
-% !split
-\subsection*{Setting up for gradient descent calculations}
-
-Using the previous relationship we can express the gradient of the cost function as
-
-\begin{align}
-	\frac{\partial \mathcal{C}_{LL}}{\partial \theta_i}
-	=& \langle \frac{ \partial E(\bm{x}; \theta_i) } { \partial \theta_i} \rangle_{data} + \frac{\partial \log Z(\theta_i)}{ \partial  \theta_i} \\
-	=& \langle \frac{ \partial E(\bm{x}; \theta_i) } { \partial \theta_i} \rangle_{data} - \langle \frac{ \partial E(\bm{x}; \theta_i) } { \partial \theta_i} \rangle_{model} \\
-	%=& \langle O_i(\bm{x}) \rangle_{data} - \langle O_i(\bm{x}) \rangle_{model}
-\end{align}
-
-This expression shows that the gradient of the log-likelihood cost
-function is a \textbf{difference of moments}, with one calculated from
-the data and one calculated from the model. The data-dependent term is
-called the \textbf{positive phase} and the model-dependent term is
-called the \textbf{negative phase} of the gradient. We see now that
-minimizing the cost function results in lowering the energy of
-configurations $\bm{x}$ near points in the training data and
-increasing the energy of configurations not observed in the training
-data. That means we increase the model's probability of configurations
-similar to those in the training data.
-
-% !split
-\subsection*{More interpretations}
-
-The gradient of the cost function also demonstrates why gradients of
-unsupervised, generative models must be computed differently from for
-those of for example FNNs. While the data-dependent expectation value
-is easily calculated based on the samples $\bm{x}_i$ in the training
-data, we must sample from the model in order to generate samples from
-which to caclulate the model-dependent term. We sample from the model
-by using MCMC-based methods. We can not sample from the model directly
-because the partition function $Z$ is generally intractable.
-
-As in supervised machine learning problems, the goal is also here to
-perform well on \textbf{unseen} data, that is to have good
-generalization from the training data. The distribution $f(x)$ we
-approximate is not the \textbf{true} distribution we wish to estimate,
-it is limited to the training data. Hence, in unsupervised training as
-well it is important to prevent overfitting to the training data. Thus
-it is common to add regularizers to the cost function in the same
-manner as we discussed for say linear regression.
-
-% ------------------- end of main content ---------------
-
-\end{document}

### 2019-01-16 doc/MachineLearning/article2.tex
changed article 2

--- 2019-01-16 doc/MachineLearning/article1.tex
@@
 
-\section{Introduction}
-
-\end{document}
-
-%%
-%% Automatically generated file from DocOnce source
-%% (https://github.com/hplgit/doconce/)
-%%
-%%
-
-%-------------------- begin preamble ----------------------
-
-\documentclass[%
-oneside,                 % oneside: electronic viewing, twoside: printing
-final,                   % draft: marks overfull hboxes, figures with paths
-10pt]{article}
-
-\listfiles               %  print all files needed to compile this document
-
-\usepackage{relsize,makeidx,color,setspace,amsmath,amsfonts,amssymb}
-\usepackage[table]{xcolor}
-\usepackage{bm,ltablex,microtype}
-
-\usepackage[pdftex]{graphicx}
-
-\usepackage{fancyvrb} % packages needed for verbatim environments
-\usepackage{minted}
-\usemintedstyle{default}
-
-\usepackage[T1]{fontenc}
-%\usepackage[latin1]{inputenc}
-\usepackage{ucs}
-\usepackage[utf8x]{inputenc}
-
-\usepackage{lmodern}         % Latin Modern fonts derived from Computer Modern
-
-% Hyperlinks in PDF:
-\definecolor{linkcolor}{rgb}{0,0,0.4}
-\usepackage{hyperref}
-\hypersetup{
-    breaklinks=true,
-    colorlinks=true,
-    linkcolor=linkcolor,
-    urlcolor=linkcolor,
-    citecolor=black,
-    filecolor=black,
-    %filecolor=blue,
-    pdfmenubar=true,
-    pdftoolbar=true,
-    bookmarksdepth=3   % Uncomment (and tweak) for PDF bookmarks with more levels than the TOC
-    }
-%\hyperbaseurl{}   % hyperlinks are relative to this root
-
-\setcounter{tocdepth}{2}  % levels in table of contents
-
-% Tricks for having figures close to where they are defined:
-% 1. define less restrictive rules for where to put figures
-\setcounter{topnumber}{2}
-\setcounter{bottomnumber}{2}
-\setcounter{totalnumber}{4}
-\renewcommand{\topfraction}{0.95}
-\renewcommand{\bottomfraction}{0.95}
-\renewcommand{\textfraction}{0}
-\renewcommand{\floatpagefraction}{0.75}
-% floatpagefraction must always be less than topfraction!
-% 2. ensure all figures are flushed before next section
-\usepackage[section]{placeins}
-% 3. enable begin{figure}[H] (often leads to ugly pagebreaks)
-%\usepackage{float}\restylefloat{figure}
-
-% --- fancyhdr package for fancy headers ---
-\usepackage{fancyhdr}
-\fancyhf{} % sets both header and footer to nothing
-\renewcommand{\headrulewidth}{0pt}
-\fancyfoot[LE,RO]{\thepage}
-% Ensure copyright on titlepage (article style) and chapter pages (book style)
-\fancypagestyle{plain}{
-  \fancyhf{}
-  \fancyfoot[C]{{\footnotesize \copyright\ 1999-2019, Morten Hjorth-Jensen. Released under CC Attribution-NonCommercial 4.0 license}}
-%  \renewcommand{\footrulewidth}{0mm}
-  \renewcommand{\headrulewidth}{0mm}
-}
-% Ensure copyright on titlepages with \thispagestyle{empty}
-\fancypagestyle{empty}{
-  \fancyhf{}
-  \fancyfoot[C]{{\footnotesize \copyright\ 1999-2019, Morten Hjorth-Jensen. Released under CC Attribution-NonCommercial 4.0 license}}
-  \renewcommand{\footrulewidth}{0mm}
-  \renewcommand{\headrulewidth}{0mm}
-}
-
-\pagestyle{fancy}
-
-\usepackage[framemethod=TikZ]{mdframed}
-
-% --- begin definitions of admonition environments ---
-
-% --- end of definitions of admonition environments ---
-
-% prevent orhpans and widows
-\clubpenalty = 10000
-\widowpenalty = 10000
-
-% --- end of standard preamble for documents ---
-
-% insert custom LaTeX commands...
-
-\raggedbottom
-\makeindex
-\usepackage[totoc]{idxlayout}   % for index in the toc
-\usepackage[nottoc]{tocbibind}  % for references/bibliography in the toc
-
 %-------------------- end preamble ----------------------
@@
 
-% matching end for #ifdef PREAMBLE
-
-\newcommand{\exercisesection}[1]{\subsection*{#1}}
-
-% ------------------- main content ----------------------
-
-% ----------------- title -------------------------
-
-\thispagestyle{empty}
-
-\begin{center}
-{\LARGE\bf
-\begin{spacing}{1.25}
-Solving quantum mechanical problems with Machine Learning
-\end{spacing}
-}
-\end{center}
-
-% ----------------- author(s) -------------------------
-
-\begin{center}
-{\bf Morten Hjorth-Jensen${}^{1, 2}$} \\ [0mm]
-\end{center}
-
-\begin{center}
-% List of all institutions:
-\centerline{{\small ${}^1$Department of Physics and Astronomy and National Superconducting Cyclotron Laboratory, Michigan State University, USA}}
-\centerline{{\small ${}^2$Department of Physics and Center for Computing in Science Education, University of Oslo, Norway}}
-\end{center}
-
-% ----------------- end author(s) -------------------------
-
-% --- begin date ---
-\begin{center}
-Jan 16, 2019 
-\end{center}
-% --- end date ---
-
-\vspace{1cm}
-
-% !split
 \subsection*{What is this talk about?}

### 2019-01-17 doc/MachineLearning/article1.tex
update

@@
 
-\pacs{32.80.Rm, 31.15.A-, 42.65.Ky}
-
 \maketitle
 
-\section{Introduction}
-
-\end{document}
-
-%%
-%% Automatically generated file from DocOnce source
-%% (https://github.com/hplgit/doconce/)
-%%
-%%
-
-%-------------------- begin preamble ----------------------
-
-\documentclass[%
-oneside,                 % oneside: electronic viewing, twoside: printing
-final,                   % draft: marks overfull hboxes, figures with paths
-10pt]{article}
-
-\listfiles               %  print all files needed to compile this document
-
-\usepackage{relsize,makeidx,color,setspace,amsmath,amsfonts,amssymb}
-\usepackage[table]{xcolor}
-\usepackage{bm,ltablex,microtype}
-
-\usepackage[pdftex]{graphicx}
-
-\usepackage{fancyvrb} % packages needed for verbatim environments
-\usepackage{minted}
-\usemintedstyle{default}
-
-\usepackage[T1]{fontenc}
-%\usepackage[latin1]{inputenc}
-\usepackage{ucs}
-\usepackage[utf8x]{inputenc}
-
-\usepackage{lmodern}         % Latin Modern fonts derived from Computer Modern
-
-% Hyperlinks in PDF:
-\definecolor{linkcolor}{rgb}{0,0,0.4}
-\usepackage{hyperref}
-\hypersetup{
-    breaklinks=true,
-    colorlinks=true,
-    linkcolor=linkcolor,
-    urlcolor=linkcolor,
-    citecolor=black,
-    filecolor=black,
-    %filecolor=blue,
-    pdfmenubar=true,
-    pdftoolbar=true,
-    bookmarksdepth=3   % Uncomment (and tweak) for PDF bookmarks with more levels than the TOC
-    }
-%\hyperbaseurl{}   % hyperlinks are relative to this root
-
-\setcounter{tocdepth}{2}  % levels in table of contents
-
-% Tricks for having figures close to where they are defined:
-% 1. define less restrictive rules for where to put figures
-\setcounter{topnumber}{2}
-\setcounter{bottomnumber}{2}
-\setcounter{totalnumber}{4}
-\renewcommand{\topfraction}{0.95}
-\renewcommand{\bottomfraction}{0.95}
-\renewcommand{\textfraction}{0}
-\renewcommand{\floatpagefraction}{0.75}
-% floatpagefraction must always be less than topfraction!
-% 2. ensure all figures are flushed before next section
-\usepackage[section]{placeins}
-% 3. enable begin{figure}[H] (often leads to ugly pagebreaks)
-%\usepackage{float}\restylefloat{figure}
-
-% --- fancyhdr package for fancy headers ---
-\usepackage{fancyhdr}
-\fancyhf{} % sets both header and footer to nothing
-\renewcommand{\headrulewidth}{0pt}
-\fancyfoot[LE,RO]{\thepage}
-% Ensure copyright on titlepage (article style) and chapter pages (book style)
-\fancypagestyle{plain}{
-  \fancyhf{}
-  \fancyfoot[C]{{\footnotesize \copyright\ 1999-2019, Morten Hjorth-Jensen. Released under CC Attribution-NonCommercial 4.0 license}}
-%  \renewcommand{\footrulewidth}{0mm}
-  \renewcommand{\headrulewidth}{0mm}
-}
-% Ensure copyright on titlepages with \thispagestyle{empty}
-\fancypagestyle{empty}{
-  \fancyhf{}
-  \fancyfoot[C]{{\footnotesize \copyright\ 1999-2019, Morten Hjorth-Jensen. Released under CC Attribution-NonCommercial 4.0 license}}
-  \renewcommand{\footrulewidth}{0mm}
-  \renewcommand{\headrulewidth}{0mm}
-}
-
-\pagestyle{fancy}
-
-\usepackage[framemethod=TikZ]{mdframed}
-
-% --- begin definitions of admonition environments ---
-
-% --- end of definitions of admonition environments ---
-
-% prevent orhpans and widows
-\clubpenalty = 10000
-\widowpenalty = 10000
-
-% --- end of standard preamble for documents ---
-
-% insert custom LaTeX commands...
-
-\raggedbottom
-\makeindex
-\usepackage[totoc]{idxlayout}   % for index in the toc
-\usepackage[nottoc]{tocbibind}  % for references/bibliography in the toc
-
-%-------------------- end preamble ----------------------
-
-\begin{document}
-
-% matching end for #ifdef PREAMBLE
-
-\newcommand{\exercisesection}[1]{\subsection*{#1}}
-
-% ------------------- main content ----------------------
-
-% ----------------- title -------------------------
-
-\thispagestyle{empty}
-
-\begin{center}
-{\LARGE\bf
-\begin{spacing}{1.25}
-Solving quantum mechanical problems with Machine Learning
-\end{spacing}
-}
-\end{center}
-
-% ----------------- author(s) -------------------------
-
-\begin{center}
-{\bf Morten Hjorth-Jensen${}^{1, 2}$} \\ [0mm]
-\end{center}
-
-\begin{center}
-% List of all institutions:
-\centerline{{\small ${}^1$Department of Physics and Astronomy and National Superconducting Cyclotron Laboratory, Michigan State University, USA}}
-\centerline{{\small ${}^2$Department of Physics and Center for Computing in Science Education, University of Oslo, Norway}}
-\end{center}
-
-% ----------------- end author(s) -------------------------
-
-% --- begin date ---
-\begin{center}
-Jan 16, 2019 
-\end{center}
-% --- end date ---
-
-\vspace{1cm}
-
-% !split
-\subsection*{What is this talk about?}
-
-% --- begin paragraph admon ---
-\paragraph{}
-The main aim is to give you a short and pedestrian introduction to  how we can use Machine Learning methods
-to solve quantum mechanical many-body problems. And why this could be of interest.
-
-The hope is that after this talk you have gotten the basic ideas to get you started.
-Peeping into \href{{https://github.com/mhjensenseminars/MachineLearningTalk}}{\nolinkurl{https://github.com/mhjensenseminars/MachineLearningTalk}}, you'll find a Jupyter notebook, slides, codes etc that will allow you to reproduce the simulations discussed here, and perhaps run your own very first calculations.
-
-Try it out and please don't hesitate to swing by if something is unclear.
-% --- end paragraph admon ---
-
-% !split
-\subsection*{Why?}
-
-How can we avoid the dimensionality curse? Many possibilities
-\begin{enumerate}
-\item smarter basis functions
-
-\item resummation of specific correlations
-
-\item stochastic sampling of high-lying states (stochastic FCI, CC and SRG/IMSRG)
-
-\item many more
-\end{enumerate}
-
-\noindent
-Machine Learning and Quantum Computing hold also great promise in tackling the 
-ever increasing dimensionalities. Here we will focus on Machine Learning.
-
-% !split
-\subsection*{Overview}
-
-% --- begin paragraph admon ---
-\paragraph{}
-\begin{itemize}
-\item Short intro to Machine Learning
-
-\item Variational Monte Carlo (Markov Chain Monte Carlo, $\mathrm{MC}^2$) and two-electron quantum dots, solving quantum mechanical problems in a stochastic way.
-It  will serve as  our motivation for switching to Machine Learning.
-
-\item From Variational Monte Carlo to Boltzmann Machines and Machine Learning
-\end{itemize}
-
-\noindent
-% --- end paragraph admon ---
-
-% !split
-\subsection*{What are the Machine Learning calculations here based on?}
+\section{Introduction and motivation}
+
+Traditional many-body methods like full configuration interaction
+theory (FCI), coupled-cluster (CC) theory, many-body perturbation
+theory (MBPT), in-medium similarity renormalization group (IMSRG),
+various Monte Carlo methods and many other many-body approaches, have
+been rather successful in describing properties of nuclear
+systems. However, essentially all of these methods face what is
+normally called the curse of dimensionality. For wave function based
+method like FCI or CC theories where the original continuous problem
+is discretized in terms selected basis functions and/or many-body
+excitations, the dimensionality of systems under study grows almost
+exponentially when larger basis sets and/or number of particles are
+included. Nuclear systems close to the limits of stability pose in
+particular a tough problem to many-body practitioners in terms of an
+increase of the number of relevant degrees of freedom. To describe say
+weakly bound nuclei within the framework of a wave function based
+approach, requires often a single-particle basis which includes bound,
+weakly bound and unbound states, increasing thereby considerably the
+number of many-body basis states. This renders often a standard
+shell-model calculation infeasible. Within the above mentioned
+many-body methods there are however several interesting theoretical
+approachs which attempt at circumventing the dimensionality
+curse. Smarter basis sets is one of these approaches, as well as the
+resummation of specific correlations and the recently proposed
+stochastic sampling of many-body states in both FCI and CC
+calculations.
+
+Recently, several authors of pointed to  possibilities within the
+broad fields of Machine Learning and Quantum Computing as approaches
+that hold great promise in tackling the ever increasing
+dimensionalities. There are several groups worldwide which now focus
+on Machine Learning and/or Quantum Computing applied to many-body
+problems.
 
@@
 
-Thanks to Jane Kim (MSU), Vilde Flugsrud (UiO), Alfred Alocias Mariadason (UiO) for many discussions and interpretations.
-
-% !split
-\subsection*{A new world}
-
-% --- begin paragraph admon ---
-\paragraph{}
+Here we try to implement these algorithms to many-body systems that mimick a nuclear physics like problem, namely electrons confined to move in harmonic oscillator like potentials.
+
 Machine learning (ML) is an extremely rich field, in spite of its young age. The
@@
 
-Popular software packages written in Python for ML are
-
-\begin{itemize}
-\item \href{{http://scikit-learn.org/stable/}}{Scikit-learn},
-
-\item \href{{https://www.tensorflow.org/}}{Tensorflow},
-
-\item \href{{http://pytorch.org/}}{PyTorch} and
-
-\item \href{{https://keras.io/}}{Keras}.
-\end{itemize}
-
-\noindent
-These are all freely available at their respective GitHub sites. They 
-encompass communities of developers in the thousands or more. And the number
-of code developers and contributors keeps increasing.
-% --- end paragraph admon ---
-
-% !split
-\subsection*{Lots of room for creativity}
-
-% --- begin paragraph admon ---
-\paragraph{}
-Not all the
-algorithms and methods can be given a rigorous mathematical
-justification, opening up thereby for experimenting
-and trial and error and thereby exciting new developments.
-% --- end paragraph admon ---
-
-% --- begin paragraph admon ---
-\paragraph{}
-A solid command of linear algebra, multivariate theory, 
-probability theory, statistical data analysis,
-understanding errors and Monte Carlo methods is important in order to understand many of the 
-various algorithms and methods.
-% --- end paragraph admon ---
-
-\textbf{Job market, a personal statement}: \href{{https://www.analyticsindiamag.com/top-countries-hiring-most-number-of-artificial-intelligence-machine-learning-experts/}}{A familiarity with ML is almost becoming a prerequisite for many of the most exciting employment opportunities}.
-And add quantum computing and there you are!
-
-% !split
-\subsection*{Knowledge of Statistical analysis and optimization of data}
-Some key elements that enter much of the discussion on ML:
-\begin{enumerate}
-\item Basic concepts, expectation values, variance, covariance, correlation functions and errors;
-
-\item Simpler models, binomial distribution, the Poisson distribution, simple and multivariate normal distributions;
-
-\item Central elements of Bayesian statistics and modeling;
-
-\item Central elements from linear algebra
-
-\item Gradient methods for data optimization
-
-\item Monte Carlo methods, Markov chains, Metropolis-Hastings algorithm;
-
-\item Estimation of errors using cross-validation, blocking, bootstrapping and jackknife methods;
-
-\item Practical optimization using Singular-value decomposition and least squares for parameterizing data.
-
-\item Principal Component Analysis.
-\end{enumerate}
-
-\noindent
-% !split 
-\subsection*{Some members of the ML family}
-
-% --- begin paragraph admon ---
-\paragraph{}
-\begin{enumerate}
-\item Linear regression and its variants, Logistic regression
-
-\item Decision tree algorithms, from simpler to more complex ones like random forests
-
-\item Bayesian statistics
-
-\item Support vector machines and finally various variants of
-
-\item Artifical neural networks and deep learning
-
-\item Convolutional NN, autoenconders
-
-\item and many more
-\end{enumerate}
-
-\noindent
-% --- end paragraph admon ---
-
-% !split
-\subsection*{What are the basic ingredients?}
-
-% --- begin paragraph admon ---
-\paragraph{}
 Almost every problem in ML and data science starts with the same ingredients:
@@
 We seek to minimize the function $\mathcal{C} (\mathbf{x}, f(\mathbf{\alpha}))$ by finding the parameter values which minimize $\mathcal{C}$. This leads to  various minimization algorithms.
-% --- end paragraph admon ---
-
-% !split
-\subsection*{What is Machine Learning?}
-
-% --- begin paragraph admon ---
-\paragraph{}
+
+\section{What is Machine Learning}
 Machine learning is the science of giving computers the ability to
@@
 You will notice however that many of the basic ideas discussed do come from Physics!
-% --- end paragraph admon ---
-
-% !split
-\subsection*{Types of Machine Learning}
-
-% --- begin paragraph admon ---
-\paragraph{}
+
 The approaches to machine learning are many, but are often split into two main categories. 
@@
 
-\noindent
-% --- end paragraph admon ---
-
-% !split
-\subsection*{References}
-
-% --- begin paragraph admon ---
-\paragraph{}
 \begin{itemize}
@@
 
-\item See also Metha and Schwab, arXiv.1410.3831, how to link Variational renormalization group theory with deep learning (recommended read)
-\end{itemize}
-
-\noindent
-% --- end paragraph admon ---
-
-% !split
-\subsection*{Another interesting article}
-
 Here we will use so-called \textbf{reduced Boltzmann Machines} to simulate quantum many-body problems.
@@
 
-% !split
-\subsection*{Courses on Data science and Machine Learning  at UiO}
-
-The \href{{https://www.mn.uio.no/english/research/about/centre-focus/innovation/data-science/studies/}}{link here}  gives an excellent overview of courses on Machine learning at UiO.
-
-\begin{enumerate}
-\item \href{{http://www.uio.no/studier/emner/matnat/math/STK2100/index-eng.html}}{STK2100 Machine learning and statistical methods for prediction and classification}.
-
-\item \href{{https://www.uio.no/studier/emner/matnat/ifi/IN3050/index-eng.html}}{IN3050 Introduction to Artificial Intelligence and Machine Learning}.
-Introductory course in machine learning and AI with an algorithmic approach.
-
-\item \href{{http://www.uio.no/studier/emner/matnat/math/STK-INF3000/index-eng.html}}{STK-INF3000/4000 Selected Topics in Data Science}.
-The course provides insight into selected contemporary relevant topics within Data Science.
-
-\item \href{{https://www.uio.no/studier/emner/matnat/ifi/IN4080/index.html}}{IN4080 Natural Language Processing}. Probabilistic and machine learning techniques applied to natural language processing.
-
-\item \href{{https://www.uio.no/studier/emner/matnat/math/STK-IN4300/index-eng.html}}{STK-IN4300 – Statistical learning methods in Data Science}.
-An advanced introduction to statistical and machine learning.
-For students with a good mathematics and statistics background.
-
-\item \href{{http://www.uio.no/studier/emner/matnat/ifi/INF4490/}}{INF4490 Biologically Inspired Computing}.
-An introduction to self-adapting methods also called artificial intelligence or machine learning.
-
-\item \href{{https://www.uio.no/studier/emner/matnat/ifi/IN-STK5000/index-eng.html}}{IN-STK5000  Adaptive Methods for Data-Based Decision Making}.
-Methods for adaptive collection and processing of data based on machine learning techniques.
-
-\item \href{{https://www.uio.no/studier/emner/matnat/ifi/IN5400/}}{IN5400/INF5860 – Machine Learning for Image Analysis}.
-An introduction to deep learning with particular emphasis on applications within Image analysis, but useful for other application areas too.
-
-\item \href{{https://www.uio.no/studier/emner/matnat/its/TEK5040/}}{TEK5040 – Deep learning}.
-The course addresses advanced algorithms and architectures for deep learning with neural networks.
-The course provides an introduction to how deep-learning techniques can be used in the construction of key parts of advanced autonomous systems that exist in physical environments and cyber environments.
-\end{enumerate}
-
-\noindent
-% !split
-\subsection*{Additional courses of interest}
-
-\begin{enumerate}
-\item \href{{https://www.uio.no/studier/emner/matnat/math/STK4051/index-eng.html}}{STK4051 Computational Statistics}
-
-\item \href{{https://www.uio.no/studier/emner/matnat/math/STK4021/index-eng.html}}{STK4021 Applied Bayesian Analysis and Numerical Methods}
-\end{enumerate}
-
-\noindent
-% !split
-\subsection*{Decision trees and Regression}
-\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=7mm]{python}
-import numpy as np
-import matplotlib.pyplot as plt
-from sklearn.preprocessing import PolynomialFeatures
-from sklearn.linear_model import LinearRegression
-
-steps=250
-
-distance=0
-x=0
-distance_list=[]
-steps_list=[]
-while x<steps:
-    distance+=np.random.randint(-1,2)
-    distance_list.append(distance)
-    x+=1
-    steps_list.append(x)
-plt.plot(steps_list,distance_list, color='green', label="Random Walk Data")
-
-steps_list=np.asarray(steps_list)
-distance_list=np.asarray(distance_list)
-
-X=steps_list[:,np.newaxis]
-
-#Polynomial fits
-
-#Degree 2
-poly_features=PolynomialFeatures(degree=2, include_bias=False)
-X_poly=poly_features.fit_transform(X)
-
-lin_reg=LinearRegression()
-poly_fit=lin_reg.fit(X_poly,distance_list)
-b=lin_reg.coef_
-c=lin_reg.intercept_
-print ("2nd degree coefficients:")
-print ("zero power: ",c)
-print ("first power: ", b[0])
-print ("second power: ",b[1])
-
-z = np.arange(0, steps, .01)
-z_mod=b[1]*z**2+b[0]*z+c
-
-fit_mod=b[1]*X**2+b[0]*X+c
-plt.plot(z, z_mod, color='r', label="2nd Degree Fit")
-plt.title("Polynomial Regression")
-
-plt.xlabel("Steps")
-plt.ylabel("Distance")
-
-#Degree 10
-poly_features10=PolynomialFeatures(degree=10, include_bias=False)
-X_poly10=poly_features10.fit_transform(X)
-
-poly_fit10=lin_reg.fit(X_poly10,distance_list)
-
-y_plot=poly_fit10.predict(X_poly10)
-plt.plot(X, y_plot, color='black', label="10th Degree Fit")
-
-plt.legend()
-plt.show()
-
-#Decision Tree Regression
-from sklearn.tree import DecisionTreeRegressor
-regr_1=DecisionTreeRegressor(max_depth=2)
-regr_2=DecisionTreeRegressor(max_depth=5)
-regr_3=DecisionTreeRegressor(max_depth=7)
-regr_1.fit(X, distance_list)
-regr_2.fit(X, distance_list)
-regr_3.fit(X, distance_list)
-
-X_test = np.arange(0.0, steps, 0.01)[:, np.newaxis]
-y_1 = regr_1.predict(X_test)
-y_2 = regr_2.predict(X_test)
-y_3=regr_3.predict(X_test)
-
-# Plot the results
-plt.figure()
-plt.scatter(X, distance_list, s=2.5, c="black", label="data")
-plt.plot(X_test, y_1, color="red",
-         label="max_depth=2", linewidth=2)
-plt.plot(X_test, y_2, color="green", label="max_depth=5", linewidth=2)
-plt.plot(X_test, y_3, color="m", label="max_depth=7", linewidth=2)
-
-plt.xlabel("Data")
-plt.ylabel("Darget")
-plt.title("Decision Tree Regression")
-plt.legend()
-plt.show()
-
-\end{minted}
-
-% !split
-\subsection*{Artificial neurons}
-
-% --- begin paragraph admon ---
-\paragraph{}
-
-The field of artificial neural networks has a long history of
-development, and is closely connected with the advancement of computer
-science and computers in general. A model of artificial neurons was
-first developed by McCulloch and Pitts in 1943 to study signal
-processing in the brain and has later been refined by others. The
-general idea is to mimic neural networks in the human brain, which is
-composed of billions of neurons that communicate with each other by
-sending electrical signals.  Each neuron accumulates its incoming
-signals, which must exceed an activation threshold to yield an
-output. If the threshold is not overcome, the neuron remains inactive,
-i.e.~has zero output.
-
-This behaviour has inspired a simple mathematical model for an artificial neuron.
-
-\[
- y = f\left(\sum_{i=1}^n w_ix_i\right) = f(u)
-\]
-Here, the output $y$ of the neuron is the value of its activation function, which have as input
-a weighted sum of signals $x_i, \dots ,x_n$ received by $n$ other neurons.
-% --- end paragraph admon ---
-
-% !split
-\subsection*{A simple perceptron model}
-
-\vspace{6mm}
-
-% inline figure
-\centerline{\includegraphics[width=0.9\linewidth]{figures/perceptron.png}}
-
-\vspace{6mm}
-
-% !split
-\subsection*{Neural network types}
-
-% --- begin paragraph admon ---
-\paragraph{}
-An artificial neural network (NN), is a computational model that consists of layers of connected neurons, or \emph{nodes}. 
-It is supposed to mimic a biological nervous system by letting each neuron interact with other neurons
-by sending signals in the form of mathematical functions between layers. 
-A wide variety of different NNs have
-been developed, but most of them consist of an input layer, an output layer and eventual layers in-between, called
-\emph{hidden layers}. All layers can contain an arbitrary number of nodes, and each connection between two nodes
-is associated with a weight variable.
-
-\vspace{6mm}
-
-% inline figure
-\centerline{\includegraphics[width=0.6\linewidth]{figures/dnn.png}}
-
-\vspace{6mm}
-% --- end paragraph admon ---
-
-% !split
 \subsection*{The system: two electrons in a harmonic oscillator trap in two dimensions}
@@
 
-% !split
 \subsection*{Quantum Monte Carlo Motivation}
 
-% --- begin paragraph admon ---
-\paragraph{Basic steps.}
-Choose a trial wave function
-$\psi_T(\bm{R})$.
-\[
-   P(\bm{R},\bm{\alpha})= \frac{\left|\psi_T(\bm{R},\bm{\alpha})\right|^2}{\int \left|\psi_T(\bm{R},\bm{\alpha})\right|^2d\bm{R}}.
-\]
-This is our model, or likelihood/probability distribution function  (PDF). It depends on some variational parameters $\bm{\alpha}$.
-The approximation to the expectation value of the Hamiltonian is now 
-\[
-   \langle E[\bm{\alpha}] \rangle = 
-   \frac{\int d\bm{R}\Psi^{\ast}_T(\bm{R},\bm{\alpha})H(\bm{R})\Psi_T(\bm{R},\bm{\alpha})}
-        {\int d\bm{R}\Psi^{\ast}_T(\bm{R},\bm{\alpha})\Psi_T(\bm{R},\bm{\alpha})}.
-\]
-% --- end paragraph admon ---
-
-% !split
 \subsection*{Quantum Monte Carlo Motivation}
@@
 
-% !split
-\subsection*{The VMC code}
-
-\begin{minted}[fontsize=\fontsize{9pt}{9pt},linenos=false,mathescape,baselinestretch=1.0,fontfamily=tt,xleftmargin=7mm]{python}
-# Importing various packages
-from math import exp, sqrt
-from random import random, seed
-import numpy as np
-import matplotlib.pyplot as plt
-from mpl_toolkits.mplot3d import Axes3D
-from matplotlib import cm
-from matplotlib.ticker import LinearLocator, FormatStrFormatter
-import sys
-
-#Trial wave function for quantum dots in two dims
-def WaveFunction(r,alpha,beta):
-    r1 = r[0,0]**2 + r[0,1]**2
-    r2 = r[1,0]**2 + r[1,1]**2
-    r12 = sqrt((r[0,0]-r[1,0])**2 + (r[0,1]-r[1,1])**2)
-    deno = r12/(1+beta*r12)
-    return exp(-0.5*alpha*(r1+r2)+deno)
-
-#Local energy  for quantum dots in two dims, using analytical local energy
-def LocalEnergy(r,alpha,beta):
-
-    r1 = (r[0,0]**2 + r[0,1]**2)
-    r2 = (r[1,0]**2 + r[1,1]**2)
-    r12 = sqrt((r[0,0]-r[1,0])**2 + (r[0,1]-r[1,1])**2)
-    deno = 1.0/(1+beta*r12)
-    deno2 = deno*deno
-    return 0.5*(1-alpha*alpha)*(r1 + r2) +2.0*alpha + 1.0/r12+deno2*(alpha*r12-deno2+2*beta*deno-1.0/r12)
-
-# The Monte Carlo sampling with the Metropolis algo
-def MonteCarloSampling():
-
-    NumberMCcycles= 100000
-    StepSize = 1.0
-    # positions
-    PositionOld = np.zeros((NumberParticles,Dimension), np.double)
-    PositionNew = np.zeros((NumberParticles,Dimension), np.double)
-    # seed for rng generator
-    seed()
-    # start variational parameter
-    alpha = 0.9
-    for ia in range(MaxVariations):
-        alpha += .025
-        AlphaValues[ia] = alpha
-        beta = 0.2 
-        for jb in range(MaxVariations):
-            beta += .01
-            BetaValues[jb] = beta
-            energy = energy2 = 0.0
-            DeltaE = 0.0
-            #Initial position
-            for i in range(NumberParticles):
-                for j in range(Dimension):
-                    PositionOld[i,j] = StepSize * (random() - .5)
-            wfold = WaveFunction(PositionOld,alpha,beta)
-
-            #Loop over MC MCcycles
-            for MCcycle in range(NumberMCcycles):
-                #Trial position
-                for i in range(NumberParticles):
-                    for j in range(Dimension):
-                        PositionNew[i,j] = PositionOld[i,j] + StepSize * (random() - .5)
-                wfnew = WaveFunction(PositionNew,alpha,beta)
-
-                #Metropolis test to see whether we accept the move
-                if random() < wfnew**2 / wfold**2:
-                   PositionOld = PositionNew.copy()
-                   wfold = wfnew
-                   DeltaE = LocalEnergy(PositionOld,alpha,beta)
-                energy += DeltaE
-                energy2 += DeltaE**2
-
-            #We calculate mean, variance and error ...
-            energy /= NumberMCcycles
-            energy2 /= NumberMCcycles
-            variance = energy2 - energy**2
-            error = sqrt(variance/NumberMCcycles)
-            Energies[ia,jb] = energy    
-    return Energies, AlphaValues, BetaValues
-
-#Here starts the main program with variable declarations
-NumberParticles = 2
-Dimension = 2
-MaxVariations = 10
-Energies = np.zeros((MaxVariations,MaxVariations))
-AlphaValues = np.zeros(MaxVariations)
-BetaValues = np.zeros(MaxVariations)
-(Energies, AlphaValues, BetaValues) = MonteCarloSampling()
-
-# Prepare for plots
-fig = plt.figure()
-ax = fig.gca(projection='3d')
-# Plot the surface.
-X, Y = np.meshgrid(AlphaValues, BetaValues)
-surf = ax.plot_surface(X, Y, Energies,cmap=cm.coolwarm,linewidth=0, antialiased=False)
-# Customize the z axis.
-zmin = np.matrix(Energies).min()
-zmax = np.matrix(Energies).max()
-ax.set_zlim(zmin, zmax)
-ax.set_xlabel(r'$\alpha$')
-ax.set_ylabel(r'$\beta$')
-ax.set_zlabel(r'$\langle E \rangle$')
-ax.zaxis.set_major_locator(LinearLocator(10))
-ax.zaxis.set_major_formatter(FormatStrFormatter('%.02f'))
-# Add a color bar which maps values to colors.
-fig.colorbar(surf, shrink=0.5, aspect=5)
-plt.show()
-
-\end{minted}
-
-% !split
-\subsection*{Technical aspect, improvements and how to define the cost function}
-
-% --- begin paragraph admon ---
-\paragraph{}
-
-\textbf{The above procedure is not the smartest one}. Looping over all variational parameters becomes expensive.
-Also, we don't use importance sampling and optimizations of the standard deviation (blocking, bootstrap, jackknife). 
-Such codes are included in the above Github address.
-
-We can also be smarter and use minimization methods to find the \textbf{optimal} variational parameters with fewer Monte Carlo cycles and then 
-fire up our heavy artillery.
-
-One way to achieve this is to minimize the energy as function of the variational parameters.
-% --- end paragraph admon ---
-
-% !split
-\subsection*{Energy derivatives}
-
-% --- begin paragraph admon ---
-\paragraph{}
-To find the derivatives of the local energy expectation value as function of the variational parameters, we can use the chain rule and the hermiticity of the Hamiltonian.
-
-Let us define 
-\[
-\bar{E}_{\alpha_i}=\frac{d\langle  E_L\rangle}{d\alpha_i}.
-\]
-as the derivative of the energy with respect to the variational parameter $\alpha_i$
-We define also the derivative of the trial function (skipping the subindex $T$) as 
-\[
-\bar{\Psi}_{i}=\frac{d\Psi}{d\alpha_i}.
-\]
-% --- end paragraph admon ---
-
-% !split
-\subsection*{Derivatives of the local energy}
-
 % --- begin paragraph admon ---
@@
 How should we define the cost function?
-% --- end paragraph admon ---
-
-% !split
-\subsection*{Meet the variance and its derivatives}
-
-% --- begin paragraph admon ---
-\paragraph{Why the variance?}
-Suppose the trial function (our model) is the exact wave function. The action of the hamiltionan
-on the wave function
-\[
-   H\Psi = \mathrm{constant}\times \Psi,
-\]
-The integral which defines various 
-expectation values involving moments of the hamiltonian becomes then
-\[
-   \langle E^n \rangle =   \langle H^n \rangle =
-   \frac{\int d\bm{R}\Psi^{\ast}(\bm{R})H^n(\bm{R})\Psi(\bm{R})}
-        {\int d\bm{R}\Psi^{\ast}(\bm{R})\Psi(\bm{R})}=
-\mathrm{constant}\times\frac{\int d\bm{R}\Psi^{\ast}(\bm{R})\Psi(\bm{R})}
-        {\int d\bm{R}\Psi^{\ast}(\bm{R})\Psi(\bm{R})}=\mathrm{constant}.
-\]
-\textbf{This gives an important information: If I want the variance, the exact wave function leads to zero variance!}
-The variance is defined as 
-\[
-\sigma_E = \langle E^2\rangle - \langle E\rangle^2.
-\]
-Variation is then performed by minimizing both the energy and the variance.
-% --- end paragraph admon ---
-
-% !split
-\subsection*{The variance defines the cost function}
-
-% --- begin paragraph admon ---
-\paragraph{}
-We can then take the derivatives of 
-\[
-\sigma_E = \langle E^2\rangle - \langle E\rangle^2,
-\]
-with respect to the variational parameters. The derivatives of the variance can then be used to defined the
-so-called Hessian matrix, which in turn allows us to use minimization methods like Newton's method or 
-standard gradient methods.
-
-This leads to however a more complicated expression, with obvious errors when evaluating integrals by Monte Carlo integration.
-Less used, see however \href{{https://journals.aps.org/prl/abstract/10.1103/PhysRevLett.94.150201}}{Filippi and Umrigar}.
-The expression becomes complicated
-\[
-\bar{E}_{ij} = 2\left[ \langle (\frac{\bar{\Psi}_{ij}}{\Psi}+\frac{\bar{\Psi}_{j}}{\Psi}\frac{\bar{\Psi}_{i}}{\Psi})(E_L-\langle E\rangle)\rangle -\langle \frac{\bar{\Psi}_{i}}{\Psi}\rangle\bar{E}_j-\langle \frac{\bar{\Psi}_{j}}{\Psi}\rangle\bar{E}_i\right] +\langle \frac{\bar{\Psi}_{i}}{\Psi}E_L{_j}\rangle +\langle \frac{\bar{\Psi}_{j}}{\Psi}E_L{_i}\rangle -\langle \frac{\bar{\Psi}_{i}}{\Psi}\rangle\langle E_L{_j}\rangle \langle \frac{\bar{\Psi}_{j}}{\Psi}\rangle\langle E_L{_i}\rangle.
-\]
-
-Evaluating the cost function means having to evaluate the above second derivative of the energy.
-% --- end paragraph admon ---
 

### 2019-01-19 doc/MachineLearning/article1.tex
update

@@
 
-% --- begin paragraph admon ---
 \paragraph{Gaussian-Binary RBM:}
@@
 \end{align}
-% --- end paragraph admon ---
 
-% !split
-\subsection*{More about RBMs}
-\begin{enumerate}
-\item Useful when we model continuous data (i.e., we wish $\mathbf{x}$ to be continuous)
-
-\item Requires a smaller learning rate, since there's no upper bound to the value a component might take in the reconstruction
-\end{enumerate}
-
-\noindent
-Other types of units include:
-\begin{enumerate}
-\item Softmax and multinomial units
-
-\item Gaussian visible and hidden units
-
-\item Binomial units
-
-\item Rectified linear units
-\end{enumerate}
-
-\noindent
-% !split
-\subsection*{Sampling: Metropolis sampling}
-In order to sample from the RBM probability distribution it is common to use Markov Chain Monte Carlo (MCMC) algorithms such as Metropolis-Hastings or Gibbs sampling.
-
-Metropolis sampling starts by suggesting a new configuration $\bm{x}^{k+1}$.
-In the brute force method this is done by some random change of the visible units.
-The new configuration is then accepted with the acceptance probability
-\begin{align}
-	A(\bm{x}^k \rightarrow \bm{x}^{k+1}) = \text{min} (1, \frac{P(\bm{x}^{k+1})}{P(\bm{x}^k)}),
-\end{align} 
-where we need the marginalized probability
-\begin{align}
-	P(\bm{x})  &= \sum_\mathbf{h} P_{rbm}(\mathbf{x}, \mathbf{h}) \\
-				&= \frac{1}{Z}\sum_\mathbf{h} e^{-E(\mathbf{x}, \mathbf{h})}.
-\end{align}
-
-% !split
-\subsection*{Sampling: Gibbs sampling}
-
-In this method we sample from the joint probability $P_{rbm} (\mathbf{x}, \mathbf{h})$ by way of a two step sampling process. We alternately update the visible and hidden units.
-New samples are generated according to the conditional probabilities $P(x_i|\mathbf{h})$ and $P(h_j|\mathbf{x})$ respectively and accepted with the probability of $1$.
-While the the visible nodes are dependent on the hidden nodes and vice versa, the nodes are independent of other nodes within the same layer.
-This is due to there being no intra layer interactions in the restricted Boltzmann machine.
-
-The conditional probabilities are often referred to as the activitation functions in the neural networks context due to their role in determining the node outputs. For the binary-binary RBM they are
-\begin{align}
-	P(h_j = 1 | \bm{x}) &= \frac{1}{1 + e^{-b_j - \sum_i x_i w_{ij}}} \\
-	P(x_i = 1 | \bm{h}) &= \frac{1}{1 + e^{-a_j - \sum_j h_j w_{ij}}},
-\end{align}
-where we recognize the logistic sigmoid function $\sigma (x) = 1/(1+exp(-x))$.
-
-% !split
-\subsection*{Gaussian RBM}
-For the Gaussian-Binary RBM the conditional probabilities are
-\begin{align}
-	P(x_i|\mathbf{h}) &= \mathcal{N}(x_i; a_i+ \sum_j h_j w_{ij}, \sigma^2) \\
-	P(h_j=1|\mathbf{x}) &=  \frac{1}{1+e^{-b_j-\frac{1}{\sigma^2} \sum_i x_i w_{ij}}},
-\end{align}
-while the visible units now follow a normal distribution, we see the hidden units again follow the logistic sigmoid function.
-
-% !split
 \subsection*{Cost function}
@@
 
-% !split
-\subsection*{Optimization / Training}
-
-The training procedure of choice often is Stochastic Gradient Descent (SGD). It consists of a series of iterations where we update the parameters according to the equation
-\begin{align}
-	\bm{\theta}_{k+1} = \bm{\theta}_k - \eta \nabla \mathcal{C} (\bm{\theta}_k)
-\end{align}
-at each $k$-th iteration. There are a range of variants of the algorithm which aim at making the learning rate $\eta$ more adaptive so the method might be more efficient while remaining stable.
-
-We now need the gradient of the cost function in order to minimize it. We find that
-\begin{align}
-	\frac{\partial \mathcal{C}(\{ \theta_i\})}{\partial \theta_i}
-	&= \langle \frac{\partial E(\bm{x}; \theta_i)}{\partial \theta_i} \rangle_{data}
-	+ \frac{\partial \text{log} Z(\{ \theta_i\})}{\partial \theta_i} \\
-	&= \langle O_i(\bm{x}) \rangle_{data} - \langle O_i(\bm{x}) \rangle_{model},
-\end{align}
-where in order to simplify notation we defined the "operator"
-\begin{align}
-	O_i(\bm{x}) = \frac{\partial E(\bm{x}; \theta_i)}{\partial \theta_i}, 
-\end{align}
-and used the statistical mechanics relationship between expectation values and the log-partition function:
-\begin{align}
-	\langle O_i(\bm{x}) \rangle_{model} = \text{Tr} P_\theta(\bm{x})O_i(\bm{x}) = - \frac{\partial \text{log} Z(\{ \theta_i\})}{\partial \theta_i}.
-\end{align}
-
-% !split 
-\subsection*{More on RBMs}
-
-The data-dependent term in the gradient is known as the positive phase of the gradient, while the model-dependent term is known as the negative phase of the gradient.
-The aim of the training is to lower the energy of configurations that are near observed data points (increasing their probability), and raising the energy of configurations that are far from observed data points (decreasing their probability).
-
-The gradient of the negative log-likelihood cost function of a Binary-Binary RBM is then
-\begin{align}
-	\frac{\partial \mathcal{C} (w_{ij}, a_i, b_j)}{\partial w_{ij}} =& \langle x_i h_j \rangle_{data} - \langle x_i h_j \rangle_{model} \\
-	\frac{\partial \mathcal{C} (w_{ij}, a_i, b_j)}{\partial a_{ij}} =& \langle x_i \rangle_{data} - \langle x_i \rangle_{model} \\
-	\frac{\partial \mathcal{C} (w_{ij}, a_i, b_j)}{\partial b_{ij}} =& \langle h_i \rangle_{data} - \langle h_i \rangle_{model}. \\
-\end{align}
-To get the expecation values with respect to the \emph{data}, we set the visible units to each of the observed samples in the training data, then update the hidden units according to the conditional probability found before.
-We then average over all samples in the training data to calculate expectation values with respect to the data.
-
-% !split
-\subsection*{Which sampling to use}
-
-To get the expectation values with respect to the \emph{model}, we use Gibbs sampling.
-We can either initialize the $\bm{x}$ randomly or with a training sample.
-While we ideally want a large number of Gibbs iterations $n\rightarrow n$, one might decide to truncate it earlier for efficiency.
-Doing this while having intialized $\bm{x}$ with a training data vector is referred to as contrastive divergence (CD), because one is then closer to approximating the gradient of this function than the negative log-likelihood.
-The contrastive divergence function is the difference between two Kullback-Leibler divergences (also called relative entropy), which measure how one probability distribution diverges from a second, expected probability distribution (in this case the estimated one from the ground truth one).
-
-% !split
 \subsection*{RBMs for the quantum many body problem}
@@
 
-% !split
-\subsection*{Running the codes}
-
-% --- begin paragraph admon ---
-\paragraph{}
 You can find the codes for the simple two-electron case at the Github repository \href{{https://github.com/mhjensenseminars/MachineLearningTalk/tree/master/doc/Programs/MLcpp/src}}{\nolinkurl{https://github.com/mhjensenseminars/MachineLearningTalk/tree/master/doc/Programs/MLcpp/src}}.
@@
 The Broyden-Fletcher-Goldfarb-Shanno algorithm was used for the minimization. We used $14$ hidden nodes in the calculations below.
-% --- end paragraph admon ---
 
-% !split
-\subsection*{Energy as function of iterations, $N=2$ electrons}
-
-% --- begin paragraph admon ---
-\paragraph{}
-
-\vspace{6mm}
-
-% inline figure
 \centerline{\includegraphics[width=0.9\linewidth]{figures/figN2.pdf}}
 
-\vspace{6mm}
-% --- end paragraph admon ---
-
-% !split
-\subsection*{Energy as function of iterations, $N=6$ electrons}
-
-% --- begin paragraph admon ---
-\paragraph{}
-
-\vspace{6mm}
-
-% inline figure
 \centerline{\includegraphics[width=0.9\linewidth]{figures/figN6.pdf}}
-
-\vspace{6mm}
-% --- end paragraph admon ---
 
@@
 
-\noindent
-% --- end paragraph admon ---
-
-% !split
-\subsection*{Additional material}
-
-% !split
-\subsection*{Kullback-Leibler relative entropy}
-
-When the goal of the training is to approximate a probability
-distribution, as it is in generative modeling, another relevant
-measure is the \textbf{Kullback-Leibler divergence}, also known as the
-relative entropy or Shannon entropy. It is a non-symmetric measure of the
-dissimilarity between two probability density functions $p$ and
-$q$. If $p$ is the unkown probability which we approximate with $q$,
-we can measure the difference by
-\begin{align}
-	\text{KL}(p||q) = \int_{-\infty}^{\infty} p (\bm{x}) \log \frac{p(\bm{x})}{q(\bm{x})}  d\bm{x}.
-\end{align}
-
-Thus, the Kullback-Leibler divergence between the distribution of the
-training data $f(\bm{x})$ and the model distribution $p(\bm{x}|
-\bm{\theta})$ is
-
-\begin{align}
-	\text{KL} (f(\bm{x})|| p(\bm{x}| \bm{\theta})) =& \int_{-\infty}^{\infty}
-	f (\bm{x}) \log \frac{f(\bm{x})}{p(\bm{x}| \bm{\theta})} d\bm{x} \\
-	=& \int_{-\infty}^{\infty} f(\bm{x}) \log f(\bm{x}) d\bm{x} - \int_{-\infty}^{\infty} f(\bm{x}) \log
-	p(\bm{x}| \bm{\theta}) d\bm{x} \\
-	%=& \mathbb{E}_{f(\bm{x})} (\log f(\bm{x})) - \mathbb{E}_{f(\bm{x})} (\log p(\bm{x}| \bm{\theta}))
-	=& \langle \log f(\bm{x}) \rangle_{f(\bm{x})} - \langle \log p(\bm{x}| \bm{\theta}) \rangle_{f(\bm{x})} \\
-	=& \langle \log f(\bm{x}) \rangle_{data} + \langle E(\bm{x}) \rangle_{data} + \log Z \\
-	=& \langle \log f(\bm{x}) \rangle_{data} + \mathcal{C}_{LL} .
-\end{align}
-
-The first term is constant with respect to $\bm{\theta}$ since $f(\bm{x})$ is independent of $\bm{\theta}$.
-Thus the Kullback-Leibler Divergence is minimal when the second term is minimal.
-The second term is the log-likelihood cost function, hence minimizing the Kullback-Leibler divergence is equivalent to maximizing the log-likelihood.
-
-% !split
-\subsection*{Optimizing the cost function}
-
-To further understand generative models it is useful to study the
-gradient of the cost function which is needed in order to minimize it
-using methods like stochastic gradient descent.
-
-The partition function is the generating function of
-expectation values, in particular there are mathematical relationships
-between expectation values and the log-partition function. In this
-case we have
-\begin{align}
-	\langle \frac{ \partial E(\bm{x}; \theta_i) } { \partial \theta_i} \rangle_{model}
-	= \int p(\bm{x}| \bm{\theta}) \frac{ \partial E(\bm{x}; \theta_i) } { \partial \theta_i} d\bm{x} 
-	= -\frac{\partial \log Z(\theta_i)}{ \partial  \theta_i} .
-\end{align}
-
-Here $\langle \cdot \rangle_{model}$ is the expectation value over the model probability distribution $p(\bm{x}| \bm{\theta})$.
-
-% !split
-\subsection*{Setting up for gradient descent calculations}
-
-Using the previous relationship we can express the gradient of the cost function as
-
-\begin{align}
-	\frac{\partial \mathcal{C}_{LL}}{\partial \theta_i}
-	=& \langle \frac{ \partial E(\bm{x}; \theta_i) } { \partial \theta_i} \rangle_{data} + \frac{\partial \log Z(\theta_i)}{ \partial  \theta_i} \\
-	=& \langle \frac{ \partial E(\bm{x}; \theta_i) } { \partial \theta_i} \rangle_{data} - \langle \frac{ \partial E(\bm{x}; \theta_i) } { \partial \theta_i} \rangle_{model} \\
-	%=& \langle O_i(\bm{x}) \rangle_{data} - \langle O_i(\bm{x}) \rangle_{model}
-\end{align}
-
-This expression shows that the gradient of the log-likelihood cost
-function is a \textbf{difference of moments}, with one calculated from
-the data and one calculated from the model. The data-dependent term is
-called the \textbf{positive phase} and the model-dependent term is
-called the \textbf{negative phase} of the gradient. We see now that
-minimizing the cost function results in lowering the energy of
-configurations $\bm{x}$ near points in the training data and
-increasing the energy of configurations not observed in the training
-data. That means we increase the model's probability of configurations
-similar to those in the training data.
-
-% !split
-\subsection*{More interpretations}
-
-The gradient of the cost function also demonstrates why gradients of
-unsupervised, generative models must be computed differently from for
-those of for example FNNs. While the data-dependent expectation value
-is easily calculated based on the samples $\bm{x}_i$ in the training
-data, we must sample from the model in order to generate samples from
-which to caclulate the model-dependent term. We sample from the model
-by using MCMC-based methods. We can not sample from the model directly
-because the partition function $Z$ is generally intractable.
-
-As in supervised machine learning problems, the goal is also here to
-perform well on \textbf{unseen} data, that is to have good
-generalization from the training data. The distribution $f(x)$ we
-approximate is not the \textbf{true} distribution we wish to estimate,
-it is limited to the training data. Hence, in unsupervised training as
-well it is important to prevent overfitting to the training data. Thus
-it is common to add regularizers to the cost function in the same
-manner as we discussed for say linear regression.
-
-% ------------------- end of main content ---------------
-
 \end{document}

### 2019-03-19 doc/MachineLearning/article2.tex
Update of paper

--- 2019-01-16 doc/MachineLearning/article2.tex
@@
 \usepackage{mdframed}
-\newtheorem{theorem}{Theorem}
 
@@
 
-\title{Machine Learning and the Many-body Problem}
+\title{Quantum Mechanical Many-body Problems with Machine Learning Algorithms}
 
@@
 
-\author{Alfred Alocias Mariadason}
+\author{Even M. Nordhagen}
 \affiliation{Department of Physics, University of Oslo, N-0316 Oslo, Norway}
@@
 
+\author{Francesco Pederiva}
+\affiliation{Dipartimento di Fisica, University of Trento, via Sommarive 14, I-38050 Povo, Trento, Italy}
+\affiliation{INFN-TIFPA, Trento Institute for Fundamental Physics and Applications, Trento, Italy}
+
 \begin{abstract}
+
+In this work we discuss differences and similarities between
+traditional many-body methods like variational and diffusion Monte
+Carlo approaches and unsupervised machine learning methods based on
+so-called Boltzmann machines. In particular we point to similarities
+between Monte Carlo approaches with shadow wave functions and
+restricted Boltzmann machines. We discuss their pros and cons and
+apply the various algorithms to studies of quantum dots. These are
+systems of interacting electrons confined to move in low-dimensional
+regions. We show also that the results obtained with restricted Boltzmann machines are of the same quality as those obtained with traditional variational Monte Carlo approaches.
 
@@
 
-\section{Introduction}
+\begin{document}
+
+\section{Introduction and motivation}
+
+Traditional many-body methods like full configuration interaction
+theory (FCI), coupled-cluster (CC) theory, many-body perturbation
+theory (MBPT), in-medium similarity renormalization group (IMSRG),
+various Monte Carlo methods and many other many-body approaches, have
+been rather successful in describing properties of interacting
+many-body systems. These methods have been widely used in condensed
+matter physics, nuclear physics, quantum chemistry and materials
+science, just to mention a few of the areas of applicability.
+
+However, essentially all of these methods face what is
+normally called the curse of dimensionality. For wave function based
+method like FCI or CC theories where the original continuous problem
+is discretized in terms selected basis functions and/or many-body
+excitations, the dimensionality of the systems under study grows almost
+exponentially when larger basis sets and/or number of particles are
+included. As an example, nuclear physics 
+systems close to the limits of stability pose in
+particular a tough problem to many-body practitioners in terms of a dramaticn
+increase of the number of relevant degrees of freedom. To describe say
+weakly bound nuclei within the framework of a wave function based
+approach, requires often a single-particle basis which includes bound,
+weakly bound and unbound states, increasing thereby considerably the
+number of many-body basis states. This renders often a standard
+FCI calculation infeasible. Within the above mentioned
+many-body methods there are however several interesting theoretical
+approachs which attempt at circumventing the dimensionality
+curse. Smarter basis sets is one of these approaches, as well as the
+resummation of specific correlations and the recently proposed
+stochastic sampling of many-body states in both FCI and CC
+calculations \cite{alavi2009,thom2017}.
+
+Recently, several authors have pointed to  possibilities within the
+broad fields of Machine Learning and Quantum Computing as approaches
+that hold great promise in tackling the ever increasing
+dimensionalities. There are several groups worldwide which now focus
+on Machine Learning and/or Quantum Computing applied to many-body
+problems.
+
+This work is inspired by the idea of representing the wave function with
+a restricted Boltzmann machine (RBM), presented recently by \href{{http://science.sciencemag.org/content/355/6325/602}}{G.
+Carleo and M.
+Troyer, Science \textbf{355}, Issue 6325, pp.
+602-606 (2017)}.
+They
+named such a wave function/network a \emph{neural network quantum state} (NQS). In their article they apply it to the quantum mechanical
+spin lattice systems of the Ising model and Heisenberg model, with
+encouraging results.
+
+Here we try to implement these algorithms to many-body systems that mimick a nuclear physics like problem, namely electrons confined to move in harmonic oscillator like potentials.
+
+Machine learning (ML) is an extremely rich field, in spite of its
+young age. The increases we have seen during the last three decades in
+computational capabilities have been followed by developments of
+methods and techniques for analyzing and handling large date sets,
+relying heavily on statistics, computer science and mathematics.  The
+field is rather new and developing rapidly.
+
+Almost every problem in ML and data science starts with the same ingredients:
+\begin{itemize}
+\item The dataset $\mathbf{x}$ (could be some observable quantity of the system we are studying)
+
+\item A model which is a function of a set of parameters $\mathbf{\alpha}$ that relates to the dataset, say a likelihood  function $p(\mathbf{x}\vert \mathbf{\alpha})$ or just a simple model $f(\mathbf{\alpha})$
+
+\item A so-called \textbf{cost} function $\mathcal{C} (\mathbf{x}, f(\mathbf{\alpha}))$ which allows us to decide how well our model represents the dataset. 
+\end{itemize}
+
+\noindent
+We seek to minimize the function $\mathcal{C} (\mathbf{x}, f(\mathbf{\alpha}))$ by finding the parameter values which minimize $\mathcal{C}$. This leads to  various minimization algorithms.
+
+\section{What is Machine Learning}
+Machine learning is the science of giving computers the ability to
+learn without being explicitly programmed.  The idea is that there
+exist generic algorithms which can be used to find patterns in a broad
+class of data sets without having to write code specifically for each
+problem. The algorithm will build its own logic based on the data.
+
+Machine learning is a subfield of computer science, and is closely
+related to computational statistics.  It evolved from the study of
+pattern recognition in artificial intelligence (AI) research, and has
+made contributions to AI tasks like computer vision, natural language
+processing and speech recognition. It has also, especially in later
+years, found applications in a wide variety of other areas, including
+bioinformatics, economy, physics, finance and marketing.
+
+You will notice however that many of the basic ideas discussed do come from Physics!
+
+The approaches to machine learning are many, but are often split into two main categories. 
+In \emph{supervised learning} we know the answer to a problem,
+and let the computer deduce the logic behind it. On the other hand, \emph{unsupervised learning}
+is a method for finding patterns and relationship in data sets without any prior knowledge of the system.
+Some authours also operate with a third category, namely \emph{reinforcement learning}. This is a paradigm 
+of learning inspired by behavioural psychology, where learning is achieved by trial-and-error, 
+solely from rewards and punishment.
+
+Another way to categorize machine learning tasks is to consider the desired output of a system.
+Some of the most common tasks are:
+
+\begin{itemize}
+  \item Classification: Outputs are divided into two or more classes.
+The goal is to   produce a model that assigns inputs into one of these classes.
+An example is to identify  digits based on pictures of hand-written ones.
+Classification is typically supervised learning.
+
+  \item Regression: Finding a functional relationship between an input data set and a reference data set.   The goal is to construct a function that maps input data to continuous output values.
+
+  \item Clustering: Data are divided into groups with certain common traits, without knowing the different groups beforehand.  It is thus a form of unsupervised learning.
+\end{itemize}
+
+\item An excellent reference, \href{{https://arxiv.org/abs/1803.08823}}{Mehta \emph{et al.}, arXiv:1803.08823 and Physics Reports in press (2018)}
+
+Here we will use so-called \textbf{reduced Boltzmann Machines} to simulate quantum many-body problems.
+For Monte Carlo aficionados, there is a very close similarity with what are called \textbf{shadow wave functions}, see the work of \href{{https://journals.aps.org/pre/abstract/10.1103/PhysRevE.90.053304}}{Pederiva and Kalos and collaborators, Phys Rev.
+E 90, 053304 (2014)}.
+
+\section{The system: two or more electrons in a harmonic oscillator trap in two dimensions}
+
+The Hamiltonian of the quantum dot is given by
+\[ \hat{H} = \hat{H}_0 + \hat{V}, 
+\]
+where $\hat{H}_0$ is the many-body HO Hamiltonian, and $\hat{V}$ is the
+inter-electron Coulomb interactions. In dimensionless units,
+\[ \hat{V}= \sum_{i < j}^N \frac{1}{r_{ij}},
+\]
+with $r_{ij}=\sqrt{\mathbf{r}_i^2 - \mathbf{r}_j^2}$.
+
+This leads to the  separable Hamiltonian, with the relative motion part given by ($r_{ij}=r$)
+\[ 
+\hat{H}_r=-\nabla^2_r + \frac{1}{4}\omega^2r^2+ \frac{1}{r},
+\]
+plus a standard Harmonic Oscillator problem  for the center-of-mass motion.
+This system has analytical solutions in two and three dimensions (\href{{https://journals.aps.org/pra/abstract/10.1103/PhysRevA.48.3561}}{M. Taut 1993 and 1994}).
+
+\subsection{Quantum Monte Carlo Motivation}
+
+Given a hamiltonian $H$ and a trial wave function $\Psi_T$, the variational principle states that the expectation value of $\langle H \rangle$, defined through 
+\[
+   \langle E \rangle =
+   \frac{\int d\bm{R}\Psi^{\ast}_T(\bm{R})H(\bm{R})\Psi_T(\bm{R})}
+        {\int d\bm{R}\Psi^{\ast}_T(\bm{R})\Psi_T(\bm{R})},
+\]
+is an upper bound to the ground state energy $E_0$ of the hamiltonian $H$, that is 
+\[
+    E_0 \le \langle H \rangle .
+\]
+
+In general, the integrals involved in the calculation of various
+expectation values are multi-dimensional ones. Traditional integration
+methods such as the Gauss-Legendre will not be adequate for say the
+computation of the energy of a many-body system.
+
+\[
+   E_L(\bm{R},\bm{\alpha})=\frac{1}{\psi_T(\bm{R},\bm{\alpha})}H\psi_T(\bm{R},\bm{\alpha}),
+\]
+called the local energy, which, together with our trial PDF yields
+\[
+  E[\bm{\alpha}]=\int P(\bm{R})E_L(\bm{R},\bm{\alpha}) d\bm{R}\approx \frac{1}{N}\sum_{i=1}^NE_L(\bm{R_i},\bm{\alpha})
+\]
+with $N$ being the number of Monte Carlo samples.
+
+We want to perform  a Variational Monte Carlo calculation of the ground state of two electrons in a quantum dot well with different oscillator energies, assuming total spin $S=0$.
+Our trial wave function has the following form
+\begin{equation}
+   \psi_{T}(\bm{r}_1,\bm{r}_2) = 
+   C\exp{\left(-\alpha_1\omega(r_1^2+r_2^2)/2\right)}
+   \exp{\left(\frac{r_{12}}{(1+\alpha_2 r_{12})}\right)}, 
+\label{eq:trial}
+\end{equation}
+where the $\alpha$s represent our variational parameters, two in this case.
+
+To find an ansatz for the correlated part of the wave function, it is
+useful to rewrite the two-particle local energy in terms of the
+relative and center-of-mass motion.  Let us denote the distance
+between the two electrons as $r_{12}$. We omit the center-of-mass
+motion since we are only interested in the case when $r_{12}
+\rightarrow 0$. The contribution from the center-of-mass (CoM)
+variable $\bm{R}_{\mathrm{CoM}}$ gives only a finite contribution.  We
+focus only on the terms that are relevant for $r_{12}$ and for three
+dimensions. The relevant local energy becomes then
+
+\[
+\lim_{r_{12} \rightarrow 0}E_L(R)=
+    \frac{1}{{\cal R}_T(r_{12})}\left(2\frac{d^2}{dr_{ij}^2}+\frac{4}{r_{ij}}\frac{d}{dr_{ij}}+
+\frac{2}{r_{ij}}-\frac{l(l+1)}{r_{ij}^2}+2E
+\right){\cal R}_T(r_{12}) = 0.
+\]
+Set $l=0$ and we have the so-called \textbf{cusp} condition
+\[
+\frac{d {\cal R}_T(r_{12})}{dr_{12}} = -\frac{1}{2(l+1)}
+{\cal R}_T(r_{12})\qquad r_{12}\to 0
+\]
+
+The above  results in
+\[
+{\cal R}_T  \propto \exp{(r_{ij}/2)}, 
+\]
+for anti-parallel spins and 
+\[
+{\cal R}_T  \propto \exp{(r_{ij}/4)}, 
+\]
+for anti-parallel spins.
+
+This is the so-called cusp condition for the relative motion,
+resulting in a minimal requirement for the correlation part of the
+wave fuction.  For general systems containing more than say two
+electrons, we have this condition for each electron pair $ij$.
+
+The elements of the gradient of the local energy are then (using the chain rule and the hermiticity of the Hamiltonian)
+\[
+\bar{E}_{i}= 2\left( \langle \frac{\bar{\Psi}_{i}}{\Psi}E_L\rangle -\langle \frac{\bar{\Psi}_{i}}{\Psi}\rangle\langle E_L \rangle\right).
+\]
+From a computational point of view it means that you need to compute the expectation values of 
+\[
+\langle \frac{\bar{\Psi}_{i}}{\Psi}E_L\rangle,
+\]
+and
+\[
+\langle \frac{\bar{\Psi}_{i}}{\Psi}\rangle\langle E_L\rangle
+\]
+These integrals are evaluted using MC intergration (with all its possible error sources). 
+We can then use methods like stochastic gradient or other minimization methods to find the optimal variational parameters (I don't discuss this topic here, but these methods are very important in ML).
+
+We have a model, our likelihood function.
+
+How should we define the cost function?
+
+\section{Why Boltzmann machines?}
+
+What is known as restricted Boltzmann Machines (RMB) have received a
+lot of attention lately.  One of the major reasons is that they can be
+stacked layer-wise to build deep neural networks that capture
+complicated statistics.
+
+The original RBMs had just one visible layer and a hidden layer, but recently so-called Gaussian-binary RBMs have gained quite some popularity in imaging since they are capable of modeling continuous data that are common to natural images.
+
+Furthermore, they have been used to solve complicated quantum mechanical many-particle problems or classical statistical physics problems like the Ising and Potts classes of models.
+
+Why use a generative model rather than the more well known discriminative deep neural networks (DNN)?
+
+\begin{itemize}
+\item Discriminitave methods have several limitations: They are mainly supervised learning methods, thus requiring labeled data.
+And there are tasks they cannot accomplish, like drawing new examples from an unknown probability distribution.
+
+\item A generative model can learn to represent and sample from a probability distribution.
+The core idea is to learn a parametric model of the probability distribution from which the training data was drawn.
+As an example
+\begin{enumerate}
+
+ \item A model for images could learn to draw new examples of cats and dogs, given a training dataset of images of cats and dogs.
+
+ \item Generate a sample of an ordered or disordered Ising model phase, having been given samples of such phases.
+
+ \item Model the trial function for Monte Carlo calculations
+\end{enumerate}
+
+\end{itemize}
+
+\begin{enumerate}
+\item Both use gradient-descent based learning procedures for minimizing cost functions
+
+\item Energy based models don't use backpropagation and automatic differentiation for computing gradients, instead turning to Markov Chain Monte Carlo methods.
+
+\item DNNs often have several hidden layers.
+A restricted Boltzmann machine has only one hidden layer, however several RBMs can be stacked to make up Deep Belief Networks, of which they constitute the building blocks.
+\end{enumerate}
+
+A standard BM network is divided into a set of observable and visible
+units $\hat{x}$ and a set of unknown hidden units/nodes $\hat{h}$.
+
+Additionally there can be bias nodes for the hidden and visible
+layers. These biases are normally set to $1$.
+
+BMs are stackable, meaning they cwe can train a BM which serves as
+input to another BM. We can construct deep networks for learning
+complex PDFs. The layers can be trained one after another, a feature
+which makes them popular in deep learning
+
+However, they are often hard to train. This leads to the introduction
+of so-called restricted BMs, or RBMS.  Here we take away all lateral
+connections between nodes in the visible layer as well as connections
+between nodes in the hidden layer. The network is illustrated in the
+figure below.
+
+\centerline{\includegraphics[width=1.0\linewidth]{figures/RBM.pdf}}
+
+\subsection{The network}
+\begin{enumerate}
+ \item A function $\mathbf{x}$ that represents the visible layer, a vector of $M$ elements (nodes).
+This layer represents both what the RBM might be given as training input, and what we want it to be able to reconstruct.
+This might for example be the pixels of an image, the spin values of the Ising model, or coefficients representing speech.
+
+ \item The function $\mathbf{h}$ represents the hidden, or latent, layer. A vector of $N$ elements (nodes). Also called "feature detectors".
+\end{enumerate}
+
+The goal of the hidden layer is to increase the model's expressive
+power. We encode complex interactions between visible variables by
+introducing additional, hidden variables that interact with visible
+degrees of freedom in a simple manner, yet still reproduce the complex
+correlations between visible degrees in the data once marginalized
+over (integrated out).
+
+Examples of this trick being employed in physics: 
+\begin{enumerate}
+ \item The Hubbard-Stratonovich transformation
+
+ \item The introduction of ghost fields in gauge theory
+
+ \item Shadow wave functions in Quantum Monte Carlo simulations
+\end{enumerate}
+
+\noindent
+\textbf{The network parameters, to be optimized/learned}:
+\begin{enumerate}
+ \item $\mathbf{a}$ represents the visible bias, a vector of same length as $\mathbf{x}$.
+
+ \item $\mathbf{b}$ represents the hidden bias, a vector of same lenght as $\mathbf{h}$.
+
+ \item $W$ represents the interaction weights, a matrix of size $M\times N$.
+\end{enumerate}
+
+\subsection{Joint distribution}
+
+The restricted Boltzmann machine is described by a Boltzmann distribution
+\begin{align}
+	P_{rbm}(\mathbf{x},\mathbf{h}) = \frac{1}{Z} e^{-\frac{1}{T_0}E(\mathbf{x},\mathbf{h})},
+\end{align}
+where $Z$ is the normalization constant or partition function, defined as 
+\begin{align}
+	Z = \int \int e^{-\frac{1}{T_0}E(\mathbf{x},\mathbf{h})} d\mathbf{x} d\mathbf{h}.
+\end{align}
+It is common to ignore $T_0$ by setting it to one.
+
+The function $E(\mathbf{x},\mathbf{h})$ gives the \textbf{energy} of a
+configuration (pair of vectors) $(\mathbf{x}, \mathbf{h})$. The lower
+the energy of a configuration, the higher the probability of it. This
+function also depends on the parameters $\mathbf{a}$, $\mathbf{b}$ and
+$W$. Thus, when we adjust them during the learning procedure, we are
+adjusting the energy function to best fit our problem.
+
+An expression for the energy function is
+\[
+E(\hat{x},\hat{h}) = -\sum_{ia}^{NA}b_i^a \alpha_i^a(x_i)-\sum_{jd}^{MD}c_j^d \beta_j^d(h_j)-\sum_{ijad}^{NAMD}b_i^a \alpha_i^a(x_i)c_j^d \beta_j^d(h_j)w_{ij}^{ad}.
+\]
+
+Here $\beta_j^d(h_j)$ and $\alpha_i^a(x_j)$ are so-called transfer functions that map a given input value to a desired feature value.
+The labels $a$ and $d$ denote that there can be multiple transfer functions per variable.
+The first sum depends only on the visible units.
+The second on the hidden ones.
+\textbf{Note} that there is no connection between nodes in a layer.
+
+The quantities $b$ and $c$ can be interpreted as the visible and hidden biases, respectively.
+
+The connection between the nodes in the two layers is given by the weights $w_{ij}$.
+
+There are different variants of RBMs, and the differences lie in the types of visible and hidden units we choose as well as in the implementation of the energy function $E(\mathbf{x},\mathbf{h})$.
+
+\paragraph{Binary-Binary RBM:}
+
+RBMs were first developed using binary units in both the visible and hidden layer. The corresponding energy function is defined as follows:
+\begin{align}
+	E(\mathbf{x}, \mathbf{h}) = - \sum_i^M x_i a_i- \sum_j^N b_j h_j - \sum_{i,j}^{M,N} x_i w_{ij} h_j,
+\end{align}
+where the binary values taken on by the nodes are most commonly 0 and 1.
+
+\paragraph{Gaussian-Binary RBM:}
+
+Another varient is the RBM where the visible units are Gaussian while the hidden units remain binary:
+\begin{align}
+	E(\mathbf{x}, \mathbf{h}) = \sum_i^M \frac{(x_i - a_i)^2}{2\sigma_i^2} - \sum_j^N b_j h_j - \sum_{i,j}^{M,N} \frac{x_i w_{ij} h_j}{\sigma_i^2}. 
+\end{align}
+
+\subsection{Cost function}
+
+When working with a training dataset, the most common training approach is maximizing the log-likelihood of the training data.
+The log likelihood characterizes the log-probability of generating the observed data using our generative model.
+Using this method our cost function is chosen as the negative log-likelihood.
+The learning then consists of trying to find parameters that maximize the probability of the dataset, and is known as Maximum Likelihood Estimation (MLE).
+Denoting the parameters as $\bm{\theta} = a_1,...,a_M,b_1,...,b_N,w_{11},...,w_{MN}$, the log-likelihood is given by
+\begin{align}
+	\mathcal{L}(\{ \theta_i \}) &= \langle \text{log} P_\theta(\bm{x}) \rangle_{data} \\
+	&= - \langle E(\bm{x}; \{ \theta_i\}) \rangle_{data} - \text{log} Z(\{ \theta_i\}),
+\end{align}
+where we used that the normalization constant does not depend on the data, $\langle \text{log} Z(\{ \theta_i\}) \rangle = \text{log} Z(\{ \theta_i\})$
+Our cost function is the negative log-likelihood, $\mathcal{C}(\{ \theta_i \}) = - \mathcal{L}(\{ \theta_i \})$
+
+\subsection{RBMs for the quantum many body problem}
+
+The idea of applying RBMs to quantum many body problems was presented by G. Carleo and M. Troyer, working with ETH Zurich and Microsoft Research.
+
+Some of their motivation included
+
+\begin{itemize}
+\item "The wave function $\Psi$ is a monolithic mathematical quantity that contains all the information on a quantum state, be it a single particle or a complex molecule.
+In principle, an exponential amount of information is needed to fully encode a generic many-body quantum state."
+
+\item There are still interesting open problems, including fundamental questions ranging from the dynamical properties of high-dimensional systems to the exact ground-state properties of strongly interacting fermions.
+
+\item The difficulty lies in finding a general strategy to reduce the exponential complexity of the full many-body wave function down to its most essential features. That is
+\begin{enumerate}
+
+\item $\rightarrow$ Dimensional reduction
+
+\item $\rightarrow$ Feature extraction
+
+\end{enumerate}
+
+\noindent
+\item Among the most successful techniques to attack these challenges, artifical neural networks play a prominent role.
+
+\item Want to understand whether an artifical neural network may adapt to describe a quantum system.
+\end{itemize}
+
+\subsection{Choose the right RBM}
+
+Carleo and Troyer applied the RBM to the quantum mechanical spin lattice systems of the Ising model and Heisenberg model, with encouraging results.
+Our goal is to test the method on systems of moving particles.
+For the spin lattice systems it was natural to use a binary-binary RBM, with the nodes taking values of 1 and -1.
+For moving particles, on the other hand, we want the visible nodes to be continuous, representing position coordinates.
+Thus, we start by choosing a Gaussian-binary RBM, where the visible nodes are continuous and hidden nodes take on values of 0 or 1.
+If eventually we would like the hidden nodes to be continuous as well the rectified linear units seem like the most relevant choice.
+
+\subsection{Representing the wave function}
+The wavefunction should be a probability amplitude depending on $\bm{x}$. The RBM model is given by the joint distribution of $\bm{x}$ and $\bm{h}$
+\begin{align}
+	F_{rbm}(\mathbf{x},\mathbf{h}) = \frac{1}{Z} e^{-\frac{1}{T_0}E(\mathbf{x},\mathbf{h})}.
+\end{align}
+To find the marginal distribution of $\bm{x}$ we set:
+\begin{align}
+	F_{rbm}(\mathbf{x}) &= \sum_\mathbf{h} F_{rbm}(\mathbf{x}, \mathbf{h}) \\
+				&= \frac{1}{Z}\sum_\mathbf{h} e^{-E(\mathbf{x}, \mathbf{h})}.
+\end{align}
+Now this is what we use to represent the wave function, calling it a neural-network quantum state (NQS)
+\begin{align}
+	\Psi (\mathbf{X}) &= F_{rbm}(\mathbf{x}) \\
+	&= \frac{1}{Z}\sum_{\bm{h}} e^{-E(\mathbf{x}, \mathbf{h})} \\
+	&= \frac{1}{Z} \sum_{\{h_j\}} e^{-\sum_i^M \frac{(x_i - a_i)^2}{2\sigma^2} + \sum_j^N b_j h_j + \sum_{i,j}^{M,N} \frac{x_i w_{ij} h_j}{\sigma^2}} \\
+	&= \frac{1}{Z} e^{-\sum_i^M \frac{(x_i - a_i)^2}{2\sigma^2}} \prod_j^N (1 + e^{b_j + \sum_i^M \frac{x_i w_{ij}}{\sigma^2}}). \\
+\end{align}
+
+\subsection{Choose the cost function}
+Now we don't necessarily have training data (unless we generate it by using some other method).
+However, what we do have is the variational principle which allows us to obtain the ground state wave function by minimizing the expectation value of the energy of a trial wavefunction (corresponding to the untrained NQS).
+Similarly to the traditional variational Monte Carlo method then, it is the local energy we wish to minimize.
+The gradient to use for the stochastic gradient descent procedure is
+\begin{align}
+	G_i = \frac{\partial \langle E_L \rangle}{\partial \theta_i}
+	= 2(\langle E_L \frac{1}{\Psi}\frac{\partial \Psi}{\partial \theta_i} \rangle - \langle E_L \rangle \langle \frac{1}{\Psi}\frac{\partial \Psi}{\partial \theta_i} \rangle ),
+\end{align}
+where the local energy is given by
+\begin{align}
+	E_L = \frac{1}{\Psi} \hat{\mathbf{H}} \Psi.
+\end{align}
+
+The trial wave function is based on the product of a Slater
+determinant with Gaussian orbitals, a simple Jastrow factor
+$\exp{(r_{ij})}$ and the reduced Boltzmann machines.
+
+\centerline{\includegraphics[width=0.9\linewidth]{figures/figN2.pdf}}
+
+\centerline{\includegraphics[width=0.9\linewidth]{figures/figN6.pdf}}
+
+\section{Conclusions and where do we stand}
+
+\begin{itemize}
+\item A simple extension of the work of \href{{http://science.sciencemag.org/content/355/6325/602}}{G.
+Carleo and M.
+Troyer, Science \textbf{355}, Issue 6325, pp.
+602-606 (2017)} gives excellent results for two-electron systems as well as good agreement with standard VMC calculations for $N=6$ and $N=12$ electrons.
+
+\item Minimization problem can be tricky.
+
+\item Anti-symmetry dealt with multiplying the trail wave function with an optimized Slater determinant.
+
+\item To come: Analysis of wave function from ML and compare with diffusion and Variational Monte Carlo calculations as well as the analytical results of Taut for the two-electron case.
+
+\item Extend to more fermions. How do we deal with the antisymmetry of the multi-fermion wave function?
+\begin{enumerate}
+
+ \item Here we used standard Hartree-Fock theory to define an optimal Slater determinant. Takes care of the antisymmetry. What about constructing an anti-symmetrized network function?
+
+ \item Use thereafter ML to determine the correlated part of the wafe function (including a standard Jastrow factor).
+
+ \item Test this for multi-fermion systems and compare with other many-body methods.
+
+\end{enumerate}
+
+\noindent
+\item Can we use ML to find out which correlations are relevant and thereby diminish the dimensionality problem in say CC or SRG theories?
+\end{itemize}
 

### 2019-03-19 doc/MachineLearning/article2.tex
update

@@
 traditional many-body methods like variational and diffusion Monte
-Carlo approaches and unsupervised machine learning methods based on
-so-called Boltzmann machines. In particular we point to similarities
-between Monte Carlo approaches with shadow wave functions and
-restricted Boltzmann machines. We discuss their pros and cons and
-apply the various algorithms to studies of quantum dots. These are
-systems of interacting electrons confined to move in low-dimensional
-regions. We show also that the results obtained with restricted Boltzmann machines are of the same quality as those obtained with traditional variational Monte Carlo approaches.
+Carlo approaches and novel unsupervised machine learning methods. The
+latter are based on so-called Boltzmann machines. In particular we
+point to similarities between Monte Carlo approaches with shadow wave
+functions and restricted Boltzmann machines. We discuss their pros and
+cons and apply the various algorithms to studies of quantum
+dots. These are systems of interacting electrons confined to move in
+low-dimensional regions. For two interacting electrons only we have
+analytic solutions for the eigenpairs, providing us the with ways to benchmark the quality of our numerical simulations.
+
+We show
+also that the results obtained with restricted Boltzmann machines are
+of the same quality as those obtained with traditional variational
+Monte Carlo approaches.
 

### 2019-03-19 doc/MachineLearning/article2.tex
update

@@
 Traditional many-body methods like full configuration interaction
-theory (FCI), coupled-cluster (CC) theory, many-body perturbation
-theory (MBPT), in-medium similarity renormalization group (IMSRG),
-various Monte Carlo methods and many other many-body approaches, have
-been rather successful in describing properties of interacting
-many-body systems. These methods have been widely used in condensed
-matter physics, nuclear physics, quantum chemistry and materials
-science, just to mention a few of the areas of applicability.
+theory (FCI) \cite{fcirefs}, coupled-cluster (CC) theory
+\cite{ccrefs}, many-body perturbation theory (MBPT) \cite{mbptrefs},
+in-medium similarity renormalization group (IMSRG) \cite{imsrgrefs},
+various Monte Carlo methods \cite{mcrefs} and many other many-body
+approaches, have been rather successful in describing properties of
+interacting many-body systems. These methods have been widely used in
+condensed matter physics, nuclear physics, quantum chemistry and
+materials science, just to mention a few of the areas of
+applicability.
 
@@
 included. As an example, nuclear physics 
-systems close to the limits of stability pose in
+systems close to the limits of stability \cite{nplims} pose in
 particular a tough problem to many-body practitioners in terms of a dramaticn
@@
 on Machine Learning and/or Quantum Computing applied to many-body
-problems.
-
-This work is inspired by the idea of representing the wave function with
-a restricted Boltzmann machine (RBM), presented recently by \href{{http://science.sciencemag.org/content/355/6325/602}}{G.
-Carleo and M.
-Troyer, Science \textbf{355}, Issue 6325, pp.
-602-606 (2017)}.
-They
-named such a wave function/network a \emph{neural network quantum state} (NQS). In their article they apply it to the quantum mechanical
-spin lattice systems of the Ising model and Heisenberg model, with
-encouraging results.
-
-Here we try to implement these algorithms to many-body systems that mimick a nuclear physics like problem, namely electrons confined to move in harmonic oscillator like potentials.
-
+problems \cite{CarleoTroyer2017}.
 Machine learning (ML) is an extremely rich field, in spite of its
@@
 field is rather new and developing rapidly.
+
+Within this setting, the aim of this work is to explore similarities
+and differences between widely used many-body methods like the
+Variational Monte Carlo or the diffusion Monte Carlo approaches and
+unsupervised Machine Learning methods using so-called Boltzmann
+machines. We will also point to similarities between Monte Carlo
+approaches with shadow wave functions and Machine Learning algorithms
+that employ Boltzmann machines. Since the field of Machine Learning is
+evolving rather rapidly and physicists may not be fully familiar with
+the plethora of Machine Learning algorithms, we will try in this
+work to keep a more pedestrian approach. In particular, we will try to outline 
+some of the similarities between Monte Carlo methods for solving quantum mechanical many-body problems and Machine Learning algorithms.
+
+We start thus with a brief review of central aspects of Monte Carlo
+methods in the next section. Our model system is that of electrons
+confined to move in two or three dimensions, so-called quantum
+dots. We will in particular focus on idealized quantum dots with
+electrons moving in a harmonic oscillator potential. There are several
+reasons for focusing on quantum dots. For two interacting electrons
+there are analytical eigenpairs, allowing us thereby to benchmark
+various many-body methods \cite{taut1993,taut1994}. For more than two
+electrons, there are extensive reference calculations based on a
+variety of many-body methods. These calculations, in addition to new
+ones presented here, will allow us to assess the relevance of our
+results obtained with Machine Learning algorithms.
+
+Thereafter, in Section \ref{sec:MLmethods}, we present and discuss the implementation of Machine Learning algorithms.
+In that section we present also a discussion of possible similarities between Boltzmann machines and so-called Shadow wave functions.
+
+\section{What is Machine Learning}
 
@@
 
-\section{What is Machine Learning}
 Machine learning is the science of giving computers the ability to

### 2019-04-17 doc/TimeDependence/Draft/Article.tex
added new draft

--- 2018-04-14 doc/TimeDependence/Tex/Article.tex
@@
 \affiliation{Department of Physics, University of Oslo, N-0316 Oslo, Norway}
+
+\author{\O yvind Sigmundson Sch\o yen}
+\email[Electronic mail:]{o.s.schoyen@fys.uio.no}
+\affiliation{Department of Physics, University of Oslo, N-0316 Oslo, Norway}
+
+\author{Sebastian G.~Winther-Larsen}
+\email[Electronic mail:]{sebastwi@fys.uio.no}
+\affiliation{Department of Physics, University of Oslo, N-0316 Oslo, Norway}
+
+\author{Morten Hjorth-Jensen}
+\email[Electronic mail:]{hjensen@msu.edu}
+\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
+\affiliation{Department of Physics and Center for Computing in Science Education, University of Oslo, N-0316 Oslo, Norway}
 

### 2019-04-17 doc/TimeDependence/Paper1/main.tex
added new draft

\documentclass[aip,jcp,reprint,floatfix]{revtex4-1}

\usepackage{graphicx}
\usepackage{comment}
\usepackage{lipsum}
\usepackage[utf8]{inputenc}
\usepackage[toc,page]{appendix}
\usepackage{braket}
\usepackage{amsmath,amssymb}
\usepackage{amsthm}
\usepackage{hyperref}
\usepackage{pgfplots}
\usepackage{bbold}

\usetikzlibrary{external}
\tikzexternalize[prefix=figures/]

\pgfplotsset{compat=1.15}

\renewcommand{\vec}[1]{\boldsymbol{#1}}
\newcommand{\hamiltonian}{\hat{H}}
\newcommand{\onehamil}{\hat{h}}
\newcommand{\twohamil}{\hat{u}}

\newcommand{\vconf}{\hat{v}_{\text{conf}}}
\newcommand{\vfield}{\hat{v}_{\text{field}}}
\newcommand{\envelope}{\vec{\mathcal{E}}}
\newcommand{\polarization}{\vec{\epsilon}}

\begin{document}

\title{Electron dynamics in quantum dot models using techniques from Quantum Chemistry}

\author{Håkon Emil Kristiansen}
\email{haakoek@uio.no}
\affiliation{Hylleraas Centre for Quantum Molecular Sciences, Department of Chemistry, University of Oslo,N-0315 Oslo, Norway}

\author{\O yvind Sigmundson Sch\o yen}
\email{o.s.schoyen@fys.uio.no}
\affiliation{Department of Physics, University of Oslo, N-0316 Oslo, Norway}

\author{Sebastian G.~Winther-Larsen}
\email{s.g.winther-larsen@fys.uio.no}
\affiliation{Department of Physics, University of Oslo, N-0316 Oslo, Norway}

\author{Morten Hjorth-Jensen}
\email{hjensen@msu.edu}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\affiliation{Department of Physics and Center for Computing in Science Education, University of Oslo, N-0316 Oslo, Norway}
\date{\today}

\begin{abstract}
    This is the abstract.
\end{abstract}

\maketitle

\section{TODO}
\begin{itemize}
    \item How do you treat the centre of mass problem???
    \item The harmonic potential theorem\cite{HarmonicPotentialTheorem}: this depends on basis size! One way to check this is to let $u^{pq}_{rs} = 0$ and use a small number of basis functions. The short story: the particle density (in time) is rigidly translated when the confining potential is harmonic and the applied field is the simple dipole laser we use! 
    \item Furthermore, if I read correctly, we should observe that the (for example) the time dependent energy is effectively the same as for a single particle problem.

    Quote from\cite{Chakraborty}: "In conclusion, the interaction of electrons in quantum
dots leads to rich structure in their energy spectrum.
However, FIR spectroscopy cannot probe it when the
confining potential is quadratic because the optical excitations are then excitations of the cm and have exactly
the same energies as single-electron excitations."

    This is (again, if I understand the implications of the harmonic potential theorem correctly) due to the fact that the Hamiltonian separates into relative and center of mass coordinates with a harmonic potential and this type of applied field. This effect is independent of the number of particles. If we can actually demonstrate this, it is very solid foundation for claiming the correctness of the implementation. This also says something about the quality of the basis set. 
    \item Basically, what I am trying to say is that the "shape" of computed values (energy, overlap, expectation value of dipole, etc) is independent of the number of particles and will look like the corresponding values of the single particle problem apart from a scaling due to the number of particles. 
    \item Thus, it is maybe sensible to include comparisons with the single particle problem (which can be solved analytically or on a grid with f. ex RK4).
\end{itemize}

\section{Introduction}
Quantum mechanics is a theory that describes/models the properties of microscopic systems. It postulates that given the so-called wavefunction,
$\Psi(\mathbf{r},t)$, we can in principle compute all that there is to know about the system.  Given suitable initial conditions $\Psi(\mathbf{r},t)$ can be determined for all future time
by solving the time-dependent Schrödinger equation (TDSE),
\begin{equation}
\label{TDSE}
 i \hbar \frac{\partial }{\partial t} \Psi(\mathbf{r},t) = \hat{H} \Psi(\mathbf{r},t).
\end{equation}
The initial condition is typically taken as some linear combination of the eigenfunctions. In particular, a prototypical choice of initial condition is the groundstate of the system. The eigenfunctions are found by solving the time-independent Schrödinger (TISE) equation
\begin{equation}
 \hat{H} \Psi(\mathbf{r}) = E \Psi(\mathbf{r}).
\end{equation}
Exact/analytic solutions to the TISE is possible only for the
simplest systems and general solutions to the TDSE are even
rarer. Thus, one must resort to approximate/numeric methods. Of
particular interest is the so-called many-body problem where one
considers systems of two or more interacting particles.

Several approaches for solving the many-body TISE
numerically/approximately has been devised such as Hartree-Fock
(HF), Density Functional Theory (DFT), Configuration Interaction (CI)
and Coupled Cluster (CC). While efficient, Hartree-Fock and DFT are
insufficient if one wants a high degree of accuracy. Configuration
Interaction and Coupled Cluster methods are hierarchical in the sense
that one can systematically construct increasingly accurate
approximations. If the CI method is not truncated we have what is
known as Full Configuration Interaction (FCI). FCI can be seen as
exact (within some finite space), however it suffers from exponential
scaling. Truncated CI methods which would achieve polynomial scaling
are problematic since they are not size-consistent and
extensive. Truncated CC methods one the other hand are size-consistent
and extensive and achieves polynomial scaling. Due to this fact
Coupled Cluster is considered the gold standard (misleading and ill-omened term as history has shown that a gold standard, in terms of monetary policy, is a bad idea - it does not work) of many-body techniques if high accuracy is desired.

Similarly, there exists time-dependent generalizations of the methods discussed above. On one end we have the time-dependent Hartree-Fock (TDHF) and time-dependent density functional theory (TD-DFT) \cite{TDHF_2004, TD_DFT_Ullrich_Book}, which in this context are considered efficient/low-cost. 
Time-dependent full configuration interaction (TDFCI) and Multiconfiguration Time-Dependent Hartree-Fock (MCTDHF) on the other hand are considered the most accurate methods \cite{Hochstuhl2014}. However, TDCI and MCTDHF suffers from exponential scaling, which makes them impractical for systems with larger particle numbers.

In a recent article \cite{OATDCC_2012}, Simen Kvaal put forward a generalization of Coupled-Cluster theory to the time domain, where the orbitals are allowed to be time-dependent. The method is referred to as the orbital-adaptive time-dependent coupled-cluster method (OATDCC), which can be viewed as a hierarchical approximation to the MCTDHF/TDFCI methods. OATDCC inherits size-consistency and extensivity from the CC method and achieves polynomial scaling. While not cheap, in the conventional sense, OATDCC alleviates some of the problems related to TDFCI and MCTDHF.

We will apply several of the methods above in order to study electron dynamics in one- and two-dimensional quantum dot models. In particular, we will make a comparative analysis of the TDHF, TDCC (OATDCC) and TDFCI methods.

\section{Theory}
We assume that the reader is familiar with second-quantization/occupation-number theory. For a comprehensive introduction see for example \cite{szabo1996modern,helgaker2008molecular, shavitt2009many}. 
\subsection{Hartree-Fock Theory}
In the Hartree-Fock approximation, the Hamiltonian reduces to a one-body term and a two-body term, written in the simplest way using second quantisation as
\begin{equation}
    \hat{H} = \hat{h} + \hat{u} 
    = h^{p}_{q}c^\dagger_p c_q + \frac{1}{4}u^{pq}_{rs}c^\dagger_p c^\dagger_q c_s c_r
\end{equation}

and the expected value of the Hamiltonian is given by
\begin{equation}
    \label{HF_expecation}
    E[\Phi] = \bra{\Phi}\hat{H}\ket{\Phi} 
    = \bra{\phi_i}\hat{h}\ket{\phi_i} + \frac{1}{2}\bra{\phi_i\phi_j}\hat{u}\ket{\phi_i\phi_j}_{\text{AS}}
\end{equation}
where the assumption is that the ground state can be represented by a single Slater determinant
\begin{equation}
    \ket{\Phi} = \prod_{i=1}^{N} c^\dagger_i \ket{\text{vac}} = \ket{\phi_1 \dots \phi_N}.
\end{equation}

It can be shown that a Slater determinant wavefunction $\ket{\Phi_\text{HF}} = \ket{\phi_i \dots \phi_N}$ is a variational minimum of the function for $E[\Phi]$ in \autoref{HF_expecation} if the single-particle functions $\ket{\phi_i}$ that make up this Slater determinant satisfy the canonical Hartree-Fock equations,
\begin{equation}
    \label{canonical_HF}
    \hat{f}(\phi_1, \dots , \phi_N)\ket{\phi_i} = \epsilon_i \ket{\phi_i}.
\end{equation}
where $\hat{f}$ is a one-body operator called the Fock operator,
\begin{equation}
    \hat{f}(\phi_1, \dots \phi_N) \equiv \hat{h} + \hat{v}_{\text{HF}},
\end{equation}
wherein the most important point of the Hartree-Fock approximation is revealed - one constrains the problem to a one-body problem where each particle only experiences a mean field from the other particles in the system. The canonical Hartree-Fock equations (\autoref{canonical_HF}) is reformulated as a matrix eigenvalue problem. This eigenvalue equation is called the Roothan-Hall equations,
\begin{equation}
    \label{roothan_hall}
    F(U)U = U\epsilon,
\end{equation}
which is solved iteratively until self-consistency is reached, i.e. the next iteration does not contribute a significant amount to the mean field of the system.

\subsubsection{Time-Dependent Hartree-Fock}

We now assume that the Hamiltonian is time-dependent, satisfying the time-dependent Schrödinger equation (\autoref{TDSE}),
\begin{equation}
    \hat{H}(t) = \hat{h}(t) + \hat{u}(t).
\end{equation}

In time dependent Hartree-Fockn(TDHF) the wavefunction is approximated by a single Slater determinant, $\ket{\Phi(t)}$, where the orbitals $\ket{\phi_p(t)}$ are time dependent. The equations of motion for the orbitals are found by demanding that the action functional 
\begin{equation}
    \mathcal{S}[\ket{\phi_p(t)}] = \int_0^T \mathcal{L} dt 
\end{equation}
is stationary with respect to variations of the orbitals, i.e $\delta \mathcal{S} = 0$. This is referred to as the Lagrange formulation of the time dependent variational principle\cite{Hochstuhl2014}.

The Lagrangian $\mathcal{L}$ is given by 
\begin{equation}
    \mathcal{L} = \bra{\Phi(t)} \hat{H} - i\hbar \frac{\partial}{\partial t} \ket{\Phi(t)} 
    - \lambda_{ij}(t)\left(\braket{\phi_i(t)|\phi_j(t)} - \delta_{ij}\right),
\end{equation}
where the Lagrange multipliers $\lambda_{ij}(t)$ enurses orthonormality of the orbitals.

Computing the variations  with respect to the orbitals (see appendix \ref{TDHF_Derivation} for details) we obtain the TDHF equations 
\begin{equation}
    i \hbar \frac{\partial}{\partial t} \ket{\phi_p(t)} = \hat{f}(t) \ket{\phi_p(t)}.
\end{equation}

\subsection{Configuration Interaction Theory}

(This section is a bit informal, but at least there is something here.)

Configuration Interaction relies on representing the wavefunction as an expansion of orthonormal slater determinants built from a set of orthonormal single-particle functions,
\begin{equation}
    \label{hierarchical_CI}
    \ket{\Psi} = A_0\ket{\Phi} + \sum_{ia}A^a_i\ket{\Phi^a_i} 
    + \frac{1}{4}\sum_{ij} \sum_{ab} A^{ab}_{ij}\ket{\Phi^{ab}_{ij}} + \dots
\end{equation}
One may choose to truncate the computation in the hierarchical manner (e.g. stopping at the dots in \autoref{hierarchical_CI} would be CISD) or include all Slater determinants generated by the first $L$ single-particle functions $\phi_0$ through $\phi_{L-1}$. This gives a space of dimension $\binom{L}{N}$, and is called the full configuration interaction space (FCI space).

Inserting the CI wavefunction into the TISE yields
\begin{align}
    \sum_{K}\hat{H} A_{KJ} \ket{\Phi_K} &= \sum_{K} E_J A_{KJ} \ket{\Phi_K} \\
    \sum_{K} A_{KJ} \bra{\Phi_I}\hat{H} \ket{\Phi_K} &= \sum_{K} E_J A_{KJ} \bra{\Phi_I}\Phi_K\rangle \\
    HA &= AE,
\end{align}
where $H$ is a matrix reprentation of the Hamiltionian, and $A$ is a matrix containing the CI expansion coefficients.

\subsubsection{Time-Dependent Configuration Interaction}
Time-development is simply a matter of inserting one's CI expansion (\autoref{hierarchical_CI}) into the TDSE (\textcolor{blue}{Haakon:} formally we can use the time dependent variational principle, minimize/compute variations wrp. to CI coeffs and get the same equation),
\begin{align}
    i\hbar \sum_I \frac{\partial A_I(t)}{\partial t} \ket{\Phi_I} &= \sum_I A_i(t) \hat{H}(t) \ket{\Phi_I} \\
    i \hbar \sum_I \frac{\partial A_I(t)}{\partial t} \bra{\Phi_J}\Phi_I\rangle &= \sum_I A_I(t) \bra{\Phi_J} \hat{H}(t) \ket{\Phi_i} \\
    i\hbar \frac{\partial A_I(t)}{\partial t} &= H(t)A(t).
\end{align}
We find a relatively simple, yet large, set of equation on matrix form. One is now free to apply one's favourite numerical integration scheme in order to compute the time-development.

\subsection{Coupled Cluster Theory}

    Full Configuration Interaction (FCI) will provide you with an exact solution, but since the FCI space grows Pascal-triangularly (Question: Is the growth exponential when the growth does not follow an exponential function?) and because we are constrained by limited amount of computing power, FCI is not practically feasible. The Hartree-Fock method, on the other hand, approximates systems by ignoring electron-electron interaction and can therefore be quite inaccurate. Coupled Cluster to the rescue! Hurray. Coupled cluster cleverly include higher-order excitations and takes electron correlation effects into account. Time-developed coupled cluster (TDCC) can be tricky, however.

    \subsubsection{Ground State coupled cluster}
    In coupled cluster theory we employ an exponential ansatz for the $N$-particle correlated wavefunction 
    \begin{equation}
       \ket{\Psi_{\text{CC}}} = e^{\hat{T}}\ket{\Phi} 
    \end{equation}
    where $\ket{\Phi}$ is a reference state, typically the HF-state. The cluster operator $\hat{T} = \sum_{k=1}^N \hat{T}_k$ is a sum of $k$-particle $k$-hole excitation operators, $\hat{T}_k$, 
    \begin{equation}
        \hat{T}_k = \left(\frac{1}{k!}\right)^2 \sum_{abc\cdots ijk\cdots} \tau^{abc\cdots}_{ijk\cdots}   c_a^\dagger  c_i c_b^\dagger c_j  c_c^\dagger c_k \cdots 
    \end{equation}
    where the unknowns $\tau^{abc\cdots}_{ijk\cdots}$ are referred to as cluster amplitudes. The operators $\hat{T}_1, \hat{T}_2, \hat{T}_3, \cdots$ and corresponding amplitudes $\tau^a_i, \tau^{ab}_{ij}, \tau^{abc}_{ijk},\cdots$ are referred to as singly, doubly, triply-, etc excited operators and amplitudes respectively.

    Insertion of the exponential ansatz into the time independent Schrödinger equation results in 
    \begin{equation}
        \hat{H}e^{\hat{T}} \ket{\Phi} = E e^{\hat{T}} \ket{\Phi}.
    \end{equation}
    Multiplying with $e^{-T}$ we have that 
    \begin{equation}
        \bar{H}_{CC} \ket{\Phi} = E \ket{\Phi} \label{SimTransSE},
    \end{equation}
    where we have defined the similarity transformed Hamiltonian 
    \begin{equation}
        \bar{H}_{CC} \equiv e^{-\hat{T}}\hat{H}e^{\hat{T}}.
    \end{equation}

    Projecting equation (\ref{SimTransSE}) with the reference or any excited determinant gives 
    \begin{align}
        E_{CC} &= \braket{\Phi|\bar{H}_{cc}|\Phi} \label{CCenergy} \\
        0 &= \braket{\Phi^\mu|\bar{H}_{cc}|\Phi} \label{AmplitudeEqsCC},
    \end{align}
    where $\mu$ denotes a generic excitation. Eq. (\ref{CCenergy}) is the coupled cluster energy, while eq. (\ref{AmplitudeEqsCC}) defines a set of nonlinear equations for the unknowns cluster amplitudes. In the limit where $\hat{T}$ is not truncated eqs. (\ref{CCenergy}) and (\ref{AmplitudeEqsCC}) is equivalent to the full configuration-interaction method\cite{VarCC_2013}.

    (BCH-expansion of $\bar{H}$)!

    It is important to note that since $\hat{T} \neq \hat{T}^\dagger$, the CC-energy is non-variational, i.e 
    \begin{equation}
        E_{CC} = \braket{\Phi|e^{-T}\hat{H}e^{T}|\Phi} \neq \braket{\Phi|e^{T^\dagger}\hat{H}e^{T}|\Phi} 
    \end{equation}
    and is not guaranteed to give an upper bound on the energy when $\hat{T}$ is truncated. Working directly with the expression $e^{T^\dagger}\hat{H}e^{T}$ is impractical since it does not truncate after four nested commutators as opposed to $\bar{H}$.

    Since CC is non-variational the question of how to compute expectation values other than the energy arises. The usual definiton of the expectation value of an operator, A, is given by 
    \begin{equation}
        \braket{A} = \braket{\Psi|\hat{A}|\Psi}.
    \end{equation}
    However, we have seen that $\ket{\Psi}^\dagger \neq \bra{\Phi}e^{-\hat{T}}$ such that $E_{CC} \neq \braket{H}$.

    The solution to this problem was introduced independently by workers in quantum chemistry and in theoretical physics\cite{VarCC_2013}.

    The idea is to introduce independent paramterizations of the ket- and bra state, $\bra{\tilde{\Psi}} \neq \ket{\Psi}^\dagger$. [Work in progress...]

    \subsubsection{The time dependent bivariational principle}
    (ASSUMING THE PREVIOUS SECTION IN PLACE)!

    The extension of CC-theory to the time domain can be done through a generalization of the time dependent variational principle\cite{BECK2001}, to a corresponding time dependent bivariational principle\cite{OATDCC_2012}.

    Consider the action-like functional 
    \begin{align}
    \mathcal{S}[\bra{\tilde{\Psi}}, \ket{\Psi}] &= \int_0^T \braket{\tilde{\Psi}(t)|i\frac{d}{dt}-\hat{H}(t)|\Psi(t)} dt  \nonumber \\ 
    &= \int_0^T \braket{\tilde{\Phi}(t)|(\mathbb{1}+\hat{\Lambda}(t))e^{-\hat{T}(t)}\left(i\frac{d}{dt}-\hat{H}(t)\right) e^{\hat{T}(t)}|\Phi(t)} dt \nonumber 
    \end{align}
    where the Hamiltonian generally is time dependent. The integrand can be written more compactly as 
    \begin{align}
        &\braket{\tilde{\Phi}(t)|(\mathbb{1}+\hat{\Lambda}(t))e^{-\hat{T}(t)}\left(i\frac{d}{dt}-\hat{H}(t)\right) e^{\hat{T}(t)}|\Phi(t)} \nonumber \\ 
        = \ \ & i \lambda_\mu \dot{\tau}^\mu + \rho^q_p \left( h^p_q - i \eta^p_q \right) + \frac{1}{4}u^{pr}_{qs}\rho^{qs}_{pr}  \\
        =\ \ & i \lambda_\mu \dot{\tau}^\mu - \mathcal{E}_{\hat{H}-iD_0}[\lambda, \tau] 
    \end{align}
    where we have defined 
    \begin{align}
        D_0 &= \braket{\tilde{\varphi}_p|\dot{\varphi}_q} c_p^\dagger \tilde{c}_q \\
        \eta^p_q &= \braket{\tilde{\varphi}_p|\dot{\varphi}_q}.
    \end{align}

    Thus, 
    \begin{align}
        \mathcal{S}[\bra{\tilde{\Psi}}, \ket{\Psi}] &= \int_0^T i \lambda_\mu \dot{\tau}^\mu + \rho^q_p \left( h^p_q - i \eta^p_q \right) + \frac{1}{4}u^{pr}_{qs}\rho^{qs}_{pr} dt \\
        &= \int_0^T i \lambda_\mu \dot{\tau}^\mu - \mathcal{E}_{\hat{H}-iD_0}[\lambda, \tau] dt
    \end{align}

    The time dependent bivariational principle states that equations of motion are obtained by demanding that $\mathcal{S}$ is stationary ($\delta S = 0$) with respect to all independent variations of the parameters. Genraly both amplitudes and single-particle function/orbitals can be varied. 
    \subsubsection{Time dependent amplitude equations}
    The equations of motions for the amplitudes read 
    \begin{align}
    i \dot{\tau}^\mu &= \frac{\partial}{\partial \lambda_\mu} \mathcal{E}_{\hat{H}-iD_0}[\lambda, \tau] \nonumber \\ 
    &=  \braket{\tilde{\Phi}_\mu|e^{-\hat{T}}(\hat{H}-iD_0)e^{\hat{T}}|\Phi} \label{OATDCC_tau_equations} \\
-i \dot{\lambda}_\mu &= \frac{\partial}{\partial \tau^\mu} \mathcal{E}_{\hat{H}-iD_0}[\lambda, \tau] \nonumber \\ 
&=  \braket{\tilde{\Phi}|(1+\Lambda)e^{-\hat{T}}[\hat{H}-iD_0,X_\mu]e^{\hat{T}}|\Phi} \label{OATDCC_lambda_equations}
\end{align}

The right-hand sides of equations [\ref{OATDCC_tau_equations}, \ref{OATDCC_lambda_equations}] are the same amplitude equations as in time independent cc-theory, plus an extra term due to the time dependence of the orbitals.

If the orbitals are chosen to be stationary\cite{Symplectic_TDCC_2018}, we have $D_0 = 0$ and the time dependence is determined by the amplitudes. This implies a static reference such as the Hartree-Fock determinant. In that case, we also have that $\bra{\tilde{\Phi}} = \bra{\Phi} $, i.e an orthonormal basis.

\subsubsection{Time dependent orbital equations}
Computing the variations of $\mathcal{S}$ with respect to the orbitals, we obtain equations of motion for the time-dependent orbitals. The time derivatives of the orbitals can, in general, be split into a P-space and a Q-space, 
\begin{align}
    \ket{\dot{\varphi}_q} &= (P+Q)\ket{\dot{\varphi}_q} = \sum_p \eta^p_q \ket{\varphi_p} + Q\ket{\dot{\varphi}_q} \\
    \bra{\dot{\tilde{\varphi}}_q} &= \bra{\dot{\tilde{\varphi}}_q}(P+Q) = -\sum_q \eta^p_q \bra{\tilde{\varphi}_q} + \bra{\dot{\tilde{\varphi}}_p}Q
\end{align}
where we have defined 
\begin{align}
    P &= \sum_{p=1}^M \ket{\varphi_p}\bra{\tilde{\varphi}_p},  \ \ M \leq N_b \\
    Q &= \mathbb{1}-P
\end{align}
with $N_b$ denoting the number of single-particle functions.

This splitting is identical to what is done in MCTDHF-theory \cite{Hochstuhl2014}, where a trunctation of the $P$-space is used to reduce the number of Slater determinants the expansion of the CI-wavefunction.

In coupled cluster methods the number of Slater determinants is reduced through the truncation of the cluster operators. Thus, we can choose to have $M=N_b$, resulting in $Q=0$. In the following we restrict our attention to the case where $Q=0$.

Note that for the MCTDHF method this would be the same as FCI with time dependent orbitals and nothing is gained in terms of reducing computational complexity.

With $Q=0$ the equations of motion for the orbitals is given by
\begin{align}
    \ket{\dot{\varphi}_q} &=  \sum_p \eta^p_q \ket{\varphi_p} \label{Reduced ket-orbital equation} \\
    \bra{\dot{\tilde{\varphi}}_q} &=  -\sum_q \eta^p_q \bra{\tilde{\varphi}_q} \label{Reduced bra-orbital equation}. 
\end{align}
The non-zero components of $\eta^p_q$ are given by the solutions of the linear equations
\begin{align}
    i \sum_{bj} A^{ib}_{aj} \eta^j_b &= \sum_j \rho^i_j h^j_a - \sum_b \rho^b_a h^i_b \nonumber \\
 + &\frac{1}{2} \left[\sum_{prs} \rho^{is}_{pr} u^{pr}_{as} - \sum_{rqs} \rho^{qs}_{ar} u^{ir}_{qs}  \right] \\
    -i \sum_{bj} A^{ja}_{bi} \eta^b_j &= \sum_b \rho^a_b h^b_i - \sum_j \rho^j_i h^a_j \nonumber \\
 + &\frac{1}{2} \left[ \sum_{prs} \rho^{as}_{pr} u^{pr}_{is} - \sum_{rqs} \rho^{qs}_{ir} u^{ar}_{qs} \right] \nonumber \\ 
 + &i \dot{\rho}_i^a \label{P-space bra equation}
\end{align}
where
\begin{align}
    A^{ib}_{aj} &= \braket{\tilde{\Psi}|[c_j^\dagger c_b,c_a^\dagger c_i]|\Psi} \nonumber \\
    &= \delta^b_a \rho^i_j - \delta^i_j \rho^b_a \nonumber  \\ 
    &= -A^{bi}_{ja}.
\end{align}

The simplest OATDCC case is the OATDCCD approximation where the cluster operators include double excitations only. One can show that the singles amplitudes are formally redundant \cite{OATDCC_2012}, leaving the one-body density matrix block diagonal, with only $\rho^i_j$ and $\rho^a_b$ being non-zero. Thus, the last term in eq. (\ref{P-space bra equation}) is zero.

Furthermore, there is no contributions to the amplitude equations (\ref{OATDCC_tau_equations}) and (\ref{OATDCC_lambda_equations}) from $D_0$ in the doubles-only approximation. This follows from that only $\eta^a_i$ and $\eta^i_a$ are non-zero.

\section{Systems}
    Relevant articles:
    \begin{itemize}
        \item Simple spin-dependent field TDHF study:\cite{TDHF_Singlet_Triplet}
        \item One dimensional time-dependent quantum dot: \cite{Zanghellini04}
        \item Harmonic single well: \cite{HjorthJensenQD_2017}
        \item Theoretical study double well: \cite{Double_Well_2000}
        \item Experiment with two fermions in double well: \cite{Fermions_Double_Well_Experiment}
        \item Coupled double well: \cite{Coupled_double_wells}
        \item Time domain two electrons in quantum dot: \cite{TimeDomainTwoElectronsQuantumDot}
        \item Time-dependency with spin-orbit coupling: \cite{Spin_orbit_coupling}
        \item Time-domain simulation of quantum spin: \cite{Time_domain_quantum_spin}
        \item Entanglement in spin-qubit quantum dots: \cite{Entanglement_spin_qubit_quantum_dots}
    \end{itemize}

    Generic information on quantum dots and why we are interested in them.

    General Hamiltonian for interacting electrons in a (time-dependent) confining potential influenced by an external electric field (laser),
    \begin{align}
    \hat{H}(t) = &\sum_{i=1}^N \left( - \frac{1}{2}\nabla^2 + \vconf(\mathbf{r}_i,t) + \vfield(\mathbf{r}_i,t) \right) \nonumber \\ 
    + &\sum_{i < j}^N \frac{1}{r_{ij}}.
    \label{eq:general_hamiltonian}
    \end{align}

    \subsection{One dimensional quantum dot}
        We will use the one-dimensional quantum dot model considered by Zanghellini et. al in \cite{Zanghellini04}, as a benchmark/validation for our implementations.

    \subsection{Two dimensional quantum dot}
        For a two dimensional quantum dot model we will consider different confining potential. There already exists studies concerning the ground state properties the electrons are confined by the harmonic oscillator potential, see for example \cite{HjorthJensenQD_2017}. We will use this as a starting point for time-dependent studies.
    \subsubsection{Single well}
        The Hamiltonian of the two-dimensional harmonic oscillator quantum dot is given by \autoref{eq:general_hamiltonian}, with the confining potential
        \begin{align}
            \vconf(\boldsymbol{r}_i) \equiv \frac{1}{2} \omega^2 \boldsymbol{r}_i^2.
        \end{align}
    \subsubsection{Double well}
    \subsection{Open system/finite potential wells}
        Maybe too ambitious in this work.

\section{Time-dependent fields}
    \subsection{Dipole laser field}
        The dipole laser field we will be looking at is given by
        \begin{align}
            \vfield(\boldsymbol{r}_i, t)
            &= \envelope(t) \cdot \boldsymbol{r}_i,
        \end{align}
        where the \emph{laser envelope} $\envelope(t)$ is given by a time-varying field, e.g., purely sinusoidal with a constant amplitude as in the article by Zanghellini et. al \cite{Zanghellini04}, and with a polarization direction $\polarization$.
        The dipole-part of the name of the laser stems from the position vector $\vec{r}_i$.
    \subsection{Magnetic field}
        Figure out how to model this field.

        Hamiltonian for a single particle in a magnetic field \cite{TimeDomainTwoElectronsQuantumDot}
\begin{equation}
    \hat{H} = \frac{1}{2m} \left( \frac{\hbar}{i} \nabla - q \vec{A} \right)^2 + V(\vec{r})
\end{equation}
For a uniform magnetic field in the z-direction, we have 
\begin{equation}
    \vec{A} = \frac{B}{2} (-y,x,0) = B_0 (-y,x,0)
\end{equation}
Thus, 
\begin{equation}
    \hat{H} \psi(x,y) = \left(-\frac{\hbar^2}{2m} \nabla^2  - \frac{2q}{2m}\vec{A} \cdot \nabla  + \frac{q^2}{2m} A^2 + V(x,y) \right) \psi(x,y).
\end{equation}
Choosing atomic units ($\hbar = q = m = 1$), we have
\begin{equation}
    \hat{H} \psi(x,y) = \left(-\frac{1}{2} \nabla^2  -\vec{A} \cdot \nabla  + \frac{1}{2}A^2 + V(x,y) \right) \psi(x,y).
\end{equation}

\section{Results \& Discussion}

    \subsection{One-dimensional quantum dot benchmark}
        Comparison of the varying methods with the results and figures from Zanghellini et. al\cite{Zanghellini04}.

        \begin{figure}
        \includegraphics[width=0.5\textwidth]{./figures/zanghellini_fig2.eps}
        \caption{\label{fig:1DZanghellini} Probability of being in the ground state
        $|\langle\Psi(x,0\ket{\Psi(x,t)}|^2$ for one-dimensional quantum dot with frequency
        $\omega = 8\Omega$ and two electrons. This plot is an exact match with figure 2 in Zanghellini
        et al.\cite{Zanghellini04}
        }
        \end{figure}

    \subsection{Two-dimensional quantum dot calculations}
        Results from different interesting time-dependent observable quantities for distinct confining potentials and time-dependent fields (maybe too ambitious on different fields).

\section{Concluding remarks}
\begin{appendices}
\section{Derivation of TDHF equations} \label{TDHF_Derivation}
By introducing functional derivatives and their action on the matrix elements (see Hochstuhl et al \cite{Hochstuhl2014}),
\begin{align}
    \frac{\delta}{\delta \phi^*_k}\bra{\phi_i}\hat{h}\ket{\phi_i} &= \hat{h}\ket{\phi_k} \\
    \frac{\delta}{\delta \phi^*_k}\bra{\phi_i}\frac{\partial}{\partial t} \ket{\phi_i} 
    &= \frac{\partial}{\partial t}\ket{\phi_k} \\
    \frac{\delta}{\delta \phi^*_k}\bra{\phi_i \phi_j} \hat{u} \ket{\phi_i, \phi_j}_{\text{AS}}
    &= 2\bra{\cdot \phi_i} \hat{u} \ket{\phi_k \phi_i}_{\text{AS}},
\end{align}
one can find the equations of motions for the system by way of the time-dependent variational principle,
\begin{equation}
    \label{td_variational_principle}
    \bra{\delta \Phi} \hat{H} - i\hbar \frac{\partial}{\partial t} \ket{\Phi} = 0.
\end{equation}

(INSERT SHORT NOTE ON EMPTY SLOT NOTATION?)

We are faced with an optimisation problem under the constraint that the orbitals are orthonormal ($\braket{\phi_i(t)}{\phi_j(t)} = \delta_{ij})$, yielding the following Lagrange functional,
\begin{equation}
    \mathcal{L} = \bra{\Phi} \hat{H} - i\hbar \frac{\partial}{\partial t} \ket{\Phi} 
    - \lambda_{ij}(\braket{\phi_i(t)}{\phi_j(t)} - \delta_{ij}),
\end{equation}
with an optimal state $\ket{\phi_k}$ satisfying
\begin{equation}
    \label{HF_optimal_phi}
    \hat{h}\ket{\phi_k} + \bra{\cdot \phi_j} \hat{u} \ket{\phi_k \phi_j}_{\text{AS}} 
    - i\hbar\frac{\partial}{\partial t} \ket{\phi_k} - \lambda_{kj} \ket{\phi_j} = 0.
\end{equation}
Left-projecting this expression onto a state $\bra{\phi_l}$ and solving for the lagrange multiplier $\lambda$ gives
\begin{equation}
    \lambda_{kl} = \bra{\phi_l} \hat{h} \ket{\phi_k} 
    + \bra{\phi_l \phi_j} \hat{u} \ket{\phi_k \phi_j}_{\text{AS}}
    - i\hbar \bra{\phi_l} \frac{\partial}{\partial t} \ket{\phi_k}.
\end{equation}
Inserting this expression into \autoref{HF_optimal_phi} yields
\begin{equation}
    i\hbar \hat{P}\frac{\partial}{\partial t} \ket{\phi_k} = \hat{P} \hat{f} \ket{\phi_k},
\end{equation}
where we have defined a projection opertor $\hat{P} = \mathbb{1} - \ket{\phi_i}\bra{\phi_i}$ and the Fock operator $\hat{f} \equiv \hat{h} + \bra{\cdot \phi_j} \hat{u} \ket{\cdot \phi_j}_{\text{AS}}$.

For unitary time-dependent operator given by $\hat{Q}(t) = i\hbar (\partial / \partial t)$,
we have
\begin{equation}
    \label{TDHF}
    i \hbar \frac{\partial}{\partial t} \ket{\phi_k(t)} = \hat{f} \ket{\phi_k(t)},
\end{equation}
if we choose $\hat{Q}(t) = \hat{f}(t)$. \autoref{TDHF} is the time-dependent Hartree-Fock equations, which can be rewritten on matrix form by expanding the molecular slater wavefunctions in an atomic basis with time-dependent coefficients and left projecting with an arbitrary atomic basis function,
\begin{align}
    i \hbar \frac{\partial}{\partial t} C_{\alpha i} (t) \ket{\chi_\alpha} &= \hat{f}(t) C_{\alpha i} (t) \ket{\chi_\alpha} \\
    \to i \hbar \frac{\partial}{\partial t} C_{\alpha i} (t) \langle{\chi_{\beta}}|{\chi_{\alpha}}\rangle &= C_{\alpha i}(t) \bra{\chi_\beta} \hat{f}(t) \ket{\chi_\alpha} \\
    \to i \hbar \mathbf{S}\dot{\mathbf{C}}(t) &= \mathbf{F}(t) \mathbf{C}(t).
\end{align}
\section{The time dependent energy}
In general the time dependent energy is given by 
\begin{equation}
    E(t) = \braket{\Psi(t)|\hat{H}(t)|\Psi(t)}.
\end{equation}
If $\hat{H}$ does not depend explicitly on time, we find that the energy is conserved
\begin{align*}
     \frac{\partial}{\partial t}E(t) &= \braket{\frac{\partial \Psi(t)}{\partial t}|\hat{H}|\Psi(t)} + \braket{\Psi(t)|\hat{H}|\frac{\partial \Psi(t)}{\partial t}} \\
     &= \braket{\Psi(t)|\hat{H}|\frac{\partial \Psi(t)}{\partial t}}^* + \braket{\Psi(t)|\hat{H}|\frac{\partial \Psi(t)}{\partial t}} \\
     &= (-i\braket{\Psi(t)|\hat{H}^2|\Psi(t)})^* -i\braket{\Psi(t)|\hat{H}^2|\Psi(t)} \\
     &= 0
\end{align*}
where we used the time dependent Schrödinger equation 
\begin{equation}
    i \frac{\partial \Psi(t)}{\partial t}  = \hat{H} \Psi(t)
\end{equation}
to obtain the last equalities.
\section{Semi-analytical solutions for two electrons confined by a harmonic oscillator potential}
We review the (semi)-analytical solution(s) of two electrons confined by a harmonic oscillator potential as discussed by Taut \cite{Taut93,Taut94}, in two- and three spatial dimensions. We also consider the analogous one dimensional problem.

Consider the Hamiltonian for two interacting particles in a harmonic oscillator potential, 
\begin{equation}
    \hat{H} = \sum_{i=1}^2 \left(\frac{1}{2} \mathbf{p}_i^2 + \frac{1}{2} \Omega^2 \mathbf{r}_i^2\right) + v_{\text{int}}(\mathbf{r}_1,\mathbf{r}_2). 
\end{equation}
In one spatial dimension the interaction between the electrons is modelled by a smoothed Coulomb potential \cite{SmoothedCoulomb} 
\begin{equation}
    v_{\text{int}}(x_1,x_2) = \frac{1}{\sqrt{(x_1-x_2)^2 + a^2}}
\end{equation}
where $a$ is a shielding parameter. In two- or three spatial dimensions the interaction is given by the usual Coulomb potential.

Following Taut \cite{Taut93} we introduce the difference vector and the center of mass as new variables,
\begin{align}
    \mathbf{r} &= \mathbf{r}_2-\mathbf{r}_1 \\
    \mathbf{R} &= \frac{1}{2} \left( \mathbf{r}_1+\mathbf{r}_2 \right). 
\end{align}
The key observation is that the Hamiltonian decouples in these variables
\begin{align}
    H &= - \nabla_\mathbf{r}^2 + \frac{1}{4} \Omega^2 \mathbf{r}^2 + v_{\text{int}}(\mathbf{r})  - \frac{1}{4}\nabla_\mathbf{R}^2 + \Omega^2 \mathbf{R}^2 \nonumber \\
    &= 2 \left( -\frac{1}{2} \nabla_\mathbf{r}^2 + \frac{1}{2} \Omega_\mathbf{r}^2 \mathbf{r}^2 + \frac{1}{2}v_{\text{int}}(\mathbf{r}) \right) \nonumber \\  &+ \frac{1}{2} \left( -\frac{1}{2}\nabla_\mathbf{R}^2 + \frac{1}{2}\Omega_\mathbf{R}^2 \mathbf{R}^2 \right) \nonumber \\
    &= 2 H_\mathbf{r} + \frac{1}{2}H_\mathbf{R} 
\end{align}
where we have defined 
\begin{align}
    \Omega_\mathbf{r} &\equiv \frac{1}{2}\Omega \\
    \Omega_\mathbf{R} &\equiv 2 \Omega \\
    H_\mathbf{r} &\equiv -\frac{1}{2} \nabla_\mathbf{r}^2   + \frac{1}{2} \Omega_\mathbf{r}^2 \mathbf{r}^2 + \frac{1}{2}v_{\text{int}}(\mathbf{r}) \\
    H_\mathbf{R} &\equiv -\frac{1}{2}\nabla_\mathbf{R}^2 + \frac{1}{2}\Omega_\mathbf{R}^2 \mathbf{R}^2
\end{align}
Since $H$ is independent of spin, the total wave function can be factorized as follows\cite{Taut93} 
\begin{equation}
    \psi(\mathbf{r},\mathbf{R},\mathbf{s}) = \varphi(\mathbf{r}) \xi(\mathbf{R}) \chi(s_1,s_2).
\end{equation}
Thus, the time-independent Schrödinger equation 
\begin{align}
    H\psi(\mathbf{r},\mathbf{R},\mathbf{s}) &= \left( 2 H_\mathbf{r} + \frac{1}{2}H_\mathbf{R} \right) \varphi(\mathbf{r}) \xi(\mathbf{R}) \chi(s_1,s_2) \\ 
    &= E\psi(\mathbf{r},\mathbf{R},\mathbf{s}) \\
    &= \left( E_\mathbf{r} + E_\mathbf{R} \right) \varphi(\mathbf{r}) \xi(\mathbf{R}) \chi(s_1,s_2)
\end{align}
separates into one equation for the relative motion
\begin{equation}
    \left[- \frac{1}{2}\nabla_\mathbf{r}^2 + \frac{1}{2} \omega_r^2 \mathbf{r}^2 + \frac{1}{2}v_{\text{int}}(\mathbf{r}) \right] \varphi(\mathbf{r}) = E^\prime_\mathbf{r} \varphi(\mathbf{r}) \label{RelativeMotionEquation}
\end{equation}
where $E^\prime_\mathbf{r} = \frac{1}{2}E_\mathbf{r}$, and one for the center of mass 
\begin{equation}
    \left[- \frac{1}{2}\nabla_\mathbf{R}^2 + \frac{1}{2}\Omega_R^2 \mathbf{R}^2 \right] = E^\prime_\mathbf{R} \xi(\mathbf{R}) \label{CenterOfMassEquation}
\end{equation}
where $E^\prime_\mathbf{R} = 2 E_\mathbf{R}$. The total energy is then given by 
\begin{equation}
    E = E_\mathbf{r} + E_\mathbf{R} = 2E^\prime_\mathbf{r} + \frac{1}{2}E^\prime_\mathbf{R}.
\end{equation}

We recognize equation (\ref{CenterOfMassEquation}) as the eigenvalue problem for a (d-dimensional) harmonic oscillator with frequency $\Omega_R$.Thus, 
\begin{equation}
    E_\mathbf{R}^\prime = \Omega_\mathbf{R} \left( n_d + \frac{d}{2} \right)
\end{equation}
for $d=1,2,3$.

In one dimension we can solve the eigenvalue problem for the relative motion (\ref{RelativeMotionEquation}) numerically, with the smoothed Coulomb potential, using for example a finite difference discretization.

In the next section we discuss how to solve the equation for the relative motion i two spatial dimensions.

\section{Solving the two-dimensional Schrödinger equation for a central symmetric potential}
Consider the Schrödinger equation in two dimensions for a central symmetric potential where we use polar coordinates 
\begin{equation}
    \left[ - \frac{1}{2} \nabla^2_{r,\theta} + V(r) \right] \psi(r,\theta) = E\psi(r,\theta). \label{SEPolar}
\end{equation}
The Laplacian in polar coordinates is given by
\begin{equation}
    \nabla_{r,\theta}^2 = \frac{1}{r} \frac{\partial}{\partial r} \left( r \frac{\partial}{\partial r} \right) + \frac{1}{r^2} \frac{\partial^2}{\partial \theta^2}
\end{equation}

Inserting the expressions for the Laplacian into equation (\ref{SEPolar}) we have that
\begin{equation}
    \left[ - \frac{1}{2} \left(\frac{1}{r} \frac{\partial}{\partial r} \left( r \frac{\partial}{\partial r} \right) + \frac{1}{r^2} \frac{\partial^2}{\partial \theta^2} \right) + V(r) \right] \psi(r,\theta) = E\psi(r,\theta).
\end{equation}

We look for solutions that are seprable on the form
\begin{equation}
    \psi(r,\theta) = R(r) \varphi(\theta)
\end{equation}
such that
\begin{equation*}
     - \frac{1}{2} \left(\frac{\varphi}{r} \frac{\partial}{\partial r} \left( r \frac{\partial R}{\partial r} \right) + \frac{R}{r^2} \frac{\partial^2 \varphi}{\partial \theta^2} \right) + V(r)R\varphi   = ER\varphi.
\end{equation*}

Multiplying with $-2r^2$ and dividing with $R\varphi$ we obtain
\begin{equation*}
      \left[\frac{r}{R}\frac{\partial}{\partial r} \left( r \frac{\partial R}{\partial r} \right) -2r^2 (V(r)-E) \right] + \frac{1}{\varphi} \frac{\partial^2 \varphi}{\partial \theta^2}     = 0,
\end{equation*}
which implies that 
\begin{equation}
    \frac{r}{R}\frac{\partial}{\partial r} \left( r \frac{\partial R}{\partial r} \right) -2r^2 (V(r)-E) = m^2 \end{equation}
    and
    \begin{equation}
    \frac{1}{\varphi} \frac{\partial^2 \varphi}{\partial \theta^2} = -m^2.
\end{equation}

\subsection{The angular equation}
The second (angular) equation is satisfied by 
\begin{equation}
    \varphi = Ne^{im\theta}
\end{equation}
with $N$ a normalization constant. Requiring 
\begin{equation}
    \varphi(\theta + 2\pi) = \varphi(\theta)
\end{equation}
implies that $m=0,\pm 1, \pm 2,\cdots$. The normalization condition 
\begin{equation}
\int_0^{2\pi} |\varphi|^2 d\theta = 2\pi|N|^2 = 1    
\end{equation} gives $N = 1/\sqrt{2\pi}$.

\subsection{The radial equation}
Next, we consider the radial equation
\begin{equation}
    r\frac{\partial}{\partial r} \left( r \frac{\partial R}{\partial r} \right) -2r^2 (V(r)-E)R = Rm^2 \label{2d-RadialEquation}
    \end{equation}
This equation simplifies if we change variables
\begin{equation}
    R(r) = \frac{u(r)}{r^{1/2}}
\end{equation}
and note that 
\begin{equation}
    r\frac{\partial}{\partial r} \left( r \frac{\partial R}{\partial r} \right) = r^{-1/2} \left[ r^2 \frac{d^2 u}{dr} + \frac{1}{4}u(r) \right].
\end{equation}
Insertion in equation (\ref{2d-RadialEquation}) yields
\begin{equation}
    \left[ -\frac{1}{2} \frac{d^2}{dr^2} + \frac{1}{2r^2} \left( m^2 - \frac{1}{4} \right) + V(r) \right] u(r) = Eu(r) \label{2d-radial-new}.
\end{equation}

For two interacting particles confined by a harmonic oscillator potential the potential is given by 
\begin{equation}
    V(r) = \frac{1}{2}\Omega_r^2 + \frac{1}{2r},
\end{equation}
for which equation (\ref{2d-radial-new}) can be solved analytically for particular oscillator frequencies\cite{Taut94}.

One can also solve the equation numerically for arbitrary oscillator frequencies. The main challenge in a numerical approach is how to deal with the singularities at $r=0$.

\section{Semi-analytical solution of a laser-driven two electron system}
Following Schwengelbeck \cite{SCHWENGELBECK_1999} we consider the problem of two interacting electrons, confined by a harmonic oscillator potential, subject to a time-dependent sinusoidal laser-field. The Hamiltonian of the system is given by 
\begin{equation}
    \hat{H} = \sum_{i=1}^2 \left(\frac{1}{2} \mathbf{p}_i^2 + \frac{1}{2} \Omega^2 r_i^2 + \mathbf{r}_i \cdot \hat{\mathbf{e}} E_0 \sin(\omega t)  \right) + v_{\text{int}}(\mathbf{r}_1,\mathbf{r}_2). 
\end{equation}
where $\omega$ and $E_0$ are the frequency and amplitude of the field and $\hat{\mathbf{e}}$ denotes a unit polarization vector.

As in the previous section, introducting the relative distance and center of mass as new variables, the Hamiltonian decouples
\begin{align}
    H &= 2 \underbrace{\left( -\frac{1}{2} \nabla_\mathbf{r}^2 + \frac{1}{2} \Omega_\mathbf{r}^2 \mathbf{r}^2 + \frac{1}{2}v_{\text{int}}(\mathbf{r}) \right)}_{\equiv H_\mathbf{r}} \nonumber \\  &+ \frac{1}{2} \underbrace{\left( -\frac{1}{2}\nabla_\mathbf{R}^2 + \frac{1}{2}\Omega_\mathbf{R}^2 \mathbf{R}^2 + 4\mathbf{R} \cdot \hat{\mathbf{e}} E_0 \sin(\omega t) \right)}_{\equiv H_\mathbf{R}}
\end{align}
and the time-dependent Schrödringer equation admits solutions of the factorized form\cite{SCHWENGELBECK_1999}
\begin{equation}
    \Psi(\mathbf{r},\mathbf{R},\mathbf{s},t) = \psi(\mathbf{R},t) \varphi(\mathbf{r},t) \chi(s_1,s_2).
\end{equation}
We use the ground state of the unperturbed system as initial condition
\begin{equation}
    \Psi(\mathbf{r},\mathbf{R},\mathbf{s},0) = \psi_0(\mathbf{R},0) \varphi_0(\mathbf{r},0) \chi(s_1,s_2),
\end{equation}
where $\varphi_0(\mathbf{r})$ and $\psi_0(\mathbf{R})$ are the ground state solutions of equations (\ref{RelativeMotionEquation}) and (\ref{CenterOfMassEquation}) respectively.

In particular, we get one equation for the relative motion 
\begin{equation}
    i \frac{\partial}{\partial t} \varphi(\mathbf{r},t) = 2H_\mathbf{r} \varphi(\mathbf{r},t)
\end{equation}
and one for the center of mass
\begin{equation}
    i \frac{\partial}{\partial t} \psi(\mathbf{R},t) = \frac{1}{2}H_\mathbf{R} \psi(\mathbf{R},t) \label{time-dependent CM eq}.
\end{equation}
Since $H_\mathbf{r}$ contains no explicit time-dependent terms, the equation for the relative motion is simply given by 
\begin{equation}
    \varphi(\mathbf{r},t) = \varphi_0(\mathbf{r},0) \exp(-i E^\prime_{0,\mathbf{r}} t)
\end{equation}
where $\{(\phi_0(\mathbf{r},0),E^\prime_{0,\mathbf{r}}\}$ is given by the lowest lying solution of equation (\ref{RelativeMotionEquation}).

\subsection{Solving the center of mass equation in time}
Eq. (\ref{time-dependent CM eq}) associated with the center of mass motion of the two electrons, can be solved analytically by means of a path integral\cite{SCHWENGELBECK_1999} or we can use a numerical approach.

\subsubsection{One spatial dimension}
Consider first the one dimensional problem, $(\mathbf{r},\mathbf{R}) = (x,X)$ with polarization vector $\hat{\mathbf{e}}$ along the positive x-direction 
\begin{align}
    i \frac{\partial \psi(X,t)}{\partial t} &= \frac{1}{2} H_X(t) \psi(X,t) \label{OneDimTimeDepEq}
\end{align}
where 
\begin{equation}
    H_X(t) =  -\frac{1}{2}\frac{d^2}{dX^2} + \frac{1}{2}\Omega_X^2 X^2 + 4X \sin(\omega t) 
\end{equation}
and $\Omega_X \equiv 2\Omega$. We use the ground state of the unperturbed harmonic oscillator as initial state for center of mass wavefunction 
\begin{equation}
    \psi(X,0) = \left( \frac{\Omega_X}{\pi} \right)^{1/4} \exp\left(-\frac{\Omega_X X^2}{2}\right).
\end{equation}
Equation (\ref{OneDimTimeDepEq}) is solved by using the standard Runge-Kutta fourth order (RK4) scheme.

Thus, we have that the total wavefunction is given by 
\begin{align}
 \Psi(x,X,t) &= \varphi(x,t) \psi(X,t) \nonumber \\
 &= \varphi_0(x,0)e^{-tiE_{0,x}^\prime} \psi(X,t).
\end{align}
where $E_{0,x}^\prime$ is the lowest eigenvalue of for the relative coordinate equation (\ref{RelativeMotionEquation}).

We will use this solution to benchmark the more general many-body solvers. In particular, we can compute the expectation value 
of the time-dependent Hamiltonian
\begin{align}
 \braket{\Psi|\hat{H}|\Psi} &= \braket{\Psi|H_x|\Psi} + \braket{\Psi|H_X(t)|\Psi} \nonumber \\
 &= E_{0,x}^\prime + \braket{\Psi|H_X(t)|\Psi}.
\end{align}

\subsubsection{Two spatial dimensions}
The two dimensional problem is almost a carbon copy of the one dimensional problem. We have that $\mathbf{R} = (X,Y)$ and we choose the polarization vector $\hat{\mathbf{e}}$ along the positive x-direction
\begin{align*}
    i \frac{\partial \psi(\mathbf{R},t)}{\partial t} &= \left( -\frac{1}{4}\nabla_\mathbf{R}^2 + \Omega^2 \mathbf{R}^2 + 2\mathbf{R} \cdot \hat{\mathbf{e}} \sin(\omega t) \right) \psi(\mathbf{R},t) \\
    &= \frac{1}{2}\underbrace{\left( -\frac{1}{2}\nabla_\mathbf{R}^2 + \frac{1}{2}\Omega_R^2 \mathbf{R}^2 + 2X \sin(\omega t) \right)}_{\equiv H_\mathbf{R}} \psi(\mathbf{R},t)
\end{align*}
where we again have that $\Omega_R = 2\Omega$.

Since we have chosen the polarization axis along the positive x-direction (or y-direction) the Hamiltonian separates
\begin{align*}
    H_\mathbf{R} &= \frac{1}{2}\left( -\frac{1}{2}\frac{\partial^2}{\partial X^2} + \frac{1}{2}\Omega_R^2 \mathbf{X}^2 + 2X \sin(\omega t) -\frac{1}{2}\frac{\partial^2}{\partial Y^2} + \frac{1}{2}\Omega_R^2 \mathbf{Y}^2 \right) \\
    &= \frac{1}{2}(H_X(t) + H_Y).
\end{align*}
Note that $H_Y$ does not contain any explicit time-dependent terms.

Then, the time-dependent Schrödinger equation for the center of mass motion admits solutions of the factorized form
\begin{equation}
    \psi(\mathbf{R},t) = \psi(X,Y,t) = \psi_X(X,t) \psi_Y(Y,t).
\end{equation}
We use the ground state of the unperturbed two-dimensional harmonic oscillator as intial condition
\begin{align}
    \psi(X,Y,0) = &\left( \frac{\Omega_R}{\pi} \right)^{1/4} \exp\left(-\frac{\Omega_R X^2}{2}\right) \nonumber \\ 
    \times &\left( \frac{\Omega_R}{\pi} \right)^{1/4} \exp\left(-\frac{\Omega_R Y^2}{2}\right)
\end{align}

Insertion of this ansatz into the time-dependent Schrödinger equation yields two independent equations 
\begin{align}
    i \frac{\partial \psi_X}{\partial t} &= \frac{1}{2} H_X(t) \psi_X \\
    i \frac{\partial \psi_Y}{\partial t} &= \frac{1}{2} H_Y \psi_Y. 
\end{align}
The first equation is now formally equivalent to the one-dimensional problem.

The second equation has the solution of a unperturbed one-dimensional harmonic oscillator 
\begin{equation}
    \psi_Y(Y,t) = \psi(y) e^{-i \epsilon_{0,Y} t}
\end{equation}
where \begin{equation}
    \epsilon_{0,y} = \frac{1}{2}\eta_Y
\end{equation} 
 and 
\begin{equation}
    \eta_Y = \frac{1}{2}\Omega_R
\end{equation}
is the ground state energy of a harmonic oscillator with oscillator frequency $\Omega_R$.

Finally, we have that the total wavefunction is given by 
\begin{align}
 \Psi(\mathbf{r},\mathbf{R},t) &= \varphi(\mathbf{r},t) \psi(\mathbf{R},t) \nonumber \\
 &= \varphi(\mathbf{r},0)e^{-i\epsilon t} \psi_X(X,t) \psi_Y(Y,t).
\end{align}
where $\epsilon$ is the lowest eigenvalue of for the relative coordinate equation (\ref{RelativeMotionEquation}) in two spatial dimensions.

In particular, we find that the expectation value of the time-dependent Hamiltonian is given by 
\begin{align}
    \braket{\Psi|\hat{H}|\Psi} &= \braket{\Psi|\hat{H}_\mathbf{r}|\Psi} + \braket{\Psi|H_\mathbf{R}|\Psi} \nonumber \\ 
    &= \epsilon + \braket{\Psi|H_Y|\Psi} + \braket{\Psi|H_X(t)|\Psi} \nonumber \\
    &= \epsilon + \epsilon_{0,y} + \braket{\Psi|H_X(t)|\Psi}.
\end{align}

\section{The harmonic potential theorem}
The harmonic potential theorem: See Dobson\cite{HarmonicPotentialTheorem} and section 6.3.2 in the book by Ullrich\cite{TD_DFT_Ullrich_Book}.

\end{appendices}
\newpage
\bibliographystyle{ieeetr}
\bibliography{refs}
\end{document}

### 2019-04-18 doc/MachineLearning/Codes/tail.f
added codes

@@ -1,0 +1,20 @@
+      subroutine tail(etail)
+      implicit real*8 (A-H,O-Z)
+      parameter (ndim=3)
+      common /const/ pi,hb,hbo4m,delta,deltai,delts,deltsi,sd,
+     +               el2,el,eli,vkey,bkeyu,bkey,bins,np,np2
+      save c6,c8,c10,sigma,rm,eps
+      data c6,c8,c10,sigma,rm,eps
+     +     /1.3732412,.4253785,.1781,2.556,2.9673,10.8/
+      rho=np/(2.*el2)**ndim
+      yfit=rm/(sigma*el2)
+      x2=yfit*yfit
+      if (ndim.eq.2) then
+         vint=(c6/4.+(c8/6.+c10*x2/8.)*x2)*x2*x2
+         etail=-pi*rho*eps*vint*(rm/sigma)**2
+      elseif (ndim.eq.3) then
+         vint=(c6/3.+(c8/5.+c10*x2/7.)*x2)*x2*yfit
+         etail=-2.*pi*rho*eps*vint*(rm/sigma)**3
+      endif
+      return
+      end

### 2019-04-18 doc/MachineLearning/Codes/vmc1.c

@@ -1,0 +1,617 @@
+/*                          
+                               
+ VMC for a quantum Liquid of Lennard-Jones particles 
+ Mass and potential constants are fixed
+*/
+
+#include <math.h>
+#include <stdio.h>
+#include <stdlib.h> /* necessary for the rand() function */
+
+
+const int npmax=500;
+const int nest=5;
+const double KB=1.;  /* Energies in degrees Kelvin, lenghts in Angstrom */
+//const double hbo2m=6.0599278;
+//  hbar^2/2m (depends on mass) 
+const double hbo2m=0.92756799; //<- This value good if lengths in LJ-sigma units 
+
+double delta,rho,sigma_lj,eps_lj,l,li,l2;
+double Pi,acc,vbox,b,tpb,tjf;
+int npart,nstack,ncx,ncy,ncz;
+
+
+struct particle
+{ 
+  double x,y,z;
+};
+
+struct table
+{
+  double point[1000];
+};
+
+
+
+double total_potential(),pair_wf(),pair_pot(),vlj(),vmod();
+double kinetic_energy(),vtail(),rgauss(),u();
+struct table grnow(),addtable();  
+
+struct table addtable(struct table table1, struct table table2)
+{
+  int i;
+  struct table temp;
+  for(i=0;i<nstack;i++)
+    {
+      temp.point[i]=table1.point[i]+table2.point[i];
+    };
+  return temp;
+}
+
+copy_points(struct particle a[npmax],struct particle b[npmax])
+{
+  int i;
+  for (i=0;i<npmax;i++)
+    {
+      b[i].x=a[i].x;
+      b[i].y=a[i].y;
+      b[i].z=a[i].z;
+
+    };
+}
+
+main() 
+{
+  struct particle walker[npmax];    
+  struct table gr,vcorr;
+  double etot,vout;
+  double dr,l22;
+  double dest[nest],estave[nest],estave2[nest];
+  double estsum[nest],estsum2[nest],estnow[nest],estvar[nest];  /* estimators */
+  int idest[nest];
+  double vbulk,accum;
+  double gr_norm,r,norm,rout,rin,tau,cvv;
+  float delta_in,rho_in,b_in;
+  int nstep1,nstep2;
+  int i,j,k;
+  FILE *out;
+
+  out=fopen("gr.out","w+");
+  Pi=4.*atan(1.);
+
+  printf("Quantum Lennard-Jones fluid in 3D: \n");
+
+/******************/
+/* Parameters of the LJ potential. These values are for atomic 4He,
[…]

### 2019-04-18 doc/MachineLearning/Codes/vpot.f

@@ -1,0 +1,27 @@
+      subroutine vpot(r,v,vp,vpp)
+      implicit real*8 (A-H,O-Z)
+      parameter (expmax=666.e0)
+      save a,alp,d,c6,c8,c10,sigma,rm,eps
+      data a,alp,d,c6,c8,c10,sigma,rm,eps /544850.4,13.353384,1.241314,
+     +1.3732412,.4253785,.1781,2.556,2.9673,10.8/
+      units=sigma/rm
+      y=rm/(sigma*r)
+      v1=a*eps*exp(-alp/y)
+      v1p=-alp*units*v1
+      v1pp=-alp*units*v1p
+      x2=y*y
+      f=eps
+      fp=0.
+      dy=d*y
+      v2=(c6+(c8+c10*x2)*x2)*x2**3
+      v2p=-(6.*c6+(8.*c8+10.*c10*x2)*x2)*x2**3/r
+      v2pp=(6.*7.*c6+(8.*9.*c8+10.*11.*c10*x2)*x2)*x2**3/r**2
+      if (dy.gt.1.) then
+         f=f*exp(-min(expmax,(dy-1.)**2))
+         fp=2.*(dy-1.)*dy*f/r
+         endif
+      v=v1-f*v2
+      vp=v1p-f*v2p-fp*v2
+      vpp=v1pp-f*v2pp-2.*fp*v2p
+      return
+      end

### 2019-04-19 doc/MachineLearning/Draft/article1.tex
added articles

--- 2019-01-19 doc/MachineLearning/article1.tex
@@
 is discretized in terms selected basis functions and/or many-body
-excitations, the dimensionality of systems under study grows almost
+excitations, the dimensionality of the systems under study grows almost
 exponentially when larger basis sets and/or number of particles are

### 2019-04-30 doc/PhysicsToday/PhysToday.tex
added physics today material

--- 2019-04-17 doc/TimeDependence/Paper1/main.tex
@@
-\documentclass[aip,jcp,reprint,floatfix]{revtex4-1}
+\documentclass[reprint,floatfix]{revtex4-1}
 
@@
 
-\usetikzlibrary{external}
-\tikzexternalize[prefix=figures/]
-
-\pgfplotsset{compat=1.15}
-
-\renewcommand{\vec}[1]{\boldsymbol{#1}}
-\newcommand{\hamiltonian}{\hat{H}}
-\newcommand{\onehamil}{\hat{h}}
-\newcommand{\twohamil}{\hat{u}}
-
-\newcommand{\vconf}{\hat{v}_{\text{conf}}}
-\newcommand{\vfield}{\hat{v}_{\text{field}}}
-\newcommand{\envelope}{\vec{\mathcal{E}}}
-\newcommand{\polarization}{\vec{\epsilon}}
-
 \begin{document}
 
-\title{Electron dynamics in quantum dot models using techniques from Quantum Chemistry}
+\title{Integrating a Computational Perspective in Physics Education}
 
-\author{Håkon Emil Kristiansen}
+\author{D.M. Caballero}
 \email{haakoek@uio.no}
-\affiliation{Hylleraas Centre for Quantum Molecular Sciences, Department of Chemistry, University of Oslo,N-0315 Oslo, Norway}
-
-\author{\O yvind Sigmundson Sch\o yen}
-\email{o.s.schoyen@fys.uio.no}
-\affiliation{Department of Physics, University of Oslo, N-0316 Oslo, Norway}
-
-\author{Sebastian G.~Winther-Larsen}
-\email{s.g.winther-larsen@fys.uio.no}
-\affiliation{Department of Physics, University of Oslo, N-0316 Oslo, Norway}
+\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
+\affiliation{Department of Physics and Center for Computing in Science Education, University of Oslo, N-0316 Oslo, Norway}
 
@@
 
-\section{TODO}
-\begin{itemize}
-    \item How do you treat the centre of mass problem???
-    \item The harmonic potential theorem\cite{HarmonicPotentialTheorem}: this depends on basis size!
-One way to check this is to let $u^{pq}_{rs} = 0$ and use a small number of basis functions.
-The short story: the particle density (in time) is rigidly translated when the confining potential is harmonic and the applied field is the simple dipole laser we use!
-    \item Furthermore, if I read correctly, we should observe that the (for example) the time dependent energy is effectively the same as for a single particle problem.
-
-    Quote from\cite{Chakraborty}: "In conclusion, the interaction of electrons in quantum
-dots leads to rich structure in their energy spectrum.
-However, FIR spectroscopy cannot probe it when the
-confining potential is quadratic because the optical excitations are then excitations of the cm and have exactly
-the same energies as single-electron excitations."
-
-    This is (again, if I understand the implications of the harmonic potential theorem correctly) due to the fact that the Hamiltonian separates into relative and center of mass coordinates with a harmonic potential and this type of applied field.
-This effect is independent of the number of particles.
-If we can actually demonstrate this, it is very solid foundation for claiming the correctness of the implementation.
-This also says something about the quality of the basis set.
-    \item Basically, what I am trying to say is that the "shape" of computed values (energy, overlap, expectation value of dipole, etc) is independent of the number of particles and will look like the corresponding values of the single particle problem apart from a scaling due to the number of particles.
-    \item Thus, it is maybe sensible to include comparisons with the single particle problem (which can be solved analytically or on a grid with f. ex RK4).
-\end{itemize}
-
-\section{Introduction}
-Quantum mechanics is a theory that describes/models the properties of microscopic systems. It postulates that given the so-called wavefunction,
-$\Psi(\mathbf{r},t)$, we can in principle compute all that there is to know about the system.  Given suitable initial conditions $\Psi(\mathbf{r},t)$ can be determined for all future time
-by solving the time-dependent Schrödinger equation (TDSE),
-\begin{equation}
-\label{TDSE}
- i \hbar \frac{\partial }{\partial t} \Psi(\mathbf{r},t) = \hat{H} \Psi(\mathbf{r},t).
-\end{equation}
-The initial condition is typically taken as some linear combination of the eigenfunctions.
-In particular, a prototypical choice of initial condition is the groundstate of the system.
-The eigenfunctions are found by solving the time-independent Schrödinger (TISE) equation
-\begin{equation}
- \hat{H} \Psi(\mathbf{r}) = E \Psi(\mathbf{r}).
-\end{equation}
-Exact/analytic solutions to the TISE is possible only for the
-simplest systems and general solutions to the TDSE are even
-rarer. Thus, one must resort to approximate/numeric methods. Of
-particular interest is the so-called many-body problem where one
-considers systems of two or more interacting particles.
-
-Several approaches for solving the many-body TISE
-numerically/approximately has been devised such as Hartree-Fock
-(HF), Density Functional Theory (DFT), Configuration Interaction (CI)
-and Coupled Cluster (CC). While efficient, Hartree-Fock and DFT are
-insufficient if one wants a high degree of accuracy. Configuration
-Interaction and Coupled Cluster methods are hierarchical in the sense
-that one can systematically construct increasingly accurate
-approximations. If the CI method is not truncated we have what is
-known as Full Configuration Interaction (FCI). FCI can be seen as
-exact (within some finite space), however it suffers from exponential
-scaling. Truncated CI methods which would achieve polynomial scaling
-are problematic since they are not size-consistent and
-extensive. Truncated CC methods one the other hand are size-consistent
-and extensive and achieves polynomial scaling. Due to this fact
-Coupled Cluster is considered the gold standard (misleading and ill-omened term as history has shown that a gold standard, in terms of monetary policy, is a bad idea - it does not work) of many-body techniques if high accuracy is desired.
-
-Similarly, there exists time-dependent generalizations of the methods discussed above.
-On one end we have the time-dependent Hartree-Fock (TDHF) and time-dependent density functional theory (TD-DFT) \cite{TDHF_2004, TD_DFT_Ullrich_Book}, which in this context are considered efficient/low-cost.
-Time-dependent full configuration interaction (TDFCI) and Multiconfiguration Time-Dependent Hartree-Fock (MCTDHF) on the other hand are considered the most accurate methods \cite{Hochstuhl2014}.
-However, TDCI and MCTDHF suffers from exponential scaling, which makes them impractical for systems with larger particle numbers.
-
-In a recent article \cite{OATDCC_2012}, Simen Kvaal put forward a generalization of Coupled-Cluster theory to the time domain, where the orbitals are allowed to be time-dependent.
-The method is referred to as the orbital-adaptive time-dependent coupled-cluster method (OATDCC), which can be viewed as a hierarchical approximation to the MCTDHF/TDFCI methods.
-OATDCC inherits size-consistency and extensivity from the CC method and achieves polynomial scaling.
-While not cheap, in the conventional sense, OATDCC alleviates some of the problems related to TDFCI and MCTDHF.
-
-We will apply several of the methods above in order to study electron dynamics in one- and two-dimensional quantum dot models.
-In particular, we will make a comparative analysis of the TDHF, TDCC (OATDCC) and TDFCI methods.
-
-\section{Theory}
-We assume that the reader is familiar with second-quantization/occupation-number theory.
-For a comprehensive introduction see for example \cite{szabo1996modern,helgaker2008molecular, shavitt2009many}.
-\subsection{Hartree-Fock Theory}
-In the Hartree-Fock approximation, the Hamiltonian reduces to a one-body term and a two-body term, written in the simplest way using second quantisation as
-\begin{equation}
-    \hat{H} = \hat{h} + \hat{u} 
-    = h^{p}_{q}c^\dagger_p c_q + \frac{1}{4}u^{pq}_{rs}c^\dagger_p c^\dagger_q c_s c_r
-\end{equation}
-
-and the expected value of the Hamiltonian is given by
-\begin{equation}
-    \label{HF_expecation}
-    E[\Phi] = \bra{\Phi}\hat{H}\ket{\Phi} 
-    = \bra{\phi_i}\hat{h}\ket{\phi_i} + \frac{1}{2}\bra{\phi_i\phi_j}\hat{u}\ket{\phi_i\phi_j}_{\text{AS}}
-\end{equation}
-where the assumption is that the ground state can be represented by a single Slater determinant
-\begin{equation}
-    \ket{\Phi} = \prod_{i=1}^{N} c^\dagger_i \ket{\text{vac}} = \ket{\phi_1 \dots \phi_N}.
-\end{equation}
-
-It can be shown that a Slater determinant wavefunction $\ket{\Phi_\text{HF}} = \ket{\phi_i \dots \phi_N}$ is a variational minimum of the function for $E[\Phi]$ in \autoref{HF_expecation} if the single-particle functions $\ket{\phi_i}$ that make up this Slater determinant satisfy the canonical Hartree-Fock equations,
-\begin{equation}
-    \label{canonical_HF}
-    \hat{f}(\phi_1, \dots , \phi_N)\ket{\phi_i} = \epsilon_i \ket{\phi_i}.
-\end{equation}
-where $\hat{f}$ is a one-body operator called the Fock operator,
-\begin{equation}
-    \hat{f}(\phi_1, \dots \phi_N) \equiv \hat{h} + \hat{v}_{\text{HF}},
-\end{equation}
-wherein the most important point of the Hartree-Fock approximation is revealed - one constrains the problem to a one-body problem where each particle only experiences a mean field from the other particles in the system.
-The canonical Hartree-Fock equations (\autoref{canonical_HF}) is reformulated as a matrix eigenvalue problem.
-This eigenvalue equation is called the Roothan-Hall equations,
-\begin{equation}
-    \label{roothan_hall}
-    F(U)U = U\epsilon,
-\end{equation}
-which is solved iteratively until self-consistency is reached, i.e. the next iteration does not contribute a significant amount to the mean field of the system.
-
-\subsubsection{Time-Dependent Hartree-Fock}
-
-We now assume that the Hamiltonian is time-dependent, satisfying the time-dependent Schrödinger equation (\autoref{TDSE}),
-\begin{equation}
-    \hat{H}(t) = \hat{h}(t) + \hat{u}(t).
-\end{equation}
-
-In time dependent Hartree-Fockn(TDHF) the wavefunction is approximated by a single Slater determinant, $\ket{\Phi(t)}$, where the orbitals $\ket{\phi_p(t)}$ are time dependent.
-The equations of motion for the orbitals are found by demanding that the action functional
-\begin{equation}
-    \mathcal{S}[\ket{\phi_p(t)}] = \int_0^T \mathcal{L} dt 
-\end{equation}
-is stationary with respect to variations of the orbitals, i.e $\delta \mathcal{S} = 0$. This is referred to as the Lagrange formulation of the time dependent variational principle\cite{Hochstuhl2014}.
-
-The Lagrangian $\mathcal{L}$ is given by 
-\begin{equation}
-    \mathcal{L} = \bra{\Phi(t)} \hat{H} - i\hbar \frac{\partial}{\partial t} \ket{\Phi(t)} 
-    - \lambda_{ij}(t)\left(\braket{\phi_i(t)|\phi_j(t)} - \delta_{ij}\right),
-\end{equation}
-where the Lagrange multipliers $\lambda_{ij}(t)$ enurses orthonormality of the orbitals.
-
-Computing the variations  with respect to the orbitals (see appendix \ref{TDHF_Derivation} for details) we obtain the TDHF equations 
-\begin{equation}
-    i \hbar \frac{\partial}{\partial t} \ket{\phi_p(t)} = \hat{f}(t) \ket{\phi_p(t)}.
-\end{equation}
-
-\subsection{Configuration Interaction Theory}
-
-(This section is a bit informal, but at least there is something here.)
-
-Configuration Interaction relies on representing the wavefunction as an expansion of orthonormal slater determinants built from a set of orthonormal single-particle functions,
-\begin{equation}
-    \label{hierarchical_CI}
-    \ket{\Psi} = A_0\ket{\Phi} + \sum_{ia}A^a_i\ket{\Phi^a_i} 
-    + \frac{1}{4}\sum_{ij} \sum_{ab} A^{ab}_{ij}\ket{\Phi^{ab}_{ij}} + \dots
-\end{equation}
-One may choose to truncate the computation in the hierarchical manner (e.g.
-stopping at the dots in \autoref{hierarchical_CI} would be CISD) or include all Slater determinants generated by the first $L$ single-particle functions $\phi_0$ through $\phi_{L-1}$.
-This gives a space of dimension $\binom{L}{N}$, and is called the full configuration interaction space (FCI space).
-
-Inserting the CI wavefunction into the TISE yields
-\begin{align}
-    \sum_{K}\hat{H} A_{KJ} \ket{\Phi_K} &= \sum_{K} E_J A_{KJ} \ket{\Phi_K} \\
-    \sum_{K} A_{KJ} \bra{\Phi_I}\hat{H} \ket{\Phi_K} &= \sum_{K} E_J A_{KJ} \bra{\Phi_I}\Phi_K\rangle \\
-    HA &= AE,
-\end{align}
-where $H$ is a matrix reprentation of the Hamiltionian, and $A$ is a matrix containing the CI expansion coefficients.
-
-\subsubsection{Time-Dependent Configuration Interaction}
-Time-development is simply a matter of inserting one's CI expansion (\autoref{hierarchical_CI}) into the TDSE (\textcolor{blue}{Haakon:} formally we can use the time dependent variational principle, minimize/compute variations wrp.
-to CI coeffs and get the same equation),
-\begin{align}
-    i\hbar \sum_I \frac{\partial A_I(t)}{\partial t} \ket{\Phi_I} &= \sum_I A_i(t) \hat{H}(t) \ket{\Phi_I} \\
-    i \hbar \sum_I \frac{\partial A_I(t)}{\partial t} \bra{\Phi_J}\Phi_I\rangle &= \sum_I A_I(t) \bra{\Phi_J} \hat{H}(t) \ket{\Phi_i} \\
-    i\hbar \frac{\partial A_I(t)}{\partial t} &= H(t)A(t).
-\end{align}
-We find a relatively simple, yet large, set of equation on matrix form. One is now free to apply one's favourite numerical integration scheme in order to compute the time-development.
-
-\subsection{Coupled Cluster Theory}
-
-    Full Configuration Interaction (FCI) will provide you with an exact solution, but since the FCI space grows Pascal-triangularly (Question: Is the growth exponential when the growth does not follow an exponential function?) and because we are constrained by limited amount of computing power, FCI is not practically feasible.
-The Hartree-Fock method, on the other hand, approximates systems by ignoring electron-electron interaction and can therefore be quite inaccurate.
-Coupled Cluster to the rescue!
-Hurray.
-Coupled cluster cleverly include higher-order excitations and takes electron correlation effects into account.
-Time-developed coupled cluster (TDCC) can be tricky, however.
-
-    \subsubsection{Ground State coupled cluster}
-    In coupled cluster theory we employ an exponential ansatz for the $N$-particle correlated wavefunction 
-    \begin{equation}
-       \ket{\Psi_{\text{CC}}} = e^{\hat{T}}\ket{\Phi} 
-    \end{equation}
-    where $\ket{\Phi}$ is a reference state, typically the HF-state. The cluster operator $\hat{T} = \sum_{k=1}^N \hat{T}_k$ is a sum of $k$-particle $k$-hole excitation operators, $\hat{T}_k$, 
-    \begin{equation}
-        \hat{T}_k = \left(\frac{1}{k!}\right)^2 \sum_{abc\cdots ijk\cdots} \tau^{abc\cdots}_{ijk\cdots}   c_a^\dagger  c_i c_b^\dagger c_j  c_c^\dagger c_k \cdots 
-    \end{equation}
-    where the unknowns $\tau^{abc\cdots}_{ijk\cdots}$ are referred to as cluster amplitudes.
-The operators $\hat{T}_1, \hat{T}_2, \hat{T}_3, \cdots$ and corresponding amplitudes $\tau^a_i, \tau^{ab}_{ij}, \tau^{abc}_{ijk},\cdots$ are referred to as singly, doubly, triply-, etc excited operators and amplitudes respectively.
-
-    Insertion of the exponential ansatz into the time independent Schrödinger equation results in 
-    \begin{equation}
-        \hat{H}e^{\hat{T}} \ket{\Phi} = E e^{\hat{T}} \ket{\Phi}.
-    \end{equation}
-    Multiplying with $e^{-T}$ we have that 
-    \begin{equation}
-        \bar{H}_{CC} \ket{\Phi} = E \ket{\Phi} \label{SimTransSE},
-    \end{equation}
-    where we have defined the similarity transformed Hamiltonian 
-    \begin{equation}
-        \bar{H}_{CC} \equiv e^{-\hat{T}}\hat{H}e^{\hat{T}}.
-    \end{equation}
-
-    Projecting equation (\ref{SimTransSE}) with the reference or any excited determinant gives 
-    \begin{align}
-        E_{CC} &= \braket{\Phi|\bar{H}_{cc}|\Phi} \label{CCenergy} \\
-        0 &= \braket{\Phi^\mu|\bar{H}_{cc}|\Phi} \label{AmplitudeEqsCC},
-    \end{align}
-    where $\mu$ denotes a generic excitation.
-Eq.
-(\ref{CCenergy}) is the coupled cluster energy, while eq.
-(\ref{AmplitudeEqsCC}) defines a set of nonlinear equations for the unknowns cluster amplitudes.
-In the limit where $\hat{T}$ is not truncated eqs.
-(\ref{CCenergy}) and (\ref{AmplitudeEqsCC}) is equivalent to the full configuration-interaction method\cite{VarCC_2013}.
-
-    (BCH-expansion of $\bar{H}$)!
-
-    It is important to note that since $\hat{T} \neq \hat{T}^\dagger$, the CC-energy is non-variational, i.e 
-    \begin{equation}
-        E_{CC} = \braket{\Phi|e^{-T}\hat{H}e^{T}|\Phi} \neq \braket{\Phi|e^{T^\dagger}\hat{H}e^{T}|\Phi} 
-    \end{equation}
-    and is not guaranteed to give an upper bound on the energy when $\hat{T}$ is truncated.
-Working directly with the expression $e^{T^\dagger}\hat{H}e^{T}$ is impractical since it does not truncate after four nested commutators as opposed to $\bar{H}$.
-
-    Since CC is non-variational the question of how to compute expectation values other than the energy arises. The usual definiton of the expectation value of an operator, A, is given by 
-    \begin{equation}
-        \braket{A} = \braket{\Psi|\hat{A}|\Psi}.
-    \end{equation}
-    However, we have seen that $\ket{\Psi}^\dagger \neq \bra{\Phi}e^{-\hat{T}}$ such that $E_{CC} \neq \braket{H}$.
-
-    The solution to this problem was introduced independently by workers in quantum chemistry and in theoretical physics\cite{VarCC_2013}.
-
-    The idea is to introduce independent paramterizations of the ket- and bra state, $\bra{\tilde{\Psi}} \neq \ket{\Psi}^\dagger$. [Work in progress...]
-
-    \subsubsection{The time dependent bivariational principle}
-    (ASSUMING THE PREVIOUS SECTION IN PLACE)!
-
-    The extension of CC-theory to the time domain can be done through a generalization of the time dependent variational principle\cite{BECK2001}, to a corresponding time dependent bivariational principle\cite{OATDCC_2012}.
-
-    Consider the action-like functional 
-    \begin{align}
-    \mathcal{S}[\bra{\tilde{\Psi}}, \ket{\Psi}] &= \int_0^T \braket{\tilde{\Psi}(t)|i\frac{d}{dt}-\hat{H}(t)|\Psi(t)} dt  \nonumber \\ 
-    &= \int_0^T \braket{\tilde{\Phi}(t)|(\mathbb{1}+\hat{\Lambda}(t))e^{-\hat{T}(t)}\left(i\frac{d}{dt}-\hat{H}(t)\right) e^{\hat{T}(t)}|\Phi(t)} dt \nonumber 
-    \end{align}
-    where the Hamiltonian generally is time dependent. The integrand can be written more compactly as 
-    \begin{align}
-        &\braket{\tilde{\Phi}(t)|(\mathbb{1}+\hat{\Lambda}(t))e^{-\hat{T}(t)}\left(i\frac{d}{dt}-\hat{H}(t)\right) e^{\hat{T}(t)}|\Phi(t)} \nonumber \\ 
-        = \ \ & i \lambda_\mu \dot{\tau}^\mu + \rho^q_p \left( h^p_q - i \eta^p_q \right) + \frac{1}{4}u^{pr}_{qs}\rho^{qs}_{pr}  \\
-        =\ \ & i \lambda_\mu \dot{\tau}^\mu - \mathcal{E}_{\hat{H}-iD_0}[\lambda, \tau] 
-    \end{align}
-    where we have defined 
-    \begin{align}
-        D_0 &= \braket{\tilde{\varphi}_p|\dot{\varphi}_q} c_p^\dagger \tilde{c}_q \\
-        \eta^p_q &= \braket{\tilde{\varphi}_p|\dot{\varphi}_q}.
-    \end{align}
-
-    Thus, 
-    \begin{align}
-        \mathcal{S}[\bra{\tilde{\Psi}}, \ket{\Psi}] &= \int_0^T i \lambda_\mu \dot{\tau}^\mu + \rho^q_p \left( h^p_q - i \eta^p_q \right) + \frac{1}{4}u^{pr}_{qs}\rho^{qs}_{pr} dt \\
-        &= \int_0^T i \lambda_\mu \dot{\tau}^\mu - \mathcal{E}_{\hat{H}-iD_0}[\lambda, \tau] dt
-    \end{align}
-
-    The time dependent bivariational principle states that equations of motion are obtained by demanding that $\mathcal{S}$ is stationary ($\delta S = 0$) with respect to all independent variations of the parameters.
-Genraly both amplitudes and single-particle function/orbitals can be varied.
-    \subsubsection{Time dependent amplitude equations}
-    The equations of motions for the amplitudes read 
-    \begin{align}
-    i \dot{\tau}^\mu &= \frac{\partial}{\partial \lambda_\mu} \mathcal{E}_{\hat{H}-iD_0}[\lambda, \tau] \nonumber \\ 
-    &=  \braket{\tilde{\Phi}_\mu|e^{-\hat{T}}(\hat{H}-iD_0)e^{\hat{T}}|\Phi} \label{OATDCC_tau_equations} \\
--i \dot{\lambda}_\mu &= \frac{\partial}{\partial \tau^\mu} \mathcal{E}_{\hat{H}-iD_0}[\lambda, \tau] \nonumber \\ 
-&=  \braket{\tilde{\Phi}|(1+\Lambda)e^{-\hat{T}}[\hat{H}-iD_0,X_\mu]e^{\hat{T}}|\Phi} \label{OATDCC_lambda_equations}
-\end{align}
-
-The right-hand sides of equations [\ref{OATDCC_tau_equations}, \ref{OATDCC_lambda_equations}] are the same amplitude equations as in time independent cc-theory, plus an extra term due to the time dependence of the orbitals.
-
-If the orbitals are chosen to be stationary\cite{Symplectic_TDCC_2018}, we have $D_0 = 0$ and the time dependence is determined by the amplitudes.
-This implies a static reference such as the Hartree-Fock determinant.
-In that case, we also have that $\bra{\tilde{\Phi}} = \bra{\Phi} $, i.e an orthonormal basis.
-
-\subsubsection{Time dependent orbital equations}
-Computing the variations of $\mathcal{S}$ with respect to the orbitals, we obtain equations of motion for the time-dependent orbitals.
-The time derivatives of the orbitals can, in general, be split into a P-space and a Q-space,
-\begin{align}
-    \ket{\dot{\varphi}_q} &= (P+Q)\ket{\dot{\varphi}_q} = \sum_p \eta^p_q \ket{\varphi_p} + Q\ket{\dot{\varphi}_q} \\
-    \bra{\dot{\tilde{\varphi}}_q} &= \bra{\dot{\tilde{\varphi}}_q}(P+Q) = -\sum_q \eta^p_q \bra{\tilde{\varphi}_q} + \bra{\dot{\tilde{\varphi}}_p}Q
-\end{align}
-where we have defined 
-\begin{align}
-    P &= \sum_{p=1}^M \ket{\varphi_p}\bra{\tilde{\varphi}_p},  \ \ M \leq N_b \\
-    Q &= \mathbb{1}-P
-\end{align}
-with $N_b$ denoting the number of single-particle functions.
-
-This splitting is identical to what is done in MCTDHF-theory \cite{Hochstuhl2014}, where a trunctation of the $P$-space is used to reduce the number of Slater determinants the expansion of the CI-wavefunction.
-
-In coupled cluster methods the number of Slater determinants is reduced through the truncation of the cluster operators.
-Thus, we can choose to have $M=N_b$, resulting in $Q=0$.
-In the following we restrict our attention to the case where $Q=0$.
-
-Note that for the MCTDHF method this would be the same as FCI with time dependent orbitals and nothing is gained in terms of reducing computational complexity.
-
-With $Q=0$ the equations of motion for the orbitals is given by
-\begin{align}
-    \ket{\dot{\varphi}_q} &=  \sum_p \eta^p_q \ket{\varphi_p} \label{Reduced ket-orbital equation} \\
-    \bra{\dot{\tilde{\varphi}}_q} &=  -\sum_q \eta^p_q \bra{\tilde{\varphi}_q} \label{Reduced bra-orbital equation}. 
-\end{align}
-The non-zero components of $\eta^p_q$ are given by the solutions of the linear equations
-\begin{align}
-    i \sum_{bj} A^{ib}_{aj} \eta^j_b &= \sum_j \rho^i_j h^j_a - \sum_b \rho^b_a h^i_b \nonumber \\
- + &\frac{1}{2} \left[\sum_{prs} \rho^{is}_{pr} u^{pr}_{as} - \sum_{rqs} \rho^{qs}_{ar} u^{ir}_{qs}  \right] \\
-    -i \sum_{bj} A^{ja}_{bi} \eta^b_j &= \sum_b \rho^a_b h^b_i - \sum_j \rho^j_i h^a_j \nonumber \\
- + &\frac{1}{2} \left[ \sum_{prs} \rho^{as}_{pr} u^{pr}_{is} - \sum_{rqs} \rho^{qs}_{ir} u^{ar}_{qs} \right] \nonumber \\ 
- + &i \dot{\rho}_i^a \label{P-space bra equation}
-\end{align}
-where
-\begin{align}
-    A^{ib}_{aj} &= \braket{\tilde{\Psi}|[c_j^\dagger c_b,c_a^\dagger c_i]|\Psi} \nonumber \\
-    &= \delta^b_a \rho^i_j - \delta^i_j \rho^b_a \nonumber  \\ 
-    &= -A^{bi}_{ja}.
-\end{align}
-
-The simplest OATDCC case is the OATDCCD approximation where the cluster operators include double excitations only.
-One can show that the singles amplitudes are formally redundant \cite{OATDCC_2012}, leaving the one-body density matrix block diagonal, with only $\rho^i_j$ and $\rho^a_b$ being non-zero.
-Thus, the last term in eq.
-(\ref{P-space bra equation}) is zero.
-
-Furthermore, there is no contributions to the amplitude equations (\ref{OATDCC_tau_equations}) and (\ref{OATDCC_lambda_equations}) from $D_0$ in the doubles-only approximation.
-This follows from that only $\eta^a_i$ and $\eta^i_a$ are non-zero.
-
-\section{Systems}
-    Relevant articles:
-    \begin{itemize}
-        \item Simple spin-dependent field TDHF study:\cite{TDHF_Singlet_Triplet}
-        \item One dimensional time-dependent quantum dot: \cite{Zanghellini04}
-        \item Harmonic single well: \cite{HjorthJensenQD_2017}
-        \item Theoretical study double well: \cite{Double_Well_2000}
-        \item Experiment with two fermions in double well: \cite{Fermions_Double_Well_Experiment}
-        \item Coupled double well: \cite{Coupled_double_wells}
-        \item Time domain two electrons in quantum dot: \cite{TimeDomainTwoElectronsQuantumDot}
-        \item Time-dependency with spin-orbit coupling: \cite{Spin_orbit_coupling}
-        \item Time-domain simulation of quantum spin: \cite{Time_domain_quantum_spin}
-        \item Entanglement in spin-qubit quantum dots: \cite{Entanglement_spin_qubit_quantum_dots}
-    \end{itemize}
-
-    Generic information on quantum dots and why we are interested in them.
-
-    General Hamiltonian for interacting electrons in a (time-dependent) confining potential influenced by an external electric field (laser),
-    \begin{align}
-    \hat{H}(t) = &\sum_{i=1}^N \left( - \frac{1}{2}\nabla^2 + \vconf(\mathbf{r}_i,t) + \vfield(\mathbf{r}_i,t) \right) \nonumber \\ 
-    + &\sum_{i < j}^N \frac{1}{r_{ij}}.
-    \label{eq:general_hamiltonian}
-    \end{align}
-
-    \subsection{One dimensional quantum dot}
-        We will use the one-dimensional quantum dot model considered by Zanghellini et. al in \cite{Zanghellini04}, as a benchmark/validation for our implementations.
-
-    \subsection{Two dimensional quantum dot}
-        For a two dimensional quantum dot model we will consider different confining potential.
-There already exists studies concerning the ground state properties the electrons are confined by the harmonic oscillator potential, see for example \cite{HjorthJensenQD_2017}.
-We will use this as a starting point for time-dependent studies.
-    \subsubsection{Single well}
-        The Hamiltonian of the two-dimensional harmonic oscillator quantum dot is given by \autoref{eq:general_hamiltonian}, with the confining potential
-        \begin{align}
-            \vconf(\boldsymbol{r}_i) \equiv \frac{1}{2} \omega^2 \boldsymbol{r}_i^2.
-        \end{align}
-    \subsubsection{Double well}
-    \subsection{Open system/finite potential wells}
-        Maybe too ambitious in this work.
-
-\section{Time-dependent fields}
-    \subsection{Dipole laser field}
-        The dipole laser field we will be looking at is given by
-        \begin{align}
-            \vfield(\boldsymbol{r}_i, t)
-            &= \envelope(t) \cdot \boldsymbol{r}_i,
-        \end{align}
-        where the \emph{laser envelope} $\envelope(t)$ is given by a time-varying field, e.g., purely sinusoidal with a constant amplitude as in the article by Zanghellini et.
-al \cite{Zanghellini04}, and with a polarization direction $\polarization$.
-        The dipole-part of the name of the laser stems from the position vector $\vec{r}_i$.
-    \subsection{Magnetic field}
-        Figure out how to model this field.
-
-        Hamiltonian for a single particle in a magnetic field \cite{TimeDomainTwoElectronsQuantumDot}
-\begin{equation}
-    \hat{H} = \frac{1}{2m} \left( \frac{\hbar}{i} \nabla - q \vec{A} \right)^2 + V(\vec{r})
-\end{equation}
-For a uniform magnetic field in the z-direction, we have 
-\begin{equation}
-    \vec{A} = \frac{B}{2} (-y,x,0) = B_0 (-y,x,0)
-\end{equation}
-Thus, 
-\begin{equation}
-    \hat{H} \psi(x,y) = \left(-\frac{\hbar^2}{2m} \nabla^2  - \frac{2q}{2m}\vec{A} \cdot \nabla  + \frac{q^2}{2m} A^2 + V(x,y) \right) \psi(x,y).
-\end{equation}
-Choosing atomic units ($\hbar = q = m = 1$), we have
-\begin{equation}
-    \hat{H} \psi(x,y) = \left(-\frac{1}{2} \nabla^2  -\vec{A} \cdot \nabla  + \frac{1}{2}A^2 + V(x,y) \right) \psi(x,y).
-\end{equation}
-
-\section{Results \& Discussion}
-
-    \subsection{One-dimensional quantum dot benchmark}
-        Comparison of the varying methods with the results and figures from Zanghellini et. al\cite{Zanghellini04}.
-
-        \begin{figure}
-        \includegraphics[width=0.5\textwidth]{./figures/zanghellini_fig2.eps}
-        \caption{\label{fig:1DZanghellini} Probability of being in the ground state
-        $|\langle\Psi(x,0\ket{\Psi(x,t)}|^2$ for one-dimensional quantum dot with frequency
-        $\omega = 8\Omega$ and two electrons. This plot is an exact match with figure 2 in Zanghellini
-        et al.\cite{Zanghellini04}
-        }
-        \end{figure}
-
-    \subsection{Two-dimensional quantum dot calculations}
-        Results from different interesting time-dependent observable quantities for distinct confining potentials and time-dependent fields (maybe too ambitious on different fields).
-
-\section{Concluding remarks}
-\begin{appendices}
-\section{Derivation of TDHF equations} \label{TDHF_Derivation}
-By introducing functional derivatives and their action on the matrix elements (see Hochstuhl et al \cite{Hochstuhl2014}),
-\begin{align}
-    \frac{\delta}{\delta \phi^*_k}\bra{\phi_i}\hat{h}\ket{\phi_i} &= \hat{h}\ket{\phi_k} \\
-    \frac{\delta}{\delta \phi^*_k}\bra{\phi_i}\frac{\partial}{\partial t} \ket{\phi_i} 
-    &= \frac{\partial}{\partial t}\ket{\phi_k} \\
-    \frac{\delta}{\delta \phi^*_k}\bra{\phi_i \phi_j} \hat{u} \ket{\phi_i, \phi_j}_{\text{AS}}
-    &= 2\bra{\cdot \phi_i} \hat{u} \ket{\phi_k \phi_i}_{\text{AS}},
-\end{align}
-one can find the equations of motions for the system by way of the time-dependent variational principle,
-\begin{equation}
-    \label{td_variational_principle}
-    \bra{\delta \Phi} \hat{H} - i\hbar \frac{\partial}{\partial t} \ket{\Phi} = 0.
-\end{equation}
-
-(INSERT SHORT NOTE ON EMPTY SLOT NOTATION?)
-
-We are faced with an optimisation problem under the constraint that the orbitals are orthonormal ($\braket{\phi_i(t)}{\phi_j(t)} = \delta_{ij})$, yielding the following Lagrange functional,
-\begin{equation}
-    \mathcal{L} = \bra{\Phi} \hat{H} - i\hbar \frac{\partial}{\partial t} \ket{\Phi} 
-    - \lambda_{ij}(\braket{\phi_i(t)}{\phi_j(t)} - \delta_{ij}),
-\end{equation}
-with an optimal state $\ket{\phi_k}$ satisfying
-\begin{equation}
-    \label{HF_optimal_phi}
-    \hat{h}\ket{\phi_k} + \bra{\cdot \phi_j} \hat{u} \ket{\phi_k \phi_j}_{\text{AS}} 
-    - i\hbar\frac{\partial}{\partial t} \ket{\phi_k} - \lambda_{kj} \ket{\phi_j} = 0.
-\end{equation}
-Left-projecting this expression onto a state $\bra{\phi_l}$ and solving for the lagrange multiplier $\lambda$ gives
-\begin{equation}
-    \lambda_{kl} = \bra{\phi_l} \hat{h} \ket{\phi_k} 
-    + \bra{\phi_l \phi_j} \hat{u} \ket{\phi_k \phi_j}_{\text{AS}}
-    - i\hbar \bra{\phi_l} \frac{\partial}{\partial t} \ket{\phi_k}.
-\end{equation}
-Inserting this expression into \autoref{HF_optimal_phi} yields
-\begin{equation}
-    i\hbar \hat{P}\frac{\partial}{\partial t} \ket{\phi_k} = \hat{P} \hat{f} \ket{\phi_k},
-\end{equation}
-where we have defined a projection opertor $\hat{P} = \mathbb{1} - \ket{\phi_i}\bra{\phi_i}$ and the Fock operator $\hat{f} \equiv \hat{h} + \bra{\cdot \phi_j} \hat{u} \ket{\cdot \phi_j}_{\text{AS}}$.
-
-For unitary time-dependent operator given by $\hat{Q}(t) = i\hbar (\partial / \partial t)$,
-we have
-\begin{equation}
-    \label{TDHF}
-    i \hbar \frac{\partial}{\partial t} \ket{\phi_k(t)} = \hat{f} \ket{\phi_k(t)},
-\end{equation}
-if we choose $\hat{Q}(t) = \hat{f}(t)$.
-\autoref{TDHF} is the time-dependent Hartree-Fock equations, which can be rewritten on matrix form by expanding the molecular slater wavefunctions in an atomic basis with time-dependent coefficients and left projecting with an arbitrary atomic basis function,
-\begin{align}
-    i \hbar \frac{\partial}{\partial t} C_{\alpha i} (t) \ket{\chi_\alpha} &= \hat{f}(t) C_{\alpha i} (t) \ket{\chi_\alpha} \\
-    \to i \hbar \frac{\partial}{\partial t} C_{\alpha i} (t) \langle{\chi_{\beta}}|{\chi_{\alpha}}\rangle &= C_{\alpha i}(t) \bra{\chi_\beta} \hat{f}(t) \ket{\chi_\alpha} \\
-    \to i \hbar \mathbf{S}\dot{\mathbf{C}}(t) &= \mathbf{F}(t) \mathbf{C}(t).
-\end{align}
-\section{The time dependent energy}
-In general the time dependent energy is given by 
-\begin{equation}
-    E(t) = \braket{\Psi(t)|\hat{H}(t)|\Psi(t)}.
-\end{equation}
-If $\hat{H}$ does not depend explicitly on time, we find that the energy is conserved
-\begin{align*}
-     \frac{\partial}{\partial t}E(t) &= \braket{\frac{\partial \Psi(t)}{\partial t}|\hat{H}|\Psi(t)} + \braket{\Psi(t)|\hat{H}|\frac{\partial \Psi(t)}{\partial t}} \\
-     &= \braket{\Psi(t)|\hat{H}|\frac{\partial \Psi(t)}{\partial t}}^* + \braket{\Psi(t)|\hat{H}|\frac{\partial \Psi(t)}{\partial t}} \\
-     &= (-i\braket{\Psi(t)|\hat{H}^2|\Psi(t)})^* -i\braket{\Psi(t)|\hat{H}^2|\Psi(t)} \\
-     &= 0
-\end{align*}
-where we used the time dependent Schrödinger equation 
-\begin{equation}
-    i \frac{\partial \Psi(t)}{\partial t}  = \hat{H} \Psi(t)
-\end{equation}
-to obtain the last equalities.
-\section{Semi-analytical solutions for two electrons confined by a harmonic oscillator potential}
-We review the (semi)-analytical solution(s) of two electrons confined by a harmonic oscillator potential as discussed by Taut \cite{Taut93,Taut94}, in two- and three spatial dimensions.
-We also consider the analogous one dimensional problem.
-
-Consider the Hamiltonian for two interacting particles in a harmonic oscillator potential, 
-\begin{equation}
-    \hat{H} = \sum_{i=1}^2 \left(\frac{1}{2} \mathbf{p}_i^2 + \frac{1}{2} \Omega^2 \mathbf{r}_i^2\right) + v_{\text{int}}(\mathbf{r}_1,\mathbf{r}_2). 
-\end{equation}
-In one spatial dimension the interaction between the electrons is modelled by a smoothed Coulomb potential \cite{SmoothedCoulomb} 
-\begin{equation}
-    v_{\text{int}}(x_1,x_2) = \frac{1}{\sqrt{(x_1-x_2)^2 + a^2}}
-\end{equation}
-where $a$ is a shielding parameter. In two- or three spatial dimensions the interaction is given by the usual Coulomb potential.
-
-Following Taut \cite{Taut93} we introduce the difference vector and the center of mass as new variables,
-\begin{align}
-    \mathbf{r} &= \mathbf{r}_2-\mathbf{r}_1 \\
-    \mathbf{R} &= \frac{1}{2} \left( \mathbf{r}_1+\mathbf{r}_2 \right). 
-\end{align}
-The key observation is that the Hamiltonian decouples in these variables
-\begin{align}
-    H &= - \nabla_\mathbf{r}^2 + \frac{1}{4} \Omega^2 \mathbf{r}^2 + v_{\text{int}}(\mathbf{r})  - \frac{1}{4}\nabla_\mathbf{R}^2 + \Omega^2 \mathbf{R}^2 \nonumber \\
-    &= 2 \left( -\frac{1}{2} \nabla_\mathbf{r}^2 + \frac{1}{2} \Omega_\mathbf{r}^2 \mathbf{r}^2 + \frac{1}{2}v_{\text{int}}(\mathbf{r}) \right) \nonumber \\  &+ \frac{1}{2} \left( -\frac{1}{2}\nabla_\mathbf{R}^2 + \frac{1}{2}\Omega_\mathbf{R}^2 \mathbf{R}^2 \right) \nonumber \\
-    &= 2 H_\mathbf{r} + \frac{1}{2}H_\mathbf{R} 
-\end{align}
-where we have defined 
-\begin{align}
-    \Omega_\mathbf{r} &\equiv \frac{1}{2}\Omega \\
-    \Omega_\mathbf{R} &\equiv 2 \Omega \\
-    H_\mathbf{r} &\equiv -\frac{1}{2} \nabla_\mathbf{r}^2   + \frac{1}{2} \Omega_\mathbf{r}^2 \mathbf{r}^2 + \frac{1}{2}v_{\text{int}}(\mathbf{r}) \\
-    H_\mathbf{R} &\equiv -\frac{1}{2}\nabla_\mathbf{R}^2 + \frac{1}{2}\Omega_\mathbf{R}^2 \mathbf{R}^2
-\end{align}
-Since $H$ is independent of spin, the total wave function can be factorized as follows\cite{Taut93} 
-\begin{equation}
-    \psi(\mathbf{r},\mathbf{R},\mathbf{s}) = \varphi(\mathbf{r}) \xi(\mathbf{R}) \chi(s_1,s_2).
-\end{equation}
-Thus, the time-independent Schrödinger equation 
-\begin{align}
-    H\psi(\mathbf{r},\mathbf{R},\mathbf{s}) &= \left( 2 H_\mathbf{r} + \frac{1}{2}H_\mathbf{R} \right) \varphi(\mathbf{r}) \xi(\mathbf{R}) \chi(s_1,s_2) \\ 
-    &= E\psi(\mathbf{r},\mathbf{R},\mathbf{s}) \\
-    &= \left( E_\mathbf{r} + E_\mathbf{R} \right) \varphi(\mathbf{r}) \xi(\mathbf{R}) \chi(s_1,s_2)
-\end{align}
-separates into one equation for the relative motion
-\begin{equation}
-    \left[- \frac{1}{2}\nabla_\mathbf{r}^2 + \frac{1}{2} \omega_r^2 \mathbf{r}^2 + \frac{1}{2}v_{\text{int}}(\mathbf{r}) \right] \varphi(\mathbf{r}) = E^\prime_\mathbf{r} \varphi(\mathbf{r}) \label{RelativeMotionEquation}
-\end{equation}
-where $E^\prime_\mathbf{r} = \frac{1}{2}E_\mathbf{r}$, and one for the center of mass 
-\begin{equation}
-    \left[- \frac{1}{2}\nabla_\mathbf{R}^2 + \frac{1}{2}\Omega_R^2 \mathbf{R}^2 \right] = E^\prime_\mathbf{R} \xi(\mathbf{R}) \label{CenterOfMassEquation}
-\end{equation}
-where $E^\prime_\mathbf{R} = 2 E_\mathbf{R}$. The total energy is then given by 
-\begin{equation}
-    E = E_\mathbf{r} + E_\mathbf{R} = 2E^\prime_\mathbf{r} + \frac{1}{2}E^\prime_\mathbf{R}.
-\end{equation}
-
-We recognize equation (\ref{CenterOfMassEquation}) as the eigenvalue problem for a (d-dimensional) harmonic oscillator with frequency $\Omega_R$.Thus, 
-\begin{equation}
-    E_\mathbf{R}^\prime = \Omega_\mathbf{R} \left( n_d + \frac{d}{2} \right)
-\end{equation}
-for $d=1,2,3$.
-
-In one dimension we can solve the eigenvalue problem for the relative motion (\ref{RelativeMotionEquation}) numerically, with the smoothed Coulomb potential, using for example a finite difference discretization.
-
-In the next section we discuss how to solve the equation for the relative motion i two spatial dimensions.
-
-\section{Solving the two-dimensional Schrödinger equation for a central symmetric potential}
-Consider the Schrödinger equation in two dimensions for a central symmetric potential where we use polar coordinates 
-\begin{equation}
-    \left[ - \frac{1}{2} \nabla^2_{r,\theta} + V(r) \right] \psi(r,\theta) = E\psi(r,\theta). \label{SEPolar}
-\end{equation}
-The Laplacian in polar coordinates is given by
-\begin{equation}
-    \nabla_{r,\theta}^2 = \frac{1}{r} \frac{\partial}{\partial r} \left( r \frac{\partial}{\partial r} \right) + \frac{1}{r^2} \frac{\partial^2}{\partial \theta^2}
-\end{equation}
-
-Inserting the expressions for the Laplacian into equation (\ref{SEPolar}) we have that
-\begin{equation}
-    \left[ - \frac{1}{2} \left(\frac{1}{r} \frac{\partial}{\partial r} \left( r \frac{\partial}{\partial r} \right) + \frac{1}{r^2} \frac{\partial^2}{\partial \theta^2} \right) + V(r) \right] \psi(r,\theta) = E\psi(r,\theta).
-\end{equation}
-
-We look for solutions that are seprable on the form
-\begin{equation}
-    \psi(r,\theta) = R(r) \varphi(\theta)
-\end{equation}
-such that
-\begin{equation*}
-     - \frac{1}{2} \left(\frac{\varphi}{r} \frac{\partial}{\partial r} \left( r \frac{\partial R}{\partial r} \right) + \frac{R}{r^2} \frac{\partial^2 \varphi}{\partial \theta^2} \right) + V(r)R\varphi   = ER\varphi.
-\end{equation*}
-
-Multiplying with $-2r^2$ and dividing with $R\varphi$ we obtain
-\begin{equation*}
-      \left[\frac{r}{R}\frac{\partial}{\partial r} \left( r \frac{\partial R}{\partial r} \right) -2r^2 (V(r)-E) \right] + \frac{1}{\varphi} \frac{\partial^2 \varphi}{\partial \theta^2}     = 0,
-\end{equation*}
-which implies that 
-\begin{equation}
-    \frac{r}{R}\frac{\partial}{\partial r} \left( r \frac{\partial R}{\partial r} \right) -2r^2 (V(r)-E) = m^2 \end{equation}
-    and
-    \begin{equation}
-    \frac{1}{\varphi} \frac{\partial^2 \varphi}{\partial \theta^2} = -m^2.
-\end{equation}
-
-\subsection{The angular equation}
-The second (angular) equation is satisfied by 
-\begin{equation}
-    \varphi = Ne^{im\theta}
-\end{equation}
-with $N$ a normalization constant. Requiring 
-\begin{equation}
-    \varphi(\theta + 2\pi) = \varphi(\theta)
-\end{equation}
-implies that $m=0,\pm 1, \pm 2,\cdots$. The normalization condition 
-\begin{equation}
-\int_0^{2\pi} |\varphi|^2 d\theta = 2\pi|N|^2 = 1    
-\end{equation} gives $N = 1/\sqrt{2\pi}$.
-
-\subsection{The radial equation}
-Next, we consider the radial equation
-\begin{equation}
-    r\frac{\partial}{\partial r} \left( r \frac{\partial R}{\partial r} \right) -2r^2 (V(r)-E)R = Rm^2 \label{2d-RadialEquation}
-    \end{equation}
-This equation simplifies if we change variables
-\begin{equation}
-    R(r) = \frac{u(r)}{r^{1/2}}
-\end{equation}
-and note that 
-\begin{equation}
-    r\frac{\partial}{\partial r} \left( r \frac{\partial R}{\partial r} \right) = r^{-1/2} \left[ r^2 \frac{d^2 u}{dr} + \frac{1}{4}u(r) \right].
-\end{equation}
-Insertion in equation (\ref{2d-RadialEquation}) yields
-\begin{equation}
-    \left[ -\frac{1}{2} \frac{d^2}{dr^2} + \frac{1}{2r^2} \left( m^2 - \frac{1}{4} \right) + V(r) \right] u(r) = Eu(r) \label{2d-radial-new}.
-\end{equation}
-
-For two interacting particles confined by a harmonic oscillator potential the potential is given by 
-\begin{equation}
-    V(r) = \frac{1}{2}\Omega_r^2 + \frac{1}{2r},
-\end{equation}
-for which equation (\ref{2d-radial-new}) can be solved analytically for particular oscillator frequencies\cite{Taut94}.
-
-One can also solve the equation numerically for arbitrary oscillator frequencies. The main challenge in a numerical approach is how to deal with the singularities at $r=0$.
-
-\section{Semi-analytical solution of a laser-driven two electron system}
-Following Schwengelbeck \cite{SCHWENGELBECK_1999} we consider the problem of two interacting electrons, confined by a harmonic oscillator potential, subject to a time-dependent sinusoidal laser-field.
-The Hamiltonian of the system is given by
-\begin{equation}
-    \hat{H} = \sum_{i=1}^2 \left(\frac{1}{2} \mathbf{p}_i^2 + \frac{1}{2} \Omega^2 r_i^2 + \mathbf{r}_i \cdot \hat{\mathbf{e}} E_0 \sin(\omega t)  \right) + v_{\text{int}}(\mathbf{r}_1,\mathbf{r}_2). 
-\end{equation}
-where $\omega$ and $E_0$ are the frequency and amplitude of the field and $\hat{\mathbf{e}}$ denotes a unit polarization vector.
-
-As in the previous section, introducting the relative distance and center of mass as new variables, the Hamiltonian decouples
-\begin{align}
-    H &= 2 \underbrace{\left( -\frac{1}{2} \nabla_\mathbf{r}^2 + \frac{1}{2} \Omega_\mathbf{r}^2 \mathbf{r}^2 + \frac{1}{2}v_{\text{int}}(\mathbf{r}) \right)}_{\equiv H_\mathbf{r}} \nonumber \\  &+ \frac{1}{2} \underbrace{\left( -\frac{1}{2}\nabla_\mathbf{R}^2 + \frac{1}{2}\Omega_\mathbf{R}^2 \mathbf{R}^2 + 4\mathbf{R} \cdot \hat{\mathbf{e}} E_0 \sin(\omega t) \right)}_{\equiv H_\mathbf{R}}
-\end{align}
-and the time-dependent Schrödringer equation admits solutions of the factorized form\cite{SCHWENGELBECK_1999}
-\begin{equation}
-    \Psi(\mathbf{r},\mathbf{R},\mathbf{s},t) = \psi(\mathbf{R},t) \varphi(\mathbf{r},t) \chi(s_1,s_2).
-\end{equation}
-We use the ground state of the unperturbed system as initial condition
-\begin{equation}
-    \Psi(\mathbf{r},\mathbf{R},\mathbf{s},0) = \psi_0(\mathbf{R},0) \varphi_0(\mathbf{r},0) \chi(s_1,s_2),
-\end{equation}
-where $\varphi_0(\mathbf{r})$ and $\psi_0(\mathbf{R})$ are the ground state solutions of equations (\ref{RelativeMotionEquation}) and (\ref{CenterOfMassEquation}) respectively.
-
-In particular, we get one equation for the relative motion 
-\begin{equation}
-    i \frac{\partial}{\partial t} \varphi(\mathbf{r},t) = 2H_\mathbf{r} \varphi(\mathbf{r},t)
-\end{equation}
-and one for the center of mass
-\begin{equation}
-    i \frac{\partial}{\partial t} \psi(\mathbf{R},t) = \frac{1}{2}H_\mathbf{R} \psi(\mathbf{R},t) \label{time-dependent CM eq}.
-\end{equation}
-Since $H_\mathbf{r}$ contains no explicit time-dependent terms, the equation for the relative motion is simply given by 
-\begin{equation}
-    \varphi(\mathbf{r},t) = \varphi_0(\mathbf{r},0) \exp(-i E^\prime_{0,\mathbf{r}} t)
-\end{equation}
-where $\{(\phi_0(\mathbf{r},0),E^\prime_{0,\mathbf{r}}\}$ is given by the lowest lying solution of equation (\ref{RelativeMotionEquation}).
-
-\subsection{Solving the center of mass equation in time}
-Eq.
-(\ref{time-dependent CM eq}) associated with the center of mass motion of the two electrons, can be solved analytically by means of a path integral\cite{SCHWENGELBECK_1999} or we can use a numerical approach.
-
-\subsubsection{One spatial dimension}
-Consider first the one dimensional problem, $(\mathbf{r},\mathbf{R}) = (x,X)$ with polarization vector $\hat{\mathbf{e}}$ along the positive x-direction 
-\begin{align}
-    i \frac{\partial \psi(X,t)}{\partial t} &= \frac{1}{2} H_X(t) \psi(X,t) \label{OneDimTimeDepEq}
-\end{align}
-where 
-\begin{equation}
-    H_X(t) =  -\frac{1}{2}\frac{d^2}{dX^2} + \frac{1}{2}\Omega_X^2 X^2 + 4X \sin(\omega t) 
-\end{equation}
-and $\Omega_X \equiv 2\Omega$. We use the ground state of the unperturbed harmonic oscillator as initial state for center of mass wavefunction 
-\begin{equation}
-    \psi(X,0) = \left( \frac{\Omega_X}{\pi} \right)^{1/4} \exp\left(-\frac{\Omega_X X^2}{2}\right).
-\end{equation}
-Equation (\ref{OneDimTimeDepEq}) is solved by using the standard Runge-Kutta fourth order (RK4) scheme.
-
-Thus, we have that the total wavefunction is given by 
-\begin{align}
- \Psi(x,X,t) &= \varphi(x,t) \psi(X,t) \nonumber \\
- &= \varphi_0(x,0)e^{-tiE_{0,x}^\prime} \psi(X,t).
-\end{align}
-where $E_{0,x}^\prime$ is the lowest eigenvalue of for the relative coordinate equation (\ref{RelativeMotionEquation}).
-
-We will use this solution to benchmark the more general many-body solvers. In particular, we can compute the expectation value 
-of the time-dependent Hamiltonian
-\begin{align}
- \braket{\Psi|\hat{H}|\Psi} &= \braket{\Psi|H_x|\Psi} + \braket{\Psi|H_X(t)|\Psi} \nonumber \\
- &= E_{0,x}^\prime + \braket{\Psi|H_X(t)|\Psi}.
-\end{align}
-
-\subsubsection{Two spatial dimensions}
-The two dimensional problem is almost a carbon copy of the one dimensional problem.
-We have that $\mathbf{R} = (X,Y)$ and we choose the polarization vector $\hat{\mathbf{e}}$ along the positive x-direction
-\begin{align*}
-    i \frac{\partial \psi(\mathbf{R},t)}{\partial t} &= \left( -\frac{1}{4}\nabla_\mathbf{R}^2 + \Omega^2 \mathbf{R}^2 + 2\mathbf{R} \cdot \hat{\mathbf{e}} \sin(\omega t) \right) \psi(\mathbf{R},t) \\
-    &= \frac{1}{2}\underbrace{\left( -\frac{1}{2}\nabla_\mathbf{R}^2 + \frac{1}{2}\Omega_R^2 \mathbf{R}^2 + 2X \sin(\omega t) \right)}_{\equiv H_\mathbf{R}} \psi(\mathbf{R},t)
-\end{align*}
-where we again have that $\Omega_R = 2\Omega$.
-
-Since we have chosen the polarization axis along the positive x-direction (or y-direction) the Hamiltonian separates
-\begin{align*}
-    H_\mathbf{R} &= \frac{1}{2}\left( -\frac{1}{2}\frac{\partial^2}{\partial X^2} + \frac{1}{2}\Omega_R^2 \mathbf{X}^2 + 2X \sin(\omega t) -\frac{1}{2}\frac{\partial^2}{\partial Y^2} + \frac{1}{2}\Omega_R^2 \mathbf{Y}^2 \right) \\
-    &= \frac{1}{2}(H_X(t) + H_Y).
-\end{align*}
-Note that $H_Y$ does not contain any explicit time-dependent terms.
-
-Then, the time-dependent Schrödinger equation for the center of mass motion admits solutions of the factorized form
-\begin{equation}
-    \psi(\mathbf{R},t) = \psi(X,Y,t) = \psi_X(X,t) \psi_Y(Y,t).
-\end{equation}
-We use the ground state of the unperturbed two-dimensional harmonic oscillator as intial condition
-\begin{align}
-    \psi(X,Y,0) = &\left( \frac{\Omega_R}{\pi} \right)^{1/4} \exp\left(-\frac{\Omega_R X^2}{2}\right) \nonumber \\ 
-    \times &\left( \frac{\Omega_R}{\pi} \right)^{1/4} \exp\left(-\frac{\Omega_R Y^2}{2}\right)
-\end{align}
-
-Insertion of this ansatz into the time-dependent Schrödinger equation yields two independent equations 
-\begin{align}
-    i \frac{\partial \psi_X}{\partial t} &= \frac{1}{2} H_X(t) \psi_X \\
-    i \frac{\partial \psi_Y}{\partial t} &= \frac{1}{2} H_Y \psi_Y. 
-\end{align}
-The first equation is now formally equivalent to the one-dimensional problem.
-
-The second equation has the solution of a unperturbed one-dimensional harmonic oscillator 
-\begin{equation}
-    \psi_Y(Y,t) = \psi(y) e^{-i \epsilon_{0,Y} t}
-\end{equation}
-where \begin{equation}
-    \epsilon_{0,y} = \frac{1}{2}\eta_Y
-\end{equation} 
- and 
-\begin{equation}
-    \eta_Y = \frac{1}{2}\Omega_R
-\end{equation}
-is the ground state energy of a harmonic oscillator with oscillator frequency $\Omega_R$.
-
-Finally, we have that the total wavefunction is given by 
-\begin{align}
- \Psi(\mathbf{r},\mathbf{R},t) &= \varphi(\mathbf{r},t) \psi(\mathbf{R},t) \nonumber \\
- &= \varphi(\mathbf{r},0)e^{-i\epsilon t} \psi_X(X,t) \psi_Y(Y,t).
-\end{align}
-where $\epsilon$ is the lowest eigenvalue of for the relative coordinate equation (\ref{RelativeMotionEquation}) in two spatial dimensions.
-
-In particular, we find that the expectation value of the time-dependent Hamiltonian is given by 
-\begin{align}
-    \braket{\Psi|\hat{H}|\Psi} &= \braket{\Psi|\hat{H}_\mathbf{r}|\Psi} + \braket{\Psi|H_\mathbf{R}|\Psi} \nonumber \\ 
-    &= \epsilon + \braket{\Psi|H_Y|\Psi} + \braket{\Psi|H_X(t)|\Psi} \nonumber \\
-    &= \epsilon + \epsilon_{0,y} + \braket{\Psi|H_X(t)|\Psi}.
-\end{align}
-
-\section{The harmonic potential theorem}
-The harmonic potential theorem: See Dobson\cite{HarmonicPotentialTheorem} and section 6.3.2 in the book by Ullrich\cite{TD_DFT_Ullrich_Book}.
-
-\end{appendices}
-\newpage
 \bibliographystyle{ieeetr}

### 2019-05-01 doc/PhysicsToday/PhysToday.tex
update

@@
 
-\author{D.M. Caballero}
-\email{haakoek@uio.no}
-\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
+\author{Daniel Marcos Caballero}
+\affiliation{Department of Physics and Astronomy, Michigan State University, East Lansing, Michigan 48824, USA}
 \affiliation{Department of Physics and Center for Computing in Science Education, University of Oslo, N-0316 Oslo, Norway}
+\author{Larry P. Engelhart}
+\affiliation{Department of Physics and Engineering, Francis Marion University, Florence, South Carolina 29502, USA}
+\author{Morten Hjorth-Jensen}
+\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, Michigan 48824, USA}
+\affiliation{Department of Physics and Center for Computing in Science Education, University of Oslo, N-0316 Oslo, Norway}
+\author{Marie Lopez del Puerto}
+\affiliation{Department of Physics, University of St. Thomas, St. Paul, Minnesota 55105, USA}
+\author{Kelly Roos}
+\affiliation{Department of Mechanical Engineering, Bradley University, Peoria, Illinois 61625, USA}
 
-\author{Morten Hjorth-Jensen}
-\email{hjensen@msu.edu}
-\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
-\affiliation{Department of Physics and Center for Computing in Science Education, University of Oslo, N-0316 Oslo, Norway}
 \date{\today}
@@
 \begin{abstract}
-    This is the abstract.
+
+Demands on computing in research continue to grow.  How can we
+integrate computing in core Physics courses in a coordinated and
+coherent way?  
+Computing competence represents a central element in
+scientific problem solving, from basic education and research to
+essentially almost all advanced problems in modern societies.
+
 \end{abstract}
@@
 
+\section{Introduction}
+
+Computational modeling and data analysis is core to a variety of
+non-academic career paths that are often pursued by physics students
+following the receipt of their Bachelor's degree.
+
+Computational modeling and data analysis is core to a variety of
+non-academic career paths that are often pursued by physics students
+following the receipt of their Bachelor's degree.
+
+A 2012 AIP survey \cite{AIPsurvery2012} of 5,000 recent physics Bachelor’s degree recipients
+indicated that 50$\%$ chose to enter the workforce immediately after
+receiving their degree.  Of these, roughly 75$\%$ go into STEM-related
+fields including engineering, software development, or information
+technology.
+
+Depending on sub-field, 60-80$\%$ of physics degree-holders in
+industry regularly need to engage in computer programming, simulation
+and modeling tasks.  Many respondents to the survey stressed the
+importance of computer skills, including programming, in increasing
+their marketability to potential employers.
+
+The use of computational modeling in the classroom setting provides
+students with insights that are complementary to those resulting from
+pencil-and-paper manipulation of equations.
+
+Computing in core physics courses allows us to bring important
+elements of scientific methods at a much earlier stage in our
+students' education.
+
+The ability to closely examine the behavior of systems that are too
+complex to be easily analytically tractable, or that have no analytic
+solutions (i.e., many systems of practical interest), helps to develop
+intuition that is unavailable to many students from analytic
+calculation.
+
+The growth in computer power over the past decades has radically
+changed science and its applications. Computational modeling, using
+computers and programming to solve, visualize, and explain phenomena, 
+is now an integrated and central part of research, development, and
+innovation. Yet, this change is sparsely reflected in most educations,
+even in science and engineering. Thus, there is a need to renew
+curricula and integrate computational modeling to ensure students are
+prepared for the 21st century workplace, a development that opens
+pedagogical challenges and opportunities. However, because the use of
+computational modeling is sparse, there is a near absence of research
+on students’ use of computational modeling in education. In addition,
+there is a growing need for students at all lveles and across disciplines for
+digital skills and computational literacy. This requires the infusion
+of aspects of computational and computer science into educational
+programs at all levels. Such a transformation of the contents of our
+education calls for new content in teacher education, new curricula
+both in computing and in disciplinary subjects with integrated
+computing, and research-based methods for instruction at all
+levels. The lack of such curricula, adequately trained teachers, and
+the supporting education research is a significant national and
+international challenge. In the US, this was recently addressed in a
+report on Computer Science education research \cite{USreport}. Similarly, in the
+European Union there is a significant focus on development of digital skills
+\cite{EUreport}.  Both reports signal an urgent
+need for curricular development, professional development programs,
+and education research on computational methods and their integration
+in various disciplines.
+
 \bibliographystyle{ieeetr}
+
 \bibliography{refs}
 \end{document}
+
+The case for improving U.S. Computer Science Education, see https://itif.org/publications/2016/05/31/case-improving-us-computer-science-education
+European Commission, New  report shows digital skills are required in all types of jobs, see https://ec.europa.eu/digital-single-market/en/news/new-report-shows-digital-skills-are-required-all-types-jobs.
+Association for Computing Machinery, 2013, see http://pathways.acm.org/executive-summary.html.
+Partnership for Integration of Computation into Undergraduate Physics, see https://www.compadre.org/PICUP/
+Marcos Daniel Caballero and Morten Hjorth-Jensen, Integrating a Computational Perspective in Physics Courses,  http://lanl.arxiv.org/abs/1802.08871, and Nova publishers, in press (2018); How to introduce computing in basic physics courses,  planned published as University Texts in Physics, Springer.

### 2019-06-24 doc/TimeDependence/Paper1/main.tex
added new version

@@
 \usepackage{bbold}
+\usepackage{physics}
+\usepackage{csquotes}
+\usepackage{commath}
 
@@
 
+%\tikzset{external/system call={latex \tikzexternalcheckshellescape -halt-on-error
+%-interaction=batchmode -jobname "\image" "\texsource" &&
+%dvips -o "\image".ps "\image".dvi}}
+
 \pgfplotsset{compat=1.15}
 
+\newcommand{\refslat}{\Phi}
+\newcommand{\oneten}{h}
+\newcommand{\twoten}{u}
 \renewcommand{\vec}[1]{\boldsymbol{#1}}
 \newcommand{\hamiltonian}{\hat{H}}
-\newcommand{\onehamil}{\hat{h}}
-\newcommand{\twohamil}{\hat{u}}
+\newcommand{\onehamil}{\hat{\oneten}}
+\newcommand{\twohamil}{\hat{\twoten}}
+\newcommand{\para}[1]{\left( #1 \right)}
+\newcommand{\brak}[1]{\left[ #1 \right]}
+\newcommand{\brac}[1]{\left\{ #1 \right\}}
+\newcommand{\ccr}[1]{\hat{c}_{#1}^{\dagger}}
+\newcommand{\can}[1]{\hat{c}_{#1}}
+\newcommand{\vac}{\text{vac}}
 
@@
 \newcommand{\polarization}{\vec{\epsilon}}
+\newcommand \mycomment[1]   {{\color{red} [{\it {#1}}]}}
+\newcommand{\todo}{$\square$}
 
@@
 \maketitle
-
-\section{TODO}
-\begin{itemize}
-    \item How do you treat the centre of mass problem???
-    \item The harmonic potential theorem\cite{HarmonicPotentialTheorem}: this depends on basis size!
-One way to check this is to let $u^{pq}_{rs} = 0$ and use a small number of basis functions.
-The short story: the particle density (in time) is rigidly translated when the confining potential is harmonic and the applied field is the simple dipole laser we use!
-    \item Furthermore, if I read correctly, we should observe that the (for example) the time dependent energy is effectively the same as for a single particle problem.
-
-    Quote from\cite{Chakraborty}: "In conclusion, the interaction of electrons in quantum
-dots leads to rich structure in their energy spectrum.
-However, FIR spectroscopy cannot probe it when the
-confining potential is quadratic because the optical excitations are then excitations of the cm and have exactly
-the same energies as single-electron excitations."
-
-    This is (again, if I understand the implications of the harmonic potential theorem correctly) due to the fact that the Hamiltonian separates into relative and center of mass coordinates with a harmonic potential and this type of applied field.
-This effect is independent of the number of particles.
-If we can actually demonstrate this, it is very solid foundation for claiming the correctness of the implementation.
-This also says something about the quality of the basis set.
-    \item Basically, what I am trying to say is that the "shape" of computed values (energy, overlap, expectation value of dipole, etc) is independent of the number of particles and will look like the corresponding values of the single particle problem apart from a scaling due to the number of particles.
-    \item Thus, it is maybe sensible to include comparisons with the single particle problem (which can be solved analytically or on a grid with f. ex RK4).
-\end{itemize}
 
@@
 \section{Theory}
+    \mycomment{I want to condense the theory section significantly and focus on the results.
+Much of the material can be referenced, however the time dependent coupled cluster part most likely needs some attention.}
+    \begin{itemize}
+        \item Suggested structure
+        \begin{itemize}
+            \item [\todo] Second quant/occ-number theory: Define notation for creation/annihilation ops, Slater dets, particle/hole indices.
+            \item[\todo] Groundstate methods: One compact subsection covering ground state methods with relevant reference(s).
+Morten et.
+al \cite{HjorthJensenQD_2017} covers all relevant methods.
+Make clear that they are used to generate the initial condition.
+            \item [\todo] Define how to compute expectation values using density matrices for all methods.
+            \item [\todo] Comment that spectrum is obtained Fourier transformation of time-dependent dipole moment after laser pulse/electric field is turned off.
+            \item [\todo] TDHF and TDFCI (extreme opposites in this context) covered in one subsection.
+Reference time dependent variational principle (both methods can be derived from this) and state the equations with relevant reference (Hochstuhl review).
+            \item [\todo] Pragmatic approach: define relevant quantities from TDCC but reference main body of theory.
+        \end{itemize}
+    \end{itemize}
+
+    \subsection{Second quantization/occupation-number theory}
     We assume that the reader is familiar with second-quantization/occupation-number theory.
 For a comprehensive introduction see for example \cite{szabo1996modern,helgaker2008molecular, shavitt2009many}.
-\subsection{Hartree-Fock Theory}
-In the Hartree-Fock approximation, the Hamiltonian reduces to a one-body term and a two-body term, written in the simplest way using second quantisation as
+
+    We will denote the general electronic Hamiltonian in its second quantized form relative to a given single-particle basis $\brac{\phi_p}_{p = 1}^{L}$ by
     \begin{equation}
-    \hat{H} = \hat{h} + \hat{u} 
-    = h^{p}_{q}c^\dagger_p c_q + \frac{1}{4}u^{pq}_{rs}c^\dagger_p c^\dagger_q c_s c_r
+        \hat{H} = \onehamil + \twohamil
+        = h^{p}_{q}\ccr{p} \can{q} + \frac{1}{4}u^{pq}_{rs}\ccr{p} \ccr{q} \can{s} \can{r}
+    \end{equation}
+    where $c_p^{\dagger},c_p$ denote creation and annihilation operators respectively.
+
+    The quantities $h^p_q$ and $u^{pq}_{rs}$ are referred to as matrix elements and are defined by
+    \begin{gather}
+        \oneten^{p}_{q} \equiv \bra{\phi_p}\onehamil\ket{\phi_q}, \\
+        \twoten^{pq}_{rs} \equiv \bra{\phi_p\phi_q}\twohamil\ket{\phi_r\phi_s}_{AS}.
+    \end{gather}
+    The subscript $AS$ denotes that $u^{pq}_{rs}$ is antisymmetrized according to
+    \begin{align}
+        \twoten^{pq}_{rs}
+        \equiv
+        \bra{\phi_p\phi_q}\twohamil\ket{\phi_r\phi_s}
+        -
+        \bra{\phi_p\phi_q}\twohamil\ket{\phi_s\phi_r}.
+    \end{align}
+    Typically $u^{pq}_{rs}$ is assumed to be antisymmetric and the subscript AS is omitted.
+
+    Einstein summation convention with summation over repeated indices is understood.
+    The position of the indices, i.e., raised or lowered, does not matter.
+
+    Furthermore we denote the reference state, also known as the \emph{Fermi vacuum}, by the single Slater determinant
+    \begin{align}
+        \ket{\refslat}
+        =
+        \prod_{i = 1}^{N}\ccr{i}\ket{\vac}
+        = \ket{\phi_1 \dots \phi_N},
+        \label{eq:ref-state}
+    \end{align}
+    where $\ket{\vac}$ is the vacuum state, and $N$ is the number of occupied single particle states.
+
+    The expectation value of the Hamiltonian with the reference state defines the so-called \emph{reference energy}
+    \begin{equation}
+        E_\text{ref} = \bra{\Phi}\hat{H}\ket{\Phi} 
+        = \oneten^{i}_{i} + \frac{1}{2} \twoten^{ij}_{ij}.
+        \label{eq:ref-energy}
     \end{equation}
 
-and the expected value of the Hamiltonian is given by
+    It is customary to use the indices $i, j, k, l, \dots$ as placeholders for the occupied states and $a, b, c, d, \dots $ for virtual states.
+A general single-particle state is denoted by $p, q, r, s, \dots$.
+
+    Using the creation and annihilation operators we can generate \emph{singly, doubly, triply,...} excited determinants defined by
+    \begin{align}
+        \ket{\Phi^a_i} &\equiv c_a^\dagger c_i \ket{\Phi} \\
+        \ket{\Phi^{ab}_{ij}} &\equiv c_a^\dagger c_i c_b^\dagger c_j \ket{\Phi} \\
+        &\vdots
+    \end{align}
+
+    Assuming that the single-particle functions are orthonormal, the reference state along with the excited determinants form an orthonormal basis for the many-body problem.
+Thus, in principle, any $N$-particle wavefunction can be expanded in the Slater determinants as
     \begin{equation}
-    \label{HF_expecation}
-    E[\Phi] = \bra{\Phi}\hat{H}\ket{\Phi} 
-    = \bra{\phi_i}\hat{h}\ket{\phi_i} + \frac{1}{2}\bra{\phi_i\phi_j}\hat{u}\ket{\phi_i\phi_j}_{\text{AS}}
-\end{equation}
-where the assumption is that the ground state can be represented by a single Slater determinant
-\begin{equation}
-    \ket{\Phi} = \prod_{i=1}^{N} c^\dagger_i \ket{\text{vac}} = \ket{\phi_1 \dots \phi_N}.
+        \ket{\Psi} = C_0 \ket{\Phi} + \sum_{ia} C^a_i \ket{\Phi^a_i} + \frac{1}{4} \sum_{ijab} C^{ab}_{ij} \ket{\Phi^{ab}_{ij}} + \cdots .
     \end{equation}
 
-It can be shown that a Slater determinant wavefunction $\ket{\Phi_\text{HF}} = \ket{\phi_i \dots \phi_N}$ is a variational minimum of the function for $E[\Phi]$ in \autoref{HF_expecation} if the single-particle functions $\ket{\phi_i}$ that make up this Slater determinant satisfy the canonical Hartree-Fock equations,
+    \subsection{Ground state methods}
+    Before we can do any time evolution, we need to define an initial condition/initial state.
+A common approach is to take ground state of the system under consideration as initial condition.
+Thus, we must first solve the time independent Schrödinger equation (TDSE)
+    \begin{equation}
+        \hat{H} \ket{\Psi} = E \ket{\Psi}
+    \end{equation}
+    for the ground state.
+
+    We will consider three of the most common many-body methods and their time dependent generalizations, namely the Hartree-Fock, coupled-cluster and configuration-interaction method.
+
+    \subsubsection{Hartree-Fock theory}
+    In Hartree-Fock theory the wavefunction is assumed to be described by the \emph{single} Slater determinant $\ket{\Phi}$ that minimizes the energy functional
+    \begin{equation}
+        E[\Phi] = \bra{\Phi}\hat{H}\ket{\Phi}.
+    \end{equation}
+    Minimization of $E[\Phi]$ subject to the constraint that the single-particle functions are orthonormal, leads to the Hartree-Fock equations 
     \begin{equation}
@@
     \end{equation}
-where $\hat{f}$ is a one-body operator called the Fock operator,
+    where $\hat{f}$ is the so-called the Fock operator,
     \begin{equation}
-    \hat{f}(\phi_1, \dots \phi_N) \equiv \hat{h} + \hat{v}_{\text{HF}},
+            \hat{f}(\phi_1, \dots \phi_N) \equiv \hat{h} + \hat{v}_{\text{HF}}.
     \end{equation}
-wherein the most important point of the Hartree-Fock approximation is revealed - one constrains the problem to a one-body problem where each particle only experiences a mean field from the other particles in the system.
-The canonical Hartree-Fock equations (\autoref{canonical_HF}) is reformulated as a matrix eigenvalue problem.
-This eigenvalue equation is called the Roothan-Hall equations,
+    The most important point of the Hartree-Fock approximation is that it constrains the problem to a one-body problem where each particle only experiences a mean field from the other particles in the system.
+
+    The single-particle functions satisfying eq.
+\ref{canonical_HF} are referred to as Hartree-Fock single-particle functions.
+In practice it is most common to expand these functions in some fixed basis $\{\chi_p\}_{p=1}^L$
     \begin{equation}
-    \label{roothan_hall}
-    F(U)U = U\epsilon,
+        \ket{\phi_i} = \sum_p U_{p i} \ket{\chi_p}.
     \end{equation}
-which is solved iteratively until self-consistency is reached, i.e. the next iteration does not contribute a significant amount to the mean field of the system.
+    Insertion of this expansion into eq. \ref{canonical_HF} leads to the eigenvalue equation known as the Roothan-Hall equation 
+    \begin{equation}
+        FU = U\mathcal{E} 
+    \end{equation}
+    where $F_{pq} = \bra{\chi_p}\hat{f}\ket{\chi_q}$ and $\mathcal{E} = \text{diag}(\epsilon_1,\cdots,\epsilon_L)$.
+
+    \subsubsection{Configuration Interaction Theory}
+
+    In configuration-interaction the wavefunction is taken as a linear combination of Slater determinant basis functions generated from a set of single-particle functions $\{\phi_p \}_{p=1}^L$
+    \begin{align}
+        \label{hierarchical_CI}
+        \ket{\Psi} &= C_0\ket{\Phi} + \sum_{ia}C^a_i\ket{\Phi^a_i} 
+        + \frac{1}{4}\sum_{ij} \sum_{ab} C^{ab}_{ij}\ket{\Phi^{ab}_{ij}} + \dots \nonumber \\
+        &= \sum_{I=1}^{N_\text{sd}} C_I \ket{\Phi_I}
+    \end{align}
+    where $\ket{\Phi_I}$ denotes a general Slater determinant.
+
+    One may choose to truncate the computation in the hierarchical manner (e.g.
+stopping at the dots in \autoref{hierarchical_CI} would be CISD) or include all Slater determinants generated by the first $L$ single-particle functions $\phi_1$ through $\phi_{L}$.
+This gives a space of dimension
+    \begin{equation}
+    N_\text{sd} = \binom{L}{N},    
+    \end{equation} and is called the full configuration interaction space (FCI space).
+
+    Minimization of the energy functional 
+    \begin{equation}
+        E[\Psi] = \bra{\Psi} \hat{H} \ket{\Psi}
+    \end{equation}
+    with respect to the expansion coefficients $C_I$ leads to the eigenvalue problem\cite{szabo1996modern}    
+    \begin{equation}
+        HC = C\mathcal{E}
+    \end{equation}
+    where $H_{IJ} = \bra{\Phi_I}\hat{H}\ket{\Phi_J}$ and $\mathcal{E}$ is a diagonal matrix containing the eigenvalues/eigenenergies.
+
+    In principle the FCI solution defines the eaxact solution in the computational space defined by the single particle functions $\{\phi_p \}$.
+However, due to the rapid growth of the number of Slater determinants, FCI calculations are possible only for (very) small systems.
+Nevertheless, FCI provides us with an invaluable way to benchmark other methods since it provides an exact solution.
+
+    One way to deal with the rapid growth of the Slater determinant space is to truncate the FCI expansion at, for example, the doubles level.
+This has the drawback that the resulting wavefunction no longer is size-consistent\cite{shavitt2009many}.
+
+    The solution to this problem is resolved by the coupled cluster method.
+
+    \subsubsection{Coupled cluster theory}
+    \mycomment{Den gamle CC seksjonen er flyttet til enden av dokumentet. Mista oversikten.}
+    \begin{itemize}
+        \item [\todo] Define exponential ansatz and cluster ops
+        \item [\todo] Define CC-energy and amplitude equations
+        \item [\todo] Possibly introduce bivar principle here, i.e CC-energy minimizes the CC-Lagrangian. Makes transition to time smoother (I think).
+        \item [\todo] Define $\lambda$-amplitudes and CC-expectation values/density operators
+        \item [\todo] Comment that CC is non-variational
+        \item [\todo] Comment that CC is size-consistent
+    \end{itemize}
+
+    \subsection{Time dependent methods}
+
+    In the following we give a brief review of the generalization of the Hartree-Fock, configuration-interaction and coupled-cluster methods to the time domain.
+A comprehensive review of time-depedent configuration interaction (TDCI) and time-dependent Hartree-Fock (TDHF) methods, including the so-called multiconfigurational time-dependent Hartree-Fock is given in the review by Hochstuhl et al.\cite{Hochstuhl2014}.
+For a detailed description of time-depedent coupled-cluster we refer to the works\cite{OATDCC_2012,Symplectic_TDCC_2018}.
+
+    Common to TDHF and TDFCI is that they can be derived from an underlying time-dependent variational principle (TDVP). Following Hochstuhl et al. the Lagrange formulation of the TDVP is given by
+    \begin{equation}
+        \delta \int_{t_0}^{t_1} L(t) = 0 \label{Lagrange TDVP}
+    \end{equation}
+    and the boundary conditions $\delta L(t_0) = \delta L(t_1) = 0$.
+    Here, $\delta$ is a variation in the wavefunction.
+    The Lagrangian $L(t)$ is given by
+    \begin{equation}
+        L(t) = \bra{\Psi(t)}\hat{H}(t)-i\partial_t\ket{\Psi(t)}
+    \end{equation}
+    Thus, we seek a wavefunction such that eq. \ref{Lagrange TDVP} is stationary with respect to all variations of the wavefunction paramters.
 
     \subsubsection{Time-Dependent Hartree-Fock}
+    In time-dependent Hartree-Fock the wavefunction is, as with the ground state Hartree-Fock method, assumed to be described by a single Slater determinant 
+    \begin{equation}
+        \ket{\Psi(t)} = \ket{\Phi(t)} \label{TDHF ansatz}
+    \end{equation}
+    where the single-particle functions now are assumed to be time-dependent.
 
-We now assume that the Hamiltonian is time-dependent, satisfying the time-dependent Schrödinger equation (\autoref{TDSE}),
+    The requirement that $\delta \int_{t_0}^{t_1} = 0$ with repsect to variations of the single-particle funtions leads to the TDHF equations\cite{Hochstuhl2014}
     \begin{equation}
-    \hat{H}(t) = \hat{h}(t) + \hat{u}(t).
+        i \frac{\partial}{\partial t} \ket{\phi_i(t)} = \hat{f}(t) \ket{\phi_i(t)},
+    \end{equation}
+    where $\hat{f}(t)$ is the Fock operator.
+
+    \subsubsection{Time-Dependent Configuration Interaction}
+    In complete analogy with the time-independent case, the TDCI wavefunction expanded in terms of Slater determinants,
+    \begin{equation}
+        \ket{\Psi(t)} = \sum_{I=1}^{N_\text{sd}} C_I(t) \ket{\Phi_I} \label{TDCI ansatz}
+    \end{equation}
+    where the expansions coefficients now are time-dependent.
+    We assume that the Slater determinants are orthogonal
+
+    Again, demanding that $\delta \int_{t_0}^{t_1} L(t) = 0$ with respect to variations of the expansion coefficients leads to the TDCI equations\cite{Hochstuhl2014}
+    \begin{equation}
+        i\frac{\partial}{\partial t} C(t) = H(t) C(t)
+    \end{equation}
+    where $H_{IJ}(t) = \bra{\Phi_I}\hat{H}(t) \ket{\Phi_J}$.
+
+\subsubsection{Time dependent coupled cluster}
+\mycomment{Den gamle TDCC seksjonen er flyttet til enden av dokumentet. Jeg mista oversikten og den ble unødvendig omfattende.}
+\begin{itemize}
+    \item [\todo] State time-depedent bivariational principle
+    \item [\todo] State $\tau,\lambda$-equations in time 
+    \item [\todo] Comment on generalization to time-dependent orbitals
+\end{itemize}
+
+\section{Systems/Quantum dots}
+    \mycomment{Motivate why we consider quantum dots}
+    \subsection{The quantum dot Hamiltonian}
+    The general Hamiltonian for interacting electrons in a (time-dependent) confining potential subject to an external electric field (e.g a laser pulse) is given by,
+    \begin{equation}
+        \begin{aligned}
+            \hat{H}(t) =& \frac{1}{2m^*} \sum_{i=1}^N (\vb{p}_i + e \vb{A}_i)^2
+                 +  \sum_{i=1}^N v_\text{conf}(\mathbf{r}_i) \\
+                 &+ \frac{1}{2}\frac{e^2}{4\pi\epsilon_r\epsilon_0} 
+                    \sum_{i \neq j} \frac{1}{\abs{\vb{r}_i - \vb{r}_j}}
+                + eE(t) \sum_{i=1}^N \vb{r}_i \cdot \hat{n}_{\text{polar}},
+        \end{aligned}
+    \end{equation}
+    where $E(t)$ is the electric field and $\hat{n}_{\text{polar}}$ a polarization vector.
+The parameters $m^*$ and $\epsilon_r$ denote the effective electron mass and relative permittivity respectively and are material dependent.
+
+    Typical parameters for the effective electron mass and relative permittivity used in the litterature are those of GaAs where $m^* = 0.067 m_e$ and $\epsilon_r = 12.4$ \mycomment{INSERT REF}.
+
+    Using $m^*$ and $\epsilon_r$ we define appropriate length, time and energy scales which is referred to as effective atomic units. The effective bohr radius is defined by 
+    \begin{equation}
+        a_0^* = \frac{4\pi \epsilon_0 \epsilon_r \hbar^2}{m^* e^2} = \frac{\epsilon_r m_e}{m^*} a_0
+    \end{equation}
+    the effective Hartree ($E_H^*$) by
+    \begin{equation}
+        1E_H^* = \frac{m^* e^4}{(4\pi \epsilon_0 \epsilon_r)^2 \hbar^2} = \frac{m^*}{\epsilon_r^2 m_e} E_H,
+    \end{equation}
+    and the effective atomic time unit by 
+    \begin{equation}
+        t_0^* = \frac{(4\pi \epsilon_0 \epsilon_r)^2 \hbar^3}{m^* e^4} = \frac{\epsilon_r^2 m_e}{m^*} t_0.
     \end{equation}
 
-In time dependent Hartree-Fockn(TDHF) the wavefunction is approximated by a single Slater determinant, $\ket{\Phi(t)}$, where the orbitals $\ket{\phi_p(t)}$ are time dependent.
-The equations of motion for the orbitals are found by demanding that the action functional
+    In the following discussion we will consider two kinds of 
+    confining potentials, the commonly used harmonic oscillator potential and a double well potential.
+
+    \subsection{Parabolic/Harmonic confinement}
+    If the confinement is quadratic/parabolic/harmonic
     \begin{equation}
-    \mathcal{S}[\ket{\phi_p(t)}] = \int_0^T \mathcal{L} dt 
+        v_\text{conf}(\mathbf{r}) = \frac{1}{2}m^* \omega_0 r^2
     \end{equation}
-is stationary with respect to variations of the orbitals, i.e $\delta \mathcal{S} = 0$. This is referred to as the Lagrange formulation of the time dependent variational principle\cite{Hochstuhl2014}.
+    the Hamiltonian can be separated in relative and center of mass coordinates. This has important implications for the properties of the solution(s) of both the stationary and dynamic problem.
 
-The Lagrangian $\mathcal{L}$ is given by 
+    In particular it can be shown that a parabolic quantum well absorbs far-infrared radiation (dipole approximation is valid) at the bare harmonic-oscillator frequency independent of electron-electron interaction and the number of electrons in the well \cite{Brey_Johnson_Halerpin, Taut_QD_arrays}.
+
+    Peeters\cite{Peeters_gen_Kohn_theorem} sums this up in what he refers to as the generalized Kohn theorem\cite{Kohn_theorem_1961}:
+
+    \begin{displayquote}
+    The resonance frequencies in the
+    magneto-optical absorption spectrum of a quantum dot
+    with parabolic confinement
+    are independent of the
+    electron-electron interaction and are given by the single
+    electron transition frequencies.
+    \end{displayquote}
+    Note that this statement is valid also when a static magnetic field is present.
+
+    This has the implication that electron-electron can not be seen and investigated by far infrared (FIR) spectroscopy \cite{Taut_Broken_Gen_Kohn_2001}.
+
+    In actual fact, it has been observed experimentally that the resonance frequencies of quantum dots are independent of the number of electrons
+    \cite{Sikorski_Merkt,Sikorski_Merkt_2}.
+
+    This serves as an important benchmark for our implementations.
+If we compute the spectrum (Fourier transform of dipole moment after laser pulse) we expect a single spectral line, at the harmonic oscillator frequency, independent of the number of particles.
+\mycomment{This gives a benchmark for $N > 2$}
+
+    \subsection{Double well}
     \begin{equation}
-    \mathcal{L} = \bra{\Phi(t)} \hat{H} - i\hbar \frac{\partial}{\partial t} \ket{\Phi(t)} 
-    - \lambda_{ij}(t)\left(\braket{\phi_i(t)|\phi_j(t)} - \delta_{ij}\right),
-\end{equation}
-where the Lagrange multipliers $\lambda_{ij}(t)$ enurses orthonormality of the orbitals.
-
-Computing the variations  with respect to the orbitals (see appendix \ref{TDHF_Derivation} for details) we obtain the TDHF equations 
-\begin{equation}
-    i \hbar \frac{\partial}{\partial t} \ket{\phi_p(t)} = \hat{f}(t) \ket{\phi_p(t)}.
+        v(x,y) = \frac{1}{2}m^* \omega_0^2 \left(x^2+y^2 -\alpha|x|+\frac{1}{4}\alpha^2 \right)
     \end{equation}
 
-\subsection{Configuration Interaction Theory}
+    \subsection{Electric field/Laser pulse}
+    \begin{itemize}
+        \item [\todo] Define sensible laser parameters, i.e. $\omega$ should correspond to far-infrared light and $\mathcal{E}_0$ should not be crazy-ass strong.
+        \item [\todo] The laser pulse model below is the most common I have seen, find a suitable reference.
+    \end{itemize}
+    \begin{equation}
+        E(t) = \mathcal{E}_0 \sin^2\left(  \frac{\pi(t-t_0)}{t_d} \right) \cos\left(\omega (t-t_0) \right)
+    \end{equation}
+    where $\mathcal{E}_0$ is the peak electric field strength, $t_d$ the pulse duration, $\omega$ is the laser/carrier frequency and $t_0$ is the start time.
+\section{Results \& Discussion}
 
-(This section is a bit informal, but at least there is something here.)
-
-Configuration Interaction relies on representing the wavefunction as an expansion of orthonormal slater determinants built from a set of orthonormal single-particle functions,
-\begin{equation}
-    \label{hierarchical_CI}
-    \ket{\Psi} = A_0\ket{\Phi} + \sum_{ia}A^a_i\ket{\Phi^a_i} 
-    + \frac{1}{4}\sum_{ij} \sum_{ab} A^{ab}_{ij}\ket{\Phi^{ab}_{ij}} + \dots
-\end{equation}
-One may choose to truncate the computation in the hierarchical manner (e.g.
-stopping at the dots in \autoref{hierarchical_CI} would be CISD) or include all Slater determinants generated by the first $L$ single-particle functions $\phi_0$ through $\phi_{L-1}$.
-This gives a space of dimension $\binom{L}{N}$, and is called the full configuration interaction space (FCI space).
-
-Inserting the CI wavefunction into the TISE yields
-\begin{align}
-    \sum_{K}\hat{H} A_{KJ} \ket{\Phi_K} &= \sum_{K} E_J A_{KJ} \ket{\Phi_K} \\
-    \sum_{K} A_{KJ} \bra{\Phi_I}\hat{H} \ket{\Phi_K} &= \sum_{K} E_J A_{KJ} \bra{\Phi_I}\Phi_K\rangle \\
-    HA &= AE,
-\end{align}
-where $H$ is a matrix reprentation of the Hamiltionian, and $A$ is a matrix containing the CI expansion coefficients.
-
-\subsubsection{Time-Dependent Configuration Interaction}
-Time-development is simply a matter of inserting one's CI expansion (\autoref{hierarchical_CI}) into the TDSE (\textcolor{blue}{Haakon:} formally we can use the time dependent variational principle, minimize/compute variations wrp.
-to CI coeffs and get the same equation),
-\begin{align}
-    i\hbar \sum_I \frac{\partial A_I(t)}{\partial t} \ket{\Phi_I} &= \sum_I A_i(t) \hat{H}(t) \ket{\Phi_I} \\
-    i \hbar \sum_I \frac{\partial A_I(t)}{\partial t} \bra{\Phi_J}\Phi_I\rangle &= \sum_I A_I(t) \bra{\Phi_J} \hat{H}(t) \ket{\Phi_i} \\
-    i\hbar \frac{\partial A_I(t)}{\partial t} &= H(t)A(t).
-\end{align}
-We find a relatively simple, yet large, set of equation on matrix form. One is now free to apply one's favourite numerical integration scheme in order to compute the time-development.
-
-\subsection{Coupled Cluster Theory}
-
-    Full Configuration Interaction (FCI) will provide you with an exact solution, but since the FCI space grows Pascal-triangularly (Question: Is the growth exponential when the growth does not follow an exponential function?) and because we are constrained by limited amount of computing power, FCI is not practically feasible.
-The Hartree-Fock method, on the other hand, approximates systems by ignoring electron-electron interaction and can therefore be quite inaccurate.
-Coupled Cluster to the rescue!
-Hurray.
-Coupled cluster cleverly include higher-order excitations and takes electron correlation effects into account.
-Time-developed coupled cluster (TDCC) can be tricky, however.
-
-    \subsubsection{Ground State coupled cluster}
-    In coupled cluster theory we employ an exponential ansatz for the $N$-particle correlated wavefunction 
-    \begin{equation}
-       \ket{\Psi_{\text{CC}}} = e^{\hat{T}}\ket{\Phi} 
-    \end{equation}
-    where $\ket{\Phi}$ is a reference state, typically the HF-state. The cluster operator $\hat{T} = \sum_{k=1}^N \hat{T}_k$ is a sum of $k$-particle $k$-hole excitation operators, $\hat{T}_k$, 
-    \begin{equation}
-        \hat{T}_k = \left(\frac{1}{k!}\right)^2 \sum_{abc\cdots ijk\cdots} \tau^{abc\cdots}_{ijk\cdots}   c_a^\dagger  c_i c_b^\dagger c_j  c_c^\dagger c_k \cdots 
-    \end{equation}
-    where the unknowns $\tau^{abc\cdots}_{ijk\cdots}$ are referred to as cluster amplitudes.
-The operators $\hat{T}_1, \hat{T}_2, \hat{T}_3, \cdots$ and corresponding amplitudes $\tau^a_i, \tau^{ab}_{ij}, \tau^{abc}_{ijk},\cdots$ are referred to as singly, doubly, triply-, etc excited operators and amplitudes respectively.
-
-    Insertion of the exponential ansatz into the time independent Schrödinger equation results in 
-    \begin{equation}
-        \hat{H}e^{\hat{T}} \ket{\Phi} = E e^{\hat{T}} \ket{\Phi}.
-    \end{equation}
-    Multiplying with $e^{-T}$ we have that 
-    \begin{equation}
-        \bar{H}_{CC} \ket{\Phi} = E \ket{\Phi} \label{SimTransSE},
-    \end{equation}
-    where we have defined the similarity transformed Hamiltonian 
-    \begin{equation}
-        \bar{H}_{CC} \equiv e^{-\hat{T}}\hat{H}e^{\hat{T}}.
-    \end{equation}
-
-    Projecting equation (\ref{SimTransSE}) with the reference or any excited determinant gives 
-    \begin{align}
-        E_{CC} &= \braket{\Phi|\bar{H}_{cc}|\Phi} \label{CCenergy} \\
-        0 &= \braket{\Phi^\mu|\bar{H}_{cc}|\Phi} \label{AmplitudeEqsCC},
-    \end{align}
-    where $\mu$ denotes a generic excitation.
-Eq.
-(\ref{CCenergy}) is the coupled cluster energy, while eq.
-(\ref{AmplitudeEqsCC}) defines a set of nonlinear equations for the unknowns cluster amplitudes.
-In the limit where $\hat{T}$ is not truncated eqs.
-(\ref{CCenergy}) and (\ref{AmplitudeEqsCC}) is equivalent to the full configuration-interaction method\cite{VarCC_2013}.
-
-    (BCH-expansion of $\bar{H}$)!
-
-    It is important to note that since $\hat{T} \neq \hat{T}^\dagger$, the CC-energy is non-variational, i.e 
-    \begin{equation}
-        E_{CC} = \braket{\Phi|e^{-T}\hat{H}e^{T}|\Phi} \neq \braket{\Phi|e^{T^\dagger}\hat{H}e^{T}|\Phi} 
-    \end{equation}
-    and is not guaranteed to give an upper bound on the energy when $\hat{T}$ is truncated.
-Working directly with the expression $e^{T^\dagger}\hat{H}e^{T}$ is impractical since it does not truncate after four nested commutators as opposed to $\bar{H}$.
-
-    Since CC is non-variational the question of how to compute expectation values other than the energy arises. The usual definiton of the expectation value of an operator, A, is given by 
-    \begin{equation}
-        \braket{A} = \braket{\Psi|\hat{A}|\Psi}.
-    \end{equation}
-    However, we have seen that $\ket{\Psi}^\dagger \neq \bra{\Phi}e^{-\hat{T}}$ such that $E_{CC} \neq \braket{H}$.
-
-    The solution to this problem was introduced independently by workers in quantum chemistry and in theoretical physics\cite{VarCC_2013}.
-
-    The idea is to introduce independent paramterizations of the ket- and bra state, $\bra{\tilde{\Psi}} \neq \ket{\Psi}^\dagger$. [Work in progress...]
-
-    \subsubsection{The time dependent bivariational principle}
-    (ASSUMING THE PREVIOUS SECTION IN PLACE)!
-
-    The extension of CC-theory to the time domain can be done through a generalization of the time dependent variational principle\cite{BECK2001}, to a corresponding time dependent bivariational principle\cite{OATDCC_2012}.
-
-    Consider the action-like functional 
-    \begin{align}
-    \mathcal{S}[\bra{\tilde{\Psi}}, \ket{\Psi}] &= \int_0^T \braket{\tilde{\Psi}(t)|i\frac{d}{dt}-\hat{H}(t)|\Psi(t)} dt  \nonumber \\ 
-    &= \int_0^T \braket{\tilde{\Phi}(t)|(\mathbb{1}+\hat{\Lambda}(t))e^{-\hat{T}(t)}\left(i\frac{d}{dt}-\hat{H}(t)\right) e^{\hat{T}(t)}|\Phi(t)} dt \nonumber 
-    \end{align}
-    where the Hamiltonian generally is time dependent. The integrand can be written more compactly as 
-    \begin{align}
-        &\braket{\tilde{\Phi}(t)|(\mathbb{1}+\hat{\Lambda}(t))e^{-\hat{T}(t)}\left(i\frac{d}{dt}-\hat{H}(t)\right) e^{\hat{T}(t)}|\Phi(t)} \nonumber \\ 
-        = \ \ & i \lambda_\mu \dot{\tau}^\mu + \rho^q_p \left( h^p_q - i \eta^p_q \right) + \frac{1}{4}u^{pr}_{qs}\rho^{qs}_{pr}  \\
-        =\ \ & i \lambda_\mu \dot{\tau}^\mu - \mathcal{E}_{\hat{H}-iD_0}[\lambda, \tau] 
-    \end{align}
-    where we have defined 
-    \begin{align}
-        D_0 &= \braket{\tilde{\varphi}_p|\dot{\varphi}_q} c_p^\dagger \tilde{c}_q \\
-        \eta^p_q &= \braket{\tilde{\varphi}_p|\dot{\varphi}_q}.
-    \end{align}
-
-    Thus, 
-    \begin{align}
-        \mathcal{S}[\bra{\tilde{\Psi}}, \ket{\Psi}] &= \int_0^T i \lambda_\mu \dot{\tau}^\mu + \rho^q_p \left( h^p_q - i \eta^p_q \right) + \frac{1}{4}u^{pr}_{qs}\rho^{qs}_{pr} dt \\
-        &= \int_0^T i \lambda_\mu \dot{\tau}^\mu - \mathcal{E}_{\hat{H}-iD_0}[\lambda, \tau] dt
-    \end{align}
-
-    The time dependent bivariational principle states that equations of motion are obtained by demanding that $\mathcal{S}$ is stationary ($\delta S = 0$) with respect to all independent variations of the parameters.
-Genraly both amplitudes and single-particle function/orbitals can be varied.
-    \subsubsection{Time dependent amplitude equations}
-    The equations of motions for the amplitudes read 
-    \begin{align}
-    i \dot{\tau}^\mu &= \frac{\partial}{\partial \lambda_\mu} \mathcal{E}_{\hat{H}-iD_0}[\lambda, \tau] \nonumber \\ 
-    &=  \braket{\tilde{\Phi}_\mu|e^{-\hat{T}}(\hat{H}-iD_0)e^{\hat{T}}|\Phi} \label{OATDCC_tau_equations} \\
--i \dot{\lambda}_\mu &= \frac{\partial}{\partial \tau^\mu} \mathcal{E}_{\hat{H}-iD_0}[\lambda, \tau] \nonumber \\ 
-&=  \braket{\tilde{\Phi}|(1+\Lambda)e^{-\hat{T}}[\hat{H}-iD_0,X_\mu]e^{\hat{T}}|\Phi} \label{OATDCC_lambda_equations}
-\end{align}
-
-The right-hand sides of equations [\ref{OATDCC_tau_equations}, \ref{OATDCC_lambda_equations}] are the same amplitude equations as in time independent cc-theory, plus an extra term due to the time dependence of the orbitals.
-
-If the orbitals are chosen to be stationary\cite{Symplectic_TDCC_2018}, we have $D_0 = 0$ and the time dependence is determined by the amplitudes.
-This implies a static reference such as the Hartree-Fock determinant.
-In that case, we also have that $\bra{\tilde{\Phi}} = \bra{\Phi} $, i.e an orthonormal basis.
-
-\subsubsection{Time dependent orbital equations}
-Computing the variations of $\mathcal{S}$ with respect to the orbitals, we obtain equations of motion for the time-dependent orbitals.
-The time derivatives of the orbitals can, in general, be split into a P-space and a Q-space,
-\begin{align}
-    \ket{\dot{\varphi}_q} &= (P+Q)\ket{\dot{\varphi}_q} = \sum_p \eta^p_q \ket{\varphi_p} + Q\ket{\dot{\varphi}_q} \\
-    \bra{\dot{\tilde{\varphi}}_q} &= \bra{\dot{\tilde{\varphi}}_q}(P+Q) = -\sum_q \eta^p_q \bra{\tilde{\varphi}_q} + \bra{\dot{\tilde{\varphi}}_p}Q
-\end{align}
-where we have defined 
-\begin{align}
-    P &= \sum_{p=1}^M \ket{\varphi_p}\bra{\tilde{\varphi}_p},  \ \ M \leq N_b \\
-    Q &= \mathbb{1}-P
-\end{align}
-with $N_b$ denoting the number of single-particle functions.
-
-This splitting is identical to what is done in MCTDHF-theory \cite{Hochstuhl2014}, where a trunctation of the $P$-space is used to reduce the number of Slater determinants the expansion of the CI-wavefunction.
-
-In coupled cluster methods the number of Slater determinants is reduced through the truncation of the cluster operators.
-Thus, we can choose to have $M=N_b$, resulting in $Q=0$.
-In the following we restrict our attention to the case where $Q=0$.
-
-Note that for the MCTDHF method this would be the same as FCI with time dependent orbitals and nothing is gained in terms of reducing computational complexity.
-
-With $Q=0$ the equations of motion for the orbitals is given by
-\begin{align}
-    \ket{\dot{\varphi}_q} &=  \sum_p \eta^p_q \ket{\varphi_p} \label{Reduced ket-orbital equation} \\
-    \bra{\dot{\tilde{\varphi}}_q} &=  -\sum_q \eta^p_q \bra{\tilde{\varphi}_q} \label{Reduced bra-orbital equation}. 
-\end{align}
-The non-zero components of $\eta^p_q$ are given by the solutions of the linear equations
-\begin{align}
-    i \sum_{bj} A^{ib}_{aj} \eta^j_b &= \sum_j \rho^i_j h^j_a - \sum_b \rho^b_a h^i_b \nonumber \\
- + &\frac{1}{2} \left[\sum_{prs} \rho^{is}_{pr} u^{pr}_{as} - \sum_{rqs} \rho^{qs}_{ar} u^{ir}_{qs}  \right] \\
-    -i \sum_{bj} A^{ja}_{bi} \eta^b_j &= \sum_b \rho^a_b h^b_i - \sum_j \rho^j_i h^a_j \nonumber \\
- + &\frac{1}{2} \left[ \sum_{prs} \rho^{as}_{pr} u^{pr}_{is} - \sum_{rqs} \rho^{qs}_{ir} u^{ar}_{qs} \right] \nonumber \\ 
- + &i \dot{\rho}_i^a \label{P-space bra equation}
-\end{align}
-where
-\begin{align}
-    A^{ib}_{aj} &= \braket{\tilde{\Psi}|[c_j^\dagger c_b,c_a^\dagger c_i]|\Psi} \nonumber \\
-    &= \delta^b_a \rho^i_j - \delta^i_j \rho^b_a \nonumber  \\ 
-    &= -A^{bi}_{ja}.
-\end{align}
-
-The simplest OATDCC case is the OATDCCD approximation where the cluster operators include double excitations only.
-One can show that the singles amplitudes are formally redundant \cite{OATDCC_2012}, leaving the one-body density matrix block diagonal, with only $\rho^i_j$ and $\rho^a_b$ being non-zero.
-Thus, the last term in eq.
-(\ref{P-space bra equation}) is zero.
-
-Furthermore, there is no contributions to the amplitude equations (\ref{OATDCC_tau_equations}) and (\ref{OATDCC_lambda_equations}) from $D_0$ in the doubles-only approximation.
-This follows from that only $\eta^a_i$ and $\eta^i_a$ are non-zero.
-
-\section{Systems}
-    Relevant articles:
     \begin{itemize}
-        \item Simple spin-dependent field TDHF study:\cite{TDHF_Singlet_Triplet}
-        \item One dimensional time-dependent quantum dot: \cite{Zanghellini04}
-        \item Harmonic single well: \cite{HjorthJensenQD_2017}
-        \item Theoretical study double well: \cite{Double_Well_2000}
-        \item Experiment with two fermions in double well: \cite{Fermions_Double_Well_Experiment}
-        \item Coupled double well: \cite{Coupled_double_wells}
-        \item Time domain two electrons in quantum dot: \cite{TimeDomainTwoElectronsQuantumDot}
-        \item Time-dependency with spin-orbit coupling: \cite{Spin_orbit_coupling}
-        \item Time-domain simulation of quantum spin: \cite{Time_domain_quantum_spin}
-        \item Entanglement in spin-qubit quantum dots: \cite{Entanglement_spin_qubit_quantum_dots}
+        \item Harmonic potential
+        \begin{itemize}
+            \item [\checkmark] Compare with Zanghellini study
+            \item [\todo] Compare single particle spectrum with $N \geq 2$.
+We could do $N=2,4,6$ for one spatial dimension (closed shell) and $N=2,6$ in two spatial dimensions.
+We expect a single spectral line at the HO-frequency independent of $N$.
+Possible noise due to the center of mass problem.
+            \item[\todo] With static magnetic field we can "manufacture" a shell structure that supports a $N=4$ closed shell system.
+The generalized Kohn theorem holds also with a static magnetic field (the same two spectral lines as in the single particle case), so in that case we could do $N=2,4,6$ in two spatial dimensions as well.
+Maybe this is better, since we then do not have to concern ourselves with 1d-model at all (apart from the Zanghellini benchmark).
+        \end{itemize}
+        \item Double well
+        \begin{itemize}
+            \item [\todo] Single particle spectrum
+            \item [\todo] Compute spectrum for $N=2,4,6$, can we "detect" many body effect compared with the single particle case?
+            \item[\todo] Compare TDHF and TDCCSD solutions. Expect "richer" result from TDCCSD.
+            \item [\todo] Convergence of dynamics as function of nr of basis functions? HO-functions centered at $x=y=0$ are not ideal for double well.
         \end{itemize}
 
-    Generic information on quantum dots and why we are interested in them.
-
-    General Hamiltonian for interacting electrons in a (time-dependent) confining potential influenced by an external electric field (laser),
-    \begin{align}
-    \hat{H}(t) = &\sum_{i=1}^N \left( - \frac{1}{2}\nabla^2 + \vconf(\mathbf{r}_i,t) + \vfield(\mathbf{r}_i,t) \right) \nonumber \\ 
-    + &\sum_{i < j}^N \frac{1}{r_{ij}}.
-    \label{eq:general_hamiltonian}
-    \end{align}
-
-    \subsection{One dimensional quantum dot}
-        We will use the one-dimensional quantum dot model considered by Zanghellini et. al in \cite{Zanghellini04}, as a benchmark/validation for our implementations.
-
-    \subsection{Two dimensional quantum dot}
-        For a two dimensional quantum dot model we will consider different confining potential.
-There already exists studies concerning the ground state properties the electrons are confined by the harmonic oscillator potential, see for example \cite{HjorthJensenQD_2017}.
-We will use this as a starting point for time-dependent studies.
-    \subsubsection{Single well}
-        The Hamiltonian of the two-dimensional harmonic oscillator quantum dot is given by \autoref{eq:general_hamiltonian}, with the confining potential
-        \begin{align}
-            \vconf(\boldsymbol{r}_i) \equiv \frac{1}{2} \omega^2 \boldsymbol{r}_i^2.
-        \end{align}
-    \subsubsection{Double well}
-    \subsection{Open system/finite potential wells}
-        Maybe too ambitious in this work.
-
-\section{Time-dependent fields}
-    \subsection{Dipole laser field}
-        The dipole laser field we will be looking at is given by
-        \begin{align}
-            \vfield(\boldsymbol{r}_i, t)
-            &= \envelope(t) \cdot \boldsymbol{r}_i,
-        \end{align}
-        where the \emph{laser envelope} $\envelope(t)$ is given by a time-varying field, e.g., purely sinusoidal with a constant amplitude as in the article by Zanghellini et.
-al \cite{Zanghellini04}, and with a polarization direction $\polarization$.
-        The dipole-part of the name of the laser stems from the position vector $\vec{r}_i$.
-    \subsection{Magnetic field}
-        Figure out how to model this field.
-
-        Hamiltonian for a single particle in a magnetic field \cite{TimeDomainTwoElectronsQuantumDot}
-\begin{equation}
-    \hat{H} = \frac{1}{2m} \left( \frac{\hbar}{i} \nabla - q \vec{A} \right)^2 + V(\vec{r})
-\end{equation}
-For a uniform magnetic field in the z-direction, we have 
-\begin{equation}
-    \vec{A} = \frac{B}{2} (-y,x,0) = B_0 (-y,x,0)
-\end{equation}
-Thus, 
-\begin{equation}
-    \hat{H} \psi(x,y) = \left(-\frac{\hbar^2}{2m} \nabla^2  - \frac{2q}{2m}\vec{A} \cdot \nabla  + \frac{q^2}{2m} A^2 + V(x,y) \right) \psi(x,y).
-\end{equation}
-Choosing atomic units ($\hbar = q = m = 1$), we have
-\begin{equation}
-    \hat{H} \psi(x,y) = \left(-\frac{1}{2} \nabla^2  -\vec{A} \cdot \nabla  + \frac{1}{2}A^2 + V(x,y) \right) \psi(x,y).
-\end{equation}
-
-\section{Results \& Discussion}
+    \end{itemize}
 
@@
         Comparison of the varying methods with the results and figures from Zanghellini et. al\cite{Zanghellini04}.
+        See \autoref{fig:1DZanghellini}.
 
@@
         \caption{\label{fig:1DZanghellini} Probability of being in the ground state
-        $|\langle\Psi(x,0\ket{\Psi(x,t)}|^2$ for one-dimensional quantum dot with frequency
-        $\omega = 8\Omega$ and two electrons. This plot is an exact match with figure 2 in Zanghellini
+        $|\langle\Psi(x,0\ket{\Psi(x,t)}|^2$ for one-dimensional quantum dot with laser frequency
+        $\omega = 8\Omega$, where $\Omega$ is the harmonic oscillator frequency. In this system there are $2$ particles and
+        20 basis sets / orbitals have been used. This plot is an exact match with figure 2 in Zanghellini
         et al.\cite{Zanghellini04}
@@
         Results from different interesting time-dependent observable quantities for distinct confining potentials and time-dependent fields (maybe too ambitious on different fields).
+
+        How many oscillator shells/HO-basis functions are needed for converged dynamics for $N=2$ and $N=6$? Start with TDHF.
+
+        \begin{figure}
+            \centering
+            \includegraphics[width=0.5\textwidth]{./figures/2Domega05.eps}
+            \caption{Probability of being in the ground state for two-dimensional
+            quantum dots for two and six particles. 
+            Potential frequency $\omega=0.5$, laser frequency is $8\omega = 4$}
+            \label{fig:2Doemga05}
+        \end{figure}
+
+        \begin{figure}
+            \centering
+            \includegraphics[width=0.5\textwidth]{./figures/2Domega10.eps}
+            \caption{Probability of being in the ground state for two-dimensional
+            quantum dots for two and six particles. 
+            Potential frequency $\omega=1.0$, laser frequency is $8\omega = 8$}
+            \label{fig:2Doemga10}
+        \end{figure}
+
+        \begin{figure}
+            \centering
+            \includegraphics[width=0.5\textwidth]{./figures/hf_vs_cc_n2.eps}
+            \caption{Time-dependent HF compared with time-dependent CC.
+                2D Quantum Dot, L=30, N=2, $\omega$=1.0.
+            }
+            \label{fig:hf_vs_cc_n2.eps}
+        \end{figure}
+
+        \begin{figure}
+            \centering
+            \includegraphics[width=0.5\textwidth]{./figures/hf_vs_cc_n6.eps}
+            \caption{Time-dependent HF compared with time-dependent CC.
+                2D Quantum Dot, L=30, N=6, $\omega$=0.5.
+            }
+            \label{fig:hf_vs_cc_n6}
+        \end{figure}
+
+        \begin{figure}
+            \centering
+            \includegraphics[width=0.5\textwidth]{./figures/cc_convergence.eps}
+            \caption{2D quantum dot with six particles for different number of
+            orbitals.}
+            \label{fig:my_label}
+        \end{figure}
 
@@
 \end{align}
-Since $H$ is independent of spin, the total wave function can be factorized as follows\cite{Taut93} 
+Since $H$ is independent of spin, the total wave function can be factorianged as follows\cite{Taut93} 
 \begin{equation}
@@
 
+% Note: This is an example of how to plot raw data files.
+% \begin{figure}
+%     \centering
+%     \begin{tikzpicture}
+%         \begin{axis}
+%             \addplot+[
+%                 scatter,
+%                 only marks,
+%             ]
+%             table {dat/foo.dat};
+%         \end{axis}
+%     \end{tikzpicture}
+%     \caption{Something new and fancy.}
+%     \label{fig:my_label_test}
+% \end{figure}
+
 \end{appendices}
@@
 \end{document}
+
+Old Text
+\begin{comment}
+    Relevant articles:
+    \begin{itemize}
+        \item Simple spin-dependent field TDHF study:\cite{TDHF_Singlet_Triplet}
+        \item One dimensional time-dependent quantum dot: \cite{Zanghellini04}
+        \item Harmonic single well: \cite{HjorthJensenQD_2017}
+        \item Theoretical study double well: \cite{Double_Well_2000}
+        \item Experiment with two fermions in double well: \cite{Fermions_Double_Well_Experiment}
+        \item Coupled double well: \cite{Coupled_double_wells}
+        \item Time domain two electrons in quantum dot: \cite{TimeDomainTwoElectronsQuantumDot}
+        \item Time-dependency with spin-orbit coupling: \cite{Spin_orbit_coupling}
+        \item Time-domain simulation of quantum spin: \cite{Time_domain_quantum_spin}
+        \item Entanglement in spin-qubit quantum dots: \cite{Entanglement_spin_qubit_quantum_dots}
+    \end{itemize}
+\end{comment}
+
+\begin{comment}
+\section{Time-dependent fields}
+    \subsection{Dipole laser field}
+        The dipole laser field we will be looking at is given by
+        \begin{align}
+            \vfield(\boldsymbol{r}_i, t)
+            &= \envelope(t) \cdot \boldsymbol{r}_i,
+        \end{align}
+        where the \emph{laser envelope} $\envelope(t)$ is given by a time-varying field, e.g., purely sinusoidal with a constant amplitude as in the article by Zanghellini et.
+al \cite{Zanghellini04}, and with a polarization direction $\polarization$.
+        The dipole-part of the name of the laser stems from the position vector $\vec{r}_i$.
+    \subsection{Magnetic field}
+        Figure out how to model this field.
+
+        Hamiltonian for a single particle in a magnetic field \cite{TimeDomainTwoElectronsQuantumDot}
+\begin{equation}
+    \hat{H} = \frac{1}{2m} \left( \frac{\hbar}{i} \nabla - q \vec{A} \right)^2 + V(\vec{r})
+\end{equation}
+For a uniform magnetic field in the z-direction, we have 
+\begin{equation}
+    \vec{A} = \frac{B}{2} (-y,x,0) = B_0 (-y,x,0)
+\end{equation}
+Thus, 
+\begin{equation}
+    \hat{H} \psi(x,y) = \left(-\frac{\hbar^2}{2m} \nabla^2  - \frac{2q}{2m}\vec{A} \cdot \nabla  + \frac{q^2}{2m} A^2 + V(x,y) \right) \psi(x,y).
+\end{equation}
+Choosing atomic units ($\hbar = q = m = 1$), we have
+\begin{equation}
+    \hat{H} \psi(x,y) = \left(-\frac{1}{2} \nabla^2  -\vec{A} \cdot \nabla  + \frac{1}{2}A^2 + V(x,y) \right) \psi(x,y).
+\end{equation}
+\end{comment}
+
+\subsection{One dimensional quantum dot}
+        We will use the one-dimensional quantum dot model considered by Zanghellini et. al in \cite{Zanghellini04}, as a benchmark/validation for our implementations.
+
+    \subsection{Two dimensional quantum dot}
+        For a two dimensional quantum dot model we will consider different confining potential.
+There already exists studies concerning the ground state properties the electrons are confined by the harmonic oscillator potential, see for example \cite{HjorthJensenQD_2017}.
+We will use this as a starting point for time-dependent studies.
+    \subsubsection{Single well}
+        The Hamiltonian of the two-dimensional harmonic oscillator quantum dot is given by \autoref{eq:general_hamiltonian}, with the confining potential
+        \begin{align}
+            \vconf(\boldsymbol{r}_i) \equiv \frac{1}{2} \omega^2 \boldsymbol{r}_i^2.
+        \end{align}
+    \subsubsection{Double well}
+
+\begin{comment}
+\section{TODO}
+\begin{itemize}
+    \item How do you treat the centre of mass problem???
+    \item The harmonic potential theorem\cite{HarmonicPotentialTheorem}: this depends on basis size!
+One way to check this is to let $u^{pq}_{rs} = 0$ and use a small number of basis functions.
+The short story: the particle density (in time) is rigidly translated when the confining potential is harmonic and the applied field is the simple dipole laser we use!
+    \item Furthermore, if I read correctly, we should observe that the (for example) the time dependent energy is effectively the same as for a single particle problem.
+
+    Quote from\cite{Chakraborty}: "In conclusion, the interaction of electrons in quantum
+dots leads to rich structure in their energy spectrum.
+However, FIR spectroscopy cannot probe it when the
+confining potential is quadratic because the optical excitations are then excitations of the cm and have exactly
+the same energies as single-electron excitations."
+
+    This is (again, if I understand the implications of the harmonic potential theorem correctly) due to the fact that the Hamiltonian separates into relative and center of mass coordinates with a harmonic potential and this type of applied field.
+This effect is independent of the number of particles.
+If we can actually demonstrate this, it is very solid foundation for claiming the correctness of the implementation.
+This also says something about the quality of the basis set.
+    \item Basically, what I am trying to say is that the "shape" of computed values (energy, overlap, expectation value of dipole, etc) is independent of the number of particles and will look like the corresponding values of the single particle problem apart from a scaling due to the number of particles.
+    \item Thus, it is maybe sensible to include comparisons with the single particle problem (which can be solved analytically or on a grid with f. ex RK4).
+\end{itemize}
+\end{comment}
+
+\section{Some ideas}
+List of ideas which is written down so that I dont forget it. Not everything here is intended for this work. 
+\begin{itemize}
+    \item Single particle problem, harmonic well: In dipole approx, state $n$ only couples with state $n-1$ and $n+1$ and one only get on spectral 
+    line (Fourier transform of time dep dipole moment after laser pulse), since $\Delta E_{n (n\pm1)} = |E_{n\pm1}-E_{n}| = \omega$.
+Is this also true when we have more than one particle?
+Can use FCI to check get excited states for $N=2$ to investigate this.
+    \item How does dynamic behavior (harmonic potential) depend on: a) The number of particles, b) The spatial dimension, i.e. 1d or 2d-dot?
+    \item TDHF vs TDCC vs TDFCI (and truncated TDCI models)
+    \item "Convergence" of dynamics as a function of the nr of basis functions. This is work in progress (Sebastian). 
+    \item Calibrate model parameters to experiments.
+    \item How does dynamic behavior depend on choice of potential compared with harmonic well? 
+    \begin{itemize}
+        \item [a)] Double dot potentials: This is already work in progress.
+        \item [b)] Gaussian potential (finite well). Do 1d case first for simplicity.
+        \item [c)] Repeat questions above regarding harmonic potential, how does dynamics depend on nr of particles and spatial dimension.
+    \end{itemize}
+    \item In what way is dynamics influenced by inclusion of a static magnetic field? This is already work in progress (Sebastian).
+    \item Spin dependent fields: If we start in, say, a singlet state ($S=0$) and the Hamiltonian is spin-independent, we can never make a transition to a state with a different spin symmetry.
+Simple dummy case is to make the "laser" act on only up or down spin, inducing transitions to other spin states.
+I have tried this on atomic/molecular systems (Håkon), and it works in the sense that I get transitions to other spin states.
+    \item Specific for double dot (possible Master thesis maybe).
+Develop a better basis set for double dots, not using just plain HO's.
+If we think of the bottom of the wells as "nuclei" we can do the quantum chemistry approach by placing basis functions on  each center, and maybe use plain HO for higher states (above double dot barrier).
+This requires overlap integrals.
+One idea is to fetch gaussian integrals from libint.
+Can be done on one-dimensional systems first.
+\end{itemize}
+
+\subsection{Coupled Cluster Theory}
+
+    Full Configuration Interaction (FCI) will provide you with an exact solution, but since the FCI space grows Pascal-triangularly (Question: Is the growth exponential when the growth does not follow an exponential function?) and because we are constrained by limited amount of computing power, FCI is not practically feasible.
+The Hartree-Fock method, on the other hand, approximates systems by ignoring electron-electron interaction and can therefore be quite inaccurate.
+Coupled Cluster to the rescue!
+Hurray.
+Coupled cluster cleverly include higher-order excitations and takes electron correlation effects into account.
+Time-developed coupled cluster (TDCC) can be tricky, however.
+
+    \subsubsection{Ground State coupled cluster}
+    In coupled cluster theory we employ an exponential ansatz for the $N$-particle correlated wavefunction 
+    \begin{equation}
+       \ket{\Psi_{\text{CC}}} = e^{\hat{T}}\ket{\Phi} 
+    \end{equation}
+    where $\ket{\Phi}$ is a reference state, typically the HF-state. The cluster operator $\hat{T} = \sum_{k=1}^N \hat{T}_k$ is a sum of $k$-particle $k$-hole excitation operators, $\hat{T}_k$, 
+    \begin{equation}
+        \hat{T}_k = \left(\frac{1}{k!}\right)^2 \sum_{abc\cdots ijk\cdots} \tau^{abc\cdots}_{ijk\cdots}   \ccr{a}  \can{i} \ccr{b} \can{j}  \ccr{c} \can{k} \cdots 
+    \end{equation}
+    where the unknowns $\tau^{abc\cdots}_{ijk\cdots}$ are referred to as cluster amplitudes.
+The operators $\hat{T}_1, \hat{T}_2, \hat{T}_3, \cdots$ and corresponding amplitudes $\tau^a_i, \tau^{ab}_{ij}, \tau^{abc}_{ijk},\cdots$ are referred to as singly, doubly, triply-, etc excited operators and amplitudes respectively.
+
+    Insertion of the exponential ansatz into the time independent Schrödinger equation results in 
+    \begin{equation}
+        \hat{H}e^{\hat{T}} \ket{\Phi} = E e^{\hat{T}} \ket{\Phi}.
+    \end{equation}
+    Multiplying with $e^{-T}$ we have that 
+    \begin{equation}
+        \bar{H}_{CC} \ket{\Phi} = E \ket{\Phi} \label{SimTransSE},
+    \end{equation}
+    where we have defined the similarity transformed Hamiltonian 
+    \begin{equation}
+        \bar{H}_{CC} \equiv e^{-\hat{T}}\hat{H}e^{\hat{T}}.
+    \end{equation}
+
+    Projecting equation (\ref{SimTransSE}) with the reference or any excited determinant gives 
+    \begin{align}
+        E_{CC} &= \braket{\Phi|\bar{H}_{cc}|\Phi} \label{CCenergy} \\
+        0 &= \braket{\Phi^\mu|\bar{H}_{cc}|\Phi} \label{AmplitudeEqsCC},
+    \end{align}
+    where $\mu$ denotes a generic excitation.
+Eq.
+(\ref{CCenergy}) is the coupled cluster energy, while eq.
+(\ref{AmplitudeEqsCC}) defines a set of nonlinear equations for the unknowns cluster amplitudes.
+In the limit where $\hat{T}$ is not truncated eqs.
+(\ref{CCenergy}) and (\ref{AmplitudeEqsCC}) is equivalent to the full configuration-interaction method\cite{VarCC_2013}.
+
+    (BCH-expansion of $\bar{H}$)!
+
+    It is important to note that since $\hat{T} \neq \hat{T}^\dagger$, the CC-energy is non-variational, i.e 
+    \begin{equation}
+        E_{CC} = \braket{\Phi|e^{-T}\hat{H}e^{T}|\Phi} \neq \braket{\Phi|e^{T^\dagger}\hat{H}e^{T}|\Phi} 
+    \end{equation}
+    and is not guaranteed to give an upper bound on the energy when $\hat{T}$ is truncated.
+Working directly with the expression $e^{T^\dagger}\hat{H}e^{T}$ is impractical since it does not truncate after four nested commutators as opposed to $\bar{H}$.
+
+    Since CC is non-variational the question of how to compute expectation values other than the energy arises. The usual definiton of the expectation value of an operator, A, is given by 
+    \begin{equation}
+        \braket{A} = \braket{\Psi|\hat{A}|\Psi}.
+    \end{equation}
+    However, we have seen that $\ket{\Psi}^\dagger \neq \bra{\Phi}e^{-\hat{T}}$ such that $E_{CC} \neq \braket{H}$.
+
+    The solution to this problem was introduced independently by workers in quantum chemistry and in theoretical physics\cite{VarCC_2013}.
+
+    The idea is to introduce independent paramterizations of the ket- and bra state, $\bra{\tilde{\Psi}} \neq \ket{\Psi}^\dagger$. [Work in progress...]
+
+    \subsubsection{The time dependent bivariational principle}
+    (ASSUMING THE PREVIOUS SECTION IN PLACE)!
+
+    The extension of CC-theory to the time domain can be done through a generalization of the time dependent variational principle\cite{BECK2001}, to a corresponding time dependent bivariational principle\cite{OATDCC_2012}.
+
+    Consider the action-like functional 
+    \begin{align}
+    \mathcal{S}[\bra{\tilde{\Psi}}, \ket{\Psi}] &= \int_0^T \braket{\tilde{\Psi}(t)|i\frac{d}{dt}-\hat{H}(t)|\Psi(t)} dt  \nonumber \\ 
+    &= \int_0^T \braket{\tilde{\Phi}(t)|(\mathbb{1}+\hat{\Lambda}(t))e^{-\hat{T}(t)}\left(i\frac{d}{dt}-\hat{H}(t)\right) e^{\hat{T}(t)}|\Phi(t)} dt \nonumber 
+    \end{align}
+    where the Hamiltonian generally is time dependent. The integrand can be written more compactly as 
+    \begin{align}
+        &\braket{\tilde{\Phi}(t)|(\mathbb{1}+\hat{\Lambda}(t))e^{-\hat{T}(t)}\left(i\frac{d}{dt}-\hat{H}(t)\right) e^{\hat{T}(t)}|\Phi(t)} \nonumber \\ 
+        = \ \ & i \lambda_\mu \dot{\tau}^\mu + \rho^q_p \left( h^p_q - i \eta^p_q \right) + \frac{1}{4}u^{pr}_{qs}\rho^{qs}_{pr}  \\
+        =\ \ & i \lambda_\mu \dot{\tau}^\mu - \mathcal{E}_{\hat{H}-iD_0}[\lambda, \tau] 
+    \end{align}
+    where we have defined 
+    \begin{align}
+        D_0 &= \braket{\tilde{\varphi}_p|\dot{\varphi}_q} \ccr{p} \tilde{c}_q \\
+        \eta^p_q &= \braket{\tilde{\varphi}_p|\dot{\varphi}_q}.
+    \end{align}
+
+    Thus, 
+    \begin{align}
+        \mathcal{S}[\bra{\tilde{\Psi}}, \ket{\Psi}] &= \int_0^T i \lambda_\mu \dot{\tau}^\mu + \rho^q_p \left( h^p_q - i \eta^p_q \right) + \frac{1}{4}u^{pr}_{qs}\rho^{qs}_{pr} dt \\
+        &= \int_0^T i \lambda_\mu \dot{\tau}^\mu - \mathcal{E}_{\hat{H}-iD_0}[\lambda, \tau] dt
+    \end{align}
+
+    The time dependent bivariational principle states that equations of motion are obtained by demanding that $\mathcal{S}$ is stationary ($\delta S = 0$) with respect to all independent variations of the parameters.
+Genraly both amplitudes and single-particle function/orbitals can be varied.
+    \subsubsection{Time dependent amplitude equations}
+    The equations of motions for the amplitudes read 
+    \begin{align}
+    i \dot{\tau}^\mu &= \frac{\partial}{\partial \lambda_\mu} \mathcal{E}_{\hat{H}-iD_0}[\lambda, \tau] \nonumber \\ 
+    &=  \braket{\tilde{\Phi}_\mu|e^{-\hat{T}}(\hat{H}-iD_0)e^{\hat{T}}|\Phi} \label{OATDCC_tau_equations} \\
+-i \dot{\lambda}_\mu &= \frac{\partial}{\partial \tau^\mu} \mathcal{E}_{\hat{H}-iD_0}[\lambda, \tau] \nonumber \\ 
+&=  \braket{\tilde{\Phi}|(1+\Lambda)e^{-\hat{T}}[\hat{H}-iD_0,X_\mu]e^{\hat{T}}|\Phi} \label{OATDCC_lambda_equations}
+\end{align}
+
+The right-hand sides of equations [\ref{OATDCC_tau_equations}, \ref{OATDCC_lambda_equations}] are the same amplitude equations as in time independent cc-theory, plus an extra term due to the time dependence of the orbitals.
+
+If the orbitals are chosen to be stationary\cite{Symplectic_TDCC_2018}, we have $D_0 = 0$ and the time dependence is determined by the amplitudes.
+This implies a static reference such as the Hartree-Fock determinant.
+In that case, we also have that $\bra{\tilde{\Phi}} = \bra{\Phi} $, i.e an orthonormal basis.
+
+\subsubsection{Time dependent orbital equations}
+Computing the variations of $\mathcal{S}$ with respect to the orbitals, we obtain equations of motion for the time-dependent orbitals.
+The time derivatives of the orbitals can, in general, be split into a P-space and a Q-space,
+\begin{align}
+    \ket{\dot{\varphi}_q} &= (P+Q)\ket{\dot{\varphi}_q} = \sum_p \eta^p_q \ket{\varphi_p} + Q\ket{\dot{\varphi}_q} \\
+    \bra{\dot{\tilde{\varphi}}_q} &= \bra{\dot{\tilde{\varphi}}_q}(P+Q) = -\sum_q \eta^p_q \bra{\tilde{\varphi}_q} + \bra{\dot{\tilde{\varphi}}_p}Q
+\end{align}
+where we have defined 
+\begin{align}
+    P &= \sum_{p=1}^M \ket{\varphi_p}\bra{\tilde{\varphi}_p},  \ \ M \leq N_b \\
+    Q &= \mathbb{1}-P
+\end{align}
+with $N_b$ denoting the number of single-particle functions.
+
+This splitting is identical to what is done in MCTDHF-theory \cite{Hochstuhl2014}, where a trunctation of the $P$-space is used to reduce the number of Slater determinants the expansion of the CI-wavefunction.
+
+In coupled cluster methods the number of Slater determinants is reduced through the truncation of the cluster operators.
+Thus, we can choose to have $M=N_b$, resulting in $Q=0$.
+In the following we restrict our attention to the case where $Q=0$.
+
+Note that for the MCTDHF method this would be the same as FCI with time dependent orbitals and nothing is gained in terms of reducing computational complexity.
+
+With $Q=0$ the equations of motion for the orbitals is given by
+\begin{align}
+    \ket{\dot{\varphi}_q} &=  \sum_p \eta^p_q \ket{\varphi_p} \label{Reduced ket-orbital equation} \\
+    \bra{\dot{\tilde{\varphi}}_q} &=  -\sum_q \eta^p_q \bra{\tilde{\varphi}_q} \label{Reduced bra-orbital equation}. 
+\end{align}
+The non-zero components of $\eta^p_q$ are given by the solutions of the linear equations
+\begin{align}
+    i \sum_{bj} A^{ib}_{aj} \eta^j_b &= \sum_j \rho^i_j h^j_a - \sum_b \rho^b_a h^i_b \nonumber \\
+ + &\frac{1}{2} \left[\sum_{prs} \rho^{is}_{pr} u^{pr}_{as} - \sum_{rqs} \rho^{qs}_{ar} u^{ir}_{qs}  \right] \\
+    -i \sum_{bj} A^{ja}_{bi} \eta^b_j &= \sum_b \rho^a_b h^b_i - \sum_j \rho^j_i h^a_j \nonumber \\
+ + &\frac{1}{2} \left[ \sum_{prs} \rho^{as}_{pr} u^{pr}_{is} - \sum_{rqs} \rho^{qs}_{ir} u^{ar}_{qs} \right] \nonumber \\ 
+ + &i \dot{\rho}_i^a \label{P-space bra equation}
+\end{align}
+where
+\begin{align}
+    A^{ib}_{aj} &= \braket{\tilde{\Psi}|[\ccr{j} \can{b},\ccr{a} \can{i}]|\Psi} \nonumber \\
+    &= \delta^b_a \rho^i_j - \delta^i_j \rho^b_a \nonumber  \\ 
+    &= -A^{bi}_{ja}.
+\end{align}
+
+The simplest OATDCC case is the OATDCCD approximation where the cluster operators include double excitations only.
+One can show that the singles amplitudes are formally redundant \cite{OATDCC_2012}, leaving the one-body density matrix block diagonal, with only $\rho^i_j$ and $\rho^a_b$ being non-zero.
+Thus, the last term in eq.
+(\ref{P-space bra equation}) is zero.
+
+Furthermore, there is no contributions to the amplitude equations (\ref{OATDCC_tau_equations}) and (\ref{OATDCC_lambda_equations}) from $D_0$ in the doubles-only approximation.
+This follows from that only $\eta^a_i$ and $\eta^i_a$ are non-zero.

### 2019-07-12 doc/MachineLearning/Survey/manybodyML.tex
added survey notes

\documentclass[aip,jcp,reprint,floatfix]{revtex4-1}

\usepackage{amsmath,amssymb}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{dcolumn}
\usepackage{bbold}
\usepackage{natbib}
\usepackage[normalem]{ulem} % to strike out plain text
\usepackage[colorlinks=true,citecolor={blue},urlcolor={blue}, linkcolor=blue]{hyperref}%
\usepackage[utf8]{inputenc}
\usepackage{amsthm}
\usepackage{braket}
\usepackage{verbatim}
\usepackage{mdframed}

\begin{document}

\title{Quantum Mechanical Many-body Problems with Machine Learning Algorithms}

\author{Scott Bogner}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}

\author{Morten Hjorth-Jensen}
\affiliation{National Superconducting Cyclotron Laboratory and Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\affiliation{Department of Physics and Center for Computing in Science Education, University of Oslo, N-0316 Oslo, Norway}

\maketitle

\section{Introduction and motivation}

Traditional many-body methods like full configuration interaction
theory (FCI), coupled-cluster (CC) theory, many-body perturbation theory (MBPT),
in-medium similarity renormalization group (IMSRG),
various Monte Carlo methods and many other many-body
approaches, have been rather successful in describing properties of
interacting many-body systems. These methods have been widely used in
condensed matter physics, nuclear physics, quantum chemistry and
materials science, just to mention a few of the areas of
applicability.

However, essentially all of these methods face what is
normally called the curse of dimensionality. For wave function based
method like FCI or CC theories where the original continuous problem
is discretized in terms selected basis functions and/or many-body
excitations, the dimensionality of the systems under study grows almost
exponentially when larger basis sets and/or number of particles are
included. As an example, nuclear physics 
systems close to the limits of stability pose in
particular a tough problem to many-body practitioners in terms of a dramaticn
increase of the number of relevant degrees of freedom. To describe say
weakly bound nuclei within the framework of a wave function based
approach, requires often a single-particle basis which includes bound,
weakly bound and unbound states, increasing thereby considerably the
number of many-body basis states. This renders often a standard
FCI calculation infeasible. Within the above mentioned
many-body methods there are however several interesting theoretical
approachs which attempt at circumventing the dimensionality
curse. Smarter basis sets is one of these approaches, as well as the
resummation of specific correlations and the recently proposed
stochastic sampling of many-body states in both FCI and CC
calculations.

Recently, several authors have pointed to  possibilities within the
broad fields of Machine Learning and Quantum Computing as approaches
that hold great promise in tackling the ever increasing
dimensionalities. There are several groups worldwide which now focus
on Machine Learning and/or Quantum Computing applied to many-body
problems.
Machine learning (ML) is an extremely rich field, in spite of its
young age. The increases we have seen during the last three decades in
computational capabilities have been followed by developments of
methods and techniques for analyzing and handling large date sets,
relying heavily on statistics, computer science and mathematics.  The
field is rather new and developing rapidly.

Machine Learning based methods offer several possibilities to
circumvent the abovementioned dimensionality problems, as well as
allowing us to model quantum mechanical systems with less a priori
knowlegde. For complex many-body systems like those which arise in
nuclear physics (in particular with the increase in the number of
degrees of freedom for nuclei close to their limits of stability).

We have recently started to explore several approaches based on deep
learning algorithms, with an emphasis on neural networks and so-called
Boltzmann machines, with several promising results for interacting
many-fermion systems, see some of the results below. Our applications
so far have been to systems of electrons confined to move in two or
three-dimensional regions (so-called quantum dots) and systems of
bosons (weakly and strongly interacting) using a mix of neural network
based algorithms and variational quantum Monte Carlo approaches.
Furthermore, we have also explored the solution of the Similarity
Renormalization Group set of equations using deep learning algorithms.

\section{Opportunities, challenges and new advances}

Neural-network quantum states (NQS) are a representation of the many-body
wave-function in terms of artificial neural networks (ANNs).
A commonly adopted choice is to parameterize
wave-function amplitudes as a feed-forward neural network.

One of the specific challenges emerging in the quantum domain is
imposing physical symmetries in the NQS representations. In the case
of a periodic arrangement of matter (often used in the simulation of
infinite systems like nuclear matter and the homogeneous electron
gas), spatial symmetries can be imposed using convolutional
architectures similar to what is used in image classification tasks.
While spatial symmetries have analogous counterparts in other ML
applications, satisfying more involved quantum symmetries often needs
a deep rethinking of ANN architectures.  The most notable case in this
sense is the \emph{exchange symmetry}.  For bosons, this amounts to
imposing the wave-function to be permutationally invariant with
respect to exchange of particle indices. The Bose-Hubbard model has
been adopted as a benchmark for ANN bosonic architectures, with
state-of-the-art results having been obtained.  The most challenging
symmetry is, however, certainly the fermionic one.  In this case, the
NQS representation needs to encode the antisymmetry of the
wave-function (exchanging two particle positions, for example, leads
to a minus sign). In this case, different approaches have been
explored, mostly expanding on existing variational ansatz for
fermions. We have for example explicitely included the fermion
antisymmetry by multiplying the NQS with an explicit Slater
determinant.  The situation for fermions is certainly the most
challenging for ML approaches at the moment, owing to the specific
nature of the symmetry.  On the applications side, NQS representations
have been used so-far along three main different research lines. These
are:

\begin{enumerate}
\item Representation of states: An active area of research concerns the general expressive power
of NQS, as also compared to other families of variational states.
Theoretical activity on the representation properties
of NQS seeks to understand how large, and how deep
should be neural networks describing interesting interacting quantum systems.
\item Learning from data: Parallel to the activity on understanding the theoretical properties of NQS,
a family of studies in this field is concerned with the
problem of understanding how hard it is, in practice, to learn
a quantum state from numerical data.
This can be realized using either synthetic data (for example coming from numerical simulations)
or directly from experiments.

This line of research has been explored in the supervised learning setting, to
understand how well NQS can represent states that are not easily expressed (in closed analytic form) as ANN.
The goal is then to train a NQS network to represent, as close as possible, a certain target state
whose amplitudes can be efficiently computed.
\item Finally, one of the main applications for the NQS representations is in the context
of variational approximations for many-body quantum problems.
The goal of these approaches is, for example, to approximately solve the Schr\"{o}dinger equation using a NQS representation for the wave-function.
In this case, the problem of finding the ground state of a given quantum Hamiltonian $H$ is formulated in variational terms as the problem of learning NQS weights $W$
minimizing $E(W)=\langle\Psi(W)|H|\Psi(W)\rangle/\langle\Psi(W)|\Psi(W)\rangle$. This is achieved using a learning scheme based on variational
Monte Carlo optimization.
Within this family of applications,
no external data representative of the quantum state is given, thus they typically demand a larger computational burden than supervised and unsupervised learning schemes for NQS.
\end{enumerate}

We have worked on the latter approach in connection with studies of
quantum dot systems (confined two-dimensional and three-dimensional
electron systems).  Using Quantum Monte Carlo samplings (QMC), a
practical issue often resulting is that of providing efficient
sampling schemes of high-dimensional spaces (path integrals,
perturbation series, etc..). This requires a careful tuning which is often
problem-dependent.  Devising general-purpose samplers for these
representations is therefore a particularly challenging
problem. Unsupervised ML methods can, however, be adopted as a tool to
speed-up Monte Carlo sampling for both classical and quantum
applications. In the figure here, we show the speed-up obtained when
using Restricted Boltzmann machines (RBM) for systems of
two-dimensional quantum dots (with $N=2,6,12,20,30,42,56,72$
electrons) compared with a Variational Monte Carlo calculation
(VMC). The results for the energies are similar. The main speed-up is
due to a simpler mathematical form for the correlated part of the
trial wave function. In the RBM calculation we used a simple Slater
determinant for the electrons using Hermite polynomials only. However,
the results for energies and other observables can be improved upon by
multiplying the NQS with a Jastrow factor (RBM+PJ/SJ). This leads
however to an increase in CPU time  since the number of parameters to optmize
increases. How to speed up the calculation of the Jastrow factor is
also an important issue.
\begin{figure}[hbtp]
        \centering
        \includegraphics[scale=0.5]{fig1.pdf}
        \caption{Speed-up obtained when
using Restricted Boltzmann machines (RBM) for systems of
two-dimensional quantum dots (with $N=2,6,12,20,30,42,56,72$
electrons) compared with a Variational Monte Carlo calculation
(VMC). The results for the energies are similar. The main speed-up is
due to a simpler mathematical form for the correlated part of the
trial wave function. In the RBM calculation we used a simple Slater
determinant for the electrons using Hermite polynomials only. However,
the results for energies and other observables can be improved upon by
multiplying the NQS with a Jastrow factor (RBM+PJ/SJ).}
\end{figure}

While exact for a large family of bosonic and fermionic systems, QMC techniques
typically incur in a severe sign problem when dealing with several
interesting fermionic models, as well as frustrated spin Hamiltonians.
In this case, it is tempting to use ML approaches to attempt a direct
or indirect reduction of the sign problem. While only in its first
stages, this family of applications has been used to infer information
about fermionic phases through hidden information in the Green's function.

Similarly, ML techniques can help reduce the burden of more subtle
manifestations of the sign problem in dynamical properties of quantum
models. In particular, the problem of reconstructing spectral functions
from imaginary-time correlations in imaginary time is also a field
in which ML can be used as an alternative to traditional maximum-entropy
techniques to perform analytical continuations of QMC data.

Applications of ML to quantum many-body problems have seen a fast-pace
progress in the past few years, touching a diverse selection of topics
ranging from numerical simulation to data analysis. The potential of
ML techniques has already surfaced in this context, already showing
improved performance with respect to existing techniques on selected
problems. To a large extent, however, the real power of ML techniques
in this domain has been only partially demonstrated, and several open
problems remain to be addressed.
In the context of variational studies with NQS, for example, the
origin of the empirical success obtained so far with different kind of
neural network quantum states is not equally well understood as for
other families of variational states, like tensor networks. Key open
challenges remain also with the representation and simulation of
fermionic systems, for which efficient neural-network representation
are still to be found.

Tensor-network representations for ML purposes, as well as
complex-valued networks like those used for NQS, play an important
role to bridge the field back to the arena of computer
science. Challenges for the future of this research direction consist
in effectively interfacing with the computer-science community, while
retaining the interests and the generality of the physics tools.

We have been able recently to study both interacting bosonic systems
and interacting fermionic systems (electrons confined to two- or
three-dimensional regions). We are planning to extend our applications to studies of infinite systems
like the homogeneous electron gas.  We plan also to address several of the
above issues in the next few years (a three to five year perspective).
The real challenge however is to extend these studies to nuclear
systems. Applying ML methods to nuclear systems based on the above
approaches is our central goal for the next five to ten years.

\end{document}

### 2019-07-12 doc/PathIntegral/CC-PathIntegral.tex
added path

\documentclass[12pt]{article}
\usepackage{geometry} % see geometry.pdf on how to lay out the page. There's lots.
\geometry{a4paper} % or letter or a5paper or ... etc
% \geometry{landscape} % rotated page geometry

% See the ``Article customise'' template for come common customisations

\title{Notes on formulating CC and IMSRG with auxiliary field path integrals}
\author{S.~K.~Bogner}
%\date{} % delete this line to display the current date

%%% BEGIN DOCUMENT
\begin{document}

\maketitle
%\tableofcontents
\section{Motivation}
Why bother?
\begin{itemize}
\item It might be fruitful to connect IMSRG/CC to functional RG approaches (e.g., can we view the SRG flow as a continuous sequence of field redefinitions in the path integral?)
\item Auxiliary fields might provide a path to extend CC/IMSRG to contain collective bosonic + fermionic DOF to handle single-particle and collective excitations on the same footing.
\item Can higher-body analogs of the Hubbard-Stratonovich transformation provide an effective way to deal with higher truncation levels in CC/IMSRG? I feel like the HS idea is reminiscent of tensor factorization, or the use of intermediates in CC-- integrating over the auxiliary field is effectively sewing the intermediates together to reconstruct the quartic (or higher) interactions. 
\item Can the arsenal of functional methods (e.g., stationary phase + loop corrections) in conjunction with the higher-body HS transformations be used to motivate/derive approximate triples, quadruples, etc.?
\item If we reformulate CC/IMSRG as a functional integral over auxiliary fields, maybe we can use stochastic methods to tractably evaluate higher truncation levels. 
\end{itemize}

\section{Auxiliary Field  Formulation for CCD }
Things are a bit simpler with CC theory, so let's start with the simplest non-trivial truncation level, CCD. Everything starts from the Hubbard-Stratonovich transformation, which is basically just a simple identity for shifted multidimensional gaussian integrals.

\begin{equation}
\label{eq:HS}
e^{-\frac{1}{4}\rho_n A_{nm}\rho_m} = \mathcal{N}\int \prod\bigl(d\sigma_n\bigr) e^{-\sigma_n A^{-1}_{nm}\sigma_m} e^{\sigma_n\rho_n}\,
\end{equation}
where the Einstein summation convention is implied and the normalization factor is given by

\begin{equation}
\mathcal{N}^{-1} = \int \prod\bigl(d\sigma_n\bigr) e^{-\sigma_n A^{-1}_{nm}\sigma_m}  = \sqrt{\frac{\pi^{N}}{\det{A} }}\,.
\end{equation}
Note that the above assumes a symmetric $N\times N$ matrix $A_{nm}=A_{mn}$ with positive-definite eigenvalues. [If the latter is violated, can we just add a large constant diagonal shift and subtract it away at the end?]

Now let's apply this to the simplest coupled cluster truncation level, CCD, where we have
\begin{equation}
\label{eq:CCD}
He^{T_2}|\phi\rangle = Ee^{T_2}|\phi\rangle\,,
\end{equation}
with the CCD correlated ground state is given by $|\psi_{CCD}\rangle = e^{T_2}|\phi\rangle$ and
\begin{equation}
T_2 = \frac{1}{4}\sum_{abij}t^{ab}_{ij}a^{\dagger}_aa_ia^{\dagger}_ba_j\equiv \frac{1}{4}\sum_{abij}t^{ab}_{ij}\hat{\rho}_{ai}\hat{\rho}_{bj}\,.
\end{equation}
Labeling pairs of particle-hole indices by a single index ($\alpha\equiv(ai)$, etc.) and suppressing indices whenever it's unambiguous, we see we can write the CCD correlated ground state using Eq.~\ref{eq:HS} as 
\begin{equation}
|\psi_{CCD}\rangle = e^{\frac{1}{4}\rho\cdot t\cdot\rho}|\phi\rangle=\int \mathcal{D}\sigma \,e^{-\sigma\cdot t^{\!-1}\!\cdot\sigma} e^{\sigma\cdot\rho}|\phi\rangle = \int \mathcal{D}\sigma \,e^{-\sigma\cdot t^{\!-1}\!\cdot\sigma}\,|\phi\{\sigma\}\rangle\,, 
\end{equation}
where I've used the Thouless theorem in the last step and defined the integration measure $\mathcal{D}\sigma \equiv \mathcal{N}\prod\bigl(d\sigma_n\bigr)$.

It's amusing that this way of writing the CCD wave function resembles the variational ansatz for the ground state wave function in the generator coordinate method (GCM). One easily sees with a couple lines of algebra that Eq.~\ref{eq:CCD} can likewise be cast as a generalized eigenvalue equation a-la the Hill-Wheeler equation
\begin{equation}
\int \mathcal{D}\sigma\, \mathcal{H}(\sigma',\sigma)f(\sigma) = E \int\mathcal{D}\sigma \mathcal{N}(\sigma',\sigma)f(\sigma)\,,
\end{equation}
where the norm-kernel $\mathcal{N}(\sigma',\sigma) \equiv \langle\phi\{\sigma'\}|\phi\{\sigma\}\rangle$, and the GCM ``wave function'' has the particular form $f(\sigma)\equiv e^{-\sigma\cdot t^{\!-1}\!\cdot\sigma}$.  This obviously isn't a practical way to solve the CCD equations-- if there are $n_pn_h$ different auxilliary fields $\sigma_{ai}$ (temporarily restoring explicit indices) and each is discretized with $N$ points, one gets a $N^{n_pn_h}$ dimensional generalized eigenvalue equation-- ouch. Nevertheless, it would be interesting if this connection can be exploited, perhaps in the reverse direction (i.e., mapping GCM calculations to a cheaper CCD-like formulation).

\subsection{Deriving the CCD energy and amplitude equations}
As a first check, let's try to derive the working CCD equations in the auxiliary field formulation.

\subsubsection{Energy equation}
Let's follow convention and start with the similarity-transformed CCD Hamiltonian.  The energy equation is
\begin{equation}
E_{CCD} = \langle\phi |e^{-T_2}He^{T_2}|\phi\rangle = \int \mathcal{D}\sigma'\int \mathcal{D}\sigma e^{-\sigma'\cdot t^{\!-1}\!\cdot\sigma'}\,e^{-\sigma\cdot t^{\!-1}\!\cdot\sigma}\langle\phi|e^{i\sigma'\cdot\rho}He^{\sigma\cdot\rho}|\phi\rangle\,.
\end{equation}
The ``$i$'' in the exponent is not a mistake-- it arises from when I use the HS transformation to write the inverse $e^{-T_2}$ transformation as
\begin{equation}
e^{-\frac{1}{4}\rho\cdot t \cdot \rho}=\int \mathcal{D}\sigma' e^{-\sigma'\cdot t^{\!-1}\!\cdot\sigma'}\,e^{i\sigma'\cdot\rho}\,.
\end{equation}
Using the Thouless theorem, $\langle\phi|e^{i\sigma'\cdot\rho}=\langle\phi|$, and that the integrals are normalized as $\int\mathcal{D}\sigma e^{-\sigma\cdot t^{\!-1}\!\cdot\sigma} = 1$ gives
\begin{equation}
E_{CCD} = \int \mathcal{D}\sigma e^{-\sigma\cdot t^{\!-1}\!\cdot\sigma}\langle\phi|H|\phi\{\sigma\}\rangle = E_{ref} + \int \mathcal{D}\sigma e^{-\sigma\cdot t^{\!-1}\!\cdot\sigma}\frac{\sigma_{ai}\sigma_{bj}}{2!}\langle\phi|H|\phi^{ab}_{ij}\rangle\,.
\end{equation}
In the last step I used that only even powers in $\sigma$ of the matrix element $\langle\phi|H|\phi\{\sigma\}\rangle$ contribute due to the gaussian weight.  Also, assuming $H$ has at most 2-body interactions, terms in the matrix element beyond quadratic order in $\sigma$ vanish.  Here and below, we'll make frequent use of
\begin{equation}
\langle \sigma_{ai}\sigma_{bj}\rangle \equiv \int \mathcal{D}\sigma e^{-\sigma\cdot t^{\!-1}\!\cdot\sigma}\sigma_{ai}\sigma_{bj} = \frac{1}{2}t^{ab}_{ij}\,,
\end{equation}
and Wick's theorem to compute gaussian averages of higher products of $\sigma$'s, e.g., 
\begin{equation}
\langle\sigma_{ai}\sigma_{bj}\sigma_{ck}\sigma_{dl}\rangle = \frac{1}{4}\bigl(t^{ab}_{ij}t^{cd}_{kl}+ t^{ad}_{il}t^{bc}_{jk} + t^{ac}_{ik}t^{bd}_{jl}\bigr)\,.
\end{equation}
Pulling it all together, we get the usual expression for the CCD energy:

\begin{equation}
E_{CCD} = E_{ref} + \frac{1}{4}t^{ab}_{ij}\langle ij||ab\rangle\,.
\end{equation}

\subsubsection{Amplitude equations}
The derivation of the CCD amplitude equation proceeds analogously, though with a few more steps. We start with
\begin{equation}
0 = \langle\phi^{ab}_{ij}|e^{-T_2}He^{T_2}|\phi\rangle = \int \mathcal{D}\sigma'\int\mathcal{D}\sigma e^{-\sigma'\cdot t^{\!-1}\!\cdot\sigma'} e^{-\sigma\cdot t^{\!-1}\!\cdot\sigma} \mathcal{M}^{ab}_{ij}(\sigma',\sigma)\,,
\end{equation}
where we've defined 
\begin{equation}
 \mathcal{M}^{ab}_{ij}(\sigma',\sigma)\equiv\langle\phi^{ab}_{ij}|e^{i\sigma'\cdot\rho}He^{\sigma\cdot\rho}|\phi\rangle\,.
\end{equation}
The general strategy then is to expand $\mathcal{M}^{ab}_{ij}(\sigma',\sigma)$ in powers of $\sigma',\sigma$, and make use of Wick's theorem with respect to the gaussian auxiliary field integrations. There's some flexibility in how you actually evaluate this. The way I proceed is to start by defining transformed creation/annihilation operators (note that the $a^{\dagger}_a$ and $a_i$ don't change under this transformation),
\begin{eqnarray}
\bar{a}^{\dagger}_i &=& e^{-i\sigma'\cdot\rho}a^{\dagger}_i e^{i\sigma'\cdot\rho} = a^{\dagger}_i - i \sigma'_{ai}a^{\dagger}_a \\
\bar{a}_a &=&  e^{-i\sigma'\cdot\rho}a_a e^{i\sigma'\cdot\rho}  = a_a + i\sigma'_{ai}a_i\,.
\end{eqnarray}
Then I invoke the Thouless theorem to give 
\begin{equation}
\mathcal{M}^{ab}_{ij}(\sigma',\sigma) = \langle\phi|\bar{a}^{\dagger}_i\bar{a}_a\bar{a}^{\dagger}_j\bar{a}_b H|\phi\{\sigma\}\rangle\,.
\end{equation}
This might seem like a mess of terms, but keep in mind that $\langle\phi|a^{\dagger}_a = 0 = \langle\phi|a_i $ and that terms with odd powers of $\sigma'$ integrate to zero. Things therefore simplify considerably to 
\begin{eqnarray}
\langle\phi|\bar{a}^{\dagger}_i\bar{a}_a\bar{a}^{\dagger}_j\bar{a}_b &=& \langle\phi|a^{\dagger}_i\bigl(a_a+i\sigma'_{ai'}a_{i'}\bigr)\bigl(a^{\dagger}_j - i\sigma'_{a'j}a^{\dagger}_{a'}\bigr)\bigl(a_b + i\sigma'_{bj'}a_{j'}\bigr)\\
&=& \langle\phi|\bigl(a^{\dagger}_ia_a + i\sigma'_{ai}\bigr)\bigl(a^{\dagger}_j - i\sigma'_{a'j}a^{\dagger}_{a'}\bigr)\bigl(a_b + i\sigma'_{bj'}a_{j'}\bigr)\\
&=& \langle\phi|\bigl(a^{\dagger}_ia_a + i\sigma'_{ai}\bigr)\bigl(a^{\dagger}_ja_b -i\sigma'_{a'j}a^{\dagger}_{a'}a_b + i \sigma'_{bj'}a^{\dagger}_ja_j' + \sigma'_{a'j}\sigma'_{bj'}a^{\dagger}_{a'}a_{j'}\bigr)\nonumber\\
&=&  \langle\phi|\bigl(a^{\dagger}_ia_aa^{\dagger}_ja_b + \sigma'_{a'j}\sigma'_{bj'}a^{\dagger}_ia_aa^{\dagger}_{a'}a_{j'} -\sigma'_{ai}\sigma'_{bj'}a^{\dagger}_{j}a_{j'}\bigr)\\
&=& \langle\phi|\bigl(a^{\dagger}_ia_aa^{\dagger}_ja_b +\sigma'_{aj}\sigma'_{bi}-\sigma'_{ai}\sigma'_{bj}\bigr)\,,
\end{eqnarray}
where I used Wick's theorem for the Fermionic operators in the last step. Therefore, the $\mathcal{M}(\sigma',\sigma)$ is given by
\begin{equation}
\mathcal{M}^{ab}_{ij}(\sigma',\sigma)= \langle\phi^{ab}_{ij}|H|\phi\{\sigma\}\rangle +\bigl(\sigma'_{aj}\sigma'_{bi} -\sigma'_{ai}\sigma'_{bj}\bigr)\langle\phi|H|\phi\{\sigma\}\rangle\,,
\end{equation}
and the $\mathcal{D}\sigma'$ integral is trivially done. The amplitude equations then become
\begin{equation}
\label{eq:AFamp}
0 = \langle\phi^{ab}_{ij}|e^{-T_2}He^{T_2}|\phi\rangle = \int \mathcal{D}\sigma e^{-\sigma\cdot t^{\!-1}\!\cdot\sigma}\biggl\{\langle\phi^{ab}_{ij}|H|\phi\{\sigma\}\rangle  - t^{ab}_{ij}\langle\phi|H|\phi\{\sigma\}\rangle\biggr\}\,.
\end{equation}
We can evaluate the above matrix elements by using the elegant generalized Wick's theorem between non-orthogonal vacua of Balian et al. (see Ch. 4 of Blaizot and Ripka), or we can proceed in a brute force manner by noting that only terms in $|\phi\{\sigma\}\rangle$ that involve even powers of $\sigma$ are non-zero after the gaussian integration. Furthermore, using that $H$ can at most change the excitation level by $\pm 2$, we have
\begin{eqnarray}
0 &=&  \int \mathcal{D}\sigma e^{-\sigma\cdot t^{\!-1}\!\cdot\sigma}\biggl\{\langle\phi^{ab}_{ij}|H|\phi\rangle + \frac{1}{2}\sigma_{a'i'}\sigma_{b'j'}\langle\phi^{ab}_{ij}|H|\phi^{a'b'}_{i'j'}\rangle+\frac{1}{24}\sigma_{a'i'}\sigma_{b'j'}\sigma_{c'k'}\sigma_{d'l'}\langle\phi^{ab}_{ij}|H|\phi^{a'b'c'd'}_{i'j'k'l'}\rangle \nonumber\\ 
&&\qquad-t^{ab}_{ij}\langle\phi|H|\phi\rangle - \frac{1}{2}t^{ab}_{ij}\sigma_{a'i'}\sigma_{b'j'}\langle\phi|H|\phi^{a'b'}_{i'j'}\rangle\biggr\}\,.
\end{eqnarray}
Applying Wick's theorem for the integral over $\mathcal{D}\sigma$, one has
\begin{eqnarray}
0 &=& \langle\phi^{ab}_{ij}|H|\phi\rangle + \frac{1}{4}t^{a'b'}_{i'j'}\langle\phi^{ab}_{ij}|H|\phi^{a'b'}_{i'j'}\rangle +\frac{1}{96}\bigl(t^{a'b'}_{i'j'}t^{c'd'}_{k'l'} + t^{a'd'}_{i'l'}t^{b'c'}_{j'k'}+t^{a'c'}_{i'k'}t^{b'd'}_{j'l'}\bigr)\langle\phi^{ab}_{ij}|H|\phi^{a'b'c'd'}_{i'j'k'l'}\rangle\nonumber\\
&&\qquad\qquad -t^{ab}_{ij}E_{ref} -\frac{1}{4}t^{ab}_{ij}t^{a'b'}_{i'j'}\langle\phi|H|\phi^{a'b'}_{i'j'}\rangle\\
&=& \langle ab||ij\rangle +  \frac{1}{4}t^{a'b'}_{i'j'}\langle\phi^{ab}_{ij}|H|\phi^{a'b'}_{i'j'}\rangle +\frac{1}{96}\bigl(t^{a'b'}_{i'j'}t^{c'd'}_{k'l'} + t^{a'd'}_{i'l'}t^{b'c'}_{j'k'}+t^{a'c'}_{i'k'}t^{b'd'}_{j'l'}\bigr)\langle\phi^{ab}_{ij}|H|\phi^{a'b'c'd'}_{i'j'k'l'}\rangle\nonumber\\
&&\qquad\qquad\qquad - t^{ab}_{ij}E_{CCD}\\
&=&\langle ab||ij\rangle +  \frac{1}{4}t^{a'b'}_{i'j'}\langle\phi^{ab}_{ij}|H|\phi^{a'b'}_{i'j'}\rangle +\frac{1}{32}t^{a'b'}_{i'j'}t^{c'd'}_{k'l'}\langle\phi^{ab}_{ij}|H|\phi^{a'b'c'd'}_{i'j'k'l'}\rangle-t^{ab}_{ij}E_{CCD}\,,
\end{eqnarray}
where we used that the 3 quadratic terms in the 2nd-to-last equation are equivalent to get to the final expression.

We could grind away and evaluate the matrix elements of $H$ between the different excited Slater determinants, but that is not necessary for our purposes. We recognize the last line to be the amplitude equation one gets when one starts from the non-similarity transformed form of the CCD equation and left projects onto $\langle\phi^{ab}_{ij}|$ (see e.g., Shavitt and Bartlett sec. 9.3.1),
\begin{equation}
\label{eq:nonST}
\langle\phi^{ab}_{ij}|He^{T_2}|\phi\rangle = E_{CCD}\langle\phi^{ab}_{ij}|e^{T_2}|\phi\rangle\,.
\end{equation}
Expanding the exponentials and only keeping the terms that make non-vanishing contributions, one gets
\begin{eqnarray}
\langle\phi^{ab}_{ij}|H|\phi\rangle + \frac{1}{4}t^{a'b'}_{i'j'}\langle\phi^{ab}_{ij}|H|\phi^{a'b'}_{i'j'}\rangle + \frac{1}{2\cdot 4^2}t^{a'b'}_{i'j'}t^{c'd'}_{k'l'}\langle\phi^{ab}_{ij}|H|\phi^{a'b'c'd'}_{i'j'k'l'}\rangle = E_{CCD}t^{ab}_{ij}\,,
\end{eqnarray}
which is clearly the same as what we get after doing our integrals over our auxiliary fields. In fact, a moment's thought reveals we did a lot of uncessary work-- starting from Eq.~\ref{eq:AFamp}, we can read the HS transformation in reverse and recognize that this is nothing else than Eq.~\ref{eq:nonST}.

In hindsight, this is sort of trivial. I started by applying a well-defined mathematical identity (the HS transformation for the exponential CCD operators), and then when I integrate over the auxilliary fields, I get back what I started from!  At the end of the day, one still has to evaluate matrix elements of H (written in terms of fermion operators) between excited Slater determinants to derive the final working amplitude equations. 
A couple speculative notes to come back to at some point:
\begin{itemize}
\item Maybe the discussion of Ch. 11 of Blaizot-Ripka, where Slater determinants are written in terms of bosonic coherent states and $H$ is mapped onto boson operators, might be worth playing around with.
\item Similarly, one might try breaking up the full $e^{T_2}=\lim_{N\rightarrow\infty}\{e^{T_2/N}\}^N$ and repeatedly inserting the resolution of the identify using Fermionic coherent states. Then one can apply the HS transformation for the quartic Grassman terms for each time slice, and finally we can ``integrate out'' the Fermionic variables a-la in Euclidean path integral formulations.
\end{itemize}

\section{Stochastic approaches}
One of the reasons for writing the CC equations as an integral over auxiliary fields is the hope that it might suggest a Monte Carlo approach. In contrast to the usual approaches, let's not work with the similarity-transformed $\bar{H}$, but instead start with the Schr\"odinger equation projected onto the various particle-hole sectors. (Is this what's called the ``projective'' approach?) Let's start off with the simplest truncation (CCD approximation) to get a feel for things.
\subsection{CCD}
We can write the energy and amplitude equations as
\begin{eqnarray}
E_{CCD} &=& \int \mathcal{D}\sigma e^{-\sigma\cdot t^{\!-1}\!\cdot\sigma} \langle\phi|H|\phi\{\sigma\}\rangle \\
t^{ab}_{ij} &=& \frac{1}{E_{CCD}}\int \mathcal{D}\sigma e^{-\sigma\cdot t^{\!-1}\!\cdot\sigma} \langle\phi^{ab}_{ij}|H|\phi\{\sigma\}\rangle\,.
\end{eqnarray}
We could then try to solve these equations iteratively:
\begin{enumerate}
\item Take an initial guess for $t$ (e.g., the 1st-order perturbative expression). 
\item Solve for $E_{CCD}$ and $t^{ab}_{ij}$ using Monte Carlo integration. 
\item Iterate
\end{enumerate}

\subsection{CCSD equations}
This proceeds analogously. We use that $[T_1,T_2]=0$, so the CCSD wave function takes the form
\begin{equation}
|\psi_{CCSD}\rangle = e^{T_1+T_2}|\phi\rangle = \int \mathcal{D}\sigma e^{-\sigma\cdot t_2^{\!-1}\cdot\sigma} |\phi\{\sigma + t_1\}\rangle\,.
\end{equation}
Then we have for the energy and amplitude equations
\begin{eqnarray}
E_{CCSD} &=& \langle\phi|He^{T_1+T_2}|\phi\rangle = \int \mathcal{D}\sigma e^{-\sigma\cdot t^{\!-1}\cdot\sigma}\langle\phi|H|\phi\{\sigma+t_1\}\rangle\\
t^{a}_i E_{CCSD}&=& \langle\phi^{a}_b|He^{T_1+T_2}|\phi\rangle = \int \mathcal{D}\sigma e^{-\sigma\cdot t^{\!-1}\cdot\sigma}\langle\phi^{a}_i|H|\phi\{\sigma+t_1\}\rangle\\
\bigl(t^{ab}_{ij}+\frac{1}{2}t^{a}_{i}t^{b}_{j}\bigr)E_{CCSD} &=& \langle\phi^{ab}_{ij}|He^{T_1+T_2}|\phi\rangle  = \int \mathcal{D}\sigma  e^{-\sigma\cdot t^{\!-1}\cdot\sigma}\langle\phi^{ab}_{ij}|H|\phi\{\sigma+t_1\}\rangle\nonumber\\
\end{eqnarray}

\begin{itemize}
\item Guess initial $t^{ab}_{ij}$ and $t^{a}_i$.
\item Evaluate integrals to get new $E_{CCSD}$, $t^{a}_i$, and $t^{ab}_{ij}$. 
\item Iterate
\end{itemize}

\subsection{CCDT equations}
Since I don't fully understand the higher-body generalizations of the HS transformation, let's try a ``mixed'' approach where we use auxiliary fields to handle the $e^{T_2}$ term, but then treat the $T_3$ term explicitly.  Starting from the Schr\"odinger equation w/a CCDT ansatz and using that $[T_2,T_3]=0$, we have for the triples amplitudes:
\begin{eqnarray}
\langle\phi^{abc}_{ijk}|He^{T_2+T_3}|\phi\rangle = \int \mathcal{D}\sigma e^{-\sigma\cdot t^{\!-1}\!\cdot\sigma} e^{T_3}\bar{H}_{T_3}|\phi\{\sigma\}\rangle \,,
\end{eqnarray}
where $\bar{H}_{T_3}\equiv e^{-T_3}He^{T_3}$. Then we have for the energy equation
\begin{equation}
E_{CCDT} = \int \mathcal{D}\sigma e^{-\sigma\cdot t^{\!-1}\!\cdot\sigma} \langle\phi|\bar{H}_{T_3}|\phi\{\sigma\}\rangle\,.
\end{equation}
This might be alarming since we know the form of the CC energy is always the same regardless of truncation level,
\begin{equation}
E_{} = E_{ref} + \frac{1}{4}t^{ab}_{ij}\langle ij||ab\rangle\,.
\end{equation}
But this discrepancy goes away when we realize that 
\begin{equation}
\langle \phi|\bar{H}_{T_3}|\phi\{\sigma\}\rangle= \langle \phi|He^{T_3}|\phi\{\sigma\}\rangle = \langle \phi|H|\phi\{\sigma\}\rangle\,,
\end{equation}
since $|\phi\{\sigma\}\rangle$ contains terms with $0,1,2,...$ particle-hole excitations, and therefore the $T_3$ term is irrelevant since the initial Hamiltonian can only change the excitation level by two if three-body interactions are ignored. Then this brings the expression back to the same form we had for CCD, so all is well.

Now onto the amplitude equations. For the double excitations,

\begin{equation}
t^{ab}_{ij} E_{CCDT} = \int \mathcal{D}\sigma e^{-\sigma\cdot t^{\!-1}\!\cdot\sigma} \langle\phi^{ab}_{ij}|He^{T_3}|\phi\{\sigma\}\rangle
\end{equation}
I don't know where I'm going with this. Let's shelve it and move on to the next idea.

\section{Approximations via loop expansions}
Path integrals are nice because they suggest novel approximation schemes. One such example is the approximate evaluation of the path integral via steepest descent (Euclidean/imaginary time) or stationary phase (Minkowski/real time), leading to so-called loop expansions. As discussed in Ch. 7 of Negele and Orland, these methods are powerful because they give a way to formulate a perturbative expansion around some non-trivial "reference" mean field solution, where the latter is itself non-perturbative in the bare coupling.

\end{document}
